### 2015-01-02 draft/providers.tex
Initial commit

\documentclass[10pt,a4paper]{article}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{multicol}
\usepackage{color}
\usepackage{verbatim}
\usepackage{semantic}
\usepackage{wasysym}
\usepackage{charter}
\usepackage{stmaryrd}
\usepackage{graphicx}

\newtheorem{definition}{Definition}

\newcommand{\kvd}[1]{\textbf{#1}}
\newcommand{\ident}[1]{\textit{#1}}
\newcommand{\val}[1]{\texttt{#1}}
\newcommand{\calv}{\mathcal{V}}
\newcommand{\calr}{\mathcal{R}}

\begin{document}

\title{Type providers}
\author{Tomas Petricek} 
\maketitle

% ==================================================================================================

\section{Introduction}
TBD
\begin{itemize}
\item \textbf{Information spaces} describes what?
\item \textbf{Schema} describes what?
\end{itemize}

\newpage
\section{Formal model}

\subsection{Information spaces}
The first problem in formalizing type providers is to find a formal model of \emph{information space}
that is simple and amenable to formal analysis, but powerful enough to model a wide range of data
sources accessible using type providers including hierarchical data (i.e.\ XML files), SQL databases, 
multi-dimensional data (such as the World Bank data set) etc.

We model information spaces simply as undirected graph with values as vertices and relations
between values as edges.

\begin{definition}
An \emph{information space} $(\calv, \calr)$ is a tuple of values $\calv$ and relations between
pairs of values $\calr \in \calv \times \calv$. The relations are undirected, meaning that 
$(v_1, v_2) \in \calr \Leftrightarrow (v_2, v_1) \in \calr$. 
\end{definition}

\subsection{Examples} 
To demonstrate that the simple definition is general enough, we look at two examples 
that demonstrate how databases and XML data can be viewed as information spaces.

\paragraph{Databases.} For simplicity, we only consider database with a simple table 
(Products) with two columns (Name and Price) where the first column is the primary key.
The mapping is demonstrated in Figure~\ref{fig:infospace-db}.

The set of values is represented as dots in the figure. It is obtained by taking all 
data and meta-data information about the database. Assuming that $P$ is a collection of
table data indexed by $I$ (in our example, pairs consisting of name and price), we 
construct the set of values by taking name of the table, names of columns and all values:

\[
\calv = \{ \val{"Products"}, \val{"Name"}, \val{"Price"} \} \cup \bigcup_{i\in I}  \{n_i | (n_1, \ldots, n_k) \in P \}
\]

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{figs/database.pdf}
\end{center}
\vspace{-1.5em}
\caption{Mapping database table to an information space}
\label{fig:infospace-db}
\end{figure}

\begin{figure}
\begin{center}
\includegraphics[scale=0.4]{figs/xml.pdf}
\end{center}
\vspace{-1.5em}
\caption{Stuff}
\label{fig:stuff2}
\end{figure}

\end{document}

### 2015-01-02 general/providers.tex
Initial commit

\documentclass[10pt]{sigplanconf}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{float}
\usepackage{pifont}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{verbatim}
\usepackage{alltt}

\newtheorem{definition}{Definition}

\newcommand{\kvd}[1]{\textbf{\sffamily #1}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\fqt}{\`{}\hspace{-0.25em}\`{}}

\newcommand{\val}[1]{\texttt{#1}}

\begin{document}

\toappear{Unpublished draft.}
\title{Type providers}
%\subtitle{\large Well-typed programs only go wrong when the schema changes.\\~}

\authorinfo
  {Authors}
  {Affliations}
  {emails@domain}

\maketitle

\begin{abstract}
~

What is the problem?

~

Why is it important?

~

What do we do?

~

What follows from that?

~
\end{abstract}

\section{Introduction}
Modern applications do not run in vacuum. They execute in \emph{information rich} environments,
access data from relational databases, external files and from increasingly important web-based 
services. Stable, organized \emph{information spaces} are now available via the web \cite{providers-techreport}.

Despite this, few strongly-typed programming languages and tools are able to seamlessly integrate 
external information sources as if they were strongly-typed components from a programmer's 
perspective.

The recent version of F\# introduced a \emph{type provider} mechanism \cite{providers-techreport} 
which makes it possible to view diverse external data sources as statically typed components. However,
this mechanism rises a number of concerns. Most importantly, if the type of the component depends on
a schema of external data source, the type safety of program is relativized with respect to possible
schema change of the data source.

~

TODO

\newpage

\begin{itemize}
\item Give a unified model of \emph{information spaces} using RDF triples and sketch how different 
  data sources (including relational databases, XML and WorldBank) can be mapped to the triple model. 
  We then distinguish between \emph{schema} and \emph{data} in the \emph{information space}.

\item We describe \emph{type-bridging mechanism} -- a formal model of F\# 3.0 type providers --
  that specifies how to map an information space into a programming language. The mechanism takes
  some information from the \emph{information space} as the \emph{schema} and based on the schema,
  maps the information space into types and values. The choice of schema is reflected in the 
  provided types.

\item We formalize the notion of \emph{relativized type safety} -- informally, a type safe program 
  using our type-bridging mechanism does not go wrong as long as the \emph{schema} in the 
  information space does not change.

\end{itemize}

\newpage

\section{F\# type providers}
F\# type providers \cite{providers-techreport} provide a bridging mechanism that exposes external
data sources as statically typed components. Technically, type providers are compile-time components
that provide the \emph{signature} (a static type) of the external component together with its 
\emph{implementation} (code that obtains data and exposes it in a format matching the signature).

The following example uses the Freebase \cite{providers-freebase} type provider to print a list of
computer scientists:
%
\begin{equation*}
\begin{array}{l}
\kvd{\#r}~\texttt{"DataStore.Freebase.dll"}\\[0.5em]
\kvd{let}~\ident{data}=\ident{DataStore.Freebase}()\\
\kvd{for}~\ident{c}~\kvd{in}~\ident{data.Science.\fqt Computer Scientists\fqt}~\kvd{do}\\
\quad\ident{printf}~\texttt{"\%s"}~\ident{c.Name}
\end{array}
\end{equation*}
%
The type provider is referenced on the first line. It provides a type \ident{DataStore.Freebase} that
is constructed on the first line. The provided type has a property \ident{Science} which returns an
instance of another type with a property \ident{\fqt Computer Scientists\fqt}. The property has a type
\ident{ComputerScientistsEntity~list} where \ident{ComputerScientistsEntity} is another provided type
with property \ident{Name} (among many others).

The provider also provides an implementation of the types and properties. Using a type erasure mechanism
(similar to the one used in Java Generics \cite{types-java-erasure}), the \ident{ComputerScientistsEntity}
type is erased to a general \ident{FreebaseEntity} type and the access to \ident{Name} property is
erased to a call \ident{c.GetStringValue(\texttt{"Name"})}.

\subsection{Questions}
The example rises a number of questions -- many of them are already answered in the F\# type 
providers technical report \cite{providers-techreport}. In this paper, we focus on the following two
themes.

\paragraph{Representation.} To implement a type provider (e.g. for the Freebase data source), the 
developer needs to write F\# code that connects to a data source and generates types based on the
information available.

This mechanism is powerful, but it is difficult to analyse formally, because a type provider may 
execute arbitrary code. Thus our first aim is to find a simpler model that is more amenable to 
formal analysis (and might be also used to simplify the implementation of certain type providers).

\paragraph{Type safety.} The above code is only correct as long as the Freebase schema does not change in 
certain ways. For example, renaming \ident{Name} property to \ident{\fqt Full name\fqt} would break
the above code, because the \ident{Name} is reflected at the type-level (and 
\ident{c.GetStringValue(\texttt{"Name"})} would fail because the value for the key
\texttt{"Name"} is not available).

On the other hand, changing the name of ``Tomas Petricek'' to ``Tom\'a\v{s} Pet\v{r}\'i\v{c}ek'' is 
not a problem, because the change is only made in part of information that is available at value 
level. After such change, the code will print different results, but it will still work.

Thus our second aim is to analyse the type provider and specify what schema changes are safe
and which are not. A type provider is type-safe with respect to a schema change if the schema
change does not break the provided types and the provided implementation.

\section{Information spaces}
The F\# type provider mechanism makes a reasonable assumption that there is no single universal
schema language for describing \emph{information spaces} -- and thus it allows arbitrary F\# code
in the type provider.

However, to model type providers formally, we need to find a model of \emph{information space} that 
is simple and amenable to formal analysis, but powerful enough to model a wide range of data
sources accessible using type providers including hierarchical data (i.e.\ XML files), SQL databases, 
multi-dimensional data (such as Freebase, the World Bank data set) etc.

\subsection{The RDF data model}
Our formal model of information spaces is based on the Resource Description Framework (RDF), which 
\emph{``is a language for representing information about resources in the World Wide Web''} \cite{semweb-rdf}.
RDF is based on the idea of modelling information using a graph. The nodes of the graph represent
resources (identified using an URI) or primitive values and the nodes represent properties.

\begin{figure}
\begin{center}
\includegraphics[scale=1]{figs/rdf.pdf}
\end{center}
\vspace{-1.5em}
\caption{Graph representation of data}
\label{fig:rdf-example}
\end{figure}

Figure~\ref{fig:rdf-example} shows a basic example adapted from \cite{semweb-rdf}. It shows a model
with three edges and two nodes. One edge represents a primitive value (a string) and two 
other edges represent URIs (unique identifier of a person and email). The nodes associate values
with the root edge and annotate them as \ident{fullName} and \ident{mailbox}. 
% The same dataset can be also stored using the following XML/RDF \cite{semweb-rdf-xml} notation:
%
% {\footnotesize \begin{verbatim}
% <rdf:RDF xmlns:rdf="http://w3.org/...rdf-syntax-ns#"  
%          xmlns:contact="http://w3.org/.../contact#">
%   <contact:Person 
%       rdf:about="http://w3.org/People/EM/contact#me">
%     <contact:fullName>Eric Miller</contact:fullName>
%     <contact:mailbox rdf:resource="mailto:em@w3.org"/>
%     <contact:personalTitle>Dr.</contact:personalTitle> 
%   </contact:Person>
% </rdf:RDF>
% \end{verbatim}
% }

The RDF model can be presented in form of \emph{triples} consisting of \emph{subject},
\emph{predicate} and \emph{object} (where subject and object are nodes and predicate
is an edge in the graph). For the purpose of this paper and our model of type providers, 
we use a simplified formal model based on triples:

\newcommand{\iprim}{\mathcal{P}}
\newcommand{\calv}{\mathcal{V}}
\newcommand{\calr}{\mathcal{R}}

\begin{definition}
Given a set of primitive values $\iprim$, a set of entity identifiers $?$ and a
set of relation identifiers $?$

an \emph{information space} $(\calv, \calr)$ is a triple of values $\calv$ and relations between
pairs of values $\calr \in \calv \times \calv$. The relations are undirected, meaning that 
$(v_1, v_2) \in \calr \Leftrightarrow (v_2, v_1) \in \calr$. 
\end{definition}

\subsection{Examples} 
The previous simple definition is general enough to model a number of real-world data sources.
We look at three examples showing how databases, XML and multi-dimensional data sets 
can be viewed as information spaces.

\paragraph{DB}

\paragraph{XML}

\newpage
~
\newpage

\bibliographystyle{abbrvnat}
\bibliography{providers}

\end{document}

### 2015-01-02 ml-2014-paper/paper.tex
Initial commit

\documentclass[submission,copyright,creativecommons]{eptcs}
\providecommand{\event}{ML 2014}

\usepackage{graphicx}
\usepackage{xcolor}
\usepackage[fleqn]{amsmath}
\usepackage[hang,flushmargin]{footmisc} 
\usepackage{graphics} 
%\usepackage{breakurl}

% =================================================================================================

\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}

\newcommand{\lang}{\begin{picture}(5,7)
\put(1.1,2.5){\rotatebox{45}{\line(1,0){6.0}}}
\put(1.1,2.5){\rotatebox{315}{\line(1,0){6.0}}}
\end{picture}}
\newcommand{\rang}{\begin{picture}(5,7)
\put(.1,2.5){\rotatebox{135}{\line(1,0){6.0}}}
\put(.1,2.5){\rotatebox{225}{\line(1,0){6.0}}}
\end{picture}}

\mathindent=1em

\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{strclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.0,0.0,0.0}

\newcommand{\ignp}{\_\kern-.5ex\_\kern-.5ex\_ }
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\ttfamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\lident}[1]{\textnormal{\sffamily 
  \`{}\hspace{-0.25em}\`{}\hspace{-0.1em}#1\`{}\hspace{-0.25em}\`{}}}
\newcommand{\cmt}[1]{\textit{\sffamily\textcolor{cmtclr}{#1}}}
\newcommand{\narrow}[1]{\hspace{-0.75em} #1 \hspace{-1.25em}~}

% =================================================================================================

\title{In the Age of Web:\\
  \textnormal{\LARGE Typed Functional-First Programming Revisited}}

\author{Tomas Petricek
\institute{University of Cambridge, UK}
\email{tomas@tomasp.net}
\and
Don Syme
\institute{Microsoft Research Cambridge, UK}
\email{don.syme@microsoft.com}
}
\def\titlerunning{In the Age of Web: Typed Functional-First Programming Revisited}
\def\authorrunning{T. Petricek, D. Syme}
\begin{document}
\maketitle

% =================================================================================================

\begin{abstract}
% What is the problem
Most programming languages were designed before the age of web.
% Why is this a problem
This matters because the web changes many assumptions that typed functional language designers take 
for granted. For example, programs do not run in a closed word, but must instead interact with 
(changing and likely unreliable) services and data sources, communication is often asynchronous 
or event-driven, and programs need to interoperate with untyped environments.

% What is my solution
In this paper, we present how F\# language and libraries face the challenges posed by the web.
Technically, this comprises using \emph{type providers} for integration with external information 
sources and for integration with untyped programming environments, using \emph{lightweight 
meta-programming} for targetting JavaScript and \emph{computation expressions} for writing 
asynchronous code.

% Why does the solution matter
In this inquiry, the holistic perspective is more important than each of the features in isolation.
We use a practical case study as a starting point and look how F\# language and libraries approach 
the challenges posed by the web. The specific lessons learned are perhaps less interesting than 
our attempt to uncover hidden assumptions that no longer hold in the age of web.
\end{abstract}

% =================================================================================================

\section{Introduction}

Among the ML family of languages, F\# often takes a pragmatic approach and emphasizes ease of use
and the ability to integrate with its execution environments\footnote{Historically, this applied 
to the .NET runtime, but the same applies to integrating with JavaScript in the web context.} over 
other aspects of language design. If you use the F\# langage as ML, you get most of the good 
well-known properties of ML\footnote{For example, F\# does not require type annotations when used 
as ML, but requires them when used with .NET objects.}. However, F\# leaves enough \emph{holes} that
let you use it \emph{not} as ML. This is the space that we explore in this paper.

This additional flexibility makes it possible to use F\# in ways that break the common assumptions
that are often taken for granted in languages such as ML and Haskell\footnote{In a way, we are trying
to uncover the hidden assumptions of the functional programming \emph{research programme} that are
normally ``\emph{rendered unfalsifiable by the methodological decisions of its protagonists}.'' 
(Lakatos \cite{philosophy-lakatos} as quoted by Chalmers \cite{philosophy-thing})}. The focus on 
the web directs our inquiry and provides an angle for reconsidering such assumptions.

Perhaps the most remarkable assumption is the idea that programs fundamentally operate in a closed
world. Although we have learned how to perform FFI and I/O \cite{haskell-ffi,haskell-imperative},
those are treated as dealing with the ``dirty real world''. For a practical solution, we argue that
we need to go much further, but deeper integration with (untyped) JavaScript libraries and (evolving) 
services inevitably breaks strict type safety requirements.

We do not claim that the F\# approach is the only possible. Rather, this paper should be seen
as a programming language experiment \cite{philosophy-pl} or an empirical observation of the 
approach used by the F\# community. We aim to provide an intriguing exploration of hidden assumptions 
and present what can be achieved using a combination of F\# features. We do so by starting with 
a simple, yet real-world problem and then exploring a solution. More specifically, the contributions 
of this position paper are:

\begin{itemize}
\item We present a case study (Section~\ref{sec:case}) showing how a combination of numerous
  F\# language features can be used for the development of modern web applications. This
  is not a toy demonstration, but an example of how F\# is used in the industry.

\item We discuss how type providers make it possible to access external information sources
  in web applications (Section~\ref{sec:tp-data}) and integration with (untyped) programming
  environments such as the JavaScript ecosystem (Section~\ref{sec:tp-lang}).

\item We show how F\# approaches the problem of compilation to JavaScript using a library 
  called FunScript (Section~\ref{sec:js-meta}), outlining important practical concerns such as 
  interoperability (Section~\ref{sec:js-lib}) and asynchronous execution (Section~\ref{sec:js-async}).

\item Throughout the paper, we discuss how the age of the web breaks the assumptions commonly 
  taken as granted in typed functional programming. We revisit the notion of type safety
  in the context of web (Section~\ref{sec:tp-relative}) and the notion of fixed language 
  semantics (Section~\ref{sec:js-relative}).
\end{itemize}

\noindent
In the first part of the paper (Section~\ref{sec:case}), we present a case study of using 
F\# for web development. The rest of the paper (Section~\ref{sec:tp} and \ref{sec:js}) discusses 
the arising issues in more depth. The source code and running demo for the case study is available 
at: \url{http://funscript.info/samples/worldbank}

% =================================================================================================

\section{Case Study: Web-based data analytics}
\label{sec:case}
In this case study, we develop a web application shown in Figure~\ref{fig:wb}, which lets the
user compare university enrolment in a number of selected countries and regions around the world. 
The resulting application runs on the client-side (as JavaScript) and fetches data dynamically 
from the World Bank \cite{data-wb-schter}.

The application is an example of a web page that could be built in the context of data journalism 
\cite{dj-handbook}. As such, it is relatively simple, works with just a single data source and 
uses a concrete indicator and a hard-coded list of countries \emph{i.e.}~to illustrate a point 
made in an accompanying article.

% -------------------------------------------------------------------------------------------------

\subsection{Accessing World Bank data with type providers}

To access the university enrollment information, we first obtain a list of countries using the 
World Bank type provider from the F\# Data library \cite{fsharp-data}. The type provider exposes 
the individual countries as members of an object (the notation \lident{Country Name} is used for 
identifiers with spaces):
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{WorldBank}~=~\ident{WorldBankData}\langl\ident{Asynchronous}=\kvd{true}\rangl
 \\[0.5em]
 \kvd{let}~\ident{data}~=~\ident{WorldBank.GetDataContext}() \\
 \kvd{let}~\ident{countries}~= \\
 \quad~\lbrack~~ \ident{data.Countries.}\lident{European Union} \\
 \qquad   \ident{data.Countries.}\lident{Czech Republic} \\
 \qquad   \ident{data.Countries.}\lident{United Kingdom} \\
 \qquad   \ident{data.Countries.}\lident{United States} ~\rbrack 
\end{array}
\end{equation*}
%
The type provider connects to the World Bank and obtains a lisf of countries at 
\emph{compile-time} and at \emph{edit-time} (when using auto-completion in an editor).
This means that the list is always up-to-date and we get a compile time error when accessing 
a country that no longer exists (a property discussed in Section~\ref{sec:tp-data}).

On the first line, we provide a static parameter \ident{Asynchronous}. Static parameters are 
resolved at compile-time (or edit-time). Here, we specify that the exposed types for accessing 
information should support only non-blocking functions. This is necessary for a web-based 
application, because JavaScript only supports non-blocking calls (using callbacks) to fetch the data.

% =================================================================================================

\begin{figure}[!t]
\hspace{3em} \includegraphics[width=35em]{worldbank.png}
\caption{Case study -- web application for comparing university enrollment in the world}
\label{fig:wb}
\end{figure}

% =================================================================================================

\subsection{Interoperating with JavaScript libraries}
To run the sample application on the client-side we use FunScript \cite{fsharp-funscript}, which 
is a library that translates F\# code to JavaScript (Section~\ref{sec:js}). Aside from 
running as JavaScript, we also want to use standard JavaScript libraries, including jQuery for DOM 
manipulation and Highcharts for charting. FunScript comes with a type provider that imports 
TypeScript \cite{ms-typescript} definitions for JavaScript libraries:
%
\begin{equation*}
\begin{array}{lll}
 \kvd{type}~\ident{j}~&\narrow{=}&~\ident{TypeScript}\langl\str{jquery.d.ts}\rangl \\
 \kvd{type}~\ident{h}~&\narrow{=}&~\ident{TypeScript}\langl\str{highcharts.d.ts}\rangl \\
\end{array}
\end{equation*}
%
The \textcolor{strclr}{\ttfamily d.ts} files are type annotations created for the TypeScript language.
Here, the type provider mechanism lets us leverage an existing effort for annotating common JavaScript 
libraries. The type provider analyses those definitions and maps them into F\# types named \ident{j} and 
\ident{h} that contain statically typed functions for calling the JavaScript libraries (we will use
them shortly). The file names are static parameters (same as \ident{Asynchronous} earlier) and are 
statically resolved and accessed at compile-time or edit-time.

Importing types for JavaScript libraries into the F\# type system has interesting implications, 
because the TypeScript language does not have the traditional type safety property \cite{ms-safets}.
We return to this topic in Section~\ref{sec:tp-lang}. Next, we generate checkboxes that appear on the 
right in Figure~\ref{fig:wb}:
%
\begin{equation*}
\begin{array}{lll}
 \kvd{let}~\ident{jQuery}~\ident{command}~=~\ident{j.jQuery.Invoke}(\ident{command}) 
 \\[0.5em]
 \kvd{let}~\ident{infos}~=~\ident{countries}~|\rang~\ident{List.map}~(\kvd{fun}~\ident{country}~\rightarrow \\
 \quad \kvd{let}~\ident{inp}~=~\ident{jQuery}(\str{<input>})\ident{.attr}(\str{type}, \str{checkbox}) \\
 \quad \ident{jQuery}(\str{\#panel})\ident{.append}(\ident{inp})\ident{.append}(\ident{country.Name}) \\
 \quad \ident{country.Name}, \ident{country.Indicators}, \ident{el}) \\
\end{array}
\end{equation*}
%
To manipulate the DOM (Document Object Model), we are using the jQuery library in a way that is
very similar to code that one would write in JavaScript. We define a helper function \ident{jQuery}
(hiding some of the complexities of the mapping) and use it to create the \str{<input>} element and
specify its attributes. Note that members like \ident{append} and \ident{attr} are standard 
jQuery patterns. The compiler sees them as ordinary object members. When writing code using F\# editors
based on the F\# Compiler Serivce \cite{fsharp-fcs}, they also appear in the auto-complete list.

Although the jQuery library is not perfect, it is a de facto standard in web development. The 
FunScript type provider makes it possible to integrate with it painlessly without explicitly 
specifying any FFI interface and without manual wrapping (see also Section~\ref{sec:js-lib}).

Note that we use a standard F\# function \ident{List.map} to iterate over the countries. This has 
a side-effect of creating the HTML elements, but it also returns a new list. The result is a list 
of $\ident{string}\ast\ident{Indicators}\ast\ident{jQuery}$ values representing the country name,
its indicators (for accessing the World Bank data) and the created DOM object representing the
checkbox.

% -------------------------------------------------------------------------------------------------

\subsection{Loading data and updating the user interface}

The main part of the sample program is a function \ident{render} that asynchronously fetches 
data for selected countries and generates a chart. To keep the code simple, we iterate over the 
\ident{infos} list from the previous section and load data for countries one by one:
%
\begin{equation*}
\begin{array}{lll}
 \kvd{let}~\ident{render}~()~=~\ident{async}~\{ \\
 \quad \kvd{let}~\ident{head}~=~\str{School enrollment, tertiary (\% gross)} \\
 \quad \kvd{let}~\ident{o}~=~\ident{h.HighchartsOptions}() \\
 \quad \ident{o.chart} \leftarrow \ident{h.HighchartsChartOptions}(\ident{renderTo}=\str{plc}) \\
 \quad \ident{o.title} \leftarrow \ident{h.HighchartsTitleOptions}(\ident{text}~=~\ident{head}) \\
 \quad \ident{o.series} \leftarrow \lbrack| ~ |\rbrack
 \\[0.5em]   
 \quad \kvd{for}~\ident{name},~\ident{ind},~\ident{check}~\kvd{in}~\ident{infos}~\kvd{do}\\
 \qquad   \kvd{if}~\ident{unbox}\langl \ident{bool} \rangl~(\ident{check.is}(\str{:checked}))~\kvd{then} \\
 \qquad\quad     \kvd{let!}~\ident{v}~=~\ident{ind.}\lident{School enrollment, tertiary (\% gross)} \\
 \qquad\quad     \kvd{let}~\ident{data}~=~\ident{vals}~
                   |\rang~\ident{Seq.map}~(\kvd{fun}~(\ident{k}, \ident{v})~\rightarrow \lbrack|~\ident{number~k};~\ident{number~v} ~|\rbrack)~
                   |\rang~\ident{Array.ofSeq} \\
 \qquad\quad     \ident{opts.series.push}(\ident{h.HighchartsSeriesOptions}(\ident{data}, \ident{name})) ~\}
\end{array}
\end{equation*}
%
Although the function looks like ordinary code, it is wrapped in the $\ident{async}~\{\ldots\}$ 
block, which is an F\# computation expression \cite{fsharp-zoo}. The F\# compiler performs 
de-sugaring similar to the CPS transformation and interprets keywods such as \kvd{let!} and 
\kvd{for} using special operations (monadic bind and other). The \ident{async} identifier 
determines that we are writing asynchronous workflow \cite{fsharp-async} that makes it 
possible to include non-blocking calls in the block.

Here, the non-blocking call is done when accessing the \lident{School enrollment, tertiary (\% gross)} 
indicator using the \kvd{let!} keyword. The indicator is a member (with a name wrapped in 
back-ticks to allow spaces) exposed as an asynchronous computation by the World Bank type provider.
The rest of the code is mostly dealing with the DOM and the Highcharts library using the API 
imported by FunScript -- we iterate over all checkboxes and generate a new chart series for each 
checked country.

Two notable points here are that \ident{async} translated to JavaScript is restricted to a single 
thread, which is not the case for ordinary F\# code (Section~\ref{sec:js-async}) and that the 
\ident{HighchartOptions} object preserves some of the underlying JavaScript semantics 
(Section~\ref{sec:js-lib}). Finally, the last part of the example code registers event handlers that 
redraw the chart when checkbox is clicked:
%
\begin{equation*}
\begin{array}{lll}
 \kvd{for}~\ignp, \ignp,~\ident{check}~\kvd{in}~\ident{infos}~\kvd{do} \\
 \quad \ident{check.click}(\kvd{fun}~\ignp~\rightarrow \ident{Async.StartImmediate}(\ident{render}()))
\end{array}
\end{equation*}
%
The \ident{click} operation (exposed by jQuery) takes a function that should be called when the 
event occurs. Calling it is a side-effectful operation that registers the handler. As \ident{render} 
is an asynchronous operation, we invoke it using the \ident{StartImmediate} primitive from the 
F\# library, which starts the computation without waiting for the result (the only way to start
a non-blocking operation in JavaScript).

% -------------------------------------------------------------------------------------------------

\subsection{Larning from the case study}

The case study shows that we can develop a simple interactive data visualization (that could
be built, for example, by data journalists) in less than 30 lines of F\# code. The code uses 
many typical functional patterns (lists, first-class functions, data types), but also uses features
that are more specific to F\# (type providers, objects, computation expressions).

Before analysing the interesting aspects of this case study, we briefly review the points that we 
find appealing and points that many would find unappealing or, at least, peculiar. First, the 
appealing points:

\begin{itemize}
\item The ML approach to types and type inference can be extended from (closed-world) data types 
  to (open-world) types for rich information sources such as World Bank. The sample code is 
  fully statically typed without explicit type annotations. Critically, types are also used for 
  exploratory programming when finding indicators using auto-complete in an editor.

\item The case study demonstrates that core ML programming style can be used in the context of
  client-side (JavaScript) web development. We used functional lists, standard higher-order 
  functions such as \ident{List.map} in much the same way as when writing ordinary F\#.

\item In addition to standard functional constructs, we were also able to reuse F\# asynchronous
  workflows to write non-blocking code that requests data from a web service (World Bank), 
  rather than using error-prone explicit callbacks that are common in JavaScript.

\item Finally, we were able to painlessly call Highcharts and jQuery. No explicit wrapping or 
  importing of individual functions and types was necessary. Moreover, despite the differences between 
  the F\# and JavaScript object model, the code is close to idiomatic F\#.
\end{itemize}

\noindent
Now, the following list looks at the aspects that appear unappealing or peculiar, especially when 
coming from the traditional functional programming background:

\begin{itemize}
\item The World Bank type provider lifts information about countries to the type level. As a result,
  we can easily write $\ident{data.Countries.}\lident{Czech Republic}$, but if Czech Republic is
  removed from the World Bank (and becomes Czechoslovakia again), the code will no longer compile
  (Section~\ref{sec:tp-data}).

\item The TypeScript language is unsound due to covariant generics \cite{ms-safets}. Thus
  importing types from TypeScript definitions introduces a potential unsoundness into the F\# 
  code (Section~\ref{sec:tp-lang}).

\item When compiling F\# to JavaScript, the FunScript library does not fully preserve the 
  semantics of F\#. For example, numerical types behave as in JavaScript (Section~\ref{sec:js-meta})
  and asynchronous workflows run on a single thread (Section~\ref{sec:js-async}).
\end{itemize}

\noindent
The most notable observation about the above points is that there is often both a positive and
a negative side: we can nicely access World Bank data, but it affects soundness properties;
we can interoperate with JavaScript libraries, but we can not fully hide undesirable JavaScript 
behaviours.

The aim of this paper is not to make value judgements and argue what is better. Using the case
study as a basis, we claim that the outlined approach is \emph{one possible} and that it
\emph{works in practice}. In the rest of the paper, we give more details about the most important 
aspects of the approach and discuss alternatives`'.

% =================================================================================================

\section{Type providers}
\label{sec:tp}
a \\
b \\
c \\
~

\subsection{Integrating data}
\label{sec:tp-data}

~

\subsection{Integrating languages}
\label{sec:tp-lang}

~

\subsection{Relativized type safety}
\label{sec:tp-relative}

~

% =================================================================================================

\section{Compiling to JavaScript}
\label{sec:js}

~

\subsection{Quotations}
\label{sec:js-meta}

We have been doing this for long. You could use compiler backend or meta-programming.
For JS, backend is perhaps better. For SQL, Freebase, CUDA, ... meta-programming is good.

This is never going to work fully -- there are different exception mechanisms, different
numbers, different casts (for example, unbox)

\subsection{Libraries}
\label{sec:js-lib}

We could ignore all JS libraries and reimplement everything (or run every Hackage library).
This works for some scenarios, but not for the case study covered here.
You need to have access to standard JS libraries.

This brings challenges though -- because the libraries are not F\# libraries.
(for example, we have to initialize the array of series)

\subsection{Async}
\label{sec:js-async}

JavaScript async is single-threaded. Should we have ``js-async''? No, that would be mess,
accept that the semantics is relativized.

\subsection{Relativized semantics}
\label{sec:js-relative}

~
% =================================================================================================

\section{Conclusions and Future Work}

\cite{websharper-piglets} \cite{websharper-guis}

\subsection{Big Picture}
Is there a grand theory? No!

Perhaps we could claim one, but it isnt there [Feyerabend] -- and that is the point 
(leave holes).

\subsection{Conclusions}

% =================================================================================================

\nocite{*}
\bibliographystyle{eptcs}
\bibliography{generic}
\end{document}

### 2015-01-02 structural/structural.tex
Initial commit

\documentclass[10pt]{sigplanconf}

\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{float}
\usepackage{pifont}
\usepackage{graphicx}
\usepackage{multirow}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{verbatim}
\usepackage{alltt}
\usepackage{semantic}
\usepackage{color}
\usepackage{setspace}
\usepackage{dblfloatfix}

\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}

\newcommand{\todo}[1]{\marginpar{\color{red}{\footnotesize \begin{spacing}{0.9} #1 \end{spacing}}}}

\newcommand{\sep}[0]{\; | \;}
\newcommand{\lsep}[0]{\;\; | \;\;}
\newcommand{\tsep}[0]{\; \triangledown \;}
\newcommand{\kvd}[1]{\textbf{\sffamily #1}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\fqt}{\`{}\hspace{-0.25em}\`{}}
\newcommand{\tytag}{\overline{\tau}}
\newcommand{\tytagof}{\ident{tag}}

\newcommand{\str}[1]{\texttt{#1}}
\newcommand{\rangl}{\hspace{-0.3em}>\hspace{-0.25em}}
\newcommand{\langl}{\hspace{-0.25em}<\hspace{-0.3em}}

\begin{document}

\toappear{Unpublished draft.}
\title{Making structured data first-class citizens}

\authorinfo
  {Tomas Petricek}
  {University of Cambridge}
  {tomas.petricek@cam.ac.uk}

\authorinfo
  {Don Syme}
  {Microsoft Research, Cambridge}
  {don.syme@microsoft.com}

\maketitle

% ==================================================================================================

\begin{abstract}
% What is the problem?
% Why is it important?
% What do we do?
% What follows from that?

Accessing data in structured formats such as XML, CSV and JSON in statically typed languages 
is difficult, because the languages do not understand the structure of the data. Dynamically
typed languages make this syntactically easier, but lead to error-prone code.
Despite numerous efforts, most of the data available on the web do not come with a schema. 
The only information available to developers is a set of examples, such as typical server
responses.

We describe an inference algorithm that infers a type of structured formats including 
CSV, XML and JSON. The algorithm is based on finding a common supertype of types
representing individual samples (or values in collections).
We use the algorithm as a basis for an F\# type provider that integrates the inference
into the F\# type system. As a result, users can access CSV, XML and JSON data in a 
statically-typed fashion just by specifying a representative sample document.
\end{abstract}

% ==================================================================================================

\section{Introduction}
\label{sec:intro}

We are witnessing an explosion of digital data. International organizations \cite{data-worldbank}
and governments \cite{data-opengovuk,data-opengovusa} expose numerous datasets; 
knowledge is collected and schematized by communities \cite{data-freebase} and numerous 
commercial services provide web services for data access\footnote{As of the time of writing the 
\url{http://www.programmableweb.com} directory lists 8357 web APIs in over 60 categories.}.

Despite this fact, few strongly-typed programming languages are able to seamlessly integrate 
external information sources as if they were strongly-typed components from a programmer's 
perspective. The \emph{type provider} mechanism of F\# 3.0 \cite{providers-techreport} 
makes it possible to integrate external data sources as statically typed components
through a compiler extensibility mechanism.

In this paper, we focus on accessing structured data formats such as CSV, XML and JSON. 
These formats are used by most of the aforementioned services. However, only a few of the
services provide a schema or other specification of the formally defines structure they use. 
Most often programmers depend on unreliable documentation and exploration of sample responses.

In this paper, we combine type inference from sample documents with F\# type provider mechanism.
The two key contributions of this paper are:

\begin{itemize}
\item We introduce F\# type providers for three most common structured document formats 
  (CSV, XML and JSON) that make structured data first-class citizens in F\# and make them
  accessible in a strongly-typed way.

\item We define types for structured data formats and present a type inference algorithm 
  based on the subtyping relation that infers type (or schema) of any structured 
  data format from a sample or a collection of samples.
\end{itemize}

We do not expect familiarity with F\# type providers, so the mechanism is introduced
along with our structural type providers in Section~\ref{sec:structural}. Type inference
is presented in Section~\ref{sec:inference}. Section~\ref{sec:operational} discusses
runtime representation of structured values and proves that it is sound with respect to
our static types. Finally, Section~\ref{sec:providers} connects the static and 
dynamic aspects by discussing how individual type providers (CSV, XML and JSON) work.

The work described in this paper is available as part of the F\# Data library, 
which is an open-source project used in practice by a number of commercial F\# users\footnote{
Available at \url{http://tpetricek.github.com/FSharp.Data}}. For this reason, we follow a 
pragmatic approach and do not ignore constraints of the real-world (such as the complexity
of the host platform or the presence of \ident{null} values).

% ==================================================================================================

\section{Structural type providers}
\label{sec:structural}

We start with a motivating example that shows how the F\# Data library simplifies working with
the JSON format and then introduce other structured type providers. At the same time, we 
introduce the F\# 3.0 type provider mechanism and highlight the key features of the type inference.

% --------------------------------------------------------------------------------------------------

\subsection{Working with JSON documents}
\label{sec:structural-json}

JSON is a popular format for data exchange on the web. It is based on data structures used in
JavaScript and uses six types (Number, String, Boolean, Array, Object and Null). Consider the
following example:
%
\begin{equation*}
\begin{array}{l}
[\; \{~\str{"name"}: \kvd{null}, \str{"age"}: 23 \}, \\
\;\;  \{~\str{"name"}: \str{"Alexander"}, \str{"age"}: 1.5 \}, \hspace{4em} \\
\;\;  \{~\str{"name"}: \str{"Tomas"} \} \; ]
\end{array}
\end{equation*}
%
In a statically typed (functional) language, JSON values would be represented using a single type. 
We could use pattern matching to check that the value matches an expected structure
and then extract the required values. Assuming \ident{data} is the above document, we print the 
names as follows:
%
\begin{equation*}
\begin{array}{l}
\kvd{match}~\ident{data}~\kvd{with}\\
|~\ident{Array items}\rightarrow\\
\quad\kvd{for}~\ident{item}~\kvd{in}~\ident{items}~\kvd{do}\\
\quad\quad\kvd{match}~\ident{item}~\kvd{with}\\
\quad\quad|~\ident{Object keys}\rightarrow\\
\quad\quad\quad\ident{printf}~\str{"\%s"}~(\ident{Map.find keys}~\str{"name"})\\
\quad\quad|~\underline{~\;}\rightarrow \ident{failwith}~\str{"Incorrect format"} \hspace{4em}\\
|~\underline{~\;}\rightarrow \ident{failwith}~\str{"Incorrect format"}
\end{array}
\end{equation*}
%
The code expects that the input is in a certain format (array of objects with the \str{"name"} field)
and it fails if the input does not match these requirements. The code is complicated for two reasons.
First, we have to repeatedly pattern match and explicitly handle failures and second, fields are
accessed indirectly (using the name as a string rather than member of a statically known type).
The first difficulty could be avoided by defining helper functions, but there is no way to avoid the 
second problem without mechanism such as type providers.

Assuming the \str{people.json} file contains the above data and \ident{data} is a string 
containing information in the same format, we can rewrite the code using \ident{JsonProvider} as 
follows (also printing the age, if it is present):
%
\begin{equation*}
\begin{array}{l}
\kvd{type}~\ident{People}~=~\ident{JsonProvider}\langl\str{"people.json"}\rangl\hspace{1em} \\[0.5em]
\kvd{let}~\ident{items}~=~\ident{People.Parse}(\ident{data})\\
\kvd{for}~\ident{item}~\kvd{in}~\ident{items}~\kvd{do}\\
\quad\ident{printf}~\str{"\%s "}~\ident{item.Name}\\
\quad\ident{Option.iter}~(\ident{printf}~\str{"(\%f)"})~\ident{item.Age}
\end{array}
\end{equation*}
%
This code achieves the same simplicity as dynamically typed programming languages, but it 
is statically type-checked.
%
\paragraph{Type providers overivew.}
The parameter \str{"people.json"} is resolved 
statically at compile-time (it has to be a constant) and is passed to the 
\ident{JsonProvider} which builds a specification of the \ident{People} type and passes it
to the F\# compiler. This type information is also available at development time allowing
advanced tooling such as code completion.

The \ident{JsonProvider} uses a type inference algorithm
(Section~\ref{sec:inference}) to infer the JSON schema from the sample file and generates
F\# types that can be used to read the data. In this case, the resulting type is a collection
of records with \ident{Name} of type \ident{string} and \ident{Age} of type \ident{decimal option}.

The type provider also specifies code that should be executed
at run-time in place of \ident{item.Name} and other operations (demonstrated in the next section).
In this example, the runtime behaviour is the same as in the version using pattern matching,
meaning that the member access throws an exception if the document does not have the expected
format. We specify this precisely in Section~\ref{sec:operational}.

\paragraph{The role of records.}
The example suggests that records play a prominent role when working with structured data. 
In fact, all three formats that we consider naturally contain a notion of record and we map 
them to F\# object types with members to support code completion.

The provider infers the types of fields (\ident{string} and \ident{decimal}) and marks the
\ident{Age} field as optional. This is not the only alternative -- we could infer a union
(sum) type consisting of two different records (with one and two fields), but that would 
complicate the processing code. In general, our inference algorithm tries to minimize the number
of unions in the inferred type.

% --------------------------------------------------------------------------------------------------

\subsection{Reading CSV files}
\label{sec:structural-csv}

The CSV format has the simplest structure of the formats that we consider. The structure is a
collection of records with fields (with names specified by the first row). When inferring the type, 
we need to infer the type of fields. Consider the following example:
%
{\small
\begin{verbatim}
  Ozone; Temp; Date
  41;    67;   2012-05-01
  36.3;  72;   2012-05-02
  12.1;  74;   3 May
\end{verbatim}
}
%
\noindent
Assuming the \str{airdata.csv} file is available locally, we can use a type provider that prints 
data obtained from a web site with live CSV data as follows:
%
\begin{equation*}
\begin{array}{l}
\kvd{type}~\ident{AirCsv}~=~\ident{CsvProvider}\langl\str{"airdata.csv"}\rangl\hspace{1em} \\[0.5em]
\kvd{let}~\ident{air}~=~\ident{AirCsv.Load}(\str{"http://data...air.csv"})\\
\kvd{for}~\ident{row}~\kvd{in}~\ident{air}~\kvd{do}\\
\quad\ident{printf}~\str{"\%s: \%d"}~\ident{row.Date row.Ozone}\\
\end{array}
\end{equation*}
%
The type of the record (\ident{row}) is inferred from the sample file, which contains three
fields. Ozone is always a numeric value, so the provider infers a \ident{decimal} type; Date uses
mixed formats, so it is inferred as \ident{string}.

\paragraph{Erasing type providers.}
At runtime, the type providers we describe use an erasure mechanism similar to Java Generics 
\cite{types-java-erasure}. When called by the compiler, a type provider also returns code that is 
executed in place of the generated types. In the above example, the compiled code looks as follows:
%
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{air}~=~\ident{CsvFile.Load}(\str{"airdata.csv"})\\
\kvd{for}~\ident{row}~\kvd{in}~\ident{air.GetRows}()~\kvd{do}\\
\quad\ident{printf}~\str{"\%s \%f"}~\ident{row.GetString}(\str{"Date"})\\
\quad\quad\quad\quad\quad\quad\quad\;\,\ident{row.GetDecimal}(\str{"Ozone"})\\
\end{array}
\end{equation*}
%
The generated type \ident{AirCsv} is erased to an underlying type \ident{CsvFile}
that represents any CSV document. Access to a named property, such as \ident{Ozone}, is compiled into a 
method call, which reads a column using its name and converts it to a desired
type (\ident{GetDecimal}). Technical report \cite{providers-techreport} provides more details on type erasure.

% --------------------------------------------------------------------------------------------------

\subsection{Parsing RSS feeds}
\label{sec:structural-xml}

XML and JSON formats often contain collections and, especially in case of XML, the collections can
be heterogeneous containing elements of different type (i.e.~differently named XML nodes). Consider
the following RSS feed:
%
{\small{
\begin{verbatim}
  <rss version="2.0"><channel>
    <title> BBC News - Europe </title>
    <item><title> Kurdish activists 
        killed in Paris </title></item>
    <item><title> German MPs warn 
        over UK EU exit </title></item>
  </channel></rss>
\end{verbatim}
}}
%
\noindent
The \ident{channel} node contains a single \ident{title} node and multiple \ident{item} nodes.
Moreover, \ident{title} nodes always contain only a single child, which is a plain text.

When extracting data from similar XML documents, we often need to get nodes of a specific name
or get the content of a node. The type provider we present is optimized for this style of access
and allows processing the feed as follows:
%
\begin{equation*}
\begin{array}{l}
\kvd{type}~\ident{RssFeed}~=~\ident{XmlProvider}\langl\str{"rss.xml"}\rangl\hspace{1em} \\[0.5em]
\kvd{let}~\ident{rss}~=~\ident{RssFeed.Load}(\str{"http://bbc.co.uk/..."})\\
\ident{printf}~\str{"\%s "}~\ident{rss.Title}\\
\kvd{for}~\ident{item}~\kvd{in}~\ident{rss.GetItems}()~\kvd{do}\\
\quad\ident{printf}~\str{"\%s "}~\ident{item.Title}\\
\end{array}
\end{equation*}
%
In this example a heterogeneous collection (\ident{rss}) is represented as a type that contains
a member (\ident{Title}) for an element that is present exactly once and a method 
(\ident{GetItems}) for elements that are present multiple times. As a further simplification,
if an element contains only a primitive value (such as \ident{Title}) it is represented as a 
value of primitive type.

\paragraph{Collections and union types.}
As an alternative, the inference could produce a collection of sum type (with a case for \ident{item}
and a case for \ident{title}). This alternative is useful for documents with complicated structure 
(such as XHTML files), but less useful for files with simpler structure such as web server responses,
data stores and configuration files.

Moreover, our approach has a nice property that the following two, semantically equivalent, XML 
nodes are exposed a type with the same public interface:
%
{\small{
\begin{verbatim}
<author name="Tomas" age="27" />
<author><name>Tomas</name><age>27</age></author>
\end{verbatim}
}}
%
\noindent
In both cases, the type is a record with two properties, typed \ident{string} and \ident{int} 
respectively. The discussion in Section~\ref{sec:inference} uses the simple model (using sum types)
and the inference described in the above examples is discussed in Section~\ref{sec:more-collections}.

% ==================================================================================================

\section{Structural type inference}
\label{sec:inference}

The type inference algorithm for structured data is based on a subtyping relation. When inferring 
the type of a specified document, we infer (the most specific) types of individual values, such as
CSV rows or JSON nodes, and then find the common supertype of values in a given dataset.

We define \emph{structural type} $\tau$ which is a type of the structured data. Note that this type
is distinct from the F\# types (type provider maps \emph{structural types} to ordinary F\# types). 
The subtyping relation is not mapped to a subtyping relation between F\# types, but it holds at 
runtime (meaning that an operation on a super-type is allowed on all subtypes). The operational 
aspects are discussed in Section~\ref{sec:operational}.

% --------------------------------------------------------------------------------------------------

\subsection{Structural types}
\label{sec:inference-types}

The grammar below defines \emph{structural type} $\tau$. Records and record fields are named with
names $\nu$. Record fields are marked with a qualifier $\delta$ which is either $?$ for optional
fields or empty for required fields.
%
\begin{equation*}
\begin{array}{rl}
\multicolumn{2}{l}{
  \tau \; ::= \, \top \lsep \kvd{null}
}\\  
\;\;  &\lsep \tau + \ldots + \tau \lsep [\tau]\\
      &\lsep \nu_{\ident{opt}} \; \{ \delta_1 \nu_1 : \tau_1, \; \ldots \; , \delta_n \nu_n : \tau_n \} \\
      &\lsep \ident{int} \lsep \ident{decimal} \lsep \ident{float} \lsep \ident{bool} \lsep \ident{string} \\
\end{array}
\end{equation*}
%
In structured documents such as CSV, XML and JSON, the record type is the most prominent. We use 
it to type rows of a CSV file, the type of XML nodes and JSON objects.

To support these scenarios, records may be optionally named with $\nu_{\ident{opt}}$. For example,
records representing XML nodes are named with the name of the node. Fields of a record
are named with names $\nu_1, \ldots, \nu_n$ and each field may be marked as optional (we write 
$\ident{?foo}$ for an optional field and $\ident{foo}$ for a mandatory field).

We start by using a simple collection type $[\tau]$ together with a union type $\tau + \ldots + \tau$.
As discussed earlier, the library implements a more sophisticated type inference for heterogeneous
collections, which is discussed in Section~\ref{sec:more-collections}. Aside from collections, a 
union type appears when the user provides multiple samples with different structure.

\paragraph{Alternative representations.}
The record type in our definition is more complex than in other systems, so it is worth discussing 
the alternatives and justifying our design.\todo{Ref. some ``other system'' here?}

Firstly, we could use an optional type to represent optional fields. However, none of the structured
formats we consider (CSV, XML, JSON) uses optional types elsewhere, so we could represent types that
are not inhabited in any our use case. Secondly, we could replace named fields and (optionally) named
records with a type assigning a name to any other type (such as $\tau~\kvd{as}~\nu$). Again, our use
cases only name record types and their fields, so types such as $[\ident{int}]~\kvd{as}~\nu$ would not
be inhabited.

Finally, one of the crucial operations in the system is finding common types for multiple record values.
For example, for values $\{ \ident{a} = 5 \}$ and $\{ \ident{a} = 10, \ident{b} = \kvd{true} \}$
we want to obtain a type $\{ \ident{a} : \ident{int}, \ident{?b} : \ident{bool} \}$. This would require
complex rules matching records of named optional types (and records of optional named types).

\begin{figure}
\begin{center}
\includegraphics[scale=0.85,trim=10mm 4mm 12mm 9mm,clip]{images/hierarchy.pdf} % left bottom right top
\end{center}
\vspace{-1em}
\caption{Subtype relation between structural types}
\label{fig:subtyping-diagram}
\end{figure}

% --------------------------------------------------------------------------------------------------

\subsection{Subtyping relation}
\label{sec:inference-subtyping}

To provide a basic intuition, the subtyping relation between structured types is
informally demonstrated in Figure~\ref{fig:subtyping-diagram}. The formal definition
looks as follows:

\begin{definition}
A subtyping relation between structural types (we write $\tau_1 :> \tau_2$ to denote
that $\tau_2$ is a subtype of $\tau_1$) is defined as (a transitive reflexive closure of):
%
\begin{align}
\label{eq:sub-num}
\ident{float}~:>~\ident{decimal}~&:>~\ident{int}\\
\label{eq:sub-null}
\tau :> \kvd{null} \quad (\textnormal{iff}~\tau \notin &\{\top, \ident{bool}, \ident{int}, \ident{decimal}, \ident{float} \})\\
\label{eq:sub-top}
\tau :> \top \quad &(\textnormal{for all}~\tau)\\
\label{eq:sub-col}
[\tau_1] :> [\tau_2] \quad &(\textnormal{if}~\tau_1 :> \tau_2)\\
\label{eq:sub-union1}
\tau_1 + \ldots + \tau_n :> \tau \quad &(\textnormal{if}~\exists i . \tau_i :> \tau)
\end{align}
\vspace{-1.5em}
\begin{equation}
\label{eq:sub-union2}
\begin{array}{l}
\tau_1 + \ldots + \tau_n :> \tau_1' + \ldots + \tau_m' \quad \\[0.2em]
\quad\quad (\textnormal{if}~\forall i \in 1 .. m. \;\exists j \in 1 .. n~\textnormal{such}~\textnormal{that}~\tau_j :> \tau_i')
\end{array}
\end{equation}
\vspace{-0.4em}
\begin{equation}
\label{eq:sub-record}
\begin{array}{l}
\nu_{\ident{opt}} \; \{ \delta_1 \nu_1 : \tau_1, \; \ldots \; , \delta_n \nu_n : \tau_n \} :>\\
\nu_{\ident{opt}}' \; \{ \delta_1' \nu_1' : \tau_1', \; \ldots \; , \delta_m' \nu_m' : \tau_m'\\[0.3em]
\;\;(\textnormal{\small if $\exists$ injective}~p : \{1 .. m\} \rightarrow \{1 .. n\}~\textnormal{\small such that}\\
\quad\forall i \in 1 .. n. (i \notin p[\{1 .. m\}] \implies \delta_i = ?)~\textnormal{\small and}\\
\quad\forall i \in 1 .. m. (\tau_{p(i)} :> \tau_i' \wedge \delta_i' = ? \implies \delta_{p(i)} = ? ) )
\end{array}
\end{equation}
\end{definition}
%
\noindent
Here is a summary of the key aspects of the definition:

% --------------------------------------------------------------------------------------------------

\begin{figure*}[t]
\vspace{-2em}
\small{
Assume that the record fields are ordered to make $k$ the largest index such that
it holds that $\nu_i = \nu'_i \Leftrightarrow i \leq k$.
}
\begin{equation*}
\inference[(record)\;]
  { \tau_i \tsep \tau'_i \vdash  \tau''_i \quad (\forall i \in 1 .. k) \quad\quad
    \textnormal{let}~\delta''_i = ?~$ iff $~\delta_i = ? ~\vee~ \delta'_i = ? }
  { \begin{array}{c}
    \nu_{\ident{opt}} \; \{ \delta_1 \nu_1 : \tau_1, \; \ldots \; , \delta_n \nu_n : \tau_n \} \tsep
    \nu_{\ident{opt}} \; \{ \delta'_1 \nu'_1 : \tau'_1, \; \ldots \; , \delta'_m \nu'_m : \tau'_m \} \vdash\\
    \nu_{\ident{opt}} \; \{ \delta''_1 \nu_1 : \tau''_1, \; \ldots \; , \delta''_n \nu_k : \tau''_k,
                            ?\nu_{k+1} : \tau_{k+1}, \ldots, ?\nu_n : \tau_n,
                            ?\nu'_{k+1} : \tau'_{k+1}, \ldots, ?\nu'_m : \tau'_m \}
    \end{array} }
\end{equation*}
\vspace{0.5em}

\small{
Assume that the union cases are ordered to make $k$ the largest index such that it holds that 
$\tytagof(\tau_i) = \tytagof(\tau'_i) \Leftrightarrow i \leq k$\newline
and assume that $m', n'$ are indices such that it holds that $\tau_i = \kvd{null} \Leftrightarrow i > n'$ and
$\tau_i' = \kvd{null} \Leftrightarrow i > m'$.
}
\begin{equation*}
\inference[(union-1a)\;]
  { \tau_i \tsep \tau'_i \vdash  \tau''_i \quad (\forall i \in 1 .. k) &
    \exists i. (\tau_i = \kvd{null} \vee \tau_i' = \kvd{null}) &
    \nexists i. (\kvd{null} :> \tau_i \vee \kvd{null} :> \tau_i') }
  { (\tau_1 + \ldots + \tau_n) \tsep (\tau'_1 + \ldots + \tau'_m) \vdash
    (\tau''_1 + \ldots + \tau''_k +
    \tau_{k+1} + \ldots \tau_{n'} + \tau'_{k+1} + \ldots + \tau'_{m'} + \kvd{null}) }
\end{equation*}
\begin{equation*}
\inference[(union-1b)\;]
  { \tau_i \tsep \tau'_i \vdash  \tau''_i \quad (\forall i \in 1 .. k) &
  	\textnormal{\footnotesize Premise of (union-1a) does not hold} }
  { (\tau_1 + \ldots + \tau_n) \tsep (\tau'_1 + \ldots + \tau'_m) \vdash
    (\tau''_1 + \ldots + \tau''_k +
    \tau_{k+1} + \ldots \tau_{n'} + \tau'_{k+1} + \ldots + \tau'_{m'}) }
\end{equation*}
\vspace{0.5em}

\small{
The remaining cases do not have any assumptions:}

\begin{equation*}
\textnormal{\footnotesize{(union-2)}}\;\;
\inference
  {\exists i . \tytagof(\tau_i) = \tytagof(\tau) & \tau \tsep \tau_i \vdash \tau_i'}
  {\tau \tsep (\tau_1 + \ldots + \tau_n) \vdash (\tau_1 + \ldots + \tau_i' + \ldots + \tau_n)}
\quad\quad
\inference
  {\nexists i . \tytagof(\tau_i) = \tytagof(\tau)}
  {\tau \tsep (\tau_1 + \ldots + \tau_n) \vdash (\tau_1 + \ldots + \tau_n + \tau)}
\end{equation*}

\begin{equation*}
\inference[(list)\;]
  {\tau \tsep \tau' \vdash \tau''}
  {[\tau] \tsep [\tau'] \vdash [\tau'']}
%
\quad\quad\quad
%
\inference[(num)\;]
  {\tau :> \tau'}
  {\tau \tsep \tau' \vdash \tau}\;
\inference
  {\tau :> \tau'}
  {\tau' \tsep \tau \vdash \tau_1}
\;(\tau, \tau' \in \{\ident{int}, \ident{decimal}, \ident{float} \})
\end{equation*}

\begin{equation*}
\textnormal{\footnotesize{(top)}}\;\;
\top \tsep \tau \vdash \tau \quad
\tau \tsep \top \vdash \tau
%
\quad\quad\quad
%
\textnormal{\footnotesize{(equal)}}\;\;
\tau \tsep \tau \vdash \tau
%
\quad\quad\quad
%
\inference[(null)\;]
  {\tau :> \kvd{null}}
  {\tau \tsep \kvd{null} \vdash \tau}\;
\inference
  {\tau :> \kvd{null}}
  {\kvd{null} \tsep \tau \vdash \tau_1}
\end{equation*}

\vspace{0.5em}
\small{
If no other inference rules applies, then the following rule is used:}
\begin{equation*}
\textnormal{\footnotesize{(union-3)}}\;\;
\tau_1 \tsep \tau_2 \vdash \tau_1 + \tau_2
\end{equation*}
\vspace{-1em}
\caption{Inference judgements that define the common supertype relation}
\label{fig:subtyping-cst}
\end{figure*}

% --------------------------------------------------------------------------------------------------

\begin{itemize}
\item Numeric types correspond to F\# types. We aim to infer a single most precise numeric type that 
  can represent all values from a sample dataset. We order types by the ranges they represent (\ref{eq:sub-num});
  \ident{int} is 32-bit integer, \ident{decimal} has higher precision than \ident{float},
  but smaller range.

\item The \kvd{null} value is a valid value of all record and union types as well as \ident{string}, but
  it is not a valid value of other primitive types, also called \emph{value types} (\ref{eq:sub-null}).

\item There is a top type (\ref{eq:sub-top}), but no bottom type. However it is possible to find common 
  supertype of any two types, because a union type $\tau_1 + \ldots + \tau_n$ is a supertype of all its 
  components (\ref{eq:sub-union1}).

\item However, given two types, there may be multiple common supertypes that are not related by
  subtyping. For example, arising from rules (\ref{eq:sub-union1}) and (\ref{eq:sub-record}).

\item A record type is a supertype of another record type if it contains all its fields (optional fields
  have to remain optional) and all other fields are optional. Records are only related if they are both 
  anonymous or if they have the same name (\ref{eq:sub-record}).
\end{itemize}

\noindent
Some of the aspects (such as \kvd{null} type and primitive numeric types) may appear overly complicated
and F\# specific. This paper aims to describe a realistic system rather than an (overly) simplified 
model.

Analogous problems will appear with most real-world languages (JVM languages have both \kvd{null}
and multiple numeric types; OCaml has multiple numeric types and would have to represent \kvd{null} with 
an explicit option type).

% --------------------------------------------------------------------------------------------------

\subsection{Common supertype relation}
\label{sec:inference-commonsuper}

As mentioned earlier, the structured type inference relies on a finding common supertype. However,
the partially ordered set of types does not have a unique greatest lower bound. For example, record types
$\{\ident{a}:\ident{int}\}$ and $\{\ident{b}:\ident{bool}\}$ have common supertypes
$\{\ident{?a}:\ident{int}, \ident{?b}:\ident{bool}\}$ and $\{\ident{a}:\ident{int}\} + \{\ident{b}:\ident{bool}\}$
which are unrelated.
Our inference algorithm aims to infer records when possible (the first case). To do that, we use
a \emph{common supertype} relation:

\begin{definition}
A \emph{common supertype} of types $\tau_1$ and $\tau_2$ is a type $\tau$ (written $\tau_1 \triangledown \tau_2 \vdash \tau$)
obtained according to the inference rules in Figure~\ref{fig:subtyping-cst}.
\end{definition}

Notable aspects of the definition including the $\tytagof$ function are discussed below.
The common supertype relation finds a single supertype (among several possibilities). 
The following theorem states that the found type is, indeed, a common supertype:

\begin{theorem}
If $\tau_1 \triangledown \tau_2 \vdash \tau$ then $\tau :> \tau_1$ and $\tau :> \tau_2$.
\end{theorem}
\begin{proof}
By analysis of the inference rules in Figure~\ref{fig:subtyping-cst}.
\end{proof}

\noindent
When finding a common supertype of two records (record), we return a record type that has the union
of fields of the two arguments. The types of shared fields are common supertypes of their respective
types. Fields that are present in only one record or are optional in one of the records are marked
as optional in the result.

Finding a common supertype of union types (union-1) is more complicated. Our definition aims to add
as few additional cases as possible, without creating nested union types (e.g. $\ident{int} + (\ident{bool} + \ident{decimal})$).
This is done by grouping types that have a common supertype which is not a union type (in the previous
example \ident{int} with \ident{decimal} and \ident{bool}). We then find common supertype of each group and create union
of these types (in our example $\ident{decimal} + \ident{bool}$).

The types are grouped by a \emph{tag} that determines the kind of type (number, record, etc.)
and is defined as:
%
\begin{equation*}
\begin{array}{rl}
\multicolumn{2}{l}{
\tytag \; ::= \; \ident{string} \lsep \ident{bool} \lsep \ident{number} 
}\\  
\;\;  &\lsep \ident{record} \lsep \ident{named}~\nu \lsep \ident{union} \lsep \ident{list}
\end{array}
\end{equation*}
%
The tag of a type is obtained using a function $\tytagof$. It holds that two types with the 
same tag (which are not unions) have a common supertype that is also not a union:
%
\begin{equation*}
\begin{array}{ll}
\tytagof : \tau \rightarrow \tytag\\
\tytagof(\ident{string}) &= \ident{string}\\
\tytagof(\ident{bool}) &= \ident{bool}\\
\tytagof([\tau]) &= \ident{list}\\
\tytagof(\tau + \ldots + \tau) &= \ident{union}\\
\tytagof(\ident{int}) = \tytagof(\ident{decimal}) = \tytagof(\ident{float}) &= \ident{number}\\
\tytagof(\{ ?_1 \nu_1 : \tau_1, \; \ldots \; , ?_n \nu_n : \tau_n \}) &= \ident{record}\\
\tytagof(\nu\; \{ ?_1 \nu_1 : \tau_1, \; \ldots \; , ?_n \nu_n : \tau_n \}) &= \ident{named}~\nu
\end{array}
\end{equation*}
%
The function is undefined for the $\top$ and $\kvd{null}$ types, but this is not a 
problem because these types are never used as arguments in Figure~\ref{fig:subtyping-cst}.
Assuming that no primitive value has a type $\top$, the top type is always eliminated in the
(top) rule and so it cannot appear as a member of union in any of the (union) rules.

The \kvd{null} type is handled explicitly. When combining two unions, we only include 
\kvd{null} type explicitly if it is present in one of the original unions and none of the 
other types permit \kvd{null} as a valid value (union-1a), otherwise \kvd{null} is excluded.

\paragraph{Minimising unions.}
We stated earlier that the common supertype relation minimises the use of union types 
(by preferring common numeric types or common record types when possible). This property
can be stated and proved formally:
%
\begin{theorem}
If $\tau :> \tau_1$ and $\tau :> \tau_2$ and $\tau$ is not a union type and
$\tau_1 \tsep \tau_2 \vdash \tau'$ then $\tau'$ is not a union type.
\end{theorem}
\begin{proof}
The only rule that introduces an union type is (union-3). By examining possible structures
of $\tau$, we see that the rule can only be used if there is no other supertype of 
both $\tau_1$ and $\tau_2$.
\end{proof}

% ==================================================================================================

\begin{figure*}[t]
\begin{equation*}
\begin{array}{l}
\kvd{let}~\ident{GetFloat value} =\\
\quad\kvd{match}~\ident{value.GetNumber}()~\kvd{with}\\
\quad|~\ident{Choice1 n} \rightarrow \ident{floatOfInt n}\\
\quad|~\ident{Choice2 d} \rightarrow \ident{floatOfDecimal d}\\
\quad|~\ident{Choice3 f} \rightarrow \ident{f}\\
\end{array}
\quad\quad
\begin{array}{l}
\kvd{let}~\ident{GetDecimal value} =\\
\quad\kvd{match}~\ident{value.GetNumber}()~\kvd{with}\\
\quad|~\ident{Choice1 n} \rightarrow \ident{decimalOfInt n}\\
\quad|~\ident{Choice2 d} \rightarrow \ident{d}\\
\quad|~\underline{~\;} \rightarrow \bot\\
\end{array}
\quad\quad
\begin{array}{l}
\kvd{let}~\ident{GetInt value} =\\
\quad\kvd{match}~\ident{value.GetNumber}()~\kvd{with}\\
\quad|~\ident{Choice1 n} \rightarrow \ident{n}\\
\quad|~\underline{~\;} \rightarrow \bot
\end{array}
\end{equation*}

\caption{Auxiliary functions for accessing numeric values}
\label{fig:runtime-functions}
\end{figure*}

\begin{figure*}[!b]
\hrulefill

\begin{equation*}
\inference
  {\ident{GetInt}(\rho)\downarrow}
  {\rho \in \ident{int}}
\quad
\inference
  {\ident{GetDecimal}(\rho)\downarrow}
  {\rho \in \ident{decimal}}
\quad
\inference
  {\ident{GetFloat}(\rho)\downarrow}
  {\rho \in \ident{float}}
\quad
\inference
  {\ident{GetBool}(\rho)\downarrow}
  {\rho \in \ident{bool}}
\quad
\inference
  {\ident{IsNull}(\rho)=\kvd{true}}
  {\rho \in \kvd{null}}
\end{equation*}\vspace{0.4em}
\begin{equation*}
\inference
  {\rho \in \kvd{null} \vee \ident{GetString}(\rho)\downarrow}
  {\rho \in \ident{string}}
\quad
\inference
  {
  \rho \in \kvd{null} \vee (\ident{GetItems}(\rho)\downarrow \wedge
  \forall \rho' \in \ident{GetItems}(\rho). \rho' \in \tau)
  }
  {\rho \in [\tau]}
\quad
\inference
  { \rho \in \kvd{null} ~\vee \exists i. \rho \in \tau_i }
  { \rho \in \tau_1 + \ldots  + \tau_n }
\end{equation*}\vspace{0.4em}
\begin{equation*}
\inference
  {
  \rho \in \kvd{null} \vee \forall i\in 1 .. n.
  \big((\ident{GetField}(\rho, \nu_{\ident{opt}}, \nu_i) = \ident{Some}(\rho') ~\wedge~ \rho' \in \tau_i) \vee 
  (\ident{GetField}(\rho, \nu_{\ident{opt}}, \nu_i) = \ident{None} ~\wedge~ \delta_i =~ ?)\big)
  }
  {\rho \in \nu_{\ident{opt}} \; \{ \delta_1 \nu_1 : \tau_1, \; \ldots \; , \delta_n \nu_n : \tau_n \}}
\end{equation*}

\caption{Inference rules that define type of a runtime value}
\label{fig:runtime-type}
\end{figure*}
\section{Runtime representation}
\label{sec:operational}

In the previous section, we defined static type of structured documents and we defined the 
subtyping relation between types. In this section, we add the operational semantics -- we 
describe runtime representation of values and discuss when a value belongs to a type. This also
clarifies what structural changes in the input document do not break programs written using 
inferred types.

As discussed earlier, the structured type providers discussed in this article use the 
\emph{type erasure} mechanism. At runtime, all values are represented using a single type that
provides a number of operations that view the value as a value with a certain structure.
If a value does not match the required structure, the operation is undefined.

The key claim that we make in this (and the next) section is that, if we have an (inferred) type $\tau_1$ and
a value that does not contain \kvd{null} and belongs to a type $\tau_2$ which is a subtype 
of $\tau_1$ ($\tau_2 :> \tau_1$), then all operations that may be called by user code to 
access the value are defined.

At runtime, a value from a structured document is represented using a single type that
we call \ident{Value}\footnote{Here we slightly diverge from the actual representation in 
the F\# Data library, which uses a different representation for every format, although they 
mostly share the structure discussed here.}. The type is a record of functions that can be 
called to obtain a value with a specified format (and may be undefined if the required format
is not available):
%
\begin{equation*}
\begin{array}{lcl}
\ident{GetBool} &:& \ident{Value} \rightarrow \ident{bool} \\
\ident{GetNumber} &:& \ident{Value} \rightarrow \ident{int}+\ident{decimal}+\ident{float} \\
\ident{GetString} &:& \ident{Value} \rightarrow \ident{string} \\
\ident{IsNull} &:& \ident{Value} \rightarrow \ident{bool}\\
\ident{GetItems} &:& \ident{Value} \rightarrow \ident{Value list}\;\\
\ident{GetField} &:& \ident{Value} \times \ident{string option} \times \ident{string} \\
                  && \quad \rightarrow \ident{Value option}
\end{array}
\end{equation*}
%
The first three operations extract a primitive value. Note that \ident{GetNumber} may return
any of the supported numeric types. The F\# runtime provides a conversion from \ident{int}
to \ident{decimal} and from \ident{decimal} to \ident{float}. These are used in Figure~\ref{fig:runtime-functions}
to define functions that get a value of a specific numeric type and convert values. The conversion
may lose precision, but it does not overflow.

The \ident{GetField} operation returns the value of a field -- it takes \ident{string option} 
which is the name of the desired record type (or \ident{None} if the record is anonymous) and
the name of the field. The operation returns \ident{None} if the value is a record with matching
name, but the field is missing.

\paragraph{Type of values.} Next, we define what does it mean for a runtime \ident{Value} to 
have a type $\tau$ and show that a value of a certain type is also a value of all its supertypes.

\begin{definition}
Given a \ident{Value}~$\rho$, the value belongs to a type $\tau$ (written
$\rho \in \tau$) as defined by inference rules in Figure~\ref{fig:runtime-type}
(we write $f(x)\downarrow$ to mean that $f(x)$ is defined).
\end{definition}

\noindent
The definition matches the intuitive understanding presented so far. For numeric types, a value is 
of a specific type (e.g.~\ident{decimal}) if \ident{GetNumber} returns the exact type or a type that
can be converted to it (e.g.~from \ident{int}). When \ident{IsNull} is true, the value belongs to 
any record and union type, as well as to the \ident{string} type.

A value is of a union type if it belongs to any of the types that form the union. In order to belong
to a specific record type, the \ident{GetField} function needs to be defined for all fields and it
has to return \ident{Some} for all required fields. Finally, no value belongs to the $\top$ type.

\paragraph{Flexibility.} When inferring document type from a sample, we can expect that future
inputs will not have exactly the same structure (for example, a web service may add additional 
field in the result).

~
\begin{itemize}
\item Using smaller number is allowed (but not bigger)
\item Using null in place of string/record/union
\item Using subtypes as elements of a collection
\item Adding fields is fine (or, in fact, having records that contain fields 
  associated with multiple different record names)
\end{itemize}

~

~

~

\paragraph{Soundness of subtyping.}
The type inference algorithm proceeds by finding a common supertype of sample values. In order for
this to be valid, a runtime value that has a type $\tau$ also needs to belong to all supertypes of
$\tau$. This is proved by the following theorem:

\begin{theorem}
Assuming $\tau_1 :> \tau_2$ and a value $\rho$ has a type $\tau_2$ 
(written $\rho \in \tau_2$) then the value $\rho$ has a type $\tau_1$
($\rho \in \tau_1$).
\end{theorem}
\begin{proof}
By analysis of the cases in the subtyping relation as defined in Section~\ref{sec:inference-subtyping}.
Assuming runtime requirements hold for $\tau_2$, the runtime requirements for $\tau_1$ hold as well.
\end{proof}

% ==================================================================================================

\section{Individual providers}
\label{sec:providers}

now we have type inference and runtime representation, so we can connect the two (somehow..)

\subsection{CSV files}
Given a CSV file

\begin{theorem}
Given a CSV file with inferred type $\tau$, and a value $\rho$ that belongs to this type,
all accessors in the code written by the user are always defined.
\end{theorem}

\subsection{JSON documents}

\section{Additional stuff}

\subsection{Collections}
\label{sec:more-collections}

\subsection{Units of measure}
\label{sec:more-units}

\newpage
~
\newpage

\bibliographystyle{abbrvnat}
\bibliography{structural}

\end{document}

### 2015-01-06 ml-2014-paper/paper.tex
Complete first version

@@
 \usepackage{graphics} 
+\usepackage{stmaryrd}
+\usepackage{amsthm}
 %\usepackage{breakurl}
@@
 
+\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
 \newcommand{\ignp}{\_\kern-.5ex\_\kern-.5ex\_ }
@@
 \newcommand{\narrow}[1]{\hspace{-0.75em} #1 \hspace{-1.25em}~}
+
+\newtheorem*{theorem*}{Theorem}
 
@@
 sources and for integration with untyped programming environments, using \emph{lightweight 
-meta-programming} for targetting JavaScript and \emph{computation expressions} for writing 
+meta-programming} for targeting JavaScript and \emph{computation expressions} for writing 
 asynchronous code.
@@
 
+% =================================================================================================
+%
+%                         #                #           
+%   #       #               #            #               
+%   # # #  #### # #  ##   ### #  #  ### #### #  ##  # #  
+%   # ## #  #   ### #  # #  # #  # #     #   # #  # ## # 
+%   # #  #  #   #   #  # #  # #  # #     #   # #  # #  # 
+%   # #  #  #   #   #  # #  # #  # #     #   # #  # #  # 
+%   # #  #  #   #   #  # # ## # ## #     #   # #  # #  # 
+%   # #  #   ## #    ##   # #  # #  ###   ## #  ##  #  # 
+%                                                    
 % =================================================================================================
@@
 normally ``\emph{rendered unfalsifiable by the methodological decisions of its protagonists}.'' 
-(Lakatos \cite{philosophy-lakatos} as quoted by Chalmers \cite{philosophy-thing})}. The focus on 
+(Lakatos \cite{philosophy-lakatos}, quoted by Chalmers \cite{philosophy-thing})}. The focus on 
 the web directs our inquiry and provides an angle for reconsidering such assumptions.
@@
 
+% =================================================================================================
+%                                                  
+%                                            #      
+%     ###                   ###  #           #      
+%    #     ##    ##  ##    #    #### #  #  ### #  # 
+%   #        #  #   #  #   #     #   #  # #  # #  # 
+%   #      ###  ##  ####    ##   #   #  # #  # #  # 
+%   #     #  #   ## #         #  #   #  # #  # #  # 
+%    #    #  #    # #         #  #   # ## # ##  ##  
+%     ###  ## # ##   ###   ###    ##  # #  # #  ##  
+%                                               #   
+%                                              ##   
 % =================================================================================================
@@
 \subsection{Loading data and updating the user interface}
+\label{sec:case-loading}
 
@@
 \emph{works in practice}. In the rest of the paper, we give more details about the most important 
-aspects of the approach and discuss alternatives`'.
+aspects of the approach and discuss alternatives.
 
 % =================================================================================================
+%
+%                                            #    # 
+%     ##                    #     #          #    # 
+%    #  #  # #   ##  # #    #  #  #  ##  # # #  ### 
+%   #    # ## # #  # ## #   #  #  # #  # ### # #  # 
+%   #    # #  # #### #  #   # # # # #  # #   # #  # 
+%   #    # #  # #    #  #    #   #  #  # #   # #  # 
+%    #  #  #  # #    #  #    #   #  #  # #   # # ## 
+%     ##   ###   ### #  #    #   #   ##  #   #  # # 
+%          #                                        
+%          #                                        
+% =================================================================================================
 
-\section{Type providers}
+\section{Integrating with the open world}
 \label{sec:tp}
-a \\
-b \\
-c \\
-~
 
-\subsection{Integrating data}
+Type providers \cite{fsharp-typeprov} is a mechanism for integrating external components into a 
+statically typed programming language. Such components include information sources (such as World
+Bank), other environments (here, JavaScript libraries via TypeScript), but type-providers can 
+also be used for limited meta-programming (the \ident{Asynchronous} parameter can be seen as
+a form of meta-programming).
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{How type providers work and fail}
+\label{sec:tp-def}
+
+In technical terms, type providers are libraries that are loaded by the compiler (and editor)
+and are executed at compile-time (or edit-time). A type provider builds information about types
+it provides and makes those available to the compiler. It is worth noting that this is done lazily,
+so the type provider does not need to provide types for the entire information space at once.
+
+In terms of programming language theory, type providers can be easily explained anecdotally. 
+They change the starting point for a type-checking of a program as follows:
+%
+\begin{equation*}
+\hspace{8em}
+\begin{array}{rclll}
+ \emptyset &\narrow{\vdash}& e : \tau 
+    &\qquad& (\textit{classical, or closed-world})\\
+ \pi~(\includegraphics[width=1em,trim=0 1.5em 0 0]{earth.png})  &\narrow{\vdash}& e : \tau
+    &\qquad& (\textit{type proiders, or open-world})
+\end{array}
+\end{equation*}
+%
+Rather than starting the type-checking with an empty context, we start type-checking with a
+context containing a \emph{projection} of some information from the world. In this paper, we do not
+attempt to formalize type providers, so we avoid the question of what, formally, is the world here.
+However, this analogy is useful for understanding how type providers work and how they can go wrong:
+
+\begin{itemize}
+\item \emph{Type provider failure.} 
+  The projection (a type provider) is implemented in F\# and can fail (for example, if internet
+  is required but unavailable). However, when it succeeds, it generates a valid F\# code and so 
+  unchecked compilation errors cannot happen.
+
+\item \emph{Runtime failure.}
+  When the world changes and we \emph{run} the program, we may get a runtime exception. It is
+  expected that the provided code will not fail as long as certain assumptions are satisfied (for
+  example, countries are not removed from a database). But if the world changes and the assumptions
+  no longer hold, the provided code may fail with a runtime exception.
+
+\item \emph{Recompilation failure.} 
+  When the world changes and we \emph{recompile} the program, the result of the projection can
+  differ and so we may not be able to type-check code that used to type-check. This can be seen as
+  negative aspect, but there is a clear benefit -- it means we discover an error that would happen
+  at runtime earlier, that is during recompilation.
+\end{itemize}
+
+\noindent
+The above points are \emph{general} points about type providers. In this paper, we are more interested
+in the \emph{specific} type providers used in the case study. We look at the World Bank and TypeScript
+type providers in the following two sections. When discussing type providers, we need to look at 
+how the provider (or the projection $\pi$) works and what assumptions it makes about the world. 
+That is, what world changes cause the provided type to change and what world changes cause a 
+runtime exception.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Integrating with World Bank data}
 \label{sec:tp-data}
 
-~
+The World Bank provider \cite{fsharp-data} is a type provider for a specific data source -- it has 
+been designed specifically for World Bank data. This contrasts with F\# Data type providers for 
+working with CSV, XML and JSON data that take sample or schema as parameter and can thus be used 
+with files of any structure.
 
-\subsection{Integrating languages}
+\subsubsection{The World Bank type provider}
+
+The projection implemented by the type provider generates a type called \ident{Countries} that contains 
+countries (and regions) as members. Each member returns a value of type \ident{Indicators} that is
+also generated and contains all indicators as members. Note that the type provider generates just 
+these two types -- in particular, it does not generate a new type for each country. This would be an 
+interesting option if we could check what indicators are available for a given country.
+
+The type provider builds the structure of the provided types, but the provided types are \emph{erased} 
+during compilation and replaced with runtime implementation. For the World Bank type provider, the 
+erasure works as follows:
+%
+\begin{equation*}
+\begin{array}{lcl}
+ \sem{~\ident{data.}\lident{Czech Republic}~} &=&
+   \ident{data.GetCountries}().\ident{GetCountry}(\str{CZE}) \\
+ \sem{~\ident{cz.}\lident{School enrollment, tertiary (\% gross)}~} &=&
+   \ident{cz.AsyncGetIndicator}(\str{SE.TER.ENRR})
+\end{array}
+\end{equation*}
+%
+The underlying operations (\ident{GetCountry}, \ident{GetCountries} and \ident{AsyncGetIndicator})
+are operations of an underlying runtime library (which is a normal non-provided library). This is
+a typical pattern -- type providers generally produce a light layer on top of a rich runtime library.
+
+An important point here is that the type provider uses the \emph{name} of a country or an indicator
+for the member name, but at runtime, it uses a \emph{code} of the country or the indicator. This has
+interesting implications for the safety properties, as discussed in the next section.
+
+\subsubsection{Safety properties of the type provider}
+
+Now that we know how the World Bank type provider generates the types and the runtime code, we can 
+look at the assumptions it makes about the world and in which cases it encounters one of the failures
+defined in Section~\ref{sec:tp-def}.
+
+\begin{itemize}
+\item If you are offline when using the type provider (compiling or editing), the type provider will
+  not be able to obtain the list of countries or indicators. This is an example of \emph{type provider
+  failure}. In case of World Bank, this is partly prevented by caching of the schema, but internet
+  access is required when using the provider for the first time.
+
+\item An interesting case is when a country or indicator is renamed. In that case, compiled code
+  using the World Bank type provider will continue to work (there is no \emph{runtime failure}), 
+  because the lookup uses the country or the indicator code. However, recompilation will fail
+  as the member name (which appears in the source code) will be different.
+
+\item Finally, if a country or indicator completely disappears, we will get \emph{runtime failure}
+  when running existing compiled code and \emph{recompilation failure}. here, the \emph{recompilation
+  failure} is a useful indicator that the assumptions made by our code have been violated by
+  the data source.
+\end{itemize}
+
+\noindent
+When contrasted with the traditional view of safety in ML family of languages, the World Bank 
+type provider certainly relaxes the usual safety conditions. However, this is not the case when 
+compared with standard developer practices. In particular, if we wrote 
+$\ident{data.GetCountries}().\ident{GetCountry}(\str{CZE})$ by hand, the \emph{runtime failure}
+behaviour would be the same, but we would lose the useful \emph{recompilation failure}. The following
+section discusses this (and other) alternatives.
+
+\subsubsection{Discussion of alternatives}
+
+There are two main alternatives that are worth discussing. First, how would the code look if 
+we did not use any type provider. Second, what are alternative designs for the World Bank type 
+provider and how would such alternatives change the safety properties.
+
+\vspace{-1em}
+\paragraph{Accessing countries without type providers.} First, consider how would we write the code to obtain a list
+of countries without using the World Bank type provider. Assuming we want only the countries and
+regions in Figure~\ref{fig:wb}, we could specify the country codes and then obtain the country objects:
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{let}~\ident{data}~=~\ident{WorldBank.GetDataContext}() \\
+ \kvd{let}~\ident{countryCodes}~=~\lbrack~\str{EU},~\str{CZE},~\str{GBP},~\str{USA}~\rbrack \\
+ \kvd{let}~\ident{countries}~=~\ident{countryCodes}~|\rangl~\ident{List.map}~(\ident{data.GetCountries}().\ident{GetCountry})
+\end{array}
+\end{equation*}
+%
+The motivation for the case study was a data journalism application with a transparent logic
+that is easy to understand (and perhaps even modify). The type provider certainly achieves these
+goals better -- a country can be added just by looking through an auto-complete list and we do not
+need to know its code or exact name. However, if we wanted to list all countries in a region, then
+the World Bank type provider would not be helpful.
+
+As for the safety properties, as discussed earlier, using explicit country codes has the same
+runtime properties as using the type provider. This could be avoided by listing all countries in
+a given region (or, according to another rule), but that is solving a different problem.
+
+In summary, the World Bank type provider fits a specific niche that we were exploring in the case 
+study. That is, when we want to create an information analysis that accesses specific data from
+an information source. With rising popularity of data journalism and Open Government Data
+initiatives, we argue that this is an important problem domainHere, we first specify the country 
+codes and then obtain the country objects.
+
+\vspace{-1em}
+\paragraph{Providing safer data accessors.} 
+The \emph{recompilation failure} when a country is removed from the dataset is an indication that
+code would not behave as expected as runtime and so we find it useful. The problematic case is 
+a \emph{runtime failure} when a country is removed after the program is compiled. One way to 
+avoid this error would be to change the type of the provided members like
+\ident{data.}\lident{Czech Republic} from \ident{Country} to \ident{Country~option}.
+
+However, that is only shifting the burden of error handling from the library to the user. If the
+author of the application does not intend to handle errors (for example, by skipping missing data),
+they would have to write the following:
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{let}~\ident{optCountries}~= \\
+ \quad~\lbrack~~ \ident{data.Countries.}\lident{European Union} \\
+ \qquad   \ident{data.Countries.}\lident{Czech Republic} \\
+ \qquad   \ident{data.Countries.}\lident{United Kingdom} \\
+ \qquad   \ident{data.Countries.}\lident{United States} ~\rbrack 
+\\[0.5em]
+ \kvd{let}~\ident{countries}~=\\
+  \quad \ident{optCountries}~|\rangl~\ident{List.map}~(\kvd{fun}~\ident{countryOpt}~\rightarrow \\
+  \qquad \kvd{match}~\ident{countryOpt}~\kvd{with}~\ident{Some~c}~\rightarrow~\ident{c}~|~\ignp~\rightarrow 
+    \ident{failwith}~\str{Country does not exist!})
+\end{array}
+\end{equation*}
+%
+For the purpose of our case study, providing option types would only make the code more complex.
+However, if we wanted to handle errors more gracefully, then option types would be a better 
+alternative. Note, however, that we can still handle exceptions using $\kvd{try}~\ldots~\kvd{with}$.
+
+Whether using option types is more desirable or not often depends on the use case (how we
+want to handle errors) and reliability of the data source (countries do not disappear often).
+This alternative can be supported by adding a static parameter to the type provider (like
+$\ident{Asynchronous}=\kvd{true}$). For example, the CSV type provider from F\# Data has a 
+parameter \ident{AssumeMissingValues} that serves exactly this purpose.
+
+\vspace{-1em}
+\paragraph{Accessing data ``as of time''.} 
+So far, we discussed how to mitigate the problems caused by the fact that the open world changes.
+Can we make the world \emph{not} change? This is a sensible question for some data sources
+(although not for the World Bank). For example Freebase \cite{google-freebase} supports the 
+notion of ``as of time''. When calling Freebase with the \ident{as\_of\_time} parameter, the API
+returns data and meta-data that was present at the specified date.
+
+This would be a desirable option for our case study -- if we want to illustrate point made 
+by an article, the accompanying application could use data available at the time when the 
+article was written. However, this feature relies on the ability of the data source (not all
+services have this), or on the possibility to create local snapshots (difficult for large data
+sources).
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Integrating with JavaScript libraries}
 \label{sec:tp-lang}
 
-~
+Type providers are often described as a technique that simplifies data access. The TypeScript
+type provider shows that there is a broader range of uses, including integration between different
+programming languages. Another example of type provider for language integration is the R
+provider \cite{fsharp-rprovider} which imports packages and functions of the statistical 
+environemnt R.
+
+\subsubsection{The TypeScript type provider}
+The projection implemented by the TypeScript type provider imports interface definitions in the
+\str{.d.ts} format specified by TypeScript \cite{ms-typescript}. This is a format of type 
+annotations for JavaScript libraries. The following example is an excerpt from the \str{jquery.d.ts}
+file that describes the \ident{jQuery} value and the \ident{attr} method used in our example:
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{declare var}~\ident{jQuery}~:~\ident{JQueryStatic}; 
+\\[0.5em]
+ \kvd{interface}~\ident{JQueryStatic}~\{ \\
+ \quad (\ident{selector}: \kvd{string}, \ident{context?}: \kvd{any}):~\ident{JQuery}; \\
+\} 
+\\[0.5em]
+\kvd{interface}~\ident{JQuery}~\{ \\
+\quad \ident{attr}(\ident{attributeName}~:~\kvd{string}):~\kvd{string}; \\
+\quad \ident{attr}(\ident{attributeName}~:~\kvd{string},~\ident{value}~:~\kvd{any}):~\ident{JQuery}; \\
+\}
+\end{array}
+\end{equation*}
+%
+The example demonstrates a number of typical problems that arise when using type providers for
+language interoperation. Not all TypeScript language constructs have direct equivalent in F\# and 
+the TypeScript type provider needs to map them to other constructs that are available:
+%
+\begin{itemize}
+\item The interface file defines \ident{jQuery} as a global variable. Although F\# supports
+  global bindings, those cannot be provided by a type provider and so the global variable
+  is exposed as a static member of the imported type, so we use it via \ident{j.jQuery}.
+
+\item The interface \ident{JQueryStatic} specifies that the \ident{jQuery} object is callable
+  (other members are omitted). This denotes that the corresponding JavaScript object is a
+  callable function with other members. F\# does not allow ``calling an object'' and so 
+  this is mapped to a method \ident{j.jQuery.Invoke}. 
+\end{itemize}
+
+\noindent
+It is also worth noting that the \ident{attr} member is overloaded and uses an optional parameter.
+However, both of these features are available in F\# and can be directly mapped. Now that we 
+covered how the type structure is generated, we also need to show what code is generated when 
+the types are erased:
+%
+\begin{equation*}
+\begin{array}{l}
+ \sem{~\ident{j.jQuery.Invoke}(\ident{command})~} ~= \\
+ \qquad \kvd{let}~\ident{jq}~=~\ident{Emit.PropertyGetImpl}(\kvd{true}, \str{jQuery}, \lbrack|~|\rbrack) \\
+ \qquad \ident{Emit.CallImpl}(\kvd{false}, \str{}, \lbrack|~ \ident{jq};~ \ident{command}~|\rbrack~) \\[0.75em]
+ \sem{~\ident{jQuery}(\str{<input>})\ident{.attr}(\str{type},\str{checkbox})~} ~= \\
+ \qquad \ident{Emit.CallImpl}(\kvd{false}, \str{attr}, \lbrack|~ \ident{jQuery}(\str{<input>});~\str{type};~\str{checkbox}~|\rbrack~)
+\end{array}
+\end{equation*}
+%
+An important point here is that the above provided code is never actually \emph{executed}. 
+As discussed in Section~\ref{sec:js-meta}, it is translated to JavaScript (which then runs
+in the web browser). So, the above code can be seen more as instructions for the translator.
+The mapping is quite simple -- member calls are translated to \ident{Emit.CallImpl} (empty
+name denotes that the object itself is called) and property getters are translated to 
+\ident{Emit.PropertyGetImpl}. In both cases the first argument denotes whether the call is
+static and the last argument is an array of arguments.
+
+\subsubsection{Safety of cross-language type providers}
+As mentioned earlier, TypeScript does not have the type soundness property \cite{ms-safets},
+because of covariant generics. This illustrates two points. First, this is a property of
+\emph{running} TypeScript code, which is not the case here -- we use TypeScript only as a
+source of annotations for JavaScript libraries. There is no guarantee that the library will 
+actually adhere to the specification, which is an issue that we return to when discussing
+translation to JavaScript (Section~\ref{sec:js-meta}).
+
+The second point is that we can not import the typing rules associated with TypeScript.
+The type provider mechanism can only provide F\# types that then behave according to the
+F\# typing rules. This means that we can import generics\footnote{Due to technical 
+limitations, this is not currently allowed by F\# type providers, but there are no
+fundamental reasons for this.}, but this will not automatically allow using them 
+covariantly. In some cases, the F\# type system is simply more strict -- and in that case,
+users have to use unsafe operations, such as $\ident{unbox}\langl \ident{bool} \rangl$ 
+in the \ident{render} function. We return to the topic of unsafe operations in 
+Section~\ref{sec:js-meta}.
+
+The above discussion highlights a broader point about using type providers for language
+interoperability. Given any language, its type system be weaker or stricter in some ways.
+Both cases make mapping difficult:
+
+\begin{itemize}
+\item First, consider a weaker system with less type information. The type provider may need
+  to map more types to \ident{object} (or similar general type). However, this will make 
+  the provided operations hard to use (because the source language is more flexible than
+  F\#). An alternative is to require explicit annotations (like \textcolor{strclr}{\ttfamily d.ts} 
+  files for JavaScript).
+
+\item Second, if the imported language has a more precise type system than F\#, the type 
+  provider hsa to drop some of the information. This can be done safely for types in 
+  contravariant positions. In the unsafe case, the runtime needs to perform dynamic checks 
+  (Section~\ref{sec:conc-interop}).
+\end{itemize}
+
+\noindent
+In case of JavaScript, we encounter both cases. In many cases, the return type is not
+statically known and is exposed as \kvd{any}, which we then map to F\# \ident{object}
+(and the developer has to use \ident{unbox}). However, TypeScript also supports limited
+form of dependent typing (overloading on constants \cite{ms-typescript-09}). This cannot
+be expressed in the F\# type system and the developer has to do more work (for example,
+choose the right overload or provide an explicit type annotation).
+
+An interesting example of type provider for a language with a weaker type system is the
+R provider \cite{fsharp-rprovider}. It solves the mismatch by using runtime reflection 
+to discover R functions and their declared parameters. However, R provider also generates
+additional function overloads that let the user specify additional parameters (not explicitly 
+declared in the signature). To make such interoperability easier, we propsed additional 
+escape mechanism \cite{fsharp-paramsdict} that provide usable syntax when static type information 
+cannot be fully recovered.
+
+\subsubsection{Discussion of alternatives}
+There are two options for calling JavaScript libraries such as jQuery. We can use annotations
+-- written either in F\# or imported into F\#, or we can let the user write inline JavaScript. 
+A different approach is to discourage the use of existing JavaScript libraries as discussed in 
+Section~\ref{sec:js-lib}.
+
+\vspace{-1em}
+\paragraph{Embedding inline JavaScript.}
+To make the interoperability mechanism general, the host language (here, F\#) should not know about 
+the syntax of the embedded language (here, JavaScript). This means that the JavaScript code can
+only be embedded as strings. FunScript provides this option and it is used for accessing various
+primitive JavaScript operations. For example, the following is one way to define a function that
+converts a value to a number using JavaScript semantics:
+%
+\begin{equation*}
+\begin{array}{l}
+ \lbrack\langl\ident{JSEmit}(\str{return \{0\}*1.0;})\rangl\rbrack \\
+ \kvd{let}~\ident{number}~(\ignp:\ident{obj}) : \ident{float} ~=~\ident{failwith}~\str{JavaScript stub should not be called.}
+\end{array}
+\end{equation*}
+%
+The example uses .NET attributes (on the first line), which are meta-data attached to a function.
+When FunScript finds a call to such annotated function, it replaces the call with the specified 
+JavaScript. The string \textcolor{strclr}{\ttfamily \{0\}} is a placeholder for the argument. Note
+that the body of the function is never actually executed and so we use a placeholder.
+
+The approach provides no guarantees about the inline JavaScript code. At compile-time, it cannot
+even check that the JavaScript code is syntactically correct. This could be improved using a 
+parameterized type provider such as $\ident{JSCode}\langl\str{return \{0\}*1.0;}\rangl$, which 
+would check that the static parameter is syntactically correct JavaScript code. So, type providers
+could be used to make this approach safer.
+
+\vspace{-1em}
+\paragraph{Other ways of writing annotations.}
+Many compilers to JavaScript provide a way for writing type signatures for JavaScript
+libraries in the host language. For example, the following snippet shows a declaration of the
+jQuery \ident{attr} method from a mapping for js\_of\_ocaml \cite{js_ocmal} (definitions in 
+SMLtoJs \cite{js_sml} are similar, but simpler):
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{class type}~\ident{jQuery}~=~\kvd{object} \\
+ \quad \kvd{method}~\ident{attr}~:~\ident{js\_string t} ~\rightarrow~ \ident{js\_string t optdef meth} \\
+ \quad \kvd{method}~\ident{attr\_set}~:~\ident{js\_string t} ~\rightarrow~ \ident{js\_string t} ~\rightarrow~ \ident{jQuery t meth} \\
+ \kvd{end}
+\end{array}
+\end{equation*}
+%
+The js\_of\_ocaml project uses the OCaml object model for calling JavaScript libraries. 
+Compared with FunScript, it is more explicit in importing JavaScript -- the \ident{meth}
+type denotes a JavaScript method; \ident{optdef} specifies that the result may be 
+undefined and \ident{js\_string} denotes a JavaScript string (which is distinct from OCaml
+strings). Also, note that the second method is called \ident{attr\_set}. This is a simple
+naming trick -- OCaml does not support overloading and js\_of\_ocaml simply ignores anything
+after underscore.
+
+Using a type provider is similar to writing a code generator that turns TypeScript 
+\textcolor{strclr}{\ttfamily d.ts} files into the above OCaml annotation. The main difference
+is that type providers provide a smoother integration. Type providers do not produce any artefacts
+(generated files) and can import entire repositories lazilly.
+
+% -------------------------------------------------------------------------------------------------
 
@@
 
-~
+Type providers for information sources, such as the World Bank provider, weaken the usual notion 
+of type safety. In general, they require that certain assumptions about the world do not change
+between the state of the world at compile-time and state of the world at run-time. The TypeScript
+provider is different and it is more interesting to consider runtime behaviour as discussed
+in Section~\ref{sec:conc-interop}.
 
+For type providers that provide access to external information sources, we can formulate a 
+\emph{relativized} notion of type safety. A full formalization is beyond the scope of this
+paper, but the following provides an outline of such theorem.
+
+\begin{theorem*}[Relativized type safety]
+Assume $\pi_{\ident{~WorldBank}}$ is the mapping implemented by the World Bank type provider,
+$w, w_0$ are models of a (read-only) world that can be seen as functions defined on country-indicator
+pairs and $\langl e, w \rangl \rightarrow e'$ is a one-step reduction on expressions that has 
+access to the World Bank data modelled by $w$. Then we define:
+
+\begin{itemize}
+\item
+Multi-step reduction $\langl e, w \rangl \rightarrow^* e'$ if either $e=e'$
+or $\langl e, w \rangl \rightarrow e''$ and $\langl e'', w \rangl \rightarrow^* e'$. 
+\item
+Functions $\ident{countries}(e)$ and $\ident{indicators}(e)$ that return sets of 
+country names \\and indicator names that appear in an expression $e$. 
+\end{itemize}
+%
+Then the following \emph{relativized type safety} implication holds:
+%
+\begin{equation*}
+\begin{array}{lcl}
+(~~\pi_{\ident{~WorldBank}}~(w_0) \vdash e : \tau 
+  && \textnormal{If $e$ is well-typed using a compile-time world $w_0$}
+\\
+\quad \wedge~ \langl e, w \rangl \rightarrow^* e'
+  && \textnormal{\ldots and the expression reduces in one or more steps}
+\\
+\quad \wedge~ (~\forall c \in \ident{countries}(e), 
+  && \textnormal{\ldots and the run-time world model $w$ contains all}
+\\
+       \qquad~~~ \forall i \in \ident{indicators}(e) . \langl c, i \rangl \in \mathit{dom}(w)))
+  && \textnormal{\;\quad country-indicator pairs that may be accessed,} 
+\\
+\Rightarrow ~\ident{value}(e') ~\vee~ (\exists e'' . \langl e', w \rangl \rightarrow e'') 
+  && \textnormal{then the result is either a value or can further reduce.}
+\end{array}
+\end{equation*}
+\end{theorem*}
+
+\noindent
+The theorem has the usual structure of a type safety theorem. The reduction is defined on pairs
+consisting of expression and a world $w$. The result of the reduction is only an expression and
+so it cannot modify the world. Even though the above is only a brief sketch, it demonstrates two
+important points about relativized type safety that are relevant to many type providers.
+
+First, we distinguish between the state of the world $w_0$ that is used at compile-time and the 
+state $w$ used at run-time. If these were the same, the additional assumption would always hold.
+Second, the additional assumptions states that all countries and indicators that \emph{may be
+accessed} need to be available. This is stricter than necessary (akin to effect systems),
+but it does not specify that the run-time world must contains \emph{all} provided countries 
+and indicators.
+
+% =================================================================================================
+%
+%                         # # #                                                            #           
+%   ###                     #                #            #                   ###                  #   
+%  #     ##  # #  #  # #  # # # # #   ###   ####  ##      #  ##   #  #  ##   #     ### # # # # #  #### 
+% #     #  # ## ## # ## # # # # ## # #  #    #   #  #     #    #  #  #    #  #    #    ### # ## #  #   
+% #     #  # #  #  # #  # # # # #  # #  #    #   #  #     #  ###  #  #  ###   ##  #    #   # #  #  #   
+% #     #  # #  #  # #  # # # # #  # #  #    #   #  #     # #  #  #  # #  #     # #    #   # #  #  #   
+%  #    #  # #  #  # #  # # # # #  # # ##    #   #  #     # #  #   ##  #  #     # #    #   # #  #  #   
+%   ###  ##  #  #  # ###  # # # #  #  # #     ##  ##      #  ## #  ##   ## # ###   ### #   # ###    ## 
+%                    #                  #                 #                                  #         
+%                    #                ##                ##                                   #         
+%
 % =================================================================================================
@@
 
-~
+JavaScript has become the \emph{lingua franca} of the web and increasing number of programming 
+languages provide a way of compiling to JavaScript. In F\#, the first project was F\# WebTools
+\cite{fsharp-webtools} in 2007. A more recent and complex framework called Websharper \cite{websharper-guis} 
+is available with full commercial support. In this paper, we use FunScript which is a lightweight
+library focused just on translating F\# to JavaScript.
 
-\subsection{Quotations}
+In this section, we discuss the options available when targeting JavaScript. The compilation to 
+JavaScript can be implemented as compiler back-end or as a library based on meta-programming 
+(Section~\ref{sec:js-meta}), there are different approaches to base libraries (Section~\ref{sec:js-lib}),
+asynchronous computing (Section~\ref{sec:js-async}) and semantic mismatch between the host language
+and JavaScript (Section~\ref{sec:js-relative}).
+
+\subsection{Lightweight meta-programming with quotations}
 \label{sec:js-meta}
 
-We have been doing this for long. You could use compiler backend or meta-programming.
-For JS, backend is perhaps better. For SQL, Freebase, CUDA, ... meta-programming is good.
+The FunScript library is based on F\# quotations \cite{fsharp-metaprog}. This means that the code
+is compiled as ordinary F\#, but the compiler also stores marked blocks of code as data. In FunScript,
+we use a launcher that, when the program runs, analyses its quotations and produces a JavaScript file
+(so, the workflow is to compile the code as usual and then run it to produce JavaScript).
 
-This is never going to work fully -- there are different exception mechanisms, different
-numbers, different casts (for example, unbox)
+In FunScript, we typically need to translate the entire source file. To allow this, we instruct the
+compiler to store code as data for the whole module using the \ident{ReflectedDefinition} attribute.
+This is done at the beginning of the file (which we omitted in Section~\ref{sec:case}):
+%
+\begin{equation*}
+\begin{array}{l}
+ \lbrack\langl\ident{ReflectedDefinition}\rangl\rbrack \\
+ \kvd{module}~\ident{Program}
+ \\[0.5em]
+ \kvd{open}~\ident{FunScript} \\
+ \kvd{open}~\ident{FunScript.TypeScript} \\
+ \kvd{open}~\ident{FSharp.Data} 
+\end{array}
+\end{equation*}
+%
+When marked with \ident{ReflectedDefinition}, the F\# compiler stores the body of all functions and 
+methods in the marked module as \emph{quotations} that can be retrieved at run-time. Note that this does
+not store the AST for the entire file -- for example, information about declared types need to be 
+obtained separately using the .NET reflection mechanism.
 
-\subsection{Libraries}
+The quotation mechanism has other uses. In the following, we use an explicit quotation $\langl @ \ldots @\rangl$,
+which returns the wrapped code as data. This can be used, for example, to compile a matrix calculation
+to run on a GPU \cite{accelerator}:
+%
+\begin{equation*}
+\begin{array}{l}
+ \langl @~ \kvd{fun}~(\ident{input}:\ident{Matrix}\langl\ident{float32}\rangl)~\rightarrow \\
+ \qquad~~ \kvd{let}~\ident{sum}~=~(\ident{shift input}~-\hspace{-0.2em}1~0) ~+~ \ident{input} ~+~ (\ident{shift input}~+\hspace{-0.2em}1~0) \\
+ \qquad~~ \ident{sum}~/~3.0\textnormal{f}~~@\rangl
+\end{array}
+\end{equation*}
+%
+Compared to writing a full compiler back-end, the lightweight approach to meta-programming makes it easy to 
+write a translator from F\# to other languages. This has been used for compiling to SQL queries, compiling
+F\# code to CUDA, for Freebase queries \cite{fsharp-data} and other applications.
+
+Using the lightweight meta-programming approach in FunScript would not be appropriate if we wanted to 
+compile arbitrary existing F\# source code to JavaScript without any modifications. For that, using a 
+compiler back-end is a better choice. However, for the task solved in our case study (compiling 
+newly written F\# code), the approach works well. A related question is providing mapping for 
+standard (in our case, F\# and .NET) libraries used by the code. We discuss this in the next section.
+
+\subsection{Accessing F\# and JavaScript libraries}
 \label{sec:js-lib}
 
-We could ignore all JS libraries and reimplement everything (or run every Hackage library).
-This works for some scenarios, but not for the case study covered here.
-You need to have access to standard JS libraries.
+The F\# code in our case study uses a number of libraries. This includes the F\# core library (for 
+example, the \ident{List.map} function), standard .NET libraries (iteration using the \kvd{for} loop
+uses .NET \ident{IEnumerable} interface) and JavaScript libraries. Accessing these libraries in 
+FunScript follows similar approach to F\# WebTools \cite{fsharp-webtools}.
 
-This brings challenges though -- because the libraries are not F\# libraries.
-(for example, we have to initialize the array of series)
+The range of possible approaches to libraries when translating to JavaScript has two extreme cases.
+We can attempt to port all libraries of our source language (F\#, Haskell or even .NET), or we 
+can ignore most standard libraries and instead provide access to as many JavaScript libraries as 
+possible. FunScript stands in the middle -- it provides access to \emph{some} F\# and .NET libraries,
+but most advanced functionality is accessed through JavaScript libraries.
 
-\subsection{Async}
+\paragraph{Mapping standard .NET and F\# libraries.}
+The F\# ecosystem relies on .NET libraries. This means that taking an F\# library and translating it 
+to JavaScript without any modification is not a viable approach as we would have to be able to also
+translate any (compiled) .NET library. This is where F\# differs, for example, from Haskell which has
+a closed ecosystem and the ghcjs project \cite{haskell-ghcjs} is thus capable of translating many 
+standard libraries.
+
+FunScript implements (explicit) mappings for standard F\# and .NET libraries that are useful in 
+JavaScript (collections, string manipulation, date and other). Those are either reimplemented in
+F\# with \ident{ReflectedDefinition} attribute, or mapped to standard JavaScript library.
+
+An interesting aspect of the case study is that we use F\# Data \cite{fsharp-data} type providers, 
+which is a standard F\# library that has not been built specifically for FunScript. This illustrates 
+how type providers and quotations interact.
+
+When we write code using type providers that provide erased types (both World Bank and TypeScript), 
+then the erasure happens before a quotation is captured. The following listing shows an example. 
+When we write code with explicit (or implicit) quotation containing a povided method (1), the actual 
+quotation (2) contains just calls to the underlying runtime library:
+%
+\begin{equation*}
+\begin{array}{llr}
+ \langl @~\ident{data.}\lident{Czech Republic}~@\rangl &\hspace{10em}& (1)\\
+ \langl @~\ident{data.GetCountries}().\ident{GetCountry}(\str{CZE}) ~@\rangl  && (2)
+\end{array}
+\end{equation*}
+%
+Thus, the FunScript component only needs to provide a client-side implementation of the underlying
+operations \ident{GetCountries}, \ident{GetCountry} and \ident{AsyncGetIndicator}. This is done in the
+same way as mappings for standard .NET libraries. Under the cover, \ident{AsyncGetIndicator} invokes
+an AJAX request to the World Bank API. We provide more details about asynchronous execution in 
+Section~\ref{sec:js-async}.
+
+\paragraph{Accessing standard JavaScript libraries.}
+As discussed earlier, client-side programs can either rely more heavily on JavaScript libraries,
+or they can reimplement most of the core functionality. The latter approach is easier when all code
+is written for the same ecosystem. This is done, for example, by ghcjs \cite{haskell-ghcjs}.
+This technique can be very effective when writing code that will be a part of a larger JavaScript
+application and implements, for example, a core computation. This can then be compiled to run
+efficiently using asm.js \cite{asm-js}.
+
+The case study presented in this paper focuses more on the plumbing code that connects multiple
+components -- F\# type providers for data access and JavaScript libraries for visualization. Thus,
+a smooth integration between F\# and JavaScript is necessary.
+
+An important problem with accessing JavaScript libraries is deciding which types should be 
+mapped to ordinary F\# types and which types should be treated as opaque JavaScript types that
+can only be manipulated by JavaScript operations and operators. For example, consider the following
+excerpt from the code in Section~\ref{sec:case-loading}:
+%
+\begin{equation*}
+\begin{array}{lll} 
+ \kvd{let}~\ident{o}~=~\ident{h.HighchartsOptions}() \\
+ \ident{o.chart} \leftarrow \ident{h.HighchartsChartOptions}(\ident{renderTo}=\str{plc}) \\
+ \ident{o.title} \leftarrow \ident{h.HighchartsTitleOptions}(\ident{text}~=~\ident{head}) \\
+ \ident{o.series} \leftarrow \lbrack| ~ |\rbrack \\
+ (\ldots)
+ \\[0.5em]   
+ \ident{opts.series.push}(\ident{h.HighchartsSeriesOptions}(\ident{data}, \ident{name}))  \\
+ (\ldots)
+\end{array}
+\end{equation*}
+%
+In ordinary F\# code, we would not initialize \ident{o.series} to an empty array before 
+accessing it and we would not expect to use a \ident{push} method to append an element --
+.NET arrays are mutable, but not resizable. Here, the first is necessary. When created, 
+the \ident{series} property of \ident{HighchartsOptions} is \kvd{undefined}. The \ident{push}
+operation is added (as an extension method) to standard .NET array, but it cannot be 
+implemented for ordinary F\# code. The following list briefly summarizes the options for
+mapping JavaScript libraries to F\# and the choices made in FunScript:
+
+\begin{itemize}
+\item Numeric types including \ident{int} and \ident{float} are mapped to JavaScript numbers.
+  However, the floating-point artihmetic in JavaScript differs from the one in F\# and so this changes 
+  the semantics. Similarly, compilation of integer operators requires additional work (for 
+  example, $1/2$ is not an integer division in JavaScript).
+
+\item FunScript maps F\# arrays to JavaScript arrays. We can define extension methods
+  such as \ident{push} that make sense for JavaScript arrays. An alternative is to use a 
+  separate type -- for example js\_of\_ocaml uses a separate \ident{js\_array} type.
+  A nice thing about the FunScript approach is that it lets us use standard F\# modules
+  and functions such as \ident{Array.map}.
+
+\item Due to its .NET heritage, there are many types in F\# that have \kvd{null} as a valid
+  value. This means that there is a reasonable precedent for allowing \kvd{undefined} values
+  on types imported from JavaScript (but not for types defined in F\#). In contrast,
+  js\_of\_ocaml is more explicit -- for example the \ident{attr} operation is marked as 
+  returning \ident{optdef} (potentially undefined) \ident{js\_string} value whereas FunScript
+  uses just .NET \ident{string} (which may be \kvd{null}).
+\end{itemize}
+
+\noindent
+In summary, the approach used by F\# is to reuse as much of F\# and .NET as possible. This
+makes writing and reusing code easier and lets developers use existing familiar libraries and
+idioms. However, it means that semantics of F\# running as JavaScript is not the same as the
+semantics of F\# running as compiled code. We return to this topic in Section~\ref{sec:js-relative}. 
+Before that, the following section gives one more example where similar approach is used.
+
+\subsection{Client-side asynchronous computations}
 \label{sec:js-async}
 
-JavaScript async is single-threaded. Should we have ``js-async''? No, that would be mess,
-accept that the semantics is relativized.
+In F\# asynchronous workflows \cite{fsharp-async} serve two purposes. First, they provide a
+way for running multiple tasks in parallel. Second, they make it possible to write long-running
+non-blocking code without the use of explicit callbacks. In our case study, we only use the 
+latter aspect. When loading data for selected countries in a loop, we fetch data asynchronously 
+(using an AJAX call). Under the cover, JavaScript triggers a callback, which then resumes the 
+asynchronous workflow.
+
+Although the recent HTML5 standard makes it possible to write multi-threaded JavaScript using
+Web Workers, the FunScript implementation of $\ident{async}~\{ \ldots \}$ does not do that. 
+Instead, it uses a simple cooperative model based on continuations (which is similar to how Links 
+\cite{web-links} compilation works).
+
+In ordinary F\#, asynchronous workflows can be started in a number of ways: \ident{Async.RunSynchronously}
+starts the work and blocks until it completes; \ident{Async.Start} starts the work in the background
+and \ident{Async.\-Start\-Immediate} starts the work on the current thread and runs callbacks on 
+in the same synchronization context (essentially on the same thread). However, only the last one
+can be mapped to JavaScript (FunScript translator will fail if the other two are used).
+
+Our approach is again to reuse standard F\# constructs, but interpret them in a way that fits
+with the new execution environment. The alternative would be to define a separate computation type
+such as $\ident{js\_async}~\{ \ldots \}$. So, developers can use existing constructs, but need
+to be aware that their code runs in a different environment.
 
@@
 
-~
+When compiling any language to JavaScript, we can treat JavaScript as a low-level runtime
+(and use the asm.js \cite{asm-js} subset for efficiency), or we can use it as 
+higher-level language and map many source language constructs to corresponding JavaScript
+constructs. In this case study, we used the latter. This enables interop with JavaScript, but 
+it makes it hard (if at all possible) to preserve the original F\# semantics in all cases.
+
+This is perhaps a controversial approach, but it fits well with F\#, because it is similar to 
+how F\# handles interop with .NET. When you do not use .NET objects, you do not need type
+annotations and you do not have to handle \kvd{null} values. If you use .NET, you have to
+accept those at some level. More generally, we could call this property \emph{relativized
+semantics}:
+
+\begin{quotation}
+\noindent
+\emph{When you use the ML subset of F\#, the program will behave in the same way regardless
+of the environment and you are guaranteed the usual ML properties. In other environments, 
+the properties are not guaranteed.}
+\end{quotation}
+
+\noindent
+Making the notion of \emph{relativized semantics} formally precise is outside of the scope
+of this paper. An important aspect is specifying the boundary (see Section~\ref{sec:conc-interop}).
+The problem is similar to writing monadic computations -- each monad is different (and has
+different properties), but there are certain laws that always hold. For full cross-compilation,
+we need a similar set of laws, but for all ML language constructs.
+
 % =================================================================================================
 
-\section{Conclusions and Future Work}
+\section{Related and further work}
+\label{sec:conc-interop}
 
-\cite{websharper-piglets} \cite{websharper-guis}
+\paragraph{Related work.}
 
-\subsection{Big Picture}
-Is there a grand theory? No!
+There is a number of projects that solve similar problems to the ones described in this paper.
+In the F\# ecosystem, the first project compiling to JavaScript was F\# WebTools \cite{fsharp-webtools},
+which also supported asynchronous computations (although using a separate computation type)
+and interop with JavaScript. More recently, WebSharper \cite{websharper-piglets,websharper-guis}
+is a more complete framework for web development that also provide composable abstractions for
+building forms (based on formlets \cite{links-formlets}) and entire applications.
 
-Perhaps we could claim one, but it isnt there [Feyerabend] -- and that is the point 
-(leave holes).
+Other statically-typed functional languages that have some way of running as JavaScript 
+include OCaml \cite{js_ocmal}, SML \cite{js_sml} and Haskell \cite{haskell-ghcjs}. Compared to
+our work, all three are stricter in preserving the semantics of the source language, but 
+do not provide as smooth JavaScript integration.
 
-\subsection{Conclusions}
+\vspace{-1em}
+\paragraph{Further work.}
+The programming model used in our case study combines dynamically typed components (JavaScript
+libraries) and statically typed components (code written in F\#). This suggests a relation with
+the work on gradual typing \cite{gradual-typing} and blame tracking \cite{blame-well}. However,
+in our work, the distinction between the statically and dynamically typed parts (and the boundary) 
+is less explicit. However, it would be interesting to see if gradual typing and blame can be
+used to formally define our informal notion of \emph{relativized semantics} introduced in 
+Section~\ref{sec:js-relative}.
+
+% =================================================================================================
+
+\section{Conclusions}
+
+The key idea of this paper is that writing modern applications for the web requires us to 
+reconsider many assumptions that functional language designers take for granted. The novelty
+of this paper is not in any single technology it presents, but in the combination it shows.
+For this reason, we used the form of a cast study, using a significant example to guide our
+discussion.
+
+\newpage
+There are two main assumptions that we reconsider. The first assumption is that programs 
+operate in a closed world. Instead, modern application access a range of external services
+and information-sources. In our case study, these are accessed using type providers.
+The second assumption is that we can fix the runtime semantics. This becomes difficult when
+the same code is compiled for diverse execution environments such as .NET, JavaScript or
+even CUDA.
+
+If there is one thing that the reader should remember from this case study, it is the 
+idea about F\# saying that ``\emph{when you use it as ML, it behaves as ML}''. That is, when
+you use the ML subset of F\# and do not access external services, information or .NET and
+JavaScript libraries, you still get all the good properties of ML. However, when you access
+external information and run your code as JavaScript, you get a weaker notion of safety that
+we called \emph{relativized type safety} and a weaker runtime guarantees that we called
+\emph{relativized semantics}. Nevertheless, the case study shows that these are enough to
+let us benefit from the functional-first statically-typed programming paradigm in the age of web.
 

### 2015-01-06 ml-2014-paper/paper.tex
Spellcheck

@@
 to the .NET runtime, but the same applies to integrating with JavaScript in the web context.} over 
-other aspects of language design. If you use the F\# langage as ML, you get most of the good 
+other aspects of language design. If you use the F\# language as ML, you get most of the good 
 well-known properties of ML\footnote{For example, F\# does not require type annotations when used 
@@
 In this case study, we develop a web application shown in Figure~\ref{fig:wb}, which lets the
-user compare university enrolment in a number of selected countries and regions around the world. 
+user compare university enrollment in a number of selected countries and regions around the world. 
 The resulting application runs on the client-side (as JavaScript) and fetches data dynamically 
@@
 %
-The type provider connects to the World Bank and obtains a lisf of countries at 
+The type provider connects to the World Bank and obtains a list of countries at 
 \emph{compile-time} and at \emph{edit-time} (when using auto-completion in an editor).
@@
 jQuery patterns. The compiler sees them as ordinary object members. When writing code using F\# editors
-based on the F\# Compiler Serivce \cite{fsharp-fcs}, they also appear in the auto-complete list.
+based on the F\# Compiler Service \cite{fsharp-fcs}, they also appear in the auto-complete list.
 
@@
 block, which is an F\# computation expression \cite{fsharp-zoo}. The F\# compiler performs 
-de-sugaring similar to the CPS transformation and interprets keywods such as \kvd{let!} and 
+de-sugaring similar to the CPS transformation and interprets keywords such as \kvd{let!} and 
 \kvd{for} using special operations (monadic bind and other). The \ident{async} identifier 
@@
 
-\subsection{Larning from the case study}
+\subsection{Learning from the case study}
 
@@
  \pi~(\includegraphics[width=1em,trim=0 1.5em 0 0]{earth.png})  &\narrow{\vdash}& e : \tau
-    &\qquad& (\textit{type proiders, or open-world})
+    &\qquad& (\textit{type providers, or open-world})
 \end{array}
@@
 an information source. With rising popularity of data journalism and Open Government Data
-initiatives, we argue that this is an important problem domainHere, we first specify the country 
+initiatives, we argue that this is an important problem domain. Here, we first specify the country 
 codes and then obtain the country objects.
@@
 provider \cite{fsharp-rprovider} which imports packages and functions of the statistical 
-environemnt R.
+environment R.
 
@@
 \item Second, if the imported language has a more precise type system than F\#, the type 
-  provider hsa to drop some of the information. This can be done safely for types in 
+  provider has to drop some of the information. This can be done safely for types in 
   contravariant positions. In the unsafe case, the runtime needs to perform dynamic checks 
@@
 additional function overloads that let the user specify additional parameters (not explicitly 
-declared in the signature). To make such interoperability easier, we propsed additional 
+declared in the signature). To make such interoperability easier, we proposed additional 
 escape mechanism \cite{fsharp-paramsdict} that provide usable syntax when static type information 
@@
 is that type providers provide a smoother integration. Type providers do not produce any artefacts
-(generated files) and can import entire repositories lazilly.
+(generated files) and can import entire repositories lazily.
 
@@
 then the erasure happens before a quotation is captured. The following listing shows an example. 
-When we write code with explicit (or implicit) quotation containing a povided method (1), the actual 
+When we write code with explicit (or implicit) quotation containing a provided method (1), the actual 
 quotation (2) contains just calls to the underlying runtime library:
@@
 \item Numeric types including \ident{int} and \ident{float} are mapped to JavaScript numbers.
-  However, the floating-point artihmetic in JavaScript differs from the one in F\# and so this changes 
+  However, the floating-point arithmetic in JavaScript differs from the one in F\# and so this changes 
   the semantics. Similarly, compilation of integer operators requires additional work (for 

### 2015-02-14 icfp-2015/paper.tex
Adding ICFP paper - abstract

\documentclass[preprint]{sigplanconf}

\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage[hang,flushmargin]{footmisc} 
\usepackage{graphics} 
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage{hyperref}

% =================================================================================================

\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}

\newcommand{\lang}{\begin{picture}(5,7)
\put(1.1,2.5){\rotatebox{45}{\line(1,0){6.0}}}
\put(1.1,2.5){\rotatebox{315}{\line(1,0){6.0}}}
\end{picture}}
\newcommand{\rang}{\begin{picture}(5,7)
\put(.1,2.5){\rotatebox{135}{\line(1,0){6.0}}}
\put(.1,2.5){\rotatebox{225}{\line(1,0){6.0}}}
\end{picture}}

\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{strclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.0,0.0,0.0}

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\ignp}{\_\kern-.5ex\_\kern-.5ex\_ }
\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\lident}[1]{\textnormal{\sffamily 
  \`{}\hspace{-0.25em}\`{}\hspace{-0.1em}#1\`{}\hspace{-0.25em}\`{}}}
\newcommand{\cmt}[1]{\textit{\sffamily\textcolor{cmtclr}{#1}}}

% =================================================================================================

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}
\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
\copyrightyear{20yy} 
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
\doi{nnnnnnn.nnnnnnn}

%\titlebanner{banner above paper title}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{F\# Data: \textnormal{Making structured data first-class citizens}}
%\subtitle{Subtitle Text, if any}

\authorinfo{Tomas Petricek}
           {University of Cambridge}
           {tomas@tomasp.net}
%\authorinfo{Name2\and Name3}
%           {Affiliation2/3}
%           {Email2/3}

\maketitle

% -------------------------------------------------------------------------------------------------

\begin{abstract}
Most statically typed languages assume that programs run in a closed world, but this is not the 
case. Modern applications interact with external services and often access data in structured 
formats such as XML, CSV and JSON. Static type systems do not understand such external data sources
and only make data access more cumbersome. Should we just give up and leave the messy world of 
external data to dynamic typing and runtime checks?

Of course, we should not give up! In this paper, we show how to integrate external data sources 
into the F\# type system. As most real-world data on the web do not come with an explicit schema, 
we develop a type inference algorithm that infers a type from representative samples. Next, we use 
the type provider mechanism for integrating the inferred structures into the F\# type
system.

The resulting library significantly reduces the amount of code that developers need to write when 
accessing data. It also provides additional safety guarantees -- arguably, as much as 
possible if we abandon the (incorrect) closed world assumption.
\end{abstract}

\category{D.3.3}{Programming Languages}{Language Constructs and Features }
\keywords F\#, Type Providers, JSON, XML, Data, Type Inference

% =================================================================================================

\section{Introduction}

\newpage

\footnote{\sffamily{http://api.openweathermap.org/data/2.5/weather?q=Prague\&units=metric}}

Bacon ipsum dolor amet biltong strip steak pancetta tenderloin. Picanha pork pig shank. Jowl ground round kielbasa capicola swine tenderloin tri-tip beef ribs brisket picanha boudin. Turkey ham hock jowl andouille hamburger pork chop shankle bacon, shank bresaola jerky chuck biltong sirloin meatball.

%
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{data}=\ident{Http.Request}(\str{http://weather.org/?q=Prague}) \\[0.25em]
 \kvd{match}~\ident{JsonValue.Parse}(\ident{data})~\kvd{with} \\[0.25em]
 |~\ident{Record}(\ident{root})\rightarrow \\[0.25em]
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\[0.25em]
 \quad |~\ident{Record}(\ident{main})\rightarrow \\[0.25em]
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\[0.25em]
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f degrees!}~\ident{num}
\end{array}
\end{equation*}
%

%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://weather.org/?q=Prague}\rangl \\[0.25em]
 \ident{printfn}~\str{Lovely \%f degrees!}~(\ident{W.GetSample().Main.Temp})
\end{array}
\end{equation*}
%

The text of the paper begins here.

% -------------------------------------------------------------------------------------------------

\newpage
~
\newpage

\appendix
\section{Appendix Title}

This is the text of the appendix, if you need one.

\acks

Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\bibliographystyle{abbrvnat}

% The bibliography should be embedded for final submission.

\begin{thebibliography}{}
\softraggedright

\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
P. Q. Smith, and X. Y. Jones. ...reference text...

\end{thebibliography}

\end{document}

%                       Revision History
%                       -------- -------
%  Date         Person  Ver.    Change
%  ----         ------  ----    ------

%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

### 2015-02-17 icfp-2015/paper.tex
Introduction and type inference algorithm

@@
 
+\usepackage{amssymb}
 \usepackage{xcolor}
@@
 \usepackage{amsthm}
+\usepackage[hyphens]{url}
 \usepackage{hyperref}
+\usepackage{ftnright}
+\usepackage[notquote]{hanging}
+\usepackage[T1]{fontenc}
+\usepackage{semantic}
 
 % =================================================================================================
+
+\urlstyle{sf}
+
+\makeatletter
+\renewcommand{\@makefntext}[1]{%
+  \parindent 1em%
+  \raggedright
+  \begin{hangparas}{0.8em}{1}
+  \noindent {$^{\@thefnmark}$~\footnotesize{#1}}
+  \end{hangparas}
+}
+\makeatother
 
@@
 \definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
+\definecolor{numclr}{rgb}{0.0,0.4,0.0}
 \definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
@@
 \newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
+\newcommand{\num}[1]{\textnormal{\textcolor{numclr}{\sffamily #1}}}
 \newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
+\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
 \newcommand{\ident}[1]{\textnormal{\sffamily #1}}
@@
 \newcommand{\cmt}[1]{\textit{\sffamily\textcolor{cmtclr}{#1}}}
+
+\newcommand{\lsep}[0]{\;\; | \;\;}
+\newcommand{\narrow}[1]{\hspace{-0.85em} #1 \hspace{-0.85em}}
+
+\newcommand{\tsep}[0]{\; \triangledown \;}
+\newcommand{\tytag}{\ident{tag}}
+\newcommand{\dropopt}[1]{\lfloor#1\rfloor}
+\newcommand{\addopt}[1]{\lceil#1\rceil}
+\newcommand{\tytagof}{\ident{tagof}}
+
+\newtheorem{definition}{Definition}
+\newtheorem{theorem}{Theorem}
 
@@
 
-% -------------------------------------------------------------------------------------------------
+% =================================================================================================
 
@@
 The resulting library significantly reduces the amount of code that developers need to write when 
-accessing data. It also provides additional safety guarantees -- arguably, as much as 
-possible if we abandon the (incorrect) closed world assumption.
+accessing data. It also provides additional safety guarantees. Arguably, as much as possible 
+if we abandon the incorrect closed world assumption.
 \end{abstract}
@@
 % =================================================================================================
+%
+%   ###                                                                       
+%    #  #    # ##### #####   ####  #####  #    #  ####  ##### #  ####  #    # 
+%    #  ##   #   #   #    # #    # #    # #    # #    #   #   # #    # ##   # 
+%    #  # #  #   #   #    # #    # #    # #    # #        #   # #    # # #  # 
+%    #  #  # #   #   #####  #    # #    # #    # #        #   # #    # #  # # 
+%    #  #   ##   #   #   #  #    # #    # #    # #    #   #   # #    # #   ## 
+%   ### #    #   #   #    #  ####  #####   ####   ####    #   #  ####  #    # 
+%
+% =================================================================================================
 
 \section{Introduction}
+\label{sec:introduction}
 
-\newpage
+The key function of many modern applications is to connect multiple data sources or services and
+present the aggregated data in some form. Mobile applications for taking notes, searching train
+schedules or finding tomorrow's weather all communicate with one or more services over the internet.
 
-\footnote{\sffamily{http://api.openweathermap.org/data/2.5/weather?q=Prague\&units=metric}}
+Increasing number of such services provide REST-based end-points that return data as CSV, XML
+or JSON. Despite numerous schematization efforts, the services typically do not come with schema.
+At best, the documentation provides a number of typical requests and sample responses.
 
-Bacon ipsum dolor amet biltong strip steak pancetta tenderloin.
-Picanha pork pig shank.
-Jowl ground round kielbasa capicola swine tenderloin tri-tip beef ribs brisket picanha boudin.
-Turkey ham hock jowl andouille hamburger pork chop shankle bacon, shank bresaola jerky chuck biltong sirloin meatball.
+For example, the OpenWeatherMap service provides an end-point for getting current weather for a 
+given city\footnote{See ``Current weather data'': \url{http://openweathermap.org/current}}. The page documents the input 
+URL parameters and shows one sample JSON response to illustrate the response structure.
+Using a standard library for working with JSON and HTTP, we might call the service and 
+read the temperature as follows \footnote{We abbreviate the full URL: 
+\url{http://api.openweathermap.org/data/2.5/weather?q=Prague\&units=metric}}:
 
+\noindent
+\begin{equation*}
+\begin{array}{l}
+ \kvd{let}~\ident{data}=\ident{Http.Request}(\str{http://weather.org/?q=Prague}) \\[0.1em]
+ \kvd{match}~\ident{JsonValue.Parse}(\ident{data})~\kvd{with} \\[0.1em]
+ |~\ident{Record}(\ident{root})\rightarrow \\[0.1em]
+ \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\[0.1em]
+ \quad |~\ident{Record}(\ident{main})\rightarrow \\[0.1em]
+ \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\[0.1em]
+ \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f degrees!}~\ident{num} \\[0.1em]
+ \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\[0.1em]
+ \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\[0.1em]
+ |~\_\rightarrow \ident{failwith}~\str{Incorrect format} 
+\end{array}
+\end{equation*}
+\vspace{0.1em}
+
+\noindent
+The code assumes that the response has a particular format described in the documentation. The
+root node must be a record with a \str{main} field, which has to be another record containing
+a \str{temp} field with numerical value. When the format is incorrect, the data access simply fails
+with an exception.
+
+Using the JSON type provider from the F\# Data library, we can write code with exactly the 
+same functionality in two lines:
+
+\vspace{-0.6em}
+\begin{equation*}
+\begin{array}{l}
+ \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://weather.org/?q=Prague}\rangl \\[0.1em]
+ \ident{printfn}~\str{Lovely \%f degrees!}~(\ident{W.GetSample().Main.Temp})
+\end{array}
+\end{equation*}
+\vspace{0.0em}
+
+\noindent
+On the first line, $\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider at 
+compile-time with the URL as a sample. The type provider infers the structure of the response
+and provides a type with a \ident{GetSample} method that returns a parsed JSON with nested
+properties \ident{Main.Temp}, returning the temperature as a number.
+
+In the rest of the paper, we give more detailed description of the mechanism and we also discuss
+theoretical safety properties of the approach. The key novel contributions of this paper are:
+
+\begin{itemize}
+\item We present type providers for XML, CSV and JSON (Section~\ref{sec:providers}) 
+  that are available in the F\# Data library and we also cover practical aspects of the 
+  implementation that contributed to their industrial adoption (Section~\ref{sec:impl}).
+
+\item We describe a predictable type inference algorithm for structured data formats that
+  underlies the type providers (Section~\ref{sec:inference}). It is based on finding common 
+  supertype of a set of examples.
+
+\item We introduce the notion of relativized type safety and show how it holds for the code
+  written using our type providers (Section~\ref{sec:safety}). Although we focus on our specific
+  case, this demonstrates a new way of thinking about ML-style type safety that is needed
+  in the context of web.
+\end{itemize}
+
+\noindent
+Although the F\# Data library has been widely adopted is one of the most downloaded F\# libraries 
+it has not yet been presented in an academic form. Additional documentation and source code can 
+be found at \url{http://fsharp.github.io/FSharp.Data}.
+
+% =================================================================================================
+%
+%   #######                                                                                  
+%      #    #   # #####  ######    #####  #####   ####  #    # # #####  ###### #####   ####  
+%      #     # #  #    # #         #    # #    # #    # #    # # #    # #      #    # #      
+%      #      #   #    # #####     #    # #    # #    # #    # # #    # #####  #    #  ####  
+%      #      #   #####  #         #####  #####  #    # #    # # #    # #      #####       # 
+%      #      #   #      #         #      #   #  #    #  #  #  # #    # #      #   #  #    # 
+%      #      #   #      ######    #      #    #  ####    ##   # #####  ###### #    #  ####  
+%
+% =================================================================================================
+
+\section{Structural type providers}
+\label{sec:providers}
+
+We start with an informal overview that shows how F\# Data type providers simplify working with 
+JSON, XML and CSV. We also introduce the necessary aspects of F\# 3.0 type providers along the 
+way. The examples in this section illustrate a number of key properties of our type inference 
+algorithm:
+
+\begin{itemize}
+\item Our mechanism is robust and predictable. This is important as the user directly works with 
+  the inferred types and should understand why a specific type was inferred from a given 
+  sample\footnote{In particular, we do not use probabilistic methods where adding additional 
+  sample could change the shape of the type.}.
+
+\item Our inference mechanism prefers records over unions. This better supports developer tooling --
+  most F\# editors provide code completion hints on ``.'' and so types with properties (records)
+  are easier to use than types that require pattern matching.
+
+\item Finally, we handle a number of practical concerns that may appear overly complicated, but
+  are important in the real world. This includes support for different numerical types, \kvd{null}
+  values and missing data and also different ways of representing Booleans in CSV files.
+\end{itemize}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Working with JSON documents}
+\label{sec:providers-json}
+
+The JSON format used in the example in Section~\ref{sec:introduction} is a popular format for data
+exchange on the web based on data structures used in JavaScript. The following is the definition 
+of the \ident{JsonValue} type used earlier to represent parsed JSON:
 %
@@
 \begin{array}{l}
- \kvd{let}~\ident{data}=\ident{Http.Request}(\str{http://weather.org/?q=Prague}) \\[0.25em]
- \kvd{match}~\ident{JsonValue.Parse}(\ident{data})~\kvd{with} \\[0.25em]
- |~\ident{Record}(\ident{root})\rightarrow \\[0.25em]
- \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\[0.25em]
- \quad |~\ident{Record}(\ident{main})\rightarrow \\[0.25em]
- \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\[0.25em]
- \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f degrees!}~\ident{num}
+ \kvd{type}~\ident{JsonValue} = \\[0.1em]
+ \quad|~\ident{Null} \\[0.1em]
+ \quad|~\ident{Number}~\kvd{of}~\ident{decimal} \\[0.1em]
+ \quad|~\ident{String}~\kvd{of}~\ident{string} \\[0.1em]
+ \quad|~\ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
+ \quad|~\ident{Record}~\kvd{of}~\ident{Map}\langl\ident{string}, \ident{JsonValue}\rangl \\[0.1em]
+ \quad|~\ident{Array}~\kvd{of}~\ident{JsonValue}[] \\[0.1em]
 \end{array}
@@
 %
+The OpenWeatherMap example in the introduction used only a (nested) record containing a numerical
+value. To demonstrate other aspects of the JSON type provider, we look at a more complex example
+that also involves \kvd{null} value and an array:
+%
+{\small{
+\begin{verbatim}
+  [ { "name": null, "age": 25 },
+    { "name": "Alexander", "age": 3.5 },
+    { "name": "Tomas" } ]
+\end{verbatim}
+}
+%
+\noindent
+Say we want to print the names of people in the list with an age if it is available. As before, 
+the standard approach would be to pattern match on the parsed \ident{JsonValue}. The code would 
+check that the top-level node is a \ident{Array}, iterate over the elements checking that each is
+a \ident{Record} with certain properties and throw an exception or skip values in incorrect format.
+The standard approach can be made nicer by defining helper functions. However, we still need to
+specify names of fields as strings, which is error prone and can not be statically checked.
 
+Assuming the \strf{people.json} file contains the above example and \ident{data} is a string value
+that contains another data set in the same format, we can print names and ages as follows:
 %
@@
 \begin{array}{l}
- \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://weather.org/?q=Prague}\rangl \\[0.25em]
- \ident{printfn}~\str{Lovely \%f degrees!}~(\ident{W.GetSample().Main.Temp})
+ \kvd{type}~\ident{People}~=~\ident{JsonProvider}\langl\str{people.json}\rangl\hspace{1em} \\[0.6em]
+ \kvd{let}~\ident{items}~=~\ident{People.Parse}(\ident{data})\\[0.1em]
+ \kvd{for}~\ident{item}~\kvd{in}~\ident{items}~\kvd{do}\\[0.1em]
+ \quad\ident{printf}~\str{\%s }~\ident{item.Name}\\[0.1em]
+ \quad\ident{Option.iter}~(\ident{printf}~\str{(\%f)})~\ident{item.Age}
 \end{array}
@@
 %
+The code achieves the same simplicity as when using dynamically typed languages, but is statically 
+type-checked. In contrast to the earlier example, we now use a local file \strf{people.json} as a 
+sample for the type inference, but then processes data from another source.
 
-The text of the paper begins here.
+\paragraph{Type providers.}
+The notation $\ident{JsonProvider}\langl\str{people.json}\rangl$ on the first line passes a 
+\emph{static parameter} to the type provider. Static parameters are resolved at compile-time, 
+so the file name has to be a constant. The provider analyzes the sample and generates
+a type that we name \ident{People}. In F\# editors that use the F\# Compiler Service [?], the 
+type provider is also executed at development-time and so the same provided types are
+used in code completion.
+
+The \ident{JsonProvider} uses a type inference algorithm (Section~\ref{sec:inference})  and
+infers the following types from the sample:
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{type}~\ident{Entity}~=  \\[0.1em]
+ \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{decimal}\rangl \\[0.1em]
+ \quad \kvd{member}~\ident{Name}~:~\ident{option}\langl \ident{string}\rangl \\[0.6em]
+ \kvd{type}~\ident{People}~=  \\[0.1em]
+ \quad \kvd{member}~\ident{GetSample}~:~\ident{unit}~\rightarrow~\ident{Entity}[] \\[0.1em]
+ \quad \kvd{member}~\ident{Parse}~:~\ident{string}~\rightarrow~\ident{Entity}[] \\[0.1em]
+\end{array}
+\end{equation*}
+%
+The type \ident{Entity} represents the person. The property \ident{Name} is optional, because
+one of the sample records contains \kvd{null} as the name. The property \ident{Age} is also optional,
+because the value is missing in one sample. The two sample ages are an integer \num{25} and a 
+decimal \num{3.5} and so the common inferred type is \ident{decimal}.
+
+The type \ident{People} provides two methods for reading data. \ident{GetSample} returns the
+sample used for the inference and \ident{Parse} parses a JSON string containing data in the same 
+format as the sample. Since the sample JSON is a collection of records, both methods return an 
+array of \ident{Entity} values.
+
+\paragraph{Error handling.}
+In addition to the structure of the types, the type provider also specifies what code should be 
+executed at run-time in place of \ident{item.Name} and other operations. This is done through a 
+\emph{type erasure} mechanism demonstrated in Section~\ref{sec:providers-csv}. In this example,
+the runtime behaviour is the same as in the hand-written sample in Section~\ref{sec:introduction} --
+a member access throws an exception if the document does not have the expected format.
+
+Informally, the safety property discussed in Section~\ref{sec:safety} states that if the inputs
+are subtypes of some of the provided samples (\emph{i.e.}~the samples are representative), then
+no exceptions will occur. In other words, we cannot avoid all failures, but we can prevent some --
+for example, if the OpenWeatherMap changes the response format, the sample in Section~\ref{sec:introduction}
+will not re-compile and the user knows that the code needs to be changed. What this means for
+the traditional ML type safety is discussed in Section~\ref{sec:safety-discuss}.
+
+\paragraph{The role of records.}
+The sample code is easy to write thanks to the fact that most F\# editors provide code completion
+when ``.'' is typed. The developer does not need to look at the sample JSON file to see what fields
+are available for a person. To support this scenario, our inference algorithm prefers records
+(we also treat XML elements and CSV rows as records).
+
+In the above example, this is demonstrated by the fact that \ident{Age} is marked as optional.
+An alternative is to provide two different record types (one with \ident{Name} and other with 
+\ident{Name} and \ident{Age}), but this would complicate the processing code.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Reading CSV files}
+\label{sec:providers-csv}
+
+In the CSV file format, the structure is a collection of rows (records) consisting of fields 
+(with names specified by the first row). The inference needs to infer the types of fields.
+For example:
+%
+{\small{
+\begin{verbatim}
+  Ozone, Temp, Date,       Autofilled
+  41,    67,   2012-05-01, 0
+  36.3,  72,   2012-05-02, 1
+  12.1,  74,   3 kveten,   0
+  17.5,  #N/A, 2012-05-04, 0
+\end{verbatim}
+}}
+%
+\noindent
+One difference between JSON and CSV formats is that in CSV, the literals have no data types.
+In JSON, strings are \str{quoted} and Booleans are \kvd{true} and \kvd{false}. This is not the
+case for CSV and so we need to infer not just the structure, but also the primitive types.
+
+Assuming the sample is saved as \strf{airdata.csv}, the following snippet prints all
+rows from another file that were not autofilled:
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{type}~\ident{AirCsv}~=~\ident{CsvProvider}\langl\str{airdata.csv}\rangl\hspace{1em} \\[0.6em]
+ \kvd{let}~\ident{air}~=~\ident{AirCsv.Parse}(\ident{data})\\[0.1em]
+ \kvd{for}~\ident{row}~\kvd{in}~\ident{air.Rows}~\kvd{do}\\[0.1em]
+ \quad\kvd{if}~\ident{not}~\ident{row.Autofilled}~\kvd{then}\\[0.1em]
+ \quad\quad \ident{printf}~\str{\%s: \%d}~\ident{row.Date row.Ozone}\\
+\end{array}
+\end{equation*}
+%
+The type of the record (\ident{row}) is, again, inferred from the sample. The \ident{Date} column
+uses mixed formats and is inferred as \ident{string} (although we support many date formats and 
+``May 3'' would work). More interestingly, we also infer \ident{Autofilled} as Boolean, because 
+the sample contains only $0$ and $1$ values and using those for Booleans is a common CSV convention.
+Finally, the fields \ident{Ozone} and \ident{Temp} have types \ident{decimal} and
+$\ident{option}\langl\ident{int}\rangl$.
+
+\paragraph{Erasing type providers.}
+At runtime, the type providers we describe use an erasure mechanism similar to Java Generics [?]. 
+A type provider also generates code that is executed in place of the members of the generated types.
+In the above example, the compiled (and actually executed) code looks as follows:
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{let}~\ident{air}~=~\ident{CsvFile}\ident{.Load}(\str{airdata.csv}, \kvd{fun}~\ident{r}\rightarrow\\[0.1em]
+ \quad \ident{asDecimal r}.[\num{0}], \ident{asIntOpt r}.[\num{1}], \ident{r}.[\num{2}], \ident{asBool r}.[\num{3}])\\[0.6em]
+ \kvd{for}~\ident{c1},\_,\ident{c3},\ident{c4} ~\kvd{in}~\ident{air.Rows}~\kvd{do}\\[0.1em]
+ \quad\kvd{if}~\ident{not}~\ident{c4}~\kvd{then}~\ident{printf}~\str{\%s: \%d}~\ident{c3 c1}\\
+\end{array}
+\end{equation*}
+%
+The generated type \ident{AirCsv} is erased to a type $\ident{CsvFile}\langl\alpha\rangl$.
+The \ident{Load} method takes the file name together with a function that turns a row represented 
+as an array of strings to a typed representation -- here, a four-element tuple 
+$\ident{decimal}*\ident{option}\langl\ident{int}\rangl*\ident{string}*\ident{bool}$ (this type
+is also used as a type argument of $\ident{CsvFile}\langl\alpha\rangl$. The properties
+such as \ident{Ozone} are then replaced with an accessor that reads the corresponding tuple
+element.
+
+More details about the full type erasure mechanism can be found in the paper on F\# type 
+providers [?]. We present a simplified model when discussing safety in Section~\ref{sec:safety}.
 
 % -------------------------------------------------------------------------------------------------
+
+\subsection{Processing XML data}
+\label{sec:providers-xml}
+
+The XML format is similar to JSON in that it uses nested elements rather than a flat CSV-like 
+structure. In our inference algorithm, we use the notion of \emph{records} for both JSON records 
+and XML elements. However, unlike in JSON, the XML elements also have a name. An element can 
+contain a collection of child elements. In case of XML, the collection can often be heterogeneous 
+containing child nodes of different names. Consider the following (simplified) RSS news feed as an example:
+%
+{\small{
+\begin{verbatim}
+  <rss version="2.0"><channel>
+    <title> BBC News - Europe </title>
+    <item><title> Kurdish activists 
+        killed in Paris </title></item>
+    <item><title> German MPs warn 
+        over UK EU exit </title></item>
+  </channel></rss>
+\end{verbatim}
+}}
+%
+\noindent
+The \ident{channel} node contains a single \ident{title} node (with the name of the feed) and multiple 
+\ident{item} nodes (individual news articles). In RSS, the \ident{title} nodes always contain plain text,
+while \ident{item} nodes contain \ident{title} (and also \ident{url} and othe properties omitted here).
+
+When extracting data from similar XML documents, we often need to get nodes of a specific name (all
+items) or extract the content of a node (feed title). The type provider we present is optimized for 
+this style of access and allows processing the feed as follows:
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{type}~\ident{RssFeed}~=~\ident{XmlProvider}\langl\str{rss-sample.xml}\rangl\hspace{1em} \\[0.5em]
+ \kvd{let}~\ident{channel}~=~\ident{RssFeed.Load}(\str{http://bbc.co.uk/...}).\ident{Channel}\\
+ \ident{printf}~\str{News from: \%s }~\ident{channel.Title}\\
+ \kvd{for}~\ident{item}~\kvd{in}~\ident{channel.Items}~\kvd{do}\\
+ \quad\ident{printf}~\str{ - \%s }~\ident{item.Title}\\
+\end{array}
+\end{equation*}
+%
+In this example, the type of the \ident{channel} value represents a heterogeneous collection
+that contains exactly one \ident{title} node and zero or more \ident{item} nodes.
+
+\paragraph{Heterogeneous collections and unions.}
+Our inference algorithms recognizes this as a \emph{heterogeneous collection} (if the same structure 
+appears in all samples) and generates a type with a member \ident{Title} for accessing the value 
+of the singleton \ident{title} element together with a member \ident{Items} that returns a 
+collection of zero or more items. As a further simplification, if an element contains only a 
+primitive value (such as \ident{title}) it is represented as a value of primitive type (so 
+\ident{channel.Title} has a type \ident{string}).
+
+As an alternative, the inference could produce a collection of union type (with a case for \ident{item}
+and a case for \ident{title}). This alternative is useful for documents with complicated structure 
+(such as XHTML files), but less useful for files with simpler structure such as web server responses,
+data stores and configuration files.
+
+Moreover, our approach based on heterogeneous collections has a nice property that it generates types
+with the same public interface for the following two ways of representing data in XML:
+%
+{\small{
+\begin{verbatim}
+  <author name="Tomas" age="27" />
+  <author><name>Tomas</name><age>27</age></author>
+\end{verbatim}
+}}
+%
+\noindent
+In both cases, the provided type for \ident{author} has two properties, typed \ident{string} and 
+\ident{int} respectively. In Section~\ref{sec:inference}, we discuss the simplified model (using 
+union types). The inference based on heterogeneous collections is described later in 
+Section~\ref{sec:impl-collections}.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Summary}
+The previous three sections demonstrated the F\# Data type providers for JSON, CSV and XML.
+A reader interested in more examples is invited to look at documentation on the F\# Data web 
+site\footnote{Available at \url{http://fsharp.github.io/FSharp.Data/}}.
+
+It should be noted that we did not attempt to present the library using well-structured ideal 
+samples, but instead used data sets that demonstrate typical issues that are frequent in real 
+world inputs (missing data, inconsistent encoding of Booleans and heterogeneous structures).
+
+From looking at just the previous three examples, these may appear arbitrary, but our experience
+suggests that they are the most common issues. The following JSON response with government debt
+information returned by the World Bank API [?] demonstrates all three problems:
+%
+{\small{
+\begin{verbatim}
+  [ { "page": 1, "pages": 5 },
+    [ { "indicator": "GC.DOD.TOTL.GD.ZS",
+        "date": "2012",
+        "value": null },
+      { "indicator": "GC.DOD.TOTL.GD.ZS",
+        "date": "2010",
+        "value": "35.1422970266502" } ] ]
+\end{verbatim}
+}}
+%
+\noindent
+First of all, the top-level element is a collection containing two values of different kind.
+The first is a record with meta-data about the current page and the second is an array with data. 
+The JSON type provider infers this as heterogeneous collection with properties \ident{Record}
+for the former and \ident{Array} for the latter. Second, the \ident{value} is \kvd{null} for some
+records. Third, numbers can be represented in JSON as numeric literals (without quotes), but
+here, they are returned as string literals instead\footnote{This is often used to avoid non-standard
+numerical types in JavaScript.}.
+
+% =================================================================================================
+%
+%   #######                                                                                  
+%      #    #   # #####  ######    # #    # ###### ###### #####  ###### #    #  ####  ###### 
+%      #     # #  #    # #         # ##   # #      #      #    # #      ##   # #    # #      
+%      #      #   #    # #####     # # #  # #####  #####  #    # #####  # #  # #      #####  
+%      #      #   #####  #         # #  # # #      #      #####  #      #  # # #      #      
+%      #      #   #      #         # #   ## #      #      #   #  #      #   ## #    # #      
+%      #      #   #      ######    # #    # #      ###### #    # ###### #    #  ####  ###### 
+%
+% =================================================================================================
+
+\section{Structural type inference}
+\label{sec:inference}
+
+Our type inference algorithm for structured data formats is based on a subtyping relation. When 
+inferring the type of a document, we infer the most specific types of individual values (CSV rows,
+JSON or XML nodes) and then find a common supertype of values in the given sample.
+
+In this section, we define the \emph{structural type} $\sigma$, which is a type of structured data.
+Note that this type is distinct from the programming language types $\tau$ (the type generates the
+latter from the former). Next, we define the subtyping relation on structural types $\sigma$ and 
+describe the algorithm for finding a common supertype.
+
+Note that the subtyping relation between structural types $\sigma$ does not map to a subtyping 
+relation between F\#  types $\tau$. However, it does hold at runtime, meaning that an operation on 
+a supertype can be performed on its subtypes. We make the type provider mechanism and runtime aspects 
+precise in Section~\ref{sec:safety} when discussing safety.
+
+\subsection{Structural types}
+\label{sec:inference-types}
+
+The grammar below defines \emph{structural type} $\sigma$. We distinguish between \emph{non-nullale types}
+that always have a valid value (written as $\hat{\sigma}$) and \emph{nullable types} that encompass missing 
+and \kvd{null} values (written as $\sigma$). We write $\nu$ for record field names and $\nu_{\ident{opt}}$
+for record names, which can be empty:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \hat{\sigma} &\narrow{=}& \nu_{\ident{opt}}\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.2em]
+                &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bit} \lsep \ident{bool} \lsep \ident{string} 
+ \\[0.6em] 
+       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \hat{\sigma}\;\,\kvd{option}\\[0.2em]
+                &\narrow{|}& [\sigma] \lsep \sigma_1 + \ldots + \sigma_n \\[0.2em]
+                &\narrow{|}& ~\top~ \lsep \kvd{null}
+\end{array}
+\end{equation*}
+%
+The non-nullable types include records (consisting of an optional name and zero or more fields with
+their types) and primitive types. Records arising from XML documents are always named, while records
+used by JSON and CSV providers are always unnamed.
+
+Next, we have three numerical types -- \ident{int} for integers, \ident{decimal} for small high-precision 
+decimal numbers and \ident{float} for floating-point numbers (these are related by sub-typing relation as
+discussed in the next section). Furthermore, the \ident{bit} type is a type of values $0$ and $1$ and 
+makes it possible to infer Boolean in the CSV processing example discussed in Section~\ref{sec:providers-csv}.
+
+Any non-nullable type is also a nullable type, but it can be wrapped in the \kvd{option}
+constructor to explicitly permit the \kvd{null} value. These are typically mapped to the standard F\# option 
+type. A simple collection type $[\sigma]$ is also nullable and missing values or \kvd{null} are treated as 
+empty collection. The type $\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded but not
+ambiguous notation) and $\top$ represents the top type.
+
+Finally, a union type in our model implicitly permits the \kvd{null} value. This is because structural union 
+types are \emph{not} mapped to F\# union types. Instead our encoding requires the user to handle the 
+cases one-by-one and so they also need to handle the situation when none of the cases matches (as discussed
+in Section~\ref{sec:impl-unions}, this is the right choice for an F\# type provider, but it would not 
+necessarilly be a good fit for other languages).
+
+% TODO: Also include reference to section with more on heterogeneous collections.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Subtyping relation}
+\label{sec:inference-subtyping}
+
+To provide a basic intuition, the subtyping relation between structured types is illustrated in
+Figure~\ref{fig:subtyping-diagram}. We split the diagram in two parts. The upper part shows 
+non-nullable types (with records and primitive types). The lower part shows nullable types with 
+\kvd{null}, collections and optional values. We omit links between the two part, but any type
+$\hat{\sigma}$ is a subtype of $\hat{\sigma}~\kvd{option}$ (in the diagram, we abbreviate 
+$\sigma~\kvd{option}$ as $\sigma?$).
+
+The diagram shows only the basic structure and it does not explain relationships between records
+with same name, but different fields and between union types. This following definition makes
+this precise.
+
+\begin{definition}
+We write $\sigma_1 :> \sigma_2$ to denote that $\sigma_2$ is a subtype of $\sigma_1$. The 
+subtyping relation is defined as a transitive reflexive closure of these rules:
+
+\vspace{0.5em}
+\noindent\boxed{\emph{Primitives, options, collections}}
+\begin{align}
+\label{eq:sub-prim}\tag{B1}
+\ident{float}\,:>\,\ident{decimal}\,:>\,\ident{int}\,:>\,\ident{bit}\quad&\textnormal{and}\quad\ident{bool}\,:>\,\ident{bit}\quad\\[-0.2em]
+\label{eq:sub-null}\tag{B2}
+\sigma :> \kvd{null} \quad &(\textnormal{iff}~\sigma \neq \hat{\sigma}) \\[-0.2em]
+\label{eq:sub-opt}\tag{B3}
+\hat{\sigma}~\kvd{option} :> \hat{\sigma} \quad &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
+\label{eq:sub-top}\tag{B4}
+\sigma :> \top \quad &(\textnormal{for all}~\sigma)\\[-0.2em]
+\label{eq:sub-col}\tag{B5}
+[\sigma_1] :> [\sigma_2] \quad &(\textnormal{if}~\sigma_1 :> \sigma_2)
+\end{align}
+
+\noindent\boxed{\emph{Union types}}
+\begin{align}
+\label{eq:sub-union1}\tag{U1}
+\sigma_1 + \ldots + \sigma_n :> \sigma_1' + \ldots + \sigma_n' \quad &(\textnormal{iff}~\forall i.\sigma_i :> \sigma_i') \\
+\label{eq:sub-union2}\tag{U2}
+\sigma_1 + \ldots + \sigma_n :> \sigma_1 + \ldots + \sigma_m \quad &(\textnormal{where}~m \leq n) \\
+\label{eq:sub-union3}\tag{U3}
+\hspace{-1em}\sigma_1 + \ldots + \sigma_n :> \sigma_{\pi(1)} + \ldots + \sigma_{\pi(n)} \quad &(\textnormal{for a permutation}~\pi) \\
+\label{eq:sub-union-el}\tag{U4}
+\sigma_1 + \ldots + \sigma_n :> \sigma_i \quad &(\textnormal{for}~i \in \{1 \ldots n\})
+\end{align}
+
+\noindent\boxed{\emph{Record types}}
+\begin{align}
+\hspace{-1em}
+\label{eq:sub-record1}\tag{R1}
+ \nu~\{ \nu_1\!:\!\sigma_1, \ldots, \nu_n\!:\!\sigma_n \} :> 
+ \nu~\{ \nu_1\!:\!\sigma_1', \ldots, \nu_n\!:\!\sigma_n' \}
+ \;\; (\sigma_i :> \sigma_i')
+\end{align}
+\vspace{-1.4em}
+\begin{align}
+\hspace{-1em}
+\label{eq:sub-record2}\tag{R2}
+ \nu~\{ \nu_1\!:\!\sigma_1, \ldots, \nu_n\!:\!\sigma_n \} :> 
+ \nu~\{ \nu_1\!:\!\sigma_1, \ldots, \nu_m\!:\!\sigma_m \}
+ \;\; (m \geq n)
+\end{align}
+\vspace{-1.0em}
+\begin{align}
+\label{eq:sub-record3}\tag{R3}
+\hspace{-1.75em}
+\begin{array}{l}
+ \nu~\{ \nu_1\!:\!\sigma_1, \ldots, \nu_n\!:\!\sigma_n \} :> \\
+ \nu~\{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, \ldots, \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
+\end{array} \qquad &(\textnormal{for a permutation}~\pi)
+\end{align}
+\vspace{-0.6em}
+\begin{align}
+\label{eq:sub-record4}\tag{R4}
+\hspace{-1.75em}
+\begin{array}{l}
+ \nu~\{ \nu_1\!:\!\sigma_1, \ldots, \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null}, \ldots, \nu_{n+m}\!:\!\kvd{null} \} :> \\
+ \nu~\{ \nu_1\!:\!\sigma_1, \ldots, \nu_n\!:\!\sigma_n \}
+\end{array}
+\end{align}
+\end{definition}
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\begin{center}
+\includegraphics[scale=0.75,trim=5mm 5mm 5mm 5mm,clip]{images/hierarchy.pdf} % left bottom right top
+\end{center}
+\vspace{-0.5em}
+\caption{Subtype relation between structural types}
+\label{fig:subtyping-diagram}
+\vspace{-0.5em}
+\end{figure}
+
+% --------------------------------------------------------------------------------------------------
+
+\noindent
+Here is a summary of the key aspects of the definition:
+\begin{itemize}
+\item For numeric types (\ref{eq:sub-prim}), we infer a single most precise numeric type that 
+  can represent all values from a sample dataset, so we order types by the ranges they represent;
+  \ident{int} is a 32-bit integer, \ident{decimal} has a range of about $-1^{29}$ to $1^{29}$ and 
+  \ident{float} is a double-precision floating-point.
+
+\item The \ident{bit} type is a subtype of both \ident{int} and \ident{bool} (\ref{eq:sub-prim}). 
+  Thus, a sample $0,1$ has a type \ident{bit}, but $0,1,\kvd{true}$ is \ident{bool} and
+  $0,1,2$ becomes \ident{int} (for a sample $0,1,2, \kvd{true}$ we would need a union type).
+
+\item The \kvd{null} type is a subtype of all nullable types (\ref{eq:sub-null}), that is all 
+  $\sigma$ types excluding non-nullable types $\hat{\sigma}$). Any non-nullable type is also a 
+  subtype of its optional version (\ref{eq:sub-opt}).
+
+\item There is a top type (\ref{eq:sub-top}), but no bottom type. However it is possible to find common 
+  supertype of any two types, because a union type $\tau_1 + \ldots + \tau_n$ is a supertype of all its 
+  components (\ref{eq:sub-union-el}).
+
+\item The remaining rules for unins are standard. Subtype can have fewer alternatives 
+  (\ref{eq:sub-union1}), elements can be reordered (\ref{eq:sub-union2}) and subtyping is covariant 
+  (\ref{eq:sub-union3}). Similarly, the collection type is also covariant (\ref{eq:sub-col}).
+
+\item As usual, the subtyping on records is covariant (\ref{eq:sub-record1}), subtype can have additional 
+  fields (\ref{eq:sub-record2}) and fields can  be reordered (\ref{eq:sub-record3}). The interesting rule 
+  is the last one (\ref{eq:sub-record4}).  Together with covariance, it states that a subtype can omit some 
+  fields, provided that their types are nullable.
+\end{itemize}
+
+\noindent
+The rule that allows subtype to have fewer record elements (\ref{eq:sub-record4}) is particularly
+important. It allows us to prefer records in some cases. For example, given two samples 
+$\{ \ident{name}:\ident{string} \}$ and $\{ \ident{name}:\ident{string}, \ident{age}:\ident{int} \}$,
+we can find a common supertype $\{ \ident{name}:\ident{string}, \ident{age}:\ident{int}~\kvd{option} \}$
+which is also a record. For usability reasons, we prefer this to another common supertype
+$\{ \ident{name}:\ident{string} \} + \{ \ident{name}:\ident{string}, \ident{age}:\ident{int} \}$.
+The next section describes precisely how our inference algorithm works.
+
+It is worth noting that some of the aspects (such as 3 different numeric types and handling of 
+the \kvd{null} type) are specific to F\#. Similar problems will appear with most real-world
+languages and systems. JVM and OCaml all haves multiple numeric types, but they would handle \kvd{null}
+differently.
+
+We could also be more or less strict about missing data -- we choose to handle missing values 
+silently when we can (\kvd{null} becomes an empty collection), but we are explicit in other cases 
+(we infer \ident{string}~\kvd{option} as a hint to the user rather than treating \kvd{null} as an 
+empty string). These choices are based on experience with using F\# Data in practice.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure*}[t]
+
+\begin{equation*}
+\hspace{3em}
+\inference[(record-1)\;]
+  { (\nu_i = \nu'_j \Leftrightarrow (i = j) \wedge (i \leq k))
+      \qquad \forall i\in\{ 1 .. k \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
+  { \begin{array}{l}
+    \nu \; \{ \nu_1 : \sigma_1,  \; \ldots \;, \nu_k : \sigma_k, \; \ldots \;, \nu_n : \tau_n \} \tsep
+    \nu \; \{ \nu'_1 : \sigma'_1, \; \ldots \;, \nu'_k : \sigma'_k, \; \ldots \;, \nu'_m : \tau'_m \} \vdash\\
+    \nu \; \{ \nu_1 : \sigma''_1, \; \ldots \; , \nu_k : \sigma''_k, 
+                            \nu_{k+1} : \addopt{\sigma_{k+1}}, \ldots, \nu_n : \addopt{\sigma_n},
+                            \nu'_{k+1} : \addopt{\sigma'_{k+1}}, \ldots, \nu'_m : \addopt{\sigma'_m} \}
+    \end{array} }
+\end{equation*}
+
+% union with something else
+\begin{equation*}
+\textnormal{\footnotesize{(union-1)}}\;\;
+\inference
+  {\exists i . \tytagof(\sigma_i) = \tytagof(\sigma) & \sigma \tsep \sigma_i \vdash \sigma_i' & \tytagof(\sigma)\neq\ident{union}}
+  {\sigma \tsep (\sigma_1 + \ldots + \sigma_n) \vdash (\sigma_1 + \ldots + \dropopt{\sigma_i'} + \ldots + \sigma_n)}
+\quad\quad
+\inference
+  {\nexists i . \tytagof(\sigma_i) = \tytagof(\sigma) & \tytagof(\sigma)\neq\ident{union}}
+  {\sigma \tsep (\sigma_1 + \ldots + \sigma_n) \vdash (\sigma_1 + \ldots + \sigma_n + \dropopt{\sigma})}
+\end{equation*}
+
+% two union types
+\begin{equation*}
+\inference[(union-2)\;]
+  { \tytagof(\sigma_i) = \tytagof(\sigma'_i) \quad 
+    \sigma_i \tsep \sigma_i' \vdash \sigma''_i \qquad (\forall i\in \{ 1 .. k \}) }
+  { \begin{array}{l}
+    (\sigma_1  + \ldots + \sigma_k +  \ldots + \sigma_n) \tsep 
+    (\sigma'_1 + \ldots + \sigma'_k + \ldots + \sigma'_m) \vdash\\
+    (\sigma''_1 + \ldots + \sigma''_k + \sigma_{k+1} + \ldots + \sigma_{n} + \sigma'_{k+1} + \ldots + \sigma'_{m}) 
+    \end{array} }
+%
+\qquad\quad
+\inference[(union-3)\;]
+  {(\textnormal{no other rule applies})}
+  {\sigma_1 \tsep \sigma_2 \vdash \dropopt{\sigma_1} + \dropopt{\sigma_2}}
+\end{equation*}
+
+\begin{equation*}
+\hspace{2em}
+\inference[(order)\;]
+  { \nu \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \tsep \sigma \vdash \sigma' }
+  { \nu \; \{ \nu_{\pi(1)} : \sigma_{\pi(1)}, \ldots, \nu_{\pi(n)} : \sigma_{\pi(n)} \} \tsep \sigma \vdash \sigma' }
+\quad
+\inference
+  { \sigma_1 + \ldots + \sigma_n \tsep \sigma \vdash \sigma' }
+  { \sigma_{\pi(1)} + \ldots + \sigma_{\pi(n)} \tsep \sigma \vdash \sigma' }
+\qquad (\pi~\textnormal{permutation})  
+\end{equation*}
+
+\begin{equation*}
+\hspace{3em}
+% collection types
+\inference[(list)\;]
+  {\sigma_1 \tsep \sigma_2 \vdash \sigma}
+  {[\sigma_1] \tsep [\sigma_2] \vdash [\sigma]}
+\qquad\qquad
+% primitive types
+\inference[(prim)\;]
+  {\sigma_1 :> \sigma_2}
+  {\sigma_1 \tsep \sigma_2 \vdash \sigma_1}\quad
+\begin{array}{l}
+  (\textnormal{when}\,~\sigma_1, \sigma_2 \in \{\ident{bit}, \ident{int}, \ident{decimal}, \ident{float} \} \\
+   \quad~~\;\textnormal{or}\,~\sigma_1, \sigma_2 \in \{\ident{bit}, \ident{bool} \})
+\end{array}   
+\end{equation*}
+
+% all rules are symmetric and reflexive
+\begin{equation*}
+\hspace{4.75em}
+\inference[(sym)\;]
+  {\sigma_1 \tsep \sigma_2 \vdash \sigma}
+  {\sigma_2 \tsep \sigma_1 \vdash \sigma}
+%
+\qquad\quad
+\begin{array}{l}
+ \textnormal{\footnotesize{(refl)}}\;\; \sigma \tsep \sigma \vdash \sigma\\[0.75em]
+ % top type
+ \textnormal{\footnotesize{(top)}}\;\; \top \tsep \sigma \vdash \sigma
+\end{array}
+%
+\qquad\quad 
+% null and nullable
+ \begin{array}{l}
+ \textnormal{\footnotesize{(null-1)}}\;\; \sigma \tsep \kvd{null} \vdash \sigma \quad(\sigma :> \kvd{null}) \\[0.75em]
+ \textnormal{\footnotesize{(null-2)}}\;\; \sigma \tsep \kvd{null} \vdash \sigma~\kvd{option} \quad(\sigma :\ngtr \kvd{null})
+ \end{array}
+\end{equation*}
+
+\caption{Inference judgements that define the common supertype relation}
+\label{fig:subtyping-cst}
+\end{figure*}
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Common supertype relation}
+\label{sec:inference-commonsuper}
+
+As mentioned earlier, the structured type inference relies on a finding common supertype. However,
+the partially ordered set of types does not have a unique greatest lower bound. For example, record types
+$\{\ident{a}:\ident{int}\}$ and $\{\ident{b}:\ident{bool}\}$ have common supertypes
+$\{\ident{?a}:\ident{int}, \ident{?b}:\ident{bool}\}$ and $\{\ident{a}:\ident{int}\} + \{\ident{b}:\ident{bool}\}$
+which are unrelated. Our inference algorithm always prefers records over unions (Theorem~\ref{thm:no-unions}).
+
+The definition of the \emph{common supertype} relation uses a number of auxiliary definitions 
+that are explained in the discussion that follows.
+
+\begin{definition}
+A \emph{common supertype} of types $\sigma_1$ and $\sigma_2$ is a type $\sigma$, written 
+$\sigma_1 \triangledown \sigma_2 \vdash \sigma$, obtained according to the inference rules in 
+Figure~\ref{fig:subtyping-cst}.
+\end{definition}
+
+\noindent
+When finding a common supertype of two records (\emph{record-1}), we return a record type that has the union
+of fields of the two arguments. We assume that the names of the first $k$ fields are the same and the
+remaining fields have different names. To get a record with the right order of elements, we provide the
+(\emph{order}) rule. The types of shared fields become common supertypes of their 
+respective types (recursively). Fields that are present in only one record are marked as optional using
+the following helper definition:
+%
+\begin{equation*}
+\begin{array}{rcll}
+ \addopt{\hat{\sigma}} &\narrow{=}& \hat{\sigma}~\kvd{option} \qquad&(\textnormal{non-nullable types})\\
+ \addopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
+\end{array}
+\end{equation*}
+%
+Finding a common supertype of when one type is a union is more complicated. Our definiion aims to 
+use a flat structure (avoid nesting unions) and limit the number of cases. This is done by grouping
+types that have a common supertype which is not a union type. For example, rather than inferring 
+$\ident{int} + (\ident{bool} + \ident{decimal})$, our algorithm will find the common supertype of
+$\ident{int}$ and $\ident{decimal}$ (which is \ident{decimal}) and it will produce just $\ident{decimal}+\ident{bool}$.
+
+To identify which types have a common supertype which is not a union, we group the types by a 
+\emph{tag}, which is defined as:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \tytag &\narrow{=}&  \ident{string} \lsep \ident{bool} \lsep \ident{number} \lsep \ident{union} \lsep \ident{collection} \\
+        &\narrow{\lsep}& \ident{rec-anon} \lsep \ident{rec-named}\;~\nu 
+\end{array}
+\end{equation*}
+%
+The tag of a type is obtained using a helper function $\tytagof{(-)} : \sigma \rightarrow \tytag$:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \tytagof(\ident{string}) &\narrow{=}& \ident{string}\\
+ \tytagof(\ident{bool}) &\narrow{=}& \ident{bool}\\
+ \tytagof(\ident{decimal}) = \tytagof(\ident{float}) &\narrow{=}& \ident{number}\\
+ \tytagof(\ident{int}) = \tytagof(\ident{bit}) &\narrow{=}& \ident{number}\\
+ \tytagof([\sigma]) &\narrow{=}& \ident{collection}\\
+ \tytagof(\sigma + \ldots + \sigma) &\narrow{=}& \ident{union}\\
+ \tytagof(\{ \nu_1 : \sigma_1, \; \ldots \; , \nu_n : \sigma_n \}) &\narrow{=}& \ident{rec-anon}\\
+ \tytagof(\nu\; \{ \nu_1 : \sigma_1, \; \ldots \; , \nu_n : \sigma_n \}) &\narrow{=}& \ident{rec-named}~\nu \\
+ \tytagof(\hat{\sigma}~\kvd{option}) &\narrow{=}& \tytagof(\hat{\sigma})
+\end{array}
+\end{equation*}
+%
+The function is undefined for the $\top$ and $\kvd{null}$ types, but this is not a 
+problem because these types never need to be used as arguments in Figure~\ref{fig:subtyping-cst}.
+The $\top$ type is always eliminated using the (\emph{top}) rule and the \kvd{null} type
+is eliminated using either (\emph{null-1}) or (\emph{null-2}).
+
+The handling of unions is specified using three rules. When adding a non-union type to a 
+union (\emph{union-1}), the union may or may not contain a case with the same tag as the new type.
+If it does, the new type is combined with the existing one, otherwise a new case is added.
+When combining two unions (\emph{union-2}), we group the cases that have a shared tags.
+Finally, the last rule (\emph{union-3}) covers the case when we are combining two non-union types.
+As discussed earlier, union implicitly permits \kvd{null} values and so we use the following 
+auxiliary function which makes nullable types non-nullable (when possible) to simplify the type:
+%
+\begin{equation*}
+\begin{array}{rcll}
+ \dropopt{\hat{\sigma}~\kvd{option}} &\narrow{=}& \hat{\sigma} &(\textnormal{option})\\
+ \dropopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
+\end{array}
+\end{equation*}
+%
+The remaining rules are mostly straightforward. For collections, we find the common supertype of 
+its elements (\emph{list}); for compatible primitive types, we choose one of the two (\emph{prim})
+and a common supertype with \kvd{null} is either the type itself (\emph{null-1}) or an option 
+type (\emph{null-2}).
+
+\paragraph{Properties.}
+The common supertype relation finds a single supertype (among multiple possible candidates).
+The following theorems specify that the found type is uniquely defined (up to reordering of 
+fields in records and union cases) and is, indeed, a common supertype.
+%
+\begin{theorem}[Uniqueness]
+\raggedright
+If $\sigma_1 \triangledown \sigma_2 \vdash \sigma$ and $\sigma_1 \triangledown \sigma_2 \vdash \sigma'$
+then it holds that $\sigma :> \sigma'$ and $\sigma' :> \sigma$.
+\end{theorem}
+\begin{proof}
+The assumptions and shapes of the types to be unified in the rules in Figure~\ref{fig:subtyping-cst} 
+are disjoint, with the exception of (\emph{order}), (\emph{sym}) and (\emph{refl}). These rules 
+always produce types that are subtypes of each other. This property is preserved by rules that use the
+common supertype relation recursively.
+\end{proof}
+
+\begin{theorem}[Common supertype]
+\raggedright
+If $\sigma_1 \triangledown \sigma_2 \vdash \sigma$ then it holds that $\sigma :> \sigma_1$ and $\sigma :> \sigma_2$.
+\end{theorem}
+\begin{proof}
+By induction over the common supertype derivation using $\vdash$.
+\end{proof}
+
+\noindent
+We stated earlier that the common supertype relation minimises the use of union types 
+(by preferring common numeric types or common record types when possible). This property
+can be stated and proved formally:
+%
+\begin{theorem}
+\label{thm:no-unions}
+Given $\sigma_1, \sigma_2$ if there is s $\sigma$ such that $\sigma :> \sigma_1$ and 
+$\sigma :> \sigma_2$ and $\sigma$ is not a union type, then $\sigma_1 \tsep \sigma_2 \vdash \sigma'$ 
+and $\sigma'$ is not a union type.
+\end{theorem}
+\begin{proof}
+The only rule that introduces an union type is (\emph{union-3}), which is used only when 
+no other inference rule can be applied.
+\end{proof}
+
+% ==================================================================================================
+
+\section{Relativized type safety}
+\label{sec:safety}
+~
+
+% - Featherweight F\#
+% - Generating FF\# from inferred types
+% - Evaluation reduction
+% - Safety property
+
+% ==================================================================================================
+
+\subsection{Discussion}
+\label{sec:safety-discuss}
+alternative - check the whole file upfront based on the inferred type
+what if the sample is online??
+
+\section{Implementation}
+\label{sec:impl}
+~
+
+\label{sec:impl-unions}
+Unions become records with fields - so having null makes sense
+
+Write API
+
+JSON - infering numbers from values (not just from actual value type - world bank example)
+
+\label{sec:impl-collections}
+
+Heterogeneous collections
+
+\section{Conclusions}
+\label{sec:conclusions}
 

### 2015-02-17 icfp-2015/paper.tex
Writing something on semantics

@@
 
+% TODO: The model here does not actually use type erasure!!
+
 \urlstyle{sf}
@@
 
+\newcommand{\reduce}{\rightarrow}
+
 \newtheorem{definition}{Definition}
@@
 
-\item We introduce the notion of relativized type safety and show how it holds for the code
-  written using our type providers (Section~\ref{sec:safety}). Although we focus on our specific
-  case, this demonstrates a new way of thinking about ML-style type safety that is needed
-  in the context of web.
+\item We provide a minimal formal model (Section~\ref{sec:formal}) and use it to prove a
+  relativized notion of type safety for our type providers (Section~\ref{sec:safety}). Although 
+  we focus on our specific case, the relativized safety property demonstrates a new way of 
+  thinking about ML-style type safety that is needed in the context of web.
 \end{itemize}
@@
 
-More details about the full type erasure mechanism can be found in the paper on F\# type 
-providers [?]. We present a simplified model when discussing safety in Section~\ref{sec:safety}.
+The type erasure mechanism is explained in our simplified formal model in Section~\ref{sec:formal}.
+More details about the full implementation in F\# can be found in the paper on F\# type 
+providers [?].
 
@@
 a supertype can be performed on its subtypes. We make the type provider mechanism and runtime aspects 
-precise in Section~\ref{sec:safety} when discussing safety.
+precise in Section~\ref{sec:formal} before discussing safety in Section~\ref{sec:safety}.
 
@@
 \begin{figure*}[t]
+{\small{
 
@@
 \end{equation*}
-
-\caption{Inference judgements that define the common supertype relation}
+}}
+\caption{\small{Inference judgements that define the common supertype relation}}
 \label{fig:subtyping-cst}
@@
 
+\begin{figure*}
+{\small{
+\begin{equation*}
+\hspace{3em}
+\begin{array}{l}
+ \ident{asBool}(i) \reduce \kvd{true}  \quad(i = 1)\\
+ \ident{asBool}(d) \reduce \kvd{true}  \quad(d = 1.0)\\
+ \ident{asBool}(f) \reduce \kvd{true}  \quad(f = 1.0)\\
+\end{array}
+\qquad
+\begin{array}{l}
+ \ident{asBool}(i) \reduce \kvd{false} \quad(i = 0)\\
+ \ident{asBool}(d) \reduce \kvd{false} \quad(d = 0.0)\\
+ \ident{asBool}(f) \reduce \kvd{false} \quad(f = 0.0)\\
+\end{array}
+\qquad
+\begin{array}{l}
+ \ident{asBool}(s) \reduce \kvd{true}  \quad(s=\str{true} \vee s=\str{1}) \\
+ \ident{asBool}(s) \reduce \kvd{false} \quad(s=\str{false} \vee s=\str{0}) \\
+ \ident{asBool}(b) \reduce b\quad(\textnormal{when}~b=\kvd{true}\vee b=\kvd{false})
+\end{array}
+\end{equation*}
+%
+\begin{equation*}
+\hspace{-1em}
+\begin{array}{l}
+ \ident{asInt}(i) \,\reduce i\\
+ \ident{asInt}(d) \reduce i\quad(i = \lfloor d \rfloor)\\
+ \ident{asInt}(f) \reduce i\quad(i = \lfloor f \rfloor) \\
+ \ident{asInt}(s) \,\reduce i\quad(s~\text{represents}~i) \\
+ \ident{asInt}(\kvd{true}) \reduce 1\\
+ \ident{asInt}(\kvd{false}) \reduce 0\\
+\end{array}
+\quad
+\begin{array}{l}
+ \ident{asDec}(i) \,\reduce d\quad(d = i)\\
+ \ident{asDec}(d) \reduce d\\
+ \ident{asDec}(f) \reduce d\quad(d = \lfloor f \rfloor) \\
+ \ident{asDec}(s) \,\reduce d\quad(s~\text{represents}~d) \\
+ \ident{asDec}(\kvd{true}) \reduce 1.0\\
+ \ident{asDec}(\kvd{false}) \reduce 0.0\\
+\end{array}
+\quad
+\begin{array}{l}
+ \ident{asFloat}(i) \,\reduce f\quad(f = i)\\
+ \ident{asFloat}(d) \reduce f\quad(f = d)\\
+ \ident{asFloat}(f) \reduce f\\
+ \ident{asFloat}(s) \,\reduce f\quad(s~\text{represents}~f) \\
+ \ident{asFloat}(\kvd{true}) \reduce 1.0\\
+ \ident{asFloat}(\kvd{false}) \reduce 0.0\\
+\end{array}
+\quad
+\begin{array}{l}
+ \ident{asStr}(i) \,\reduce s\quad(s~\text{represents}~i)\\
+ \ident{asStr}(d) \reduce s\quad(s~\text{represents}~d)\\
+ \ident{asStr}(f) \reduce s\quad(s~\text{represents}~f)\\
+ \ident{asStr}(s) \,\reduce s\\
+ \ident{asStr}(\kvd{true}) \reduce \str{true}\\
+ \ident{asStr}(\kvd{false}) \reduce \str{false}\\
+\end{array}
+\end{equation*}
+%
+\begin{equation*}
+\hspace{8em}
+\begin{array}{l}
+ \ident{getChildren}([s_1; \ldots; s_n]) \reduce [s_1; \ldots; s_n]\\
+ \ident{getChildren}(\kvd{null}) \reduce [~]\\
+\end{array}
+\qquad\qquad\qquad
+\begin{array}{l}
+ \ident{getField}(\nu,\nu_i,\nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
+ \ident{getField}(\epsilon,\nu_i,\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
+\end{array}
+\end{equation*}
+}}
+\caption{\small{Reduction rules for conversion functions}}
+\label{fig:op-conversions}
+\end{figure*}
+
+% TODO: We'll need more things here for working with options 
+% Also, 'getField' should return 'null' if the field does not exist
+% (which we will presumably handle nicely thanks to handling of options)
+
+% ==================================================================================================
+
+\section{Formalising type providers}
+\label{sec:formal}
+
+In this section, we build the necessary theoretical framework for proving the relativised type safety 
+property of F\# Data type providers in Section~\ref{sec:safety}. We start by discussing the runtime 
+representation of structured values and runtime conversions (Section~\ref{sec:formal-convert}). Then 
+we embed these into a formal model of an F\# subset that is necessary for discussing type providers 
+(Section~\ref{sec:formal-ff}) and we describe how F\# Data type providers turn an inferred 
+structural type into actual F\# code (Section~\ref{sec:formal-tp}).
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Structural values and conversions}
+\label{sec:formal-convert}
+
+In the model presented here, we represent JSON, XML and CSV documents using the same 
+\emph{structural value}\footnote{Here, we diverge from the actual implementation in F\# Data which 
+uses a different implementation for each format (reusing existing libraries).}. 
+A structural value are first-order and can be one of the primitive values 
+(Boolean, string, integer, decimal and float), a missing value (\kvd{null}), a record 
+(named or unnamed) and a collection:
+%
+\begin{equation*}
+\begin{array}{lcl}
+ s  &\narrow{=}& i \lsep d \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
+    &\narrow{|}& [s_1; \ldots; s_n] \\[0.1em]
+    &\narrow{|}& \{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \}\\[0.1em]
+    &\narrow{|}& \nu~\{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \} \\
+\end{array}
+\end{equation*}
+
+\noindent
+As suggested by subtyping, our system permits certain runtime conversions (\emph{e.g.}~a numerical 
+value \num{1} can be treated as \kvd{true}). To capture these, the operational semantics accesses 
+values using the following operations:
+%
+\begin{equation*}
+\begin{array}{lcl}
+ op &\narrow{=}& \ident{asInt}(s) \lsep \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{asBool}(s) \\
+    &\narrow{|}& \ident{asStr}(s) \lsep \ident{getChildren}(s) \lsep \ident{getField}(\nu_{\ident{opt}}, \nu, s)
+\end{array}
+\end{equation*}
+%
+The reduction rules for these operations are defined in Figure~\ref{fig:op-conversions}. It is important
+to note that the relativised property we intend to prove about F\# Data provides \emph{sufficient}, but
+not \emph{necessary} conditions. This means that a type provided based on sample is guaranteed to work 
+correctly on certain inputs, but may also handle several additional inputs.
+
+The conversion functions follow this approach. They perform conversions required by the subtyping relation
+(\emph{e.g.}~integer is converted to decimal or a float) but also attempt additional conversions.
+
+% TODO: Finish this
+
+\subsection{Featherweight F\#}
+\label{sec:formal-ff}
+
+Types are primitive types or a class or a StructVal, which represents different structured values
+
+L
+\begin{equation*}
+\begin{array}{l}
+ \tau = \ident{int}\lsep\ident{decimal}\lsep\ident{float}\lsep\ident{bool}\lsep\ident{string}\lsep \ident{StructVal} \lsep C \\[0.2em]
+ L = \kvd{type}~C(\overline{f:\tau}) = \overline{M} \\[0.2em]
+ M = \kvd{member}~N:\tau=e \\[0.2em]
+ v = ...
+\end{array}
+\end{equation*}
+
+.. todo -- evaluation goes here too
+
+\subsection{Type providers}
+\label{sec:formal-tp}
+
+How we turn inferred structural type into F\# type
+
+% ==================================================================================================
+
 \section{Relativized type safety}
 \label{sec:safety}
-~
-
-% - Featherweight F\#
-% - Generating FF\# from inferred types
-% - Evaluation reduction
+
 % - Safety property
@@
 
+Error handling - solve issues just add another example
+
 \label{sec:impl-unions}
@@
 \label{sec:conclusions}
+
+  - Prime example of type providers, explores relativized type safety
+  - Used by industry, downloads, tested, etc. 
+  - Predictable yet powerful type inference
 

### 2015-02-23 icfp-2015/paper.tex
Adding more in semantics

@@
 \usepackage{amssymb}
+\usepackage{multicol}
 \usepackage{xcolor}
@@
 % TODO: The model here does not actually use type erasure!!
+% TODO: Maybe get rid of \mu_opt for records everywhere
+% TODO: Disambigurate 's' as structural value and 's' as string
 
@@
   \begin{hangparas}{0.8em}{1}
-  \noindent {$^{\@thefnmark}$~\footnotesize{#1}}
+  \noindent {$^{\@thefnmark}$~#1}
   \end{hangparas}
@@
 
-\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
-\newcommand{\ignp}{\_\kern-.5ex\_\kern-.5ex\_ }
 \newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
@@
 \newcommand{\ident}[1]{\textnormal{\sffamily #1}}
-\newcommand{\lident}[1]{\textnormal{\sffamily 
-  \`{}\hspace{-0.25em}\`{}\hspace{-0.1em}#1\`{}\hspace{-0.25em}\`{}}}
+\newcommand{\lident}[1]{\textnormal{\sffamily\`{}\hspace{-0.25em}\`{}\hspace{-0.1em}#1\`{}\hspace{-0.25em}\`{}}}
 \newcommand{\cmt}[1]{\textit{\sffamily\textcolor{cmtclr}{#1}}}
@@
 \newcommand{\tytagof}{\ident{tagof}}
-
-\newcommand{\reduce}{\rightarrow}
+\newcommand{\nameoftag}{\ident{nameof}}
+
+\newcommand{\reduce}{\rightsquigarrow}
+
+\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
 
@@
 \end{verbatim}
-}
+}}
 %
@@
 
-When extracting data from similar XML documents, we often need to get nodes of a specific name (all
-items) or extract the content of a node (feed title). The type provider we present is optimized for 
+When extracting data from such XML documents, we often need to get nodes of a specific name (all
+items) or extract the content of a node (feed title). The XML type provider is optimized for 
 this style of access and allows processing the feed as follows:
-%
+
+\noindent
 \begin{equation*}
@@
 
-Our type inference algorithm for structured data formats is based on a subtyping relation. When 
+Our type inference algorithm for structured formats is based on a subtyping relation. When 
 inferring the type of a document, we infer the most specific types of individual values (CSV rows,
-JSON or XML nodes) and then find a common supertype of values in the given sample.
+JSON or XML nodes) and find a common supertype of values in the sample.
 
@@
 \begin{array}{rcl}
- \hat{\sigma} &\narrow{=}& \nu_{\ident{opt}}\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.2em]
+ \hat{\sigma} &\narrow{=}& \nu_{\ident{opt}}\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                 &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bit} \lsep \ident{bool} \lsep \ident{string} 
  \\[0.6em] 
-       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \hat{\sigma}\;\,\kvd{option}\\[0.2em]
-                &\narrow{|}& [\sigma] \lsep \sigma_1 + \ldots + \sigma_n \\[0.2em]
-                &\narrow{|}& ~\top~ \lsep \kvd{null}
+       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \hat{\sigma}\;\,\kvd{option} \lsep [\sigma] \\[0.1em]
+              &\narrow{|}& \sigma_1 + \ldots + \sigma_n \lsep ~\top~ \lsep \kvd{null}
 \end{array}
@@
 The diagram shows only the basic structure and it does not explain relationships between records
-with same name, but different fields and between union types. This following definition makes
-this precise.
+with the same name and between union types. This following definition makes this precise.
 
@@
 \label{eq:sub-union3}\tag{U3}
-\hspace{-1em}\sigma_1 + \ldots + \sigma_n :> \sigma_{\pi(1)} + \ldots + \sigma_{\pi(n)} \quad &(\textnormal{for a permutation}~\pi) \\
+\hspace{-1em}\sigma_1 + \ldots + \sigma_n :> \sigma_{\pi(1)} + \ldots + \sigma_{\pi(n)} \quad &(\textnormal{permutation}~\pi) \\
 \label{eq:sub-union-el}\tag{U4}
-\sigma_1 + \ldots + \sigma_n :> \sigma_i \quad &(\textnormal{for}~i \in \{1 \ldots n\})
+\sigma_1 + \ldots + \sigma_n :> \sigma_i \quad &(\textnormal{for}~i \in 1 \ldots n)
 \end{align}
@@
 \label{eq:sub-record1}\tag{R1}
- \nu~\{ \nu_1\!:\!\sigma_1, \ldots, \nu_n\!:\!\sigma_n \} :> 
- \nu~\{ \nu_1\!:\!\sigma_1', \ldots, \nu_n\!:\!\sigma_n' \}
+ \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> 
+ \nu~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
  \;\; (\sigma_i :> \sigma_i')
@@
 \label{eq:sub-record2}\tag{R2}
- \nu~\{ \nu_1\!:\!\sigma_1, \ldots, \nu_n\!:\!\sigma_n \} :> 
- \nu~\{ \nu_1\!:\!\sigma_1, \ldots, \nu_m\!:\!\sigma_m \}
+ \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> 
+ \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_m\!:\!\sigma_m \}
  \;\; (m \geq n)
@@
 \begin{array}{l}
- \nu~\{ \nu_1\!:\!\sigma_1, \ldots, \nu_n\!:\!\sigma_n \} :> \\
- \nu~\{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, \ldots, \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
-\end{array} \qquad &(\textnormal{for a permutation}~\pi)
+ \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
+ \nu~\{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, .., \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
+\end{array} \qquad &(\textnormal{permutation}~\pi)
 \end{align}
@@
 \begin{array}{l}
- \nu~\{ \nu_1\!:\!\sigma_1, \ldots, \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null}, \ldots, \nu_{n+m}\!:\!\kvd{null} \} :> \\
- \nu~\{ \nu_1\!:\!\sigma_1, \ldots, \nu_n\!:\!\sigma_n \}
+ \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null}, .., \nu_{n+m}\!:\!\kvd{null} \} :> \\
+ \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
 \end{array}
@@
 
-% --------------------------------------------------------------------------------------------------
-
-\noindent
-Here is a summary of the key aspects of the definition:
-\begin{itemize}
-\item For numeric types (\ref{eq:sub-prim}), we infer a single most precise numeric type that 
-  can represent all values from a sample dataset, so we order types by the ranges they represent;
-  \ident{int} is a 32-bit integer, \ident{decimal} has a range of about $-1^{29}$ to $1^{29}$ and 
-  \ident{float} is a double-precision floating-point.
-
-\item The \ident{bit} type is a subtype of both \ident{int} and \ident{bool} (\ref{eq:sub-prim}). 
-  Thus, a sample $0,1$ has a type \ident{bit}, but $0,1,\kvd{true}$ is \ident{bool} and
-  $0,1,2$ becomes \ident{int} (for a sample $0,1,2, \kvd{true}$ we would need a union type).
-
-\item The \kvd{null} type is a subtype of all nullable types (\ref{eq:sub-null}), that is all 
-  $\sigma$ types excluding non-nullable types $\hat{\sigma}$). Any non-nullable type is also a 
-  subtype of its optional version (\ref{eq:sub-opt}).
-
-\item There is a top type (\ref{eq:sub-top}), but no bottom type. However it is possible to find common 
-  supertype of any two types, because a union type $\tau_1 + \ldots + \tau_n$ is a supertype of all its 
-  components (\ref{eq:sub-union-el}).
-
-\item The remaining rules for unins are standard. Subtype can have fewer alternatives 
-  (\ref{eq:sub-union1}), elements can be reordered (\ref{eq:sub-union2}) and subtyping is covariant 
-  (\ref{eq:sub-union3}). Similarly, the collection type is also covariant (\ref{eq:sub-col}).
-
-\item As usual, the subtyping on records is covariant (\ref{eq:sub-record1}), subtype can have additional 
-  fields (\ref{eq:sub-record2}) and fields can  be reordered (\ref{eq:sub-record3}). The interesting rule 
-  is the last one (\ref{eq:sub-record4}).  Together with covariance, it states that a subtype can omit some 
-  fields, provided that their types are nullable.
-\end{itemize}
-
-\noindent
-The rule that allows subtype to have fewer record elements (\ref{eq:sub-record4}) is particularly
-important. It allows us to prefer records in some cases. For example, given two samples 
-$\{ \ident{name}:\ident{string} \}$ and $\{ \ident{name}:\ident{string}, \ident{age}:\ident{int} \}$,
-we can find a common supertype $\{ \ident{name}:\ident{string}, \ident{age}:\ident{int}~\kvd{option} \}$
-which is also a record. For usability reasons, we prefer this to another common supertype
-$\{ \ident{name}:\ident{string} \} + \{ \ident{name}:\ident{string}, \ident{age}:\ident{int} \}$.
-The next section describes precisely how our inference algorithm works.
-
-It is worth noting that some of the aspects (such as 3 different numeric types and handling of 
-the \kvd{null} type) are specific to F\#. Similar problems will appear with most real-world
-languages and systems. JVM and OCaml all haves multiple numeric types, but they would handle \kvd{null}
-differently.
-
-We could also be more or less strict about missing data -- we choose to handle missing values 
-silently when we can (\kvd{null} becomes an empty collection), but we are explicit in other cases 
-(we infer \ident{string}~\kvd{option} as a hint to the user rather than treating \kvd{null} as an 
-empty string). These choices are based on experience with using F\# Data in practice.
-
 % -------------------------------------------------------------------------------------------------
@@
 \begin{figure*}[t]
-{\small{
-
 \begin{equation*}
@@
 \end{equation*}
-}}
-\caption{\small{Inference judgements that define the common supertype relation}}
+
+\caption{Inference judgements that define the common supertype relation}
 \label{fig:subtyping-cst}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+
+\noindent
+Here is a summary of the key aspects of the definition:
+\begin{itemize}
+\item For numeric types (\ref{eq:sub-prim}), we infer a single most precise numeric type that 
+  can represent all values from a sample dataset, so we order types by the ranges they represent;
+  \ident{int} is a 32-bit integer, \ident{decimal} has a range of about $-1^{29}$ to $1^{29}$ and 
+  \ident{float} is a double-precision floating-point.
+
+\item The \ident{bit} type is a subtype of both \ident{int} and \ident{bool} (\ref{eq:sub-prim}). 
+  Thus, a sample $0,1$ has a type \ident{bit}; $0,1,\kvd{true}$ is \ident{bool} and
+  $0,1,2$ becomes \ident{int} (for a sample $0,1,2, \kvd{true}$ we would need a union type).
+
+\item The \kvd{null} type is a subtype of all nullable types (\ref{eq:sub-null}), that is all 
+  $\sigma$ types excluding non-nullable types $\hat{\sigma}$). Any non-nullable type is also a 
+  subtype of its optional version (\ref{eq:sub-opt}).
+
+\item There is a top type (\ref{eq:sub-top}), but no bottom type. It is possible to find common 
+  supertype of any two types, because a union type $\tau_1 + \ldots + \tau_n$ is a supertype of all its 
+  components (\ref{eq:sub-union-el}).
+
+\item The remaining rules for unins are standard. Subtype can have fewer alternatives 
+  (\ref{eq:sub-union1}), elements can be reordered (\ref{eq:sub-union2}) and subtyping is covariant 
+  (\ref{eq:sub-union3}). Similarly, the collection type is also covariant (\ref{eq:sub-col}).
+
+\item As usual, the subtyping on records is covariant (\ref{eq:sub-record1}), subtype can have additional 
+  fields (\ref{eq:sub-record2}) and fields can  be reordered (\ref{eq:sub-record3}). The interesting rule 
+  is the last one (\ref{eq:sub-record4}).  Together with covariance, it states that a subtype can omit some 
+  fields, provided that their types are nullable.
+\end{itemize}
+
+\noindent
+The rule that allows subtype to have fewer record elements (\ref{eq:sub-record4}) is particularly
+important. It allows us to prefer records in some cases. For example, given two samples 
+$\{ \ident{name}:\ident{string} \}$ and $\{ \ident{name}:\ident{string}, \ident{age}:\ident{int} \}$,
+we can find a common supertype $\{ \ident{name}:\ident{string}, \ident{age}:\ident{int}~\kvd{option} \}$
+which is also a record. For usability reasons, we prefer this to another common supertype
+$\{ \ident{name}:\ident{string} \} + \{ \ident{name}:\ident{string}, \ident{age}:\ident{int} \}$.
+The next section describes precisely how our inference algorithm works.
+
+It is worth noting that some of the aspects (such as 3 different numeric types and handling of 
+the \kvd{null} type) are specific to F\#. Similar problems will appear with most real-world
+languages and systems. JVM and OCaml all haves multiple numeric types, but they would handle \kvd{null}
+differently.
+
+We could also be more or less strict about missing data -- we choose to handle missing values 
+silently when we can (\kvd{null} becomes an empty collection), but we are explicit in other cases 
+(we infer \ident{string}~\kvd{option} as a hint to the user rather than treating \kvd{null} as an 
+empty string). These choices are based on experience with using F\# Data in practice.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure*}
+\begin{equation*}
+\begin{array}{l}
+ \ident{asInt}(i) \,\reduce i\\
+ \ident{asInt}(d) \reduce i\quad(i = \lfloor d \rfloor)\\
+ \ident{asInt}(f) \reduce i\quad(i = \lfloor f \rfloor) \\
+ \ident{asInt}(t) \,\reduce i\quad(t~\text{\small represents}~i) \\
+ \ident{asInt}(\kvd{true}) \reduce 1\\
+ \ident{asInt}(\kvd{false}) \reduce 0\\
+\end{array}
+\;
+\begin{array}{l}
+ \ident{asDec}(i) \,\reduce d\quad(d = i)\\
+ \ident{asDec}(d) \reduce d\\
+ \ident{asDec}(f) \reduce d\quad(d = \lfloor f \rfloor) \\
+ \ident{asDec}(t) \,\reduce d\quad(t~\text{\small represents}~d) \\
+ \ident{asDec}(\kvd{true}) \reduce 1.0\\
+ \ident{asDec}(\kvd{false}) \reduce 0.0\\
+\end{array}
+\;
+\begin{array}{l}
+ \ident{asFloat}(i) \,\reduce f\quad(f = i)\\
+ \ident{asFloat}(d) \reduce f\quad(f = d)\\
+ \ident{asFloat}(f) \reduce f\\
+ \ident{asFloat}(t) \,\reduce f\quad(t~\text{\small represents}~f) \\
+ \ident{asFloat}(\kvd{true}) \reduce 1.0\\
+ \ident{asFloat}(\kvd{false}) \reduce 0.0\\
+\end{array}
+\;
+\begin{array}{l}
+ \ident{asStr}(i) \,\reduce t\quad(t~\text{\small represents}~i)\\
+ \ident{asStr}(d) \reduce t\quad(t~\text{\small represents}~d)\\
+ \ident{asStr}(f) \reduce t\quad(t~\text{\small represents}~f)\\
+ \ident{asStr}(t) \,\reduce t\\
+ \ident{asStr}(\kvd{true}) \reduce \str{true}\\
+ \ident{asStr}(\kvd{false}) \reduce \str{false}\\
+\end{array}
+\end{equation*}
+%
+\begin{equation*}
+\hspace{-1.5em}
+\begin{array}{l}
+ \ident{asBool}(1) \reduce \kvd{true}\\
+ \ident{asBool}(1.0) \reduce \kvd{true}\\
+ \ident{asBool}(\str{true}) \reduce \kvd{true}\\
+ \ident{asBool}(\str{1}) \reduce \kvd{true}
+ \\[0.6em]
+ \ident{isNull}(\kvd{null}) \reduce \kvd{true} \\
+ \ident{isNull}(\_) \reduce \kvd{false} 
+\end{array}
+\hspace{2.6em}
+\begin{array}{l}
+ \ident{asBool}(0) \reduce \kvd{false}\\
+ \ident{asBool}(0.0) \reduce \kvd{false}\\
+ \ident{asBool}(\str{false}) \reduce \kvd{false}\\
+ \ident{asBool}(\str{0}) \reduce \kvd{false}\\
+ \ident{asBool}(b) \reduce b
+ \\[0.6em]
+ \ident{getChildren}(\kvd{null}) \reduce [~]
+\end{array}
+\hspace{3.0em}
+\begin{array}{l}
+ \ident{getField}_{\nu,\nu_i}(\nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
+ \ident{getField}_{\bot, \nu_i}(\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
+ \ident{getField}_{\nu,\nu'}(\nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )\\
+ \ident{getField}_{\bot, \nu'}(\{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )
+ \\[0.6em]
+ \ident{getChildren}([s_1; \ldots; s_n]) \reduce [s_1; \ldots; s_n]
+\end{array}
+\end{equation*}
+%
+\begin{equation*}
+\hspace{-1.5em}
+\begin{array}{l}
+ \ident{isTag}_\ident{string}(t) \reduce \kvd{true} \\
+ \ident{isTag}_\ident{bool}(v) \reduce \kvd{true} \quad(\textnormal{when}~v\in{0,1,0.0, 1.0, \kvd{true},\kvd{false}} )\\
+ \ident{isTag}_\ident{number}(v) \reduce \kvd{true} \quad(\textnormal{when}~v=i, v=d, v=f) \\
+ \ident{isTag}_\ident{collection}([s_1; \ldots; s_n]) \reduce \kvd{true} \\
+\end{array}
+\hspace{2.5em}
+\begin{array}{l}
+ \ident{isTag}_\ident{collection}(\kvd{null}) \reduce \kvd{true} \\
+ \ident{isTag}_\ident{rec-anon}(\{ \nu_1\mapsto s_1, \ldots, \nu_n\mapsto s_n \}) \reduce \kvd{true} \\
+ \ident{isTag}_{\ident{rec-named}~\nu} (\nu\;\{ \nu_1\mapsto s_1, \ldots, \nu_n\mapsto s_n \}) \reduce \kvd{true} \\
+ \ident{isTag}_{\_}(\_) \reduce \kvd{false} \\
+\end{array}
+\end{equation*}
+
+\caption{Reduction rules for conversion functions}
+\label{fig:op-conversions}
 \end{figure*}
@@
 which are unrelated. Our inference algorithm always prefers records over unions (Theorem~\ref{thm:no-unions}). 
-
 The definition of the \emph{common supertype} relation uses a number of auxiliary definitions 
@@
 %
-The tag of a type is obtained using a helper function $\tytagof{(-)} : \sigma \rightarrow \tytag$:
+The tag of a type is obtained using a function $\tytagof{(-)} : \sigma \rightarrow \tytag$:
 %
@@
  \tytagof([\sigma]) &\narrow{=}& \ident{collection}\\
+\end{array}
+\end{equation*}
+\begin{equation*}
+\begin{array}{rcl}
  \tytagof(\sigma + \ldots + \sigma) &\narrow{=}& \ident{union}\\
@@
 %
-The remaining rules are mostly straightforward. For collections, we find the common supertype of 
+The remaining rules are straightforward. For collections, we find the common supertype of 
 its elements (\emph{list}); for compatible primitive types, we choose one of the two (\emph{prim})
 and a common supertype with \kvd{null} is either the type itself (\emph{null-1}) or an option 
-type (\emph{null-2}).
+ (\emph{null-2}).
 
 \paragraph{Properties.}
-The common supertype relation finds a single supertype (among multiple possible candidates).
+The common supertype relation finds a single supertype (among multiple candidates).
 The following theorems specify that the found type is uniquely defined (up to reordering of 
@@
 \begin{proof}
-By induction over the common supertype derivation using $\vdash$.
+By induction over the common supertype derivation $\vdash$.
 \end{proof}
@@
 
-\begin{figure*}
-{\small{
-\begin{equation*}
-\hspace{3em}
-\begin{array}{l}
- \ident{asBool}(i) \reduce \kvd{true}  \quad(i = 1)\\
- \ident{asBool}(d) \reduce \kvd{true}  \quad(d = 1.0)\\
- \ident{asBool}(f) \reduce \kvd{true}  \quad(f = 1.0)\\
-\end{array}
-\qquad
-\begin{array}{l}
- \ident{asBool}(i) \reduce \kvd{false} \quad(i = 0)\\
- \ident{asBool}(d) \reduce \kvd{false} \quad(d = 0.0)\\
- \ident{asBool}(f) \reduce \kvd{false} \quad(f = 0.0)\\
-\end{array}
-\qquad
-\begin{array}{l}
- \ident{asBool}(s) \reduce \kvd{true}  \quad(s=\str{true} \vee s=\str{1}) \\
- \ident{asBool}(s) \reduce \kvd{false} \quad(s=\str{false} \vee s=\str{0}) \\
- \ident{asBool}(b) \reduce b\quad(\textnormal{when}~b=\kvd{true}\vee b=\kvd{false})
-\end{array}
-\end{equation*}
-%
-\begin{equation*}
-\hspace{-1em}
-\begin{array}{l}
- \ident{asInt}(i) \,\reduce i\\
- \ident{asInt}(d) \reduce i\quad(i = \lfloor d \rfloor)\\
- \ident{asInt}(f) \reduce i\quad(i = \lfloor f \rfloor) \\
- \ident{asInt}(s) \,\reduce i\quad(s~\text{represents}~i) \\
- \ident{asInt}(\kvd{true}) \reduce 1\\
- \ident{asInt}(\kvd{false}) \reduce 0\\
-\end{array}
-\quad
-\begin{array}{l}
- \ident{asDec}(i) \,\reduce d\quad(d = i)\\
- \ident{asDec}(d) \reduce d\\
- \ident{asDec}(f) \reduce d\quad(d = \lfloor f \rfloor) \\
- \ident{asDec}(s) \,\reduce d\quad(s~\text{represents}~d) \\
- \ident{asDec}(\kvd{true}) \reduce 1.0\\
- \ident{asDec}(\kvd{false}) \reduce 0.0\\
-\end{array}
-\quad
-\begin{array}{l}
- \ident{asFloat}(i) \,\reduce f\quad(f = i)\\
- \ident{asFloat}(d) \reduce f\quad(f = d)\\
- \ident{asFloat}(f) \reduce f\\
- \ident{asFloat}(s) \,\reduce f\quad(s~\text{represents}~f) \\
- \ident{asFloat}(\kvd{true}) \reduce 1.0\\
- \ident{asFloat}(\kvd{false}) \reduce 0.0\\
-\end{array}
-\quad
-\begin{array}{l}
- \ident{asStr}(i) \,\reduce s\quad(s~\text{represents}~i)\\
- \ident{asStr}(d) \reduce s\quad(s~\text{represents}~d)\\
- \ident{asStr}(f) \reduce s\quad(s~\text{represents}~f)\\
- \ident{asStr}(s) \,\reduce s\\
- \ident{asStr}(\kvd{true}) \reduce \str{true}\\
- \ident{asStr}(\kvd{false}) \reduce \str{false}\\
-\end{array}
-\end{equation*}
-%
-\begin{equation*}
-\hspace{8em}
-\begin{array}{l}
- \ident{getChildren}([s_1; \ldots; s_n]) \reduce [s_1; \ldots; s_n]\\
- \ident{getChildren}(\kvd{null}) \reduce [~]\\
-\end{array}
-\qquad\qquad\qquad
-\begin{array}{l}
- \ident{getField}(\nu,\nu_i,\nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
- \ident{getField}(\epsilon,\nu_i,\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
-\end{array}
-\end{equation*}
-}}
-\caption{\small{Reduction rules for conversion functions}}
-\label{fig:op-conversions}
-\end{figure*}
-
-% TODO: We'll need more things here for working with options 
-% Also, 'getField' should return 'null' if the field does not exist
-% (which we will presumably handle nicely thanks to handling of options)
-
-% ==================================================================================================
-
 \section{Formalising type providers}
@@
 \emph{structural value}\footnote{Here, we diverge from the actual implementation in F\# Data which 
-uses a different implementation for each format (reusing existing libraries).}. 
-A structural value are first-order and can be one of the primitive values 
-(Boolean, string, integer, decimal and float), a missing value (\kvd{null}), a record 
-(named or unnamed) and a collection:
+uses a different implementation for each format (reusing existing libraries).}. Structural values are 
+first-order and can be one of the primitive values (Boolean, string, integer, decimal and float), a 
+missing value (\kvd{null}), a collection and a record (named or unnamed):
 %
@@
 \begin{array}{lcl}
- s  &\narrow{=}& i \lsep d \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
+ s  &\narrow{=}& i \lsep d \lsep f \lsep t \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
     &\narrow{|}& [s_1; \ldots; s_n] \\[0.1em]
@@
 \noindent
+The first few cases represent primitive values ($i$ for integers, $d$ for decimals, $f$ for floating
+point numbers and $t$ for strings). A collection is written as a list of values in square brackets, 
+separated by semicolons. A record can optionally start with its name $\nu$, followed by a sequence of 
+field assignments $\nu_i \mapsto s_i$ written in curly brackets.
+
 As suggested by subtyping, our system permits certain runtime conversions (\emph{e.g.}~a numerical 
@@
 \begin{array}{lcl}
- op &\narrow{=}& \ident{asInt}(s) \lsep \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{asBool}(s) \\
-    &\narrow{|}& \ident{asStr}(s) \lsep \ident{getChildren}(s) \lsep \ident{getField}(\nu_{\ident{opt}}, \nu, s)
+ op &\narrow{=}& \ident{asInt}(s) \lsep \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{asBool}(s) \lsep \ident{asStr}(s) \\
+    &\narrow{|}& \ident{getChildren}(s) \lsep \ident{getField}_{\nu_{\ident{opt}}, \nu}(s) \lsep \ident{isNull}(s) \lsep \ident{isTag}_t(s)
 \end{array}
@@
 to note that the relativised property we intend to prove about F\# Data provides \emph{sufficient}, but
-not \emph{necessary} conditions. This means that a type provided based on sample is guaranteed to work 
+not \emph{necessary} conditions. This means that a type provided based on a sample is guaranteed to work 
 correctly on certain inputs, but may also handle several additional inputs.
@@
 
-% TODO: Finish this
+\begin{itemize}
+\item \textbf{Widening operations.} The widening operations are required by the subtyping. These include
+  converting to a wider numerical type (integer to decimal and float, decimal to float, converting 0 and 
+  1 to Boolean).
+
+\item \textbf{Conversions.} Additional conversions are not strictly required, but prove useful in practice.
+  These include converting to a smaller numerical type (such as float to integer) written as $i=\lfloor f\rfloor$,
+  which is not defined in case of arithmetic overflow; conversion to and from string values written as
+  ``$t$~represents~$i$'', which parses or formats a non-string value and also trating of decimals 
+  0.0 and 1.0 as Booleans.
+
+\item \textbf{Structural operations.} Aside from primitives, the \ident{getChildren} operation returns treats
+  \kvd{null} as an empty collection. The \ident{getField} function is defined separately for named and unnamed
+  records (we write $\bot$ for a name of an unnamed record). Accessing a field of a named record checks that 
+  the expected record name matches the actual.
+
+\item \textbf{Helper functions.} The semantics also defines two helper functions. The \ident{isNull} function
+  is used to check whether value is \kvd{null} and \ident{isTag} is used to test whether a value can be treated 
+  as a value of type associated with the specified tag (as defined in Section~\ref{sec:inference-commonsuper}).
+  The last line defines a ``catch all'' pattern, which returns \kvd{false} for all remaining cases.
+\end{itemize}
+
+% -------------------------------------------------------------------------------------------------
 
@@
 
-Types are primitive types or a class or a StructVal, which represents different structured values
-
-L
-\begin{equation*}
-\begin{array}{l}
- \tau = \ident{int}\lsep\ident{decimal}\lsep\ident{float}\lsep\ident{bool}\lsep\ident{string}\lsep \ident{StructVal} \lsep C \\[0.2em]
- L = \kvd{type}~C(\overline{f:\tau}) = \overline{M} \\[0.2em]
- M = \kvd{member}~N:\tau=e \\[0.2em]
- v = ...
-\end{array}
-\end{equation*}
-
-.. todo -- evaluation goes here too
+The semantics fragment introduced in the previous section discusses values and operations that are used by
+F\# Data. This section adds a minimal subset of the standard F\# language. We focus on the object system
+(which is what F\# type providers produce) and so the following is based on Featherweight Java. However, we
+only need class types (without inheritance) with properties. A class has a single implicit constructor and 
+the declaration closes over constructor parameters. In addition, we include special constructs for working
+with option types and lists.
+
+\begin{equation*}
+\begin{array}{rcl}
+ \tau &\narrow{=}& \ident{int} \lsep \ident{decimal} \lsep \ident{float} \lsep \ident{bool} \lsep \ident{string} \\[0.0em]
+      &\narrow{|}& C \lsep \ident{StructVal} \lsep \ident{list}\langl\tau\rangl \lsep \ident{option}\langl\tau\rangl \\[0.6em]
+ L &\narrow{=}& \kvd{type}~C(\overline{x:\tau}) = \overline{M} \\[0.0em]
+ M &\narrow{=}& \kvd{member}~N:\tau=e \\[0.6em]
+ v &\narrow{=}& s \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \\[0.0em]
+ e &\narrow{=}& op \lsep e.N \lsep \kvd{new}~C(\overline{e}) \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3}\\
+   &\narrow{|}& \ident{Some}(e)\lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
+   &\narrow{|}& [e_1;\ldots;e_n]\lsep\kvd{match}~e~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 \\
+\end{array}
+\end{equation*}
+%
+In the above definition, $\tau$ denotes types including a special type \ident{StructVal} for all structural 
+values $s$; a class definition $L$ consists of a constructor and zero or more members. Values $v$ include 
+previously defined structural values $s$ and values for the option type; finally expressions $e$ include all 
+previously defined operations $op$, class construction, member access, conditionals and expressions for working 
+with option values and lists.
+
+In this paper, we do not define the typing of the F\# subset introduced above as our goal is not to prove
+type safety for the core language. For the relativized soundness proof in Section~\ref{sec:safety}, it is
+sufficient to discuss reduction.
+
+\paragraph{Reduction.} The reduction relation is of the form $e \reduce e'$. We also write 
+$e \reduce^{*} e'$ to denote the reflexive and transitive closure of $\reduce$. The reduction rules
+for primitive functions $op$ were discussed earlier in Figure~\ref{fig:op-conversions}. The rules 
+for other expressions defined above are provided in Figure~\ref{fig:ff-reduction}.
+
+The (\emph{ctx}) rule performs a reduction inside a sub-expression specified by an evaluation context.
+This models the eager evaluation order of F\#. An evaluation context $E$ is defined as:
+
+\vspace{-0.5em}
+\begin{equation*}
+\begin{array}{rcl}
+ E &\narrow{=}& [\cdot] \lsep f(E) \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e}) \lsep \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2 \\[0.1em]
+   &\narrow{|}& \ident{Some}(E) \lsep \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
+   &\narrow{|}& [\overline{v};E;\overline{e}] \lsep \kvd{match}~E~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 
+ \\[0.6em]
+ f &\narrow{\in}& \{\; \ident{asInt}, \ident{asDec}, \ident{asFloat}, \ident{asBool}, \ident{asStr}, \\[0.1em]
+              && ~~\; \ident{getField}, \ident{getChildren}, \ident{isNull}, \ident{isTag}_t \;\}
+\end{array} 
+\end{equation*}
+
+\noindent
+The evaluation first reduces arguments of functions and the evaluation proceeds from left to right 
+as denoted by $\overline{v}, E, \overline{e}$ in constructor arguments or $\overline{v};E;\overline{e}$
+in list initialization.
+
+We write $e[\overline{x} \leftarrow \overline{v}]$ for the result of replacing variables $\overline{x}$ by
+values $\overline{v}$ in expression. The (\emph{member}) rule reduces a member access using a class 
+definition in the assumption to obtain the body of the expression. Finally, the remaining six rules
+provide standard reductions for conditionals and pattern matching.
+
+The language presented here is very simple. In particular, it does not provide any form of recursion
+and so all expressions reduce to a value in a finite number of steps or get stuck due to an error
+condition. An error condition can be a wrong argument passed to conditional, pattern matching or
+one of the conversion functions from Figure~\ref{fig:op-conversions}.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\begin{equation*}
+\begin{array}{rl}
+ \textnormal{\footnotesize{(ctx)}}&
+ \hspace{-0.4em}
+ \inference
+  { e \reduce e' }
+  { E[e] \reduce E[e']}\\
+ \\
+ \textnormal{\footnotesize{(member)}}&
+ \hspace{-0.4em}
+ \inference
+ { \kvd{type}~C(\overline{x:\tau})=\ldots \kvd{member}~N_i : \tau_i = e_i \ldots }
+ { (\kvd{new}~C(\overline{v})).N_i \reduce e_i[\overline{x} \leftarrow \overline{v}] }\\
+ \\
+ \textnormal{\footnotesize{(cond1)}}&
+ \hspace{-0.4em}
+ \kvd{if}~\kvd{true}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_1 \\
+ \\
+ \textnormal{\footnotesize{(cond2)}}&
+ \hspace{-0.4em}
+ \kvd{if}~\kvd{false}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_2 \\
+ \\
+ \textnormal{\footnotesize{(match1)}}&
+ \hspace{-1em}
+ \begin{array}{l}
+  \kvd{match}~\ident{None}~\kvd{with} \\
+  \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
+ \end{array} \hspace{-0.5em} ~\reduce~ e_2 \\
+ \\
+ \textnormal{\footnotesize{(match2)}}&
+ \hspace{-1em}
+ \begin{array}{l}
+    \kvd{match}~\ident{Some}(v)~\kvd{with} \\
+    \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
+ \end{array} \hspace{-0.5em} ~\reduce~ e_1[x\leftarrow v]\\
+ \\
+ \textnormal{\footnotesize{(match3)}}&
+ \hspace{-1em}
+ \begin{array}{l}
+  \kvd{match}~[v_1;\ldots;v_n]~\kvd{with} \\[0em]
+  [x_1;\ldots;x_m ] \rightarrow e_1 \,|\, \_ \rightarrow e_2
+ \end{array} \hspace{-0.5em} ~\reduce~ e_2\quad(m\neq n) \\
+ \\
+ \textnormal{\footnotesize{(match4)}}&
+ \hspace{-1em}
+ \begin{array}{l}
+  \kvd{match}~[v_1;\ldots;v_n]~\kvd{with} \\[0em]
+  [x_1;\ldots;x_n ] \rightarrow e_1 \,|\, \_ \rightarrow e_2
+ \end{array} \hspace{-0.5em} ~\reduce~ e_1[\overline{x}\leftarrow\overline{v}] \\
+\end{array}
+\end{equation*}
+
+\caption{Remaining reduction rules for Featherweight F\#}
+\label{fig:ff-reduction}
+\end{figure}
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure*}
+\begin{multicols}{2}
+% Primitive structural types become corresponding F# types and conversion is inserted
+\begin{equation*}
+\begin{array}{l}
+ \sem{\sigma_p}_e = \tau_p,op(e),\emptyset\quad\textnormal{where} \\[0.6em]
+\quad\sigma_p, \tau_p, op\in  \{~ (\ident{bit}, \ident{bool}, \ident{asBool}),  (\ident{bool}, \ident{bool}, \ident{asBool})\\[0.2em]
+\hspace{2.9em} (\ident{int}, \ident{int}, \ident{asInt}), (\ident{decimal},\ident{decimal},\ident{asDec}),\\[0.2em]
+\hspace{2.9em} (\ident{float},\ident{float},\ident{asFloat}), (\ident{string},\ident{string},\ident{asStr}) ~\}\\[0.6em]
+\end{array}
+\end{equation*}
+%
+% Records become classes
+\begin{equation*}
+\begin{array}{l}
+ \sem{\nu_{\ident{opt}}\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}}_e = \\[0.1em]
+ \quad C, \kvd{new}~C(e), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
+ \qquad \;\;C~\textnormal{is a fresh class name} \\[0.1em]
+ \qquad \;\,\,L = \kvd{type}~C(v:\ident{StructVal})~=~M_1\ldots M_n  \\[0.1em]
+ \qquad M_i = \kvd{member}~\nu_i:\tau_i=e_i\\[0.1em]
+ \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_{e'},\;\; e'=\ident{getField}_{\nu_{\ident{opt}}, \nu_i}(e)\\[0.6em]
+\end{array}
+\end{equation*}
+%
+% Collections
+\begin{equation*}
+\begin{array}{l}
+ \sem{\,[\sigma]\,}_e = \ident{list}\langl\tau\rangl, \ident{getChildren}(e'), L \;\;\textnormal{where}~ \tau, e', L = \sem{\hat{\sigma}}_e
+\end{array}
+\end{equation*}
+
+% Option values
+\begin{equation*}
+\begin{array}{l}
+ \sem{\hat{\sigma}\;\,\kvd{option}}_e = \\[0.2em]
+ \hspace{1.25em} \ident{option}\langl\tau\rangl, \kvd{if}~\ident{isNull}~e'~\kvd{then}~\ident{None}~\kvd{else}~\ident{Some}(e'), L\\[0.2em] 
+ \hspace{1.25em} \textnormal{where}~\tau, e', L = \sem{\hat{\sigma}}_e
+\end{array}
+\end{equation*}
+%
+% Sum type
+\begin{equation*}
+\begin{array}{l}
+ \sem{\sigma_1 + \ldots + \sigma_n}_e = \\[0.1em]
+ \quad C, \kvd{new}~C(e), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
+ \qquad \;\;C~\textnormal{is a fresh class name} \\[0.1em]
+ \qquad \;\,\,L = \kvd{type}~C(v:\ident{StructVal})~=~M_1\ldots M_n \\[0.1em]
+ \qquad M_i = \kvd{member}~\nu_i:\ident{option}\langl\tau_i\rangl=e_i'\\[0.1em]
+ \qquad \;\,\,e_i' = \kvd{if}~\ident{isNull}(e_i)~\kvd{then}~\ident{None}~\kvd{else} \\
+ \hspace{4.8em}     \kvd{if}~\ident{isTag}_{t_i}{(e_i)}~\kvd{then}~
+     \ident{Some}(e_1)~\kvd{else}~\ident{None} \\[0.1em]
+ \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_e,\;\; t_i = \tytagof{(\sigma_i)},\;\; \nu_i=\nameoftag{(t_i)}
+\end{array}
+\end{equation*}
+%
+% Anything else
+\begin{equation*}
+\begin{array}{l}
+ \sem{\top}_v = \sem{\kvd{null}}_v = \ident{StructVal}, v, \emptyset
+\end{array}
+\end{equation*}
+\end{multicols}
+
+\noindent
+\begin{equation*}
+\qquad
+\qquad
+\begin{array}{rcl}
+ \nameoftag(\ident{string}) &\narrow{=}& \ident{String} \\
+ \nameoftag(\ident{bool}) &\narrow{=}& \ident{Boolean} \\
+\end{array}
+\qquad
+\begin{array}{rcl}
+ \nameoftag(\ident{number}) &\narrow{=}& \ident{Number} \\
+ \nameoftag(\ident{collection}) &\narrow{=}& \ident{List} \\
+\end{array}
+\qquad
+\begin{array}{rcl}
+ \nameoftag(\ident{rec-anon}) &\narrow{=}& \ident{Record} \\
+ \nameoftag(\ident{rec-named}\;\nu) &\narrow{=}& \nu \\
+\end{array}
+\end{equation*}
+
+\caption{Type provider -- generation of featherweight F\# types from inferred structural types}
+\label{fig:tp-generation}
+\end{figure*}
+
+% -------------------------------------------------------------------------------------------------
 
@@
 
-How we turn inferred structural type into F\# type
+So far, we defined the type inference algorithm for structured data formats which produces a structured
+type $\sigma$ based on one or more sample documents (Section~\ref{sec:inference}). In Section~\ref{sec:formal-ff},
+we defined a simplified model of F\# language and evaluation, which also included conversions that are
+modelled based on the F\# Data library runtime (Section~\ref{sec:formal-tp}). In this section, we define how
+the type providers work, which links the two aspects discussed so far.
+
+The type providers for XML, CSV and JSON all work in the same way. They take sample documents and infer a common
+supertype $\sigma$ from the samples. From $\sigma$, they then generate F\# types that are then exposed to the
+programmer\footnote{The actual implementation provides \emph{erased types} as described in [?]. Here, we treat
+the code as actually generated. This is an acceptable simplification, because F\# Data type providers do not 
+rely on laziness that is available through erased types.}.
+
+When generating the F\# types, the type provider needs to produce the corresponding F\# type, expression that
+wraps a structured document value (of type \ident{StructVal}) and a collection of class definitions. We express
+it using the following mapping:
+%
+\begin{equation*}
+\sem{-}_e : \sigma \rightarrow \tau \times e' \times L
+\end{equation*}
+%
+The semantics is parameterized by an expression $e$, which represents code to obtain a structured value that is
+being wrapped. Then, given an inferred structured type $\sigma$, the type provider produces an F\# type $\tau$,
+an expression $e'$ which constructs a value of type $\tau$ from $e$ and also a set of class definitions $L$.
+
+The rules that define $\sem{-}_e$ are shown in Figure~\ref{fig:tp-generation}.
 
 % ==================================================================================================
+
+\newpage
 
@@
 \label{sec:safety}
+~
+
+We presented the system together with its syntactic equational theory and categorical semantics.
+The equational theory is presented in order to explain how the systems work, but it also
+provides a basis for an operational semantics for concrete systems. Exploring these, 
+and their connection to the denotational semantics, is further work.
 

### 2015-02-23 icfp-2015/paper.tex
Adding more stuff to semantics

@@
 % TODO: Disambigurate 's' as structural value and 's' as string
+% TODO: Explain why we can only provide simple classes (!)
+% TODO: Replace \nu_opt with \nu_? or something else that does not look shit
 
@@
 
+\newcommand{\llangl}{\langl\hspace{-0.35em}\langl}
+\newcommand{\rrangl}{\rangl\hspace{-0.35em}\rangl}
+
 \definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
@@
 \newcommand{\sem}[1]{\llbracket #1 \rrbracket}
+\newcommand{\semalt}[1]{\llangl #1 \rrangl}
 
@@
            {tomas@tomasp.net}
-%\authorinfo{Name2\and Name3}
-%           {Affiliation2/3}
-%           {Email2/3}
-
+\authorinfo{Gustavo Guerra}
+           {Microsoft Corporation, London}
+           {gustavo@codebeside.org}
+\authorinfo{Don Syme}
+           {Microsoft Research, Cambridge}
+           {dsyme@microsoft.com}
 \maketitle
@@
 \begin{array}{l}
- \ident{getField}_{\nu,\nu_i}(\nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
- \ident{getField}_{\bot, \nu_i}(\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
- \ident{getField}_{\nu,\nu'}(\nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )\\
- \ident{getField}_{\bot, \nu'}(\{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )
+ \ident{getField}(\nu,\nu_i, \nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
+ \ident{getField}(\bot, \nu_i, \{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
+ \ident{getField}(\nu,\nu', \nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )\\
+ \ident{getField}(\bot, \nu', \{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )
  \\[0.6em]
@@
 \begin{array}{l}
- \ident{isTag}_\ident{string}(t) \reduce \kvd{true} \\
- \ident{isTag}_\ident{bool}(v) \reduce \kvd{true} \quad(\textnormal{when}~v\in{0,1,0.0, 1.0, \kvd{true},\kvd{false}} )\\
- \ident{isTag}_\ident{number}(v) \reduce \kvd{true} \quad(\textnormal{when}~v=i, v=d, v=f) \\
- \ident{isTag}_\ident{collection}([s_1; \ldots; s_n]) \reduce \kvd{true} \\
-\end{array}
-\hspace{2.5em}
-\begin{array}{l}
- \ident{isTag}_\ident{collection}(\kvd{null}) \reduce \kvd{true} \\
- \ident{isTag}_\ident{rec-anon}(\{ \nu_1\mapsto s_1, \ldots, \nu_n\mapsto s_n \}) \reduce \kvd{true} \\
- \ident{isTag}_{\ident{rec-named}~\nu} (\nu\;\{ \nu_1\mapsto s_1, \ldots, \nu_n\mapsto s_n \}) \reduce \kvd{true} \\
- \ident{isTag}_{\_}(\_) \reduce \kvd{false} \\
+ \ident{isTag}(\ident{string}, t) \reduce \kvd{true} \\
+ \ident{isTag}(\ident{bool}, v) \reduce \kvd{true} \quad(\textnormal{when}~v\in{0,1,0.0, 1.0, \kvd{true},\kvd{false}} )\\
+ \ident{isTag}(\ident{number}, v) \reduce \kvd{true} \quad(\textnormal{when}~v=i, v=d, v=f) \\
+ \ident{isTag}(\ident{collection}, [s_1; \ldots; s_n]) \reduce \kvd{true} \\
+\end{array}
+\hspace{1.8em}
+\begin{array}{l}
+ \ident{isTag}(\ident{collection}, \kvd{null}) \reduce \kvd{true} \\
+ \ident{isTag}(\ident{rec-anon}, \{ \nu_1\mapsto s_1, \ldots, \nu_n\mapsto s_n \}) \reduce \kvd{true} \\
+ \ident{isTag}(\ident{rec-named}~\nu, \nu\;\{ \nu_1\mapsto s_1, \ldots, \nu_n\mapsto s_n \}) \reduce \kvd{true} \\
+ \ident{isTag}(\_, \_) \reduce \kvd{false} \\
 \end{array}
@@
 As suggested by subtyping, our system permits certain runtime conversions (\emph{e.g.}~a numerical 
-value \num{1} can be treated as \kvd{true}). To capture these, the operational semantics accesses 
-values using the following operations:
+value \num{1} can be treated as \kvd{true}). These are captured by the following operations:
 %
@@
 \begin{array}{lcl}
- op &\narrow{=}& \ident{asInt}(s) \lsep \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{asBool}(s) \lsep \ident{asStr}(s) \\
-    &\narrow{|}& \ident{getChildren}(s) \lsep \ident{getField}_{\nu_{\ident{opt}}, \nu}(s) \lsep \ident{isNull}(s) \lsep \ident{isTag}_t(s)
+ op &\narrow{=}& \ident{asInt}(s) \lsep \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{asBool}(s)  \\
+    &\narrow{|}& \ident{asStr}(s) \lsep \ident{getChildren}(s) \lsep \ident{getField}(\nu_{\ident{opt}}, \nu, s)\\
+    &\narrow{|}& \ident{isNull}(s) \lsep \ident{isTag}(t, s)
 \end{array}
@@
 The reduction rules for these operations are defined in Figure~\ref{fig:op-conversions}. It is important
-to note that the relativised property we intend to prove about F\# Data provides \emph{sufficient}, but
-not \emph{necessary} conditions. This means that a type provided based on a sample is guaranteed to work 
+to note that the relativised property we prove about F\# Data specifies \emph{sufficient}, but
+not \emph{necessary} conditions. That is, a type provided based on a sample is guaranteed to work 
 correctly on certain inputs, but may also handle several additional inputs.
@@
 
-\subsection{Featherweight F\#}
-\label{sec:formal-ff}
-
-The semantics fragment introduced in the previous section discusses values and operations that are used by
-F\# Data. This section adds a minimal subset of the standard F\# language. We focus on the object system
-(which is what F\# type providers produce) and so the following is based on Featherweight Java. However, we
-only need class types (without inheritance) with properties. A class has a single implicit constructor and 
-the declaration closes over constructor parameters. In addition, we include special constructs for working
-with option types and lists.
-
-\begin{equation*}
-\begin{array}{rcl}
- \tau &\narrow{=}& \ident{int} \lsep \ident{decimal} \lsep \ident{float} \lsep \ident{bool} \lsep \ident{string} \\[0.0em]
-      &\narrow{|}& C \lsep \ident{StructVal} \lsep \ident{list}\langl\tau\rangl \lsep \ident{option}\langl\tau\rangl \\[0.6em]
- L &\narrow{=}& \kvd{type}~C(\overline{x:\tau}) = \overline{M} \\[0.0em]
- M &\narrow{=}& \kvd{member}~N:\tau=e \\[0.6em]
- v &\narrow{=}& s \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \\[0.0em]
- e &\narrow{=}& op \lsep e.N \lsep \kvd{new}~C(\overline{e}) \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3}\\
-   &\narrow{|}& \ident{Some}(e)\lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
-   &\narrow{|}& [e_1;\ldots;e_n]\lsep\kvd{match}~e~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 \\
-\end{array}
-\end{equation*}
-%
-In the above definition, $\tau$ denotes types including a special type \ident{StructVal} for all structural 
-values $s$; a class definition $L$ consists of a constructor and zero or more members. Values $v$ include 
-previously defined structural values $s$ and values for the option type; finally expressions $e$ include all 
-previously defined operations $op$, class construction, member access, conditionals and expressions for working 
-with option values and lists.
-
-In this paper, we do not define the typing of the F\# subset introduced above as our goal is not to prove
-type safety for the core language. For the relativized soundness proof in Section~\ref{sec:safety}, it is
-sufficient to discuss reduction.
-
-\paragraph{Reduction.} The reduction relation is of the form $e \reduce e'$. We also write 
-$e \reduce^{*} e'$ to denote the reflexive and transitive closure of $\reduce$. The reduction rules
-for primitive functions $op$ were discussed earlier in Figure~\ref{fig:op-conversions}. The rules 
-for other expressions defined above are provided in Figure~\ref{fig:ff-reduction}.
-
-The (\emph{ctx}) rule performs a reduction inside a sub-expression specified by an evaluation context.
-This models the eager evaluation order of F\#. An evaluation context $E$ is defined as:
-
-\vspace{-0.5em}
-\begin{equation*}
-\begin{array}{rcl}
- E &\narrow{=}& [\cdot] \lsep f(E) \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e}) \lsep \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2 \\[0.1em]
-   &\narrow{|}& \ident{Some}(E) \lsep \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
-   &\narrow{|}& [\overline{v};E;\overline{e}] \lsep \kvd{match}~E~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 
- \\[0.6em]
- f &\narrow{\in}& \{\; \ident{asInt}, \ident{asDec}, \ident{asFloat}, \ident{asBool}, \ident{asStr}, \\[0.1em]
-              && ~~\; \ident{getField}, \ident{getChildren}, \ident{isNull}, \ident{isTag}_t \;\}
-\end{array} 
-\end{equation*}
-
-\noindent
-The evaluation first reduces arguments of functions and the evaluation proceeds from left to right 
-as denoted by $\overline{v}, E, \overline{e}$ in constructor arguments or $\overline{v};E;\overline{e}$
-in list initialization.
-
-We write $e[\overline{x} \leftarrow \overline{v}]$ for the result of replacing variables $\overline{x}$ by
-values $\overline{v}$ in expression. The (\emph{member}) rule reduces a member access using a class 
-definition in the assumption to obtain the body of the expression. Finally, the remaining six rules
-provide standard reductions for conditionals and pattern matching.
-
-The language presented here is very simple. In particular, it does not provide any form of recursion
-and so all expressions reduce to a value in a finite number of steps or get stuck due to an error
-condition. An error condition can be a wrong argument passed to conditional, pattern matching or
-one of the conversion functions from Figure~\ref{fig:op-conversions}.
-
-% -------------------------------------------------------------------------------------------------
-
 \begin{figure}
@@
 \begin{array}{rl}
- \textnormal{\footnotesize{(ctx)}}&
- \hspace{-0.4em}
- \inference
-  { e \reduce e' }
-  { E[e] \reduce E[e']}\\
- \\
  \textnormal{\footnotesize{(member)}}&
@@
  { (\kvd{new}~C(\overline{v})).N_i \reduce e_i[\overline{x} \leftarrow \overline{v}] }\\
+ \\
+ \textnormal{\footnotesize{(ctx)}}&
+ \hspace{-0.4em}
+  E[e] \reduce E[e'] \qquad\qquad(\textnormal{when}~e \reduce e')\\
  \\
@@
  \end{array} \hspace{-0.5em} ~\reduce~ e_1[\overline{x}\leftarrow\overline{v}] \\
+ \\
+ \textnormal{\footnotesize{(map)}}&
+ \hspace{-0.4em}
+ \ident{List.map}~(\lambda x.e)~[v_1; \ldots] ~\reduce~ [e[x\leftarrow v_1]; \ldots] \\
 \end{array}
@@
 \label{fig:ff-reduction}
+\vspace{-0.5em}
 \end{figure}
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Featherweight F\#}
+\label{sec:formal-ff}
+
+The semantics fragment introduced in the previous section discusses values and operations that are used by
+F\# Data. This section adds a minimal subset of the standard F\# language. We focus on the object system
+(which is what F\# type providers produce) and so the following is based on Featherweight Java. However, we
+only need class types (without inheritance) with properties. A class has a single implicit constructor and 
+the declaration closes over constructor parameters. In addition, we include special constructs for working
+with option types and lists.
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \tau &\narrow{=}& \ident{int} \lsep \ident{decimal} \lsep \ident{float} \lsep \ident{bool} \lsep \ident{string} \\[0.0em]
+      &\narrow{|}& C \lsep \ident{StructVal} \lsep \ident{list}\langl\tau\rangl \lsep \ident{option}\langl\tau\rangl \\[0.6em]
+ L &\narrow{=}& \kvd{type}~C(\overline{x:\tau}) = \overline{M} \\[0.0em]
+ M &\narrow{=}& \kvd{member}~N:\tau=e \\[0.6em]
+ v &\narrow{=}& s \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \\[0.0em]
+ e &\narrow{=}& op \lsep e.N \lsep \kvd{new}~C(\overline{e}) \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3}\\
+   &\narrow{|}& \ident{Some}(e)\lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
+   &\narrow{|}& \kvd{match}~e~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 \\
+   &\narrow{|}& [e_1;\ldots;e_n]\lsep \ident{List.map}~(\lambda x\rightarrow e_1)~e_2
+\end{array}
+\end{equation*}
+
+\noindent
+In the above definition, $\tau$ denotes types including a special type \ident{StructVal} for all structural 
+values $s$; a class definition $L$ consists of a constructor and zero or more members. Values $v$ include 
+previously defined structural values $s$ and values for the option type; finally expressions $e$ include all 
+previously defined operations $op$, class construction, member access, conditionals and expressions for working 
+with option values and lists. We include \ident{List.map} as a special construct to avoid making the language too complex.
+
+In this paper, we do not define the typing of the F\# subset introduced above as our goal is not to prove
+type safety for the core language. For the relativized soundness proof in Section~\ref{sec:safety}, it is
+sufficient to discuss reduction.
+
+\paragraph{Reduction.} The reduction relation is of the form $e \reduce e'$. We also write 
+$e \reduce^{*} e'$ to denote the reflexive and transitive closure of $\reduce$. The reduction rules
+for primitive functions $op$ were discussed earlier in Figure~\ref{fig:op-conversions}. The rules 
+for other expressions defined above are provided in Figure~\ref{fig:ff-reduction}.
+
+The (\emph{ctx}) rule performs a reduction inside a sub-expression specified by an evaluation context.
+This models the eager evaluation order of F\#. An evaluation context $E$ is defined as:
+
+\noindent
+\begin{equation*}
+\begin{array}{rcl}
+ E &\narrow{=}& f(E) \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e}) \lsep \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2 \\[0.1em]
+   &\narrow{|}& \ident{Some}(E) \lsep \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
+   &\narrow{|}& [\overline{v};E;\overline{e}] \lsep \kvd{match}~E~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 \\[0.1em]
+   &\narrow{|}& \ident{map}~(\lambda x\rightarrow e)~E \lsep [\cdot] 
+ \\[0.6em]
+ f &\narrow{\in}& \{\; \ident{asInt}, \ident{asDec}, \ident{asFloat}, \ident{asBool}, \ident{asStr}, \\[0.1em]
+              && ~~\; \ident{getField}, \ident{getChildren}, \ident{isNull}, \ident{isTag} \;\}
+\end{array} 
+\end{equation*}
+
+\noindent
+The evaluation first reduces arguments of functions and the evaluation proceeds from left to right 
+as denoted by $\overline{v}, E, \overline{e}$ in constructor arguments or $\overline{v};E;\overline{e}$
+in list initialization.
+
+We write $e[\overline{x} \leftarrow \overline{v}]$ for the result of replacing variables $\overline{x}$ by
+values $\overline{v}$ in expression. The (\emph{member}) rule reduces a member access using a class 
+definition in the assumption to obtain the body of the expression. Finally, the remaining six rules
+provide standard reductions for conditionals and pattern matching.
+
+The language is sufficient for our purpose, but simple. It does not provide recursion
+and so all expressions reduce to a value in a finite number of steps or get stuck due to an error
+condition. An error condition can be a wrong argument passed to conditional, pattern matching or
+one of the conversion functions from Figure~\ref{fig:op-conversions}.
 
@@
 % Primitive structural types become corresponding F# types and conversion is inserted
-\begin{equation*}
-\begin{array}{l}
- \sem{\sigma_p}_e = \tau_p,op(e),\emptyset\quad\textnormal{where} \\[0.6em]
+\noindent
+\begin{equation*}
+\begin{array}{l}
+ \sem{\sigma_p}_e = \tau_p,op(e),\emptyset\quad\textnormal{where} \\[0.4em]
 \quad\sigma_p, \tau_p, op\in  \{~ (\ident{bit}, \ident{bool}, \ident{asBool}),  (\ident{bool}, \ident{bool}, \ident{asBool})\\[0.2em]
@@
  \qquad M_i = \kvd{member}~\nu_i:\tau_i=e_i\\[0.1em]
- \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_{e'},\;\; e'=\ident{getField}_{\nu_{\ident{opt}}, \nu_i}(e)\\[0.6em]
+ \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_{e'},\;\; e'=\ident{getField}(\nu_{\ident{opt}}, \nu_i, e)\\[0.6em]
 \end{array}
@@
 \begin{array}{l}
- \sem{\,[\sigma]\,}_e = \ident{list}\langl\tau\rangl, \ident{getChildren}(e'), L \;\;\textnormal{where}~ \tau, e', L = \sem{\hat{\sigma}}_e
+ \sem{\,[\sigma]\,}_e = \ident{list}\langl\tau\rangl, \ident{List.map}~(\lambda x\rightarrow e')~(\ident{getChildren}(e)), L\\[0.4em]
+ \quad \textnormal{where}~~\tau, e', L = \sem{\hat{\sigma}}_x
 \end{array}
@@
 % Option values
+\noindent
 \begin{equation*}
@@
  \sem{\hat{\sigma}\;\,\kvd{option}}_e = \\[0.2em]
- \hspace{1.25em} \ident{option}\langl\tau\rangl, \kvd{if}~\ident{isNull}~e'~\kvd{then}~\ident{None}~\kvd{else}~\ident{Some}(e'), L\\[0.2em] 
+ \hspace{1.25em} \ident{option}\langl\tau\rangl, \kvd{if}~\ident{isNull}~e~\kvd{then}~\ident{None}~\kvd{else}~\ident{Some}(e'), L\\[0.2em] 
  \hspace{1.25em} \textnormal{where}~\tau, e', L = \sem{\hat{\sigma}}_e
@@
  \qquad \;\,\,L = \kvd{type}~C(v:\ident{StructVal})~=~M_1\ldots M_n \\[0.1em]
- \qquad M_i = \kvd{member}~\nu_i:\ident{option}\langl\tau_i\rangl=e_i'\\[0.1em]
- \qquad \;\,\,e_i' = \kvd{if}~\ident{isNull}(e_i)~\kvd{then}~\ident{None}~\kvd{else} \\
- \hspace{4.8em}     \kvd{if}~\ident{isTag}_{t_i}{(e_i)}~\kvd{then}~
-     \ident{Some}(e_1)~\kvd{else}~\ident{None} \\[0.1em]
+ \qquad M_i = \kvd{member}~\nu_i:\ident{option}\langl\tau_i\rangl=\\[0.1em]
+ \hspace{5.8em}  \kvd{if}~\ident{isNull}(v)~\kvd{then}~\ident{None}~\kvd{else} \\
+ \hspace{5.8em}  \kvd{if}~\ident{isTag}(t_i, v)~\kvd{then}~
+     \ident{Some}(e_i)~\kvd{else}~\ident{None} \\[0.1em]
  \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_e,\;\; t_i = \tytagof{(\sigma_i)},\;\; \nu_i=\nameoftag{(t_i)}
@@
 
+\vspace{-0.5em}
 \noindent
@@
 \label{fig:tp-generation}
+\vspace{-0.5em}
 \end{figure*}
@@
 type $\sigma$ based on one or more sample documents (Section~\ref{sec:inference}). In Section~\ref{sec:formal-ff},
-we defined a simplified model of F\# language and evaluation, which also included conversions that are
-modelled based on the F\# Data library runtime (Section~\ref{sec:formal-tp}). In this section, we define how
+we defined a simplified model of F\# language and evaluation, which also included conversions that
+model the F\# Data library runtime (Section~\ref{sec:formal-tp}). In this section, we define how
 the type providers work, which links the two aspects discussed so far.
@@
 
-When generating the F\# types, the type provider needs to produce the corresponding F\# type, expression that
-wraps a structured document value (of type \ident{StructVal}) and a collection of class definitions. We express
-it using the following mapping:
+\paragraph{Type provider mapping.}
+When generating types, the type provider produces an F\# type $\tau$, expression that wraps a structured 
+document value (of type \ident{StructVal}) as a value of type $\tau$ and a collection of class definitions. 
+We express it using the following mapping:
 %
@@
 being wrapped. Then, given an inferred structured type $\sigma$, the type provider produces an F\# type $\tau$,
-an expression $e'$ which constructs a value of type $\tau$ from $e$ and also a set of class definitions $L$.
-
-The rules that define $\sem{-}_e$ are shown in Figure~\ref{fig:tp-generation}.
+an expression $e'$ which constructs a value of type $\tau$ using $e$ and also a set of class definitions $L$.
+
+Figure~\ref{fig:tp-generation} shows the rules that define $\sem{-}_e$. Primitive types are all
+handled by a single rule. For a given structural type, it returns the corresponding F\# type -- the types are
+mapped directly with the exception of \ident{bit}, which becomes an F\# \ident{bool}. The generated code calls
+an appropriate conversion function from Figure~\ref{fig:op-conversions} on the input.
+
+Handling of records is more interesting. We generate a new class $C$ that takes a structured value as constructor
+parameter. For each record field, we generate a new member with the same name as the field\footnote{The actual
+F\# Data implementation also capitalizes the names.} The body of the member calls \ident{getField} and then
+passes this expression to $\sem{\sigma_i}$ which adds additional wrapping that maps the field (structured value
+of type $\sigma_i$) into an F\# type $\tau_i$. The returned expression creates a new instance of $C$ and
+the mapping returns the class $C$ together with all recursively generated definitions.
+
+A collection type becomes an F\# $\ident{list}\langl\tau\rangl$. The returned expression calls \ident{getChildren}
+(which turns \kvd{null} values into empty lists) and then uses \ident{List.map} to convert all nested values to 
+an F\# type $\tau$. The handling of option type is similar -- it checks if the original value is \kvd{null} and
+if no, it wraps the recursively generated conversion expression $e'$ in the \ident{Some} constructor.
+
+As discussed earlier, union types are also generated as classes with properties. Given a union type $\sigma_1 + \ldots + \sigma_n$,
+we get corresponding F\# types $\tau_i$ and generate $n$ members of type $\ident{option}\langl \tau_i\rangl$.
+Each member return \ident{Some} when the value is not \kvd{null} and has the right structure (checked by 
+\ident{isTag}). The type inference algorithm also guarantees that there is only one case for each type tag 
+(Section~\ref{sec:inference-commonsuper}) and there are no nested union types. Thus, checking for a tag is
+sufficient and we can also use the tag to identify the name of the generated member (using the
+\ident{nameof} function).
+
+\paragraph{Example 1.}
+To illustrate how the type provision mechanism works, we now consider two simple examples. First, assume 
+that the inferred type is a record with two fields (one optional) such as 
+$\ident{Person}~\{~\ident{Age}:\ident{int}~\kvd{option}, \ident{Name}:\ident{string}~ \}$. Applying the
+rules from Figure~\ref{fig:tp-generation} produces the following class:
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{type}~\ident{Person}(\ident{v}:\ident{StructVal})~= \\[0.1em]
+ \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl\ident{int}\rangl~= \\[0.1em]
+ \qquad \kvd{if}~\ident{isNull}~(\ident{getField}(\ident{Person},\ident{Age}, \ident{v}))~\kvd{then}~\ident{None} \\[0.1em]
+ \qquad \kvd{else}~\ident{Some}(\ident{asInt}(\ident{getField}(\ident{Person},\ident{Age}, \ident{v}))) \\[0.1em]
+ \quad \kvd{member}~\ident{Name}~:~\ident{string}~= \\[0.1em]
+ \qquad \ident{asStr}(\ident{getField}(\ident{Person},\ident{Name}, \ident{v}))
+\end{array}
+\end{equation*}
+%
+The body of the \ident{Age} member is produced by the case for optional types applied to an expression
+$\ident{getField}(\ident{Person},\ident{Age},v)$. If the returned field is not \kvd{null}, then the member
+calls \ident{asInt} (produced by the case for primitive types) and wraps the result in the \ident{Some}
+constructor. Note that \ident{getField} is defined even when the field does not exist, but returns \ident{null}.
+This lets us treat missing fields as optional fields. The \ident{Name} member is similar, but does not 
+perform any checks.
+
+For completeness, a type corresponding to the record is \ident{Person} and given a 
+structural document $s$, we create a \ident{Person} value by calling $\kvd{new}~\ident{Person}(s)$.
+
+\paragraph{Example 2.} The second example illustrates the remaining parts of the type provision,
+including collections and union types. Teusing the \ident{Person} type from the previous example,
+consider a collection $[\ident{Person}+\ident{string}]$, which contains a mix of \ident{Person} and string values.
+
+\noindent
+\begin{equation*}
+\begin{array}{l}
+ \kvd{type}~\ident{PersonOrString}(\ident{v}:\ident{StructVal})~= \\[0.1em]
+ \quad \kvd{member}~\ident{Person}~:~\ident{option}\langl\ident{Person}\rangl~= \\[0.1em]
+ \qquad \kvd{if}~\ident{isNull}(v)~\kvd{then}~\ident{None}~\kvd{else} \\[0.1em]
+ \qquad \kvd{if}~\ident{isTag}(\ident{rec-named Person}, v)~\kvd{then}~\\[0.1em]
+ \qquad\quad \ident{Some}(\kvd{new}~\ident{Person}(v))~\kvd{else}~\ident{None} \\[0.1em]
+ \quad \kvd{member}~\ident{String}~:~\ident{option}\langl\ident{string}\rangl~= \\[0.1em]
+ \qquad \kvd{if}~\ident{isNull}(v)~\kvd{then}~\ident{None}~\kvd{else} \\[0.1em]
+ \qquad \kvd{if}~\ident{isTag}(\ident{string}, v)~\kvd{then}~\\[0.1em]
+ \qquad\quad \ident{Some}(\ident{asStr}(v))~\kvd{else}~\ident{None}
+\end{array}
+\end{equation*}
+
+\noindent
+The type provider generates the above type and the collection type is mapped to an F\# type 
+$\ident{list}\langl\ident{PersonOrString}\rangl$. Given a structural document value $s$, the code
+to obtain the wrapped F\# value is:
+%
+\begin{equation*}
+\ident{List.map}~(\lambda x\rightarrow\kvd{new}~\ident{PersonOrString}(x))~(\ident{getChildren}(s))
+\end{equation*}
+%
+The \ident{PersonOrString} type contains one property for each of the union case. In the body, they
+check that the value is not \kvd{null} and that it has the right structure (using the \ident{isTag}
+function). This checks that the value is a record named \ident{Person} or a string, respectively.
+If the conditions are satisfied, the value is converted to the F\# type corresponding to the case
+and wrapped in \ident{Some}.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Infering types from values}
+
+Before concluding the section on formalizing type providers,
+
+\begin{equation*}
+\begin{array}{rclcrcl}
+ \semalt{0} &\narrow{=}& \ident{bit} &\qquad& \semalt{1} &\narrow{=}& \ident{bit} \\
+ \semalt{i} &\narrow{=}& \ident{int} && \semalt{d} &\narrow{=}& \ident{decimal} \\
+ \semalt{f} &\narrow{=}& \ident{float} && \semalt{t} &\narrow{=}& \ident{string} \\
+ \semalt{b} &\narrow{=}& \ident{bool} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} \\[0em]
+\end{array}
+\end{equation*}
+\noindent
+\begin{equation*}
+\begin{array}{l}
+ \semalt{[s_1; \ldots; s_n]} = [\sigma_n] \quad\textnormal{where}\\[0.1em]
+ \qquad\sigma_0 = \top,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{s_i} \vdash \sigma_i
+ \\[0.6em]
+ \semalt{\nu_{\ident{opt}}~\{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \}} =\\[0.1em]
+ \qquad\nu_{\ident{opt}}~\{ \nu_1:\semalt{s_1}, \ldots, \nu_n :\semalt{s_n} \}
+\end{array}
+\end{equation*}
 
 % ==================================================================================================
-
-\newpage
 
@@
 
+Turning JSON, CSV or XML to the idealized structured value presented here (esp. XML)
+
 Error handling - solve issues just add another example

### 2015-02-24 icfp-2015/paper.tex
Add some type checking rules

@@
 % TODO: Replace \nu_opt with \nu_? or something else that does not look shit
+% TODO: Cleanup "strucutred" vs "structural"
+% TODO: Call it "relativized safety" or "relativized soundness" consistently
 
@@
 
-\caption{Remaining reduction rules for Featherweight F\#}
+\caption{Featherweight F\# -- Remaining reduction rules}
 \label{fig:ff-reduction}
@@
 The semantics fragment introduced in the previous section discusses values and operations that are used by
-F\# Data. This section adds a minimal subset of the standard F\# language. We focus on the object system
-(which is what F\# type providers produce) and so the following is based on Featherweight Java. However, we
-only need class types (without inheritance) with properties. A class has a single implicit constructor and 
-the declaration closes over constructor parameters. In addition, we include special constructs for working
-with option types and lists.
+F\# Data. This section adds a minimal subset of the standard F\# language.
+
+We focus on the object system (which is what F\# type providers produce) and so the following combines 
+standard ML [?] features with Featherweight Java [?] classes. However, we only need classes with properties and
+without inheritance. A class has a single implicit constructor and the declaration closes over 
+constructor parameters. To avoid including all of ML, we only select constructs for working with options and lists
+that we need later.
 %
@@
  M &\narrow{=}& \kvd{member}~N:\tau=e \\[0.6em]
- v &\narrow{=}& s \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \\[0.0em]
+ v &\narrow{=}& s \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \lsep [v_1; \ldots; v_n] \\[0.0em]
  e &\narrow{=}& op \lsep e.N \lsep \kvd{new}~C(\overline{e}) \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3}\\
@@
 \noindent
-In the above definition, $\tau$ denotes types including a special type \ident{StructVal} for all structural 
-values $s$; a class definition $L$ consists of a constructor and zero or more members. Values $v$ include 
-previously defined structural values $s$ and values for the option type; finally expressions $e$ include all 
+In the above definition, $\tau$ denotes types including a type of all structured values $s$ called
+\ident{StructVal}. A class definition $L$ consists of a constructor and zero or more members. Values $v$ include 
+previously defined structural values $s$ and values for the option and list type; finally expressions $e$ include 
 previously defined operations $op$, class construction, member access, conditionals and expressions for working 
@@
 
-In this paper, we do not define the typing of the F\# subset introduced above as our goal is not to prove
-type safety for the core language. For the relativized soundness proof in Section~\ref{sec:safety}, it is
-sufficient to discuss reduction.
+Next, we define the reduction relation and (a fragment of) type checking for Featherweight F\#.
+The language presented here is intentionally incomplete. We only define parts needed to prove
+the relativized soundness property in Section~\ref{sec:safety}.
 
@@
 one of the conversion functions from Figure~\ref{fig:op-conversions}.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure}
+\begin{equation*}
+\qquad
+\inference
+  {~}
+  {\Gamma \vdash s : \ident{StructVal}}
+\qquad  
+\inference
+  {~}
+  {\Gamma \vdash n : \ident{int}}
+\end{equation*}
+
+\begin{equation*}
+\inference
+  {\Gamma \vdash e : C & \kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..}
+  {\Gamma \vdash e.N_i:\tau_i}
+\end{equation*}
+
+\begin{equation*}
+\inference
+  {\Gamma \vdash e_i : \tau_i & \kvd{type}~C(x_1:\tau_1, \ldots, x_n:\tau_n) = \ldots}
+  {\Gamma \vdash \kvd{new}~C(e_1, \ldots, e_n):C}
+\end{equation*}
+
+\caption{Featherweight F\# -- Fragment of type checking}
+\label{fig:ff-typecheck}
+\end{figure}
+
+% -------------------------------------------------------------------------------------------------
+
+\paragraph{Type checking.} 
+Selected type checking rules for Featherweight F\# are shown in Figure~\ref{fig:ff-typecheck}.
+The rules demonstrate the key difference between our language and standard ML or Featherweight Java:
+%
+\begin{itemize}
+\item All structured values $s$ have a type \ident{StructVal}. Some of those have other types
+  (primitive integers, decimals, floats, strings and Booleans as illustrated by the rule for $n$).
+  For other values, \ident{StructVal} is the only type -- this includes records and \kvd{null}.
+\item A list containing other structural values $[s_1; \ldots; s_n]$ has a type \ident{StructVal}.  
+  but can also have the $\ident{list}\langl\tau\rangl$ type. Conversely, lists that contain
+  non-structural values like objects or options are not of type \ident{StructVal}.
+\item The two rules for checking class construction and member access are similar to corresponding
+  rules of Featherweight Java.  
+\end{itemize}
+%
+An important part of Featherweight Java that is omitted here is the checking of type declarations
+(ensuring the bodies of members are well-typed). We omit this, because it is not needed for the
+relativized safety theorem, which considers only classes generated by our type provider (which 
+are correct by construction).
 
@@
 
-Before concluding the section on formalizing type providers,
-
+Before concluding the section on formalizing type providers, there is one more missing piece. The
+common supertype algorithm in Section~\ref{sec:inference} discusses the core of the type inference for
+structural values, but we have not yet clarified how exactly it is used. We address this in the present
+section.
+
+Given a JSON, XML or CSV document, the F\# Data implementation constructs a structural value $s$
+from the sample (this is straightforward, but Section~\ref{sec:impl-parsing} discusses some interesting 
+aspects). The following defines a mapping $\semalt{-}$ which turns a sample value $s$ into a structural
+type $\sigma$:
+%
 \begin{equation*}
@@
 \noindent
-\begin{equation*}
-\begin{array}{l}
- \semalt{[s_1; \ldots; s_n]} = [\sigma_n] \quad\textnormal{where}\\[0.1em]
- \qquad\sigma_0 = \top,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{s_i} \vdash \sigma_i
- \\[0.6em]
+\vspace{-0.2em}
+\begin{equation*}
+\begin{array}{l}
+ \semalt{[s_1; \ldots; s_n]} = [\semalt{s_1, \ldots, s_n}]
+ \\[0.5em]
  \semalt{\nu_{\ident{opt}}~\{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \}} =\\[0.1em]
  \qquad\nu_{\ident{opt}}~\{ \nu_1:\semalt{s_1}, \ldots, \nu_n :\semalt{s_n} \}
-\end{array}
-\end{equation*}
+ \\[0.5em]
+ \semalt{s_1, \ldots, s_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
+ \qquad\sigma_0 = \top,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{s_i} \vdash \sigma_i
+\end{array}
+\end{equation*}
+%
+Primitive values are mapped to their corresponding types, with the exception of $0$ and $1$, which
+are inferred as \ident{bit}. For records, we return a type with field types inferred based on 
+individual values.
+
+The interesting part is inferring type based on multiple samples. We overload
+the notation and write $\semalt{s_1, \ldots, s_n}$ for a type inferred from multiple samples.
+This uses the common supertype relation to find a common type for all values (starting with $\top$).
+This operation is used at the top-level (when calling type provider with multiple samples)
+and also when inferring the type of a collection.
 
@@
 \label{sec:safety}
-~
-
-We presented the system together with its syntactic equational theory and categorical semantics.
-The equational theory is presented in order to explain how the systems work, but it also
-provides a basis for an operational semantics for concrete systems. Exploring these, 
-and their connection to the denotational semantics, is further work.
-
+
+Now that we have
+
+\begin{theorem}[Relativized safety]
+Given samples $s_1, \ldots, s_n$, inferred type $\sigma=\semalt{s_1, \ldots, s_n}$ and generated 
+$\tau,e,L = \sem{\sigma}_x$ then:
+
+For all $s'$ such that $\exists i.s_i :> s'$, then for any $e$ such that $x:\ident{StructVal} \vdash e:\tau$
+it is the case that $e[x\leftarrow s'] \reduce^{*} v$ and $\vdash v : \tau$.
+
+\end{theorem}
 % - Safety property
@@
 
+\label{sec:impl-parsing}
+
 Turning JSON, CSV or XML to the idealized structured value presented here (esp. XML)

### 2015-02-24 icfp-2015/paper.tex
Write few more things about proofs

@@
 \newtheorem{theorem}{Theorem}
+\newtheorem{lemma}{Lemma}
 
@@
 
+\begin{lemma}[Correctness of provided types]
+Given a rule from the type provider, when we have a structural value that
+
+Given $\semalt{s} :> \semalt{s'}$ and $\tau, e, L = \sem{\semalt{s}}_{s'}$ then
+either $e \reduce v$ or $\tau=C$ and for all fields $e.N \reduce v$.
+\end{lemma}
+\begin{proof}
+?
+\end{proof}
+
+\begin{lemma}[Preservation]
+Preservation
+\end{lemma}
+\begin{proof}
+by induction over the reduction $\reduce$ (this should be fairly standard)
+\end{proof}
+
+\begin{lemma}[Progress]
+Progress
+\end{lemma}
+\begin{proof}
+by induction over the typing of $e$ (this is standard bit) 
+also using the observation that all provided classes are well-typed
+(by handwaving). The interesting case is when we do anything on StructVal, in which case
+we need:
+
+All access on StructVals happens in the code generated by type provider
+and we can use the lemma...
+\end{proof}
+
+Now we can finally write
+
 \begin{theorem}[Relativized safety]
-Given samples $s_1, \ldots, s_n$, inferred type $\sigma=\semalt{s_1, \ldots, s_n}$ and generated 
-$\tau,e,L = \sem{\sigma}_x$ then:
-
-For all $s'$ such that $\exists i.s_i :> s'$, then for any $e$ such that $x:\ident{StructVal} \vdash e:\tau$
-it is the case that $e[x\leftarrow s'] \reduce^{*} v$ and $\vdash v : \tau$.
-
+Assume $s_1, \ldots, s_n$ are samples, $\sigma=\semalt{s_1, \ldots, s_n}$ is an inferred
+type and $\tau,e,L = \sem{\sigma}_x$ are a type, expression and class definitions generated by a 
+type provider.
+
+Then for all new inputs $s'$ such that $\exists i.(\semalt{s_i} :> \semalt{s'})$, and for any $e$ 
+such that $x:\ident{StructVal} \vdash e:\tau$ using class definitions $L$, it is the case that 
+$e[x\leftarrow s'] \reduce^{*} v$ and $\vdash v : \tau$.
 \end{theorem}
-% - Safety property
+\begin{proof}
+Using progress and preservation.
+\end{proof}
 

### 2015-02-24 icfp-2015/paper.tex
Finishing the proofs and discussions

--- 2015-02-24 icfp-2015/paper.tex
@@
 \usepackage{semantic}
+\usepackage{enumitem}
 
@@
 % TODO: Cleanup "strucutred" vs "structural"
-% TODO: Call it "relativized safety" or "relativized soundness" consistently
+% TODO: Insert a few more mentions of why "s1 + s2 <: s1 + s2 + s3" is valid (cannot do complete pattern match)
 
@@
 \label{eq:sub-prim}\tag{B1}
-\ident{float}\,:>\,\ident{decimal}\,:>\,\ident{int}\,:>\,\ident{bit}\quad&\textnormal{and}\quad\ident{bool}\,:>\,\ident{bit}\quad\\[-0.2em]
+\ident{float}\,:>\,\ident{decimal}\,:>\,\ident{int}\,:>\,\ident{bit}\quad&\textnormal{and}\quad\ident{bool}\,:>\,\ident{bit}\\[-0.2em]
 \label{eq:sub-null}\tag{B2}
@@
 \label{eq:sub-union3}\tag{U3}
+\sigma_1 + \ldots + \sigma_n :> \sigma_1 + \ldots + \sigma_m \quad &(\textnormal{where}~m \geq n) \\
+\label{eq:sub-union4}\tag{U4}
 \hspace{-1em}\sigma_1 + \ldots + \sigma_n :> \sigma_{\pi(1)} + \ldots + \sigma_{\pi(n)} \quad &(\textnormal{permutation}~\pi) \\
-\label{eq:sub-union-el}\tag{U4}
+\label{eq:sub-union-el}\tag{U5}
 \sigma_1 + \ldots + \sigma_n :> \sigma_i \quad &(\textnormal{for}~i \in 1 \ldots n)
@@
 \begin{figure*}[t]
+\noindent
 \begin{equation*}
@@
 % all rules are symmetric and reflexive
+\noindent
 \begin{equation*}
@@
 \begin{array}{l}
- \textnormal{\footnotesize{(refl)}}\;\; \sigma \tsep \sigma \vdash \sigma\\[0.75em]
+ \textnormal{\footnotesize{(refl)}}\;\; \sigma \tsep \sigma \vdash \sigma\\[0.6em]
  % top type
@@
  \begin{array}{l}
- \textnormal{\footnotesize{(null-1)}}\;\; \sigma \tsep \kvd{null} \vdash \sigma \quad(\sigma :> \kvd{null}) \\[0.75em]
+ \textnormal{\footnotesize{(null-1)}}\;\; \sigma \tsep \kvd{null} \vdash \sigma \quad(\sigma :> \kvd{null}) \\[0.6em]
  \textnormal{\footnotesize{(null-2)}}\;\; \sigma \tsep \kvd{null} \vdash \sigma~\kvd{option} \quad(\sigma :\ngtr \kvd{null})
@@
 
-\item The remaining rules for unins are standard. Subtype can have fewer alternatives 
+\item For unions, we include the standard rules. Subtype can have fewer alternatives 
   (\ref{eq:sub-union1}), elements can be reordered (\ref{eq:sub-union2}) and subtyping is covariant 
-  (\ref{eq:sub-union3}). Similarly, the collection type is also covariant (\ref{eq:sub-col}).
+  (\ref{eq:sub-union4}). An unusual rule (\ref{eq:sub-union3}) says that subtype can also have 
+  \emph{more} cases. This is valid because the type provider exposes unions as objects that do not
+  allow complete pattern matching.
 
@@
 \label{fig:ff-reduction}
-\vspace{-0.5em}
+\vspace{-1em}
 \end{figure}
@@
 The language presented here is intentionally incomplete. We only define parts needed to prove
-the relativized soundness property in Section~\ref{sec:safety}.
+the relativized safety property in Section~\ref{sec:safety}.
 
@@
 
+\vspace{0.5em}
 \noindent
@@
   {~}
-  {\Gamma \vdash s : \ident{StructVal}}
+  {L; \Gamma \vdash s : \ident{StructVal}}
 \qquad  
@@
   {~}
-  {\Gamma \vdash n : \ident{int}}
+  {L; \Gamma \vdash n : \ident{int}}
 \end{equation*}
@@
 \inference
-  {\Gamma \vdash e : C & \kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;..}
-  {\Gamma \vdash e.N_i:\tau_i}
+  {L; \Gamma \vdash e : C & \kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;.. \in L}
+  {L; \Gamma \vdash e.N_i:\tau_i}
 \end{equation*}
@@
 \inference
-  {\Gamma \vdash e_i : \tau_i & \kvd{type}~C(x_1:\tau_1, \ldots, x_n:\tau_n) = \ldots}
-  {\Gamma \vdash \kvd{new}~C(e_1, \ldots, e_n):C}
+  {L; \Gamma \vdash e_i : \tau_i & \kvd{type}~C(x_1:\tau_1, \ldots, x_n:\tau_n) = \ldots \in L}
+  {L; \Gamma \vdash \kvd{new}~C(e_1, \ldots, e_n):C}
 \end{equation*}
@@
 \paragraph{Type checking.} 
-Selected type checking rules for Featherweight F\# are shown in Figure~\ref{fig:ff-typecheck}.
+The type checking rules in Figure~\ref{fig:ff-typecheck} are written using a judgement
+$L; \Gamma \vdash e : \tau$ where the context also contains a set of class declarations $L$.
 The rules demonstrate the key difference between our language and standard ML or Featherweight Java:
 %
-\begin{itemize}
-\item All structured values $s$ have a type \ident{StructVal}. Some of those have other types
+\begin{itemize}[noitemsep]
+\item[--] All structured values $s$ have a type \ident{StructVal}. Some of those have other types
   (primitive integers, decimals, floats, strings and Booleans as illustrated by the rule for $n$).
   For other values, \ident{StructVal} is the only type -- this includes records and \kvd{null}.
-\item A list containing other structural values $[s_1; \ldots; s_n]$ has a type \ident{StructVal}.  
+\item[--] A list containing other structural values $[s_1; \ldots; s_n]$ has a type \ident{StructVal}.  
   but can also have the $\ident{list}\langl\tau\rangl$ type. Conversely, lists that contain
   non-structural values like objects or options are not of type \ident{StructVal}.
-\item The two rules for checking class construction and member access are similar to corresponding
+\item[--] Primitive operations $op$ are treated as functions, accepting \ident{StructVal} and producing
+  an appropriate type as the result.
+\item[--] The two rules for checking class construction and member access are similar to corresponding
   rules of Featherweight Java.  
@@
 perform any checks. 
-
 For completeness, a type corresponding to the record is \ident{Person} and given a 
@@
 
-Now that we have
+Informally, the safety property of F\# Data type providers states that, given representative sample
+documents, any code that can be written using the provided types is guaranteed to work. We call this 
+\emph{relativized safety}, because we cannot avoid \emph{all} errors. In particular, the user can 
+always use the type provider with an input that has a different structure than any of the samples -- and in 
+this case, it is expected that the code will fail at runtime (throw an exception in the actual 
+implementation or by getting stuck in our model).
+
+The key question is, what is a representative sample? Given a set of sample documents, 
+the provided type is guaranteed to work if the inferred type of the input is a subtype of any of the
+sample documents. Going back to Section~\ref{sec:inference-subtyping}, this means that:
+%
+\begin{itemize}[noitemsep]
+\item[--] Input can contain smaller numerical values (for example, if a sample contains float, the input can contain an integer).
+\item[--] Records in the actual input can have additional fields
+\item[--] Records in the actual input can have fewer fileds, provided that the type of the fields is marked as optional in the sample
+\item[--] Union types in the input can have both fewer or more cases
+\item[--] When we have a union type in the sample, the actual input can also contain just values of one of the union cases
+\end{itemize}
+%
+The following lemma states that the provided code (generated in Figure~\ref{fig:tp-generation})
+works correctly on inputs $s'$ that is a subtype of the sample $s$. More formally, we require that the 
+provided expression (using $s'$ as the input) can be reduced to a value and, if it is a class,
+all its members can also be reduced to values.
+
+\begin{lemma}[Correctness of provided types]
+\label{thm:tp-correctness}
+Given a sample value $s$ and an input value $s'$ such that $\semalt{s} :> \semalt{s'}$
+and provided type, expression and class declarations $\tau, e, L = \sem{\semalt{s}}_{s'}$, 
+then $e \reduce^{*} v$ and if $\tau$ is a class ($\tau=C$) then for all members $N_i$ of the 
+class $C$, it holds that $e.N_i \reduce^{*} v$.
+\end{lemma}
+\begin{proof}
+By induction over the structure of $\sem{-}_e$. For primitives, the conversion functions accept all subtypes.
+For other cases, analyse the the provided code to see that it can work on all subtypes (\emph{e.g.}~\ident{getChildren}
+works on \kvd{null} values, \ident{getField} returns \kvd{null} when a field is missing and, when \ident{isTag} returns
+\ident{true}, then a value has the structure required by the corresponding case).
+\end{proof}
+
+\noindent
+Now that we know that the provided types are correct with respect to the subtyping relation, we can
+look at the main theorem of the paper. It statest that, for any input (which is a subtype of any of the
+samples) and any expression $e$, a well-typed program that uses the provided types does not ``go wrong''.
+
+Using the standard approach to syntactic type safety [?], we prove the type preservation 
+(reduction does not change type) and progress (an expression that is not a value can be reduced).
 
 \begin{theorem}[Relativized safety]
-Given samples $s_1, \ldots, s_n$, inferred type $\sigma=\semalt{s_1, \ldots, s_n}$ and generated 
-$\tau,e,L = \sem{\sigma}_x$ then:
-
-For all $s'$ such that $\exists i.s_i :> s'$, then for any $e$ such that $x:\ident{StructVal} \vdash e:\tau$
-it is the case that $e[x\leftarrow s'] \reduce^{*} v$ and $\vdash v : \tau$.
-
+\label{thm:safety}
+Assume $s_1, \ldots, s_n$ are samples, $\sigma=\semalt{s_1, \ldots, s_n}$ is an inferred
+type and $\tau,e,L = \sem{\sigma}_x$ are a type, expression and class definitions generated by a 
+type provider.
+
+Then for all new inputs $s'$ such that $\exists i.(\semalt{s_i} :> \semalt{s'})$, let $e_s=e[x\leftarrow s']$
+be an expression (of type $\tau$) that wraps the input in a provided type. Then, for any expression $e_c$
+(user code) such that $\emptyset; y:\tau \vdash e_c:\tau'$ and that does not contain any structural values
+as sub-expressions, it is the case that $e_c[y\leftarrow e_s] \reduce^{*} v$ for some value $v$ and
+also $\emptyset; \vdash v : \tau$.
 \end{theorem}
-% - Safety property
-
-% ==================================================================================================
+\begin{proof}
+We discuss the two parts of the proof separately as type prservation (Lemma~\ref{thm:rs-preservation})
+and progress (Lemma~\ref{thm:rs-progress}).
+\end{proof}
+
+\begin{lemma}[Preservation]
+\label{thm:rs-preservation}
+Given the class definitions $L$ generated by a type provider as specified in
+the assumptions of Theorem~\ref{thm:safety}, then if $\Gamma \vdash e : \tau$ and 
+$e \reduce^{*} e'$ then $\Gamma \vdash e' : \tau$.
+\end{lemma}
+\begin{proof}
+By induction over the reduction $\reduce$. The cases for the ML subset of Featherweight F\# 
+are standard. For (\emph{member}), we check that code generated by type providers
+in Figure~\ref{fig:tp-generation} is well-typed.
+\end{proof}
+
+\noindent
+The progress lemma states that evaluation of a well-typed program does not reach an undefined state. 
+This is not a problem for the ML subset and object-oriented subset of the calculus. The problematic
+part are the conversion functions (Figure~\ref{fig:op-conversions}). Given a structural value 
+(which has a type \ident{StructVal} in our language), the reduction can get stuck if the value does
+not have a structure required by a specific conversion function used.
+
+The Lemma~\ref{thm:tp-correctness} guarantees that this does not happen. In Theorem~\ref{thm:safety},
+we carefuly state that we only consider expressions $e_c$ which ``[do] not contain any structural 
+values as sub-expressions''. This makes sure that the only code working with structural values is the
+code generated by the type provider.
+
+\begin{lemma}[Progress]
+\label{thm:rs-progress}
+Given the assumptions and definitions from Theorem~\ref{thm:safety}, it is the case that
+$e_c[y\leftarrow e_s] \reduce^{*} v$.
+\end{lemma}
+\begin{proof}
+By induction over the typing derivation of $L; \emptyset \vdash e_c[y\leftarrow e_s] : \tau'$. 
+The cases for the ML subset are standard. For member access, we rely on Lemma~\ref{thm:tp-correctness}.
+\end{proof}
+
+% -------------------------------------------------------------------------------------------------
 
@@
 \label{sec:safety-discuss}
-alternative - check the whole file upfront based on the inferred type
-what if the sample is online??
+
+The \emph{relativized safety} property does not guarantee the same amount of safety as standard type safety
+for programming langauges without type providers. However, it reflects the reality of programming with external
+data sources that is increasingly important in the age of web [?]. So, type providers do not reduce the safety
+-- they simply make the existing issues visible.
+
+The actual implementation in F\# Data throws an exception for invalid inputs. In contrast, the calculus
+presented here simply gets stuck. We could extend the calculus with exceptions, but that would obscure the 
+purpose -- precisely specifying when the code using type providers does not ``go wrong.''
+
+As mentioned earlier, \emph{relativized safety} specifies a sufficient, but not a necessary condition.
+This is also reflected in the conversion functions which allow additional conversions not required by 
+the subtyping relation. In practice, the additional flexibility proves useful (as minor variations in 
+input formats are common). However, an interesting stricter alternative would be to check that an input 
+has the required type when it is loaded -- and throw an exception immediately when loading data rather than 
+on member access.
+
+Finally, the formal model presented here ignores two aspects of the real type provider mechanisms. As discussed
+here, we do not use the \emph{type erasure} mechanism, but treat type providers as if they were actually 
+generating code. For XML, CSV and JSON, both models can be used. However, one interesting aspect of erasing 
+type providers is that the types can be generated lazilly -- only classes that are actually used in the type
+checked code are generated. Extending our formalism to capture this aspect is an interesting future work
+relevant to other type providers -- including the World Bank type provider [?] that is also available
+in the F\# Data library.
+
+% ==================================================================================================
 
@@
 
+\subsection{Parsing}
 \label{sec:impl-parsing}
-
 Turning JSON, CSV or XML to the idealized structured value presented here (esp. XML)
 
+JSON - infering numbers from values (not just from actual value type - world bank example)
+
+\subsection{Unions}
+\label{sec:impl-unions}
+Mainly because type providers can only provide objects
+
+Unions become records with fields - so hiding null makes sense
+also it means the open-world semantics
+
+\subsection{Heterogeneous collections}
+\label{sec:impl-collections}
+
+Heterogeneous collections
+
+\subsection{Practical experience}
 Error handling - solve issues just add another example
 
-\label{sec:impl-unions}
-Unions become records with fields - so having null makes sense
+% ==================================================================================================
+
+\section{Related and future work}
+
+~
 
@@
 
-JSON - infering numbers from values (not just from actual value type - world bank example)
-
-\label{sec:impl-collections}
-
-Heterogeneous collections
+~
+
+F\# and Agda type providers
+
+Kathleen Fisher's work
+
+JSON inference stuff
+
+LINQ and XML stuff
 
@@
 
-\newpage
-~
-\newpage
-
-\appendix
-\section{Appendix Title}
-
-This is the text of the appendix, if you need one.
-
 \acks
-
 Acknowledgments, if needed.
@@
 \end{document}
-
-%                       Revision History
-%                       -------- -------
-%  Date         Person  Ver.    Change
-%  ----         ------  ----    ------
-
-%  2013.06.29   TU      0.1--4  comments on permission/copyright notices

### 2015-02-25 icfp-2015/paper.tex
First half of implementation

@@
 \subsection{Infering types from values}
+\label{sec:formal-inferval}
 
@@
 \begin{proof}
-We discuss the two parts of the proof separately as type prservation (Lemma~\ref{thm:rs-preservation})
+We discuss the two parts of the proof separately as type preservation (Lemma~\ref{thm:rs-preservation})
 and progress (Lemma~\ref{thm:rs-progress}).
@@
 % ==================================================================================================
+%
+%   ###                                                                                     
+%    #  #    # #####  #      ###### #    # ###### #    # #####   ##   ##### #  ####  #    # 
+%    #  ##  ## #    # #      #      ##  ## #      ##   #   #    #  #    #   # #    # ##   # 
+%    #  # ## # #    # #      #####  # ## # #####  # #  #   #   #    #   #   # #    # # #  # 
+%    #  #    # #####  #      #      #    # #      #  # #   #   ######   #   # #    # #  # # 
+%    #  #    # #      #      #      #    # #      #   ##   #   #    #   #   # #    # #   ## 
+%   ### #    # #      ###### ###### #    # ###### #    #   #   #    #   #   #  ####  #    # 
+%
+% ==================================================================================================
 
@@
 \label{sec:impl}
-~
-
-\subsection{Parsing}
+
+The theoretical model in Section~\ref{sec:formal} presents the core ideas behind the type 
+providers for structured data formats. However, an important part of the success of the F\# Data
+library is also its pragmatic approach, which makes it suitable for use with real-world data.
+
+The formal model discusses some of the pragmatic choices, such as the preference for records over 
+unions and the ability to treat $0$ and $1$ as both Booleans and numbers. In this section, we briefly
+discuss some of the remaining practical concerns, starting with the handling of collections.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Parsing strucutred values}
 \label{sec:impl-parsing}
-Turning JSON, CSV or XML to the idealized structured value presented here (esp. XML)
-
-JSON - infering numbers from values (not just from actual value type - world bank example)
-
-\subsection{Unions}
-\label{sec:impl-unions}
-Mainly because type providers can only provide objects
-
-Unions become records with fields - so hiding null makes sense
-also it means the open-world semantics
+
+In this paper, we treat the XML, JSON and CSV formats uniformly as \emph{structural values}.
+The definition of structural values is, indeed, rich enough to capture all three formats. The
+structure is similar to JSON (it has primitive values, records and collections). We also added
+\emph{named} records, which are needed for XML.
+%
+\begin{itemize}
+\item When reading JSON, we directly create a corresponding structural value (with all records unnamed).
+  Optionally, we also read numerical values (and dates) stored in strings.
+\item When reading CSV, we read each row as an unnamed record and return them in a collection.
+  Optionally, the read values are parsed as numbers or Booleans and missing values become \kvd{null}.
+\item When reading XML, we create a named record for each node. Attributes become record fields and
+  body becomes a special field named $\bullet$
+\end{itemize}
+%
+The reading of XML documents is perhaps the most interesting. To demonstrate how the body is treated,
+consider the following:
+%
+{\small{
+\begin{verbatim}
+  <root id="1">
+    <item>Hello!</item>
+  </root>    
+\end{verbatim}
+}}
+%
+\noindent
+This XML becomes a record named \ident{root} with fields \ident{id} and $\bullet$. The nested element
+contains only the $\bullet$ field containing the inner text:
+%
+\begin{equation*}
+\ident{root}~\{ \ident{id} \mapsto 1, \bullet \mapsto [ \ident{item}~\{ \bullet \mapsto \str{Hello!} \}] \}
+\end{equation*}
+%
+When generating types for types inferred from XML, we also include a special case to remove the $\bullet$
+node using the fact that this can be only a collection (of elements) or a primitive value.
+
+Finally, the XML type provider in F\# Data includes an additional option to use \emph{global inference}.
+In that case, the inference algorithm from Section~\ref{sec:formal-inferval} is replaced with an alternative
+that unifies the types of all records with the same name. This is useful when the sample is, for example,
+an XHTML document -- in this case, all occurrences of an element (\emph{e.g.}~{\small\ttfamily <a>})
+are treated as the same type.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure*}
+\noindent
+\begin{equation*}
+\inference[(col)\;]
+  { \tytagof(\sigma_i) = \tytagof(\sigma'_i) \quad 
+    \sigma_i \tsep \sigma'_i \vdash \sigma''_i \quad
+    \phi_i \tsep \phi'_i \vdash \phi''_i \qquad (\forall i\in \{ 1 .. k \}) }
+  { \begin{array}{c}
+    [\sigma_1,\psi_1 | \ldots | \sigma_k,\psi_k | \ldots | \sigma_n,\psi_n ] \tsep 
+    [\sigma'_1,\psi'_1 | \ldots | \sigma'_k, \psi'_k | \ldots | \sigma'_m, \psi'_m ] \vdash\\[0.1em]
+    [\sigma''_1, \psi''_1 | \ldots | \sigma''_k, \psi''_k | 
+        \sigma_{k+1},\addopt{\psi_{k+1}} | \ldots | \sigma_{n},\addopt{\psi_{n}} |
+        \sigma'_{k+1},\addopt{\psi'_{k+1}} | \ldots | \sigma'_{m}, \addopt{\psi'_{m}} ]
+    \end{array} }
+\quad
+\begin{array}{rcl}
+ \\
+ \addopt{\phi} = \phi'&\narrow{\textnormal{such that}}&\phi \tsep 1? \vdash \phi' \\[0.1em]
+ \phi \tsep \phi' \vdash \phi&\narrow{\textnormal{when}}&\phi :> \phi' \\[0.1em]
+ \phi \tsep \phi' \vdash \phi'&\narrow{\textnormal{when}}&\phi' :> \phi
+\end{array}    
+\end{equation*}
+\caption{Inference judgement that defines the common supertype of two heterogeneous collections}
+\label{fig:subtyping-hetcol}
+\end{figure*}
+
+% -------------------------------------------------------------------------------------------------
 
@@
 
-Heterogeneous collections
-
-\subsection{Practical experience}
+When introducing the XML type provider (Section~\ref{sec:providers-xml}), we briefly mentioned 
+that F\# Data implements a special handling of heterogeneous collections. In the example, the provider
+generated a type with \ident{Title} member of type \ident{string} (corresponding to a nested element
+{\ttfamily\small <title>} that appears exactly once) and \ident{Items} member (returning a list of
+values obtained from multiple {\ttfamily\small <item>} elements).
+
+To capture this behaviour, we need to extend our earlier treatment of collections. Rather than storing
+a single type for the elements as in $[\sigma]$, we store multiple possible element types. However,
+this is not just a union type as we also store \emph{inferred multiplicity} of elements for 
+each of the types:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \psi &\narrow{=}& 1 \lsep 1? \lsep \ast \\
+ \sigma &\narrow{=}& ~\ldots \lsep [\sigma_1, \psi_1 | \ldots | \sigma_n, \psi_n ]
+\end{array}
+\end{equation*}
+%
+The multiplicities $1, 1?$ and $\ast$ represent \emph{exactly one}, \emph{zero or one} and \emph{zero or more},
+respectively. Thus, the inferred type of the collection in the RSS feed would be
+$[\ident{title}~\{ \ldots \}\;,1|\;\ident{item}~\{ \ldots \},\ast]$, which reads as a collection
+containing exactly one \ident{title} record and any number of \ident{item} records.
+
+The subtyping relation on heterogeneous collections specifies that a subtype can have fewer cases 
+provided that they do not have to appear exactly once. A subtype can also have a stricter specification
+of multiplicity (the ordering is $\ast :> 1? :> 1$).
+
+Finding a common supertype of heterogeneous collections is analogous to the handling of union types.
+The key rule is (\emph{col}) in Figure~\ref{fig:subtyping-hetcol}. It merges cases with the same tag
+(by merging both the type and the multiplicity). For cases that appear only in one collection, we ensure
+that the multiplicity is \emph{zero or one} or \emph{zero or more} using the auxiliary definition
+$\addopt{-}$.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Inferring and generating unions}
+\label{sec:impl-unions}
+
+The F\# Data type providers prefer types that do not contain unions (Theorem~\ref{thm:no-unions}).
+The motivation for this is twofold. First, the current support for type providers in the F\# compiler
+does not allow type providers to generate F\#-specific types (including discriminated unions). This
+means that the provided type cannot use idiomatic F\# representation. Second, when accessing data,
+using records aids the explorability -- when a value is a record, users can type ``.'' and most modern
+F\# editors provide auto-completion list with members.
+
+This choice has an important desirable consequence -- it allows treating a union with additional 
+cases as a subtype of another union and, hence, the provided code is guaranteed to work on \emph{more}
+inputs. Consider a type that is either a \ident{Person} record or just a string. 
+The type provider exposes it as the following F\# class:
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{type}~\ident{PersonOrString}~=  \\[0.1em]
+ \quad \kvd{member}~\ident{Person}~:~\ident{option}\langl \ident{Person}\rangl \\[0.1em]
+ \quad \kvd{member}~\ident{String}~:~\ident{option}\langl \ident{string}\rangl \\[0.1em]
+\end{array}
+\end{equation*}
+%
+If we provided an algebraic data type (discriminated union), the consumer would have to use pattern
+matching that would cover \emph{two} cases. The above type forces the user to also handle a third
+case, when both properties return \ident{None}. This also means that union types can silently skip
+over \kvd{null} values.
+
+That said, exposing a discriminated union (perhaps together with better tooling) is certainly an
+attractive alternative that we plan to consider in the future. This can also be combined with type 
+inference that chooses between records and union types based on some statistical metric (see
+related work in Section~\ref{sec:related}).
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Practical experiences}
+
+Sample is captured at compile time. It can come from the web or not (cache locally if you want
+but from the web has nice implications).
+
+We dont know what to do about schema change at runtime
+(for some data sources, we could do more clever things though)
+
+In practice..
 Error handling - solve issues just add another example
@@
 \section{Related and future work}
-
+\label{sec:related}
 ~
+
+Statistical metric makes the safety even messier (probabilistic!)
 

### 2015-02-26 icfp-2015/paper.tex
Adding references and related work

@@
 \usepackage{enumitem}
+\usepackage{flushend}
 
@@
 \noindent
-Although the F\# Data library has been widely adopted is one of the most downloaded F\# libraries 
+Although the F\# Data library \cite{fsharp-data} has been widely adopted is one of the most downloaded F\# libraries 
 it has not yet been presented in an academic form. Additional documentation and source code can 
@@
 
-Sample is captured at compile time. It can come from the web or not (cache locally if you want
-but from the web has nice implications).
-
-We dont know what to do about schema change at runtime
-(for some data sources, we could do more clever things though)
-
-In practice..
-Error handling - solve issues just add another example
+An important concern about using F\# Data in practice is the handling of schema change. When using a
+type provider, the sample is captured at compile-time. If the schema changes later (so that the actual
+input is no longer a subtype of the samples), the program using the type provider fails at run-time
+and it is the developer's responsibility to handle the exception. However, this is the same problem that
+happens when reading data using any other library.
+
+F\# Data can help discover such errors earlier. The example in Section~\ref{sec:introduction}
+points the JSON type provider to a sample using a live URL. This has the advantage that a re-compilation 
+fails when the schema changes, which is an indication that the program needs to be updated to reflect the
+change. If this is undesirable, it is always possible to cache the sample locally.
+
+In general, there is no better solution for plain XML, CSV and JSON data sources. Some data sources provide 
+versioning support (with meta-data about how the schema changed). For those, a type 
+provider could adapt automatically, but we leave this for future.
+
+An approach that we find useful in practice is to keep a set of samples. When the program fails at run-time
+(because a service returned data in an unexpected format), the new input can be added as an additional sample,
+which then shows what parts of code need to be modified.
 
@@
 \label{sec:related}
-~
-
-Statistical metric makes the safety even messier (probabilistic!)
-
-Write API
-
-~
-
-F\# and Agda type providers
-
-Kathleen Fisher's work
-
-JSON inference stuff
-
-LINQ and XML stuff
+
+The F\# Data library connects two lines of research that have been previously disconnected. The first is 
+extending the type systems of programming languages to accommodate external data sources and the second
+is inferring types for real-world data sources.
+
+The type provider mechanism has been introduced in F\# \cite{fsharp-typeprov,fsharp-typeprov-ddfp}, 
+added to Agda \cite{agda-tp} and used in areas such as semantic web \cite{liteq}. Our paper is novel 
+in that it shows the programming language theory behind a concrete type providers.
+
+\paragraph{Extending the type systems.} 
+A number of systems integrate external data formats into a programming language. Those include 
+XML \cite{xduce,xduce-ml} and databases \cite{links}. In both of these, the system either requires
+the user to explicitly define the schema (using the host language) or it has an ad-hoc extension 
+that reads the schema (\emph{e.g.}~from a database). LINQ \cite{linq} is more general, but relies
+on code generation when importing the schema.
+
+The work that is most similar to F\# Data is the XML and SQL integration in C$\omega$ \cite{comega-xs}.
+It extends a C\# like object-oriented language with types similar to our structured types 
+(including nullable types, choices with subtyping and heterogeneous collections with multiplicities).
+However, C$\omega$ does not infer the types from samples and extends the type system of the host
+language (rather than using a general purpose embedding mechanism).
+
+\paragraph{Advanced type systems.}
+Aside from type providers, a number of other advanced type system features could be used to
+tackle the problem discussed in this paper. The Ur \cite{ur} language has a rich system for working
+with records; meta-programming \cite{template-hask}, \cite{th-camlp4} and multi-stage programming \cite{multi-stage}
+could be used to generate code for the provided types. However, as far as we are aware, none of these 
+systems have been used to provide the same level of integration with XML, CSV and JSON.
+
+Another approach would be to use gradual typing \cite{gradual,gradual-js} and add types for structured
+data formats to existing dynamic language to check, for example, JSON manipulation in JavaScript.
+
+\paragraph{Typing real-world data.}
+The second line of research related to our work focuses on inferring structure of real-world data sets.
+A recent work on JSON \cite{typing-json} infers a succinct type using MapReduce to handle large number
+of samples. It fuses similar types based on a type similarity measure. This is more sophisticated than
+our technique, but it would make formally specifying the safety properties (Theorem~\ref{thm:safety}) difficult.
+Extending our \emph{relativized safety} property to a \emph{probabilistic safety} is an interesting 
+future work.
+
+The PADS project \cite{pads-dsl,pads-ml} tackles a more general problem of handling \emph{any} data format.
+The schema definitions in PADS are similar to our structural type. The structure inference for LearnPADS
+\cite{pads-learn} infers the data format from a flat input stream. A type provider for PADS could follow
+many of the patterns we explore in this paper, but formally specifying the safety property would be
+challenging.
 
@@
 
-% We recommend abbrvnat bibliography style.
+\cite{age-of-web}
 
 \bibliographystyle{abbrvnat}
-
-% The bibliography should be embedded for final submission.
-
-\begin{thebibliography}{}
-\softraggedright
-
-\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
-P. Q. Smith, and X. Y. Jones. ...reference text...
-
-\end{thebibliography}
+\bibliography{paper}
 

### 2015-02-26 icfp-2015/build.sh

@@ -1,0 +1,3 @@
+latex paper.tex --non-interactive
+bibtex paper
+pdflatex paper.tex

### 2015-02-27 icfp-2015/paper.tex
Adding conclusion and more refs

@@
 so the file name has to be a constant. The provider analyzes the sample and generates
-a type that we name \ident{People}. In F\# editors that use the F\# Compiler Service [?], the 
+a type that we name \ident{People}. In F\# editors that use the F\# Compiler Service, the 
 type provider is also executed at development-time and so the same provided types are
@@
 \paragraph{Erasing type providers.}
-At runtime, the type providers we describe use an erasure mechanism similar to Java Generics [?]. 
+At runtime, the type providers we describe use an erasure mechanism similar to Java Generics \cite{java-erasure}. 
 A type provider also generates code that is executed in place of the members of the generated types.
@@
 More details about the full implementation in F\# can be found in the paper on F\# type 
-providers [?].
+providers \cite{fsharp-typeprov}.
 
@@
 suggests that they are the most common issues. The following JSON response with government debt
-information returned by the World Bank API [?] demonstrates all three problems:
+information returned by the World Bank API demonstrates all three problems:
 %
@@
 We focus on the object system (which is what F\# type providers produce) and so the following combines 
-standard ML [?] features with Featherweight Java [?] classes. However, we only need classes with properties and
+standard ML \cite{sml} features with Featherweight Java \cite{fwjava} classes. However, we only need classes with properties and
 without inheritance. A class has a single implicit constructor and the declaration closes over 
@@
 supertype $\sigma$ from the samples. From $\sigma$, they then generate F\# types that are then exposed to the
-programmer\footnote{The actual implementation provides \emph{erased types} as described in [?]. Here, we treat
-the code as actually generated. This is an acceptable simplification, because F\# Data type providers do not 
+programmer\footnote{The actual implementation provides \emph{erased types} as described in \cite{fsharp-typeprov}. 
+Here, we treat the code as actually generated. This is an acceptable simplification, because F\# Data type providers do not 
 rely on laziness that is available through erased types.}.
@@
 
-Using the standard approach to syntactic type safety [?], we prove the type preservation 
+Using the standard approach to syntactic type safety  \cite{syntactic}, we prove the type preservation 
 (reduction does not change type) and progress (an expression that is not a value can be reduced).
@@
 for programming langauges without type providers. However, it reflects the reality of programming with external
-data sources that is increasingly important in the age of web [?]. So, type providers do not reduce the safety
--- they simply make the existing issues visible.
+data sources that is increasingly important in the age of web \cite{age-of-web}. So, type providers do not
+reduce the safety -- they simply make the existing issues visible.
 
@@
 checked code are generated. Extending our formalism to capture this aspect is an interesting future work
-relevant to other type providers -- including the World Bank type provider [?] that is also available
+relevant to other type providers -- including the World Bank type provider that is also available
 in the F\# Data library.
@@
 
-  - Prime example of type providers, explores relativized type safety
-  - Used by industry, downloads, tested, etc. 
-  - Predictable yet powerful type inference
+In this paper, we explored the F\# Data type providers for structured data formats such as XML, CSV and JSON.
+As most real-world data do not come with an explicit schema, the library uses \emph{type inference} that
+deduces a type from a set of samples. Next, it uses the \emph{type provider} mechanism to integrate the 
+inferred type directly into the F\# type system.
+
+The type inference algorithm we use is based on a common supertype relation. For usability reasons, the
+algorithm attempts to infer type that does not contain unions and prefers records with optional fields instead.
+The algorithm is simple and predictable, which is important as developers need to understand how changing the
+samples affects the resulting types.
+
+In the second part of the paper, we explore the programming language theory behind type providers. F\# Data
+is a prime example of type providers, but our work also demonstrates a more general point. The types generated
+by type providers can depend on external input (such as samples) and so we can only formulate \emph{relativized
+safety property}, which says that a program is safe only if the actual inputs satisfy additional conditions --
+in our case, they have to be subtypes of one of the samples.
+
+The type provider mechanism has been described before, but this paper is novel in that it explores concrete
+type providers from the perspective of programming language theory. This is particularly important as the
+F\# Data library is becoming de-facto standard tool for data access in F\#\footnote{At the time of writing, 
+the library has 37,000 downloads on NuGet; over 1600 commits and 36 contributors on GitHub.} and other 
+languages are begining to adopt mechanisms similar to F\# type providers.
 
 \acks
-Acknowledgments, if needed.
-
-\cite{age-of-web}
+We would like to thank to the F\# Data contributors on GitHub and other colleagues working on type providers,
+including Jomo Fisher, Keith Battocchi and Kenji Takeda.
 

### 2015-02-27 icfp-2015/paper.tex
Spellechecking

@@
 \ident{item} nodes (individual news articles). In RSS, the \ident{title} nodes always contain plain text,
-while \ident{item} nodes contain \ident{title} (and also \ident{url} and othe properties omitted here).
+while \ident{item} nodes contain \ident{title} (and also \ident{url} and other properties omitted here).
 
@@
 
-The grammar below defines \emph{structural type} $\sigma$. We distinguish between \emph{non-nullale types}
+The grammar below defines \emph{structural type} $\sigma$. We distinguish between \emph{non-nullable types}
 that always have a valid value (written as $\hat{\sigma}$) and \emph{nullable types} that encompass missing 
@@
 in Section~\ref{sec:impl-unions}, this is the right choice for an F\# type provider, but it would not 
-necessarilly be a good fit for other languages).
+necessarily be a good fit for other languages).
 
@@
 %
-Finding a common supertype of when one type is a union is more complicated. Our definiion aims to 
+Finding a common supertype of when one type is a union is more complicated. Our definition aims to 
 use a flat structure (avoid nesting unions) and limit the number of cases. This is done by grouping
@@
   which is not defined in case of arithmetic overflow; conversion to and from string values written as
-  ``$t$~represents~$i$'', which parses or formats a non-string value and also trating of decimals 
+  ``$t$~represents~$i$'', which parses or formats a non-string value and also treating of decimals 
   0.0 and 1.0 as Booleans.
@@
 \paragraph{Example 2.} The second example illustrates the remaining parts of the type provision,
-including collections and union types. Teusing the \ident{Person} type from the previous example,
+including collections and union types. Reusing the \ident{Person} type from the previous example,
 consider a collection $[\ident{Person}+\ident{string}]$, which contains a mix of \ident{Person} and string values.
@@
 
-\subsection{Infering types from values}
+\subsection{Inferring types from values}
 \label{sec:formal-inferval}
@@
 \item[--] Records in the actual input can have additional fields
-\item[--] Records in the actual input can have fewer fileds, provided that the type of the fields is marked as optional in the sample
+\item[--] Records in the actual input can have fewer fields, provided that the type of the fields is marked as optional in the sample
 \item[--] Union types in the input can have both fewer or more cases
@@
 Now that we know that the provided types are correct with respect to the subtyping relation, we can
-look at the main theorem of the paper. It statest that, for any input (which is a subtype of any of the
+look at the main theorem of the paper. It states that, for any input (which is a subtype of any of the
 samples) and any expression $e$, a well-typed program that uses the provided types does not ``go wrong''.
@@
 The Lemma~\ref{thm:tp-correctness} guarantees that this does not happen. In Theorem~\ref{thm:safety},
-we carefuly state that we only consider expressions $e_c$ which ``[do] not contain any structural 
+we carefully state that we only consider expressions $e_c$ which ``[do] not contain any structural 
 values as sub-expressions''. This makes sure that the only code working with structural values is the
@@
 The \emph{relativized safety} property does not guarantee the same amount of safety as standard type safety
-for programming langauges without type providers. However, it reflects the reality of programming with external
+for programming languages without type providers. However, it reflects the reality of programming with external
 data sources that is increasingly important in the age of web \cite{age-of-web}. So, type providers do not
@@
 generating code. For XML, CSV and JSON, both models can be used. However, one interesting aspect of erasing 
-type providers is that the types can be generated lazilly -- only classes that are actually used in the type
+type providers is that the types can be generated lazily -- only classes that are actually used in the type
 checked code are generated. Extending our formalism to capture this aspect is an interesting future work
@@
 
-\subsection{Parsing strucutred values}
+\subsection{Parsing structured values}
 \label{sec:impl-parsing}
@@
 the library has 37,000 downloads on NuGet; over 1600 commits and 36 contributors on GitHub.} and other 
-languages are begining to adopt mechanisms similar to F\# type providers.
+languages are beginning to adopt mechanisms similar to F\# type providers.
 

### 2015-02-27 icfp-2015/paper.tex
structured format/data vs structural value/type

@@
 % TODO: Maybe get rid of \mu_opt for records everywhere
-% TODO: Disambigurate 's' as structural value and 's' as string
 % TODO: Explain why we can only provide simple classes (!)
 % TODO: Replace \nu_opt with \nu_? or something else that does not look shit
-% TODO: Cleanup "strucutred" vs "structural"
 % TODO: Insert a few more mentions of why "s1 + s2 <: s1 + s2 + s3" is valid (cannot do complete pattern match)
@@
 
-To provide a basic intuition, the subtyping relation between structured types is illustrated in
+To provide a basic intuition, the subtyping relation between structural types is illustrated in
 Figure~\ref{fig:subtyping-diagram}. We split the diagram in two parts. The upper part shows 
@@
 
-As mentioned earlier, the structured type inference relies on a finding common supertype. However,
+As mentioned earlier, the structural type inference relies on a finding common supertype. However,
 the partially ordered set of types does not have a unique greatest lower bound. For example, record types
@@
 property of F\# Data type providers in Section~\ref{sec:safety}. We start by discussing the runtime 
-representation of structured values and runtime conversions (Section~\ref{sec:formal-convert}). Then 
+representation of structural values and runtime conversions (Section~\ref{sec:formal-convert}). Then 
 we embed these into a formal model of an F\# subset that is necessary for discussing type providers 
@@
 uses a different implementation for each format (reusing existing libraries).}. Structural values are 
-first-order and can be one of the primitive values (Boolean, string, integer, decimal and float), a 
-missing value (\kvd{null}), a collection and a record (named or unnamed):
+first-order and can be one of the primitive values (Boolean \kvd{true} or \kvd{false}, string $t$, 
+integer $i$, decimal $d$ and floating-point number $f$), a missing value (\kvd{null}), a collection 
+and a record (named or unnamed):
 %
@@
     &\narrow{|}& \ident{asStr}(s) \lsep \ident{getChildren}(s) \lsep \ident{getField}(\nu_{\ident{opt}}, \nu, s)\\
-    &\narrow{|}& \ident{isNull}(s) \lsep \ident{isTag}(t, s)
+    &\narrow{|}& \ident{isNull}(s) \lsep \ident{isTag}(\ident{tag}, s)
 \end{array}
@@
 \noindent
-In the above definition, $\tau$ denotes types including a type of all structured values $s$ called
+In the above definition, $\tau$ denotes types including a type of all structural values $s$ called
 \ident{StructVal}. A class definition $L$ consists of a constructor and zero or more members. Values $v$ include 
@@
 \begin{itemize}[noitemsep]
-\item[--] All structured values $s$ have a type \ident{StructVal}. Some of those have other types
+\item[--] All structural values $s$ have a type \ident{StructVal}. Some of those have other types
   (primitive integers, decimals, floats, strings and Booleans as illustrated by the rule for $n$).
@@
 
-So far, we defined the type inference algorithm for structured data formats which produces a structured
+So far, we defined the type inference algorithm for structured data formats which produces a structural
 type $\sigma$ based on one or more sample documents (Section~\ref{sec:inference}). In Section~\ref{sec:formal-ff},
@@
 \paragraph{Type provider mapping.}
-When generating types, the type provider produces an F\# type $\tau$, expression that wraps a structured 
+When generating types, the type provider produces an F\# type $\tau$, expression that wraps a structural 
 document value (of type \ident{StructVal}) as a value of type $\tau$ and a collection of class definitions. 
@@
 %
-The semantics is parameterized by an expression $e$, which represents code to obtain a structured value that is
-being wrapped. Then, given an inferred structured type $\sigma$, the type provider produces an F\# type $\tau$,
+The semantics is parameterized by an expression $e$, which represents code to obtain a structural value that is
+being wrapped. Then, given an inferred structural type $\sigma$, the type provider produces an F\# type $\tau$,
 an expression $e'$ which constructs a value of type $\tau$ using $e$ and also a set of class definitions $L$.
@@
 
-Handling of records is more interesting. We generate a new class $C$ that takes a structured value as constructor
+Handling of records is more interesting. We generate a new class $C$ that takes a structural value as constructor
 parameter. For each record field, we generate a new member with the same name as the field\footnote{The actual
 F\# Data implementation also capitalizes the names.} The body of the member calls \ident{getField} and then
-passes this expression to $\sem{\sigma_i}$ which adds additional wrapping that maps the field (structured value
+passes this expression to $\sem{\sigma_i}$ which adds additional wrapping that maps the field (structural value
 of type $\sigma_i$) into an F\# type $\tau_i$. The returned expression creates a new instance of $C$ and
@@
 For completeness, a type corresponding to the record is \ident{Person} and given a 
-structural document $s$, we create a \ident{Person} value by calling $\kvd{new}~\ident{Person}(s)$.
+structural value $s$, we create a \ident{Person} value by calling $\kvd{new}~\ident{Person}(s)$.
 
@@
 
-\subsection{Parsing structured values}
+\subsection{Parsing structured data}
 \label{sec:impl-parsing}
@@
 The work that is most similar to F\# Data is the XML and SQL integration in C$\omega$ \cite{comega-xs}.
-It extends a C\# like object-oriented language with types similar to our structured types 
+It extends a C\# like object-oriented language with types similar to our structural types 
 (including nullable types, choices with subtyping and heterogeneous collections with multiplicities).
@@
 The schema definitions in PADS are similar to our structural type. The structure inference for LearnPADS
-\cite{pads-learn} infers the data format from a flat input stream. A type provider for PADS could follow
+\cite{pads-learn} infers the data format from a flat input stream. A PADS type provider could follow
 many of the patterns we explore in this paper, but formally specifying the safety property would be

### 2015-02-27 icfp-2015/paper.tex
Use \nu? rather than \nu_opt for optional record names

@@
 that always have a valid value (written as $\hat{\sigma}$) and \emph{nullable types} that encompass missing 
-and \kvd{null} values (written as $\sigma$). We write $\nu$ for record field names and $\nu_{\ident{opt}}$
+and \kvd{null} values (written as $\sigma$). We write $\nu$ for record field names and $\nu?$
 for record names, which can be empty:
@@
 \begin{array}{rcl}
- \hat{\sigma} &\narrow{=}& \nu_{\ident{opt}}\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
+ \hat{\sigma} &\narrow{=}& \nu?\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                 &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bit} \lsep \ident{bool} \lsep \ident{string} 
@@
 \label{eq:sub-record1}\tag{R1}
- \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> 
- \nu~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
- \;\; (\sigma_i :> \sigma_i')
+ \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \hspace{-0.1em}:> \hspace{-0.1em}
+ \nu?~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
+ \; (\sigma_i \hspace{-0.25em}:>\hspace{-0.25em} \sigma_i')
 \end{align}
@@
 \label{eq:sub-record2}\tag{R2}
- \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> 
- \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_m\!:\!\sigma_m \}
- \;\; (m \geq n)
+ \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \hspace{-0.1em} :> \hspace{-0.1em}
+ \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_m\!:\!\sigma_m \}
+ \; (m \hspace{-0.1em}\geq\hspace{-0.1em} n)
 \end{align}
@@
 \begin{array}{l}
- \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
- \nu~\{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, .., \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
+ \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
+ \nu?~\{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, .., \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
 \end{array} \qquad &(\textnormal{permutation}~\pi)
@@
 \begin{array}{l}
- \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null}, .., \nu_{n+m}\!:\!\kvd{null} \} :> \\
- \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
+ \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null}, .., \nu_{n+m}\!:\!\kvd{null} \} :> \\
+ \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
 \end{array}
@@
   { \begin{array}{l}
-    \nu \; \{ \nu_1 : \sigma_1,  \; \ldots \;, \nu_k : \sigma_k, \; \ldots \;, \nu_n : \tau_n \} \tsep
-    \nu \; \{ \nu'_1 : \sigma'_1, \; \ldots \;, \nu'_k : \sigma'_k, \; \ldots \;, \nu'_m : \tau'_m \} \vdash\\
-    \nu \; \{ \nu_1 : \sigma''_1, \; \ldots \; , \nu_k : \sigma''_k, 
+    \nu? \; \{ \nu_1 : \sigma_1,  \; \ldots \;, \nu_k : \sigma_k, \; \ldots \;, \nu_n : \tau_n \} \tsep
+    \nu? \; \{ \nu'_1 : \sigma'_1, \; \ldots \;, \nu'_k : \sigma'_k, \; \ldots \;, \nu'_m : \tau'_m \} \vdash\\
+    \nu? \; \{ \nu_1 : \sigma''_1, \; \ldots \; , \nu_k : \sigma''_k, 
                             \nu_{k+1} : \addopt{\sigma_{k+1}}, \ldots, \nu_n : \addopt{\sigma_n},
@@
 \begin{equation*}
-\hspace{2em}
-\inference[(order)\;]
-  { \nu \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \tsep \sigma \vdash \sigma' }
-  { \nu \; \{ \nu_{\pi(1)} : \sigma_{\pi(1)}, \ldots, \nu_{\pi(n)} : \sigma_{\pi(n)} \} \tsep \sigma \vdash \sigma' }
+\hspace{1em}
+\inference[(order-1)\;]
+  { \nu? \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \tsep \sigma \vdash \sigma' }
+  { \nu? \; \{ \nu_{\pi(1)} : \sigma_{\pi(1)}, \ldots, \nu_{\pi(n)} : \sigma_{\pi(n)} \} \tsep \sigma \vdash \sigma' }
 \quad
-\inference
+\inference[(order-2)\;]
   { \sigma_1 + \ldots + \sigma_n \tsep \sigma \vdash \sigma' }
@@
  op &\narrow{=}& \ident{asInt}(s) \lsep \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{asBool}(s)  \\
-    &\narrow{|}& \ident{asStr}(s) \lsep \ident{getChildren}(s) \lsep \ident{getField}(\nu_{\ident{opt}}, \nu, s)\\
+    &\narrow{|}& \ident{asStr}(s) \lsep \ident{getChildren}(s) \lsep \ident{getField}(\nu?, \nu, s)\\
     &\narrow{|}& \ident{isNull}(s) \lsep \ident{isTag}(\ident{tag}, s)
@@
 \begin{array}{l}
- \sem{\nu_{\ident{opt}}\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}}_e = \\[0.1em]
+ \sem{\nu?\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}}_e = \\[0.1em]
  \quad C, \kvd{new}~C(e), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
@@
  \qquad M_i = \kvd{member}~\nu_i:\tau_i=e_i\\[0.1em]
- \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_{e'},\;\; e'=\ident{getField}(\nu_{\ident{opt}}, \nu_i, e)\\[0.6em]
+ \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_{e'},\;\; e'=\ident{getField}(\nu?, \nu_i, e)\\[0.6em]
 \end{array}
@@
  \\[0.5em]
- \semalt{\nu_{\ident{opt}}~\{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \}} =\\[0.1em]
- \qquad\nu_{\ident{opt}}~\{ \nu_1:\semalt{s_1}, \ldots, \nu_n :\semalt{s_n} \}
+ \semalt{\nu?~\{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \}} =\\[0.1em]
+ \qquad\nu?~\{ \nu_1:\semalt{s_1}, \ldots, \nu_n :\semalt{s_n} \}
  \\[0.5em]

### 2015-04-13 icfp-2014-src/finals/paper.tex
Add ICFP SRC finals, also add some forgotten changes

\documentclass[preprint]{sigplanconf}

\usepackage{amssymb}
\usepackage{multicol}
\usepackage{xcolor}
\usepackage{graphicx}
\usepackage[fleqn]{amsmath}
\usepackage[hang,flushmargin]{footmisc}
\usepackage{graphics}
\usepackage{stmaryrd}
\usepackage{amsthm}
\usepackage[hyphens]{url}
\usepackage{hyperref}
\usepackage[notquote]{hanging}
\usepackage[T1]{fontenc}
\usepackage{semantic}
\usepackage{enumitem}
\usepackage{flushend}

% =================================================================================================

% TODO: The model here does not actually use type erasure!!
% TODO: Explain why we can only provide simple classes (!)
% TODO: Insert a few more mentions of why "s1 + s2 <: s1 + s2 + s3" is valid (cannot do complete pattern match)

\urlstyle{sf}

\makeatletter
\renewcommand{\@makefntext}[1]{%
  \parindent 1em%
  \raggedright
  \begin{hangparas}{0.8em}{1}
  \noindent {$^{\@thefnmark}$~#1}
  \end{hangparas}
}
\makeatother

\newcommand{\langl}{\begin{picture}(4.5,7)
\put(1.1,2.5){\rotatebox{60}{\line(1,0){5.5}}}
\put(1.1,2.5){\rotatebox{300}{\line(1,0){5.5}}}
\end{picture}}
\newcommand{\rangl}{\begin{picture}(4.5,7)
\put(.9,2.5){\rotatebox{120}{\line(1,0){5.5}}}
\put(.9,2.5){\rotatebox{240}{\line(1,0){5.5}}}
\end{picture}}

\newcommand{\lang}{\begin{picture}(5,7)
\put(1.1,2.5){\rotatebox{45}{\line(1,0){6.0}}}
\put(1.1,2.5){\rotatebox{315}{\line(1,0){6.0}}}
\end{picture}}
\newcommand{\rang}{\begin{picture}(5,7)
\put(.1,2.5){\rotatebox{135}{\line(1,0){6.0}}}
\put(.1,2.5){\rotatebox{225}{\line(1,0){6.0}}}
\end{picture}}

\newcommand{\llangl}{\langl\hspace{-0.35em}\langl}
\newcommand{\rrangl}{\rangl\hspace{-0.35em}\rangl}

\definecolor{cmtclr}{rgb}{0.0,0.6,0.0}
\definecolor{numclr}{rgb}{0.0,0.4,0.0}
\definecolor{kvdclr}{rgb}{0.0,0.0,0.6}
\definecolor{strclr}{rgb}{0.5,0.1,0.0}
\definecolor{prepclr}{rgb}{0.0,0.0,0.0}

\newcommand{\kvd}[1]{\textnormal{\textcolor{kvdclr}{\sffamily #1}}}
\newcommand{\num}[1]{\textnormal{\textcolor{numclr}{\sffamily #1}}}
\newcommand{\str}[1]{\textnormal{\textcolor{strclr}{\sffamily "#1"}}}
\newcommand{\strf}[1]{\textnormal{\textcolor{strclr}{\sffamily #1}}}
\newcommand{\ident}[1]{\textnormal{\sffamily #1}}
\newcommand{\lident}[1]{\textnormal{\sffamily\`{}\hspace{-0.25em}\`{}\hspace{-0.1em}#1\`{}\hspace{-0.25em}\`{}}}
\newcommand{\cmt}[1]{\textit{\sffamily\textcolor{cmtclr}{#1}}}

\newcommand{\lsep}[0]{\;\; | \;\;}
\newcommand{\narrow}[1]{\hspace{-0.85em} #1 \hspace{-0.85em}}

\newcommand{\tsep}[0]{\; \triangledown \;}
\newcommand{\tytag}{\ident{tag}}
\newcommand{\dropopt}[1]{\lfloor#1\rfloor}
\newcommand{\addopt}[1]{\lceil#1\rceil}
\newcommand{\tytagof}{\ident{tagof}}
\newcommand{\nameoftag}{\ident{nameof}}

\newcommand{\reduce}{\rightsquigarrow}

\newcommand{\sem}[1]{\llbracket #1 \rrbracket}
\newcommand{\semalt}[1]{\llangl #1 \rrangl}

\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}

% =================================================================================================

\makeatletter
\def\@copyrightspace{\relax}
\makeatother
\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}
\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country}
\copyrightyear{20yy}
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\doi{nnnnnnn.nnnnnnn}

%\titlebanner{Unpublished draft, March 2015}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{F\# Data: \textnormal{Accessing structured data made easy}}
%\subtitle{Subtitle Text, if any}

\authorinfo{Tomas Petricek}
           {University of Cambridge}
           {tomas@tomasp.net}
\maketitle

% =================================================================================================

\section{Introduction}

Programming may not be the new \emph{literacy}, but it is finding its way into many areas
of modern society. For example, making sense of large amounts of data that are increasingly made
available through open government data initiatives\footnote{In the US (\url{http://data.gov}) and
in the UK (\url{http://data.gov.uk}).} is almost impossible without some programming skills.
In media, \emph{data journalism} reflects this development. Data journalists still write articles
and focus on stories, but programming is at the core of their work.

Improving support for data access in programming language can make understanding data simpler,
more usable and reproducible:

\begin{itemize}
  \item \textbf{Simpler.} Many programming languages treat data as foreign entities that have to
    be parsed or processed. Instead, data should be treated as first-class entities fully integrated
    with the rest of the programming language.
  \item \textbf{Usability.} Modern developer tools make coding easier with auto-complete and early
    error checking. Unfortunately, these typically rely on static types which, in turn, make
    programming with standard untyped data harder.
  \item \textbf{Reproducible.} Data journalists often use a wide range of tools (including Excel,
    scripts and other ad-hoc tools). This makes it hard to reproduce the analysis and detect
    errors when the input data changes.
\end{itemize}

\noindent
The presented work reconcilles the \emph{simplicity} of data access in dynamically-typed programming
languages with the \emph{usability} and \emph{reproducibility} provided by statically-typed
languages. More specifically, we develop F\# type providers for accessing data in structured
data formats such as CSV, XML and JSON, which are frequently used by open government data initiatives
as well as other web-based data sources.

% =================================================================================================

\section{Motivation: Accessing structured data}

Despite numerous schematization efforts, most data on the web is available without an explicit
schema. At best, the documentation provides a number of typical requests and sample responses.
For simplicity, we demonstrate the problem using the OpenWeatherMap service, which can be used to get
the current weather for a given city\footnote{See ``Current weather data'': \url{http://openweathermap.org/current}}.
The page documents the URL parameters and shows one sample JSON response to illustrate the
response structure.

\paragraph{Statically-typed.}
In a statically typed functional language like F\#, we could use a library for working with HTTP
and parsing JSON to call the service and read the temperature. Here, the parsing library returns a
value of a \ident{JsonValue} data type and we use pattern matching to extract the value we
need\footnote{We abbreviate the full URL:
\url{http://api.openweathermap.org/data/2.5/weather?q=Prague\&units=metric}}:

\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{data}=\ident{Http.Request}(\str{http://weather.org/?q=Prague}) \\[0.1em]
 \kvd{match}~\ident{JsonValue.Parse}(\ident{data})~\kvd{with} \\[0.1em]
 |~\ident{Record}(\ident{root})\rightarrow \\[0.1em]
 \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\[0.1em]
 \quad |~\ident{Record}(\ident{main})\rightarrow \\[0.1em]
 \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\[0.1em]
 \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f degrees!}~\ident{num} \\[0.1em]
 \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\[0.1em]
 \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\[0.1em]
 |~\_\rightarrow \ident{failwith}~\str{Incorrect format}
\end{array}
\end{equation*}
\vspace{0.1em}

\noindent
The pattern matching assumes that the response has a
particular format (as described in the documentation). The root node must be a record with a
\str{main} field, which has to be another record containing a \str{temp} field with a numerical
value. When the format is incorrect, the data access simply fails with an exception.

The code is complicated, because data is parsed into a fully general data structure that we then
process. The code is not benefiting from the generality of the data structure -- quite the opposite!

\paragraph{Dynamically-typed.}
Doing the same in JavaScript is shorter and simpler (not surprisingly, as JSON has been designed
after a subset of JavaScript). Using jQuery to perform the request, we can write:
%
\begin{equation*}
\begin{array}{l}
\ident{jQuery.ajax}(\str{http://weather.org/?q=Prague}, \\
\quad \kvd{function}(\ident{data})~\{ \\
\qquad \kvd{var}~\ident{obj} = \ident{JSON.parse}(\ident{data}); \\
\qquad \ident{write}(\str{Lovely }, \ident{obj.main.temp}, \str{ degrees!}); \\
\quad \});
\end{array}
\end{equation*}
%
Although the code is shorter, writing it is not \emph{easier} than writing the original
statically-typed version. Even though some JavaScript editors provide auto-completion, they
will fail to help us here, because they have no knowledge of the shape of the object \ident{obj}.
So, the author will have to open the documentation and guess the available fields from the
provided sample.

\paragraph{Type providers.}
This paper presents the F\# Data library that implements \emph{type providers} for accessing
structured  data formats such as XML, JSON and CSV. Using the JSON type provider, we can write code
with the same functionality in three lines, but with full editor support including auto-complete
on the \ident{obj} object:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://weather.org/?q=Prague}\rangl \\
 \kvd{let}~\ident{obj} = \ident{W.GetSample}() \\
 \ident{printfn}~\str{Lovely \%f degrees!}~\ident{obj.main.temp}
\end{array}
\end{equation*}
\vspace{0.0em}

\noindent
On the first line, $\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider at
compile-time with the URL as a sample. The type provider infers the structure of the response
from the sample and provides a type that has a statically known property \ident{main}, returning
an object with a property \ident{temp} that provides the temperature as a number.

This gives us the best of both worlds -- the \emph{simplicity} of dynamic typing with the
\emph{usability}, \emph{safety} and associated tooling common in statically-typed languages.

% =================================================================================================

\section{Background: Type providers}

This paper presents a collection of \emph{type providers} for integrating structured data into the
F\# programming language. As outlined in the previous example, our key technical contribution is the
algorithm that infers appropriate type from an example document and and the type providers that
expose the type. In this section, we give a brief overview of the type provider mechanims and of
related approaches to integrating data into programming languages.

\subsection{How type providers work}

Documents in the JSON format consists of several possible kinds of values. The OpenWeatherMap
example in the introduction used only (nested) record and a numerical value. To demonstrate other
aspects, we look at a more complex example that also invloves collections and strings:
%
{\small{
\begin{verbatim}
  [ { "name": "Jan", "age": 25 },
    { "name": "Alexander", "age": 3.5 },
    { "name": "Tomas" } ]
\end{verbatim}
}}
%
\noindent
Say we want to print the names of people in the list with an age if it is available. Assuming
\strf{people.json} contains the above sample and \ident{data} is a string value
that contains another data set in the same format, we can use \ident{JsonProvider} as follows:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{People}~=~\ident{JsonProvider}\langl\str{people.json}\rangl\hspace{1em} \\ [0.6em]
 \kvd{let}~\ident{items}~=~\ident{People.Parse}(\ident{data})\\
 \kvd{for}~\ident{item}~\kvd{in}~\ident{items}~\kvd{do}\\
 \quad\ident{printf}~\str{\%s }~\ident{item.name}\\
 \quad\ident{Option.iter}~(\ident{printf}~\str{(\%f)})~\ident{item.age}
\end{array}
\end{equation*}
%
In contrast to the earlier example, the example now uses a local file \strf{people.json} as a
\emph{representative sample} for the type inference, but then processes data (available at
run-time) from another source.

\paragraph{Type providers.}
The notation $\ident{JsonProvider}\langl\str{people.json}\rangl$ on the first line passes a
\emph{static parameter} to the type provider. Static parameters are resolved at compile-time,
so the file name has to be a constant. The provider analyzes the sample and generates
a type that we name \ident{People}. In F\# editors, the type provider is also executed at
development-time and so the same provided types are used in code completion.

The \ident{JsonProvider} uses a type inference algorithm discussed below and infers the
following types from the sample:
%
\begin{equation*}
\begin{array}{l}
 \kvd{type}~\ident{Entity}~=  \\
 \quad \kvd{member}~\ident{name}~:~\ident{string} \\
 \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{decimal}\rangl \\[0.6em]
 \kvd{type}~\ident{People}~=  \\
 \quad \kvd{member}~\ident{GetSample}~:~\ident{unit}~\rightarrow~\ident{Entity}[] \\
 \quad \kvd{member}~\ident{Parse}~:~\ident{string}~\rightarrow~\ident{Entity}[] \\
\end{array}
\end{equation*}
%
The type \ident{Entity} represents the person. The field \ident{Name} is available for all
sample values and is inferred as \ident{string}. The field \ident{Age} is marked as optional,
because the value is missing in one sample. The two sample ages are an integer $25$ and a
decimal $3.5$ and so the common inferred type is \ident{decimal}.

The type \ident{People} provides two methods for reading data. \ident{GetSample} returns the
sample used for the inference and \ident{Parse} parses a JSON string containing data in the same
format as the sample. Since the sample JSON is a collection of records, both methods return an
array of \ident{Entity} values.

\paragraph{Erasing type providers.}
At compile-time, F\# type providers use an erasure mechanism similar to Java Generics \cite{java-erasure}.
A type provider generates types and code that should be executed when members are accessed. In compiled
code, the types are erased and the program directly executes the generated code.
In the above example, the compiled (and actually executed) code looks as follows:

\noindent
\begin{equation*}
\begin{array}{l}
 \kvd{let}~\ident{items}~=~\ident{asArray}(\ident{JsonValue.Parse}(\ident{data}))\\
 \kvd{for}~\ident{item}~\kvd{in}~\ident{items}~\kvd{do}\\
 \quad\ident{printf}~\str{\%s }~\ident{asString}~(\ident{getProp}~\str{name}~\ident{item})\\
 \quad\ident{Option.iter}~(\ident{printf}~\str{(\%f)})\\
 \qquad(\ident{Option.map}~\ident{asFloat}~(\ident{tryGetProp}~\str{age}~\ident{item}))
\end{array}
\end{equation*}
%
The generated type \ident{Entity} is erased to a type $\ident{JsonValue}$, which represents any
JSON value and is returned by the \ident{Parse} method. The remaining properties are erased to
calls to various operations of the type provider runtime such as \ident{asArray}, \ident{getProp}
or \ident{asFloat} that attempt to convert a JSON value into the required structure (and produce
a run-time exception if this is not possible).

The (hidden) type erasure process turns the static provided types into code that we might write
without type providers. In particular, checked member names become unchecked strings. A type
provider cannot remove all possibilities for a failure -- indeed, an exception still occurs if the
input does not have the right format, but it simplifies writing code and removes most errors when
a representative sample is provided.

\subsection{Type systems and data integration}
The F\# Data library connects two lines of research that have been previously disconnected. The
first is extending the type systems of programming languages and the second is inferring the
structure of real-world data sources.

The type provider mechanism has been introduced in F\# \cite{fsharp-typeprov,fsharp-typeprov-ddfp}
and used in areas such as semantic web \cite{liteq}. The library presented in this paper is the
most widely used library of type providers and it is also novel in that it shows the programming
language theory behind a concrete type provider.

\paragraph{Extending the type systems.}
A number of systems integrate external data formats into a programming language. Those include
XML \cite{xduce,xduce-ml} and databases \cite{links}. In both of these, the system either requires
the user to explicitly define the schema (using the host language) or it has an ad-hoc extension
that reads the schema (\emph{e.g.}~from a database). LINQ \cite{linq} is more general, but relies
on code generation when importing the schema.

The work that is most similar to F\# Data is the XML and SQL integration in C$\omega$ \cite{comega-xs}.
It extends C\# with types capable of representing structured data formats, but it does not infer
the types from samples and it modifies the C\# language (rather than using a general purpose
embedding mechanism).

Aside from type providers, a number of other advanced type system features could be used to
tackle the problem discussed in this paper. The Ur \cite{ur} language has a rich system for working
with records; meta-programming \cite{template-hask}, \cite{th-camlp4} and multi-stage programming
\cite{multi-stage} could be used to generate code for the provided types. However, as far as we
are aware, none of these systems have been used to provide the same level of integration with
XML, CSV and JSON.

\paragraph{Typing real-world data.}
The second line of research related to our work focuses on inferring structure of real-world data sets.
A recent work on JSON \cite{typing-json} infers a succinct type using MapReduce to handle large number
of samples. It fuses similar types based on a type similarity measure. This is more sophisticated than
our technique, but it would make formally specifying the safety properties difficult.

The PADS project \cite{pads-dsl,pads-ml} tackles a more general problem of handling \emph{any} data format.
The schema definitions in PADS are similar to our structural type. The structure inference for LearnPADS
\cite{pads-learn} infers the data format from a flat input stream. A PADS type provider could follow
many of the patterns we explore in our work, but formally specifying the safety property would be
again challenging.

% =================================================================================================

\section{Approach and uniqueness}

The F\# Data library relies on two key techniques. First, it implements a type inference algorithm
that generates a suitable type from one or more sample documents. Second, it implements a type
provider that turns this \emph{structural type} into an ordinary F\# type that is then used by
the programmers. The mapping preserves an important property that we call \emph{relativized type
safety}. Informally, we prove that when the provided sample is \emph{representative}, code written
using the type provider will not fail.

Whether a sample is \emph{representative} is formally described using a subtyping relation
on structural types $\sigma$. The full definition of the relationship can be found in our recent
report \cite{fsharp-data-paper}. In this paper, we provide a brief overview and discuss one
interesting case in detail.

\subsection{Structural types}
\label{sec:inference-types}

The grammar below defines a \emph{structural type} $\sigma$. We distinguish between \emph{non-nullable types}
that always have a valid value (written as $\hat{\sigma}$) and \emph{nullable types} that encompass missing
and \kvd{null} values (written as $\sigma$). We write $\nu$ for record field:
%
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
              &\narrow{|}& \ident{float} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string}
 \\[0.6em]
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \hat{\sigma}\;\,\kvd{option} \lsep [\sigma] \\[0.1em]
              &\narrow{|}& \sigma_1 + \ldots + \sigma_n \lsep ~\top~ \lsep \kvd{null}
\end{array}
\end{equation*}
%
The non-nullable types include records (consisting of zero or more fields with their types) and
primitive types (\ident{int} for integers, \ident{float} for floating-point numbers, strings and
Booleans).

Any non-nullable type is also a nullable type, but it can be wrapped in the \kvd{option}
constructor to explicitly permit the \kvd{null} value. These are typically mapped to the standard
F\# option type. A simple collection type $[\sigma]$ is also nullable and missing values or
\kvd{null} are treated as empty collection. The type $\kvd{null}$ is inhabited by the $\kvd{null}$
value (using an overloaded but not ambiguous notation) and $\top$ represents the top type.

Finally, a union type in our model implicitly permits the \kvd{null} value. This is because the
type provided for unions requires the user to handle the situation when none of the case matches
(and so developers always provide code-path that can be run when the value is missing).

% --------------------------------------------------------------------------------------------------

\subsection{Subtyping relation}
\label{sec:inference-subtyping}

The subtyping relation between structural types is illustrated in Figure~\ref{fig:subtyping-diagram}.
We split the diagram in two parts. The upper part shows non-nullable types (with records and
primitive types). The lower part shows nullable types with \kvd{null}, collections and optional
values. We omit links between the two part, but any type $\hat{\sigma}$ is a subtype of
$\hat{\sigma}~\kvd{option}$ (in the diagram, we abbreviate $\sigma~\kvd{option}$ as $\sigma?$).
The following excerpt specifies some of the relationships formally:

\begin{definition}
$\sigma_1 :> \sigma_2$ denote that $\sigma_2$ is a subtype of $\sigma_1$. The
subtyping relation is defined as a transitive reflexive closure of:

\begin{align}
\ident{float}\,:>\,\ident{int}\\[-0.2em]
\sigma :> \kvd{null} \quad &(\textnormal{iff}~\sigma \neq \hat{\sigma}) \\[-0.2em]
\hat{\sigma}~\kvd{option} :> \hat{\sigma} \quad &(\textnormal{for all}~\hat{\sigma})
\end{align}
\begin{align}
 \{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} &:>  \{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
   \;\,\quad (\sigma_i \hspace{-0.25em}:>\hspace{-0.25em} \sigma_i') \\
 \{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} &:>  \{ \nu_1\!:\!\sigma_1, .., \nu_m\!:\!\sigma_m \}
   \quad (m \hspace{-0.1em}\geq\hspace{-0.1em} n) \\
 \{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} &:> \{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, .., \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
\end{align}
\vspace{-1.5em}
\begin{align}
\hspace{-0.6em}
\begin{array}{l}
 \{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null}, .., \nu_{n+m}\!:\!\kvd{null} \} :> \\
 \{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
\end{array}
\end{align}
\end{definition}

% --------------------------------------------------------------------------------------------------

\begin{figure}
\begin{center}
\includegraphics[scale=0.75,trim=5mm 5mm 5mm 5mm,clip]{images/hierarchy.pdf} % left bottom right top
\end{center}
\vspace{-0.5em}
\caption{Subtype relation between structural types}
\label{fig:subtyping-diagram}
\vspace{-0.5em}
\end{figure}

% -------------------------------------------------------------------------------------------------
\noindent
Here is a summary of the key aspects of the definition:

\begin{itemize}
\item For numeric types (1), we infer the most precise numeric type that can represent all values
  from a sample dataset (even though this means loss of precision in some cases).

\item The \kvd{null} type is a subtype of all nullable types (2), that is all
  $\sigma$ types excluding non-nullable types $\hat{\sigma}$). Any non-nullable type is also a
  subtype of its optional version (2).

\item The subtyping on records is covariant (4), subtype can have
  additional fields (5) and fields can  be reordered (6). The interesting rule
  is (7) -- together with covariance, it states that a subtype can omit some
  fields, provided that their types are nullable.
\end{itemize}

\noindent
The rule that allows subtype to have fewer record elements (8) is particularly
important. It allows us to prefer records in some cases. For example, given two samples
$\{ \ident{name}:\ident{string} \}$ and $\{ \ident{name}:\ident{string}, \ident{age}:\ident{int} \}$,
we can find a common supertype $\{ \ident{name}:\ident{string}, \ident{age}:\ident{int}~\kvd{option} \}$
which is also a record. For usability reasons, we prefer this to another common supertype
$\{ \ident{name}:\ident{string} \} + \{ \ident{name}:\ident{string}, \ident{age}:\ident{int} \}$.
Working with records does not require pattern matching and it makes it easier to explore data
in editors that provide auto-completion.

% -------------------------------------------------------------------------------------------------

\begin{figure*}[t]
\noindent
\begin{equation*}
\hspace{3em}
\inference[(record)\;]
  { (\nu_i = \nu'_j \Leftrightarrow (i = j) \wedge (i \leq k))
      \qquad \forall i\in\{ 1 .. k \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
  { \begin{array}{l}
    \{ \nu_1 : \sigma_1,  \; \ldots \;, \nu_k : \sigma_k, \; \ldots \;, \nu_n : \tau_n \} \tsep
    \{ \nu'_1 : \sigma'_1, \; \ldots \;, \nu'_k : \sigma'_k, \; \ldots \;, \nu'_m : \tau'_m \} \vdash\\
    \{ \nu_1 : \sigma''_1, \; \ldots \; , \nu_k : \sigma''_k,
                            \nu_{k+1} : \addopt{\sigma_{k+1}}, \ldots, \nu_n : \addopt{\sigma_n},
                            \nu'_{k+1} : \addopt{\sigma'_{k+1}}, \ldots, \nu'_m : \addopt{\sigma'_m} \}
    \end{array} }
\end{equation*}

\begin{equation*}
\hspace{3em}
% collection types
\inference[(list)\;]
  {\sigma_1 \tsep \sigma_2 \vdash \sigma}
  {[\sigma_1] \tsep [\sigma_2] \vdash [\sigma]}
\qquad
% primitive types
\inference[(prim)\;]
  { }
  {\ident{int} \tsep \ident{float} \vdash \ident{float}}\quad
\qquad
% null and nullable
 \begin{array}{l}
 \textnormal{\footnotesize{(null-1)}}\;\; \sigma \tsep \kvd{null} \vdash \sigma \quad(\sigma :> \kvd{null}) \\[0.6em]
 \textnormal{\footnotesize{(null-2)}}\;\; \sigma \tsep \kvd{null} \vdash \sigma~\kvd{option} \quad(\sigma :\ngtr \kvd{null})
 \end{array}
\end{equation*}

\caption{Selected inference judgements that define the common supertype relation}
\label{fig:subtyping-cst}
\end{figure*}

% --------------------------------------------------------------------------------------------------

\subsection{Common supertype relation}
\label{sec:inference-commonsuper}

As demonstrated by the example in the last section, structured types do not have a unique greatest
lower bound. Our inference algorithm prefers records over unions and is defined in terms of the
\emph{common supertype} relation. The type infernce obtains a type for each sample (or each element
of a collection) and then uses the relation to finds their common type. To demonstrate the idea,
we show the definition for primitive types and records:

\begin{definition}
A \emph{common supertype} of types $\sigma_1$ and $\sigma_2$ is a type $\sigma$, written
$\sigma_1 \triangledown \sigma_2 \vdash \sigma$, obtained according to the inference rules in
Figure~\ref{fig:subtyping-cst} (remaining rules can be found in the report \cite{fsharp-data-paper}).
\end{definition}

\noindent
When finding a common supertype of two records (\emph{record}), we return a record type that has the union
of fields of the two arguments. We assume that the names of the first $k$ fields are the same and the
remaining fields have different names (other rules permit reordering of fields). The types of shared
fields become common supertypes of their respective types (recursively). Fields that are present in
only one record are marked as optional using the following helper definition:

\begin{equation*}
\begin{array}{rcll}
 \addopt{\hat{\sigma}} &\narrow{=}& \hat{\sigma}~\kvd{option} \qquad&(\textnormal{non-nullable types})\\
 \addopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
\end{array}
\end{equation*}

\noindent
Discussing the full system is out of the scope of this paper, but the system has a number of desirable
properties -- the common supertype is uniquely defined and it is a supertype of both of the
provided examples. Furthermore (to aid usability) our algorithm finds a common supertype that is not
a union, if such type exists.

% =================================================================================================

\section{Results and contribution}

The contributions and results of the presented work fall in three categories. First, our work is
interesting in that it looks at a concrete type provider from the perspective of programming language
theory and we introduce a novel notion of \emph{relativized type safety}. Second, the F\# Data library
has become the most frequently downloaded F\# library and is de-facto standard for data access in F\#.
Third, the work is also interesting from philosophical perspective because it illustrates an important
change in thinking about types that we need to adopt when developing types for systems in the
modern age of the web.

\subsection{Relativized type safety}

The safety property of F\# Data type providers states that, given representative sample documents,
any code that can be written using the provided types is guaranteed to work. We call this
\emph{relativized safety}, because we cannot avoid \emph{all} errors. In particular, the user can
always use the type provider with an input that has a different structure than any of the samples --
and in this case, it is expected that the code will fail at runtime. Our formal model
\cite{fsharp-data-paper} states precisely what inputs can be handled by a provided type:
%
\begin{itemize}[noitemsep]
\item[--] Input can contain smaller numerical values (for example, if a sample contains float, the input can contain an integer).
\item[--] Records in the actual input can have additional fields.
\item[--] Records in the actual input can have fewer fields, provided that the type of the fields is marked as optional in the sample.
\item[--] Union types in the input can have both fewer or more cases.
\item[--] When we have a union type in the sample, the actual input can contain values of any of the union cases.
\end{itemize}
%
This is proved by our safety theorem, which relies on a simplified model of type providers and the
F\# language. We follow the approach of syntactic type safety \cite{syntactic} and show the type
preservation (reducing expression does not change type) and progress (an expression that is not a
value can be reduced).

\begin{theorem}[Relativized safety]
\label{thm:safety}
Assume $s_1, \ldots, s_n$ are samples, $\sigma$ is a common supertype of the types of $s_i$.
Next, assume that the type provider maps structural type $\sigma$ to an F\# type $\tau$ and that
it provides a function $f$ that takes an input $s$ and produces a value $\tau$.

Then for all new inputs $s'$ that are subtypes of one of the samples $s_i$, and for any expression
$e_c$ (user code) such that $x:\tau \vdash e_c:\tau'$, it is the case that
$e_c[x\leftarrow f(s')] \rightarrow^{*} v$ for value $v$ and $\vdash v:\tau'$.
\end{theorem}

\subsection{New perspective on static typing}

The \emph{relativized safety} property does not guarantee the same amount of safety as standard
type safety for programming languages without type providers. However, it reflects the reality of
programming with external data sources that is increasingly important in the age of the web.
In short, type providers do not \emph{reduce} the safety -- they simply \emph{reveal} an existing issue.

In a separate paper \cite{age-of-web}, we argue that this is an important futrue direction for
programming languages. Programs increasingly access external data sources (be it open government
data accessed by data journalists or services consumed by mobile applications). Most programming
language theory treats programs as closed expressions with no external dependencies -- this gives
us a formally tractable model, but it ``throws out the baby with the bath water''. Our experience
with F\# Data suggests that we can accept the reality of working with data \emph{and} still
provide clear formal model with provable safety properties.

\subsection{Practical experience}

The F\# Data library has become the standard library for accessing data in F\# and is widely used
by both open-source projects and commercial users of F\#. At the time of writing, it has 38
contributors and it has over 43,000 downloads\footnote{The first version (providing the functionality
discussed here) has been created by the author of this paper; further contributions added new features
and significantly improved the portability and overall quality of the library.
See \url{https://github.com/fsharp/FSharp.Data}.}.

The success of the library provides an evidence that some of the pragmatic (and perhaps
controversial) choices made in the F\# Data design work well in practice:

\begin{itemize}
\item The fact that a provided type may fail at run-time is not a problem as developers expect this
  when working with external data. In practice, the input that caused the issue can be added as
  another sample -- the new compilation errors show which part of the program need to be modified
  to handle the input correctly.

\item The F\# Data library has been designed to work well with modern F\# editors that use the
  F\# compiler to provide auto-completion based on static types\footnote{This includes Visual
  Studio, Emacs, Xamarin Studio, Atom and others.}. The success of F\# Data confirms that editting
  experience and interactive development are important aspect of modern programming environments.
\end{itemize}

% =================================================================================================

\section{Conclusions}

The F\# Data library presented in this paper simplifies working with structured data formats such
as XML, JSON and CSV. This is achieved by using the type provider mechanism to integrate external
structured data into the programming language. We briefly outlined the programming language theory
behind F\# Data as well as its practical usability properties that contributed to its adoption.

Accessing and processing data is becoming an important ability in the modern information-rich
society and we believe that tools such as F\# Data can make programming simpler, more usable
and more trustworthy -- and, for example, enable data journalists produce analyses that their
readers can not just understand, but also verify and modify.

\bibliographystyle{plain}
\bibliography{paper}

\end{document}

### 2015-04-13 icfp-2015/paper.tex
Add ICFP SRC finals, also add some forgotten changes

@@
 % TODO: The model here does not actually use type erasure!!
-% TODO: Maybe get rid of \mu_opt for records everywhere
 % TODO: Explain why we can only provide simple classes (!)
-% TODO: Replace \nu_opt with \nu_? or something else that does not look shit
 % TODO: Insert a few more mentions of why "s1 + s2 <: s1 + s2 + s3" is valid (cannot do complete pattern match)
@@
 
-%\titlebanner{banner above paper title}        % These are ignored unless
+\titlebanner{Unpublished draft, March 2015}        % These are ignored unless
 %\preprintfooter{short description of paper}   % 'preprint' option specified.
@@
 \begin{abstract}
-Most statically typed languages assume that programs run in a closed world, but this is not the 
+Most static type systems assume that a program runs in a closed world, but this is not the 
 case. Modern applications interact with external services and often access data in structured 
 formats such as XML, CSV and JSON. Static type systems do not understand such external data sources
-and only make data access more cumbersome. Should we just give up and leave the messy world of 
+and only make accessing data more cumbersome. Should we just give up and leave the messy world of 
 external data to dynamic typing and runtime checks?
@@
 The resulting library significantly reduces the amount of code that developers need to write when 
-accessing data. It also provides additional safety guarantees. Arguably, as much as possible 
+accessing data. It also provides additional safety guarantees. Arguably, as much safety as possible 
 if we abandon the incorrect closed world assumption.
@@
 
-The key function of many modern applications is to connect multiple data sources or services and
-present the aggregated data in some form. Mobile applications for taking notes, searching train
-schedules or finding tomorrow's weather all communicate with one or more services over the internet.
+The purpose of many modern applications is to connect multiple data sources or services and
+present the aggregated data in a new form. Mobile applications for taking notes, searching train
+schedules or finding tomorrow's weather all communicate with one or more services over the network.
 
 Increasing number of such services provide REST-based end-points that return data as CSV, XML
-or JSON. Despite numerous schematization efforts, the services typically do not come with schema.
-At best, the documentation provides a number of typical requests and sample responses.
+or JSON. Despite numerous schematization efforts, the services typically do not come with an explicit 
+schema. At best, the documentation provides a number of typical requests and sample responses.
 
@@
 Using a standard library for working with JSON and HTTP, we might call the service and 
-read the temperature as follows \footnote{We abbreviate the full URL: 
+read the temperature as follows\footnote{We abbreviate the full URL: 
 \url{http://api.openweathermap.org/data/2.5/weather?q=Prague\&units=metric}}:
@@
 root node must be a record with a \str{main} field, which has to be another record containing
-a \str{temp} field with numerical value. When the format is incorrect, the data access simply fails
+a \str{temp} field with a numerical value. When the format is incorrect, the data access simply fails
 with an exception.
@@
 
-In the rest of the paper, we give more detailed description of the mechanism and we also discuss
-theoretical safety properties of the approach. The key novel contributions of this paper are:
+In the rest of the paper, we give a detailed description of the mechanism and we also discuss
+safety properties of the approach. The key novel contributions of this paper are:
 
@@
 \noindent
-Although the F\# Data library \cite{fsharp-data} has been widely adopted is one of the most downloaded F\# libraries 
+Although the F\# Data library \cite{fsharp-data} has been widely adopted and is one of the most downloaded F\# libraries 
 it has not yet been presented in an academic form. Additional documentation and source code can 
@@
 \begin{verbatim}
-  [ { "name": null, "age": 25 },
+  [ { "name": "Jan", "age": 25 },
     { "name": "Alexander", "age": 3.5 },
@@
 %
-The code achieves the same simplicity as when using dynamically typed languages, but is statically 
+The code achieves the same simplicity as when using dynamically typed languages, but it is statically 
 type-checked. In contrast to the earlier example, we now use a local file \strf{people.json} as a 
@@
 so the file name has to be a constant. The provider analyzes the sample and generates
-a type that we name \ident{People}. In F\# editors that use the F\# Compiler Service, the 
+a type that we name \ident{People}. In editors that use the F\# Compiler Service, the 
 type provider is also executed at development-time and so the same provided types are
@@
  \kvd{type}~\ident{Entity}~=  \\[0.1em]
- \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{decimal}\rangl \\[0.1em]
- \quad \kvd{member}~\ident{Name}~:~\ident{option}\langl \ident{string}\rangl \\[0.6em]
+ \quad \kvd{member}~\ident{Name}~:~\ident{string} \\[0.1em]
+ \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{decimal}\rangl \\[0.6em]
  \kvd{type}~\ident{People}~=  \\[0.1em]
@@
 %
-The type \ident{Entity} represents the person. The property \ident{Name} is optional, because
-one of the sample records contains \kvd{null} as the name. The property \ident{Age} is also optional,
-because the value is missing in one sample. The two sample ages are an integer \num{25} and a 
-decimal \num{3.5} and so the common inferred type is \ident{decimal}.
+The type \ident{Entity} represents the person. The field \ident{Name} is available for all
+sample values and is inferred as \ident{string}. The field \ident{Age} is marked as optional,
+because the value is missing in one sample. The two sample ages are an integer $25$ and a 
+decimal $3.5$ and so the common inferred type is \ident{decimal}.
 
@@
 are subtypes of some of the provided samples (\emph{i.e.}~the samples are representative), then
-no exceptions will occur. In other words, we cannot avoid all failures, but we can prevent some --
-for example, if the OpenWeatherMap changes the response format, the sample in Section~\ref{sec:introduction}
+no exceptions will occur. In other words, we cannot avoid all failures, but we can prevent some.
+If the OpenWeatherMap changes the response format, the sample in Section~\ref{sec:introduction}
 will not re-compile and the user knows that the code needs to be changed. What this means for
@@
 \paragraph{Erasing type providers.}
-At runtime, the type providers we describe use an erasure mechanism similar to Java Generics \cite{java-erasure}. 
-A type provider also generates code that is executed in place of the members of the generated types.
+At compile-time, F\# type providers use an erasure mechanism similar to Java Generics \cite{java-erasure}. 
+A type provider generates types and code that should be executed when members are accessed. In compiled
+code, the types are erased and the program directly executes the generated code.
 In the above example, the compiled (and actually executed) code looks as follows:
@@
 as an array of strings to a typed representation -- here, a four-element tuple 
-$\ident{decimal}*\ident{option}\langl\ident{int}\rangl*\ident{string}*\ident{bool}$ (this type
-is also used as a type argument of $\ident{CsvFile}\langl\alpha\rangl$. The properties
+$\ident{decimal}*\ident{option}\langl\ident{int}\rangl*\ident{string}*\ident{bool}$. The properties
 such as \ident{Ozone} are then replaced with an accessor that reads the corresponding tuple
@@
 
-The type erasure mechanism is explained in our simplified formal model in Section~\ref{sec:formal}.
-More details about the full implementation in F\# can be found in the paper on F\# type 
-providers \cite{fsharp-typeprov}.
+The formal model in Section~\ref{sec:formal} does not discuss the erasure mechanism and instead 
+models evaluation of the provided types. We do not lose anything here -- the erasure is mainly
+useful when the provided types are very large (and cannot be efficiently generated all at once).
+For more, see the F\# type providers report \cite{fsharp-typeprov}.
 
@@
 
-The XML format is similar to JSON in that it uses nested elements rather than a flat CSV-like 
-structure. In our inference algorithm, we use the notion of \emph{records} for both JSON records 
-and XML elements. However, unlike in JSON, the XML elements also have a name. An element can 
-contain a collection of child elements. In case of XML, the collection can often be heterogeneous 
-containing child nodes of different names. Consider the following (simplified) RSS news feed as an example:
+The XML format is similar to JSON in that it uses nested elements rather than a flat 
+structure. Our inference algorithm uses the notion of \emph{records} for both JSON records 
+and XML elements. However, unlike in JSON, the XML elements also have a name.
+
+An XML element can contain a collection of child elements. The collection is often heterogeneous 
+containing child nodes of different names. Consider the following (simplified) RSS feed:
 %
@@
 
-When extracting data from such XML documents, we often need to get nodes of a specific name (all
+When extracting data from XML, we often need to get nodes of a specific name (all
 items) or extract the content of a node (feed title). The XML type provider is optimized for 
-this style of access and allows processing the feed as follows:
+this style of access:
 
@@
 \paragraph{Heterogeneous collections and unions.}
-Our inference algorithms recognizes this as a \emph{heterogeneous collection} (if the same structure 
+Our inference algorithms recognizes the \ident{channel} element as a \emph{heterogeneous collection} (if the same structure 
 appears in all samples) and generates a type with a member \ident{Title} for accessing the value 
@@
 %
-The \ident{PersonOrString} type contains one property for each of the union case. In the body, they
+The \ident{PersonOrString} type contains one member for each of the union case. In the body, they
 check that the value is not \kvd{null} and that it has the right structure (using the \ident{isTag}

### 2015-05-01 ml-2014-paper/paper.tex
Fix typos (ML 2014 post)

@@
 This matters because the web changes many assumptions that typed functional language designers take
-for granted. For example, programs do not run in a closed word, but must instead interact with 
+for granted. For example, programs do not run in a closed world, but must instead interact with
 (changing and likely unreliable) services and data sources, communication is often asynchronous
@@
 % What is my solution
-In this paper, we present how F\# language and libraries face the challenges posed by the web.
+In this paper, we present how the F\# language and libraries face the challenges posed by the web.
 Technically, this comprises using \emph{type providers} for integration with external information
@@
 In this inquiry, the holistic perspective is more important than each of the features in isolation.
-We use a practical case study as a starting point and look how F\# language and libraries approach 
-the challenges posed by the web. The specific lessons learned are perhaps less interesting than 
-our attempt to uncover hidden assumptions that no longer hold in the age of web.
+We use a practical case study as a starting point and look at how F\# language and libraries
+approach the challenges posed by the web. The specific lessons learned are perhaps less interesting
+than our attempt to uncover hidden assumptions that no longer hold in the age of web.
 \end{abstract}
@@
 
-We do not claim that the F\# approach is the only possible. Rather, this paper should be seen
+We do not claim that the F\# approach is the only possible one. Rather, this paper should be seen
 as a programming language experiment \cite{philosophy-pl} or an empirical observation of the
@@
   F\# language features can be used for the development of modern web applications. This
-  is not a toy demonstration, but an example of how F\# is used in the industry.
+  is not a toy demonstration, but an example of how F\# is used in industry.
 
@@
 \item Throughout the paper, we discuss how the age of the web breaks the assumptions commonly
-  taken as granted in typed functional programming. We revisit the notion of type safety
-  in the context of web (Section~\ref{sec:tp-relative}) and the notion of fixed language 
+  taken for granted in typed functional programming. We revisit the notion of type safety
+  in the context of the web (Section~\ref{sec:tp-relative}) and the notion of fixed language
   semantics (Section~\ref{sec:js-relative}).
@@
 de-sugaring similar to the CPS transformation and interprets keywords such as \kvd{let!} and
-\kvd{for} using special operations (monadic bind and other). The \ident{async} identifier 
+\kvd{for} using special operations (monadic bind and others). The \ident{async} identifier
 determines that we are writing asynchronous workflow \cite{fsharp-async} that makes it
@@
 (Section~\ref{sec:js-lib}). Finally, the last part of the example code registers event handlers that
-redraw the chart when checkbox is clicked:
+redraw the chart when the checkbox is clicked:
 %
@@
 The aim of this paper is not to make value judgements and argue what is better. Using the case
-study as a basis, we claim that the outlined approach is \emph{one possible} and that it
+study as a basis, we claim that the outlined approach is just \emph{one possible} and that it
 \emph{works in practice}. In the rest of the paper, we give more details about the most important
@@
 
-Type providers \cite{fsharp-typeprov} is a mechanism for integrating external components into a 
+Type providers \cite{fsharp-typeprov} are a mechanism for integrating external components into a
 statically typed programming language. Such components include information sources (such as World
@@
 \item \emph{Type provider failure.}
-  The projection (a type provider) is implemented in F\# and can fail (for example, if internet
-  is required but unavailable). However, when it succeeds, it generates a valid F\# code and so 
+  The projection (a type provider) is implemented in F\# and can fail (for example, if the internet
+  is required but unavailable). However, when it succeeds, it generates valid F\# code and so
   unchecked compilation errors cannot happen.
@@
 been designed specifically for World Bank data. This contrasts with F\# Data type providers for
-working with CSV, XML and JSON data that take sample or schema as parameter and can thus be used 
+working with CSV, XML and JSON data that take a sample or schema as a parameter and can thus be used
 with files of any structure.
@@
   not be able to obtain the list of countries or indicators. This is an example of \emph{type provider
-  failure}. In case of World Bank, this is partly prevented by caching of the schema, but internet
+  failure}. In case of the World Bank, this is partly prevented by caching of the schema, but internet
   access is required when using the provider for the first time.
@@
 \noindent
-When contrasted with the traditional view of safety in ML family of languages, the World Bank 
+When contrasted with the traditional view of safety in the ML family of languages, the World Bank
 type provider certainly relaxes the usual safety conditions. However, this is not the case when
@@
 \vspace{-1em}
-\paragraph{Accessing countries without type providers.} First, consider how would we write the code to obtain a list
+\paragraph{Accessing countries without type providers.} First, consider how we would write the code to obtain a list
 of countries without using the World Bank type provider. Assuming we want only the countries and
@@
 The \emph{recompilation failure} when a country is removed from the dataset is an indication that
-code would not behave as expected as runtime and so we find it useful. The problematic case is 
+code would not behave as expected at runtime and so we find it useful. The problematic case is
 a \emph{runtime failure} when a country is removed after the program is compiled. One way to
@@
 $\ident{Asynchronous}=\kvd{true}$). For example, the CSV type provider from F\# Data has a
-parameter \ident{AssumeMissingValues} that serves exactly this purpose.
+parameter \ident{AssumeMissingValues} that instructs the type provider to provide option values,
+even if the sample does not contain them.
 
@@
 \emph{relativized} notion of type safety. A full formalization is beyond the scope of this
-paper, but the following provides an outline of such theorem.
+paper, but the following provides an outline of such a theorem.
 
@@
 
-JavaScript has become the \emph{lingua franca} of the web and increasing number of programming 
+JavaScript has become the \emph{lingua franca} of the web and an increasing number of programming
 languages provide a way of compiling to JavaScript. In F\#, the first project was F\# WebTools
@@
 (Section~\ref{sec:js-meta}), there are different approaches to base libraries (Section~\ref{sec:js-lib}),
-asynchronous computing (Section~\ref{sec:js-async}) and semantic mismatch between the host language
-and JavaScript (Section~\ref{sec:js-relative}).
+asynchronous computing (Section~\ref{sec:js-async}) and the semantic mismatch between the host
+language and JavaScript (Section~\ref{sec:js-relative}).
 
@@
 methods in the marked module as \emph{quotations} that can be retrieved at run-time. Note that this does
-not store the AST for the entire file -- for example, information about declared types need to be 
+not store the AST for the entire file -- for example, information about declared types needs to be
 obtained separately using the .NET reflection mechanism.
@@
 compiler back-end is a better choice. However, for the task solved in our case study (compiling
-newly written F\# code), the approach works well. A related question is providing mapping for 
+newly written F\# code), the approach works well. A related question is providing mappings for
 standard (in our case, F\# and .NET) libraries used by the code. We discuss this in the next section.
@@
 uses .NET \ident{IEnumerable} interface) and JavaScript libraries. Accessing these libraries in
-FunScript follows similar approach to F\# WebTools \cite{fsharp-webtools}.
+FunScript follows a similar approach to F\# WebTools \cite{fsharp-webtools}.
 
@@
 JavaScript (collections, string manipulation, date and other). Those are either reimplemented in
-F\# with \ident{ReflectedDefinition} attribute, or mapped to standard JavaScript library.
+F\# with the \ident{ReflectedDefinition} attribute, or mapped to standard JavaScript library.
 
@@
 makes writing and reusing code easier and lets developers use existing familiar libraries and
-idioms. However, it means that semantics of F\# running as JavaScript is not the same as the
+idioms. However, it means that the semantics of F\# running as JavaScript is not the same as the
 semantics of F\# running as compiled code. We return to this topic in Section~\ref{sec:js-relative}.
@@
 
-In F\# asynchronous workflows \cite{fsharp-async} serve two purposes. First, they provide a
+In F\#, asynchronous workflows \cite{fsharp-async} serve two purposes. First, they provide a
 way for running multiple tasks in parallel. Second, they make it possible to write long-running
@@
 starts the work and blocks until it completes; \ident{Async.Start} starts the work in the background
-and \ident{Async.\-Start\-Immediate} starts the work on the current thread and runs callbacks on 
+and \ident{Async.\-Start\-Immediate} starts the work on the current thread and runs callbacks
 in the same synchronization context (essentially on the same thread). However, only the last one
@@
 
-There is a number of projects that solve similar problems to the ones described in this paper.
+There are a number of projects that solve similar problems to the ones described in this paper.
 In the F\# ecosystem, the first project compiling to JavaScript was F\# WebTools \cite{fsharp-webtools},
@@
 and interop with JavaScript. More recently, WebSharper \cite{websharper-piglets,websharper-guis}
-is a more complete framework for web development that also provide composable abstractions for
+is a more complete framework for web development that also provides composable abstractions for
 building forms (based on formlets \cite{links-formlets}) and entire applications.
@@
 of this paper is not in any single technology it presents, but in the combination it shows.
-For this reason, we used the form of a cast study, using a significant example to guide our
+For this reason, we used the form of a case study, using a significant example to guide our
 discussion.

### 2015-10-25 ml-2014-paper/paper.tex
Camera ready version of ML2014 post-proceedings

@@
 \email{don.syme@microsoft.com}
+\and
+Zach Bray
+\institute{Type Inferred Ltd}
+\email{zachbray@gmail.com}
 }
 \def\titlerunning{In the Age of Web: Typed Functional-First Programming Revisited}
-\def\authorrunning{T. Petricek, D. Syme}
+\def\authorrunning{T. Petricek, D. Syme, Z. Bray}
 \begin{document}
@@
 those are treated as dealing with the ``dirty real world''. For a practical solution, we argue that
-we need to go much further, but deeper integration with (untyped) JavaScript libraries and (evolving)
-services inevitably breaks strict type safety requirements.
+we need to go much further -- to the extent that deeper integration with (untyped) JavaScript
+libraries and (evolving) services inevitably breaks some of the strict type safety requirements.
 
@@
 
-Note that we use a standard F\# function \ident{List.map} to iterate over the countries. This has
-a side-effect of creating the HTML elements, but it also returns a new list. The result is a list
-of $\ident{string}\ast\ident{Indicators}\ast\ident{jQuery}$ values representing the country name,
-its indicators (for accessing the World Bank data) and the created DOM object representing the
-checkbox.
+Note that we use a standard F\# function \ident{List.map} to iterate over the countries. The
+function passed as an argument has a side-effect of creating the HTML elements, but it also returns
+a new list. The result is a list of $\ident{string}\ast\ident{Indicators}\ast\ident{jQuery}$ values
+representing the country name, its indicators (for accessing the World Bank data) and the created
+DOM object representing the checkbox.
 
@@
 study as a basis, we claim that the outlined approach is just \emph{one possible} and that it
-\emph{works in practice}. In the rest of the paper, we give more details about the most important
+\emph{works in practice}. The rest of the paper gives more details about the most important
 aspects of the approach and discuss alternatives.
@@
 
-In technical terms, type providers are libraries that are loaded by the compiler (and editor)
-and are executed at compile-time (or edit-time). A type provider builds information about types
-it provides and makes those available to the compiler. It is worth noting that this is done lazily,
-so the type provider does not need to provide types for the entire information space at once.
-
-In terms of programming language theory, type providers can be easily explained anecdotally.
-They change the starting point for a type-checking of a program as follows:
+Type providers are libraries that are loaded by the compiler (and editor) and are executed at
+compile-time (or edit-time). A type provider builds information about types and makes those
+available to the compiler. This is done lazily, so the type provider does not need to provide
+types for the entire information space at once. In terms of programming language theory, type
+providers change the starting point for a type-checking of a program as follows:
 %
@@
 Rather than starting the type-checking with an empty context, we start type-checking with a
-context containing a \emph{projection} of some information from the world. In this paper, we do not
-attempt to formalize type providers, so we avoid the question of what, formally, is the world here.
-However, this analogy is useful for understanding how type providers work and how they can go wrong:
+context containing a \emph{projection} of some information from the world. This analogy is useful
+for understanding how type providers work and how they can go wrong:
 
@@
   The projection (a type provider) is implemented in F\# and can fail (for example, if the internet
-  is required but unavailable). However, when it succeeds, it generates valid F\# code and so
-  unchecked compilation errors cannot happen.
+  is required but unavailable). However, when the type provider succeeds, it generates valid F\#
+  code and so unchecked compilation errors cannot happen.
 
@@
   negative aspect, but there is a clear benefit -- it means we discover an error that would happen
-  at runtime earlier, that is during recompilation.
+  at runtime earlier during recompilation.
 \end{itemize}
@@
 \noindent
-The above points are \emph{general} points about type providers. In this paper, we are more interested
-in the \emph{specific} type providers used in the case study. We look at the World Bank and TypeScript
-type providers in the following two sections. When discussing type providers, we need to look at
-how the provider (or the projection $\pi$) works and what assumptions it makes about the world.
-That is, what world changes cause the provided type to change and what world changes cause a
-runtime exception.
+The above points are \emph{general} points about type providers. In this paper, we look at two
+\emph{specific} type providers for World Bank and TypeScript. When discussing type providers, we
+look at how the provider (or the projection $\pi$) works and what assumptions it makes
+about the world. That is, what world changes cause the provided type to change and what world
+changes cause a runtime exception.
 
@@
 
-The World Bank provider \cite{fsharp-data} is a type provider for a specific data source -- it has
-been designed specifically for World Bank data. This contrasts with F\# Data type providers for
-working with CSV, XML and JSON data that take a sample or schema as a parameter and can thus be used
-with files of any structure.
-
-\subsubsection{The World Bank type provider}
+The World Bank provider \cite{fsharp-data} is a type provider designed specifically for a single
+data source. This contrasts with F\# Data type providers for working with CSV, XML and JSON data
+that take a sample or schema as a parameter and can thus be used with files of any structure.
 
 The projection implemented by the type provider generates a type called \ident{Countries} that contains
-countries (and regions) as members. Each member returns a value of type \ident{Indicators} that is
-also generated and contains all indicators as members. Note that the type provider generates just
-these two types -- in particular, it does not generate a new type for each country. This would be an
-interesting option if we could check what indicators are available for a given country.
-
-The type provider builds the structure of the provided types, but the provided types are \emph{erased}
-during compilation and replaced with runtime implementation. For the World Bank type provider, the
-erasure works as follows:
+countries as members. Each member returns a value of type \ident{Indicators} that is
+also generated and contains all indicators as members. The type provider provides types that are
+\emph{erased} during compilation and replaced with runtime implementation. For the World Bank type
+provider, the erasure works as follows:
 %
@@
 a typical pattern -- type providers generally produce a light layer on top of a rich runtime library.
-
 An important point here is that the type provider uses the \emph{name} of a country or an indicator
 for the member name, but at runtime, it uses a \emph{code} of the country or the indicator. This has
-interesting implications for the safety properties, as discussed in the next section.
+interesting implications for the safety properties.
 
@@
 
-Now that we know how the World Bank type provider generates the types and the runtime code, we can
-look at the assumptions it makes about the world and in which cases it encounters one of the failures
-defined in Section~\ref{sec:tp-def}.
+Given the mapping, we can now look at the assumptions that the World Bank type provider makes about
+the world and in which cases it encounters one of the failures defined in Section~\ref{sec:tp-def}.
 
@@
 \item If you are offline when using the type provider (compiling or editing), the type provider will
-  not be able to obtain the list of countries or indicators. This is an example of \emph{type provider
-  failure}. In case of the World Bank, this is partly prevented by caching of the schema, but internet
-  access is required when using the provider for the first time.
-
-\item An interesting case is when a country or indicator is renamed. In that case, compiled code
-  using the World Bank type provider will continue to work (there is no \emph{runtime failure}),
-  because the lookup uses the country or the indicator code. However, recompilation will fail
-  as the member name (which appears in the source code) will be different.
-
-\item Finally, if a country or indicator completely disappears, we will get \emph{runtime failure}
-  when running existing compiled code and \emph{recompilation failure}. here, the \emph{recompilation
+  not be able to obtain the list of countries or indicators. This is a \emph{type provider failure}.
+  The World Bank provider caches the schema, but it still requires internet access for the first time.
+
+\item When a country or indicator is renamed, the compiled code will continue to work (there is no
+  \emph{runtime failure}), because the lookup uses the country or the indicator \emph{code}. However,
+  recompilation will fail as the member name (which appears in the source code) will be different.
+
+\item Finally, if a country or indicator disappears, we will get \emph{runtime failure} when running
+  existing compiled code and \emph{recompilation failure}. Here, the \emph{recompilation
   failure} is a useful indicator that the assumptions made by our code have been violated by
@@
 \noindent
-When contrasted with the traditional view of safety in the ML family of languages, the World Bank
-type provider certainly relaxes the usual safety conditions. However, this is not the case when
-compared with standard developer practices. In particular, if we wrote
-$\ident{data.GetCountries}().\ident{GetCountry}(\str{CZE})$ by hand, the \emph{runtime failure}
-behaviour would be the same, but we would lose the useful \emph{recompilation failure}. The following
-section discusses this (and other) alternatives.
+When contrasted with the traditional view of ML type-safety, the World Bank type provider relaxes
+the usual safety conditions. However, this is not the case when compared with standard developer
+practices. In particular, if we wrote $\ident{data.GetCountries}().\ident{GetCountry}(\str{CZE})$,
+the \emph{runtime failure} behaviour would be the same, but we would lose the useful
+\emph{recompilation failure}.
 
@@
 
-There are two main alternatives that are worth discussing. First, how would the code look if
+There are two main alternatives worth discussing. First, how would the code look if
 we did not use any type provider. Second, what are alternative designs for the World Bank type
@@
 \vspace{-1em}
-\paragraph{Accessing countries without type providers.} First, consider how we would write the code to obtain a list
-of countries without using the World Bank type provider. Assuming we want only the countries and
-regions in Figure~\ref{fig:wb}, we could specify the country codes and then obtain the country objects:
+\paragraph{Accessing countries without type providers.} Assuming we want the countries
+and regions in Figure~\ref{fig:wb}, we could build a list of country codes and use projection:
 %
@@
  \kvd{let}~\ident{countryCodes}~=~\lbrack~\str{EU},~\str{CZE},~\str{GBP},~\str{USA}~\rbrack \\
- \kvd{let}~\ident{countries}~=~\ident{countryCodes}~|\rangl~\ident{List.map}~(\ident{data.GetCountries}().\ident{GetCountry})
+ \kvd{let}~\ident{countries}~=~\ident{countryCodes}~|\rangl~\ident{List.map}~(\ident{data.GetCountries}().\ident{GetCountry}())
 \end{array}
@@
 The motivation for the case study was a data journalism application with a transparent logic
-that is easy to understand (and perhaps even modify). The type provider certainly achieves these
-goals better -- a country can be added just by looking through an auto-complete list and we do not
-need to know its code or exact name. However, if we wanted to list all countries in a region, then
-the World Bank type provider would not be helpful.
-
-As for the safety properties, as discussed earlier, using explicit country codes has the same
-runtime properties as using the type provider. This could be avoided by listing all countries in
-a given region (or, according to another rule), but that is solving a different problem.
+that is easy to understand and modify. The type provider achieves these goals better -- a country
+can be added just by looking through an auto-complete list without knowing its code.
+As for the safety properties, using explicit country codes has the same runtime properties as using
+the type provider. This could be avoided by listing all countries in a given region, but that is
+solving a different problem.
 
@@
 an information source. With rising popularity of data journalism and Open Government Data
-initiatives, we argue that this is an important problem domain. Here, we first specify the country
-codes and then obtain the country objects.
+initiatives, we argue that this is an important problem domain.
 
@@
 author of the application does not intend to handle errors (for example, by skipping missing data),
-they would have to write the following:
-%
-\begin{equation*}
-\begin{array}{l}
- \kvd{let}~\ident{optCountries}~= \\
- \quad~\lbrack~~ \ident{data.Countries.}\lident{European Union} \\
- \qquad   \ident{data.Countries.}\lident{Czech Republic} \\
- \qquad   \ident{data.Countries.}\lident{United Kingdom} \\
- \qquad   \ident{data.Countries.}\lident{United States} ~\rbrack
-\\[0.5em]
- \kvd{let}~\ident{countries}~=\\
-  \quad \ident{optCountries}~|\rangl~\ident{List.map}~(\kvd{fun}~\ident{countryOpt}~\rightarrow \\
-  \qquad \kvd{match}~\ident{countryOpt}~\kvd{with}~\ident{Some~c}~\rightarrow~\ident{c}~|~\ignp~\rightarrow
-    \ident{failwith}~\str{Country does not exist!})
-\end{array}
-\end{equation*}
-%
-For the purpose of our case study, providing option types would only make the code more complex.
-However, if we wanted to handle errors more gracefully, then option types would be a better
-alternative. Note, however, that we can still handle exceptions using $\kvd{try}~\ldots~\kvd{with}$.
-
-Whether using option types is more desirable or not often depends on the use case (how we
-want to handle errors) and reliability of the data source (countries do not disappear often).
-This alternative can be supported by adding a static parameter to the type provider (like
-$\ident{Asynchronous}=\kvd{true}$). For example, the CSV type provider from F\# Data has a
-parameter \ident{AssumeMissingValues} that instructs the type provider to provide option values,
-even if the sample does not contain them.
+they would have to add code that explicitly throws an exception. For the purpose of our case study,
+we prefer to keep the end-user code simple -- we can still handle the error using
+$\kvd{try}~\ldots~\kvd{with}$.
+
+Whether using option types is more desirable depends on the use case (how we want to handle errors)
+and reliability of the data source (countries do not disappear often). The choice can be left to
+the user by adding a static parameter (like $\ident{Asynchronous}=\kvd{true}$). For example, the CSV
+type provider from F\# Data has a parameter \ident{AssumeMissingValues} that instructs the type
+provider to always provide options.
 
@@
 \paragraph{Accessing data ``as of time''.}
-So far, we discussed how to mitigate the problems caused by the fact that the open world changes.
+We discussed how to mitigate the problems caused by the fact that the open world changes.
 Can we make the world \emph{not} change? This is a sensible question for some data sources
-(although not for the World Bank). For example Freebase \cite{google-freebase} supports the
-notion of ``as of time''. When calling Freebase with the \ident{as\_of\_time} parameter, the API
-returns data and meta-data that was present at the specified date.
+(but not the World Bank). Freebase \cite{google-freebase} supports the notion of ``as of time''.
+When calling the API with the \ident{as\_of\_time} parameter it returns data and meta-data that
+was present at the specified date.
 
@@
 by an article, the accompanying application could use data available at the time when the
-article was written. However, this feature relies on the ability of the data source (not all
-services have this), or on the possibility to create local snapshots (difficult for large data
-sources).
+article was written. However, this feature relies on the ability of the data source -- not all
+services have this option.
 
@@
 
-\subsubsection{The TypeScript type provider}
-The projection implemented by the TypeScript type provider imports interface definitions in the
-\str{.d.ts} format specified by TypeScript \cite{ms-typescript}. This is a format of type
-annotations for JavaScript libraries. The following example is an excerpt from the \str{jquery.d.ts}
-file that describes the \ident{jQuery} value and the \ident{attr} method used in our example:
+The projection implemented by the TypeScript type provider imports type annotations for JavaScript
+libraries sa specified by TypeScript \cite{ms-typescript}. The following is an excerpt from the
+\str{jquery.d.ts} file that describes the \ident{jQuery} value and the \ident{attr} method used
+in our example:
 %
@@
 \}
-\\[0.5em]
+\\
 \kvd{interface}~\ident{JQuery}~\{ \\
@@
 %
-The example demonstrates a number of typical problems that arise when using type providers for
-language interoperation. Not all TypeScript language constructs have direct equivalent in F\# and
+The example demonstrates typical problems that arise when using type providers for language
+interoperation. Not all TypeScript language constructs have direct equivalent in F\# and
 the TypeScript type provider needs to map them to other constructs that are available:
@@
 \begin{itemize}
-\item The interface file defines \ident{jQuery} as a global variable. Although F\# supports
-  global bindings, those cannot be provided by a type provider and so the global variable
-  is exposed as a static member of the imported type, so we use it via \ident{j.jQuery}.
+\item The interface file defines \ident{jQuery} as a global variable. F\# type providers cannot
+  provide global bindings and so the variable is exposed as a static member of the imported type
+  (\emph{i.e.}~\ident{j.jQuery}).
 
@@
 It is also worth noting that the \ident{attr} member is overloaded and uses an optional parameter.
-However, both of these features are available in F\# and can be directly mapped. Now that we
-covered how the type structure is generated, we also need to show what code is generated when
-the types are erased:
-%
+However, both of these features are available in F\# and can be directly mapped. The following
+snippet shows the runtime code that is generated when the types are erased:
+
 \begin{equation*}
@@
 %
-An important point here is that the above provided code is never actually \emph{executed}.
-As discussed in Section~\ref{sec:js-meta}, it is translated to JavaScript (which then runs
-in the web browser). So, the above code can be seen more as instructions for the translator.
-The mapping is quite simple -- member calls are translated to \ident{Emit.CallImpl} (empty
-name denotes that the object itself is called) and property getters are translated to
-\ident{Emit.PropertyGetImpl}. In both cases the first argument denotes whether the call is
-static and the last argument is an array of arguments.
+The provided code is never actually \emph{executed}. As discussed in Section~\ref{sec:js-meta}, it
+is translated to JavaScript (which then runs in the web browser). So, the above code can be seen
+more as instructions for the translator. The mapping is straightforward -- member calls are
+translated to \ident{Emit.CallImpl} (empty name denotes that the object itself is called) and
+property getters are translated to \ident{Emit.PropertyGetImpl}. The first argument
+denotes whether the call is static and the last argument is an array of arguments.
 
 \subsubsection{Safety of cross-language type providers}
-As mentioned earlier, TypeScript does not have the type soundness property \cite{ms-safets},
-because of covariant generics. This illustrates two points. First, this is a property of
-\emph{running} TypeScript code, which is not the case here -- we use TypeScript only as a
-source of annotations for JavaScript libraries. There is no guarantee that the library will
-actually adhere to the specification, which is an issue that we return to when discussing
-translation to JavaScript (Section~\ref{sec:js-meta}).
-
-The second point is that we can not import the typing rules associated with TypeScript.
+TypeScript does not have the traditional type soundness property \cite{ms-safets},
+because of covariant generics. However, this is a property of \emph{running} TypeScript code, while
+we use TypeScript only as a source of annotations for JavaScript libraries. Still, there is only a
+weak guarantee that the library will adhere to the specification. We return to this issue
+when discussing translation to JavaScript (Section~\ref{sec:js-meta}).
+
+More importantly, we do not import the typing rules associated with TypeScript.
 The type provider mechanism can only provide F\# types that then behave according to the
@@
 The above discussion highlights a broader point about using type providers for language
-interoperability. Given any language, its type system be weaker or stricter in some ways.
-Both cases make mapping difficult:
+interoperability. For any language, its type system may be weaker or stricter in some ways:
 
 \begin{itemize}
-\item First, consider a weaker system with less type information. The type provider may need
-  to map more types to \ident{object} (or similar general type). However, this will make
-  the provided operations hard to use (because the source language is more flexible than
-  F\#). An alternative is to require explicit annotations (like \textcolor{strclr}{\ttfamily d.ts}
-  files for JavaScript).
-
-\item Second, if the imported language has a more precise type system than F\#, the type
+\item When interoperating with a weaker system, the type provider may need
+  to map more types to a general type like \ident{object}. This makes the provided operations hard
+  to use (as the source language is more flexible than F\#). An alternative is to use explicit
+  annotations (like \textcolor{strclr}{\ttfamily d.ts} files for JavaScript).
+
+\item If the imported language has a more precise type system than F\#, the type
   provider has to drop some of the information. This can be done safely for types in
@@
 \noindent
-In case of JavaScript, we encounter both cases. In many cases, the return type is not
+In case of TypeScript, we encounter both cases. In many cases, the return type is not
 statically known and is exposed as \kvd{any}, which we then map to F\# \ident{object}
@@
 form of dependent typing (overloading on constants \cite{ms-typescript-09}). This cannot
-be expressed in the F\# type system and the developer has to do more work (for example,
-choose the right overload or provide an explicit type annotation).
-
-An interesting example of type provider for a language with a weaker type system is the
-R provider \cite{fsharp-rprovider}. It solves the mismatch by using runtime reflection
-to discover R functions and their declared parameters. However, R provider also generates
-additional function overloads that let the user specify additional parameters (not explicitly
-declared in the signature). To make such interoperability easier, we proposed additional
-escape mechanism \cite{fsharp-paramsdict} that provide usable syntax when static type information
-cannot be fully recovered.
+be expressed in the F\# type system and the developer has to choose the right provided overload.
 
@@
 
-\vspace{-1em}
 \paragraph{Embedding inline JavaScript.}
-To make the interoperability mechanism general, the host language (here, F\#) should not know about
-the syntax of the embedded language (here, JavaScript). This means that the JavaScript code can
-only be embedded as strings. FunScript provides this option and it is used for accessing various
-primitive JavaScript operations. For example, the following is one way to define a function that
-converts a value to a number using JavaScript semantics:
+In general, the host language should not know about the sytnax of the embedded language. This can
+be hanlded via quasi-quotations \cite{haskell-templ}, or by embedding JavaScript as strings (perhaps
+with syntax checking at a later stage). In FunScript, the \ident{number} function is defined as:
 %
@@
 %
-The example uses .NET attributes (on the first line), which are meta-data attached to a function.
-When FunScript finds a call to such annotated function, it replaces the call with the specified
-JavaScript. The string \textcolor{strclr}{\ttfamily \{0\}} is a placeholder for the argument. Note
-that the body of the function is never actually executed and so we use a placeholder.
-
-The approach provides no guarantees about the inline JavaScript code. At compile-time, it cannot
-even check that the JavaScript code is syntactically correct. This could be improved using a
+The example uses .NET attributes, which are meta-data attached to a function. When FunScript finds
+a call to such annotated function, it replaces the call with the specified JavaScript.
+The approach provides no guarantees about the inline JavaScript code. In practice, this feature
+is used only for minimal JavaScript runtime, but the approach could be improved using a
 parameterized type provider such as $\ident{JSCode}\langl\str{return \{0\}*1.0;}\rangl$, which
@@
 Many compilers to JavaScript provide a way for writing type signatures for JavaScript
-libraries in the host language. For example, the following snippet shows a declaration of the
+libraries in the host language. For example, the following shows a declaration of the
 jQuery \ident{attr} method from a mapping for js\_of\_ocaml \cite{js_ocmal} (definitions in
-SMLtoJs \cite{js_sml} are similar, but simpler):
+SMLtoJs \cite{js_sml} are simpler):
 %
@@
 type denotes a JavaScript method; \ident{optdef} specifies that the result may be
-undefined and \ident{js\_string} denotes a JavaScript string (which is distinct from OCaml
-strings). Also, note that the second method is called \ident{attr\_set}. This is a simple
-naming trick -- OCaml does not support overloading and js\_of\_ocaml simply ignores anything
-after underscore.
+undefined and \ident{js\_string} denotes a JavaScript string. Also, note that the second method
+is called \ident{attr\_set}. This is a simple naming trick -- OCaml does not support overloading
+and js\_of\_ocaml simply ignores anything after underscore.
 
@@
 \textcolor{strclr}{\ttfamily d.ts} files into the above OCaml annotation. The main difference
-is that type providers provide a smoother integration. Type providers do not produce any artefacts
+is that type providers do not produce any artefacts
 (generated files) and can import entire repositories lazily.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Relativized type safety}
-\label{sec:tp-relative}
-
-Type providers for information sources, such as the World Bank provider, weaken the usual notion
-of type safety. In general, they require that certain assumptions about the world do not change
-between the state of the world at compile-time and state of the world at run-time. The TypeScript
-provider is different and it is more interesting to consider runtime behaviour as discussed
-in Section~\ref{sec:conc-interop}.
-
-For type providers that provide access to external information sources, we can formulate a
-\emph{relativized} notion of type safety. A full formalization is beyond the scope of this
-paper, but the following provides an outline of such a theorem.
-
-\begin{theorem*}[Relativized type safety]
-Assume $\pi_{\ident{~WorldBank}}$ is the mapping implemented by the World Bank type provider,
-$w, w_0$ are models of a (read-only) world that can be seen as functions defined on country-indicator
-pairs and $\langl e, w \rangl \rightarrow e'$ is a one-step reduction on expressions that has
-access to the World Bank data modelled by $w$. Then we define:
-
-\begin{itemize}
-\item
-Multi-step reduction $\langl e, w \rangl \rightarrow^* e'$ if either $e=e'$
-or $\langl e, w \rangl \rightarrow e''$ and $\langl e'', w \rangl \rightarrow^* e'$.
-\item
-Functions $\ident{countries}(e)$ and $\ident{indicators}(e)$ that return sets of
-country names \\and indicator names that appear in an expression $e$.
-\end{itemize}
-%
-Then the following \emph{relativized type safety} implication holds:
-%
-\begin{equation*}
-\begin{array}{lcl}
-(~~\pi_{\ident{~WorldBank}}~(w_0) \vdash e : \tau
-  && \textnormal{If $e$ is well-typed using a compile-time world $w_0$}
-\\
-\quad \wedge~ \langl e, w \rangl \rightarrow^* e'
-  && \textnormal{\ldots and the expression reduces in one or more steps}
-\\
-\quad \wedge~ (~\forall c \in \ident{countries}(e),
-  && \textnormal{\ldots and the run-time world model $w$ contains all}
-\\
-       \qquad~~~ \forall i \in \ident{indicators}(e) . \langl c, i \rangl \in \mathit{dom}(w)))
-  && \textnormal{\;\quad country-indicator pairs that may be accessed,}
-\\
-\Rightarrow ~\ident{value}(e') ~\vee~ (\exists e'' . \langl e', w \rangl \rightarrow e'')
-  && \textnormal{then the result is either a value or can further reduce.}
-\end{array}
-\end{equation*}
-\end{theorem*}
-
-\noindent
-The theorem has the usual structure of a type safety theorem. The reduction is defined on pairs
-consisting of expression and a world $w$. The result of the reduction is only an expression and
-so it cannot modify the world. Even though the above is only a brief sketch, it demonstrates two
-important points about relativized type safety that are relevant to many type providers.
-
-First, we distinguish between the state of the world $w_0$ that is used at compile-time and the
-state $w$ used at run-time. If these were the same, the additional assumption would always hold.
-Second, the additional assumptions states that all countries and indicators that \emph{may be
-accessed} need to be available. This is stricter than necessary (akin to effect systems),
-but it does not specify that the run-time world must contains \emph{all} provided countries
-and indicators.
 
@@
 
-In this section, we discuss the options available when targeting JavaScript. The compilation to
-JavaScript can be implemented as compiler back-end or as a library based on meta-programming
-(Section~\ref{sec:js-meta}), there are different approaches to base libraries (Section~\ref{sec:js-lib}),
-asynchronous computing (Section~\ref{sec:js-async}) and the semantic mismatch between the host
-language and JavaScript (Section~\ref{sec:js-relative}).
+A wide range of choices is available when targeting JavaScript. The compilation to JavaScript can be
+implemented as compiler back-end or as a library (Section~\ref{sec:js-meta}) and there are different
+approaches to base libraries (Section~\ref{sec:js-lib}) and asynchronous computing
+(Section~\ref{sec:js-async}).
 
@@
 The FunScript library is based on F\# quotations \cite{fsharp-metaprog}. This means that the code
-is compiled as ordinary F\#, but the compiler also stores marked blocks of code as data. In FunScript,
-we use a launcher that, when the program runs, analyses its quotations and produces a JavaScript file
-(so, the workflow is to compile the code as usual and then run it to produce JavaScript).
-
+is compiled as ordinary F\#, but the compiler also stores marked blocks of code as data.
 In FunScript, we typically need to translate the entire source file. To allow this, we instruct the
 compiler to store code as data for the whole module using the \ident{ReflectedDefinition} attribute.
-This is done at the beginning of the file (which we omitted in Section~\ref{sec:case}):
-%
+This is done at the beginning of the file:
+
 \begin{equation*}
@@
 When marked with \ident{ReflectedDefinition}, the F\# compiler stores the body of all functions and
-methods in the marked module as \emph{quotations} that can be retrieved at run-time. Note that this does
-not store the AST for the entire file -- for example, information about declared types needs to be
-obtained separately using the .NET reflection mechanism.
-
-The quotation mechanism has other uses. In the following, we use an explicit quotation $\langl @ \ldots @\rangl$,
-which returns the wrapped code as data. This can be used, for example, to compile a matrix calculation
-to run on a GPU \cite{accelerator}:
-%
-\begin{equation*}
-\begin{array}{l}
- \langl @~ \kvd{fun}~(\ident{input}:\ident{Matrix}\langl\ident{float32}\rangl)~\rightarrow \\
- \qquad~~ \kvd{let}~\ident{sum}~=~(\ident{shift input}~-\hspace{-0.2em}1~0) ~+~ \ident{input} ~+~ (\ident{shift input}~+\hspace{-0.2em}1~0) \\
- \qquad~~ \ident{sum}~/~3.0\textnormal{f}~~@\rangl
-\end{array}
-\end{equation*}
-%
+methods in the marked module as \emph{quotations} that can be retrieved at run-time.
 Compared to writing a full compiler back-end, the lightweight approach to meta-programming makes it easy to
 write a translator from F\# to other languages. This has been used for compiling to SQL queries, compiling
-F\# code to CUDA, for Freebase queries \cite{fsharp-data} and other applications.
+F\# code to CUDA or for Freebase queries \cite{fsharp-data}. For example, explicit quotations such as
+$\langl @~ ((\ident{shift input}~-\hspace{-0.2em}1) ~+~ \ident{input})/2 ~@\rangl$
+have been used for translating F\# code to GPU \cite{accelerator}.
 
@@
 compiler back-end is a better choice. However, for the task solved in our case study (compiling
-newly written F\# code), the approach works well. A related question is providing mappings for
-standard (in our case, F\# and .NET) libraries used by the code. We discuss this in the next section.
+newly written F\# code), the approach works well.
 
@@
 uses .NET \ident{IEnumerable} interface) and JavaScript libraries. Accessing these libraries in
-FunScript follows a similar approach to F\# WebTools \cite{fsharp-webtools}.
+FunScript follows the F\# WebTools \cite{fsharp-webtools}.
 
 The range of possible approaches to libraries when translating to JavaScript has two extreme cases.
-We can attempt to port all libraries of our source language (F\#, Haskell or even .NET), or we
-can ignore most standard libraries and instead provide access to as many JavaScript libraries as
-possible. FunScript stands in the middle -- it provides access to \emph{some} F\# and .NET libraries,
-but most advanced functionality is accessed through JavaScript libraries.
+We can attempt to port all libraries of the source language (F\#, Haskell or even .NET), or we
+can ignore standard libraries and instead provide access to all JavaScript libraries. FunScript
+stands in the middle -- it provides access to \emph{some} F\# and .NET libraries,
+but relies  on JavaScript for more advanced functionality.
 
 \paragraph{Mapping standard .NET and F\# libraries.}
-The F\# ecosystem relies on .NET libraries. This means that taking an F\# library and translating it
-to JavaScript without any modification is not a viable approach as we would have to be able to also
-translate any (compiled) .NET library. This is where F\# differs, for example, from Haskell which has
-a closed ecosystem and the ghcjs project \cite{haskell-ghcjs} is thus capable of translating many
-standard libraries.
-
-FunScript implements (explicit) mappings for standard F\# and .NET libraries that are useful in
-JavaScript (collections, string manipulation, date and other). Those are either reimplemented in
-F\# with the \ident{ReflectedDefinition} attribute, or mapped to standard JavaScript library.
+The F\# ecosystem relies on .NET, so taking an F\# library and translating it to JavaScript without
+any modification would require translating any (compiled) .NET library. This is where F\# differs,
+for example, from Haskell which has a closed ecosystem and the ghcjs project \cite{haskell-ghcjs}
+is thus capable of translating most standard libraries.
 
 An interesting aspect of the case study is that we use F\# Data \cite{fsharp-data} type providers,
-which is a standard F\# library that has not been built specifically for FunScript. This illustrates
-how type providers and quotations interact.
-
+which is a standard F\# library that has not been built specifically for FunScript. Thus it is worth
+explaining how type providers and quotations interact.
 When we write code using type providers that provide erased types (both World Bank and TypeScript),
-then the erasure happens before a quotation is captured. The following listing shows an example.
-When we write code with explicit (or implicit) quotation containing a provided method (1), the actual
+then the erasure happens before a quotation is captured. When we write code with explicit (or
+implicit) quotation containing a provided method (1), the actual
 quotation (2) contains just calls to the underlying runtime library:
@@
 %
-Thus, the FunScript component only needs to provide a client-side implementation of the underlying
+As a result, the FunScript component only needs to provide a client-side implementation of the underlying
 operations \ident{GetCountries}, \ident{GetCountry} and \ident{AsyncGetIndicator}. This is done in the
@@
 \paragraph{Accessing standard JavaScript libraries.}
-As discussed earlier, client-side programs can either rely more heavily on JavaScript libraries,
-or they can reimplement most of the core functionality. The latter approach is easier when all code
-is written for the same ecosystem. This is done, for example, by ghcjs \cite{haskell-ghcjs}.
-This technique can be very effective when writing code that will be a part of a larger JavaScript
-application and implements, for example, a core computation. This can then be compiled to run
-efficiently using asm.js \cite{asm-js}.
-
-The case study presented in this paper focuses more on the plumbing code that connects multiple
-components -- F\# type providers for data access and JavaScript libraries for visualization. Thus,
-a smooth integration between F\# and JavaScript is necessary.
+As discussed earlier, the focus of this case study is on integrating multiple ecosystems -- using
+F\# type providers for data access and JavaScript libraries for visualization. In this approach
+a high degree of integration between F\# and JavaScript is necessary.
 
@@
 mapped to ordinary F\# types and which types should be treated as opaque JavaScript types that
-can only be manipulated by JavaScript operations and operators. For example, consider the following
-excerpt from the code in Section~\ref{sec:case-loading}:
+can only be manipulated by JavaScript operations. Consider the following excerpt from the code
+in Section~\ref{sec:case-loading}:
 %
@@
  \\[0.5em]
- \ident{opts.series.push}(\ident{h.HighchartsSeriesOptions}(\ident{data}, \ident{name}))  \\
- (\ldots)
+ \ident{opts.series.push}(\ident{h.HighchartsSeriesOptions}(\ident{data}, \ident{name}))
 \end{array}
@@
 the \ident{series} property of \ident{HighchartsOptions} is \kvd{undefined}. The \ident{push}
-operation is added (as an extension method) to standard .NET array, but it cannot be
-implemented for ordinary F\# code. The following list briefly summarizes the options for
-mapping JavaScript libraries to F\# and the choices made in FunScript:
+operation is added to standard .NET array, but it cannot be implemented for ordinary F\# code.
+In summary, the options for mapping JavaScript libraries to F\# and the choices made by FunScript
+are:
 
@@
   However, the floating-point arithmetic in JavaScript differs from the one in F\# and so this changes
-  the semantics. Similarly, compilation of integer operators requires additional work (for
-  example, $1/2$ is not an integer division in JavaScript).
+  the semantics. Compilation of integer operators requires additional work (as $1/2$ is not an
+  integer division in JavaScript).
 
 \item FunScript maps F\# arrays to JavaScript arrays. We can define extension methods
-  such as \ident{push} that make sense for JavaScript arrays. An alternative is to use a
-  separate type -- for example js\_of\_ocaml uses a separate \ident{js\_array} type.
-  A nice thing about the FunScript approach is that it lets us use standard F\# modules
-  and functions such as \ident{Array.map}.
+  such as \ident{push} that make sense for JavaScript arrays. This lets us use standard
+  F\# modules and functions such as \ident{Array.map}. An alternative is to use a
+  separate type -- js\_of\_ocaml uses a separate type \ident{js\_array}.
 
@@
   on types imported from JavaScript (but not for types defined in F\#). In contrast,
-  js\_of\_ocaml is more explicit -- for example the \ident{attr} operation is marked as
-  returning \ident{optdef} (potentially undefined) \ident{js\_string} value whereas FunScript
-  uses just .NET \ident{string} (which may be \kvd{null}).
+  js\_of\_ocaml is more explicit and uses \ident{optdef} type to denote potentially undefined
+  values.
 \end{itemize}
@@
 \noindent
-In summary, the approach used by F\# is to reuse as much of F\# and .NET as possible. This
+In summary, the approach used by F\# is to reuse as much of F\# and .NET as easily possible. This
 makes writing and reusing code easier and lets developers use existing familiar libraries and
-idioms. However, it means that the semantics of F\# running as JavaScript is not the same as the
-semantics of F\# running as compiled code. We return to this topic in Section~\ref{sec:js-relative}.
-Before that, the following section gives one more example where similar approach is used.
-
-\subsection{Client-side asynchronous computations}
+idioms. However, it means that the semantics of F\# running as JavaScript is not always strictly
+the same as the semantics of F\# running as compiled code.
+
+\paragraph{Client-side asynchronous computations.}
 \label{sec:js-async}
-
 In F\#, asynchronous workflows \cite{fsharp-async} serve two purposes. First, they provide a
@@
 latter aspect. When loading data for selected countries in a loop, we fetch data asynchronously
-(using an AJAX call). Under the cover, JavaScript triggers a callback, which then resumes the
+(using AJAX). Under the cover, JavaScript triggers a callback, which then resumes the
 asynchronous workflow.
 
-Although the recent HTML5 standard makes it possible to write multi-threaded JavaScript using
-Web Workers, the FunScript implementation of $\ident{async}~\{ \ldots \}$ does not do that.
-Instead, it uses a simple cooperative model based on continuations (which is similar to how Links
-\cite{web-links} compilation works).
-
-In ordinary F\#, asynchronous workflows can be started in a number of ways: \ident{Async.RunSynchronously}
-starts the work and blocks until it completes; \ident{Async.Start} starts the work in the background
-and \ident{Async.\-Start\-Immediate} starts the work on the current thread and runs callbacks
-in the same synchronization context (essentially on the same thread). However, only the last one
-can be mapped to JavaScript (FunScript translator will fail if the other two are used).
-
-Our approach is again to reuse standard F\# constructs, but interpret them in a way that fits
-with the new execution environment. The alternative would be to define a separate computation type
-such as $\ident{js\_async}~\{ \ldots \}$. So, developers can use existing constructs, but need
-to be aware that their code runs in a different environment.
+FunScript uses a simple cooperative model based on continuations (which is similar to how Links
+\cite{web-links} compilation works). In ordinary F\#, asynchronous workflows can be started in a
+number of ways: \ident{Async.RunSynchronously} starts the work and blocks until it completes;
+\ident{Async.Start} starts the work in the background and \ident{Async.StartImmediate} starts the
+work on the current thread and runs callbacks on the same thread. Only the last one
+can be mapped to JavaScript (FunScript translator will fail at translation-time if the other two are used).
+Our approach is again to reuse standard F\# constructs, but interpret them in a different way that
+fits better with the new execution environment.
+
+% =================================================================================================
+%
+% =================================================================================================
+
+\section{Relativized safety and semantics}
+
+If there is one key message that the reader should take from this paper, it is the idea that both
+the type safety property and the langauge semantics can be relativized. That is, when we write code
+that uses the safe ML subset of F\#, we get all the good ML properties. However, to be able to
+interoperate with the open world and other ecosystems, the core language can be reinterpreted -- in
+a practically useful way but with semantics and safety that is relative with respect to the new
+environment.
+
+\subsection{Relativized type safety}
+\label{sec:tp-relative}
+
+Type providers for information sources, such as the World Bank provider, weaken the usual notion
+of type safety. In general, they require that certain assumptions about the world do not change
+between the state of the world at compile-time and state of the world at run-time. The TypeScript
+provider is different and it is more interesting to consider runtime behaviour as discussed
+in Section~\ref{sec:conc-interop}.
+
+For type providers that provide access to external information sources, we can formulate a
+\emph{relativized} notion of type safety. A full formalization is beyond the scope of this
+paper, but the following provides an outline of such a theorem.
+
+\begin{theorem*}[Relativized type safety]
+Assume $\pi_{\ident{~WorldBank}}$ is the mapping implemented by the World Bank type provider,
+$w_0, w_1$ are models of a (read-only) world that can be seen as functions defined on country-indicator
+pairs and $\langl e, w \rangl \rightarrow e'$ is a one-step reduction on expressions that has
+access to the World Bank data modelled by $w$.
+
+Then, if $e$ is well-typed using a compile-time model of the world $w_0$ and the model of the world
+used at run-time $w_1$ contains all country indicator pairs that may be accessed by $e$, then $e$
+can make a reduction step, \emph{i.e.} $\langl e, w_1 \rangl \rightarrow e'$ and the type of $e'$
+is the same as the type of $e$.
+\end{theorem*}
+
+\noindent
+The theorem has the usual structure of a type safety theorem. The reduction is defined on pairs
+consisting of expression and a world $w$. The result of the reduction is only an expression and
+so it cannot modify the world. Even though we give only a brief sketch, it demonstrates two
+important points about relativized type safety that are relevant to many type providers.
+
+First, we distinguish between the state of the world $w_0$ that is used at compile-time and the
+state $w_1$ used at run-time. If these were the same, the additional assumption would always hold.
+Second, we require that all countries and indicators that \emph{may be accessed} need to be
+available. This is an over-approximation, but it does \emph{not} say that
+the run-time world must contains \emph{all} countries and indicators.
 
@@
 constructs. In this case study, we used the latter. This enables interop with JavaScript, but
-it makes it hard (if at all possible) to preserve the original F\# semantics in all cases.
+it makes it hard (if at all possible) to preserve the original F\# semantics.
 
@@
 libraries) and statically typed components (code written in F\#). This suggests a relation with
-the work on gradual typing \cite{gradual-typing} and blame tracking \cite{blame-well}. However,
+the work on gradual typing \cite{gradual-fun,gradual-oop} and blame tracking \cite{blame-well}. However,
 in our work, the distinction between the statically and dynamically typed parts (and the boundary)
@@
 
-\newpage
 There are two main assumptions that we reconsider. The first assumption is that programs
@@
 let us benefit from the functional-first statically-typed programming paradigm in the age of web.
+\newpage
 

### 2015-10-25 icfp-2015/build.sh

@@ -1,3 +1,26 @@
-latex paper.tex --non-interactive
-bibtex paper
-pdflatex paper.tex
+#!/bin/bash
+if test "$OS" = "Windows_NT"
+then
+  MONO=""
+else
+  # Mono fix for https://github.com/fsharp/FAKE/issues/805
+  export MONO_MANAGED_WATCHER=false
+  MONO="mono"
+fi
+
+$MONO .paket/paket.bootstrapper.exe
+exit_code=$?
+if [ $exit_code -ne 0 ]; then
+  exit $exit_code
+fi
+if [ -e "paket.lock" ]
+then
+  $MONO .paket/paket.exe restore
+else
+  $MONO .paket/paket.exe install
+fi
+exit_code=$?
+if [ $exit_code -ne 0 ]; then
+  exit $exit_code
+fi
+$MONO packages/FAKE/tools/FAKE.exe $@ --fsiargs build.fsx

### 2015-10-25 icfp-2015/paper.tex
Adding alternative XML sample (with variants)

@@
 \usepackage[fleqn]{amsmath}
-\usepackage[hang,flushmargin]{footmisc} 
 \usepackage{graphics} 
@@
 \usepackage{ftnright}
-\usepackage[notquote]{hanging}
 \usepackage[T1]{fontenc}
@@
 \usepackage{enumitem}
-\usepackage{flushend}
+%\usepackage[hang,flushmargin]{footmisc} 
+%\usepackage[notquote]{hanging}
+%\usepackage{flushend}
 
@@
 
-% TODO: The model here does not actually use type erasure!!
-% TODO: Explain why we can only provide simple classes (!)
-% TODO: Insert a few more mentions of why "s1 + s2 <: s1 + s2 + s3" is valid (cannot do complete pattern match)
+% TODO: Adapt XML example 
+% TODO: Switch from sum types to variant types
+% TODO: Discuss 'stability' of the inference (minor change in schema does not cause dramatic cahnges)
+% TODO: Say that any variant is a subtype of any other variant 
+% TODO: Say that anything is a subtype of variant
+% TODO: Say that variant is a last-resort thing
+% TODO: Double check that bottom/top are used correctly
+% TODO: Explain (even more) that there is no least upper bound, which is why Theorem 3 is good enough
+% TODO: Avoid "no other rule applies" in the formal system
+% TODO: Practical evaluation (lots of downloads, lots of Github - how often people customize samples?)
 
@@
 \ident{Name} and \ident{Age}), but this would complicate the processing code.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Processing XML data, alternative}
+
+In XML, documents are formed by (possibly) nested elements with attributes. We can view elements
+as records (unlike in JSON, the records are named) with a field for each attribute and an additional
+special field for the nested contents (which may be a collection of elements).
+
+As an example, consider a simple extensible document format where a root element
+{\ttfamily\small <doc>} can contain a number of document element, one of which is
+{\ttfamily\small <heading>} representing headings:
+%
+{\small{
+\begin{verbatim}
+  <doc>
+    <heading>Working with JSON</heading>
+    <p>Type providers make this easy.</p>
+    <heading>Working with XML</heading>
+    <p>Processing XML is as easy as JSON.</p>
+    <image source="xml.png" />
+  </doc>
+\end{verbatim}
+}}
+%
+\noindent
+The F\# Data library has been primarilly designed to allow easy reading of data. For example,
+say we want to print all headings that appear in the document. The sample shows a part of the 
+possible document structure (in particular the {\ttfamily\small <heading>} element we want to
+read), but it does not show all possible elements (say, {\ttfamily\small <table>}). Assuming the
+above document is \strf{sample.doc}, we can write:
+%
+\noindent
+\begin{equation*}
+\begin{array}{l}
+ \kvd{type}~\ident{Document}~=~\ident{XmlProvider}\langl\str{sample.doc}\rangl\hspace{1em} \\[0.5em]
+ \kvd{let}~\ident{root}~=~\ident{Document.Load}(\str{http://.../another.doc})\\
+ \kvd{for}~\ident{elem}~\kvd{in}~\ident{root.Doc}~\kvd{do}\\
+ \quad\kvd{match}~\ident{elem.Heading}~\kvd{with} \\
+ \quad|~\ident{Some text} \rightarrow \ident{printf}~\str{ - \%s }~\ident{text}\\
+ \quad|~\ident{None} \rightarrow ()\\
+\end{array}
+\end{equation*}
+%
+The example iterates over a collection of elements \ident{elem} returned by \ident{root.Doc}. 
+Our system provides a typed access to elements known from the sample document and so we can
+write \ident{elem.Heading}.
+
+\paragraph{Open world and variant types.}
+In the above example, The system does not infer a union type representing a choice between
+heading, paragraph and image. By its nature, the XML format is extensible and the sample cannot 
+include all possible examples\footnote{Even when the document structure is defined using XML 
+Schema, documents may contain elements prefixed with other namespaces.}. The type of \ident{elem}
+is thus a \emph{variant} type annotated with the three statically known possibilities (heading,
+paragraph, image). The variant is mapped to the following F\# type:
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{type}~\ident{Element}~=  \\[0.1em]
+ \quad \kvd{member}~\ident{Heading}~:~\ident{option}\langl \ident{string} \rangl\\
+ \quad \kvd{member}~\ident{Paragraph}~:~\ident{option}\langl \ident{string} \rangl\\
+ \quad \kvd{member}~\ident{Image}~:~\ident{option}\langl \ident{Image} \rangl\\
+\end{array}
+\end{equation*}
+%
+The type provides a nice access to the elements known statically from the sample, but it encodes
+the \emph{open-world assumption} -- the user also needs to handle the case when the value is none 
+of the statically known elements.
+
+One perhaps surprising consequence of the use of variant types is that the above code will work
+correctly on document that contains elements other than those in the sample document. This would 
+arguably not be desirable design for data types in a programming language, but it is extremely
+useful when reading data from external data sources. It is also worth noting that our type 
+inference attempts to minimze the number of variant types that appear in the resulting type --
+a variant is not used if there is another option.
 
@@
 \acks
+?
 We would like to thank to the F\# Data contributors on GitHub and other colleagues working on type providers,

### 2015-10-25 icfp-2015/paper.tex
Delete old XML samples and Erased type providers

@@
 $\ident{option}\langl\ident{int}\rangl$.
-
-\paragraph{Erasing type providers.}
-At compile-time, F\# type providers use an erasure mechanism similar to Java Generics \cite{java-erasure}. 
-A type provider generates types and code that should be executed when members are accessed. In compiled
-code, the types are erased and the program directly executes the generated code.
-In the above example, the compiled (and actually executed) code looks as follows:
-%
-\begin{equation*}
-\begin{array}{l}
- \kvd{let}~\ident{air}~=~\ident{CsvFile}\ident{.Load}(\str{airdata.csv}, \kvd{fun}~\ident{r}\rightarrow\\[0.1em]
- \quad \ident{asDecimal r}.[\num{0}], \ident{asIntOpt r}.[\num{1}], \ident{r}.[\num{2}], \ident{asBool r}.[\num{3}])\\[0.6em]
- \kvd{for}~\ident{c1},\_,\ident{c3},\ident{c4} ~\kvd{in}~\ident{air.Rows}~\kvd{do}\\[0.1em]
- \quad\kvd{if}~\ident{not}~\ident{c4}~\kvd{then}~\ident{printf}~\str{\%s: \%d}~\ident{c3 c1}\\
-\end{array}
-\end{equation*}
-%
-The generated type \ident{AirCsv} is erased to a type $\ident{CsvFile}\langl\alpha\rangl$.
-The \ident{Load} method takes the file name together with a function that turns a row represented 
-as an array of strings to a typed representation -- here, a four-element tuple 
-$\ident{decimal}*\ident{option}\langl\ident{int}\rangl*\ident{string}*\ident{bool}$. The properties
-such as \ident{Ozone} are then replaced with an accessor that reads the corresponding tuple
-element.
-
-The formal model in Section~\ref{sec:formal} does not discuss the erasure mechanism and instead 
-models evaluation of the provided types. We do not lose anything here -- the erasure is mainly
-useful when the provided types are very large (and cannot be efficiently generated all at once).
-For more, see the F\# type providers report \cite{fsharp-typeprov}.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Processing XML data}
-\label{sec:providers-xml}
-
-The XML format is similar to JSON in that it uses nested elements rather than a flat 
-structure. Our inference algorithm uses the notion of \emph{records} for both JSON records 
-and XML elements. However, unlike in JSON, the XML elements also have a name.
-
-An XML element can contain a collection of child elements. The collection is often heterogeneous 
-containing child nodes of different names. Consider the following (simplified) RSS feed:
-%
-{\small{
-\begin{verbatim}
-  <rss version="2.0"><channel>
-    <title> BBC News - Europe </title>
-    <item><title> Kurdish activists 
-        killed in Paris </title></item>
-    <item><title> German MPs warn 
-        over UK EU exit </title></item>
-  </channel></rss>
-\end{verbatim}
-}}
-%
-\noindent
-The \ident{channel} node contains a single \ident{title} node (with the name of the feed) and multiple 
-\ident{item} nodes (individual news articles). In RSS, the \ident{title} nodes always contain plain text,
-while \ident{item} nodes contain \ident{title} (and also \ident{url} and other properties omitted here).
-
-When extracting data from XML, we often need to get nodes of a specific name (all
-items) or extract the content of a node (feed title). The XML type provider is optimized for 
-this style of access:
-
-\noindent
-\begin{equation*}
-\begin{array}{l}
- \kvd{type}~\ident{RssFeed}~=~\ident{XmlProvider}\langl\str{rss-sample.xml}\rangl\hspace{1em} \\[0.5em]
- \kvd{let}~\ident{channel}~=~\ident{RssFeed.Load}(\str{http://bbc.co.uk/...}).\ident{Channel}\\
- \ident{printf}~\str{News from: \%s }~\ident{channel.Title}\\
- \kvd{for}~\ident{item}~\kvd{in}~\ident{channel.Items}~\kvd{do}\\
- \quad\ident{printf}~\str{ - \%s }~\ident{item.Title}\\
-\end{array}
-\end{equation*}
-%
-In this example, the type of the \ident{channel} value represents a heterogeneous collection
-that contains exactly one \ident{title} node and zero or more \ident{item} nodes.
-
-\paragraph{Heterogeneous collections and unions.}
-Our inference algorithms recognizes the \ident{channel} element as a \emph{heterogeneous collection} (if the same structure 
-appears in all samples) and generates a type with a member \ident{Title} for accessing the value 
-of the singleton \ident{title} element together with a member \ident{Items} that returns a 
-collection of zero or more items. As a further simplification, if an element contains only a 
-primitive value (such as \ident{title}) it is represented as a value of primitive type (so 
-\ident{channel.Title} has a type \ident{string}).
-
-As an alternative, the inference could produce a collection of union type (with a case for \ident{item}
-and a case for \ident{title}). This alternative is useful for documents with complicated structure 
-(such as XHTML files), but less useful for files with simpler structure such as web server responses,
-data stores and configuration files.
-
-Moreover, our approach based on heterogeneous collections has a nice property that it generates types
-with the same public interface for the following two ways of representing data in XML:
-%
-{\small{
-\begin{verbatim}
-  <author name="Tomas" age="27" />
-  <author><name>Tomas</name><age>27</age></author>
-\end{verbatim}
-}}
-%
-\noindent
-In both cases, the provided type for \ident{author} has two properties, typed \ident{string} and 
-\ident{int} respectively. In Section~\ref{sec:inference}, we discuss the simplified model (using 
-union types). The inference based on heterogeneous collections is described later in 
-Section~\ref{sec:impl-collections}.
 

### 2015-10-31 icfp-2015/paper.tex
Switch to use variants

@@
 % TODO: Explain (even more) that there is no least upper bound, which is why Theorem 3 is good enough
+% TODO:  -- but maybe we can have lub w.r.t variants? or something?
 % TODO: Avoid "no other rule applies" in the formal system
@@
 \subsection{Processing XML data, alternative}
+\label{sec:providers-xml}
 
@@
 that always have a valid value (written as $\hat{\sigma}$) and \emph{nullable types} that encompass missing 
-and \kvd{null} values (written as $\sigma$). We write $\nu$ for record field names and $\nu?$
+and \kvd{null} values (written as $\sigma$). We write $\nu$ for record field names and $\nu_?$
 for record names, which can be empty:
@@
 \begin{array}{rcl}
- \hat{\sigma} &\narrow{=}& \nu?\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
+ \hat{\sigma} &\narrow{=}& \nu_?\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                 &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bit} \lsep \ident{bool} \lsep \ident{string} 
  \\[0.6em] 
-       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \hat{\sigma}\;\,\kvd{option} \lsep [\sigma] \\[0.1em]
-              &\narrow{|}& \sigma_1 + \ldots + \sigma_n \lsep ~\top~ \lsep \kvd{null}
+       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{option}\langl \hat{\sigma} \rangl \lsep [\sigma] \\[0.1em]
+              &\narrow{|}& \kvd{var}\langl\sigma_1, \ldots, \sigma_n\rangl \lsep ~\top~ \lsep \kvd{null}
 \end{array}
@@
 
-Finally, a union type in our model implicitly permits the \kvd{null} value. This is because structural union 
-types are \emph{not} mapped to F\# union types. Instead our encoding requires the user to handle the 
-cases one-by-one and so they also need to handle the situation when none of the cases matches (as discussed
-in Section~\ref{sec:impl-unions}, this is the right choice for an F\# type provider, but it would not 
-necessarily be a good fit for other languages).
-
-% TODO: Also include reference to section with more on heterogeneous collections.
+Finally, a variant type \kvd{var} is annotated with the statically known types it may represent based
+on the sample document. As discussed earlier (Section~\ref{sec:providers-xml}), the variant type forces
+the user to handle the case when a value is of a different type than the ones statically known. For
+this reason, the variant type also implicitly permits the \kvd{null} value.
 
@@
 We write $\sigma_1 :> \sigma_2$ to denote that $\sigma_2$ is a subtype of $\sigma_1$. The 
-subtyping relation is defined as a transitive reflexive closure of these rules:
+subtyping relation is defined as a transitive reflexive closure of the following rules:
 
@@
 
-\noindent\boxed{\emph{Union types}}
+\noindent\boxed{\emph{Variant types}}
 \begin{align}
-\label{eq:sub-union1}\tag{U1}
-\sigma_1 + \ldots + \sigma_n :> \sigma_1' + \ldots + \sigma_n' \quad &(\textnormal{iff}~\forall i.\sigma_i :> \sigma_i') \\
-\label{eq:sub-union2}\tag{U2}
-\sigma_1 + \ldots + \sigma_n :> \sigma_1 + \ldots + \sigma_m \quad &(\textnormal{where}~m \leq n) \\
-\label{eq:sub-union3}\tag{U3}
-\sigma_1 + \ldots + \sigma_n :> \sigma_1 + \ldots + \sigma_m \quad &(\textnormal{where}~m \geq n) \\
-\label{eq:sub-union4}\tag{U4}
-\hspace{-1em}\sigma_1 + \ldots + \sigma_n :> \sigma_{\pi(1)} + \ldots + \sigma_{\pi(n)} \quad &(\textnormal{permutation}~\pi) \\
-\label{eq:sub-union-el}\tag{U5}
-\sigma_1 + \ldots + \sigma_n :> \sigma_i \quad &(\textnormal{for}~i \in 1 \ldots n)
+\label{eq:sub-var-bot}\tag{U1}
+\kvd{var}\langl \sigma_1, \ldots, \sigma_n\rangl :>&~ \sigma \\
+\label{eq:sub-var-eq}\tag{U2}
+\kvd{var}\langl \sigma_1, \ldots, \sigma_n \rangl :>&~ \kvd{var}\langl \sigma_1', \ldots, \sigma_n' \rangl
 \end{align}
@@
 \label{eq:sub-record1}\tag{R1}
- \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \hspace{-0.1em}:> \hspace{-0.1em}
- \nu?~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
+ \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \hspace{-0.1em}:> \hspace{-0.1em}
+ \nu_?~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
  \; (\sigma_i \hspace{-0.25em}:>\hspace{-0.25em} \sigma_i')
@@
 \label{eq:sub-record2}\tag{R2}
- \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \hspace{-0.1em} :> \hspace{-0.1em}
- \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_m\!:\!\sigma_m \}
+ \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \hspace{-0.1em} :> \hspace{-0.1em}
+ \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_m\!:\!\sigma_m \}
  \; (m \hspace{-0.1em}\geq\hspace{-0.1em} n)
@@
 \label{eq:sub-record3}\tag{R3}
-\hspace{-1.75em}
-\begin{array}{l}
- \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
- \nu?~\{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, .., \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
+\hspace{-1.5em}
+\begin{array}{l}
+ \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
+ \nu_?~\{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, .., \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
 \end{array} \qquad &(\textnormal{permutation}~\pi)
@@
 \label{eq:sub-record4}\tag{R4}
-\hspace{-1.75em}
-\begin{array}{l}
- \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null}, .., \nu_{n+m}\!:\!\kvd{null} \} :> \\
- \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
+\hspace{-1.5em}
+\begin{array}{l}
+ \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null}, .., \nu_{n+m}\!:\!\kvd{null} \} :> \\
+ \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
 \end{array}
@@
   Thus, a sample $0,1$ has a type \ident{bit}; $0,1,\kvd{true}$ is \ident{bool} and
-  $0,1,2$ becomes \ident{int} (for a sample $0,1,2, \kvd{true}$ we would need a union type).
+  $0,1,2$ becomes \ident{int} (for a sample $0,1,2, \kvd{true}$ we would need a variant type).
 
@@
 
-\item There is a top type (\ref{eq:sub-top}), but no bottom type. It is possible to find common 
-  supertype of any two types, because a union type $\tau_1 + \ldots + \tau_n$ is a supertype of all its 
-  components (\ref{eq:sub-union-el}).
-
-\item For unions, we include the standard rules. Subtype can have fewer alternatives 
-  (\ref{eq:sub-union1}), elements can be reordered (\ref{eq:sub-union2}) and subtyping is covariant 
-  (\ref{eq:sub-union4}). An unusual rule (\ref{eq:sub-union3}) says that subtype can also have 
-  \emph{more} cases. This is valid because the type provider exposes unions as objects that do not
-  allow complete pattern matching.
+\item There is a top type (\ref{eq:sub-top}) and variants can be seen as the bottom because
+  any type $\sigma$ is a subtype of a variant (\ref{eq:sub-var-bot}). There is no \emph{single} 
+  bottom type -- there is a range of distinct variant types, but any variant is a subtype
+  of any other variant (\ref{eq:sub-var-eq}).
 
@@
 \noindent
-The rule that allows subtype to have fewer record elements (\ref{eq:sub-record4}) is particularly
-important. It allows us to prefer records in some cases. For example, given two samples 
-$\{ \ident{name}:\ident{string} \}$ and $\{ \ident{name}:\ident{string}, \ident{age}:\ident{int} \}$,
-we can find a common supertype $\{ \ident{name}:\ident{string}, \ident{age}:\ident{int}~\kvd{option} \}$
-which is also a record. For usability reasons, we prefer this to another common supertype
-$\{ \ident{name}:\ident{string} \} + \{ \ident{name}:\ident{string}, \ident{age}:\ident{int} \}$.
+If we were proposing a type system for type-checking programs, the use of multiple variant types
+(which all are subtypes of each other) would not add any value and we could introduce a single 
+bottom type instead. However, our purpose is different -- we aim to make reading structured data
+easier and, as shown in Section~\ref{sec:providers-xml}, the types forming a variant provide a
+useful information about the sample document.
+
+Another important rule is (\ref{eq:sub-record4}), which allows subtype to have fewer record 
+elements. In some cases, this allows us to infer records rather than variants. For example, given 
+$\{ \ident{name}\!:\!\ident{string} \}$ and $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int} \}$,
+we can find a common supertype $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int}~\kvd{option} \}$
+which is also a record. This is more precise and more usable type than another common supertype
+$\kvd{var}\langl \{ \ident{name}:\ident{string} \}, \{ \ident{name}:\ident{string}, \ident{age}:\ident{int} \}\rangl$.
 The next section describes precisely how our inference algorithm works.
@@
 (we infer \ident{string}~\kvd{option} as a hint to the user rather than treating \kvd{null} as an 
-empty string). These choices are based on experience with using F\# Data in practice.
+empty string). These choices are based on experience with using F\# Data in practice
+and might differ for other languages.
 
@@
 the partially ordered set of types does not have a unique greatest lower bound. For example, record types
-$\{\ident{a}:\ident{int}\}$ and $\{\ident{b}:\ident{bool}\}$ have common supertypes
-$\{\ident{?a}:\ident{int}, \ident{?b}:\ident{bool}\}$ and $\{\ident{a}:\ident{int}\} + \{\ident{b}:\ident{bool}\}$
-which are unrelated. Our inference algorithm always prefers records over unions (Theorem~\ref{thm:no-unions}). 
+$\{\ident{a}\!:\!\ident{int}\}$ and $\{\ident{b}\!:\!\ident{bool}\}$ have common supertypes
+$\{\ident{?a}\!:\!\ident{int}, \ident{?b}\!:\!\ident{bool}\}$ and 
+$\kvd{var}\langl\{\ident{a}\!:\!\ident{int}\}, \{\ident{b}:\ident{bool}\}\rangl$
+
+NOT TRUE
+which are unrelated.
+
+ Our inference algorithm always prefers records over unions (Theorem~\ref{thm:no-unions}). 
 The definition of the \emph{common supertype} relation uses a number of auxiliary definitions 

### 2015-11-14 pldi-2016/paper.tex
Edits in TP formalization

--- 2015-10-31 icfp-2015/paper.tex
@@
-\documentclass[preprint]{sigplanconf}
+\documentclass[10pt,preprint,clearpagebib]{sigplanconf}
+
+\usepackage{listings}
+\usepackage{enumitem}
+\usepackage[hyphens]{url}
+\usepackage[svgnames]{xcolor}
+\definecolor{lcl}{RGB}{160,0,128}
+\usepackage[colorlinks=true,breaklinks,draft=false]{hyperref}
+\hypersetup{urlcolor=lcl,linkcolor=lcl}
+\newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}
 
@@
 \usepackage{amsthm}
-\usepackage[hyphens]{url}
-\usepackage{hyperref}
 \usepackage{ftnright}
@@
 \usepackage{enumitem}
-%\usepackage[hang,flushmargin]{footmisc} 
-%\usepackage[notquote]{hanging}
-%\usepackage{flushend}
+\usepackage[hang,flushmargin]{footmisc} 
+\usepackage[notquote]{hanging}
+\usepackage{flushend}
 
@@
 % TODO: Practical evaluation (lots of downloads, lots of Github - how often people customize samples?)
+% TODO: Explain why we need option type (because T + null is a variant)
+
+% TODO: In the PLDI version, we delete all the unnecessary runtime conversions
+% but we should say that those happen for practical reasons (e.g. float -> int, or anyhing to string)
+
+% T :> S means that S is a subtype of T (and S can be used where T is expected)
+% TOP is the universal supertype
+% BOTTOM is the universal subtype
+
+% variant is the top type 
+%  (variant is a suprtype of everything; everything is a subtype of variant; 
+%   everything can be used where variant is expected)
+%
+% bottom is the bottom type
+%  (bottom is a subtype of everything; everything is a supertype of bottom
+%   bottom can be used where anything else is expected - but it has no values)
+%
+% define :=: when <: and :>, then take equivalence classes
+
+% least upper bound of T1 and T2 is a supertype of both (T :> T1 and T :> T2)
+% and for all S such that S :> T1 and S :> T2, it holds that S :> T
 
@@
 \newcommand{\lsep}[0]{\;\; | \;\;}
-\newcommand{\narrow}[1]{\hspace{-0.85em} #1 \hspace{-0.85em}}
+\newcommand{\narrow}[1]{\hspace{-0.7em} #1 \hspace{-0.7em}}
 
@@
 \newtheorem{theorem}{Theorem}
-\newtheorem{lemma}{Lemma}
+\newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{lemma}[theorem]{Lemma}
 
@@
 \copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
-\doi{nnnnnnn.nnnnnnn}
-
-\titlebanner{Unpublished draft, March 2015}        % These are ignored unless
+%\doi{nnnnnnn.nnnnnnn}
+
+%\titlebanner{Unpublished draft, March 2015}        % These are ignored unless
 %\preprintfooter{short description of paper}   % 'preprint' option specified.
@@
 \begin{abstract}
-Most static type systems assume that a program runs in a closed world, but this is not the 
-case. Modern applications interact with external services and often access data in structured 
-formats such as XML, CSV and JSON. Static type systems do not understand such external data sources
-and only make accessing data more cumbersome. Should we just give up and leave the messy world of 
-external data to dynamic typing and runtime checks?
-
-Of course, we should not give up! In this paper, we show how to integrate external data sources 
-into the F\# type system. As most real-world data on the web do not come with an explicit schema, 
-we develop a type inference algorithm that infers a type from representative samples. Next, we use 
-the type provider mechanism for integrating the inferred structures into the F\# type
-system.
-
-The resulting library significantly reduces the amount of code that developers need to write when 
-accessing data. It also provides additional safety guarantees. Arguably, as much safety as possible 
-if we abandon the incorrect closed world assumption.
+Most modern applications interact with external services and access data in formats such as 
+XML, JSON and CSV. Static type systems do not understand external data sources and only 
+make accessing data more cumbersome. Should we give up and leave the messy world of external data 
+to dynamic typing and runtime checks? Of course, not!
+
+We show how to integrate external data sources into the F\# type system. As most real-world data
+do not come with an explicit schema, we develop a type inference algorithm that infers a type from 
+representative sample documents and integrate it into the F\# type system using type providers.
+
+Our library significantly reduces the amount of code that developers need to write when 
+accessing data. It also provides additional safety guarantees, arguably, as much as possible if 
+we abandon the closed world assumption.
 \end{abstract}
@@
 \category{D.3.3}{Programming Languages}{Language Constructs and Features }
-\keywords F\#, Type Providers, JSON, XML, Data, Type Inference
+\keywords F\#, Type Providers, Inference, JSON, XML
 
@@
 
-The purpose of many modern applications is to connect multiple data sources or services and
-present the aggregated data in a new form. Mobile applications for taking notes, searching train
-schedules or finding tomorrow's weather all communicate with one or more services over the network.
-
+Applications for taking notes, searching train schedules or finding tomorrow's weather all 
+communicate with one or more services over the network and present the aggregated data. 
 Increasing number of such services provide REST-based end-points that return data as CSV, XML
-or JSON. Despite numerous schematization efforts, the services typically do not come with an explicit 
-schema. At best, the documentation provides a number of typical requests and sample responses.
-
-For example, the OpenWeatherMap service provides an end-point for getting current weather for a 
-given city\footnote{See ``Current weather data'': \url{http://openweathermap.org/current}}. The page documents the input 
-URL parameters and shows one sample JSON response to illustrate the response structure.
-Using a standard library for working with JSON and HTTP, we might call the service and 
-read the temperature as follows\footnote{We abbreviate the full URL: 
-\url{http://api.openweathermap.org/data/2.5/weather?q=Prague\&units=metric}}:
-
-\noindent
-\begin{equation*}
-\begin{array}{l}
- \kvd{let}~\ident{data}=\ident{Http.Request}(\str{http://weather.org/?q=Prague}) \\[0.1em]
- \kvd{match}~\ident{JsonValue.Parse}(\ident{data})~\kvd{with} \\[0.1em]
- |~\ident{Record}(\ident{root})\rightarrow \\[0.1em]
- \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\[0.1em]
- \quad |~\ident{Record}(\ident{main})\rightarrow \\[0.1em]
- \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\[0.1em]
- \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f degrees!}~\ident{num} \\[0.1em]
- \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\[0.1em]
- \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\[0.1em]
+or JSON. Despite numerous schematization efforts, most services do not come with an 
+explicit schema. At best, the documentation provides sample responses for typical requests.
+
+For example, OpenWeatherMap provides an end-point for getting current weather for a 
+given city\footnote{See ``Current weather data'': \url{http://openweathermap.org/current}}. 
+The documentation describes the URL parameters and shows one sample JSON to illustrate the typical
+response structure. Using a standard library for working with JSON and HTTP, we might call the 
+service and read the temperature as follows:
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://weather.org/?q=NYC}) \\
+ \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
+ |~\ident{Record}(\ident{root})\rightarrow \\
+ \quad \kvd{match}~\ident{Map.find}~\str{main}~\ident{root}~\kvd{with} \\
+ \quad |~\ident{Record}(\ident{main})\rightarrow \\
+ \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\
+ \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Nice \%f degrees!}~\ident{num} \\
+ \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
+ \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
  |~\_\rightarrow \ident{failwith}~\str{Incorrect format} 
@@
 \end{equation*}
-\vspace{0.1em}
-
-\noindent
+%
 The code assumes that the response has a particular format described in the documentation. The
 root node must be a record with a \str{main} field, which has to be another record containing
-a \str{temp} field with a numerical value. When the format is incorrect, the data access simply fails
+a numerical \str{temp} field. When the format is incorrect, the data access simply fails
 with an exception.
 
-Using the JSON type provider from the F\# Data library, we can write code with exactly the 
+Using the JSON type provider from F\# Data, we can write code with exactly the 
 same functionality in two lines:
-
-\vspace{-0.6em}
-\begin{equation*}
-\begin{array}{l}
- \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://weather.org/?q=Prague}\rangl \\[0.1em]
- \ident{printfn}~\str{Lovely \%f degrees!}~(\ident{W.GetSample().Main.Temp})
-\end{array}
-\end{equation*}
-\vspace{0.0em}
-
-\noindent
-On the first line, $\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider at 
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://weather.org/?q=NYC}\rangl \\[0.1em]
+ \ident{printfn}~\str{Nice \%f degrees!}~(\ident{W.GetSample().Main.Temp})
+\end{array}
+\end{equation*}
+%
+Here, $\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider at 
 compile-time with the URL as a sample. The type provider infers the structure of the response
@@
 
-In the rest of the paper, we give a detailed description of the mechanism and we also discuss
-safety properties of the approach. The key novel contributions of this paper are:
+The rest of the paper describes the mechanism and discusses its safety properties. 
+The key novel contributions are:
 
 \begin{itemize}
-\item We present type providers for XML, CSV and JSON (Section~\ref{sec:providers}) 
-  that are available in the F\# Data library and we also cover practical aspects of the 
-  implementation that contributed to their industrial adoption (Section~\ref{sec:impl}).
+\item We present F\# Data type providers for XML, CSV and JSON (\S\ref{sec:providers}) 
+  and practical aspects of their implementation that contributed to their industrial 
+  adoption (\S\ref{sec:impl}).
 
 \item We describe a predictable type inference algorithm for structured data formats that
-  underlies the type providers (Section~\ref{sec:inference}). It is based on finding common 
-  supertype of a set of examples.
-
-\item We provide a minimal formal model (Section~\ref{sec:formal}) and use it to prove a
-  relativized notion of type safety for our type providers (Section~\ref{sec:safety}). Although 
-  we focus on our specific case, the relativized safety property demonstrates a new way of 
-  thinking about ML-style type safety that is needed in the context of web.
+  underlies the type providers (\S\ref{sec:inference}) based on a \emph{common supertype}
+  relation.
+
+\item We give a formal model (\S\ref{sec:formal}) and use it to prove
+  \emph{relativized type safety} for the type providers (\S\ref{sec:safety}).
+  This adapts the ML-style type safety for the context of the web.
 \end{itemize}
 
-\noindent
-Although the F\# Data library \cite{fsharp-data} has been widely adopted and is one of the most downloaded F\# libraries 
-it has not yet been presented in an academic form. Additional documentation and source code can 
-be found at \url{http://fsharp.github.io/FSharp.Data}.
+% Although the F\# Data library \cite{fsharp-data} has been widely adopted and is one of the most downloaded F\# libraries 
+% it has not yet been presented in an academic form. Additional documentation and source code can 
+% be found at \url{http://fsharp.github.io/FSharp.Data}.
 
@@
 
+\newpage
+~
+\newpage
 \section{Structural type inference}
@@
 inferring the type of a document, we infer the most specific types of individual values (CSV rows,
-JSON or XML nodes) and find a common supertype of values in the sample.
+JSON or XML nodes) and recursively find a common supertype of all child nodes or sample documents.
 
 In this section, we define the \emph{structural type} $\sigma$, which is a type of structured data.
-Note that this type is distinct from the programming language types $\tau$ (the type generates the
-latter from the former). Next, we define the subtyping relation on structural types $\sigma$ and 
+Note that this type is distinct from the programming language types $\tau$ (type providers generate 
+the latter from the former). Next, we define the subtyping relation on structural types $\sigma$ and 
 describe the algorithm for finding a common supertype.
-
-Note that the subtyping relation between structural types $\sigma$ does not map to a subtyping 
-relation between F\#  types $\tau$. However, it does hold at runtime, meaning that an operation on 
-a supertype can be performed on its subtypes. We make the type provider mechanism and runtime aspects 
-precise in Section~\ref{sec:formal} before discussing safety in Section~\ref{sec:safety}.
 
@@
 
-The grammar below defines \emph{structural type} $\sigma$. We distinguish between \emph{non-nullable types}
-that always have a valid value (written as $\hat{\sigma}$) and \emph{nullable types} that encompass missing 
-and \kvd{null} values (written as $\sigma$). We write $\nu$ for record field names and $\nu_?$
-for record names, which can be empty:
+We distinguish between \emph{non-nullable types} that always have a valid value (written as 
+$\hat{\sigma}$) and \emph{nullable types} that encompass missing and \kvd{null} values 
+(written as $\sigma$). We write $\nu$ for record field names and $\nu_?$ for names that can be empty:
 %
@@
  \hat{\sigma} &\narrow{=}& \nu_?\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
-                &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bit} \lsep \ident{bool} \lsep \ident{string} 
+                &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
  \\[0.6em] 
        \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{option}\langl \hat{\sigma} \rangl \lsep [\sigma] \\[0.1em]
-              &\narrow{|}& \kvd{var}\langl\sigma_1, \ldots, \sigma_n\rangl \lsep ~\top~ \lsep \kvd{null}
-\end{array}
-\end{equation*}
-%
-The non-nullable types include records (consisting of an optional name and zero or more fields with
-their types) and primitive types. Records arising from XML documents are always named, while records
-used by JSON and CSV providers are always unnamed.
-
-Next, we have three numerical types -- \ident{int} for integers, \ident{decimal} for small high-precision 
-decimal numbers and \ident{float} for floating-point numbers (these are related by sub-typing relation as
-discussed in the next section). Furthermore, the \ident{bit} type is a type of values $0$ and $1$ and 
-makes it possible to infer Boolean in the CSV processing example discussed in Section~\ref{sec:providers-csv}.
-
-Any non-nullable type is also a nullable type, but it can be wrapped in the \kvd{option}
-constructor to explicitly permit the \kvd{null} value. These are typically mapped to the standard F\# option 
-type. A simple collection type $[\sigma]$ is also nullable and missing values or \kvd{null} are treated as 
-empty collection. The type $\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded but not
-ambiguous notation) and $\top$ represents the top type.
-
-Finally, a variant type \kvd{var} is annotated with the statically known types it may represent based
-on the sample document. As discussed earlier (Section~\ref{sec:providers-xml}), the variant type forces
-the user to handle the case when a value is of a different type than the ones statically known. For
-this reason, the variant type also implicitly permits the \kvd{null} value.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Subtyping relation}
-\label{sec:inference-subtyping}
-
-To provide a basic intuition, the subtyping relation between structural types is illustrated in
-Figure~\ref{fig:subtyping-diagram}. We split the diagram in two parts. The upper part shows 
-non-nullable types (with records and primitive types). The lower part shows nullable types with 
-\kvd{null}, collections and optional values. We omit links between the two part, but any type
-$\hat{\sigma}$ is a subtype of $\hat{\sigma}~\kvd{option}$ (in the diagram, we abbreviate 
-$\sigma~\kvd{option}$ as $\sigma?$).
-
-The diagram shows only the basic structure and it does not explain relationships between records
-with the same name and between union types. This following definition makes this precise.
-
-\begin{definition}
-We write $\sigma_1 :> \sigma_2$ to denote that $\sigma_2$ is a subtype of $\sigma_1$. The 
-subtyping relation is defined as a transitive reflexive closure of the following rules:
-
-\vspace{0.5em}
-\noindent\boxed{\emph{Primitives, options, collections}}
-\begin{align}
-\label{eq:sub-prim}\tag{B1}
-\ident{float}\,:>\,\ident{decimal}\,:>\,\ident{int}\,:>\,\ident{bit}\quad&\textnormal{and}\quad\ident{bool}\,:>\,\ident{bit}\\[-0.2em]
-\label{eq:sub-null}\tag{B2}
-\sigma :> \kvd{null} \quad &(\textnormal{iff}~\sigma \neq \hat{\sigma}) \\[-0.2em]
-\label{eq:sub-opt}\tag{B3}
-\hat{\sigma}~\kvd{option} :> \hat{\sigma} \quad &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
-\label{eq:sub-top}\tag{B4}
-\sigma :> \top \quad &(\textnormal{for all}~\sigma)\\[-0.2em]
-\label{eq:sub-col}\tag{B5}
-[\sigma_1] :> [\sigma_2] \quad &(\textnormal{if}~\sigma_1 :> \sigma_2)
-\end{align}
-
-\noindent\boxed{\emph{Variant types}}
-\begin{align}
-\label{eq:sub-var-bot}\tag{U1}
-\kvd{var}\langl \sigma_1, \ldots, \sigma_n\rangl :>&~ \sigma \\
-\label{eq:sub-var-eq}\tag{U2}
-\kvd{var}\langl \sigma_1, \ldots, \sigma_n \rangl :>&~ \kvd{var}\langl \sigma_1', \ldots, \sigma_n' \rangl
-\end{align}
-
-\noindent\boxed{\emph{Record types}}
-\begin{align}
-\hspace{-1em}
-\label{eq:sub-record1}\tag{R1}
- \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \hspace{-0.1em}:> \hspace{-0.1em}
- \nu_?~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
- \; (\sigma_i \hspace{-0.25em}:>\hspace{-0.25em} \sigma_i')
-\end{align}
-\vspace{-1.4em}
-\begin{align}
-\hspace{-1em}
-\label{eq:sub-record2}\tag{R2}
- \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \hspace{-0.1em} :> \hspace{-0.1em}
- \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_m\!:\!\sigma_m \}
- \; (m \hspace{-0.1em}\geq\hspace{-0.1em} n)
-\end{align}
-\vspace{-1.0em}
-\begin{align}
-\label{eq:sub-record3}\tag{R3}
-\hspace{-1.5em}
-\begin{array}{l}
- \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
- \nu_?~\{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, .., \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
-\end{array} \qquad &(\textnormal{permutation}~\pi)
-\end{align}
-\vspace{-0.6em}
-\begin{align}
-\label{eq:sub-record4}\tag{R4}
-\hspace{-1.5em}
-\begin{array}{l}
- \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null}, .., \nu_{n+m}\!:\!\kvd{null} \} :> \\
- \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
-\end{array}
-\end{align}
-\end{definition}
+              &\narrow{|}& \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \lsep ~\bot~ \lsep \kvd{null}
+\end{array}
+\end{equation*}
+%
+Non-nullable types include records (consisting of an optional name and fields with their types) 
+and primitive types. Records arising from XML documents are named, while records used by 
+JSON and CSV providers are unnamed.
+
+The three numerical types, \ident{int} for integers, \ident{decimal} for small precise decimal 
+numbers and \ident{float} for floating-point numbers, are related by sub-typing as discussed in 
+\S\ref{sec:inference-subtyping}.
+
+Any non-nullable type can be wrapped in the \kvd{option} constructor to explicitly permit the 
+\kvd{null} value. Type providers map \kvd{option} to standard F\# option type. A collection 
+type $[\sigma]$ is also nullable and missing values or \kvd{null} are treated as empty collection. 
+The type $\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded notation) and 
+$\bot$ is the bottom type.
+
+A variant type \kvd{any} is annotated with the types it may represent in the sample document.
+As discussed earlier (\S\ref{sec:providers-xml}), variants force the user to handle the case when 
+a value is of a different type than the statically known ones and so it can be seen as a family of 
+top types. For the same reason, the variants also implicitly permit the \kvd{null} value.
 
@@
 \begin{center}
-\includegraphics[scale=0.75,trim=5mm 5mm 5mm 5mm,clip]{images/hierarchy.pdf} % left bottom right top
+\includegraphics[scale=0.80,trim=5mm 5mm 5mm 5mm,clip]{images/hierarchy.pdf} % left bottom right top
 \end{center}
@@
 
+\subsection{Subtyping relation}
+\label{sec:inference-subtyping}
+
+Figure~\ref{fig:subtyping-diagram} provides a basic intuition about the subtyping between
+structural types. The upper part shows non-nullable types (with records and primitive types) and 
+the lower part shows nullable types with \kvd{null}, collections and optional values. We omit links 
+between the two part, but any type $\hat{\sigma}$ is a subtype of $\kvd{option}\langl\hat{\sigma}\rangl$ 
+and we abbreviate $\kvd{option}\langl\sigma\rangl$ as $\sigma?$.
+
+\begin{definition}
+We write $\sigma_1 :> \sigma_2$ to denote that $\sigma_2$ is a subtype of $\sigma_1$. The 
+subtyping relation is defined as a transitive reflexive closure of the following rules:
+
+\vspace{0.5em}
+\noindent\boxed{\emph{Primitives, options, collections}}
+\begin{align}
+\label{eq:sub-prim}\tag{B1}
+\ident{float}\,&:>\,\ident{decimal}\,:>\,\ident{int}&\\[-0.2em]
+\label{eq:sub-null}\tag{B2}
+\sigma &:> \kvd{null}  &(\textnormal{iff}~\sigma \neq \hat{\sigma}) \\[-0.2em]
+\label{eq:sub-bot}\tag{B3}
+\sigma &:> \bot  &(\textnormal{for all}~\sigma)\\[-0.2em]
+\label{eq:sub-opt}\tag{B4}
+\kvd{option}\langl\hat{\sigma}\rangl &:> \hat{\sigma}  &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
+\label{eq:sub-opt-cov}\tag{B5}
+\kvd{option}\langl\hat{\sigma_1}\rangl &:> 
+  \kvd{option}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} :> \hat{\sigma_2})\\[-0.2em]
+\label{eq:sub-col}\tag{B6}
+[\sigma_1] &:> [\sigma_2]  &(\textnormal{if}~\sigma_1 :> \sigma_2)
+\end{align}
+
+\noindent\boxed{\emph{Variant types}}
+\begin{align}
+\label{eq:sub-var-bot}\tag{U1}
+\kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl :>&~ \sigma \\
+\label{eq:sub-var-eq}\tag{U2}
+\kvd{any}\langl \sigma_1, \ldots, \sigma_n \rangl :>&~ \kvd{any}\langl \sigma_1', \ldots, \sigma_n' \rangl
+\end{align}
+
+\noindent\boxed{\emph{Record types}}
+\begin{align}
+\label{eq:sub-record1}\tag{R1}
+\begin{array}{l}
+ \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
+ \nu_?~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
+\end{array} \qquad ~~~~(\textnormal{if}~\sigma_i :> \sigma_i')
+\end{align}
+\vspace{-0.8em}
+\begin{align}
+\label{eq:sub-record2}\tag{R2}
+\begin{array}{l}
+ \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
+ \nu_?~\{ \nu_1\!:\!\sigma_1, .., .., \nu_m\!:\!\sigma_m \}
+\end{array} \qquad (\textnormal{when}~m \geq n)
+\end{align}
+\vspace{-0.8em}
+\begin{align}
+\label{eq:sub-record3}\tag{R3}
+\begin{array}{l}
+ \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
+ \nu_?~\{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, .., \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
+\end{array} ~~(\pi~\textnormal{perm.})
+\end{align}
+\vspace{-0.8em}
+\begin{align}
+\label{eq:sub-record4}\tag{R4}
+\begin{array}{l}
+ \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null} \} :> \\
+ \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
+\end{array}
+\end{align}
+\end{definition}
+
+% --------------------------------------------------------------------------------------------------
+
 \begin{figure*}[t]
 \noindent
+\begin{equation*}
+\boxed{
+\begin{array}{c}
+\begin{array}{rclcl}
+ \tytag &\narrow{=}& \ident{bool} \\
+        &\narrow{\lsep}& \ident{string} &\narrow{\lsep}& \ident{number}  \\
+        &\narrow{\lsep}& \ident{any}  &\narrow{\lsep}& \ident{collection} \\
+        &\narrow{\lsep}& \ident{record} &\narrow{\lsep}& \ident{named}\;~\nu 
+\end{array}
+%
+\quad\;\;\;
+%
+\begin{array}{rcl}
+ \tytagof(\ident{string}) &\narrow{=}& \ident{string}\\
+ \tytagof(\ident{bool}) &\narrow{=}& \ident{bool}\\
+ \tytagof([\sigma]) &\narrow{=}& \ident{collection}\\
+ \tytagof(\hat{\sigma}~\kvd{option}) &\narrow{=}& \tytagof(\hat{\sigma})
+\end{array}
+%
+\;\;
+%
+\begin{array}{rcl}
+ \tytagof(\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl) &\narrow{=}& \ident{any}\\
+ \tytagof(\{ \nu_1 : \sigma_1, \; \ldots \; , \nu_n : \sigma_n \}) &\narrow{=}& \ident{rec-anon}\\
+ \tytagof(\nu\; \{ \nu_1 : \sigma_1, \; \ldots \; , \nu_n : \sigma_n \}) &\narrow{=}& \ident{rec-named}~\nu \\
+ \multicolumn{3}{c}{ \tytagof(\sigma) = \ident{number}\quad \sigma \in\{ \ident{int},\ident{decimal},\ident{float} \}  } 
+\end{array}
+%
+\\[3em]
+%
+\begin{array}{rcll}
+ \addopt{\hat{\sigma}} &\narrow{=}& \hat{\sigma}~\kvd{option} &(\textnormal{non-nullable types})\\
+ \addopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
+\end{array}
+\qquad\qquad
+\begin{array}{rcll}
+ \dropopt{\hat{\sigma}~\kvd{option}} &\narrow{=}& \hat{\sigma} &(\textnormal{option})\\
+ \dropopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
+\end{array}
+\end{array}
+}
+\end{equation*}
+
 \begin{equation*}
@@
 \inference[(record-1)\;]
-  { (\nu_i = \nu'_j \Leftrightarrow (i = j) \wedge (i \leq k))
+  { (\nu_i = \nu'_j) \Leftrightarrow (i = j) \wedge (i \leq k)
       \qquad \forall i\in\{ 1 .. k \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
   { \begin{array}{l}
-    \nu? \; \{ \nu_1 : \sigma_1,  \; \ldots \;, \nu_k : \sigma_k, \; \ldots \;, \nu_n : \tau_n \} \tsep
-    \nu? \; \{ \nu'_1 : \sigma'_1, \; \ldots \;, \nu'_k : \sigma'_k, \; \ldots \;, \nu'_m : \tau'_m \} \vdash\\
-    \nu? \; \{ \nu_1 : \sigma''_1, \; \ldots \; , \nu_k : \sigma''_k, 
-                            \nu_{k+1} : \addopt{\sigma_{k+1}}, \ldots, \nu_n : \addopt{\sigma_n},
-                            \nu'_{k+1} : \addopt{\sigma'_{k+1}}, \ldots, \nu'_m : \addopt{\sigma'_m} \}
+    \nu_? \; \{ \nu_1 \!:\! \sigma_1,  \; \ldots \;, \nu_k \!:\! \sigma_k, \; \ldots \;, \nu_n \!:\! \tau_n \} \tsep
+    \nu_? \; \{ \nu'_1 \!:\! \sigma'_1, \; \ldots \;, \nu'_k \!:\! \sigma'_k, \; \ldots \;, \nu'_m \!:\! \tau'_m \} \vdash\\
+    \nu_? \; \{ \nu_1 \!:\! \sigma''_1, \; \ldots \; , \nu_k \!:\! \sigma''_k, 
+                            \nu_{k+1} \!:\! \addopt{\sigma_{k+1}}, \ldots, \nu_n \!:\! \addopt{\sigma_n},
+                            \nu'_{k+1} \!:\! \addopt{\sigma'_{k+1}}, \ldots, \nu'_m \!:\! \addopt{\sigma'_m} \}
     \end{array} }
 \end{equation*}
+\vspace{-1.75em}
 
@@
 \begin{equation*}
-\textnormal{\footnotesize{(union-1)}}\;\;
+\textnormal{\footnotesize{(var-1)}}\;\;
 \inference
-  {\exists i . \tytagof(\sigma_i) = \tytagof(\sigma) & \sigma \tsep \sigma_i \vdash \sigma_i' & \tytagof(\sigma)\neq\ident{union}}
-  {\sigma \tsep (\sigma_1 + \ldots + \sigma_n) \vdash (\sigma_1 + \ldots + \dropopt{\sigma_i'} + \ldots + \sigma_n)}
-\quad\quad
+  {\exists i . \tytagof(\sigma_i) = \tytagof(\sigma) & \sigma \tsep \sigma_i \vdash \sigma_i' & \tytagof(\sigma)\neq\ident{any}}
+  {\sigma \tsep \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \vdash 
+    \kvd{any}\langl\sigma_1, \ldots, \dropopt{\sigma_i'}, \ldots, \sigma_n\rangl}
+\;\;
 \inference
-  {\nexists i . \tytagof(\sigma_i) = \tytagof(\sigma) & \tytagof(\sigma)\neq\ident{union}}
-  {\sigma \tsep (\sigma_1 + \ldots + \sigma_n) \vdash (\sigma_1 + \ldots + \sigma_n + \dropopt{\sigma})}
-\end{equation*}
+  {\nexists i . \tytagof(\sigma_i) = \tytagof(\sigma) & \tytagof(\sigma)\neq\ident{any}}
+  {\sigma \tsep \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \vdash 
+    \kvd{any}\langl\sigma_1, \ldots, \sigma_n, \dropopt{\sigma}\rangl}
+\end{equation*}
+\vspace{-1.75em}
 
@@
 \begin{equation*}
-\inference[(union-2)\;]
-  { \tytagof(\sigma_i) = \tytagof(\sigma'_i) \quad 
-    \sigma_i \tsep \sigma_i' \vdash \sigma''_i \qquad (\forall i\in \{ 1 .. k \}) }
+\begin{array}{l}
+\inference[(var-2)\;]
+  { (\forall i\in \{ 1 .. k \})\quad \tytagof(\sigma_i) = \tytagof(\sigma'_i) \quad 
+    \sigma_i \tsep \sigma_i' \vdash \sigma''_i}
   { \begin{array}{l}
-    (\sigma_1  + \ldots + \sigma_k +  \ldots + \sigma_n) \tsep 
-    (\sigma'_1 + \ldots + \sigma'_k + \ldots + \sigma'_m) \vdash\\
-    (\sigma''_1 + \ldots + \sigma''_k + \sigma_{k+1} + \ldots + \sigma_{n} + \sigma'_{k+1} + \ldots + \sigma'_{m}) 
+    \kvd{any}\langl \sigma_1, \ldots, \sigma_k,  \ldots, \sigma_n\rangl \tsep 
+    \kvd{any}\langl \sigma'_1, \ldots, \sigma'_k, \ldots, \sigma'_m\rangl \vdash\\
+    \kvd{any}\langl \sigma''_1, \ldots, \sigma''_k, \sigma_{k+1}, \ldots, \sigma_{n}, \sigma'_{k+1}, \ldots, \sigma'_{m}\rangl
     \end{array} }
-%
-\qquad\quad
-\inference[(union-3)\;]
-  {(\textnormal{no other rule applies})}
-  {\sigma_1 \tsep \sigma_2 \vdash \dropopt{\sigma_1} + \dropopt{\sigma_2}}
-\end{equation*}
-
-\begin{equation*}
-\hspace{1em}
+\end{array}    
+\begin{array}{l}
+\inference[(var-3)\;]
+  {(\forall i\in\{1,2\}) & \tytagof(\sigma_1) \neq \tytagof(\sigma_2) \\ 
+   \tytagof(\sigma_i) \neq \ident{any} & \nexists\sigma_i' . (\sigma_i = \kvd{option}\langl\sigma_i'\rangl)  }
+  {\sigma_1 \tsep \sigma_2 \vdash \kvd{any}\langle\dropopt{\sigma_1}, \dropopt{\sigma_2}\rangle}
+\end{array}  
+\end{equation*}
+\vspace{-1.75em}
+
+\begin{equation*}
 \inference[(order-1)\;]
-  { \nu? \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \tsep \sigma \vdash \sigma' }
-  { \nu? \; \{ \nu_{\pi(1)} : \sigma_{\pi(1)}, \ldots, \nu_{\pi(n)} : \sigma_{\pi(n)} \} \tsep \sigma \vdash \sigma' }
+  { \nu_? \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \tsep \sigma \vdash \sigma' }
+  { \nu_? \; \{ \nu_{\pi(1)} : \sigma_{\pi(1)}, \ldots, \nu_{\pi(n)} : \sigma_{\pi(n)} \} \tsep \sigma \vdash \sigma' }
 \quad
 \inference[(order-2)\;]
-  { \sigma_1 + \ldots + \sigma_n \tsep \sigma \vdash \sigma' }
-  { \sigma_{\pi(1)} + \ldots + \sigma_{\pi(n)} \tsep \sigma \vdash \sigma' }
-\qquad (\pi~\textnormal{permutation})  
-\end{equation*}
+  { \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \tsep \sigma \vdash \sigma' }
+  { \kvd{any}\langl\sigma_{\pi(1)}, \ldots, \sigma_{\pi(n)}\rangl \tsep \sigma \vdash \sigma' }
+\quad (\pi~\textnormal{perm.})  
+\end{equation*}
+\vspace{-1.75em}
 
@@
 \hspace{3em}
-% collection types
+% option types
+\inference[(opt)\;]
+  {\hat{\sigma_1} \tsep \sigma_2 \vdash \sigma}
+  {\kvd{option}\langl\hat{\sigma_1}\rangl \tsep \sigma_2 \vdash \kvd{option}\langl\dropopt{\sigma}\rangl}
+\qquad
+% primitive types
+\inference[(prim)\;]
+  {\sigma_1 :> \sigma_2 &
+   \tytagof(\sigma_1) = \tytagof(\sigma_2) = \ident{number} }
+  {\sigma_1 \tsep \sigma_2 \vdash \sigma_1}\quad
+\end{equation*}
+\vspace{-2em}
+
+% all rules are symmetric and reflexive
+\begin{equation*}
 \inference[(list)\;]
@@
   {[\sigma_1] \tsep [\sigma_2] \vdash [\sigma]}
-\qquad\qquad
-% primitive types
-\inference[(prim)\;]
-  {\sigma_1 :> \sigma_2}
-  {\sigma_1 \tsep \sigma_2 \vdash \sigma_1}\quad
-\begin{array}{l}
-  (\textnormal{when}\,~\sigma_1, \sigma_2 \in \{\ident{bit}, \ident{int}, \ident{decimal}, \ident{float} \} \\
-   \quad~~\;\textnormal{or}\,~\sigma_1, \sigma_2 \in \{\ident{bit}, \ident{bool} \})
-\end{array}   
-\end{equation*}
-
-% all rules are symmetric and reflexive
-\noindent
-\begin{equation*}
-\hspace{4.75em}
+\quad
 \inference[(sym)\;]
@@
 %
-\qquad\quad
+\quad
 \begin{array}{l}
  \textnormal{\footnotesize{(refl)}}\;\; \sigma \tsep \sigma \vdash \sigma\\[0.6em]
- % top type
- \textnormal{\footnotesize{(top)}}\;\; \top \tsep \sigma \vdash \sigma
-\end{array}
-%
-\qquad\quad 
+ % bottom type
+ \textnormal{\footnotesize{(bot)}}\;\; \bot \tsep \sigma \vdash \sigma
+\end{array}
+%
+\quad
 % null and nullable
@@
 \begin{itemize}
-\item For numeric types (\ref{eq:sub-prim}), we infer a single most precise numeric type that 
-  can represent all values from a sample dataset, so we order types by the ranges they represent;
+\item Numeric types (\ref{eq:sub-prim}) ar ordered by the range of values they represent;
   \ident{int} is a 32-bit integer, \ident{decimal} has a range of about $-1^{29}$ to $1^{29}$ and 
-  \ident{float} is a double-precision floating-point.
-
-\item The \ident{bit} type is a subtype of both \ident{int} and \ident{bool} (\ref{eq:sub-prim}). 
-  Thus, a sample $0,1$ has a type \ident{bit}; $0,1,\kvd{true}$ is \ident{bool} and
-  $0,1,2$ becomes \ident{int} (for a sample $0,1,2, \kvd{true}$ we would need a variant type).
-
-\item The \kvd{null} type is a subtype of all nullable types (\ref{eq:sub-null}), that is all 
-  $\sigma$ types excluding non-nullable types $\hat{\sigma}$). Any non-nullable type is also a 
-  subtype of its optional version (\ref{eq:sub-opt}).
-
-\item There is a top type (\ref{eq:sub-top}) and variants can be seen as the bottom because
-  any type $\sigma$ is a subtype of a variant (\ref{eq:sub-var-bot}). There is no \emph{single} 
-  bottom type -- there is a range of distinct variant types, but any variant is a subtype
-  of any other variant (\ref{eq:sub-var-eq}).
+  \ident{float} is a floating-point number.
+
+\item The \kvd{null} type is a subtype of all nullable types (\ref{eq:sub-null}), i.e. all 
+  types excluding non-nullable types $\hat{\sigma}$. Any non-nullable type is a 
+  subtype of its optional version (\ref{eq:sub-opt}) and both options and collections are 
+  covariant (\ref{eq:sub-opt-cov}, \ref{eq:sub-col}).
+
+\item There is a bottom type (\ref{eq:sub-bot}) and variants behave as top types, because
+  any type $\sigma$ is a subtype of any variant (\ref{eq:sub-var-bot}). There is a range of
+  variants (no \emph{single} top type), but any variant is a subtype of any other variant 
+  (\ref{eq:sub-var-eq}).
 
@@
 \noindent
-If we were proposing a type system for type-checking programs, the use of multiple variant types
-(which all are subtypes of each other) would not add any value and we could introduce a single 
-bottom type instead. However, our purpose is different -- we aim to make reading structured data
-easier and, as shown in Section~\ref{sec:providers-xml}, the types forming a variant provide a
-useful information about the sample document.
-
-Another important rule is (\ref{eq:sub-record4}), which allows subtype to have fewer record 
-elements. In some cases, this allows us to infer records rather than variants. For example, given 
+If we were type-checking programs, defining variants as a family of top types (which are all 
+subtypes of each other) would not add any value over having a single top type. Our purpose 
+is different. We simplify reading structured data and, as shown in \S\ref{sec:providers-xml}, 
+the types forming a variant provide a valuable information about the sample document.
+
+A particularly important rule is (\ref{eq:sub-record4}), which allows subtype to have fewer record 
+elements. This allows us to prefer records over variants. For example, given 
 $\{ \ident{name}\!:\!\ident{string} \}$ and $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int} \}$,
-we can find a common supertype $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int}~\kvd{option} \}$
-which is also a record. This is more precise and more usable type than another common supertype
-$\kvd{var}\langl \{ \ident{name}:\ident{string} \}, \{ \ident{name}:\ident{string}, \ident{age}:\ident{int} \}\rangl$.
-The next section describes precisely how our inference algorithm works.
-
-It is worth noting that some of the aspects (such as 3 different numeric types and handling of 
-the \kvd{null} type) are specific to F\#. Similar problems will appear with most real-world
-languages and systems. JVM and OCaml all haves multiple numeric types, but they would handle \kvd{null}
-differently.
-
-We could also be more or less strict about missing data -- we choose to handle missing values 
-silently when we can (\kvd{null} becomes an empty collection), but we are explicit in other cases 
-(we infer \ident{string}~\kvd{option} as a hint to the user rather than treating \kvd{null} as an 
-empty string). These choices are based on experience with using F\# Data in practice
-and might differ for other languages.
+we find a common supertype $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int}~\kvd{option} \}$.
+This is a least upper bound (clarified below) and more usable than another common supertype 
+$\kvd{var}\langl \{ \ident{name}\!:\!\ident{string} \}, \{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int} \}\rangl$.
+
+Some of the aspects of our system (numeric types, handling of \kvd{null} and treatment of
+missing data) are based on common F\# practices and might differ for other languages. For example, 
+OCaml also has multiple numeric types, but would likely always handle \kvd{null} explicitly.
 
@@
 \begin{array}{l}
- \ident{asInt}(i) \,\reduce i\\
- \ident{asInt}(d) \reduce i\quad(i = \lfloor d \rfloor)\\
- \ident{asInt}(f) \reduce i\quad(i = \lfloor f \rfloor) \\
- \ident{asInt}(t) \,\reduce i\quad(t~\text{\small represents}~i) \\
- \ident{asInt}(\kvd{true}) \reduce 1\\
- \ident{asInt}(\kvd{false}) \reduce 0\\
-\end{array}
-\;
+\ident{getField}(\nu,\nu_i, \nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
+\ident{getField}(\bullet, \nu_i, \{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
+\ident{getField}(\nu,\nu', \nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )\\
+\ident{getField}(\bullet, \nu', \{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )
+\\[0.5em]
+\hspace{-0.5em}
 \begin{array}{l}
 \ident{asDec}(i) \,\reduce d\quad(d = i)\\
- \ident{asDec}(d) \reduce d\\
- \ident{asDec}(f) \reduce d\quad(d = \lfloor f \rfloor) \\
- \ident{asDec}(t) \,\reduce d\quad(t~\text{\small represents}~d) \\
- \ident{asDec}(\kvd{true}) \reduce 1.0\\
- \ident{asDec}(\kvd{false}) \reduce 0.0\\
-\end{array}
-\;
-\begin{array}{l}
+\ident{asDec}(d) \reduce d
+\\[0.5em]
+\ident{isNull}(\kvd{null}) \reduce \kvd{true} \\
+\ident{isNull}(\_) \reduce \kvd{false} 
+\end{array}
+\begin{array}{l}
+\\
 \ident{asFloat}(i) \,\reduce f\quad(f = i)\\
@@
 \ident{asFloat}(f) \reduce f\\
- \ident{asFloat}(t) \,\reduce f\quad(t~\text{\small represents}~f) \\
- \ident{asFloat}(\kvd{true}) \reduce 1.0\\
- \ident{asFloat}(\kvd{false}) \reduce 0.0\\
-\end{array}
-\;
-\begin{array}{l}
- \ident{asStr}(i) \,\reduce t\quad(t~\text{\small represents}~i)\\
- \ident{asStr}(d) \reduce t\quad(t~\text{\small represents}~d)\\
- \ident{asStr}(f) \reduce t\quad(t~\text{\small represents}~f)\\
- \ident{asStr}(t) \,\reduce t\\
- \ident{asStr}(\kvd{true}) \reduce \str{true}\\
- \ident{asStr}(\kvd{false}) \reduce \str{false}\\
-\end{array}
-\end{equation*}
-%
-\begin{equation*}
-\hspace{-1.5em}
-\begin{array}{l}
- \ident{asBool}(1) \reduce \kvd{true}\\
- \ident{asBool}(1.0) \reduce \kvd{true}\\
- \ident{asBool}(\str{true}) \reduce \kvd{true}\\
- \ident{asBool}(\str{1}) \reduce \kvd{true}
- \\[0.6em]
- \ident{isNull}(\kvd{null}) \reduce \kvd{true} \\
- \ident{isNull}(\_) \reduce \kvd{false} 
-\end{array}
-\hspace{2.6em}
-\begin{array}{l}
- \ident{asBool}(0) \reduce \kvd{false}\\
- \ident{asBool}(0.0) \reduce \kvd{false}\\
- \ident{asBool}(\str{false}) \reduce \kvd{false}\\
- \ident{asBool}(\str{0}) \reduce \kvd{false}\\
- \ident{asBool}(b) \reduce b
- \\[0.6em]
+\end{array}
+\end{array}
+~
+\begin{array}{l}
+\ident{isTag}(\ident{string}, t) \reduce \kvd{true} \\
+\ident{isTag}(\ident{bool}, s) \reduce \kvd{true} \quad(\textnormal{when}~s\in{\kvd{true},\kvd{false}} )\\
+\ident{isTag}(\ident{number}, s) \reduce \kvd{true} \quad(\textnormal{when}~s=i, s=d ~\textnormal{or}~ s=f) \\
+\ident{isTag}(\ident{collection}, [s_1; \ldots; s_n]) \reduce \kvd{true} \\
+\ident{isTag}(\ident{collection}, \kvd{null}) \reduce \kvd{true} \\
+\ident{isTag}(\ident{record}, \{ \nu_1\mapsto s_1, \ldots, \nu_n\mapsto s_n \}) \reduce \kvd{true} \\
+\ident{isTag}(\ident{named}~\nu, \nu\;\{ \nu_1\mapsto s_1, \ldots, \nu_n\mapsto s_n \}) \reduce \kvd{true} \\
+\ident{isTag}(\_, \_) \reduce \kvd{false} \\
+\end{array}
+\end{equation*}
+\vspace{-1em}
+\begin{equation*}
+ \hspace{5.6em}
+ \ident{getChildren}([s_1; \ldots; s_n]) \reduce [s_1; \ldots; s_n]  \qquad\quad
  \ident{getChildren}(\kvd{null}) \reduce [~] 
-\end{array}
-\hspace{3.0em}
-\begin{array}{l}
- \ident{getField}(\nu,\nu_i, \nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
- \ident{getField}(\bot, \nu_i, \{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
- \ident{getField}(\nu,\nu', \nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )\\
- \ident{getField}(\bot, \nu', \{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )
- \\[0.6em]
- \ident{getChildren}([s_1; \ldots; s_n]) \reduce [s_1; \ldots; s_n]
-\end{array}
-\end{equation*}
-%
-\begin{equation*}
-\hspace{-1.5em}
-\begin{array}{l}
- \ident{isTag}(\ident{string}, t) \reduce \kvd{true} \\
- \ident{isTag}(\ident{bool}, v) \reduce \kvd{true} \quad(\textnormal{when}~v\in{0,1,0.0, 1.0, \kvd{true},\kvd{false}} )\\
- \ident{isTag}(\ident{number}, v) \reduce \kvd{true} \quad(\textnormal{when}~v=i, v=d, v=f) \\
- \ident{isTag}(\ident{collection}, [s_1; \ldots; s_n]) \reduce \kvd{true} \\
-\end{array}
-\hspace{1.8em}
-\begin{array}{l}
- \ident{isTag}(\ident{collection}, \kvd{null}) \reduce \kvd{true} \\
- \ident{isTag}(\ident{rec-anon}, \{ \nu_1\mapsto s_1, \ldots, \nu_n\mapsto s_n \}) \reduce \kvd{true} \\
- \ident{isTag}(\ident{rec-named}~\nu, \nu\;\{ \nu_1\mapsto s_1, \ldots, \nu_n\mapsto s_n \}) \reduce \kvd{true} \\
- \ident{isTag}(\_, \_) \reduce \kvd{false} \\
-\end{array}
 \end{equation*}
@@
 
-As mentioned earlier, the structural type inference relies on a finding common supertype. However,
-the partially ordered set of types does not have a unique greatest lower bound. For example, record types
-$\{\ident{a}\!:\!\ident{int}\}$ and $\{\ident{b}\!:\!\ident{bool}\}$ have common supertypes
-$\{\ident{?a}\!:\!\ident{int}, \ident{?b}\!:\!\ident{bool}\}$ and 
-$\kvd{var}\langl\{\ident{a}\!:\!\ident{int}\}, \{\ident{b}:\ident{bool}\}\rangl$
-
-NOT TRUE
-which are unrelated.
-
- Our inference algorithm always prefers records over unions (Theorem~\ref{thm:no-unions}). 
-The definition of the \emph{common supertype} relation uses a number of auxiliary definitions 
-that are explained in the discussion that follows.
+Given two structural types, the \emph{common supertype} relation finds a least upper bound
+(Theorem~\ref{thm:lub}). When possible, it avoid variant types and prefers records 
+(Corollary~\ref{thm:no-unions}), which is important for usability as discussed earlier
+(\S\ref{sec:providers-json}).
 
@@
 \noindent
-When finding a common supertype of two records (\emph{record-1}), we return a record type that has the union
-of fields of the two arguments. We assume that the names of the first $k$ fields are the same and the
-remaining fields have different names. To get a record with the right order of elements, we provide the
-(\emph{order}) rule. The types of shared fields become common supertypes of their 
-respective types (recursively). Fields that are present in only one record are marked as optional using
-the following helper definition:
-%
-\begin{equation*}
-\begin{array}{rcll}
- \addopt{\hat{\sigma}} &\narrow{=}& \hat{\sigma}~\kvd{option} \qquad&(\textnormal{non-nullable types})\\
- \addopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
-\end{array}
-\end{equation*}
-%
-Finding a common supertype of when one type is a union is more complicated. Our definition aims to 
-use a flat structure (avoid nesting unions) and limit the number of cases. This is done by grouping
-types that have a common supertype which is not a union type. For example, rather than inferring 
-$\ident{int} + (\ident{bool} + \ident{decimal})$, our algorithm will find the common supertype of
-$\ident{int}$ and $\ident{decimal}$ (which is \ident{decimal}) and it will produce just $\ident{decimal}+\ident{bool}$.
-
-To identify which types have a common supertype which is not a union, we group the types by a 
-\emph{tag}, which is defined as:
-%
-\begin{equation*}
-\begin{array}{rcl}
- \tytag &\narrow{=}&  \ident{string} \lsep \ident{bool} \lsep \ident{number} \lsep \ident{union} \lsep \ident{collection} \\
-        &\narrow{\lsep}& \ident{rec-anon} \lsep \ident{rec-named}\;~\nu 
-\end{array}
-\end{equation*}
-%
-The tag of a type is obtained using a function $\tytagof{(-)} : \sigma \rightarrow \tytag$:
-%
-\begin{equation*}
-\begin{array}{rcl}
- \tytagof(\ident{string}) &\narrow{=}& \ident{string}\\
- \tytagof(\ident{bool}) &\narrow{=}& \ident{bool}\\
- \tytagof(\ident{decimal}) = \tytagof(\ident{float}) &\narrow{=}& \ident{number}\\
- \tytagof(\ident{int}) = \tytagof(\ident{bit}) &\narrow{=}& \ident{number}\\
- \tytagof([\sigma]) &\narrow{=}& \ident{collection}\\
-\end{array}
-\end{equation*}
-\begin{equation*}
-\begin{array}{rcl}
- \tytagof(\sigma + \ldots + \sigma) &\narrow{=}& \ident{union}\\
- \tytagof(\{ \nu_1 : \sigma_1, \; \ldots \; , \nu_n : \sigma_n \}) &\narrow{=}& \ident{rec-anon}\\
- \tytagof(\nu\; \{ \nu_1 : \sigma_1, \; \ldots \; , \nu_n : \sigma_n \}) &\narrow{=}& \ident{rec-named}~\nu \\
- \tytagof(\hat{\sigma}~\kvd{option}) &\narrow{=}& \tytagof(\hat{\sigma})
-\end{array}
-\end{equation*}
-%
-The function is undefined for the $\top$ and $\kvd{null}$ types, but this is not a 
-problem because these types never need to be used as arguments in Figure~\ref{fig:subtyping-cst}.
-The $\top$ type is always eliminated using the (\emph{top}) rule and the \kvd{null} type
-is eliminated using either (\emph{null-1}) or (\emph{null-2}).
-
-The handling of unions is specified using three rules. When adding a non-union type to a 
-union (\emph{union-1}), the union may or may not contain a case with the same tag as the new type.
-If it does, the new type is combined with the existing one, otherwise a new case is added.
-When combining two unions (\emph{union-2}), we group the cases that have a shared tags.
-Finally, the last rule (\emph{union-3}) covers the case when we are combining two non-union types.
-As discussed earlier, union implicitly permits \kvd{null} values and so we use the following 
-auxiliary function which makes nullable types non-nullable (when possible) to simplify the type:
-%
-\begin{equation*}
-\begin{array}{rcll}
- \dropopt{\hat{\sigma}~\kvd{option}} &\narrow{=}& \hat{\sigma} &(\textnormal{option})\\
- \dropopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
-\end{array}
-\end{equation*}
-%
-The remaining rules are straightforward. For collections, we find the common supertype of 
-its elements (\emph{list}); for compatible primitive types, we choose one of the two (\emph{prim})
+When finding a common supertype of records (\emph{record-1}), we return a record that has the 
+union of their fields. The types of shared fields become common supertypes of their respective 
+types while fields that are present in only one record are marked as optional. The (\emph{order-1})
+rule lets us reorder fields.
+
+Although all variants are mutual subtypes, we find one that best represents the sample. We avoid
+nested variants and limit the number of cases. This is done by grouping types 
+that have a common supertype which is not a variant. For example, rather than inferring 
+$\kvd{any}\langl\ident{int}, \kvd{any}\langl\ident{bool}, \ident{decimal}\rangl\rangl$, 
+our algorithm finds the common supertype of $\ident{int}$ and $\ident{decimal}$ and produces 
+$\kvd{any}\langl\ident{decimal},\ident{bool}\rangl$.
+
+To identify types that have a common supertype which is not a variant, we group the types by
+a tag. The tag of a type is obtained using a function $\tytagof{(-)} : \sigma \rightarrow \tytag$.
+The function is not defined for $\bot$ and $\kvd{null}$ as those are hanled by the 
+(\emph{bot}) and (\emph{null-1}) or (\emph{null-2}) rules, respectively.
+
+The handling of variants is specified using three rules. When combining a non-variant and a
+variant (\emph{var-1}), the variant may or may not already contain a case with the tag of the other
+type. If it does, the two types are combined, otherwise a new case is added.
+When combining two variants (\emph{var-2}), we group the cases that have a shared tags.
+Finally, (\emph{var-3}) covers the case when we are combining two distinct non-variant types.
+As unions implicitly permit \kvd{null} values, we use an auxiliary function $\dropopt{-}$ to
+make nullable types non-nullable (when possible) to simplify the type.
+
+The remaining rules are straightforward. For collections and options, we find the common supertype 
+of the contained value(s); for compatible primitive types, we choose their supertype (\emph{prim})
 and a common supertype with \kvd{null} is either the type itself (\emph{null-1}) or an option 
@@
 \paragraph{Properties.}
-The common supertype relation finds a single supertype (among multiple candidates).
-The following theorems specify that the found type is uniquely defined (up to reordering of 
-fields in records and union cases) and is, indeed, a common supertype.
-%
-\begin{theorem}[Uniqueness]
+
+We say that two types are equivalent if they are mutual subtypes. The partially ordered set of 
+types does not have a \emph{unique} least upper bound, but it does have a least upper bound
+with respect to this equivalence. The common supertype relation is a function 
+(Theorem~\ref{thm:func}) and finds the least upper bound (Theorem~\ref{thm:lub}).
+
+\begin{definition}
+Types $\sigma_1$, $\sigma_2$ are \emph{equivalent}, written $\sigma_1 \equiv \sigma_2$ iff
+they are mutual subtypes, i.e. $\sigma_1 <: \sigma_2 \wedge \sigma_1 :> \sigma_2$.
+An equivalence class of a type $\sigma$ is $\mathcal{E}(\sigma) = \{ \sigma' \;|\; \sigma\equiv\sigma' \}$.
+\end{definition}
+
+\noindent
+The equivalence class $\mathcal{E}$ sheds light on the structure of structural types. It defines
+a lattice with bottom $\{ \bot \}$ and top (set of all variant types). It also joins records with
+reordered fields (of same names and types) and types that contain other equivalent types (e.g.~a
+records with variants as fields).
+
+\begin{theorem}[Function]
+\label{thm:func}
 \raggedright
-If $\sigma_1 \triangledown \sigma_2 \vdash \sigma$ and $\sigma_1 \triangledown \sigma_2 \vdash \sigma'$
+For all $\sigma_1$ and $\sigma_2$ there exists exactly one $\sigma$ such that 
+$\sigma_1 \triangledown \sigma_2 \vdash \sigma$.
+Furthermore, if $\sigma_1 \triangledown \sigma_2 \vdash \sigma'$
 then it holds that $\sigma :> \sigma'$ and $\sigma' :> \sigma$.
@@
 \begin{proof}
-The assumptions and shapes of the types to be unified in the rules in Figure~\ref{fig:subtyping-cst} 
-are disjoint, with the exception of (\emph{order}), (\emph{sym}) and (\emph{refl}). These rules 
-always produce types that are subtypes of each other. This property is preserved by rules that use the
-common supertype relation recursively.
+The pre-conditions of rules in Figure~\ref{fig:subtyping-cst} are disjoint, with the exception 
+of (\emph{order}), (\emph{sym}) and (\emph{refl}). These rules produce types that are subtypes
+of each other and this property is preserved by rules that use the common supertype relation 
+recursively.
 \end{proof}
 
-\begin{theorem}[Common supertype]
+\begin{theorem}[Least upper bound]
+\label{thm:lub}
 \raggedright
-If $\sigma_1 \triangledown \sigma_2 \vdash \sigma$ then it holds that $\sigma :> \sigma_1$ and $\sigma :> \sigma_2$.
+If $\sigma_1 \triangledown \sigma_2 \vdash \sigma$ then $\sigma$ is a least upper bound, i.e. 
+$\sigma :> \sigma_1$ and $\sigma :> \sigma_2$ and for all $\sigma'$ such that $\sigma' :> \sigma_1$
+and $\sigma' :> \sigma_2$, it holds that $\sigma' :> \sigma$.
 \end{theorem}
 \begin{proof}
-By induction over the common supertype derivation $\vdash$.
+By induction over $\vdash$. Note that the algorithm never produces nested variant or nested
+option. When one type is variant, the only common supertype is also a variant (\emph{var-2}); 
+for (\emph{var-3}) no other common supertype exists because the two types have distinct 
+tags and are not options.
 \end{proof}
 
-\noindent
-We stated earlier that the common supertype relation minimises the use of union types 
-(by preferring common numeric types or common record types when possible). This property
-can be stated and proved formally:
-%
-\begin{theorem}
+\begin{corollary}
 \label{thm:no-unions}
-Given $\sigma_1, \sigma_2$ if there is s $\sigma$ such that $\sigma :> \sigma_1$ and 
-$\sigma :> \sigma_2$ and $\sigma$ is not a union type, then $\sigma_1 \tsep \sigma_2 \vdash \sigma'$ 
-and $\sigma'$ is not a union type.
-\end{theorem}
+Given $\sigma_1, \sigma_2, \sigma$ such that $\sigma :> \sigma_1$ and $\sigma :> \sigma_2$ and 
+$\sigma$ is not a variant, then $\sigma_1 \tsep \sigma_2 \vdash \sigma'$ 
+and $\sigma' \equiv \sigma$.
+\end{corollary}
 \begin{proof}
-The only rule that introduces an union type is (\emph{union-3}), which is used only when 
-no other inference rule can be applied.
+Consequence of Theorem~\ref{thm:lub} and the fact that the set of all variants is the
+top element of $\mathcal{E}$.
 \end{proof}
 
+% ==================================================================================================
+% 
+%    #######
+%    #        ####  #####  #    #   ##   #      # ######   ##   ##### #  ####  #    #
+%    #       #    # #    # ##  ##  #  #  #      #     #   #  #    #   # #    # ##   #
+%    #####   #    # #    # # ## # #    # #      #    #   #    #   #   # #    # # #  #
+%    #       #    # #####  #    # ###### #      #   #    ######   #   # #    # #  # #
+%    #       #    # #   #  #    # #    # #      #  #     #    #   #   # #    # #   ##
+%    #        ####  #    # #    # #    # ###### # ###### #    #   #   #  ####  #    #
+%
 % ==================================================================================================
@@
 
-In this section, we build the necessary theoretical framework for proving the relativised type safety 
-property of F\# Data type providers in Section~\ref{sec:safety}. We start by discussing the runtime 
-representation of structural values and runtime conversions (Section~\ref{sec:formal-convert}). Then 
-we embed these into a formal model of an F\# subset that is necessary for discussing type providers 
-(Section~\ref{sec:formal-ff}) and we describe how F\# Data type providers turn an inferred 
-structural type into actual F\# code (Section~\ref{sec:formal-tp}).
+In this section, we build the necessary theoretical framework for proving relativised type safety 
+of our type providers (\S\ref{sec:safety}). We start by discussing the runtime representation of 
+structural values and runtime conversions (\S\ref{sec:formal-convert}), we embed these into a 
+formal model of an F\# subset (\S\ref{sec:formal-ff}) and we describe how our type providers 
+turn an inferred structural type into actual F\# code (\S\ref{sec:formal-tp}).
 
@@
 
-In the model presented here, we represent JSON, XML and CSV documents using the same 
-\emph{structural value}\footnote{Here, we diverge from the actual implementation in F\# Data which 
-uses a different implementation for each format (reusing existing libraries).}. Structural values are 
-first-order and can be one of the primitive values (Boolean \kvd{true} or \kvd{false}, string $t$, 
-integer $i$, decimal $d$ and floating-point number $f$), a missing value (\kvd{null}), a collection 
-and a record (named or unnamed):
+We represent JSON, XML and CSV documents using the same \emph{structural value}\footnote{Here, we 
+diverge from the actual implementation in F\# Data which uses a different implementation for each 
+format (reusing existing libraries).}. Structural values are first-order and can be one of the 
+following cases:
 %
@@
  s &\narrow{=}& i \lsep d \lsep f \lsep t \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
-    &\narrow{|}& [s_1; \ldots; s_n] \\[0.1em]
-    &\narrow{|}& \{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \}\\[0.1em]
-    &\narrow{|}& \nu~\{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \} \\
-\end{array}
-\end{equation*}
-
-\noindent
+   &\narrow{|}& [s_1; \ldots; s_n] \lsep \nu_?\{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \}
+\end{array}
+\end{equation*}
+%
 The first few cases represent primitive values ($i$ for integers, $d$ for decimals, $f$ for floating
-point numbers and $t$ for strings). A collection is written as a list of values in square brackets, 
-separated by semicolons. A record can optionally start with its name $\nu$, followed by a sequence of 
-field assignments $\nu_i \mapsto s_i$ written in curly brackets.
-
-As suggested by subtyping, our system permits certain runtime conversions (\emph{e.g.}~a numerical 
-value \num{1} can be treated as \kvd{true}). These are captured by the following operations:
+point numbers and $t$ for strings) and the missing value \kvd{null}. A collection is written as a 
+list of values in square brackets, separated by semicolons. A record starts with an optional name 
+$\nu_?$, followed by a sequence of field assignments $\nu_i \mapsto s_i$ written in curly brackets.
+
+As indicated by the subtyping, our system permits certain runtime conversions (e.g.~an integer
+\num{1} can be treated as a floating-point \num{1.0}). The following primitive operations 
+implement the conversions and other helpers:
 %
@@
 \begin{array}{lcl}
- op &\narrow{=}& \ident{asInt}(s) \lsep \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{asBool}(s)  \\
-    &\narrow{|}& \ident{asStr}(s) \lsep \ident{getChildren}(s) \lsep \ident{getField}(\nu?, \nu, s)\\
-    &\narrow{|}& \ident{isNull}(s) \lsep \ident{isTag}(\ident{tag}, s)
-\end{array}
-\end{equation*}
-%
-The reduction rules for these operations are defined in Figure~\ref{fig:op-conversions}. It is important
-to note that the relativised property we prove about F\# Data specifies \emph{sufficient}, but
-not \emph{necessary} conditions. That is, a type provided based on a sample is guaranteed to work 
-correctly on certain inputs, but may also handle several additional inputs.
-
-The conversion functions follow this approach. They perform conversions required by the subtyping relation
-(\emph{e.g.}~integer is converted to decimal or a float) but also attempt additional conversions.
-
-\begin{itemize}
-\item \textbf{Widening operations.} The widening operations are required by the subtyping. These include
-  converting to a wider numerical type (integer to decimal and float, decimal to float, converting 0 and 
-  1 to Boolean).
-
-\item \textbf{Conversions.} Additional conversions are not strictly required, but prove useful in practice.
-  These include converting to a smaller numerical type (such as float to integer) written as $i=\lfloor f\rfloor$,
-  which is not defined in case of arithmetic overflow; conversion to and from string values written as
-  ``$t$~represents~$i$'', which parses or formats a non-string value and also treating of decimals 
-  0.0 and 1.0 as Booleans.
-
-\item \textbf{Structural operations.} Aside from primitives, the \ident{getChildren} operation returns treats
-  \kvd{null} as an empty collection. The \ident{getField} function is defined separately for named and unnamed
-  records (we write $\bot$ for a name of an unnamed record). Accessing a field of a named record checks that 
-  the expected record name matches the actual.
-
-\item \textbf{Helper functions.} The semantics also defines two helper functions. The \ident{isNull} function
-  is used to check whether value is \kvd{null} and \ident{isTag} is used to test whether a value can be treated 
-  as a value of type associated with the specified tag (as defined in Section~\ref{sec:inference-commonsuper}).
+ op  &\narrow{=}& \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{getChildren}(s) \\
+     &\narrow{|}& \ident{getField}(\nu?, \nu, s) \lsep \ident{isNull}(s) \lsep \ident{isTag}(\ident{tag}, s)
+\end{array}
+\end{equation*}
+%
+The type providers generate code that uses these operations at runtime. Their behavior is defined
+by the reduction rules in Figure~\ref{fig:op-conversions}. The conversion functions (\ident{asDec}
+and \ident{asFloat}) only perform conversions required by the subtyping relation. As noted later
+(\S\ref{?}), our actual implementation is more lax.
+
+The \ident{getField} operation is used to access a field of a record. The operation ensures that
+the actual record name matches the expected name (we write $\bullet$ for the name of an unnamed 
+record). The \ident{getChildren} operation will used to turn \kvd{null} into an empty collction.
+
+Finally, we also define two helper functions. The \ident{isNull} operation is used to check whether 
+value is \kvd{null} and \ident{isTag} is used to test whether a value can be treated as a value of 
+a type associated with the specified tag (as defined in Figure~\ref{fig:subtyping-cst}).
 The last line defines a ``catch all'' pattern, which returns \kvd{false} for all remaining cases.
-\end{itemize}
 
@@
  { (\kvd{new}~C(\overline{v})).N_i \reduce e_i[\overline{x} \leftarrow \overline{v}] }\\
- \\
- \textnormal{\footnotesize{(ctx)}}&
- \hspace{-0.4em}
-  E[e] \reduce E[e'] \qquad\qquad(\textnormal{when}~e \reduce e')\\
  \\
@@
  \ident{List.map}~(\lambda x.e)~[v_1; \ldots] ~\reduce~ [e[x\leftarrow v_1]; \ldots] \\
+ \\
+ \textnormal{\footnotesize{(ctx)}}&
+ \hspace{-0.4em}
+  E[e] \reduce E[e'] \qquad\qquad(\textnormal{when}~e \reduce e')\\
 \end{array}
@@
 
-The semantics fragment introduced in the previous section discusses values and operations that are used by
-F\# Data. This section adds a minimal subset of the standard F\# language.
-
-We focus on the object system (which is what F\# type providers produce) and so the following combines 
-standard ML \cite{sml} features with Featherweight Java \cite{fwjava} classes. However, we only need classes with properties and
-without inheritance. A class has a single implicit constructor and the declaration closes over 
-constructor parameters. To avoid including all of ML, we only select constructs for working with options and lists
-that we need later.
-%
+The semantics fragment in the previous section discusses values and operations that are used by
+F\# Data. This section adds a minimal subset of F\#. Type providers provide classes and so we focus
+on the F\# object model and combine aspects of standard ML \cite{sml} and Featherweight Java \cite{fwjava}.
+
+We only need classes with parameter-less members and without inheritance. A class has a single 
+implicit constructor and the declaration closes over constructor parameters. To avoid including all 
+of ML, we only pick constructs for working with options and lists that we need later.
+
 \begin{equation*}
@@
  v &\narrow{=}& s \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \lsep [v_1; \ldots; v_n] \\[0.0em]
- e &\narrow{=}& op \lsep e.N \lsep \kvd{new}~C(\overline{e}) \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3}\\
-   &\narrow{|}& \ident{Some}(e)\lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
-   &\narrow{|}& \kvd{match}~e~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 \\
-   &\narrow{|}& [e_1;\ldots;e_n]\lsep \ident{List.map}~(\lambda x\rightarrow e_1)~e_2
-\end{array}
-\end{equation*}
-
-\noindent
-In the above definition, $\tau$ denotes types including a type of all structural values $s$ called
-\ident{StructVal}. A class definition $L$ consists of a constructor and zero or more members. Values $v$ include 
-previously defined structural values $s$ and values for the option and list type; finally expressions $e$ include 
-previously defined operations $op$, class construction, member access, conditionals and expressions for working 
-with option values and lists. We include \ident{List.map} as a special construct to avoid making the language too complex.
+ e &\narrow{=}& s \lsep op \lsep e.N \lsep \kvd{new}~C(\overline{e}) \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3}\\
+   &\narrow{|}& \ident{None} \lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
+   &\narrow{|}& \ident{Some}(e) \lsep [e_1;\ldots;e_n]\lsep \ident{List.map}~(\lambda x\rightarrow e_1)~e_2\\
+   &\narrow{|}& \kvd{match}~e~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2
+\end{array}
+\end{equation*}
+
+\noindent
+The type \ident{StructVal} is a type of all structural values $s$. A class
+definition $L$ consists of a constructor and zero or more members. Values $v$ include 
+previously defined structural values $s$ and values for the option and list type; finally 
+expressions $e$ include previously defined operations $op$, class construction, member access, 
+conditionals and expressions for working with option values and lists. We include 
+\ident{List.map} as a special construct to avoid making the language too complex.
 
@@
 The language presented here is intentionally incomplete. We only define parts needed to prove
-the relativized safety property in Section~\ref{sec:safety}.
-
-\paragraph{Reduction.} The reduction relation is of the form $e \reduce e'$. We also write 
+the relativized safety property (\S\ref{sec:safety}).
+
+\paragraph{Reduction.}
+The reduction relation is of the form $e \reduce e'$. We also write 
 $e \reduce^{*} e'$ to denote the reflexive and transitive closure of $\reduce$. The reduction rules
-for primitive functions $op$ were discussed earlier in Figure~\ref{fig:op-conversions}. The rules 
-for other expressions defined above are provided in Figure~\ref{fig:ff-reduction}.
+for operations $op$ were discussed earlier. Figure~\ref{fig:ff-reduction} shows the remaining
+rules.
 
 The (\emph{ctx}) rule performs a reduction inside a sub-expression specified by an evaluation context.
-This models the eager evaluation order of F\#. An evaluation context $E$ is defined as:
+This models the eager evaluation of F\#. An evaluation context $E$ is defined as:
 
@@
 \begin{equation*}
+\begin{array}{l}
+\hspace{-0.5em}
 \begin{array}{rcl}
- E &\narrow{=}& f(E) \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e}) \lsep \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2 \\[0.1em]
-   &\narrow{|}& \ident{Some}(E) \lsep \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
-   &\narrow{|}& [\overline{v};E;\overline{e}] \lsep \kvd{match}~E~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2 \\[0.1em]
-   &\narrow{|}& \ident{map}~(\lambda x\rightarrow e)~E \lsep [\cdot] 
- \\[0.6em]
- f &\narrow{\in}& \{\; \ident{asInt}, \ident{asDec}, \ident{asFloat}, \ident{asBool}, \ident{asStr}, \\[0.1em]
-              && ~~\; \ident{getField}, \ident{getChildren}, \ident{isNull}, \ident{isTag} \;\}
+ E &\narrow{=}& [\overline{v};E;\overline{e}] \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e}) \lsep \ident{Some}(E) \\[0.1em]
+   &\narrow{|}& \ident{map}~(\lambda x\rightarrow e)~E \lsep \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2 \lsep f(E)\\[0.1em]
+   &\narrow{|}& \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
+   &\narrow{|}& \kvd{match}~E~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2
+\end{array} \\[2.5em]
+f \in \{ \ident{asDec}, \ident{asFloat}, \ident{getField}, \ident{getChildren}, \ident{isNull}, \ident{isTag} \}
 \end{array}
@@
 values $\overline{v}$ in expression. The (\emph{member}) rule reduces a member access using a class 
-definition in the assumption to obtain the body of the expression. Finally, the remaining six rules
-provide standard reductions for conditionals and pattern matching.
-
-The language is sufficient for our purpose, but simple. It does not provide recursion
-and so all expressions reduce to a value in a finite number of steps or get stuck due to an error
-condition. An error condition can be a wrong argument passed to conditional, pattern matching or
-one of the conversion functions from Figure~\ref{fig:op-conversions}.
+definition in the assumption to obtain the body of a member. The remaining six rules
+give standard reductions for conditionals and pattern matching.
+
+The language is simple, but sufficient for our purpose. All expressions reduce to a value in a 
+finite number of steps or get stuck due to an error condition. An error condition can be a wrong 
+argument passed to conditional, pattern matching or one of the conversion functions from 
+Figure~\ref{fig:op-conversions}.
 
@@
 \begin{equation*}
-\qquad
+\inference
+  {L; \Gamma \vdash e_1 : \ident{StructVal}}
+  {L; \Gamma \vdash [e_1; \ldots; e_n]:\ident{StructVal}}
+\quad  
+\inference
+  {~}
+  {L; \Gamma \vdash n : \ident{int}}
+\end{equation*}
+\begin{equation*}
+\inference
+  {L; \Gamma \vdash e_1 : \tau}
+  {L; \Gamma \vdash [e_1; \ldots; e_n]:\ident{list}\langl\tau\rangl}
+\quad
 \inference
@@
   {L; \Gamma \vdash s : \ident{StructVal}}
-\qquad  
+\end{equation*}
+\vspace{0.5em}
+\begin{equation*}
 \inference
-  {~}
-  {L; \Gamma \vdash n : \ident{int}}
-\end{equation*}
-
-\begin{equation*}
-\inference
-  {L; \Gamma \vdash e : C & \kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;.. \in L}
+  {L; \Gamma \vdash e : C \\ \kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;.. \in L}
   {L; \Gamma \vdash e.N_i:\tau_i}
 \end{equation*}
-
+\vspace{0.25em}
 \begin{equation*}
@@
 \paragraph{Type checking.} 
-The type checking rules in Figure~\ref{fig:ff-typecheck} are written using a judgement
+Typing rules in Figure~\ref{fig:ff-typecheck} are written using a judgement
 $L; \Gamma \vdash e : \tau$ where the context also contains a set of class declarations $L$.
-The rules demonstrate the key difference between our language and standard ML or Featherweight Java:
-%
-\begin{itemize}[noitemsep]
-\item[--] All structural values $s$ have a type \ident{StructVal}. Some of those have other types
-  (primitive integers, decimals, floats, strings and Booleans as illustrated by the rule for $n$).
+The rules demonstrate the key differences from Standard ML and Featherweight Java:
+%
+\begin{itemize}
+\item[--] All structural values $s$ have the type \ident{StructVal}, but some have other types
+  (Booleans, strings, integers, etc.) as illustrated by the rule for $n$.
   For other values, \ident{StructVal} is the only type -- this includes records and \kvd{null}.
-\item[--] A list containing other structural values $[s_1; \ldots; s_n]$ has a type \ident{StructVal}.  
+\item[--] A list containing other structural values $[s_1; \ldots; s_n]$ has a type \ident{StructVal},
   but can also have the $\ident{list}\langl\tau\rangl$ type. Conversely, lists that contain
   non-structural values like objects or options are not of type \ident{StructVal}.
-\item[--] Primitive operations $op$ are treated as functions, accepting \ident{StructVal} and producing
-  an appropriate type as the result.
-\item[--] The two rules for checking class construction and member access are similar to corresponding
+\item[--] Operations $op$ (omitted) are treated as functions, accepting 
+  \ident{StructVal} and producing an appropriate result type.
+\item[--] Rules for checking class construction and member access are similar to corresponding
   rules of Featherweight Java.  
@@
 An important part of Featherweight Java that is omitted here is the checking of type declarations
-(ensuring the bodies of members are well-typed). We omit this, because it is not needed for the
-relativized safety theorem, which considers only classes generated by our type provider (which 
-are correct by construction).
+(ensuring the bodies of members are well-typed). We omit this, because we consider only classes 
+generated by our type providers (which are correct by construction).
 
@@
 \begin{figure*}
+\begin{equation*}
+\quad\qquad\boxed{
+\begin{array}{rcl}
+ \nameoftag(\ident{string}) &\narrow{=}& \ident{String} \\
+ \nameoftag(\ident{bool}) &\narrow{=}& \ident{Boolean} \\
+\end{array}
+\qquad
+\begin{array}{rcl}
+ \nameoftag(\ident{number}) &\narrow{=}& \ident{Number} \\
+ \nameoftag(\ident{collection}) &\narrow{=}& \ident{List} \\
+\end{array}
+\qquad
+\begin{array}{rcl}
+ \nameoftag(\ident{record}) &\narrow{=}& \ident{Record} \\
+ \nameoftag(\ident{named}\;\nu) &\narrow{=}& \nu \\
+\end{array}
+}
+\end{equation*}
+
 \begin{multicols}{2}
@@
 \begin{array}{l}
- \sem{\sigma_p}_e = \tau_p,op(e),\emptyset\quad\textnormal{where} \\[0.4em]
-\quad\sigma_p, \tau_p, op\in  \{~ (\ident{bit}, \ident{bool}, \ident{asBool}),  (\ident{bool}, \ident{bool}, \ident{asBool})\\[0.2em]
-\hspace{2.9em} (\ident{int}, \ident{int}, \ident{asInt}), (\ident{decimal},\ident{decimal},\ident{asDec}),\\[0.2em]
-\hspace{2.9em} (\ident{float},\ident{float},\ident{asFloat}), (\ident{string},\ident{string},\ident{asStr}) ~\}\\[0.6em]
+ \sem{\sigma_p}_e = \tau_p,op(e),\emptyset\quad\textnormal{where}~\tau_p = \sigma_p \\[0.2em]
+\quad\sigma_p, op\in  \{~(\ident{decimal},\ident{asDec}), (\ident{float},\ident{asFloat})~\}
 \end{array}
@@
 \begin{array}{l}
- \sem{\nu?\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}}_e = \\[0.1em]
+ \sem{\,\nu_? \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}\,}_e = \\[0.1em]
  \quad C, \kvd{new}~C(e), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
@@
  \qquad M_i = \kvd{member}~\nu_i:\tau_i=e_i\\[0.1em]
- \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_{e'},\;\; e'=\ident{getField}(\nu?, \nu_i, e)\\[0.6em]
+ \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_{e'},\;\; e'=\ident{getField}(\nu_?, \nu_i, e)\\[0.6em]
 \end{array}
@@
 \begin{array}{l}
- \sem{\,[\sigma]\,}_e = \ident{list}\langl\tau\rangl, \ident{List.map}~(\lambda x\rightarrow e')~(\ident{getChildren}(e)), L\\[0.4em]
- \quad \textnormal{where}~~\tau, e', L = \sem{\hat{\sigma}}_x
-\end{array}
-\end{equation*}
-
+ \sem{\,[\sigma]\,}_e = \ident{list}\langl\tau\rangl, e_b, L \;\;\textnormal{where}\\[0.4em]
+ \qquad e_b = \ident{List.map}~(\lambda x\rightarrow e')~(\ident{getChildren}(e)) \\[0.1em]
+ \qquad \tau, e', L = \sem{\hat{\sigma}}_x
+\end{array}
+\end{equation*}
+%
+% Anything else
+\begin{equation*}
+\begin{array}{l}
+ \sem{\bot}_e = \sem{\kvd{null}}_e = \ident{StructVal}, e, \emptyset
+\end{array}
+\end{equation*}
+
+\noindent
+\begin{equation*}
+\hspace{-2.5em}
+\begin{array}{l}
+ \sem{\sigma_p}_e = \tau_p,e,\emptyset\quad\textnormal{where}~\tau_p = \sigma_p \\[0.2em]
+\quad\sigma_p \in  \{~ \ident{string},\ident{bool},\ident{int}~\}\\[0.6em]
+\end{array}
+\end{equation*}
+%
+% Sum type
+\begin{equation*}
+\hspace{-2.5em}
+\begin{array}{l}
+ \sem{\,\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl\,}_e = \\[0.1em]
+ \quad C, \kvd{new}~C(e), L_1\cup\ldots\cup L_n\cup\{ L \}\;\;\textnormal{where}\\[0.6em]
+ \qquad \;\;C~\textnormal{is a fresh class name} \\[0.1em]
+ \qquad \;\,\,L = \kvd{type}~C(v:\ident{StructVal})~=~M_1\ldots M_n \\[0.1em]
+ \qquad M_i = \kvd{member}~\nu_i:\ident{option}\langl\tau_i\rangl=\\[0.1em]
+ \hspace{5.8em}  \kvd{if}~\ident{isTag}(t_i, v)~\kvd{then}~
+     \ident{Some}(e_i)~\kvd{else}~\ident{None} \\[0.1em]
+ \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_e,\;\; t_i = \tytagof{(\sigma_i)},\;\; \nu_i=\nameoftag{(t_i)}
+\end{array}
+\end{equation*}
+%
 % Option values
-\noindent
-\begin{equation*}
+\begin{equation*}
+\hspace{-2.5em}
 \begin{array}{l}
@@
 \end{equation*}
-%
-% Sum type
-\begin{equation*}
-\begin{array}{l}
- \sem{\sigma_1 + \ldots + \sigma_n}_e = \\[0.1em]
- \quad C, \kvd{new}~C(e), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
- \qquad \;\;C~\textnormal{is a fresh class name} \\[0.1em]
- \qquad \;\,\,L = \kvd{type}~C(v:\ident{StructVal})~=~M_1\ldots M_n \\[0.1em]
- \qquad M_i = \kvd{member}~\nu_i:\ident{option}\langl\tau_i\rangl=\\[0.1em]
- \hspace{5.8em}  \kvd{if}~\ident{isNull}(v)~\kvd{then}~\ident{None}~\kvd{else} \\
- \hspace{5.8em}  \kvd{if}~\ident{isTag}(t_i, v)~\kvd{then}~
-     \ident{Some}(e_i)~\kvd{else}~\ident{None} \\[0.1em]
- \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_e,\;\; t_i = \tytagof{(\sigma_i)},\;\; \nu_i=\nameoftag{(t_i)}
-\end{array}
-\end{equation*}
-%
-% Anything else
-\begin{equation*}
-\begin{array}{l}
- \sem{\top}_v = \sem{\kvd{null}}_v = \ident{StructVal}, v, \emptyset
-\end{array}
-\end{equation*}
 \end{multicols}
-
-\vspace{-0.5em}
-\noindent
-\begin{equation*}
-\qquad
-\qquad
-\begin{array}{rcl}
- \nameoftag(\ident{string}) &\narrow{=}& \ident{String} \\
- \nameoftag(\ident{bool}) &\narrow{=}& \ident{Boolean} \\
-\end{array}
-\qquad
-\begin{array}{rcl}
- \nameoftag(\ident{number}) &\narrow{=}& \ident{Number} \\
- \nameoftag(\ident{collection}) &\narrow{=}& \ident{List} \\
-\end{array}
-\qquad
-\begin{array}{rcl}
- \nameoftag(\ident{rec-anon}) &\narrow{=}& \ident{Record} \\
- \nameoftag(\ident{rec-named}\;\nu) &\narrow{=}& \nu \\
-\end{array}
-\end{equation*}
 
@@
 
-So far, we defined the type inference algorithm for structured data formats which produces a structural
-type $\sigma$ based on one or more sample documents (Section~\ref{sec:inference}). In Section~\ref{sec:formal-ff},
-we defined a simplified model of F\# language and evaluation, which also included conversions that
-model the F\# Data library runtime (Section~\ref{sec:formal-tp}). In this section, we define how
-the type providers work, which links the two aspects discussed so far.
-
-The type providers for XML, CSV and JSON all work in the same way. They take sample documents and infer a common
-supertype $\sigma$ from the samples. From $\sigma$, they then generate F\# types that are then exposed to the
-programmer\footnote{The actual implementation provides \emph{erased types} as described in \cite{fsharp-typeprov}. 
-Here, we treat the code as actually generated. This is an acceptable simplification, because F\# Data type providers do not 
-rely on laziness that is available through erased types.}.
+So far, we defined the type inference algorithm which produces a structural type $\sigma$ from one 
+or more sample documents (\S\ref{sec:inference}) and we defined a simplified model of evaluation
+of F\# language (\S\ref{sec:formal-ff}) and F\# Data runtime (\S\ref{sec:formal-tp}). Next, we 
+define how the type providers work, linking the two parts.
+
+All F\# Data type providers take (one or more) sample documents, infer a common supertype $\sigma$
+and then use $\sigma$ to generate F\# types that are exposed to the programmer\footnote{The actual 
+implementation provides \emph{erased types} as described in \cite{fsharp-typeprov}. Here, we treat 
+the code as actually generated. This is an acceptable simplification, because F\# Data type providers 
+do not rely on laziness that is available through erased types.}.
 
 \paragraph{Type provider mapping.}
-When generating types, the type provider produces an F\# type $\tau$, expression that wraps a structural 
-document value (of type \ident{StructVal}) as a value of type $\tau$ and a collection of class definitions. 
-We express it using the following mapping:
-%
-\begin{equation*}
-\sem{-}_e : \sigma \rightarrow \tau \times e' \times L
-\end{equation*}
-%
-The semantics is parameterized by an expression $e$, which represents code to obtain a structural value that is
-being wrapped. Then, given an inferred structural type $\sigma$, the type provider produces an F\# type $\tau$,
-an expression $e'$ which constructs a value of type $\tau$ using $e$ and also a set of class definitions $L$.
-
-Figure~\ref{fig:tp-generation} shows the rules that define $\sem{-}_e$. Primitive types are all
+When generating types, the type provider produces an F\# type $\tau$, an expression that wraps 
+the input document (of type \ident{StructVal}) as a value of type $\tau$ and a collection of class 
+definitions. We express it using the following mapping:
+%
+\begin{equation*}
+\sem{-}_{-} : (\sigma\times e) \rightarrow (\tau \times e' \times L)
+\end{equation*}
+%
+The mapping $\sem{\sigma}_e$ takes a inferred structural type $\sigma$ and an expression $e$, 
+which represents code to obtain a structural value that is being wrapped. It returns an F\# type 
+$\tau$, an expression $e'$ which constructs a value of type $\tau$ using $e$ and also a set of 
+generated class definitions $L$.
+
+Figure~\ref{fig:tp-generation} shows the rules that define $\sem{-}_{-}$. Primitive types are all
 handled by a single rule. For a given structural type, it returns the corresponding F\# type -- the types are
@@
 
+% ==================================================================================================
+%
+%    #######
+%       #    #   # #####  ######     ####    ##   ###### ###### ##### #   #
+%       #     # #  #    # #         #       #  #  #      #        #    # #
+%       #      #   #    # #####      ####  #    # #####  #####    #     #
+%       #      #   #####  #              # ###### #      #        #     #
+%       #      #   #      #         #    # #    # #      #        #     #
+%       #      #   #      ######     ####  #    # #      ######   #     #
+%
 % ==================================================================================================

### 2015-11-17 pldi-2016/paper.tex
More edits and add practical experience

@@
-\documentclass[10pt,preprint,clearpagebib]{sigplanconf}
+\documentclass[10pt,preprint,blind,clearpagebib]{sigplanconf}
 
@@
 \usepackage[svgnames]{xcolor}
-\definecolor{lcl}{RGB}{160,0,128}
+\definecolor{lcl}{RGB}{140,0,100}
 \usepackage[colorlinks=true,breaklinks,draft=false]{hyperref}
-\hypersetup{urlcolor=lcl,linkcolor=lcl}
+\hypersetup{urlcolor=lcl,linkcolor=lcl,citecolor=lcl}
 \newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}
-
 \usepackage{amssymb}
@@
 
-% TODO: Adapt XML example 
-% TODO: Switch from sum types to variant types
 % TODO: Discuss 'stability' of the inference (minor change in schema does not cause dramatic cahnges)
-% TODO: Say that any variant is a subtype of any other variant 
-% TODO: Say that anything is a subtype of variant
 % TODO: Say that variant is a last-resort thing
-% TODO: Double check that bottom/top are used correctly
-% TODO: Explain (even more) that there is no least upper bound, which is why Theorem 3 is good enough
-% TODO:  -- but maybe we can have lub w.r.t variants? or something?
-% TODO: Avoid "no other rule applies" in the formal system
 % TODO: Practical evaluation (lots of downloads, lots of Github - how often people customize samples?)
 % TODO: Explain why we need option type (because T + null is a variant)
-
 % TODO: In the PLDI version, we delete all the unnecessary runtime conversions
 % but we should say that those happen for practical reasons (e.g. float -> int, or anyhing to string)
-
-% T :> S means that S is a subtype of T (and S can be used where T is expected)
-% TOP is the universal supertype
-% BOTTOM is the universal subtype
-
-% variant is the top type 
-%  (variant is a suprtype of everything; everything is a subtype of variant; 
-%   everything can be used where variant is expected)
-%
-% bottom is the bottom type
-%  (bottom is a subtype of everything; everything is a supertype of bottom
-%   bottom can be used where anything else is expected - but it has no values)
-%
-% define :=: when <: and :>, then take equivalence classes
-
-% least upper bound of T1 and T2 is a supertype of both (T :> T1 and T :> T2)
-% and for all S such that S :> T1 and S :> T2, it holds that S :> T
 
@@
 \begin{abstract}
-Most modern applications interact with external services and access data in formats such as 
-XML, JSON and CSV. Static type systems do not understand external data sources and only 
-make accessing data more cumbersome. Should we give up and leave the messy world of external data 
-to dynamic typing and runtime checks? Of course, not!
+Most modern applications interact with external services and access data in structured formats such 
+as XML, JSON and CSV. Static type systems do not understand such formats making accessing data more 
+cumbersome. Should we give up and leave the messy world of external data to dynamic typing and 
+runtime checks? Of course, not!
 
@@
 \begin{array}{l}
- \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://weather.org/?q=NYC}) \\
+ \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC}) \\
  \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
@@
 \begin{array}{l}
- \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://weather.org/?q=NYC}\rangl \\[0.1em]
+ \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://api.owm.org/?q=NYC}\rangl \\[0.1em]
  \ident{printfn}~\str{Nice \%f degrees!}~(\ident{W.GetSample().Main.Temp})
@@
 %
-Here, $\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider at 
+$\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider at 
 compile-time with the URL as a sample. The type provider infers the structure of the response
@@
 \end{itemize}
-
-% Although the F\# Data library \cite{fsharp-data} has been widely adopted and is one of the most downloaded F\# libraries 
-% it has not yet been presented in an academic form. Additional documentation and source code can 
-% be found at \url{http://fsharp.github.io/FSharp.Data}.
 
@@
 We start with an informal overview that shows how F\# Data type providers simplify working with 
-JSON, XML and CSV. We also introduce the necessary aspects of F\# 3.0 type providers along the 
+JSON, XML and CSV. We introduce the necessary aspects of F\# type providers along the 
 way. The examples in this section illustrate a number of key properties of our type inference 
@@
 \begin{itemize}
-\item Our mechanism is robust and predictable. This is important as the user directly works with 
-  the inferred types and should understand why a specific type was inferred from a given 
-  sample\footnote{In particular, we do not use probabilistic methods where adding additional 
-  sample could change the shape of the type.}.
-
-\item Our inference mechanism prefers records over unions. This better supports developer tooling --
+\item Our mechanism is predictable. The user directly works with the inferred types and should 
+  understand why a specific type was inferred from a given sample\footnote{In particular, we do 
+  not use probabilistic methods where adding additional sample could change the shape of the type.}.
+
+\item Our inference mechanism prefers records. This supports developer tooling --
   most F\# editors provide code completion hints on ``.'' and so types with properties (records)
@@
 
-\item Finally, we handle a number of practical concerns that may appear overly complicated, but
-  are important in the real world. This includes support for different numerical types, \kvd{null}
-  values and missing data and also different ways of representing Booleans in CSV files.
+\item Finally, we handle a number of practical concerns important in the real world. This includes 
+  support for different numerical types, \kvd{null} values and missing data.
 \end{itemize}
@@
 
-The JSON format used in the example in Section~\ref{sec:introduction} is a popular format for data
-exchange on the web based on data structures used in JavaScript. The following is the definition 
-of the \ident{JsonValue} type used earlier to represent parsed JSON:
+The JSON format used in \S\ref{sec:introduction} is a populardata exchange format based on 
+JavaScript data structures. The following is the definition of \ident{JsonValue} 
+used earlier to represent JSON:
 %
@@
  \kvd{type}~\ident{JsonValue} = \\[0.1em]
- \quad|~\ident{Null} \\[0.1em]
- \quad|~\ident{Number}~\kvd{of}~\ident{decimal} \\[0.1em]
- \quad|~\ident{String}~\kvd{of}~\ident{string} \\[0.1em]
- \quad|~\ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
- \quad|~\ident{Record}~\kvd{of}~\ident{Map}\langl\ident{string}, \ident{JsonValue}\rangl \\[0.1em]
- \quad|~\ident{Array}~\kvd{of}~\ident{JsonValue}[] \\[0.1em]
-\end{array}
-\end{equation*}
-%
-The OpenWeatherMap example in the introduction used only a (nested) record containing a numerical
-value. To demonstrate other aspects of the JSON type provider, we look at a more complex example
-that also involves \kvd{null} value and an array:
+ \quad|~ \ident{Number}~\kvd{of}~\ident{decimal} ~~|~ \ident{Null} \\[0.1em]
+ \quad|~ \ident{String}~\kvd{of}~\ident{string} \qquad\, |~ \ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
+ \quad|~ \ident{Record}~\kvd{of}~\ident{Map}\langl\ident{string}, \ident{JsonValue}\rangl \\[0.1em]
+ \quad|~ \ident{Array}~\kvd{of}~\ident{JsonValue}[] \\[0.1em]
+\end{array}
+\end{equation*}
+%
+The earlier example used only a nested record containing a number. To demonstrate other 
+aspects of the JSON type provider, we look at an example that also involves an array:
 %
@@
 \begin{verbatim}
-  [ { "name": "Jan", "age": 25 },
-    { "name": "Alexander", "age": 3.5 },
-    { "name": "Tomas" } ]
+  [ { "name":"Jan", "age":25 }, { "name":"Tomas" },
+    { "name":"Alexander", "age":3.5 } ]
 \end{verbatim}
@@
 \noindent
-Say we want to print the names of people in the list with an age if it is available. As before, 
-the standard approach would be to pattern match on the parsed \ident{JsonValue}. The code would 
-check that the top-level node is a \ident{Array}, iterate over the elements checking that each is
-a \ident{Record} with certain properties and throw an exception or skip values in incorrect format.
-The standard approach can be made nicer by defining helper functions. However, we still need to
-specify names of fields as strings, which is error prone and can not be statically checked.
-
-Assuming the \strf{people.json} file contains the above example and \ident{data} is a string value
-that contains another data set in the same format, we can print names and ages as follows:
+Say we want to print the names with an age if it is available. The standard approach would 
+be to pattern match on the parsed \ident{JsonValue}, check that the top-level node is a \ident{Array}
+and iterate over the elements checking that each is a \ident{Record} with certain properties. We
+would throw an exception or skip values in incorrect format. Even if we defined helper functions, 
+we would still need to specify field names as strings, which is error prone and can not be 
+statically checked.
+
+Assuming \strf{people.json} is the above example and \ident{data} is a string containing
+JSON in the same format, we can write:
 %
@@
  \kvd{type}~\ident{People}~=~\ident{JsonProvider}\langl\str{people.json}\rangl\hspace{1em} \\[0.6em]
- \kvd{let}~\ident{items}~=~\ident{People.Parse}(\ident{data})\\[0.1em]
- \kvd{for}~\ident{item}~\kvd{in}~\ident{items}~\kvd{do}\\[0.1em]
+ \kvd{for}~\ident{item}~\kvd{in}~\ident{People.Parse}(\ident{data})~\kvd{do}\\[0.1em]
  \quad\ident{printf}~\str{\%s }~\ident{item.Name}\\[0.1em]
@@
 \paragraph{Type providers.}
-The notation $\ident{JsonProvider}\langl\str{people.json}\rangl$ on the first line passes a 
-\emph{static parameter} to the type provider. Static parameters are resolved at compile-time, 
-so the file name has to be a constant. The provider analyzes the sample and generates
-a type that we name \ident{People}. In editors that use the F\# Compiler Service, the 
-type provider is also executed at development-time and so the same provided types are
-used in code completion.
-
-The \ident{JsonProvider} uses a type inference algorithm (Section~\ref{sec:inference})  and
+The notation $\ident{JsonProvider}\langl\str{people.json}\rangl$ passes a \emph{static parameter} 
+to the type provider. Static parameters are resolved at compile-time and have to be constant. 
+The provider analyzes the sample and generates a type that we name \ident{People}. In most F\# 
+editors, the type provider is also executed at development-time and the provided types are 
+used in auto-completion.
+
+The \ident{JsonProvider} uses a type inference algorithm (\S\ref{sec:inference})  and
 infers the following types from the sample:
@@
 \begin{array}{l}
- \kvd{type}~\ident{Entity}~=  \\[0.1em]
- \quad \kvd{member}~\ident{Name}~:~\ident{string} \\[0.1em]
- \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{decimal}\rangl \\[0.6em]
- \kvd{type}~\ident{People}~=  \\[0.1em]
- \quad \kvd{member}~\ident{GetSample}~:~\ident{unit}~\rightarrow~\ident{Entity}[] \\[0.1em]
- \quad \kvd{member}~\ident{Parse}~:~\ident{string}~\rightarrow~\ident{Entity}[] \\[0.1em]
+ \kvd{type}~\ident{Entity}~=  \\
+ \quad \kvd{member}~\ident{Name}~:~\ident{string} \\
+ \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{decimal}\rangl \\[0.5em]
+ \kvd{type}~\ident{People}~=  \\
+ \quad \kvd{member}~\ident{GetSample}~:~\ident{unit}~\rightarrow~\ident{Entity}[] \\
+ \quad \kvd{member}~\ident{Parse}~:~\ident{string}~\rightarrow~\ident{Entity}[] \\
 \end{array}
@@
 
-The type \ident{People} provides two methods for reading data. \ident{GetSample} returns the
+The type \ident{People} has two methods for reading data. \ident{GetSample} parses the
 sample used for the inference and \ident{Parse} parses a JSON string containing data in the same 
-format as the sample. Since the sample JSON is a collection of records, both methods return an 
-array of \ident{Entity} values.
+format as the sample. Both methods return an array of \ident{Entity} values.
 
@@
 In addition to the structure of the types, the type provider also specifies what code should be 
-executed at run-time in place of \ident{item.Name} and other operations. This is done through a 
-\emph{type erasure} mechanism demonstrated in Section~\ref{sec:providers-csv}. In this example,
-the runtime behaviour is the same as in the hand-written sample in Section~\ref{sec:introduction} --
-a member access throws an exception if the document does not have the expected format.
-
-Informally, the safety property discussed in Section~\ref{sec:safety} states that if the inputs
-are subtypes of some of the provided samples (\emph{i.e.}~the samples are representative), then
+executed at run-time in place of \ident{item.Name} and other operations. The runtime behaviour is 
+the same as in the earlier hand-written sample (\S\ref{sec:introduction}). A member access 
+throws an exception if the document does not have the expected format.
+
+Informally, the safety property (\S\ref{sec:safety}) states that if the inputs
+are subtypes of some of the provided samples (i.e.~the samples are representative), then
 no exceptions will occur. In other words, we cannot avoid all failures, but we can prevent some.
-If the OpenWeatherMap changes the response format, the sample in Section~\ref{sec:introduction}
+If the OpenWeatherMap changes the response format, the sample in \S\ref{sec:introduction}
 will not re-compile and the user knows that the code needs to be changed. What this means for
-the traditional ML type safety is discussed in Section~\ref{sec:safety-discuss}.
+the traditional ML type safety is discussed in \S\ref{sec:safety-discuss}.
 
@@
 are available for a person. To support this scenario, our inference algorithm prefers records 
-(we also treat XML elements and CSV rows as records).
+(we also treat XML elements and CSV rows as records) and only infers types that require pattern 
+matching as the last resort.
 
@@
 
-\subsection{Processing XML data, alternative}
+\subsection{Processing XML documents}
 \label{sec:providers-xml}
 
-In XML, documents are formed by (possibly) nested elements with attributes. We can view elements
-as records (unlike in JSON, the records are named) with a field for each attribute and an additional
-special field for the nested contents (which may be a collection of elements).
-
-As an example, consider a simple extensible document format where a root element
+XML documents are formed by nested elements with attributes. We can view elements as records with 
+a field for each attribute and an additional special field for the nested contents (which is a 
+collection of elements).
+
+Consider a simple extensible document format where a root element
 {\ttfamily\small <doc>} can contain a number of document element, one of which is
@@
 \noindent
-The F\# Data library has been primarilly designed to allow easy reading of data. For example,
-say we want to print all headings that appear in the document. The sample shows a part of the 
-possible document structure (in particular the {\ttfamily\small <heading>} element we want to
-read), but it does not show all possible elements (say, {\ttfamily\small <table>}). Assuming the
-above document is \strf{sample.doc}, we can write:
+The F\# Data library has been designed primarilly to simplify reading of data. For example,
+say we want to print all headings in the document. The sample shows a part of the document structure 
+(in particular the {\ttfamily\small <heading>} element), but it does not show all possible elements 
+(say, {\ttfamily\small <table>}). Assuming the above document is \strf{sample.doc}, we can write:
 %
@@
 The example iterates over a collection of elements \ident{elem} returned by \ident{root.Doc}. 
-Our system provides a typed access to elements known from the sample document and so we can
-write \ident{elem.Heading}.
+The type provides a typed access to elements known from the sample document and so we can
+write \ident{elem.Heading}, which returns an optional value.
 
 \paragraph{Open world and variant types.}
-In the above example, The system does not infer a union type representing a choice between
-heading, paragraph and image. By its nature, the XML format is extensible and the sample cannot 
-include all possible examples\footnote{Even when the document structure is defined using XML 
-Schema, documents may contain elements prefixed with other namespaces.}. The type of \ident{elem}
-is thus a \emph{variant} type annotated with the three statically known possibilities (heading,
-paragraph, image). The variant is mapped to the following F\# type:
+In the above example, we do not infer a union type representing a choice between 
+heading, paragraph and image. By its nature, XML is extensible and the sample cannot include all 
+possible nodes\footnote{Even when the document structure is defined using XML Schema, documents may 
+contain elements prefixed with other namespaces.}. The type of \ident{elem} is thus a top type (we 
+call it \emph{variant}) annotated with the statically known possibilities (heading, paragraph, image). 
+It is mapped to the following F\# type:
 %
@@
 
-One perhaps surprising consequence of the use of variant types is that the above code will work
-correctly on document that contains elements other than those in the sample document. This would 
-arguably not be desirable design for data types in a programming language, but it is extremely
-useful when reading data from external data sources. It is also worth noting that our type 
+As a consequence, the variant type can be seen as the top type. The above code will work
+correctly on document that contains any elements, including those not listed in the sample. This
+is extremely useful when reading data from external data sources, because the sample document only
+needs to contain values relevant to the application. It is also worth noting that our type 
 inference attempts to minimze the number of variant types that appear in the resulting type --
-a variant is not used if there is another option.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Reading CSV files}
-\label{sec:providers-csv}
-
-In the CSV file format, the structure is a collection of rows (records) consisting of fields 
-(with names specified by the first row). The inference needs to infer the types of fields.
-For example:
-%
-{\small{
-\begin{verbatim}
-  Ozone, Temp, Date,       Autofilled
-  41,    67,   2012-05-01, 0
-  36.3,  72,   2012-05-02, 1
-  12.1,  74,   3 kveten,   0
-  17.5,  #N/A, 2012-05-04, 0
-\end{verbatim}
-}}
-%
-\noindent
-One difference between JSON and CSV formats is that in CSV, the literals have no data types.
-In JSON, strings are \str{quoted} and Booleans are \kvd{true} and \kvd{false}. This is not the
-case for CSV and so we need to infer not just the structure, but also the primitive types.
-
-Assuming the sample is saved as \strf{airdata.csv}, the following snippet prints all
-rows from another file that were not autofilled:
-%
-\begin{equation*}
-\begin{array}{l}
- \kvd{type}~\ident{AirCsv}~=~\ident{CsvProvider}\langl\str{airdata.csv}\rangl\hspace{1em} \\[0.6em]
- \kvd{let}~\ident{air}~=~\ident{AirCsv.Parse}(\ident{data})\\[0.1em]
- \kvd{for}~\ident{row}~\kvd{in}~\ident{air.Rows}~\kvd{do}\\[0.1em]
- \quad\kvd{if}~\ident{not}~\ident{row.Autofilled}~\kvd{then}\\[0.1em]
- \quad\quad \ident{printf}~\str{\%s: \%d}~\ident{row.Date row.Ozone}\\
-\end{array}
-\end{equation*}
-%
-The type of the record (\ident{row}) is, again, inferred from the sample. The \ident{Date} column
-uses mixed formats and is inferred as \ident{string} (although we support many date formats and 
-``May 3'' would work). More interestingly, we also infer \ident{Autofilled} as Boolean, because 
-the sample contains only $0$ and $1$ values and using those for Booleans is a common CSV convention.
-Finally, the fields \ident{Ozone} and \ident{Temp} have types \ident{decimal} and
-$\ident{option}\langl\ident{int}\rangl$.
+a variant is not used if there is another option (Corollary~\ref{thm:no-unions}).
 
@@
 \subsection{Summary}
-The previous three sections demonstrated the F\# Data type providers for JSON, CSV and XML.
-A reader interested in more examples is invited to look at documentation on the F\# Data web 
-site\footnote{Available at \url{http://fsharp.github.io/FSharp.Data/}}.
-
-It should be noted that we did not attempt to present the library using well-structured ideal 
-samples, but instead used data sets that demonstrate typical issues that are frequent in real 
-world inputs (missing data, inconsistent encoding of Booleans and heterogeneous structures).
-
-From looking at just the previous three examples, these may appear arbitrary, but our experience
-suggests that they are the most common issues. The following JSON response with government debt
-information returned by the World Bank API demonstrates all three problems:
+In addition to JSON and XML (\S\ref{sec:providers-json}, \S\ref{sec:providers-xml}), F\# Data also 
+implements a type provider for CSV. We treat CSV data as lists of records (with field for each 
+column) and so handling of CSV is just a special case of our inference algorithm.
+
+Throughout the introduction, we used data sets that demonstrate typical issues that are frequent
+in real-world inputs (missing data, inconsistent encoding of primitive values and heterogeneous 
+structures). In our experience, these are the most common issues. The following JSON response with 
+government debt information returned by the World Bank API\footnote{Available at 
+\url{http://data.worldbank.org}} demonstrates all three problems:
 %
@@
     [ { "indicator": "GC.DOD.TOTL.GD.ZS",
-        "date": "2012",
-        "value": null },
+        "date": "2012", "value": null },
       { "indicator": "GC.DOD.TOTL.GD.ZS",
-        "date": "2010",
-        "value": "35.1422970266502" } ] ]
+        "date": "2010", "value": "35.14229" } ] ]
 \end{verbatim}
@@
 The first is a record with meta-data about the current page and the second is an array with data. 
-The JSON type provider infers this as heterogeneous collection with properties \ident{Record}
-for the former and \ident{Array} for the latter. Second, the \ident{value} is \kvd{null} for some
-records. Third, numbers can be represented in JSON as numeric literals (without quotes), but
-here, they are returned as string literals instead\footnote{This is often used to avoid non-standard
-numerical types in JavaScript.}.
+The actual implementation supports a concept of heterogeneous collections (omitted in this paper
+for space reasons) and provides a type with properties \ident{Record} for the former and 
+\ident{Array} for the latter. Second, the \ident{value} is \kvd{null} for some records. Third, 
+numbers can be represented in JSON as numeric literals (without quotes), but here, they are 
+returned as string literals instead\footnote{This is often used to avoid non-standard numerical 
+types of JavaScript.}.
 
@@
 
-\newpage
-~
-\newpage
 \section{Structural type inference}
@@
 Our type inference algorithm for structured formats is based on a subtyping relation. When 
-inferring the type of a document, we infer the most specific types of individual values (CSV rows,
-JSON or XML nodes) and recursively find a common supertype of all child nodes or sample documents.
-
-In this section, we define the \emph{structural type} $\sigma$, which is a type of structured data.
-Note that this type is distinct from the programming language types $\tau$ (type providers generate 
-the latter from the former). Next, we define the subtyping relation on structural types $\sigma$ and 
-describe the algorithm for finding a common supertype.
+inferring the type, we infer the most specific types of individual values (CSV rows, JSON or XML 
+nodes) and recursively find a common supertype of all child nodes or all sample documents.
+
+We first define the type of structured data $\sigma$. Note that this type is distinct 
+from the programming language types $\tau$ (type providers generate the latter from the former). 
+Next, we define the subtyping relation on structural types $\sigma$ and describe the algorithm 
+for finding a common supertype.
 
@@
 (written as $\sigma$). We write $\nu$ for record field names and $\nu_?$ for names that can be empty:
-%
+
+\noindent
 \begin{equation*}
@@
 Any non-nullable type can be wrapped in the \kvd{option} constructor to explicitly permit the 
-\kvd{null} value. Type providers map \kvd{option} to standard F\# option type. A collection 
-type $[\sigma]$ is also nullable and missing values or \kvd{null} are treated as empty collection. 
-The type $\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded notation) and 
-$\bot$ is the bottom type.
+\kvd{null} value. Type providers map \kvd{option} to standard F\# option. A collection $[\sigma]$ 
+is also nullable and missing values or \kvd{null} are treated as empty collections. The type 
+$\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded notation) and $\bot$ 
+is the bottom type.
 
@@
 a value is of a different type than the statically known ones and so it can be seen as a family of 
-top types. For the same reason, the variants also implicitly permit the \kvd{null} value.
+top types. For the same reason, variants also implicitly permit the \kvd{null} value.
 
@@
 structural types. The upper part shows non-nullable types (with records and primitive types) and 
-the lower part shows nullable types with \kvd{null}, collections and optional values. We omit links 
-between the two part, but any type $\hat{\sigma}$ is a subtype of $\kvd{option}\langl\hat{\sigma}\rangl$ 
-and we abbreviate $\kvd{option}\langl\sigma\rangl$ as $\sigma?$.
+the lower part shows nullable types with \kvd{null}, collections and optional values. We 
+abbreviate $\kvd{option}\langl\sigma\rangl$ as $\sigma?$ and we omit links between the two parts;
+any type $\hat{\sigma}$ is a subtype of $\kvd{option}\langl\hat{\sigma}\rangl$.
 
@@
 
-\vspace{0.5em}
-\noindent\boxed{\emph{Primitives, options, collections}}
+\noindent
 \begin{align}
-\label{eq:sub-prim}\tag{B1}
-\ident{float}\,&:>\,\ident{decimal}\,:>\,\ident{int}&\\[-0.2em]
-\label{eq:sub-null}\tag{B2}
-\sigma &:> \kvd{null}  &(\textnormal{iff}~\sigma \neq \hat{\sigma}) \\[-0.2em]
-\label{eq:sub-bot}\tag{B3}
+  \label{eq:sub-prim}\tag{P1}
+  \ident{float}\,:>\,\ident{decimal}\,&:>\,\ident{int}&\\[-0.2em]
+  \label{eq:sub-null}\tag{P2}
+  \sigma &:> \kvd{null}  &(\textnormal{iff}~\sigma \neq \hat{\sigma}) 
+\end{align}
+
+\noindent
+\begin{align}
+  \label{eq:sub-opt}\tag{P3}
+  \kvd{option}\langl\hat{\sigma}\rangl &:> \hat{\sigma}  &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
+  \label{eq:sub-opt-cov}\tag{P4}
+  \kvd{option}\langl\hat{\sigma_1}\rangl &:> 
+    \kvd{option}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} :> \hat{\sigma_2}) \\[-0.2em]
+  \label{eq:sub-col}\tag{P5}
+  [\sigma_1] &:> [\sigma_2]  &(\textnormal{if}~\sigma_1 :> \sigma_2) \\[-0.2em]
+  \label{eq:sub-bot}\tag{P6}
   \sigma &:> \bot  &(\textnormal{for all}~\sigma)\\[-0.2em]
-\label{eq:sub-opt}\tag{B4}
-\kvd{option}\langl\hat{\sigma}\rangl &:> \hat{\sigma}  &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
-\label{eq:sub-opt-cov}\tag{B5}
-\kvd{option}\langl\hat{\sigma_1}\rangl &:> 
-  \kvd{option}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} :> \hat{\sigma_2})\\[-0.2em]
-\label{eq:sub-col}\tag{B6}
-[\sigma_1] &:> [\sigma_2]  &(\textnormal{if}~\sigma_1 :> \sigma_2)
+  \label{eq:sub-var-top}\tag{P7}
+  \kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl &:> \sigma 
 \end{align}
 
-\noindent\boxed{\emph{Variant types}}
-\begin{align}
-\label{eq:sub-var-bot}\tag{U1}
-\kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl :>&~ \sigma \\
-\label{eq:sub-var-eq}\tag{U2}
-\kvd{any}\langl \sigma_1, \ldots, \sigma_n \rangl :>&~ \kvd{any}\langl \sigma_1', \ldots, \sigma_n' \rangl
-\end{align}
-
-\noindent\boxed{\emph{Record types}}
+\noindent
 \begin{align}
@@
 \end{equation*}
-\vspace{-1.75em}
+\vspace{-2em}
 
@@
 \begin{itemize}
-\item Numeric types (\ref{eq:sub-prim}) ar ordered by the range of values they represent;
-  \ident{int} is a 32-bit integer, \ident{decimal} has a range of about $-1^{29}$ to $1^{29}$ and 
-  \ident{float} is a floating-point number.
+\item Numeric types (\ref{eq:sub-prim}) are ordered by the range they represent;
+  \ident{int} is a 32-bit integer, \ident{decimal} has a range of $-1^{29}$ to $1^{29}$ and 
+  \ident{float} is a floating-point number (but imprecise).
 
@@
 \item There is a bottom type (\ref{eq:sub-bot}) and variants behave as top types, because
-  any type $\sigma$ is a subtype of any variant (\ref{eq:sub-var-bot}). There is a range of
-  variants (no \emph{single} top type), but any variant is a subtype of any other variant 
-  (\ref{eq:sub-var-eq}).
-
-\item As usual, the subtyping on records is covariant (\ref{eq:sub-record1}), subtype can have additional 
-  fields (\ref{eq:sub-record2}) and fields can  be reordered (\ref{eq:sub-record3}). The interesting rule 
-  is the last one (\ref{eq:sub-record4}).  Together with covariance, it states that a subtype can omit some 
-  fields, provided that their types are nullable.
+  any type $\sigma$ is a subtype of any variant (\ref{eq:sub-var-top}). There is a range of
+  variants (no \emph{single} top type), but any variant is a subtype of any other variant.
+
+\item As usual, the subtyping on records is covariant (\ref{eq:sub-record1}), subtype can have 
+  additional fields (\ref{eq:sub-record2}) and fields can  be reordered (\ref{eq:sub-record3}). 
+  The interesting rule is (\ref{eq:sub-record4}); together with covariance, it states that a 
+  subtype can omit some fields, provided that their types are nullable.
 \end{itemize}
@@
 If we were type-checking programs, defining variants as a family of top types (which are all 
-subtypes of each other) would not add any value over having a single top type. Our purpose 
-is different. We simplify reading structured data and, as shown in \S\ref{sec:providers-xml}, 
+mutual subtypes) would not add any value over having a single top type, but our purpose is 
+different. We simplify reading structured data and, as shown earlier (\S\ref{sec:providers-xml}), 
 the types forming a variant provide a valuable information about the sample document.
@@
 \begin{figure*}
-\begin{equation*}
-\begin{array}{l}
+\noindent
+\begin{equation*}
+\begin{array}{l}
+\ident{asFloat}(i) \,\reduce f\quad(f = i)\\
+\ident{asFloat}(d) \reduce f\quad(f = d)\\
+\ident{asFloat}(f) \reduce f
+\\[0.5em]
+\ident{asDec}(i) \,\reduce d\quad(d = i)\\
+\ident{asDec}(d) \reduce d
+\end{array}
+\qquad
+\begin{array}{l}
+\ident{hasType}(\nu_? \{ \nu_1 \!:\! \sigma_1, \ldots, \nu_n \!:\! \sigma_n \}, \nu'_? \{ \nu'_1\mapsto s_1, \ldots, \nu'_m\mapsto s_m \}) \reduce (\nu_? = \nu'_?) ~\wedge \\
+  \quad (~ ((\nu_1 = \nu'_1) \wedge \ident{hasType}(\sigma_1, s_1)) \vee\ldots\vee ((\nu_1 = \nu'_m) \wedge \ident{hasType}(\sigma_1, s_m)) \vee \ldots \vee\\
+  \quad ~\; ((\nu_n = \nu'_1) \wedge \ident{hasType}(\sigma_n, s_1)) \vee\ldots\vee ((\nu_n = \nu'_m) \wedge \ident{hasType}(\sigma_n, s_m))~)
+\\[0.5em]
+\ident{hasType}([\sigma], [s_1; \ldots; s_n]) \reduce \ident{hasType}(\sigma, s_1)\wedge\ldots\wedge\ident{hasType}(\sigma, s_n) \\
+\ident{hasType}([\sigma], \kvd{null}) \reduce \kvd{true} \\  
+\end{array}  
+\end{equation*}
+%
+\vspace{-0.5em}
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{isNull}(\kvd{null}) \reduce \kvd{true} \\
+\ident{isNull}(\_) \reduce \kvd{false} 
+\\[0.5em]
 \ident{getField}(\nu,\nu_i, \nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
@@
 \\[0.5em]
-\hspace{-0.5em}
-\begin{array}{l}
-\ident{asDec}(i) \,\reduce d\quad(d = i)\\
-\ident{asDec}(d) \reduce d
-\\[0.5em]
-\ident{isNull}(\kvd{null}) \reduce \kvd{true} \\
-\ident{isNull}(\_) \reduce \kvd{false} 
-\end{array}
-\begin{array}{l}
-\\
-\ident{asFloat}(i) \,\reduce f\quad(f = i)\\
-\ident{asFloat}(d) \reduce f\quad(f = d)\\
-\ident{asFloat}(f) \reduce f\\
-\end{array}
+\ident{getChildren}([s_1; \ldots; s_n]) \reduce [s_1; \ldots; s_n]  \\  
+\ident{getChildren}(\kvd{null}) \reduce [~] 
 \end{array}
@@
 \begin{array}{l}
-\ident{isTag}(\ident{string}, t) \reduce \kvd{true} \\
-\ident{isTag}(\ident{bool}, s) \reduce \kvd{true} \quad(\textnormal{when}~s\in{\kvd{true},\kvd{false}} )\\
-\ident{isTag}(\ident{number}, s) \reduce \kvd{true} \quad(\textnormal{when}~s=i, s=d ~\textnormal{or}~ s=f) \\
-\ident{isTag}(\ident{collection}, [s_1; \ldots; s_n]) \reduce \kvd{true} \\
-\ident{isTag}(\ident{collection}, \kvd{null}) \reduce \kvd{true} \\
-\ident{isTag}(\ident{record}, \{ \nu_1\mapsto s_1, \ldots, \nu_n\mapsto s_n \}) \reduce \kvd{true} \\
-\ident{isTag}(\ident{named}~\nu, \nu\;\{ \nu_1\mapsto s_1, \ldots, \nu_n\mapsto s_n \}) \reduce \kvd{true} \\
-\ident{isTag}(\_, \_) \reduce \kvd{false} \\
-\end{array}
-\end{equation*}
-\vspace{-1em}
-\begin{equation*}
- \hspace{5.6em}
- \ident{getChildren}([s_1; \ldots; s_n]) \reduce [s_1; \ldots; s_n]  \qquad\quad
- \ident{getChildren}(\kvd{null}) \reduce [~] 
+\ident{hasType}(\ident{string}, t) \reduce \kvd{true} \\
+\ident{hasType}(\ident{bool}, s) \reduce \kvd{true} \quad(\textnormal{when}~s\in{\kvd{true},\kvd{false}} )\\
+\ident{hasType}(\ident{float}, s) \reduce \kvd{true} \quad(\textnormal{when}~s=i, s=d ~\textnormal{or}~ s=f) \\
+\ident{hasType}(\ident{decimal}, s) \reduce \kvd{true} \quad(\textnormal{when}~s=i ~\textnormal{or}~ s=f) \\
+\ident{hasType}(\ident{int}, i) \reduce \kvd{true}\\
+\ident{hasType}(\kvd{option}\langl\sigma\rangl, s) \reduce \ident{hasType}(\sigma,s) \\
+\ident{hasType}(\kvd{option}\langl\sigma\rangl, s) \reduce \kvd{true} \\
+\ident{hasType}(\kvd{any}\langl\ldots\rangl, s) \reduce \kvd{true} \\
+\ident{hasType}(\_, \_) \reduce \kvd{false} \\
+\end{array}
 \end{equation*}
@@
 Given two structural types, the \emph{common supertype} relation finds a least upper bound
-(Theorem~\ref{thm:lub}). When possible, it avoid variant types and prefers records 
+(Theorem~\ref{thm:lub}). When possible, it avoids variant types and prefers records 
 (Corollary~\ref{thm:no-unions}), which is important for usability as discussed earlier
@@
 our algorithm finds the common supertype of $\ident{int}$ and $\ident{decimal}$ and produces 
-$\kvd{any}\langl\ident{decimal},\ident{bool}\rangl$.
-
-To identify types that have a common supertype which is not a variant, we group the types by
-a tag. The tag of a type is obtained using a function $\tytagof{(-)} : \sigma \rightarrow \tytag$.
-The function is not defined for $\bot$ and $\kvd{null}$ as those are hanled by the 
-(\emph{bot}) and (\emph{null-1}) or (\emph{null-2}) rules, respectively.
+$\kvd{any}\langl\ident{decimal},\ident{bool}\rangl$. To identify types that have a common supertype 
+which is not a variant, we group the types by a tag. The tag of a type is obtained using a function 
+\tytagof{(-)}. The function is not defined on $\bot$ and $\kvd{null}$ as those are hanled by the 
+(\emph{bot}) and (\emph{null-1}) or (\emph{null-2}).
 
@@
 Finally, (\emph{var-3}) covers the case when we are combining two distinct non-variant types.
-As unions implicitly permit \kvd{null} values, we use an auxiliary function $\dropopt{-}$ to
+As variants implicitly permit \kvd{null} values, we use an auxiliary function $\dropopt{-}$ to
 make nullable types non-nullable (when possible) to simplify the type.
@@
 \paragraph{Properties.}
-
-We say that two types are equivalent if they are mutual subtypes. The partially ordered set of 
-types does not have a \emph{unique} least upper bound, but it does have a least upper bound
-with respect to this equivalence. The common supertype relation is a function 
-(Theorem~\ref{thm:func}) and finds the least upper bound (Theorem~\ref{thm:lub}).
+The partially ordered set of types does not have a \emph{unique} least upper bound, but it does 
+have a least upper bound with respect to an equivalence between mutual subtypes. The common 
+supertype relation is well-defined (Theorem~\ref{thm:func}) and finds the least upper bound 
+(Theorem~\ref{thm:lub}).
 
@@
 reordered fields (of same names and types) and types that contain other equivalent types (e.g.~a
-records with variants as fields).
+lists containing variant values).
 
@@
 of (\emph{order}), (\emph{sym}) and (\emph{refl}). These rules produce types that are subtypes
-of each other and this property is preserved by rules that use the common supertype relation 
-recursively.
+of each other and this property is preserved by rules that use $\triangledown$ recursively.
 \end{proof}
@@
 
-In this section, we build the necessary theoretical framework for proving relativised type safety 
-of our type providers (\S\ref{sec:safety}). We start by discussing the runtime representation of 
-structural values and runtime conversions (\S\ref{sec:formal-convert}), we embed these into a 
-formal model of an F\# subset (\S\ref{sec:formal-ff}) and we describe how our type providers 
-turn an inferred structural type into actual F\# code (\S\ref{sec:formal-tp}).
+In this section, we build the theoretical framework for proving relativised type safety of 
+structural type providers. We discuss the runtime representation of documents and 
+primitive operations (\S\ref{sec:formal-convert}), we embed these into a formal model of an F\# 
+subset (\S\ref{sec:formal-ff}) and we describe how structural type providers turn
+inferred structural types into F\# types (\S\ref{sec:formal-tp}).
 
@@
 
-We represent JSON, XML and CSV documents using the same \emph{structural value}\footnote{Here, we 
-diverge from the actual implementation in F\# Data which uses a different implementation for each 
-format (reusing existing libraries).}. Structural values are first-order and can be one of the 
-following cases:
+We represent JSON, XML and CSV documents using the same \emph{structural value}. Structural 
+values are first-order and can be one of the following cases:
 %
@@
 The first few cases represent primitive values ($i$ for integers, $d$ for decimals, $f$ for floating
-point numbers and $t$ for strings) and the missing value \kvd{null}. A collection is written as a 
-list of values in square brackets, separated by semicolons. A record starts with an optional name 
-$\nu_?$, followed by a sequence of field assignments $\nu_i \mapsto s_i$ written in curly brackets.
+point numbers and $t$ for strings) and the \kvd{null} value. A collection is written as a 
+list of values in square brackets. A record starts with an optional name $\nu_?$, followed by a 
+sequence of field assignments $\nu_i \mapsto s_i$.
 
@@
  op  &\narrow{=}& \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{getChildren}(s) \\
-     &\narrow{|}& \ident{getField}(\nu?, \nu, s) \lsep \ident{isNull}(s) \lsep \ident{isTag}(\ident{tag}, s)
-\end{array}
-\end{equation*}
-%
-The type providers generate code that uses these operations at runtime. Their behavior is defined
-by the reduction rules in Figure~\ref{fig:op-conversions}. The conversion functions (\ident{asDec}
-and \ident{asFloat}) only perform conversions required by the subtyping relation. As noted later
-(\S\ref{?}), our actual implementation is more lax.
+     &\narrow{|}& \ident{getField}(\nu?, \nu, s) \lsep \ident{isNull}(s) \lsep \ident{hasType}(\sigma, s)
+\end{array}
+\end{equation*}
+%
+The operations are used in code generated by type providers. Their behavior is defined
+by the rules in Figure~\ref{fig:op-conversions}. The conversion functions (\ident{asDec}
+and \ident{asFloat}) only perform conversions required by the subtyping relation (our actual 
+implementation is more lax and performs additional conversions).
 
@@
 the actual record name matches the expected name (we write $\bullet$ for the name of an unnamed 
-record). The \ident{getChildren} operation will used to turn \kvd{null} into an empty collction.
+record). The \ident{getChildren} operation returns elements of a list and turns \kvd{null} into 
+an empty collction.
 
 Finally, we also define two helper functions. The \ident{isNull} operation is used to check whether 
-value is \kvd{null} and \ident{isTag} is used to test whether a value can be treated as a value of 
-a type associated with the specified tag (as defined in Figure~\ref{fig:subtyping-cst}).
-The last line defines a ``catch all'' pattern, which returns \kvd{false} for all remaining cases.
+value is \kvd{null} and \ident{hasType} is a runtime type test. The most complex case is handling 
+of records where we check that for each field $\nu_1, \ldots, \nu_n$ in the type, the actual record 
+value has a field of the same name with a matching type. The last line defines a ``catch all'' 
+pattern, which returns \kvd{false} for all remaining cases. Although not elenegantly, the rules 
+can be expressed using just Featherweight F\#, discussed in the next section.
 
@@
 \begin{figure}
+\noindent
 \begin{equation*}
@@
  { (\kvd{new}~C(\overline{v})).N_i \reduce e_i[\overline{x} \leftarrow \overline{v}] }\\
+ \\
+ \textnormal{\footnotesize{(eq1)}}&
+ v=v\reduce\kvd{true} \qquad \textnormal{\footnotesize{(eq2)}}~~v=v'\reduce\kvd{false}\\
  \\
@@
  \hspace{-0.4em}
- \ident{List.map}~(\lambda x.e)~[v_1; \ldots] ~\reduce~ [e[x\leftarrow v_1]; \ldots] \\
+ \ident{map}~(\lambda x.e)~[v_1; \ldots] ~\reduce~ [e[x\leftarrow v_1]; \ldots] \\
  \\
@@
 
-The semantics fragment in the previous section discusses values and operations that are used by
-F\# Data. This section adds a minimal subset of F\#. Type providers provide classes and so we focus
-on the F\# object model and combine aspects of standard ML \cite{sml} and Featherweight Java \cite{fwjava}.
+We now extend the semantics discussed so far and add a minimal subset of F\# needed to model F\# 
+Data. Type providers use classes and so we focus on the F\# object model and combine aspects of SML 
+\cite{sml} and Featherweight Java \cite{fwjava}.
 
@@
 of ML, we only pick constructs for working with options and lists that we need later.
-
+%
 \begin{equation*}
@@
  L &\narrow{=}& \kvd{type}~C(\overline{x:\tau}) = \overline{M} \\[0.0em]
- M &\narrow{=}& \kvd{member}~N:\tau=e \\[0.6em]
+ M &\narrow{=}& \kvd{member}~N:\tau=e
+\end{array}
+\end{equation*}
+
+\noindent
+\begin{equation*}
+\begin{array}{rcl}
  v &\narrow{=}& s \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \lsep [v_1; \ldots; v_n] \\[0.0em]
@@
    &\narrow{|}& \ident{None} \lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
-   &\narrow{|}& \ident{Some}(e) \lsep [e_1;\ldots;e_n]\lsep \ident{List.map}~(\lambda x\rightarrow e_1)~e_2\\
-   &\narrow{|}& \kvd{match}~e~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2
+   &\narrow{|}& \ident{Some}(e) \lsep [e_1;\ldots;e_n]\lsep \ident{map}~(\lambda x\rightarrow e_1)~e_2\\
+   &\narrow{|}& e = e \lsep \kvd{match}~e~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2
 \end{array}
@@
 conditionals and expressions for working with option values and lists. We include 
-\ident{List.map} as a special construct to avoid making the language too complex.
+\ident{map} as a special construct to avoid making the language too complex.
 
@@
 \begin{array}{rcl}
- E &\narrow{=}& [\overline{v};E;\overline{e}] \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e}) \lsep \ident{Some}(E) \\[0.1em]
-   &\narrow{|}& \ident{map}~(\lambda x\rightarrow e)~E \lsep \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2 \lsep f(E)\\[0.1em]
+ E &\narrow{=}& [\overline{v};E;\overline{e}] \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e})\\[0.1em]
+   &\narrow{|}& \ident{map}~(\lambda x\rightarrow e)~E \lsep \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2 \\[0.1em]
+   &\narrow{|}& \ident{Some}(E) \lsep f(E) \lsep E = e \lsep v = E \\[0.1em]
    &\narrow{|}& \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
    &\narrow{|}& \kvd{match}~E~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2
-\end{array} \\[2.5em]
-f \in \{ \ident{asDec}, \ident{asFloat}, \ident{getField}, \ident{getChildren}, \ident{isNull}, \ident{isTag} \}
+\end{array} \\[3em]
+f \in \{ \ident{asDec}, \ident{asFloat}, \ident{getField}, \ident{getChildren}, \ident{isNull}, \ident{hasType} \}
 \end{array}
@@
 We write $e[\overline{x} \leftarrow \overline{v}]$ for the result of replacing variables $\overline{x}$ by
-values $\overline{v}$ in expression. The (\emph{member}) rule reduces a member access using a class 
+values $\overline{v}$ in an expression. The (\emph{member}) rule reduces a member access using a class 
 definition in the assumption to obtain the body of a member. The remaining six rules
@@
 
-The language is simple, but sufficient for our purpose. All expressions reduce to a value in a 
+We omit expressions $e \vee e$ and $e \wedge e$ that are used in Figure~\ref{fig:op-conversions}, 
+but those can be easily defined as syntactic sugar using \kvd{if}. All expressions reduce to a value in a 
 finite number of steps or get stuck due to an error condition. An error condition can be a wrong 
-argument passed to conditional, pattern matching or one of the conversion functions from 
-Figure~\ref{fig:op-conversions}.
+argument passed to conditional, pattern matching or one of the primitives from Figure~\ref{fig:op-conversions}.
 
@@
 \begin{figure}
+\noindent  
 \begin{equation*}
@@
 An important part of Featherweight Java that is omitted here is the checking of type declarations
-(ensuring the bodies of members are well-typed). We omit this, because we consider only classes 
-generated by our type providers (which are correct by construction).
+(ensuring the members are well-typed). We consider only classes generated by our type providers 
+and those are well-typed by construction.
 
@@
 \begin{figure*}
+\noindent
 \begin{equation*}
@@
  \sem{\,[\sigma]\,}_e = \ident{list}\langl\tau\rangl, e_b, L \;\;\textnormal{where}\\[0.4em]
- \qquad e_b = \ident{List.map}~(\lambda x\rightarrow e')~(\ident{getChildren}(e)) \\[0.1em]
+ \qquad e_b = \ident{map}~(\lambda x\rightarrow e')~(\ident{getChildren}(e)) \\[0.1em]
  \qquad \tau, e', L = \sem{\hat{\sigma}}_x
@@
  \qquad M_i = \kvd{member}~\nu_i:\ident{option}\langl\tau_i\rangl=\\[0.1em]
- \hspace{5.8em}  \kvd{if}~\ident{isTag}(t_i, v)~\kvd{then}~
+ \hspace{5.8em}  \kvd{if}~\ident{hasType}(\sigma_i, v)~\kvd{then}~
      \ident{Some}(e_i)~\kvd{else}~\ident{None} \\[0.1em]
@@
 or more sample documents (\S\ref{sec:inference}) and we defined a simplified model of evaluation
-of F\# language (\S\ref{sec:formal-ff}) and F\# Data runtime (\S\ref{sec:formal-tp}). Next, we 
+of F\# (\S\ref{sec:formal-ff}) and F\# Data runtime (\S\ref{sec:formal-tp}). In this section, we 
 define how the type providers work, linking the two parts.
@@
 All F\# Data type providers take (one or more) sample documents, infer a common supertype $\sigma$
-and then use $\sigma$ to generate F\# types that are exposed to the programmer\footnote{The actual 
+and then use it to generate F\# types that are exposed to the programmer\footnote{The actual 
 implementation provides \emph{erased types} as described in \cite{fsharp-typeprov}. Here, we treat 
@@
 \paragraph{Type provider mapping.}
-When generating types, the type provider produces an F\# type $\tau$, an expression that wraps 
-the input document (of type \ident{StructVal}) as a value of type $\tau$ and a collection of class 
-definitions. We express it using the following mapping:
+The type provider produces an F\# type $\tau$ together with an expression that wraps an input
+value (of type \ident{StructVal}) as a value of type $\tau$ and a collection of class definitions. 
+We express it using the following mapping:
 %
@@
 %
-The mapping $\sem{\sigma}_e$ takes a inferred structural type $\sigma$ and an expression $e$, 
+The mapping $\sem{\sigma}_e$ takes an inferred structural type $\sigma$ and an expression $e$, 
 which represents code to obtain a structural value that is being wrapped. It returns an F\# type 
@@
 Figure~\ref{fig:tp-generation} shows the rules that define $\sem{-}_{-}$. Primitive types are all
-handled by a single rule. For a given structural type, it returns the corresponding F\# type -- the types are
-mapped directly with the exception of \ident{bit}, which becomes an F\# \ident{bool}. The generated code calls
-an appropriate conversion function from Figure~\ref{fig:op-conversions} on the input.
-
-Handling of records is more interesting. We generate a new class $C$ that takes a structural value as constructor
+handled by two rules -- for \ident{decimal} and \ident{float}, we insert a call to an appropriate 
+conversion function from Figure~\ref{fig:op-conversions}.
+Handling of records is more interesting. We generate a class $C$ that takes a structural value as constructor
 parameter. For each record field, we generate a new member with the same name as the field\footnote{The actual
-F\# Data implementation also capitalizes the names.} The body of the member calls \ident{getField} and then
-passes this expression to $\sem{\sigma_i}$ which adds additional wrapping that maps the field (structural value
-of type $\sigma_i$) into an F\# type $\tau_i$. The returned expression creates a new instance of $C$ and
-the mapping returns the class $C$ together with all recursively generated definitions.
+F\# Data implementation also capitalizes the names.}. The body of the member calls \ident{getField} and then
+wraps this expression using $\sem{\sigma_i}$ which maps the field (structural value of type $\sigma_i$) into 
+an F\# value of type $\tau_i$. The returned expression creates a new instance of $C$ and the mapping returns 
+the class $C$ together with all recursively generated definitions. Note that the class
+name $C$ is not directly accessed by the user and so we can use arbitrary name, although the actual 
+implementation in F\# Data attempts to infer a reasonable name\footnote{For example, in 
+\ident{\{\str{person}:\{\str{name}:\str{Tomas}\}\}}, the nested record will be named \ident{Person}
+based on the name of the parent record field.}.
 
 A collection type becomes an F\# $\ident{list}\langl\tau\rangl$. The returned expression calls \ident{getChildren}
-(which turns \kvd{null} values into empty lists) and then uses \ident{List.map} to convert all nested values to 
-an F\# type $\tau$. The handling of option type is similar -- it checks if the original value is \kvd{null} and
-if no, it wraps the recursively generated conversion expression $e'$ in the \ident{Some} constructor.
-
-As discussed earlier, union types are also generated as classes with properties. Given a union type $\sigma_1 + \ldots + \sigma_n$,
-we get corresponding F\# types $\tau_i$ and generate $n$ members of type $\ident{option}\langl \tau_i\rangl$.
-Each member return \ident{Some} when the value is not \kvd{null} and has the right structure (checked by 
-\ident{isTag}). The type inference algorithm also guarantees that there is only one case for each type tag 
-(Section~\ref{sec:inference-commonsuper}) and there are no nested union types. Thus, checking for a tag is
-sufficient and we can also use the tag to identify the name of the generated member (using the
-\ident{nameof} function).
+(which turns \kvd{null} into an empty list) and then uses \ident{map} to convert nested values to 
+values of an F\# type $\tau$. The handling of option type is similar; if the value is not \kvd{null},
+we wrap the recursively generated conversion expression $e'$ in the \ident{Some} constructor.
+
+As discussed earlier, variant types are also generated as F\# classes with properties. Given a variant
+$\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl$, we get corresponding F\# types $\tau_i$ and generate 
+$n$ members of type $\ident{option}\langl \tau_i\rangl$. When the member is accessed, we need to perform
+a runtime type test using \ident{hasType} to ensure that the value has the right type (similarly to runtime 
+type conversions from the top type in languages like Java). If the type matches, a \ident{Some} value is 
+returned. The type inference algorithm also guarantees that there is only one case for each type tag 
+(\S\ref{sec:inference-commonsuper}) and so we can use the tag when naming the generated member 
+(using the \ident{nameof} function).
 
 \paragraph{Example 1.}
-To illustrate how the type provision mechanism works, we now consider two simple examples. First, assume 
-that the inferred type is a record with two fields (one optional) such as 
-$\ident{Person}~\{~\ident{Age}:\ident{int}~\kvd{option}, \ident{Name}:\ident{string}~ \}$. Applying the
-rules from Figure~\ref{fig:tp-generation} produces the following class:
+To illustrate how the mechanism works, we consider two examples. First, assume 
+that the inferred type is a record  
+$\{~\ident{Age}\!:\!\kvd{option}\langl\ident{int}\rangl, \ident{Name}\!:\!\ident{string}~ \}$. 
+The rules from Figure~\ref{fig:tp-generation} produce the following class:
 %
@@
 $\ident{getField}(\ident{Person},\ident{Age},v)$. If the returned field is not \kvd{null}, then the member
-calls \ident{asInt} (produced by the case for primitive types) and wraps the result in the \ident{Some}
-constructor. Note that \ident{getField} is defined even when the field does not exist, but returns \ident{null}.
-This lets us treat missing fields as optional fields. The \ident{Name} member is similar, but does not 
-perform any checks. 
-For completeness, a type corresponding to the record is \ident{Person} and given a 
-structural value $s$, we create a \ident{Person} value by calling $\kvd{new}~\ident{Person}(s)$.
-
-\paragraph{Example 2.} The second example illustrates the remaining parts of the type provision,
-including collections and union types. Reusing the \ident{Person} type from the previous example,
-consider a collection $[\ident{Person}+\ident{string}]$, which contains a mix of \ident{Person} and string values.
-
-\noindent
-\begin{equation*}
-\begin{array}{l}
- \kvd{type}~\ident{PersonOrString}(\ident{v}:\ident{StructVal})~= \\[0.1em]
+calls \ident{asInt} (produced by the case for primitive types) and wraps the result in \ident{Some}.
+Note that \ident{getField} is defined even when the field does not exist, but it returns 
+\kvd{null}. This lets us treat missing fields as optional. An F\# type corresponding to the (structural)
+record is \ident{Person} and given a structural value $s$, we create a \ident{Person} value by calling 
+$\kvd{new}~\ident{Person}(s)$.
+
+\paragraph{Example 2.} The second example illustrates the remaining types including collections and 
+variants. Reusing \ident{Person} from the previous example, consider 
+$[\kvd{any}\langl\ident{Person},\ident{string}\rangl]$, a collection which contains a mix 
+of \ident{Person} and string values:
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{type}~\ident{PersonOrString}(v:\ident{StructVal})~= \\[0.1em]
  \quad \kvd{member}~\ident{Person}~:~\ident{option}\langl\ident{Person}\rangl~= \\[0.1em]
- \qquad \kvd{if}~\ident{isNull}(v)~\kvd{then}~\ident{None}~\kvd{else} \\[0.1em]
- \qquad \kvd{if}~\ident{isTag}(\ident{rec-named Person}, v)~\kvd{then}~\\[0.1em]
- \qquad\quad \ident{Some}(\kvd{new}~\ident{Person}(v))~\kvd{else}~\ident{None} \\[0.1em]
+ \qquad \kvd{if}~\ident{hasType}(\{\ident{Age}\!:\!\kvd{option}\langl\ident{int}\rangl, \ident{Name}\!:\!\ident{string}\}, v)~\\[0.1em]
+ \qquad\quad \kvd{then}~\ident{Some}(\kvd{new}~\ident{Person}(v))~\kvd{else}~\ident{None} \\[0.1em]
  \quad \kvd{member}~\ident{String}~:~\ident{option}\langl\ident{string}\rangl~= \\[0.1em]
- \qquad \kvd{if}~\ident{isNull}(v)~\kvd{then}~\ident{None}~\kvd{else} \\[0.1em]
- \qquad \kvd{if}~\ident{isTag}(\ident{string}, v)~\kvd{then}~\\[0.1em]
- \qquad\quad \ident{Some}(\ident{asStr}(v))~\kvd{else}~\ident{None}
-\end{array}
-\end{equation*}
-
-\noindent
-The type provider generates the above type and the collection type is mapped to an F\# type 
-$\ident{list}\langl\ident{PersonOrString}\rangl$. Given a structural document value $s$, the code
-to obtain the wrapped F\# value is:
-%
-\begin{equation*}
-\ident{List.map}~(\lambda x\rightarrow\kvd{new}~\ident{PersonOrString}(x))~(\ident{getChildren}(s))
-\end{equation*}
-%
-The \ident{PersonOrString} type contains one member for each of the union case. In the body, they
-check that the value is not \kvd{null} and that it has the right structure (using the \ident{isTag}
-function). This checks that the value is a record named \ident{Person} or a string, respectively.
-If the conditions are satisfied, the value is converted to the F\# type corresponding to the case
-and wrapped in \ident{Some}.
+ \qquad \kvd{if}~\ident{hasType}(\ident{string}, v)~\\[0.1em]
+ \qquad\quad \kvd{then}~\ident{Some}(\ident{asStr}(v))~\kvd{else}~\ident{None}
+\end{array}
+\end{equation*}
+%
+The type provider maps the structural type to an F\# type $\ident{list}\langl\ident{PersonOrString}\rangl$. 
+Given a structural document value $s$, the code to obtain the wrapped F\# value is:
+%
+\begin{equation*}
+\ident{map}~(\lambda x\rightarrow\kvd{new}~\ident{PersonOrString}(x))~(\ident{getChildren}(s))
+\end{equation*}
+%
+The \ident{PersonOrString} type contains one member for each of the variant case. In the body, they
+check that the value has the correct type using \ident{hasType}. This also implicitly handles \kvd{null}
+by returning \kvd{false}. As discussed earlier, variants provide easy access to the known cases
+(\ident{string} or \ident{Person}), but they are robust with respect to unexpected inputs.
 
@@
 
-Before concluding the section on formalizing type providers, there is one more missing piece. The
-common supertype algorithm in Section~\ref{sec:inference} discusses the core of the type inference for
-structural values, but we have not yet clarified how exactly it is used. We address this in the present
-section.
-
-Given a JSON, XML or CSV document, the F\# Data implementation constructs a structural value $s$
-from the sample (this is straightforward, but Section~\ref{sec:impl-parsing} discusses some interesting 
-aspects). The following defines a mapping $\semalt{-}$ which turns a sample value $s$ into a structural
-type $\sigma$:
-%
-\begin{equation*}
-\begin{array}{rclcrcl}
- \semalt{0} &\narrow{=}& \ident{bit} &\qquad& \semalt{1} &\narrow{=}& \ident{bit} \\
- \semalt{i} &\narrow{=}& \ident{int} && \semalt{d} &\narrow{=}& \ident{decimal} \\
- \semalt{f} &\narrow{=}& \ident{float} && \semalt{t} &\narrow{=}& \ident{string} \\
- \semalt{b} &\narrow{=}& \ident{bool} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} \\[0em]
-\end{array}
-\end{equation*}
-\noindent
-\vspace{-0.2em}
-\begin{equation*}
-\begin{array}{l}
+The common supertype algorithm (\S\ref{sec:inference}) is at the core of the type inference 
+for structural values, but we have not yet clarified how it is used. Given a JSON, XML or CSV 
+document, F\# Data constructs a structural value $s$ from the sample. The following defines a mapping 
+$\semalt{s_1,\ldots,s_n}$ which turns a collection of sample values $s_1, \ldots, s_n$ into a 
+structural type $\sigma$:
+
+\noindent
+\begin{equation*}
+\begin{array}{rclcrclcrcl}
+ \semalt{i} &\narrow{=}& \ident{int} && \semalt{d} &\narrow{=}& \ident{decimal} && \semalt{b} &\narrow{=}& \ident{bool}\\
+ \semalt{f} &\narrow{=}& \ident{float} && \semalt{t} &\narrow{=}& \ident{string} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null}\\
+\end{array}
+\end{equation*}
+\noindent
+\vspace{-0.5em}
+\begin{equation*}
+\begin{array}{l}
+ \semalt{s_1, \ldots, s_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
+ \qquad\sigma_0 = \bot,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{s_i} \vdash \sigma_i
+ \\[0.5em]
  \semalt{[s_1; \ldots; s_n]} = [\semalt{s_1, \ldots, s_n}]
@@
  \qquad\nu?~\{ \nu_1:\semalt{s_1}, \ldots, \nu_n :\semalt{s_n} \}
- \\[0.5em]
- \semalt{s_1, \ldots, s_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
- \qquad\sigma_0 = \top,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{s_i} \vdash \sigma_i
-\end{array}
-\end{equation*}
-%
-Primitive values are mapped to their corresponding types, with the exception of $0$ and $1$, which
-are inferred as \ident{bit}. For records, we return a type with field types inferred based on 
-individual values.
-
-The interesting part is inferring type based on multiple samples. We overload
-the notation and write $\semalt{s_1, \ldots, s_n}$ for a type inferred from multiple samples.
-This uses the common supertype relation to find a common type for all values (starting with $\top$).
-This operation is used at the top-level (when calling type provider with multiple samples)
-and also when inferring the type of a collection.
+\end{array}
+\end{equation*}
+%
+We overload the notation and write $\semalt{s}$ when inferring type from a single value. Primitive 
+values are mapped to their corresponding types. For records, we return a type with field types 
+inferred based on individual values.
+
+When infering a value from multiple samples, we use the common supertype relation to find a common 
+type for all values (starting with $\bot$). This operation is used at the top-level (when calling 
+a type provider with multiple samples) and also when inferring the type of a collection.
 
@@
 
-Informally, the safety property of F\# Data type providers states that, given representative sample
+Informally, the safety property for structural type providers states that, given representative sample
 documents, any code that can be written using the provided types is guaranteed to work. We call this 
-\emph{relativized safety}, because we cannot avoid \emph{all} errors. In particular, the user can 
-always use the type provider with an input that has a different structure than any of the samples -- and in 
-this case, it is expected that the code will fail at runtime (throw an exception in the actual 
-implementation or by getting stuck in our model).
-
-The key question is, what is a representative sample? Given a set of sample documents, 
-the provided type is guaranteed to work if the inferred type of the input is a subtype of any of the
-sample documents. Going back to Section~\ref{sec:inference-subtyping}, this means that:
+\emph{relativized safety}, because we cannot avoid \emph{all} errors. In particular, one can always
+provide an input that has a different structure than any of the samples. In this case, it is expected 
+that the code will throw an exception in the implementation (or get stuck in our model).
+
+More formally, given a set of sample documents, code using the provided type is guaranteed to work if 
+the inferred type of the input is a subtype of any of the samples. Going back to 
+\S\ref{sec:inference-subtyping}, this means that:
 %
 \begin{itemize}[noitemsep]
-\item[--] Input can contain smaller numerical values (for example, if a sample contains float, the input can contain an integer).
-\item[--] Records in the actual input can have additional fields
-\item[--] Records in the actual input can have fewer fields, provided that the type of the fields is marked as optional in the sample
-\item[--] Union types in the input can have both fewer or more cases
-\item[--] When we have a union type in the sample, the actual input can also contain just values of one of the union cases
+\item[--] Input can contain smaller numerical values (e.g., if a sample contains float, the input can contain an integer)
+\item[--] Records in the input can have additional fields
+\item[--] Records in the input can have fewer fields, provided that the type of the field is missing in some of the samples
+\item[--] When a variant is inferred from the sample, the actual input can also contain any other value  
 \end{itemize}
@@
 The following lemma states that the provided code (generated in Figure~\ref{fig:tp-generation})
-works correctly on inputs $s'$ that is a subtype of the sample $s$. More formally, we require that the 
-provided expression (using $s'$ as the input) can be reduced to a value and, if it is a class,
+works correctly on an input $s'$ that is a subtype of the sample $s$. More formally, the provided
+provided expression (with input $s'$) can be reduced to a value and, if it is a class,
 all its members can also be reduced to values.
@@
 \begin{proof}
-By induction over the structure of $\sem{-}_e$. For primitives, the conversion functions accept all subtypes.
-For other cases, analyse the the provided code to see that it can work on all subtypes (\emph{e.g.}~\ident{getChildren}
-works on \kvd{null} values, \ident{getField} returns \kvd{null} when a field is missing and, when \ident{isTag} returns
-\ident{true}, then a value has the structure required by the corresponding case).
+By induction over the structure of $\sem{-}_-$. For primitives, the conversion functions accept all subtypes.
+For other cases, analyse the provided code to see that it can work on all subtypes (\emph{e.g.}~\ident{getChildren}
+works on \kvd{null} values, \ident{getField} returns \kvd{null} when a field is missing and, for variants,
+the value has the correct type as guaranteed by \ident{hasType}.
 \end{proof}
@@
 \noindent
-Now that we know that the provided types are correct with respect to the subtyping relation, we can
-look at the main theorem of the paper. It states that, for any input (which is a subtype of any of the
-samples) and any expression $e$, a well-typed program that uses the provided types does not ``go wrong''.
-
-Using the standard approach to syntactic type safety  \cite{syntactic}, we prove the type preservation 
+This shows that the provided types are correct with respect to the subtyping relation. Next, the
+main theorem of the paper states that, for any input (which is a subtype of any of the samples) and 
+any expression $e$, a well-typed program that uses the provided types does not ``go wrong''.
+
+Using the standard syntactic type safety  \cite{syntactic}, we prove type preservation 
 (reduction does not change type) and progress (an expression that is not a value can be reduced).
@@
 
-Then for all new inputs $s'$ such that $\exists i.(\semalt{s_i} :> \semalt{s'})$, let $e_s=e[x\leftarrow s']$
+For all new inputs $s'$ such that $\exists i.(\semalt{s_i} :> \semalt{s'})$, let $e_s=e[x\leftarrow s']$
 be an expression (of type $\tau$) that wraps the input in a provided type. Then, for any expression $e_c$
-(user code) such that $\emptyset; y:\tau \vdash e_c:\tau'$ and that does not contain any structural values
-as sub-expressions, it is the case that $e_c[y\leftarrow e_s] \reduce^{*} v$ for some value $v$ and
+(user code) such that $\emptyset; y:\tau \vdash e_c:\tau'$ and $e_c$ does not contain primitive operations
+$op$ as a sub-expression, it is the case that $e_c[y\leftarrow e_s] \reduce^{*} v$ for some value $v$ and
 also $\emptyset; \vdash v : \tau$.
@@
 The progress lemma states that evaluation of a well-typed program does not reach an undefined state. 
-This is not a problem for the ML subset and object-oriented subset of the calculus. The problematic
-part are the conversion functions (Figure~\ref{fig:op-conversions}). Given a structural value 
-(which has a type \ident{StructVal} in our language), the reduction can get stuck if the value does
-not have a structure required by a specific conversion function used.
-
-The Lemma~\ref{thm:tp-correctness} guarantees that this does not happen. In Theorem~\ref{thm:safety},
-we carefully state that we only consider expressions $e_c$ which ``[do] not contain any structural 
-values as sub-expressions''. This makes sure that the only code working with structural values is the
-code generated by the type provider.
+This is not a problem for the (standard) ML subset and object-oriented subset of the calculus. The 
+problematic part are the primitive functions (Figure~\ref{fig:op-conversions}). Given a structural 
+value (of type \ident{StructVal}), the reduction can get stuck if the value does not have a structure 
+required by a specific primitive.
+
+The Lemma~\ref{thm:tp-correctness} guarantees that this does not happen inside the provided type.
+In Theorem~\ref{thm:safety}, we carefully state that we only consider expressions $e_c$ which 
+``[do] not contain primitive operations $op$ as sub-expressions''. This makes sure that the only 
+code working with structural values is the code generated by the type provider.
 
@@
 Given the assumptions and definitions from Theorem~\ref{thm:safety}, it is the case that
-$e_c[y\leftarrow e_s] \reduce^{*} v$.
+$e_c[y\leftarrow e_s] \reduce e_c'$.
 \end{lemma}
 \begin{proof}
-By induction over the typing derivation of $L; \emptyset \vdash e_c[y\leftarrow e_s] : \tau'$. 
+Proceed by induction over the typing derivation of $L; \emptyset \vdash e_c[y\leftarrow e_s] : \tau'$. 
 The cases for the ML subset are standard. For member access, we rely on Lemma~\ref{thm:tp-correctness}.
@@
 
-% -------------------------------------------------------------------------------------------------
+% ==================================================================================================
+%
+%   ###                                                                                     
+%    #  #    # #####  #      ###### #    # ###### #    # #####   ##   ##### #  ####  #    # 
+%    #  ##  ## #    # #      #      ##  ## #      ##   #   #    #  #    #   # #    # ##   # 
+%    #  # ## # #    # #      #####  # ## # #####  # #  #   #   #    #   #   # #    # # #  # 
+%    #  #    # #####  #      #      #    # #      #  # #   #   ######   #   # #    # #  # # 
+%    #  #    # #      #      #      #    # #      #   ##   #   #    #   #   # #    # #   ## 
+%   ### #    # #      ###### ###### #    # ###### #    #   #   #    #   #   #  ####  #    # 
+%
+% ==================================================================================================
+
+\section{Practical experience}
+\label{sec:impl}
+
+The F\# Data library has been widely adopted by the industry and is one of the most downloaded
+F\# libraries\footnote{At the time of writing, the library has over 80,000 downloads on NuGet 
+(package repository), 1,821 commits and 44 contributors on GitHub.}. A practical demonstration of 
+development using the library can be seen in an attached video and additional documentation can be
+found at \url{http://fsharp.github.io/FSharp.Data}.
+
+In this section, we discuss our practical experience with the safety guarantees provided by the
+F\# Data type providers and other notable aspects of the implementation.
 
@@
 
-The \emph{relativized safety} property does not guarantee the same amount of safety as standard type safety
-for programming languages without type providers. However, it reflects the reality of programming with external
-data sources that is increasingly important in the age of web \cite{age-of-web}. So, type providers do not
-reduce the safety -- they simply make the existing issues visible.
+The \emph{relativized safety} property does not guarantee the same amount of safety as standard 
+ML type safety. However, it reflects the reality of programming with external data sources that is 
+increasingly important \cite{age-of-web}. Type providers do not reduce the safety -- they simply 
+illuminate the existing issues.
 
@@
 
-% ==================================================================================================
-%
-%   ###                                                                                     
-%    #  #    # #####  #      ###### #    # ###### #    # #####   ##   ##### #  ####  #    # 
-%    #  ##  ## #    # #      #      ##  ## #      ##   #   #    #  #    #   # #    # ##   # 
-%    #  # ## # #    # #      #####  # ## # #####  # #  #   #   #    #   #   # #    # # #  # 
-%    #  #    # #####  #      #      #    # #      #  # #   #   ######   #   # #    # #  # # 
-%    #  #    # #      #      #      #    # #      #   ##   #   #    #   #   # #    # #   ## 
-%   ### #    # #      ###### ###### #    # ###### #    #   #   #    #   #   #  ####  #    # 
-%
-% ==================================================================================================
-
-\section{Implementation}
-\label{sec:impl}
+\subsection{Practical experiences}
+
+An important concern about using F\# Data in practice is the handling of schema change. When using a
+type provider, the sample is captured at compile-time. If the schema changes later (so that the actual
+input is no longer a subtype of the samples), the program using the type provider fails at run-time
+and it is the developer's responsibility to handle the exception. However, this is the same problem that
+happens when reading data using any other library.
+
+F\# Data can help discover such errors earlier. The example in Section~\ref{sec:introduction}
+points the JSON type provider to a sample using a live URL. This has the advantage that a re-compilation 
+fails when the schema changes, which is an indication that the program needs to be updated to reflect the
+change. If this is undesirable, it is always possible to cache the sample locally.
+
+In general, there is no better solution for plain XML, CSV and JSON data sources. Some data sources provide 
+versioning support (with meta-data about how the schema changed). For those, a type 
+provider could adapt automatically, but we leave this for future.
+
+An approach that we find useful in practice is to keep a set of samples. When the program fails at run-time
+(because a service returned data in an unexpected format), the new input can be added as an additional sample,
+which then shows what parts of code need to be modified.
+
+\subsection{Reading CSV files}
+\label{sec:providers-csv}
+
+In the CSV file format, the structure is a collection of rows (records) consisting of fields 
+(with names specified by the first row). The inference needs to infer the types of fields.
+For example:
+%
+{\small{
+\begin{verbatim}
+  Ozone, Temp, Date,       Autofilled
+  41,    67,   2012-05-01, 0
+  36.3,  72,   2012-05-02, 1
+  12.1,  74,   3 kveten,   0
+  17.5,  #N/A, 2012-05-04, 0
+\end{verbatim}
+}}
+%
+\noindent
+One difference between JSON and CSV formats is that in CSV, the literals have no data types.
+In JSON, strings are \str{quoted} and Booleans are \kvd{true} and \kvd{false}. This is not the
+case for CSV and so we need to infer not just the structure, but also the primitive types.
+
+Assuming the sample is saved as \strf{airdata.csv}, the following snippet prints all
+rows from another file that were not autofilled:
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{type}~\ident{AirCsv}~=~\ident{CsvProvider}\langl\str{airdata.csv}\rangl\hspace{1em} \\[0.6em]
+ \kvd{let}~\ident{air}~=~\ident{AirCsv.Parse}(\ident{data})\\[0.1em]
+ \kvd{for}~\ident{row}~\kvd{in}~\ident{air.Rows}~\kvd{do}\\[0.1em]
+ \quad\kvd{if}~\ident{not}~\ident{row.Autofilled}~\kvd{then}\\[0.1em]
+ \quad\quad \ident{printf}~\str{\%s: \%d}~\ident{row.Date row.Ozone}\\
+\end{array}
+\end{equation*}
+%
+The type of the record (\ident{row}) is, again, inferred from the sample. The \ident{Date} column
+uses mixed formats and is inferred as \ident{string} (although we support many date formats and 
+``May 3'' would work). More interestingly, we also infer \ident{Autofilled} as Boolean, because 
+the sample contains only $0$ and $1$ values and using those for Booleans is a common CSV convention.
+Finally, the fields \ident{Ozone} and \ident{Temp} have types \ident{decimal} and
+$\ident{option}\langl\ident{int}\rangl$.
 
@@
 
-% -------------------------------------------------------------------------------------------------
+% ==================================================================================================
+%
+%    ######
+%    #     # ###### #        ##   ##### ###### #####     #    #  ####  #####  #    #
+%    #     # #      #       #  #    #   #      #    #    #    # #    # #    # #   #
+%    ######  #####  #      #    #   #   #####  #    #    #    # #    # #    # ####
+%    #   #   #      #      ######   #   #      #    #    # ## # #    # #####  #  #
+%    #    #  #      #      #    #   #   #      #    #    ##  ## #    # #   #  #   #
+%    #     # ###### ###### #    #   #   ###### #####     #    #  ####  #    # #    #
+%
+% ==================================================================================================
+
+\section{Related and future work}
+\label{sec:related}
+
+The F\# Data library connects two lines of research that have been previously disconnected. The first is 
+extending the type systems of programming languages to accommodate external data sources and the second
+is inferring types for real-world data sources.
+
+The type provider mechanism has been introduced in F\# \cite{fsharp-typeprov,fsharp-typeprov-ddfp}, 
+added to Agda \cite{agda-tp} and used in areas such as semantic web \cite{liteq}. Our paper is novel 
+in that it shows the programming language theory behind a concrete type providers.
+
+\paragraph{Extending the type systems.} 
+A number of systems integrate external data formats into a programming language. Those include 
+XML \cite{xduce,xduce-ml} and databases \cite{links}. In both of these, the system either requires
+the user to explicitly define the schema (using the host language) or it has an ad-hoc extension 
+that reads the schema (\emph{e.g.}~from a database). LINQ \cite{linq} is more general, but relies
+on code generation when importing the schema.
+
+The work that is most similar to F\# Data is the XML and SQL integration in C$\omega$ \cite{comega-xs}.
+It extends a C\# like object-oriented language with types similar to our structural types 
+(including nullable types, choices with subtyping and heterogeneous collections with multiplicities).
+However, C$\omega$ does not infer the types from samples and extends the type system of the host
+language (rather than using a general purpose embedding mechanism).
+
+\paragraph{Advanced type systems.}
+Aside from type providers, a number of other advanced type system features could be used to
+tackle the problem discussed in this paper. The Ur \cite{ur} language has a rich system for working
+with records; meta-programming \cite{template-hask}, \cite{th-camlp4} and multi-stage programming \cite{multi-stage}
+could be used to generate code for the provided types. However, as far as we are aware, none of these 
+systems have been used to provide the same level of integration with XML, CSV and JSON.
+
+Another approach would be to use gradual typing \cite{gradual,gradual-js} and add types for structured
+data formats to existing dynamic language to check, for example, JSON manipulation in JavaScript.
+
+\paragraph{Typing real-world data.}
+The second line of research related to our work focuses on inferring structure of real-world data sets.
+A recent work on JSON \cite{typing-json} infers a succinct type using MapReduce to handle large number
+of samples. It fuses similar types based on a type similarity measure. This is more sophisticated than
+our technique, but it would make formally specifying the safety properties (Theorem~\ref{thm:safety}) difficult.
+Extending our \emph{relativized safety} property to a \emph{probabilistic safety} is an interesting 
+future work.
+
+The PADS project \cite{pads-dsl,pads-ml} tackles a more general problem of handling \emph{any} data format.
+The schema definitions in PADS are similar to our structural type. The structure inference for LearnPADS
+\cite{pads-learn} infers the data format from a flat input stream. A PADS type provider could follow
+many of the patterns we explore in this paper, but formally specifying the safety property would be
+challenging.
+
+\section{Conclusions}
+\label{sec:conclusions}
+
+In this paper, we explored the F\# Data type providers for structured data formats such as XML, CSV and JSON.
+As most real-world data do not come with an explicit schema, the library uses \emph{type inference} that
+deduces a type from a set of samples. Next, it uses the \emph{type provider} mechanism to integrate the 
+inferred type directly into the F\# type system.
+
+The type inference algorithm we use is based on a common supertype relation. For usability reasons, the
+algorithm attempts to infer type that does not contain unions and prefers records with optional fields instead.
+The algorithm is simple and predictable, which is important as developers need to understand how changing the
+samples affects the resulting types.
+
+In the second part of the paper, we explore the programming language theory behind type providers. F\# Data
+is a prime example of type providers, but our work also demonstrates a more general point. The types generated
+by type providers can depend on external input (such as samples) and so we can only formulate \emph{relativized
+safety property}, which says that a program is safe only if the actual inputs satisfy additional conditions --
+in our case, they have to be subtypes of one of the samples.
+
+The type provider mechanism has been described before, but this paper is novel in that it explores concrete
+type providers from the perspective of programming language theory. This is particularly important as the
+F\# Data library is becoming de-facto standard tool for data access in F\#\footnote{At the time of writing, 
+the library has 37,000 downloads on NuGet; over 1600 commits and 36 contributors on GitHub.} and other 
+languages are beginning to adopt mechanisms similar to F\# type providers.
+
+\acks
+?
+We would like to thank to the F\# Data contributors on GitHub and other colleagues working on type providers,
+including Jomo Fisher, Keith Battocchi and Kenji Takeda.
+
+\bibliographystyle{abbrvnat}
+\bibliography{paper}
+
+\appendix
+% --------------------------------------------------------------------------------------------------
 
@@
 
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Practical experiences}
-
-An important concern about using F\# Data in practice is the handling of schema change. When using a
-type provider, the sample is captured at compile-time. If the schema changes later (so that the actual
-input is no longer a subtype of the samples), the program using the type provider fails at run-time
-and it is the developer's responsibility to handle the exception. However, this is the same problem that
-happens when reading data using any other library.
-
-F\# Data can help discover such errors earlier. The example in Section~\ref{sec:introduction}
-points the JSON type provider to a sample using a live URL. This has the advantage that a re-compilation 
-fails when the schema changes, which is an indication that the program needs to be updated to reflect the
-change. If this is undesirable, it is always possible to cache the sample locally.
-
-In general, there is no better solution for plain XML, CSV and JSON data sources. Some data sources provide 
-versioning support (with meta-data about how the schema changed). For those, a type 
-provider could adapt automatically, but we leave this for future.
-
-An approach that we find useful in practice is to keep a set of samples. When the program fails at run-time
-(because a service returned data in an unexpected format), the new input can be added as an additional sample,
-which then shows what parts of code need to be modified.
-
-% ==================================================================================================
-
-\section{Related and future work}
-\label{sec:related}
-
-The F\# Data library connects two lines of research that have been previously disconnected. The first is 
-extending the type systems of programming languages to accommodate external data sources and the second
-is inferring types for real-world data sources.
-
-The type provider mechanism has been introduced in F\# \cite{fsharp-typeprov,fsharp-typeprov-ddfp}, 
-added to Agda \cite{agda-tp} and used in areas such as semantic web \cite{liteq}. Our paper is novel 
-in that it shows the programming language theory behind a concrete type providers.
-
-\paragraph{Extending the type systems.} 
-A number of systems integrate external data formats into a programming language. Those include 
-XML \cite{xduce,xduce-ml} and databases \cite{links}. In both of these, the system either requires
-the user to explicitly define the schema (using the host language) or it has an ad-hoc extension 
-that reads the schema (\emph{e.g.}~from a database). LINQ \cite{linq} is more general, but relies
-on code generation when importing the schema.
-
-The work that is most similar to F\# Data is the XML and SQL integration in C$\omega$ \cite{comega-xs}.
-It extends a C\# like object-oriented language with types similar to our structural types 
-(including nullable types, choices with subtyping and heterogeneous collections with multiplicities).
-However, C$\omega$ does not infer the types from samples and extends the type system of the host
-language (rather than using a general purpose embedding mechanism).
-
-\paragraph{Advanced type systems.}
-Aside from type providers, a number of other advanced type system features could be used to
-tackle the problem discussed in this paper. The Ur \cite{ur} language has a rich system for working
-with records; meta-programming \cite{template-hask}, \cite{th-camlp4} and multi-stage programming \cite{multi-stage}
-could be used to generate code for the provided types. However, as far as we are aware, none of these 
-systems have been used to provide the same level of integration with XML, CSV and JSON.
-
-Another approach would be to use gradual typing \cite{gradual,gradual-js} and add types for structured
-data formats to existing dynamic language to check, for example, JSON manipulation in JavaScript.
-
-\paragraph{Typing real-world data.}
-The second line of research related to our work focuses on inferring structure of real-world data sets.
-A recent work on JSON \cite{typing-json} infers a succinct type using MapReduce to handle large number
-of samples. It fuses similar types based on a type similarity measure. This is more sophisticated than
-our technique, but it would make formally specifying the safety properties (Theorem~\ref{thm:safety}) difficult.
-Extending our \emph{relativized safety} property to a \emph{probabilistic safety} is an interesting 
-future work.
-
-The PADS project \cite{pads-dsl,pads-ml} tackles a more general problem of handling \emph{any} data format.
-The schema definitions in PADS are similar to our structural type. The structure inference for LearnPADS
-\cite{pads-learn} infers the data format from a flat input stream. A PADS type provider could follow
-many of the patterns we explore in this paper, but formally specifying the safety property would be
-challenging.
-
-\section{Conclusions}
-\label{sec:conclusions}
-
-In this paper, we explored the F\# Data type providers for structured data formats such as XML, CSV and JSON.
-As most real-world data do not come with an explicit schema, the library uses \emph{type inference} that
-deduces a type from a set of samples. Next, it uses the \emph{type provider} mechanism to integrate the 
-inferred type directly into the F\# type system.
-
-The type inference algorithm we use is based on a common supertype relation. For usability reasons, the
-algorithm attempts to infer type that does not contain unions and prefers records with optional fields instead.
-The algorithm is simple and predictable, which is important as developers need to understand how changing the
-samples affects the resulting types.
-
-In the second part of the paper, we explore the programming language theory behind type providers. F\# Data
-is a prime example of type providers, but our work also demonstrates a more general point. The types generated
-by type providers can depend on external input (such as samples) and so we can only formulate \emph{relativized
-safety property}, which says that a program is safe only if the actual inputs satisfy additional conditions --
-in our case, they have to be subtypes of one of the samples.
-
-The type provider mechanism has been described before, but this paper is novel in that it explores concrete
-type providers from the perspective of programming language theory. This is particularly important as the
-F\# Data library is becoming de-facto standard tool for data access in F\#\footnote{At the time of writing, 
-the library has 37,000 downloads on NuGet; over 1600 commits and 36 contributors on GitHub.} and other 
-languages are beginning to adopt mechanisms similar to F\# type providers.
-
-\acks
-?
-We would like to thank to the F\# Data contributors on GitHub and other colleagues working on type providers,
-including Jomo Fisher, Keith Battocchi and Kenji Takeda.
-
-\bibliographystyle{abbrvnat}
-\bibliography{paper}
-
 \end{document}

### 2015-11-17 pldi-2016/paper.tex
Fit everything in 11 pages

@@
 Most modern applications interact with external services and access data in structured formats such 
-as XML, JSON and CSV. Static type systems do not understand such formats making accessing data more 
-cumbersome. Should we give up and leave the messy world of external data to dynamic typing and 
-runtime checks? Of course, not!
-
-We show how to integrate external data sources into the F\# type system. As most real-world data
+as XML, JSON and CSV. Static type systems do not understand such formats, often making data access
+more cumbersome. Should we give up and leave the messy world of external data to dynamic typing 
+and runtime checks? Of course, not!
+
+In this paper, we integrate external structured data into the F\# type system. As most real-world data
 do not come with an explicit schema, we develop a type inference algorithm that infers a type from 
@@
 
-Applications for taking notes, searching train schedules or finding tomorrow's weather all 
-communicate with one or more services over the network and present the aggregated data. 
+Social network clients, applications for finding tomorrow's weather or searching train schedules
+all communicate with one or more services over the network and display the resulting information. 
 Increasing number of such services provide REST-based end-points that return data as CSV, XML
@@
 given city\footnote{See ``Current weather data'': \url{http://openweathermap.org/current}}. 
-The documentation describes the URL parameters and shows one sample JSON to illustrate the typical
-response structure. Using a standard library for working with JSON and HTTP, we might call the 
-service and read the temperature as follows:
+The documentation shows one sample to illustrate the typical response. Using standard 
+JSON and web libraries, we might write:
 %
@@
  \quad \quad \kvd{match}~\ident{Map.find}~\str{temp}~\ident{main}~\kvd{with} \\
- \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Nice \%f degrees!}~\ident{num} \\
+ \quad \quad |~\ident{Number}(\ident{num})\rightarrow \ident{printfn}~\str{Lovely \%f!}~\ident{num} \\
  \quad \quad |~\_\rightarrow \ident{failwith}~\str{Incorrect format} \\
@@
  \kvd{type}~\ident{W} = \ident{JsonProvider}\langl\str{http://api.owm.org/?q=NYC}\rangl \\[0.1em]
- \ident{printfn}~\str{Nice \%f degrees!}~(\ident{W.GetSample().Main.Temp})
+ \ident{printfn}~\str{Lovely \%f!}~(\ident{W.GetSample().Main.Temp})
 \end{array}
@@
 
-\item We describe a predictable type inference algorithm for structured data formats that
-  underlies the type providers (\S\ref{sec:inference}) based on a \emph{common supertype}
-  relation.
+\item We describe a predictable type inference algorithm for structured data formats, 
+  based on a \emph{common supertype} relation, that underlies the type providers 
+  (\S\ref{sec:inference}).
 
@@
   This adapts the ML-style type safety for the context of the web.
+
+\item The supplementary screencast show developer experience using 
+  F\# Data with JSON, XML and CSV.
 \end{itemize}
@@
 We start with an informal overview that shows how F\# Data type providers simplify working with 
-JSON, XML and CSV. We introduce the necessary aspects of F\# type providers along the 
+JSON and XML. We introduce the necessary aspects of F\# type providers along the 
 way. The examples in this section illustrate a number of key properties of our type inference 
@@
 \item Our mechanism is predictable. The user directly works with the inferred types and should 
-  understand why a specific type was inferred from a given sample\footnote{In particular, we do 
-  not use probabilistic methods where adding additional sample could change the shape of the type.}.
+  understand why a specific type was inferred from a given sample.\footnote{In particular, we do 
+  not use probabilistic methods where adding an additional sample could completely change the 
+  shape of the type.}
 
@@
 
-The JSON format used in \S\ref{sec:introduction} is a populardata exchange format based on 
+The JSON format used in \S\ref{sec:introduction} is a popular data exchange format based on 
 JavaScript data structures. The following is the definition of \ident{JsonValue} 
@@
  \kvd{type}~\ident{JsonValue} = \\[0.1em]
- \quad|~ \ident{Number}~\kvd{of}~\ident{decimal} ~~|~ \ident{Null} \\[0.1em]
- \quad|~ \ident{String}~\kvd{of}~\ident{string} \qquad\, |~ \ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
+ \quad|~ \ident{Number}~\kvd{of}~\ident{decimal} ~~~~|~ \ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
+ \quad|~ \ident{String}~\kvd{of}~\ident{string} ~~\qquad\, |~ \ident{Null} \\[0.1em]
  \quad|~ \ident{Record}~\kvd{of}~\ident{Map}\langl\ident{string}, \ident{JsonValue}\rangl \\[0.1em]
@@
 \noindent
-Say we want to print the names with an age if it is available. The standard approach would 
+Say we want to print the names and an age if it is available. The standard approach would 
 be to pattern match on the parsed \ident{JsonValue}, check that the top-level node is a \ident{Array}
-and iterate over the elements checking that each is a \ident{Record} with certain properties. We
-would throw an exception or skip values in incorrect format. Even if we defined helper functions, 
+and iterate over the elements checking that each element is a \ident{Record} with certain properties. 
+We would throw an exception for values in an incorrect format. Even with helper functions, 
 we would still need to specify field names as strings, which is error prone and can not be 
@@
 %
-The code achieves the same simplicity as when using dynamically typed languages, but it is statically 
-type-checked. In contrast to the earlier example, we now use a local file \strf{people.json} as a 
-sample for the type inference, but then processes data from another source.
+In contrast to the earlier example, we now use a local file \strf{people.json} as a sample for 
+the type inference, but then processes data from another source. The code achieves the same 
+simplicity as when using dynamically typed languages, but it is statically type-checked.
 
@@
 to the type provider. Static parameters are resolved at compile-time and have to be constant. 
-The provider analyzes the sample and generates a type that we name \ident{People}. In most F\# 
-editors, the type provider is also executed at development-time and the provided types are 
-used in auto-completion.
+The provider analyzes the sample and generates a type  \ident{People}. Most F\# editors also 
+execute the type provider in the background at development-time and use the provided types 
+in auto-completion.
 
@@
 sample values and is inferred as \ident{string}. The field \ident{Age} is marked as optional,
-because the value is missing in one sample. The two sample ages are an integer $25$ and a 
+because the value is missing in one sample. The two age values are an integer $25$ and a 
 decimal $3.5$ and so the common inferred type is \ident{decimal}.
@@
 The type \ident{People} has two methods for reading data. \ident{GetSample} parses the
-sample used for the inference and \ident{Parse} parses a JSON string containing data in the same 
-format as the sample. Both methods return an array of \ident{Entity} values.
+sample used for the inference and \ident{Parse} parses a JSON string. This lets us read
+data at runtime, provided that it is in the same format as the static sample.
 
@@
 executed at run-time in place of \ident{item.Name} and other operations. The runtime behaviour is 
-the same as in the earlier hand-written sample (\S\ref{sec:introduction}). A member access 
+the same as in the earlier hand-written sample (\S\ref{sec:introduction}) -- a member access 
 throws an exception if the document does not have the expected format.
@@
 no exceptions will occur. In other words, we cannot avoid all failures, but we can prevent some.
-If the OpenWeatherMap changes the response format, the sample in \S\ref{sec:introduction}
-will not re-compile and the user knows that the code needs to be changed. What this means for
-the traditional ML type safety is discussed in \S\ref{sec:safety-discuss}.
+Moreover, if the OpenWeatherMap changes the response format, the sample in \S\ref{sec:introduction}
+will not re-compile and the user knows that the code needs to be changed.
 
@@
 The sample code is easy to write thanks to the fact that most F\# editors provide code completion
-when ``.'' is typed. The developer does not need to look at the sample JSON file to see what fields
-are available for a person. To support this scenario, our inference algorithm prefers records 
-(we also treat XML elements and CSV rows as records) and only infers types that require pattern 
-matching as the last resort.
+when ``.'' is typed (see the supplementary screencast). The developer does not need to look at the 
+sample JSON file to see what fields are available. To support this scenario, our inference algorithm 
+prefers records (we also treat XML elements and CSV rows as records) and only infers types that 
+require pattern matching as the last resort.
 
@@
 (in particular the {\ttfamily\small <heading>} element), but it does not show all possible elements 
-(say, {\ttfamily\small <table>}). Assuming the above document is \strf{sample.doc}, we can write:
-%
-\noindent
-\begin{equation*}
-\begin{array}{l}
- \kvd{type}~\ident{Document}~=~\ident{XmlProvider}\langl\str{sample.doc}\rangl\hspace{1em} \\[0.5em]
- \kvd{let}~\ident{root}~=~\ident{Document.Load}(\str{http://.../another.doc})\\
+(say, {\ttfamily\small <table>}). Assuming the above document is \strf{sample.xml}, we can write:
+%
+\noindent
+\begin{equation*}
+\begin{array}{l}
+ \kvd{type}~\ident{Document}~=~\ident{XmlProvider}\langl\str{sample.xml}\rangl\hspace{1em} \\[0.5em]
+ \kvd{let}~\ident{root}~=~\ident{Document.Load}(\str{c:/pldi/another.xml})\\
  \kvd{for}~\ident{elem}~\kvd{in}~\ident{root.Doc}~\kvd{do}\\
- \quad\kvd{match}~\ident{elem.Heading}~\kvd{with} \\
- \quad|~\ident{Some text} \rightarrow \ident{printf}~\str{ - \%s }~\ident{text}\\
- \quad|~\ident{None} \rightarrow ()\\
-\end{array}
-\end{equation*}
-%
-The example iterates over a collection of elements \ident{elem} returned by \ident{root.Doc}. 
-The type provides a typed access to elements known from the sample document and so we can
-write \ident{elem.Heading}, which returns an optional value.
+ \quad\ident{Option.iter}~(\ident{printf}~\str{ - \%s})~\ident{elem.Heading}\\
+\end{array}
+\end{equation*}
+%
+The example iterates over a collection of elements returned by \ident{root.Doc}. The type of element 
+provides a typed access to elements known from the sample and so we can write \ident{elem.Heading}, 
+which returns an optional string value.
 
 \paragraph{Open world and variant types.}
-In the above example, we do not infer a union type representing a choice between 
-heading, paragraph and image. By its nature, XML is extensible and the sample cannot include all 
-possible nodes\footnote{Even when the document structure is defined using XML Schema, documents may 
-contain elements prefixed with other namespaces.}. The type of \ident{elem} is thus a top type (we 
-call it \emph{variant}) annotated with the statically known possibilities (heading, paragraph, image). 
-It is mapped to the following F\# type:
+We do not infer a union type representing a choice between heading, paragraph and image. By its 
+nature, XML is extensible and the sample cannot include all possible nodes.\footnote{Even when the 
+document structure is defined using XML Schema, documents may contain elements prefixed with other 
+namespaces.} The type of \ident{elem} is thus a top type (we call it a \emph{variant}) annotated 
+with the statically known possibilities. It is mapped to the following F\# type:
 %
@@
 %
-The type provides a nice access to the elements known statically from the sample, but it encodes
-the \emph{open-world assumption} -- the user also needs to handle the case when the value is none 
+The type provides a nice access to the statically known elements, but it encodes the 
+\emph{open-world assumption}. The user also needs to handle the case when the value is none 
 of the statically known elements.
@@
 correctly on document that contains any elements, including those not listed in the sample. This
-is extremely useful when reading data from external data sources, because the sample document only
+is extremely useful when reading data from external sources, because the sample document only
 needs to contain values relevant to the application. It is also worth noting that our type 
 inference attempts to minimze the number of variant types that appear in the resulting type --
-a variant is not used if there is another option (Corollary~\ref{thm:no-unions}).
+a variant is not used if there is another option (see Corollary~\ref{thm:no-unions} and 
+\S\ref{sec:impl-hetero}).
 
@@
 \subsection{Summary}
-In addition to JSON and XML (\S\ref{sec:providers-json}, \S\ref{sec:providers-xml}), F\# Data also 
-implements a type provider for CSV. We treat CSV data as lists of records (with field for each 
-column) and so handling of CSV is just a special case of our inference algorithm.
+\label{sec:providers-sum}
+In addition to type providers for JSON and XM, F\# Data also implements a type provider for CSV 
+(\S\ref{sec:impl-parsing}). We treat CSV files as lists of records (with field for each column) 
+and so CSV is handled directly by our inference algorithm.
 
 Throughout the introduction, we used data sets that demonstrate typical issues that are frequent
-in real-world inputs (missing data, inconsistent encoding of primitive values and heterogeneous 
+in the real-world (missing data, inconsistent encoding of primitive values and heterogeneous 
 structures). In our experience, these are the most common issues. The following JSON response with 
@@
 The first is a record with meta-data about the current page and the second is an array with data. 
-The actual implementation supports a concept of heterogeneous collections (omitted in this paper
-for space reasons) and provides a type with properties \ident{Record} for the former and 
-\ident{Array} for the latter. Second, the \ident{value} is \kvd{null} for some records. Third, 
+The actual F\# Data implementation supports a concept of heterogeneous collections (briefly outlined
+in \S\ref{sec:impl-hetero}) and provides a type with properties \ident{Record} for the former and 
+\ident{Array} for the latter. Second, the \ident{value} field is \kvd{null} for some records. Third, 
 numbers can be represented in JSON as numeric literals (without quotes), but here, they are 
-returned as string literals instead\footnote{This is often used to avoid non-standard numerical 
-types of JavaScript.}.
+returned as string literals instead.\footnote{This is often used to avoid non-standard numerical 
+types of JavaScript.}
 
@@
 (written as $\sigma$). We write $\nu$ for record field names and $\nu_?$ for names that can be empty:
-
-\noindent
+%
 \begin{equation*}
@@
 \end{equation*}
-%
+
+\noindent
 Non-nullable types include records (consisting of an optional name and fields with their types) 
@@
 \kvd{null} value. Type providers map \kvd{option} to standard F\# option. A collection $[\sigma]$ 
-is also nullable and missing values or \kvd{null} are treated as empty collections. The type 
+is also nullable and \kvd{null} values are treated as empty collections. The type 
 $\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded notation) and $\bot$ 
@@
   \label{eq:sub-null}\tag{P2}
-  \sigma &:> \kvd{null}  &(\textnormal{iff}~\sigma \neq \hat{\sigma}) 
-\end{align}
-
-\noindent
-\begin{align}
+  \sigma &:> \kvd{null}  &(\textnormal{iff}~\sigma \neq \hat{\sigma})  \\[-0.2em]
   \label{eq:sub-opt}\tag{P3}
@@
   \kvd{option}\langl\hat{\sigma_1}\rangl &:> 
-    \kvd{option}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} :> \hat{\sigma_2}) \\[-0.2em]
+    \kvd{option}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} :> \hat{\sigma_2})\\[-0.2em]
   \label{eq:sub-col}\tag{P5}
@@
 \end{align}
-\vspace{-0.8em}
+\vspace{-1em}
 \begin{align}
@@
   additional fields (\ref{eq:sub-record2}) and fields can  be reordered (\ref{eq:sub-record3}). 
-  The interesting rule is (\ref{eq:sub-record4}); together with covariance, it states that a 
-  subtype can omit some fields, provided that their types are nullable.
+  The interesting rule is (\ref{eq:sub-record4}). It implies that subtype can omit fields, 
+  provided that \kvd{null} is a valid value for the field (i.e. the field is optional).
 \end{itemize}
@@
 mutual subtypes) would not add any value over having a single top type, but our purpose is 
-different. We simplify reading structured data and, as shown earlier (\S\ref{sec:providers-xml}), 
+different. We focus on reading structured data and, as shown earlier (\S\ref{sec:providers-xml}), 
 the types forming a variant provide a valuable information about the sample document.
@@
 
-Some of the aspects of our system (numeric types, handling of \kvd{null} and treatment of
-missing data) are based on common F\# practices and might differ for other languages. For example, 
-OCaml also has multiple numeric types, but would likely always handle \kvd{null} explicitly.
+Some of the aspects of our system (numeric types and handling of missing data) offer a range of
+choices. For example, OCaml also has different numerical types and would likely always handle 
+\kvd{null} explicitly.
 
@@
 \label{fig:op-conversions}
+\vspace{-0.5em}
 \end{figure*}
@@
 type. If it does, the two types are combined, otherwise a new case is added.
-When combining two variants (\emph{var-2}), we group the cases that have a shared tags.
+When combining two variants (\emph{var-2}), we group the cases that have shared tags.
 Finally, (\emph{var-3}) covers the case when we are combining two distinct non-variant types.
@@
 
-We represent JSON, XML and CSV documents using the same \emph{structural value}. Structural 
-values are first-order and can be one of the following cases:
+We represent JSON, XML and CSV documents using the same first-order \emph{structural values} defined
+as follows:
 %
@@
 by the rules in Figure~\ref{fig:op-conversions}. The conversion functions (\ident{asDec}
-and \ident{asFloat}) only perform conversions required by the subtyping relation (our actual 
+and \ident{asFloat}) only perform conversions required by the subtyping (our actual 
 implementation is more lax and performs additional conversions).
@@
 value has a field of the same name with a matching type. The last line defines a ``catch all'' 
-pattern, which returns \kvd{false} for all remaining cases. Although not elenegantly, the rules 
+pattern, which returns \kvd{false} for all remaining cases. Although not elegant, the rules 
 can be expressed using just Featherweight F\#, discussed in the next section.
@@
 \label{fig:ff-reduction}
-\vspace{-1em}
 \end{figure}
@@
  L &\narrow{=}& \kvd{type}~C(\overline{x:\tau}) = \overline{M} \\[0.0em]
- M &\narrow{=}& \kvd{member}~N:\tau=e
-\end{array}
-\end{equation*}
-
-\noindent
-\begin{equation*}
-\begin{array}{rcl}
+ M &\narrow{=}& \kvd{member}~N:\tau=e \\[0.6em]
  v &\narrow{=}& s \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \lsep [v_1; \ldots; v_n] \\[0.0em]
@@
 
-Next, we define the reduction relation and (a fragment of) type checking for Featherweight F\#.
-The language presented here is intentionally incomplete. We only define parts needed to prove
-the relativized safety property (\S\ref{sec:safety}).
-
 \paragraph{Reduction.}
@@
 This models the eager evaluation of F\#. An evaluation context $E$ is defined as:
-
-\vspace{0.5em}
-\noindent
-\begin{equation*}
-\begin{array}{l}
-\hspace{-0.5em}
+%
+\begin{equation*}
 \begin{array}{rcl}
@@
    &\narrow{|}& \ident{map}~(\lambda x\rightarrow e)~E \lsep \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2 \\[0.1em]
-   &\narrow{|}& \ident{Some}(E) \lsep f(E) \lsep E = e \lsep v = E \\[0.1em]
+   &\narrow{|}& \ident{Some}(E) \lsep op(E) \lsep E = e \lsep v = E \\[0.1em]
    &\narrow{|}& \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
    &\narrow{|}& \kvd{match}~E~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2
-\end{array} \\[3em]
-f \in \{ \ident{asDec}, \ident{asFloat}, \ident{getField}, \ident{getChildren}, \ident{isNull}, \ident{hasType} \}
 \end{array} 
@@
 We omit expressions $e \vee e$ and $e \wedge e$ that are used in Figure~\ref{fig:op-conversions}, 
-but those can be easily defined as syntactic sugar using \kvd{if}. All expressions reduce to a value in a 
+as those can be easily defined as syntactic sugar using \kvd{if}. All expressions reduce to a value in a 
 finite number of steps or get stuck due to an error condition. An error condition can be a wrong 
@@
 $L; \Gamma \vdash e : \tau$ where the context also contains a set of class declarations $L$.
-The rules demonstrate the key differences from Standard ML and Featherweight Java:
+The fragment demonstrates the key differences from Standard ML and Featherweight Java:
 %
@@
 All F\# Data type providers take (one or more) sample documents, infer a common supertype $\sigma$
-and then use it to generate F\# types that are exposed to the programmer\footnote{The actual 
+and then use it to generate F\# types that are exposed to the programmer.\footnote{The actual 
 implementation provides \emph{erased types} as described in \cite{fsharp-typeprov}. Here, we treat 
 the code as actually generated. This is an acceptable simplification, because F\# Data type providers 
-do not rely on laziness that is available through erased types.}.
+do not rely on laziness that is available through erased types.}
 
 \paragraph{Type provider mapping.}
-The type provider produces an F\# type $\tau$ together with an expression that wraps an input
+A type provider produces an F\# type $\tau$ together with an expression that wraps an input
 value (of type \ident{StructVal}) as a value of type $\tau$ and a collection of class definitions. 
@@
 conversion function from Figure~\ref{fig:op-conversions}.
-Handling of records is more interesting. We generate a class $C$ that takes a structural value as constructor
-parameter. For each record field, we generate a new member with the same name as the field\footnote{The actual
-F\# Data implementation also capitalizes the names.}. The body of the member calls \ident{getField} and then
+
+For records, we generate a class $C$ that takes a structural value as constructor
+parameter. For each record field, we generate a member with the same name as the field.\footnote{The actual
+F\# Data implementation also capitalizes the names.} The body of the member calls \ident{getField} and then
 wraps this expression using $\sem{\sigma_i}$ which maps the field (structural value of type $\sigma_i$) into 
 an F\# value of type $\tau_i$. The returned expression creates a new instance of $C$ and the mapping returns 
-the class $C$ together with all recursively generated definitions. Note that the class
+the class $C$ together with all recursively generated classes. Note that the class
 name $C$ is not directly accessed by the user and so we can use arbitrary name, although the actual 
-implementation in F\# Data attempts to infer a reasonable name\footnote{For example, in 
+implementation in F\# Data attempts to infer a reasonable name.\footnote{For example, in 
 \ident{\{\str{person}:\{\str{name}:\str{Tomas}\}\}}, the nested record will be named \ident{Person}
-based on the name of the parent record field.}.
+based on the name of the parent record field.}
 
@@
 $\ident{getField}(\ident{Person},\ident{Age},v)$. If the returned field is not \kvd{null}, then the member
-calls \ident{asInt} (produced by the case for primitive types) and wraps the result in \ident{Some}.
-Note that \ident{getField} is defined even when the field does not exist, but it returns 
-\kvd{null}. This lets us treat missing fields as optional. An F\# type corresponding to the (structural)
-record is \ident{Person} and given a structural value $s$, we create a \ident{Person} value by calling 
-$\kvd{new}~\ident{Person}(s)$.
+calls \ident{asInt} and wraps the result in \ident{Some}. Note that \ident{getField} is defined even 
+when the field does not exist, but it returns \kvd{null}. This lets us treat missing fields as 
+optional. An F\# type corresponding to the (structural) record is \ident{Person} and a structural 
+value $s$ is wrapped by by calling $\kvd{new}~\ident{Person}(s)$.
 
@@
 \end{equation*}
-%
+
 The \ident{PersonOrString} type contains one member for each of the variant case. In the body, they
@@
 structural type $\sigma$:
-
-\noindent
+%
 \begin{equation*}
@@
 \noindent
-This shows that the provided types are correct with respect to the subtyping relation. Next, the
-main theorem of the paper states that, for any input (which is a subtype of any of the samples) and 
+This shows that the provided types are correct with respect to subtyping. Our key
+theorem states that, for any input (which is a subtype of any of the samples) and 
 any expression $e$, a well-typed program that uses the provided types does not ``go wrong''.
-
-Using the standard syntactic type safety  \cite{syntactic}, we prove type preservation 
+Using standard syntactic type safety  \cite{syntactic}, we prove type preservation 
 (reduction does not change type) and progress (an expression that is not a value can be reduced).
@@
 The F\# Data library has been widely adopted by the industry and is one of the most downloaded
-F\# libraries\footnote{At the time of writing, the library has over 80,000 downloads on NuGet 
-(package repository), 1,821 commits and 44 contributors on GitHub.}. A practical demonstration of 
-development using the library can be seen in an attached video and additional documentation can be
+F\# libraries.\footnote{At the time of writing, the library has over 80,000 downloads on NuGet 
+(package repository), 1,821 commits and 44 contributors on GitHub.} A practical demonstration of 
+development using the library can be seen in an attached screencast and additional documentation can be
 found at \url{http://fsharp.github.io/FSharp.Data}.
@@
 
-\subsection{Discussion}
+\subsection{Relativized safety in practicve}
 \label{sec:safety-discuss}
 
-The \emph{relativized safety} property does not guarantee the same amount of safety as standard 
-ML type safety. However, it reflects the reality of programming with external data sources that is 
+The \emph{relativized safety} property does not guarantee the same amount of safety as traditional
+ML type safety, but it reflects the reality of programming with external data sources that is 
 increasingly important \cite{age-of-web}. Type providers do not reduce the safety -- they simply 
@@
 
-The actual implementation in F\# Data throws an exception for invalid inputs. In contrast, the calculus
-presented here simply gets stuck. We could extend the calculus with exceptions, but that would obscure the 
-purpose -- precisely specifying when the code using type providers does not ``go wrong.''
-
-As mentioned earlier, \emph{relativized safety} specifies a sufficient, but not a necessary condition.
-This is also reflected in the conversion functions which allow additional conversions not required by 
-the subtyping relation. In practice, the additional flexibility proves useful (as minor variations in 
-input formats are common). However, an interesting stricter alternative would be to check that an input 
-has the required type when it is loaded -- and throw an exception immediately when loading data rather than 
-on member access.
-
-Finally, the formal model presented here ignores two aspects of the real type provider mechanisms. As discussed
-here, we do not use the \emph{type erasure} mechanism, but treat type providers as if they were actually 
-generating code. For XML, CSV and JSON, both models can be used. However, one interesting aspect of erasing 
-type providers is that the types can be generated lazily -- only classes that are actually used in the type
-checked code are generated. Extending our formalism to capture this aspect is an interesting future work
-relevant to other type providers -- including the World Bank type provider that is also available
-in the F\# Data library.
-
-\subsection{Practical experiences}
-
-An important concern about using F\# Data in practice is the handling of schema change. When using a
-type provider, the sample is captured at compile-time. If the schema changes later (so that the actual
-input is no longer a subtype of the samples), the program using the type provider fails at run-time
-and it is the developer's responsibility to handle the exception. However, this is the same problem that
-happens when reading data using any other library.
-
-F\# Data can help discover such errors earlier. The example in Section~\ref{sec:introduction}
-points the JSON type provider to a sample using a live URL. This has the advantage that a re-compilation 
+One such issue is the handling of schema change. With type providers, the sample is captured 
+at compile-time. If the schema changes later (so that the input is no longer a subtype of the 
+sample), the program fails at runtime and developers have to handle the exception. This is the 
+same problem that happens when reading data using any other library.
+
+F\# Data can help discover such errors earlier. Our first example (\S\ref{sec:introduction})
+points the JSON type provider at a sample using a live URL. This has the advantage that a re-compilation 
 fails when the schema changes, which is an indication that the program needs to be updated to reflect the
-change. If this is undesirable, it is always possible to cache the sample locally.
-
-In general, there is no better solution for plain XML, CSV and JSON data sources. Some data sources provide 
-versioning support (with meta-data about how the schema changed). For those, a type 
-provider could adapt automatically, but we leave this for future.
-
-An approach that we find useful in practice is to keep a set of samples. When the program fails at run-time
-(because a service returned data in an unexpected format), the new input can be added as an additional sample,
-which then shows what parts of code need to be modified.
-
-\subsection{Reading CSV files}
-\label{sec:providers-csv}
-
-In the CSV file format, the structure is a collection of rows (records) consisting of fields 
-(with names specified by the first row). The inference needs to infer the types of fields.
-For example:
+change.
+
+In general, there is no better solution for plain XML, CSV and JSON data sources. However, some data 
+sources provide versioning support (with meta-data about how the schema changed). For those, a type 
+provider could adapt automatically, but we leave this for future work.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Parsing structured data}
+\label{sec:impl-parsing}
+
+In our formalization, we treat XML, JSON and CSV uniformly as \emph{structural values}. With the addition of 
+named records (for XML nodes), the definition of structural values is rich enough to capture all 
+three formats\footnote{The same mechanism has later been used by the HTML type provider 
+(\url{http://fsharp.github.io/FSharp.Data/HtmlProvider.html}), which provides similarly easy 
+access to data in HTML tables and lists.}. However, parsing real-world data poses a number of practical issues.
+
+\paragraph{Reading CSV data.}
+When reading CSV data, we read each row as an unnamed record and return a collection of rows.
+One difference between JSON and CSV is that in CSV, the literals have no data types and so 
+we also need to infer the type of primitive values. For example:
 %
@@
 \noindent
-One difference between JSON and CSV formats is that in CSV, the literals have no data types.
-In JSON, strings are \str{quoted} and Booleans are \kvd{true} and \kvd{false}. This is not the
-case for CSV and so we need to infer not just the structure, but also the primitive types.
-
-Assuming the sample is saved as \strf{airdata.csv}, the following snippet prints all
-rows from another file that were not autofilled:
-%
-\begin{equation*}
-\begin{array}{l}
- \kvd{type}~\ident{AirCsv}~=~\ident{CsvProvider}\langl\str{airdata.csv}\rangl\hspace{1em} \\[0.6em]
- \kvd{let}~\ident{air}~=~\ident{AirCsv.Parse}(\ident{data})\\[0.1em]
- \kvd{for}~\ident{row}~\kvd{in}~\ident{air.Rows}~\kvd{do}\\[0.1em]
- \quad\kvd{if}~\ident{not}~\ident{row.Autofilled}~\kvd{then}\\[0.1em]
- \quad\quad \ident{printf}~\str{\%s: \%d}~\ident{row.Date row.Ozone}\\
-\end{array}
-\end{equation*}
-%
-The type of the record (\ident{row}) is, again, inferred from the sample. The \ident{Date} column
-uses mixed formats and is inferred as \ident{string} (although we support many date formats and 
-``May 3'' would work). More interestingly, we also infer \ident{Autofilled} as Boolean, because 
-the sample contains only $0$ and $1$ values and using those for Booleans is a common CSV convention.
-Finally, the fields \ident{Ozone} and \ident{Temp} have types \ident{decimal} and
-$\ident{option}\langl\ident{int}\rangl$.
-
-The theoretical model in Section~\ref{sec:formal} presents the core ideas behind the type 
-providers for structured data formats. However, an important part of the success of the F\# Data
-library is also its pragmatic approach, which makes it suitable for use with real-world data.
-
-The formal model discusses some of the pragmatic choices, such as the preference for records over 
-unions and the ability to treat $0$ and $1$ as both Booleans and numbers. In this section, we briefly
-discuss some of the remaining practical concerns, starting with the handling of collections.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Parsing structured data}
-\label{sec:impl-parsing}
-
-In this paper, we treat the XML, JSON and CSV formats uniformly as \emph{structural values}.
-The definition of structural values is, indeed, rich enough to capture all three formats. The
-structure is similar to JSON (it has primitive values, records and collections). We also added
-\emph{named} records, which are needed for XML.
-%
-\begin{itemize}
-\item When reading JSON, we directly create a corresponding structural value (with all records unnamed).
-  Optionally, we also read numerical values (and dates) stored in strings.
-\item When reading CSV, we read each row as an unnamed record and return them in a collection.
-  Optionally, the read values are parsed as numbers or Booleans and missing values become \kvd{null}.
-\item When reading XML, we create a named record for each node. Attributes become record fields and
-  body becomes a special field named $\bullet$
-\end{itemize}
-%
-The reading of XML documents is perhaps the most interesting. To demonstrate how the body is treated,
-consider the following:
+The value {\small\ttfamily \#N/A} is commonly used to represent missing values in CSV and is treated
+as \kvd{null}. The \ident{Date} column uses mixed formats and is inferred as \ident{string} 
+(we support many date formats and ``May 3'' would be parsed correctly). More interestingly,
+we also infer \ident{Autofiled} as Boolean, because the sample contains only $0$ and $1$.
+This is handled by adding a \ident{bit} type which is a subtype of both \ident{int} and \ident{bool}.
+
+\paragraph{Reading XML documents.}
+Mapping XML documents to structural values is perhaps the most interesting. For each node, we
+create a named record. Attributes become record fields and the body becomes a field with a special
+name:
 %
@@
 \noindent
-This XML becomes a record named \ident{root} with fields \ident{id} and $\bullet$. The nested element
-contains only the $\bullet$ field containing the inner text:
+This XML becomes a record \ident{root} with fields \ident{id} and $\bullet$ for the body. 
+The nested element contains only the $\bullet$ field with the inner text. As with CSV, we
+infer type of primitive values:
 %
@@
 %
-When generating types for types inferred from XML, we also include a special case to remove the $\bullet$
-node using the fact that this can be only a collection (of elements) or a primitive value.
-
-Finally, the XML type provider in F\# Data includes an additional option to use \emph{global inference}.
-In that case, the inference algorithm from Section~\ref{sec:formal-inferval} is replaced with an alternative
-that unifies the types of all records with the same name. This is useful when the sample is, for example,
-an XHTML document -- in this case, all occurrences of an element (\emph{e.g.}~{\small\ttfamily <a>})
-are treated as the same type.
+When generating types for XML documents, we also lifts the members nested under the $\bullet$ 
+field into the parent type to simplify the structure of the generated type.
+
+The XML type provider also includes an option to use \emph{global inference}. In that case, 
+the inference from values (\ref{sec:formal-inferval}) unifies the types of all records with the 
+same name. This is useful because, for example, in XHTML all {\small\ttfamily <table>} elements
+will be treated as values of the same type.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Heterogeneous collections} 
+\label{sec:impl-hetero}
+
+When introducing type providers (\S\ref{sec:providers-sum}), we mentioned that F\# Data 
+implements a special handling of heterogeneous collections. This allows us to avoid inferring a 
+variant types in many common scenarios. In the earlier example, a sample collection 
+contains a record (with \strf{pages} and \strf{page} fields) and a nested collection with values.
+
+Rather than storing a single type for the collection elements as in $[\sigma]$, heterogeneous
+collections store multiple possible element types together with their \emph{  inferred multiplicity} 
+(zero or one, exactly one, zero or more):
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \psi &\narrow{=}& 1? \lsep 1 \lsep \ast \\
+ \sigma &\narrow{=}& ~\ldots \lsep [\sigma_1, \psi_1 | \ldots | \sigma_n, \psi_n ]
+\end{array}
+\end{equation*}
+%
+We omit the details, but finding a common supertype of two heterogeneous 
+collections is analogous to the handling of variants. We merge cases with the same tag (by finding 
+their common super-type) and calculate their new shared multiplicity (for example, by turning 
+$1$ and $1?$ into $1?$).
 
@@
 
-The F\# Data library connects two lines of research that have been previously disconnected. The first is 
-extending the type systems of programming languages to accommodate external data sources and the second
-is inferring types for real-world data sources.
-
-The type provider mechanism has been introduced in F\# \cite{fsharp-typeprov,fsharp-typeprov-ddfp}, 
-added to Agda \cite{agda-tp} and used in areas such as semantic web \cite{liteq}. Our paper is novel 
-in that it shows the programming language theory behind a concrete type providers.
+We connect two lines of research: integration of external data into statically-typed 
+programming languages and inferring types for real-world data sources. We build on F\# type 
+providers \cite{fsharp-typeprov,fsharp-typeprov-ddfp,idris-tp,liteq}, but our paper is novel 
+in that it discusses the theory and safety of a concrete type provider.
 
 \paragraph{Extending the type systems.} 
-A number of systems integrate external data formats into a programming language. Those include 
-XML \cite{xduce,xduce-ml} and databases \cite{links}. In both of these, the system either requires
-the user to explicitly define the schema (using the host language) or it has an ad-hoc extension 
-that reads the schema (\emph{e.g.}~from a database). LINQ \cite{linq} is more general, but relies
-on code generation when importing the schema.
-
-The work that is most similar to F\# Data is the XML and SQL integration in C$\omega$ \cite{comega-xs}.
-It extends a C\# like object-oriented language with types similar to our structural types 
-(including nullable types, choices with subtyping and heterogeneous collections with multiplicities).
-However, C$\omega$ does not infer the types from samples and extends the type system of the host
-language (rather than using a general purpose embedding mechanism).
+Previous work that integrates external data, such as XML \cite{xduce,xduce-ml} and databases 
+\cite{links,linq}, into a programming language, required the user to define the schema
+or it has an ad-hoc extension that reads the schema.
+
+C$\omega$ \cite{comega-xs} is the most similar to F\# Data. It extends C\# with types similar to our
+structural types (including similar heterogeneous collections), but it does not infer the types
+from a sample and extends the type system of the host language (rather than using general purpose
+embedding).
 
 \paragraph{Advanced type systems.}
-Aside from type providers, a number of other advanced type system features could be used to
-tackle the problem discussed in this paper. The Ur \cite{ur} language has a rich system for working
-with records; meta-programming \cite{template-hask}, \cite{th-camlp4} and multi-stage programming \cite{multi-stage}
-could be used to generate code for the provided types. However, as far as we are aware, none of these 
+A number of other advanced type system features could be used to tackle the problem discussed 
+in this paper. The Ur \cite{ur} language has a rich system for working with records; 
+meta-programming \cite{template-hask}, \cite{th-camlp4} and multi-stage programming \cite{multi-stage}
+could be used to generate code for the provided types; and gradual typing \cite{gradual,gradual-js} 
+can add typing to existing dynamic languages. As far as we are aware, none of these 
 systems have been used to provide the same level of integration with XML, CSV and JSON.
 
-Another approach would be to use gradual typing \cite{gradual,gradual-js} and add types for structured
-data formats to existing dynamic language to check, for example, JSON manipulation in JavaScript.
-
 \paragraph{Typing real-world data.}
-The second line of research related to our work focuses on inferring structure of real-world data sets.
-A recent work on JSON \cite{typing-json} infers a succinct type using MapReduce to handle large number
-of samples. It fuses similar types based on a type similarity measure. This is more sophisticated than
-our technique, but it would make formally specifying the safety properties (Theorem~\ref{thm:safety}) difficult.
-Extending our \emph{relativized safety} property to a \emph{probabilistic safety} is an interesting 
-future work.
+Recent work \cite{typing-json} infers a succinct type of large JSON datasets using MapReduce.
+It fuses similar types based on similarity. This is more sophisticated than our technique, but it 
+makes formal specification of safety (Theorem~\ref{thm:safety}) difficult. Extending our 
+\emph{relativized safety} to \emph{probabilistic safety} is an interesting future work.
 
@@
 
-In this paper, we explored the F\# Data type providers for structured data formats such as XML, CSV and JSON.
-As most real-world data do not come with an explicit schema, the library uses \emph{type inference} that
-deduces a type from a set of samples. Next, it uses the \emph{type provider} mechanism to integrate the 
-inferred type directly into the F\# type system.
-
-The type inference algorithm we use is based on a common supertype relation. For usability reasons, the
-algorithm attempts to infer type that does not contain unions and prefers records with optional fields instead.
-The algorithm is simple and predictable, which is important as developers need to understand how changing the
-samples affects the resulting types.
-
-In the second part of the paper, we explore the programming language theory behind type providers. F\# Data
-is a prime example of type providers, but our work also demonstrates a more general point. The types generated
-by type providers can depend on external input (such as samples) and so we can only formulate \emph{relativized
-safety property}, which says that a program is safe only if the actual inputs satisfy additional conditions --
+We explored the F\# Data type providers for XML, CSV and JSON. As most real-world data do not come 
+with an explicit schema, the library uses \emph{type inference} that deduces a type from a set of 
+samples. Our type inference algorithm is based on a common supertype relation. For usability 
+reasons, it prefers records and avoids variant types. The algorithm is predictable, which is 
+important as developers need to understand how changing the samples affects the resulting types.
+
+We explore the programming language theory behind type providers. F\# Data is a prime example of 
+type providers, but our work also demonstrates a more general point. The types generated by type 
+providers can depend on external input and so we can only guarantee \emph{relativized safety}, 
+which says that a program is safe only if the actual inputs satisfy additional conditions --
 in our case, they have to be subtypes of one of the samples.
 
-The type provider mechanism has been described before, but this paper is novel in that it explores concrete
-type providers from the perspective of programming language theory. This is particularly important as the
-F\# Data library is becoming de-facto standard tool for data access in F\#\footnote{At the time of writing, 
-the library has 37,000 downloads on NuGet; over 1600 commits and 36 contributors on GitHub.} and other 
-languages are beginning to adopt mechanisms similar to F\# type providers.
-
-\acks
-?
-We would like to thank to the F\# Data contributors on GitHub and other colleagues working on type providers,
-including Jomo Fisher, Keith Battocchi and Kenji Takeda.
+Type providers have been described before, but this paper is novel in that it explores concrete 
+type providers from the perspective of programming language theory. This is particularly important
+for F\# Data type providers, which have been widely adopted by the industry.
+
+% \acks
+% We would like to thank to the F\# Data contributors on GitHub and other colleagues working on type providers,
+% including Jomo Fisher, Keith Battocchi and Kenji Takeda.
 
@@
 
-\appendix
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure*}
-\noindent
-\begin{equation*}
-\inference[(col)\;]
-  { \tytagof(\sigma_i) = \tytagof(\sigma'_i) \quad 
-    \sigma_i \tsep \sigma'_i \vdash \sigma''_i \quad
-    \phi_i \tsep \phi'_i \vdash \phi''_i \qquad (\forall i\in \{ 1 .. k \}) }
-  { \begin{array}{c}
-    [\sigma_1,\psi_1 | \ldots | \sigma_k,\psi_k | \ldots | \sigma_n,\psi_n ] \tsep 
-    [\sigma'_1,\psi'_1 | \ldots | \sigma'_k, \psi'_k | \ldots | \sigma'_m, \psi'_m ] \vdash\\[0.1em]
-    [\sigma''_1, \psi''_1 | \ldots | \sigma''_k, \psi''_k | 
-        \sigma_{k+1},\addopt{\psi_{k+1}} | \ldots | \sigma_{n},\addopt{\psi_{n}} |
-        \sigma'_{k+1},\addopt{\psi'_{k+1}} | \ldots | \sigma'_{m}, \addopt{\psi'_{m}} ]
-    \end{array} }
-\quad
-\begin{array}{rcl}
- \\
- \addopt{\phi} = \phi'&\narrow{\textnormal{such that}}&\phi \tsep 1? \vdash \phi' \\[0.1em]
- \phi \tsep \phi' \vdash \phi&\narrow{\textnormal{when}}&\phi :> \phi' \\[0.1em]
- \phi \tsep \phi' \vdash \phi'&\narrow{\textnormal{when}}&\phi' :> \phi
-\end{array}    
-\end{equation*}
-\caption{Inference judgement that defines the common supertype of two heterogeneous collections}
-\label{fig:subtyping-hetcol}
-\end{figure*}
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Heterogeneous collections}
-\label{sec:impl-collections}
-
-When introducing the XML type provider (Section~\ref{sec:providers-xml}), we briefly mentioned 
-that F\# Data implements a special handling of heterogeneous collections. In the example, the provider
-generated a type with \ident{Title} member of type \ident{string} (corresponding to a nested element
-{\ttfamily\small <title>} that appears exactly once) and \ident{Items} member (returning a list of
-values obtained from multiple {\ttfamily\small <item>} elements).
-
-To capture this behaviour, we need to extend our earlier treatment of collections. Rather than storing
-a single type for the elements as in $[\sigma]$, we store multiple possible element types. However,
-this is not just a union type as we also store \emph{inferred multiplicity} of elements for 
-each of the types:
-%
-\begin{equation*}
-\begin{array}{rcl}
- \psi &\narrow{=}& 1 \lsep 1? \lsep \ast \\
- \sigma &\narrow{=}& ~\ldots \lsep [\sigma_1, \psi_1 | \ldots | \sigma_n, \psi_n ]
-\end{array}
-\end{equation*}
-%
-The multiplicities $1, 1?$ and $\ast$ represent \emph{exactly one}, \emph{zero or one} and \emph{zero or more},
-respectively. Thus, the inferred type of the collection in the RSS feed would be
-$[\ident{title}~\{ \ldots \}\;,1|\;\ident{item}~\{ \ldots \},\ast]$, which reads as a collection
-containing exactly one \ident{title} record and any number of \ident{item} records.
-
-The subtyping relation on heterogeneous collections specifies that a subtype can have fewer cases 
-provided that they do not have to appear exactly once. A subtype can also have a stricter specification
-of multiplicity (the ordering is $\ast :> 1? :> 1$).
-
-Finding a common supertype of heterogeneous collections is analogous to the handling of union types.
-The key rule is (\emph{col}) in Figure~\ref{fig:subtyping-hetcol}. It merges cases with the same tag
-(by merging both the type and the multiplicity). For cases that appear only in one collection, we ensure
-that the multiplicity is \emph{zero or one} or \emph{zero or more} using the auxiliary definition
-$\addopt{-}$.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Inferring and generating unions}
-\label{sec:impl-unions}
-
-The F\# Data type providers prefer types that do not contain unions (Theorem~\ref{thm:no-unions}).
-The motivation for this is twofold. First, the current support for type providers in the F\# compiler
-does not allow type providers to generate F\#-specific types (including discriminated unions). This
-means that the provided type cannot use idiomatic F\# representation. Second, when accessing data,
-using records aids the explorability -- when a value is a record, users can type ``.'' and most modern
-F\# editors provide auto-completion list with members.
-
-This choice has an important desirable consequence -- it allows treating a union with additional 
-cases as a subtype of another union and, hence, the provided code is guaranteed to work on \emph{more}
-inputs. Consider a type that is either a \ident{Person} record or just a string. 
-The type provider exposes it as the following F\# class:
-%
-\begin{equation*}
-\begin{array}{l}
- \kvd{type}~\ident{PersonOrString}~=  \\[0.1em]
- \quad \kvd{member}~\ident{Person}~:~\ident{option}\langl \ident{Person}\rangl \\[0.1em]
- \quad \kvd{member}~\ident{String}~:~\ident{option}\langl \ident{string}\rangl \\[0.1em]
-\end{array}
-\end{equation*}
-%
-If we provided an algebraic data type (discriminated union), the consumer would have to use pattern
-matching that would cover \emph{two} cases. The above type forces the user to also handle a third
-case, when both properties return \ident{None}. This also means that union types can silently skip
-over \kvd{null} values.
-
-That said, exposing a discriminated union (perhaps together with better tooling) is certainly an
-attractive alternative that we plan to consider in the future. This can also be combined with type 
-inference that chooses between records and union types based on some statistical metric (see
-related work in Section~\ref{sec:related}).
-
 \end{document}

### 2015-11-19 pldi-2016/paper.tex
Final edits, page 1

@@
 
-\title{F\# Data: \textnormal{Making structured data first-class citizens}}
+\title{Types from data: \textnormal{Making structured data first-class citizens in F\#}}
 %\subtitle{Subtitle Text, if any}
@@
 In this paper, we integrate external structured data into the F\# type system. As most real-world data
-do not come with an explicit schema, we develop a type inference algorithm that infers a type from 
+does not come with an explicit schema, we develop a shape inference algorithm that infers a shape from 
 representative sample documents and integrate it into the F\# type system using type providers.
 
-Our library significantly reduces the amount of code that developers need to write when 
-accessing data. It also provides additional safety guarantees, arguably, as much as possible if 
-we abandon the closed world assumption.
+Our library significantly reduces the amount of data access code and it provides additional 
+safety guarantees when contrasted with the widely used weakly typed techniques.
 \end{abstract}
@@
 Social network clients, applications for finding tomorrow's weather or searching train schedules
-all communicate with one or more services over the network and display the resulting information. 
-Increasing number of such services provide REST-based end-points that return data as CSV, XML
-or JSON. Despite numerous schematization efforts, most services do not come with an 
-explicit schema. At best, the documentation provides sample responses for typical requests.
-
-For example, OpenWeatherMap provides an end-point for getting current weather for a 
-given city\footnote{See ``Current weather data'': \url{http://openweathermap.org/current}}. 
-The documentation shows one sample to illustrate the typical response. Using standard 
-JSON and web libraries, we might write:
-%
+all communicate with external services. Increasing number of such services provide REST-based 
+end-points that return data as CSV, XML or JSON. Despite many schematization efforts, most 
+services do not come with an explicit schema. At best, the documentation provides sample responses 
+for typical requests.
+
+For example, \url{http://openweathermap.org/current} provides an end-point to get the current weather.
+The documentation shows one sample to illustrate the typical response. Using 
+standard JSON and web libraries, we might write:
+
+\noindent
 \begin{equation*}
@@
 %
-The code assumes that the response has a particular format described in the documentation. The
+The code assumes that the response has a particular shape described in the documentation. The
 root node must be a record with a \str{main} field, which has to be another record containing
-a numerical \str{temp} field. When the format is incorrect, the data access simply fails
-with an exception.
+a numerical \str{temp} field. When the shape is different, the data access simply fails
+with an exception. The code, while not immediately unsound, is manifestly prone to errors if
+strings are misspelled or incorrect shape assumed.
 
@@
 %
+\vspace{-0.1em}
 \begin{equation*}
@@
 
-The rest of the paper describes the mechanism and discusses its safety properties. 
-The key novel contributions are:
+In short, \emph{the types come from the sample data}. In our experience, this technique is 
+immensely practical and surprisingly effective in achieving sound information interchange 
+in heterogeneous systems. This paper presents our approach:
 
@@
 
-\item We describe a predictable type inference algorithm for structured data formats, 
-  based on a \emph{common supertype} relation, that underlies the type providers 
+\item We describe a predictable shape inference algorithm for structured data formats, 
+  based on a \emph{preferred shape} relation, that underlies the type providers 
   (\S\ref{sec:inference}).
@@
   This adapts the ML-style type safety for the context of the web.
-
-\item The supplementary screencast show developer experience using 
+\end{itemize}
+
+\vspace{-0.1em}
+\noindent
+The supplementary screencast illustrates the practical developer experience using 
   F\# Data with JSON, XML and CSV.
-\end{itemize}
 

### 2015-11-19 pldi-2016/paper.tex
Final edits in Section 2

@@
 root node must be a record with a \str{main} field, which has to be another record containing
-a numerical \str{temp} field. When the shape is different, the data access simply fails
-with an exception. The code, while not immediately unsound, is manifestly prone to errors if
-strings are misspelled or incorrect shape assumed.
+a numerical \str{temp} field. When the shape is different, the code simply fails with an exception. 
+While not immediately unsound, the code is manifestly prone to errors if strings are misspelled 
+or incorrect shape assumed.
 
@@
 
-\section{Structural type providers}
+\section{Type providers for structured data}
 \label{sec:providers}
@@
 We start with an informal overview that shows how F\# Data type providers simplify working with 
-JSON and XML. We introduce the necessary aspects of F\# type providers along the 
-way. The examples in this section illustrate a number of key properties of our type inference 
-algorithm:
+JSON and XML. We introduce the necessary aspects of F\# type providers along the way. The examples 
+in this section illustrate the key design principles of the shape inference algorithm:
 
 \begin{itemize}
-\item Our mechanism is predictable. The user directly works with the inferred types and should 
-  understand why a specific type was inferred from a given sample.\footnote{In particular, we do 
+\item The mechanism is predictable. The user directly works with the provided types and should 
+  understand why a specific type was produced from a given sample.\footnote{In particular, we do 
   not use probabilistic methods where adding an additional sample could completely change the 
@@
 
-\item Our inference mechanism prefers records. This supports developer tooling --
-  most F\# editors provide code completion hints on ``.'' and so types with properties (records)
-  are easier to use than types that require pattern matching.
-
-\item Finally, we handle a number of practical concerns important in the real world. This includes 
+\item The type providers prefer F\# objects with properties. This is to allow extensible 
+  (open-world) data formats (\S\ref{sec:providers-xml}). It also interacts well with developer tooling
+  as most F\# editors provide auto-completion on ``.'' and so objects are easier to use than 
+  types that require pattern matching.
+
+\item Finally, the mechanism handles practical concerns important in the real world. This includes 
   support for different numerical types, \kvd{null} values and missing data.
@@
 
-The JSON format used in \S\ref{sec:introduction} is a popular data exchange format based on 
+The JSON format is a popular data exchange format based on 
 JavaScript data structures. The following is the definition of \ident{JsonValue} 
-used earlier to represent JSON:
+used earlier (\S\ref{sec:introduction}) to represent JSON data:
 %
@@
 \noindent
-Say we want to print the names and an age if it is available. The standard approach would 
-be to pattern match on the parsed \ident{JsonValue}, check that the top-level node is a \ident{Array}
-and iterate over the elements checking that each element is a \ident{Record} with certain properties. 
-We would throw an exception for values in an incorrect format. Even with helper functions, 
-we would still need to specify field names as strings, which is error prone and can not be 
-statically checked.
+The standard approach to print the names and ages would be to pattern match on the parsed 
+\ident{JsonValue}, check that the top-level node is a \ident{Array} and iterate over the elements 
+checking that each element is a \ident{Record} with certain properties. We would throw an exception 
+for values of an incorrect shape. As before, the code would specify field names as strings, which 
+is error prone and can not be statically checked.
 
 Assuming \strf{people.json} is the above example and \ident{data} is a string containing
-JSON in the same format, we can write:
+JSON of the same shape, we can write:
 %
@@
 In contrast to the earlier example, we now use a local file \strf{people.json} as a sample for 
-the type inference, but then processes data from another source. The code achieves the same 
+the type inference, but then processes data from another source. The code achieves a similar 
 simplicity as when using dynamically typed languages, but it is statically type-checked.
@@
 execute the type provider in the background at development-time and use the provided types 
-in auto-completion.
-
-The \ident{JsonProvider} uses a type inference algorithm (\S\ref{sec:inference})  and
-infers the following types from the sample:
+in auto-completion and background type-checking.
+
+The \ident{JsonProvider} uses a shape inference algorithm and provides
+the following F\# types for the sample:
 %
@@
 sample used for the inference and \ident{Parse} parses a JSON string. This lets us read
-data at runtime, provided that it is in the same format as the static sample.
+data at runtime, provided that it has the same shape as the static sample.
 
@@
 the same as in the earlier hand-written sample (\S\ref{sec:introduction}) -- a member access 
-throws an exception if the document does not have the expected format.
-
-Informally, the safety property (\S\ref{sec:safety}) states that if the inputs
-are subtypes of some of the provided samples (i.e.~the samples are representative), then
-no exceptions will occur. In other words, we cannot avoid all failures, but we can prevent some.
-Moreover, if the OpenWeatherMap changes the response format, the sample in \S\ref{sec:introduction}
-will not re-compile and the user knows that the code needs to be changed.
-
-\paragraph{The role of records.}
-The sample code is easy to write thanks to the fact that most F\# editors provide code completion
+throws an exception if data does not have the expected shape.
+
+Informally, the safety property (\S\ref{sec:safety}) states that if the inputs are compatible
+with one of the static samples (i.e.~the samples are representative), then no exceptions will 
+occur. In other words, we cannot avoid all failures, but we can prevent some. Moreover, if 
+\url{http://openweathermap.org} changes the shape of the response, the code in \S\ref{sec:introduction}
+will not re-compile and the developer knows that the code needs to be changed.
+
+\paragraph{The role of objects with properties.}
+The sample code is easy to write thanks to the fact that most F\# editors provide auto-completion
 when ``.'' is typed (see the supplementary screencast). The developer does not need to look at the 
-sample JSON file to see what fields are available. To support this scenario, our inference algorithm 
-prefers records (we also treat XML elements and CSV rows as records) and only infers types that 
-require pattern matching as the last resort.
-
-In the above example, this is demonstrated by the fact that \ident{Age} is marked as optional.
+sample JSON file to see what fields are available. To support this scenario, our type providers 
+map the inferred shapes to F\# objects with (possibly optional) properties.
+
+This is demonstrated by the fact that \ident{Age} becomes an optional member.
 An alternative is to provide two different record types (one with \ident{Name} and other with 
 \ident{Name} and \ident{Age}), but this would complicate the processing code.
+It is worth noting that languages with stronger tooling around pattern matching
+such as Idris \cite{idris-tools} might have different preferences.
 
@@
 
-Consider a simple extensible document format where a root element
-{\ttfamily\small <doc>} can contain a number of document element, one of which is
-{\ttfamily\small <heading>} representing headings:
+Consider a simple extensible document format where a root element {\ttfamily\small <doc>} can 
+contain a number of document elements, one of which is {\ttfamily\small <heading>} representing 
+headings:
 %
@@
 \noindent
-The F\# Data library has been designed primarilly to simplify reading of data. For example,
+The F\# Data library has been designed primarily to simplify reading of data. For example,
 say we want to print all headings in the document. The sample shows a part of the document structure 
@@
 %
-The example iterates over a collection of elements returned by \ident{root.Doc}. The type of element 
+The example iterates over a collection of elements returned by \ident{root.Doc}. The type of \ident{elem} 
 provides a typed access to elements known from the sample and so we can write \ident{elem.Heading}, 
@@
 
-\paragraph{Open world and variant types.}
-We do not infer a union type representing a choice between heading, paragraph and image. By its 
-nature, XML is extensible and the sample cannot include all possible nodes.\footnote{Even when the 
-document structure is defined using XML Schema, documents may contain elements prefixed with other 
-namespaces.} The type of \ident{elem} is thus a top type (we call it a \emph{variant}) annotated 
-with the statically known possibilities. It is mapped to the following F\# type:
+\paragraph{Open world.}
+By its nature, XML is extensible and the sample cannot include all possible nodes.\footnote{Even 
+when the document structure is defined using XML Schema, documents may contain elements prefixed 
+with other namespaces.} This is the important \emph{open world assumption} about external data. 
+Actual input might be an element about which nothing is known.
+
+For this reason, we do not infer a closed choice between heading, paragraph and image. In the 
+subsequent formalization, we introduce a \emph{top shape} (\S\ref{?}) and extend it with labels 
+capturing the statically known possibilities (\S\ref{?}). The \emph{labelled top shape} is mapped 
+to the following type:
 %
@@
 %
-The type provides a nice access to the statically known elements, but it encodes the 
-\emph{open-world assumption}. The user also needs to handle the case when the value is none 
-of the statically known elements.
-
-As a consequence, the variant type can be seen as the top type. The above code will work
-correctly on document that contains any elements, including those not listed in the sample. This
-is extremely useful when reading data from external sources, because the sample document only
-needs to contain values relevant to the application. It is also worth noting that our type 
-inference attempts to minimze the number of variant types that appear in the resulting type --
-a variant is not used if there is another option (see Corollary~\ref{thm:no-unions} and 
-\S\ref{sec:impl-hetero}).
+This provides access to the elements known statically from the sample. However the type is 
+deliberately `weak', because the user needs to explicitly handle the case when a value is not 
+a statically known element. The above code uses \ident{Option.iter} to skip all unknown elements.
+
+The provided type is also consistent with our design principles, which prefers optional properties. 
+The gain is that the provided types support both open-world data and developer tooling. It is also 
+worth noting that our shape inference uses labelled top shapes only as the last resort 
+(Corollary~\ref{thm:no-unions}, \S\ref{sec:impl-hetero}).
 
@@
 \label{sec:providers-sum}
-In addition to type providers for JSON and XM, F\# Data also implements a type provider for CSV 
-(\S\ref{sec:impl-parsing}). We treat CSV files as lists of records (with field for each column) 
-and so CSV is handled directly by our inference algorithm.
-
-Throughout the introduction, we used data sets that demonstrate typical issues that are frequent
-in the real-world (missing data, inconsistent encoding of primitive values and heterogeneous 
-structures). In our experience, these are the most common issues. The following JSON response with 
-government debt information returned by the World Bank API\footnote{Available at 
+Throughout the introduction, we used data sets that demonstrate the typical problems that are frequent
+in the real-world (\emph{missing data}, \emph{inconsistent encoding} of primitive values and 
+\emph{heterogeneous shapes}). In our experience, these are the most common issues. The following 
+JSON response with government debt information returned by the World Bank\footnote{Available at 
 \url{http://data.worldbank.org}} demonstrates all three problems:
@@
 types of JavaScript.}
+
+In addition to type providers for JSON and XM, F\# Data also implements a type provider for CSV 
+(\S\ref{sec:impl-parsing}). We treat CSV files as lists of records (with field for each column) 
+and so CSV is handled directly by our inference algorithm.
 
@@
 
-\subsection{Relativized safety in practicve}
+\subsection{Relativized safety in practice}
 \label{sec:safety-discuss}

### 2015-11-20 pldi-2016/paper.tex
Final edits in Section 3

@@
  \kvd{type}~\ident{JsonValue} = \\[0.1em]
- \quad|~ \ident{Number}~\kvd{of}~\ident{decimal} ~~~~|~ \ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
+ \quad|~ \ident{Number}~\kvd{of}~\ident{float} ~~~~|~ \ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
  \quad|~ \ident{String}~\kvd{of}~\ident{string} ~~\qquad\, |~ \ident{Null} \\[0.1em]
@@
  \quad \kvd{member}~\ident{Name}~:~\ident{string} \\
- \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{decimal}\rangl \\[0.5em]
+ \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{float}\rangl \\[0.5em]
  \kvd{type}~\ident{People}~=  \\
@@
 because the value is missing in one sample. The two age values are an integer $25$ and a 
-decimal $3.5$ and so the common inferred type is \ident{decimal}.
+float $3.5$ and so the common inferred type is \ident{float}.
 
@@
 For this reason, we do not infer a closed choice between heading, paragraph and image. In the 
-subsequent formalization, we introduce a \emph{top shape} (\S\ref{?}) and extend it with labels 
-capturing the statically known possibilities (\S\ref{?}). The \emph{labelled top shape} is mapped 
-to the following type:
+subsequent formalization, we introduce a \emph{top shape} (\S\ref{sec:inference-types}) and extend 
+it with labels capturing the statically known possibilities (\S\ref{sec:inference-vars}). The 
+\emph{labelled top shape} is mapped to the following type:
 %
@@
 worth noting that our shape inference uses labelled top shapes only as the last resort 
-(Corollary~\ref{thm:no-unions}, \S\ref{sec:impl-hetero}).
+(Theorem~\ref{thm:lub}, \S\ref{sec:impl-hetero}).
 
@@
 %
-%   #######                                                                                  
-%      #    #   # #####  ######    # #    # ###### ###### #####  ###### #    #  ####  ###### 
-%      #     # #  #    # #         # ##   # #      #      #    # #      ##   # #    # #      
-%      #      #   #    # #####     # # #  # #####  #####  #    # #####  # #  # #      #####  
-%      #      #   #####  #         # #  # # #      #      #####  #      #  # # #      #      
-%      #      #   #      #         # #   ## #      #      #   #  #      #   ## #    # #      
-%      #      #   #      ######    # #    # #      ###### #    # ###### #    #  ####  ###### 
+%     #####
+%    #     # #    #   ##   #####  ######    # #    # ###### ###### #####  ###### #    #  ####  ######
+%    #       #    #  #  #  #    # #         # ##   # #      #      #    # #      ##   # #    # #
+%     #####  ###### #    # #    # #####     # # #  # #####  #####  #    # #####  # #  # #      #####
+%          # #    # ###### #####  #         # #  # # #      #      #####  #      #  # # #      #
+%    #     # #    # #    # #      #         # #   ## #      #      #   #  #      #   ## #    # #
+%     #####  #    # #    # #      ######    # #    # #      ###### #    # ###### #    #  ####  ######
 %
@@
 
-\section{Structural type inference}
+\section{Shape inference for structured data}
 \label{sec:inference}
 
-Our type inference algorithm for structured formats is based on a subtyping relation. When 
-inferring the type, we infer the most specific types of individual values (CSV rows, JSON or XML 
-nodes) and recursively find a common supertype of all child nodes or all sample documents.
-
-We first define the type of structured data $\sigma$. Note that this type is distinct 
-from the programming language types $\tau$ (type providers generate the latter from the former). 
-Next, we define the subtyping relation on structural types $\sigma$ and describe the algorithm 
-for finding a common supertype.
-
-\subsection{Structural types}
-\label{sec:inference-types}
-
-We distinguish between \emph{non-nullable types} that always have a valid value (written as 
-$\hat{\sigma}$) and \emph{nullable types} that encompass missing and \kvd{null} values 
-(written as $\sigma$). We write $\nu$ for record field names and $\nu_?$ for names that can be empty:
-%
-\begin{equation*}
-\begin{array}{rcl}
- \hat{\sigma} &\narrow{=}& \nu_?\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
-                &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
- \\[0.6em] 
-       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{option}\langl \hat{\sigma} \rangl \lsep [\sigma] \\[0.1em]
-              &\narrow{|}& \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \lsep ~\bot~ \lsep \kvd{null}
-\end{array}
-\end{equation*}
-
-\noindent
-Non-nullable types include records (consisting of an optional name and fields with their types) 
-and primitive types. Records arising from XML documents are named, while records used by 
-JSON and CSV providers are unnamed.
-
-The three numerical types, \ident{int} for integers, \ident{decimal} for small precise decimal 
-numbers and \ident{float} for floating-point numbers, are related by sub-typing as discussed in 
-\S\ref{sec:inference-subtyping}.
-
-Any non-nullable type can be wrapped in the \kvd{option} constructor to explicitly permit the 
-\kvd{null} value. Type providers map \kvd{option} to standard F\# option. A collection $[\sigma]$ 
-is also nullable and \kvd{null} values are treated as empty collections. The type 
-$\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded notation) and $\bot$ 
-is the bottom type.
-
-A variant type \kvd{any} is annotated with the types it may represent in the sample document.
-As discussed earlier (\S\ref{sec:providers-xml}), variants force the user to handle the case when 
-a value is of a different type than the statically known ones and so it can be seen as a family of 
-top types. For the same reason, variants also implicitly permit the \kvd{null} value.
+The shape inference algorithm for structured data is based on a shape preference relation. When 
+inferring the shape, it infers the most specific shapes of individual values (CSV rows, JSON or XML 
+nodes) and recursively finds a common shape of all child nodes or all sample documents.
+
+We first define the shape of structured data $\sigma$. We use the term \emph{shape} to distinguish 
+shapes from programming language \emph{types} $\tau$ (type providers generate the latter from the former). 
+Next, we define the preference relation on shapes $\sigma$ and describe the algorithm 
+for finding a common shape.
 
@@
 
-\subsection{Subtyping relation}
+\subsection{Inferred shapes}
+\label{sec:inference-types}
+
+We distinguish between \emph{non-nullable shapes} that always have a valid value (written as 
+$\hat{\sigma}$) and \emph{nullable shapes} that encompass missing and \kvd{null} values 
+(written as $\sigma$). We write $\nu$ for record names and record field names. In the rest of the
+paper, we assume that record fields can be freely reordered:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \hat{\sigma} &\narrow{=}& \nu \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
+                &\narrow{|}& \ident{float} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
+ \\[0.6em] 
+       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{nullable}\langl \hat{\sigma} \rangl \lsep [\sigma] \lsep \kvd{any} \lsep \kvd{null}  \lsep ~\bot~
+\end{array}
+\end{equation*}
+
+\noindent
+Non-nullable shapes include records (consisting of a name and fields with their shapes) and 
+primitives. Names of records arising from XML are the names of the XML elements while
+JSON records always use a single name $\bullet$.
+
+We include two numerical primitives, \ident{int} for integers and \ident{float} for floating-point 
+numbers. The two are related by the perference relation and we prefer \ident{int}.
+
+Any non-nullable shape $\hat{\sigma}$ can be wrapped as $\kvd{nullable}\langl\hat{\sigma}\rangl$ to 
+explicitly permit the \kvd{null} value. Type providers map \kvd{nullable} shapes to the F\# option
+type. A collection $[\sigma]$ is also nullable and \kvd{null} values are treated as empty 
+collections. The shape $\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded 
+notation) and $\bot$ is the bottom shape. The \kvd{any} shape is the top shape, but we revisit it 
+later by adding labels for statically known alternative shapes (\S\ref{sec:inference-vars}) as 
+discussed earlier (\S\ref{sec:providers-xml}).
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Preferred shape relation}
 \label{sec:inference-subtyping}
 
-Figure~\ref{fig:subtyping-diagram} provides a basic intuition about the subtyping between
-structural types. The upper part shows non-nullable types (with records and primitive types) and 
-the lower part shows nullable types with \kvd{null}, collections and optional values. We 
-abbreviate $\kvd{option}\langl\sigma\rangl$ as $\sigma?$ and we omit links between the two parts;
-any type $\hat{\sigma}$ is a subtype of $\kvd{option}\langl\hat{\sigma}\rangl$.
+Figure~\ref{fig:subtyping-diagram} provides an intuition about the preference between 
+shapes. The upper part shows non-nullable shapes (with records and primitives) and the lower part 
+shows nullable shapes with \kvd{null}, collections and nullable shapes. In the diagram, we 
+abbreviate $\kvd{nullable}\langl\sigma\rangl$ as $\sigma?$ and we omit links between the two parts;
+a shape $\hat{\sigma}$ is preferred over $\kvd{nullable}\langl\hat{\sigma}\rangl$.
 
 \begin{definition}
-We write $\sigma_1 :> \sigma_2$ to denote that $\sigma_2$ is a subtype of $\sigma_1$. The 
-subtyping relation is defined as a transitive reflexive closure of the following rules:
+We write $\sigma_1 \sqsupseteq \sigma_2$ to denote that $\sigma_2$ is preferred over $\sigma_1$. 
+The shape preference relation is defined as a transitive reflexive closure of the following rules:
 
@@
   \label{eq:sub-prim}\tag{P1}
-  \ident{float}\,:>\,\ident{decimal}\,&:>\,\ident{int}&\\[-0.2em]
+  \ident{float}\,&\sqsupseteq\,\ident{int}&\\[-0.2em]
   \label{eq:sub-null}\tag{P2}
-  \sigma &:> \kvd{null}  &(\textnormal{iff}~\sigma \neq \hat{\sigma})  \\[-0.2em]
+  \sigma &\sqsupseteq \kvd{null}  &(\textnormal{iff}~\sigma \neq \hat{\sigma})  \\[-0.2em]
   \label{eq:sub-opt}\tag{P3}
-  \kvd{option}\langl\hat{\sigma}\rangl &:> \hat{\sigma}  &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
+  \kvd{nullable}\langl\hat{\sigma}\rangl &\sqsupseteq \hat{\sigma}  &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
   \label{eq:sub-opt-cov}\tag{P4}
-  \kvd{option}\langl\hat{\sigma_1}\rangl &:> 
-    \kvd{option}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} :> \hat{\sigma_2})\\[-0.2em]
+  \kvd{nullable}\langl\hat{\sigma_1}\rangl &\sqsupseteq 
+    \kvd{nullable}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} \sqsupseteq \hat{\sigma_2})\\[-0.2em]
   \label{eq:sub-col}\tag{P5}
-  [\sigma_1] &:> [\sigma_2]  &(\textnormal{if}~\sigma_1 :> \sigma_2) \\[-0.2em]
+  [\sigma_1] &\sqsupseteq [\sigma_2]  &(\textnormal{if}~\sigma_1 \sqsupseteq \sigma_2) \\[-0.2em]
   \label{eq:sub-bot}\tag{P6}
-  \sigma &:> \bot  &(\textnormal{for all}~\sigma)\\[-0.2em]
+  \sigma &\sqsupseteq \bot  &(\textnormal{for all}~\sigma)\\[-0.2em]
   \label{eq:sub-var-top}\tag{P7}
-  \kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl &:> \sigma 
+  \kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl &\sqsupseteq \sigma 
 \end{align}
+\vspace{-2em}
 
@@
 \begin{array}{l}
- \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
- \nu_?~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
-\end{array} \qquad ~~~~(\textnormal{if}~\sigma_i :> \sigma_i')
+ \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsupseteq \\
+ \quad\nu~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
+\end{array} \qquad ~~~~(\textnormal{if}~\sigma_i \sqsupseteq \sigma_i')
 \end{align}
-\vspace{-1em}
+\vspace{-1.5em}
 \begin{align}
@@
 \begin{array}{l}
- \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
- \nu_?~\{ \nu_1\!:\!\sigma_1, .., .., \nu_m\!:\!\sigma_m \}
-\end{array} \qquad (\textnormal{when}~m \geq n)
+ \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsupseteq \\
+ \quad\nu~\{ \nu_1\!:\!\sigma_1, .., .., \nu_m\!:\!\sigma_m \}
+\end{array} \quad~~\, (\textnormal{when}~m \geq n)
 \end{align}
-\vspace{-1em}
+\vspace{-1.5em}
 \begin{align}
@@
 \begin{array}{l}
- \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
- \nu_?~\{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, .., \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
-\end{array} ~~(\pi~\textnormal{perm.})
-\end{align}
-\vspace{-1em}
-\begin{align}
-\label{eq:sub-record4}\tag{R4}
-\begin{array}{l}
- \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null} \} :> \\
- \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
+ \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null} \} \sqsupseteq \\
+ \quad\nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
 \end{array}
@@
 \begin{equation*}
-\boxed{
-\begin{array}{c}
 \begin{array}{rclcl}
- \tytag &\narrow{=}& \ident{bool} \\
-        &\narrow{\lsep}& \ident{string} &\narrow{\lsep}& \ident{number}  \\
-        &\narrow{\lsep}& \ident{any}  &\narrow{\lsep}& \ident{collection} \\
-        &\narrow{\lsep}& \ident{record} &\narrow{\lsep}& \ident{named}\;~\nu 
+ \tytag &\narrow{=}& \ident{collection}  &\narrow{\lsep}& \ident{number} \\
+        &\narrow{\lsep}& \ident{nullable} &\narrow{\lsep}& \ident{string}  \\
+        &\narrow{\lsep}& \nu ~\lsep \ident{any} &\narrow{\lsep}& \ident{bool}
 \end{array}
@@
  \tytagof(\ident{bool}) &\narrow{=}& \ident{bool}\\
- \tytagof([\sigma]) &\narrow{=}& \ident{collection}\\
- \tytagof(\hat{\sigma}~\kvd{option}) &\narrow{=}& \tytagof(\hat{\sigma})
+ \tytagof(\ident{int}) &\narrow{=}& \ident{number}\\
+ \tytagof(\ident{float}) &\narrow{=}& \ident{number}\\
 \end{array}
@@
  \tytagof(\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl) &\narrow{=}& \ident{any}\\
- \tytagof(\{ \nu_1 : \sigma_1, \; \ldots \; , \nu_n : \sigma_n \}) &\narrow{=}& \ident{rec-anon}\\
- \tytagof(\nu\; \{ \nu_1 : \sigma_1, \; \ldots \; , \nu_n : \sigma_n \}) &\narrow{=}& \ident{rec-named}~\nu \\
- \multicolumn{3}{c}{ \tytagof(\sigma) = \ident{number}\quad \sigma \in\{ \ident{int},\ident{decimal},\ident{float} \}  } 
-\end{array}
-%
-\\[3em]
-%
+ \tytagof(\nu\; \{ \nu_1 : \sigma_1, \; \ldots \; , \nu_n : \sigma_n \}) &\narrow{=}& \nu \\
+ \tytagof(\kvd{nullable}\langl\hat{\sigma}\rangl) &\narrow{=}& \ident{nullable}\\
+ \tytagof([\sigma]) &\narrow{=}& \ident{collection}\\
+\end{array}
+\end{equation*}
+
+\vspace{-0.5em}
+\begin{equation*}
+\quad\qquad\qquad
 \begin{array}{rcll}
- \addopt{\hat{\sigma}} &\narrow{=}& \hat{\sigma}~\kvd{option} &(\textnormal{non-nullable types})\\
+ \addopt{\hat{\sigma}} &\narrow{=}& \kvd{nullable}\langl\hat{\sigma}\rangl &(\textnormal{non-nullable shapes})\\
  \addopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
@@
 \begin{array}{rcll}
- \dropopt{\hat{\sigma}~\kvd{option}} &\narrow{=}& \hat{\sigma} &(\textnormal{option})\\
+ \dropopt{\kvd{nullable}\langl\hat{\sigma}\rangl} &\narrow{=}& \hat{\sigma} &(\textnormal{nullable shape})\\
  \dropopt{\sigma} &\narrow{=}& \sigma &(\textnormal{otherwise})
 \end{array}
-\end{array}
-}
-\end{equation*}
-
-\begin{equation*}
-\hspace{3em}
-\inference[(record-1)\;]
+\end{equation*}
+
+\vspace{-0.5em}
+\begin{equation*}
+% option types
+\inference[(nullable)\;]
+  {\hat{\sigma_1} \tsep \sigma_2 \vdash \sigma}
+  {\kvd{nullable}\langl\hat{\sigma_1}\rangl \tsep \sigma_2 \vdash \kvd{nullable}\langl\dropopt{\sigma}\rangl}
+\qquad
+\inference[(any)\;]
+  { \tytagof(\sigma_1) \neq \tytagof(\sigma_2) & \tytagof(\sigma_1) \neq \ident{nullable} \neq \tytagof{\sigma_2}}
+  { \sigma_1 \tsep \sigma_2 \vdash \kvd{any} }  
+\end{equation*}
+\vspace{-2em}
+
+\begin{equation*}
+\hspace{4em}
+\inference[(record)\;]
   { (\nu_i = \nu'_j) \Leftrightarrow (i = j) \wedge (i \leq k)
@@
   { \begin{array}{l}
-    \nu_? \; \{ \nu_1 \!:\! \sigma_1,  \; \ldots \;, \nu_k \!:\! \sigma_k, \; \ldots \;, \nu_n \!:\! \tau_n \} \tsep
-    \nu_? \; \{ \nu'_1 \!:\! \sigma'_1, \; \ldots \;, \nu'_k \!:\! \sigma'_k, \; \ldots \;, \nu'_m \!:\! \tau'_m \} \vdash\\
-    \nu_? \; \{ \nu_1 \!:\! \sigma''_1, \; \ldots \; , \nu_k \!:\! \sigma''_k, 
+    \nu \; \{ \nu_1 \!:\! \sigma_1,  \; \ldots \;, \nu_k \!:\! \sigma_k, \; \ldots \;, \nu_n \!:\! \tau_n \} \tsep
+    \nu \; \{ \nu'_1 \!:\! \sigma'_1, \; \ldots \;, \nu'_k \!:\! \sigma'_k, \; \ldots \;, \nu'_m \!:\! \tau'_m \} \vdash\\
+    \nu \; \{ \nu_1 \!:\! \sigma''_1, \; \ldots \; , \nu_k \!:\! \sigma''_k, 
                             \nu_{k+1} \!:\! \addopt{\sigma_{k+1}}, \ldots, \nu_n \!:\! \addopt{\sigma_n},
@@
 
-% union with something else
-\begin{equation*}
-\textnormal{\footnotesize{(var-1)}}\;\;
-\inference
-  {\exists i . \tytagof(\sigma_i) = \tytagof(\sigma) & \sigma \tsep \sigma_i \vdash \sigma_i' & \tytagof(\sigma)\neq\ident{any}}
-  {\sigma \tsep \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \vdash 
-    \kvd{any}\langl\sigma_1, \ldots, \dropopt{\sigma_i'}, \ldots, \sigma_n\rangl}
-\;\;
-\inference
-  {\nexists i . \tytagof(\sigma_i) = \tytagof(\sigma) & \tytagof(\sigma)\neq\ident{any}}
-  {\sigma \tsep \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \vdash 
-    \kvd{any}\langl\sigma_1, \ldots, \sigma_n, \dropopt{\sigma}\rangl}
-\end{equation*}
-\vspace{-2em}
-
-% two union types
-\begin{equation*}
-\begin{array}{l}
-\inference[(var-2)\;]
-  { (\forall i\in \{ 1 .. k \})\quad \tytagof(\sigma_i) = \tytagof(\sigma'_i) \quad 
-    \sigma_i \tsep \sigma_i' \vdash \sigma''_i}
-  { \begin{array}{l}
-    \kvd{any}\langl \sigma_1, \ldots, \sigma_k,  \ldots, \sigma_n\rangl \tsep 
-    \kvd{any}\langl \sigma'_1, \ldots, \sigma'_k, \ldots, \sigma'_m\rangl \vdash\\
-    \kvd{any}\langl \sigma''_1, \ldots, \sigma''_k, \sigma_{k+1}, \ldots, \sigma_{n}, \sigma'_{k+1}, \ldots, \sigma'_{m}\rangl
-    \end{array} }
-\end{array}    
-\begin{array}{l}
-\inference[(var-3)\;]
-  {(\forall i\in\{1,2\}) & \tytagof(\sigma_1) \neq \tytagof(\sigma_2) \\ 
-   \tytagof(\sigma_i) \neq \ident{any} & \nexists\sigma_i' . (\sigma_i = \kvd{option}\langl\sigma_i'\rangl)  }
-  {\sigma_1 \tsep \sigma_2 \vdash \kvd{any}\langle\dropopt{\sigma_1}, \dropopt{\sigma_2}\rangle}
-\end{array}  
-\end{equation*}
-\vspace{-2em}
-
-\begin{equation*}
-\inference[(order-1)\;]
-  { \nu_? \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \tsep \sigma \vdash \sigma' }
-  { \nu_? \; \{ \nu_{\pi(1)} : \sigma_{\pi(1)}, \ldots, \nu_{\pi(n)} : \sigma_{\pi(n)} \} \tsep \sigma \vdash \sigma' }
-\quad
-\inference[(order-2)\;]
-  { \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \tsep \sigma \vdash \sigma' }
-  { \kvd{any}\langl\sigma_{\pi(1)}, \ldots, \sigma_{\pi(n)}\rangl \tsep \sigma \vdash \sigma' }
-\quad (\pi~\textnormal{perm.})  
-\end{equation*}
-\vspace{-2em}
-
-\begin{equation*}
-\hspace{3em}
-% option types
-\inference[(opt)\;]
-  {\hat{\sigma_1} \tsep \sigma_2 \vdash \sigma}
-  {\kvd{option}\langl\hat{\sigma_1}\rangl \tsep \sigma_2 \vdash \kvd{option}\langl\dropopt{\sigma}\rangl}
-\qquad
-% primitive types
-\inference[(prim)\;]
-  {\sigma_1 :> \sigma_2 &
-   \tytagof(\sigma_1) = \tytagof(\sigma_2) = \ident{number} }
-  {\sigma_1 \tsep \sigma_2 \vdash \sigma_1}\quad
-\end{equation*}
-\vspace{-2em}
-
 % all rules are symmetric and reflexive
 \begin{equation*}
+\hspace{2em}
 \inference[(list)\;]
@@
 \quad
-% null and nullable
- \begin{array}{l}
- \textnormal{\footnotesize{(null-1)}}\;\; \sigma \tsep \kvd{null} \vdash \sigma \quad(\sigma :> \kvd{null}) \\[0.6em]
- \textnormal{\footnotesize{(null-2)}}\;\; \sigma \tsep \kvd{null} \vdash \sigma~\kvd{option} \quad(\sigma :\ngtr \kvd{null})
- \end{array}
-\end{equation*}
-
-\caption{Inference judgements that define the common supertype relation}
+% null and primitives
+\begin{array}{l}
+ \textnormal{\footnotesize{(null)}}\;\; \sigma \tsep \kvd{null} \vdash \addopt{\sigma} \quad (\sigma \neq \bot)\\[0.6em]
+ \textnormal{\footnotesize{(prim)}}\;\; \kvd{int} \tsep \kvd{float} \vdash \kvd{float}
+\end{array}
+\end{equation*}
+
+\caption{Inference judgements that define the common preferred shape relation}
 \label{fig:subtyping-cst}
@@
 \begin{itemize}
-\item Numeric types (\ref{eq:sub-prim}) are ordered by the range they represent;
-  \ident{int} is a 32-bit integer, \ident{decimal} has a range of $-1^{29}$ to $1^{29}$ and 
-  \ident{float} is a floating-point number (but imprecise).
-
-\item The \kvd{null} type is a subtype of all nullable types (\ref{eq:sub-null}), i.e. all 
-  types excluding non-nullable types $\hat{\sigma}$. Any non-nullable type is a 
-  subtype of its optional version (\ref{eq:sub-opt}) and both options and collections are 
+\item Numeric shape with smaller range is preferred (\ref{eq:sub-prim}) and we choose 32-bit 
+\ident{int} over \ident{float} when possible.
+
+\item The \kvd{null} shape is preferred over all nullable shapes (\ref{eq:sub-null}), i.e. 
+  all shapes excluding non-nullable shapes $\hat{\sigma}$. Any non-nullable shape is preferred
+  over its nullable version (\ref{eq:sub-opt}); nullable shapes and collections are 
   covariant (\ref{eq:sub-opt-cov}, \ref{eq:sub-col}).
 
-\item There is a bottom type (\ref{eq:sub-bot}) and variants behave as top types, because
-  any type $\sigma$ is a subtype of any variant (\ref{eq:sub-var-top}). There is a range of
-  variants (no \emph{single} top type), but any variant is a subtype of any other variant.
-
-\item As usual, the subtyping on records is covariant (\ref{eq:sub-record1}), subtype can have 
-  additional fields (\ref{eq:sub-record2}) and fields can  be reordered (\ref{eq:sub-record3}). 
-  The interesting rule is (\ref{eq:sub-record4}). It implies that subtype can omit fields, 
-  provided that \kvd{null} is a valid value for the field (i.e. the field is optional).
+\item There is a bottom shape (\ref{eq:sub-bot}) and \kvd{any} behaves as the top shape, because
+  any shape $\sigma$ is preferred over \kvd{any} (\ref{eq:sub-var-top}).
+
+\item The record shapes are covariant (\ref{eq:sub-record1}) and preferred record can have 
+  additional fields (\ref{eq:sub-record2}). The rule (\ref{eq:sub-record3}) implies that records 
+  can omit fields, provided that \kvd{null} is a valid value for the 
+  field (i.e. the field is optional).
 \end{itemize}
@@
 \noindent
-If we were type-checking programs, defining variants as a family of top types (which are all 
-mutual subtypes) would not add any value over having a single top type, but our purpose is 
-different. We focus on reading structured data and, as shown earlier (\S\ref{sec:providers-xml}), 
-the types forming a variant provide a valuable information about the sample document.
-
-A particularly important rule is (\ref{eq:sub-record4}), which allows subtype to have fewer record 
-elements. This allows us to prefer records over variants. For example, given 
-$\{ \ident{name}\!:\!\ident{string} \}$ and $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int} \}$,
-we find a common supertype $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int}~\kvd{option} \}$.
-This is a least upper bound (clarified below) and more usable than another common supertype 
-$\kvd{var}\langl \{ \ident{name}\!:\!\ident{string} \}, \{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int} \}\rangl$.
+The rule that allows preferred shapes to have fewer fields (\ref{eq:sub-record3}) is particularly
+important. This is important, because it allows us to prefer records over the \kvd{any} 
+shape. Given $\{ \ident{name}\!:\!\ident{string} \}$ and $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int} \}$,
+we find a common shape $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int}~\kvd{option} \}$.
+This is a least upper bound and more usable than the top shape $\kvd{var}$.
 
 Some of the aspects of our system (numeric types and handling of missing data) offer a range of
-choices. For example, OCaml also has different numerical types and would likely always handle 
+choices. For example, OCaml also has different numerical types, but would likely always handle 
 \kvd{null} explicitly.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Common preferred shape relation}
+\label{sec:inference-commonsuper}
+
+Given two shapes, the \emph{common preferred shape} relation finds a least upper bound
+(Theorem~\ref{thm:lub}). It prefers records, which is important for usability as discussed
+earlier.
+
+\begin{definition}
+A \emph{common preferred shape} of shapes $\sigma_1$ and $\sigma_2$ is a shape $\sigma$, written 
+$\sigma_1 \tsep \sigma_2 \vdash \sigma$, obtained according to the inference rules in 
+Figure~\ref{fig:subtyping-cst}.
+\end{definition}
+
+\noindent
+We define shape \emph{tags} to identify shapes that have a common preferred shape which is not the
+top shape. When finding a common shape of two records (\emph{record}), we return a record that has 
+the union of their fields. The shape of shared fields becomes the common preferred shape of their 
+respective shapes; fields that are present in only one record are marked as nullable using 
+the function $\addopt{-}$.
+
+We can find a common shape of two different numbers (\emph{prim}); for two collections,
+we combine their elements (\emph{list}). When one shape is nullable (\emph{nullable}), we find
+the common non-nullable shape and wrap it in \kvd{nullable}; if one of the shapes is \kvd{null}, 
+we make the other one nullable (\emph{null}). Finally, if the two shapes do not have matching tags 
+(and they are not nullable), we have to infer the \kvd{any} shape as there is no better alternative. 
+The remaining rules for reflexivity, symmetry and the bottom shape are standard.
+
+\paragraph{Properties.}
+The set of shapes does not have a \emph{unique} least upper bound, but it 
+has a least upper bound with respect to an equivalence between mutually preferred shapes. 
+The common preferred shape relation defines a function (Theorem~\ref{thm:func}) that finds a
+least upper bound (Theorem~\ref{thm:lub}).
+
+The equivalence on shapes groups records by their non-nullable fields. Given 
+$\sigma_1 = \ident{A}~\{ \ident{a}\!:\!\ident{int}, \ident{b}\!:\!\kvd{nullable}\langl\ident{int}\rangl \}$,
+$\sigma_2 = \ident{A}~\{ \ident{a}\!:\!\ident{int} \}$ and
+$\sigma_3 = \ident{A}~\{ \ident{a}\!:\!\ident{int}, \ident{b}\!:\!\ident{int} \}$,
+it is the case that $\sigma_1 \sqsubseteq \sigma_2$ and $\sigma_1 \sqsupseteq \sigma_2$ but
+$\sigma_3 \sqsubseteq \sigma_1$ and not vice versa.
+
+Similarly to labels on the labelled top type (\S\ref{sec:inference-vars}), the nullable fields 
+of records serve as annotations that provide access to additional values that may or may not be 
+available in the input. In the provided code, these always have to be protected by a runtime check.
+
+\begin{theorem}[Preferred shape function]
+\label{thm:func}
+For all $\sigma_1$ and $\sigma_2$ there exists exactly one $\sigma$ such that 
+$\sigma_1 \tsep \sigma_2 \vdash \sigma$. 
+\end{theorem}
+\begin{proof}
+The pre-conditions of rules in Figure~\ref{fig:subtyping-cst} are disjoint, with the exception 
+of (\emph{sym}) and (\emph{refl}), but applying those does not affect the result. Reflexivity is
+preserved recursively and all rules, symmetry does not enable different derrivations, even for
+non-symmetric (\emph{nullable}) and (\emph{prim}).
+\end{proof}
+
+\begin{theorem}[Least upper bound]
+\label{thm:lub}
+If $\sigma_1 \triangledown \sigma_2 \vdash \sigma$ then $\sigma$ is a least upper bound, i.e. 
+$\sigma \sqsupseteq \sigma_1$ and $\sigma \sqsupseteq \sigma_2$ and for all $\sigma'$ such that $\sigma' \sqsupseteq \sigma_1$
+and $\sigma' \sqsupseteq \sigma_2$, it holds that $\sigma' \sqsupseteq \sigma$.
+\end{theorem}
+\begin{proof}
+By induction over $\vdash$. The algorithm only infers the top shape \kvd{any} when for non-nullable
+shapes of distinct tags and so no better preferred shape exists.
+\end{proof}
+
+\noindent
+Next, we extend the core model (sufficient for the relativized safety) 
+with \emph{labelled top types} discussed earleir.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure*}
+  \noindent
+  \begin{equation*}
+  \textnormal{\footnotesize{(top-1)}}\;\;
+  \inference
+    {\exists i . \tytagof(\sigma_i) = \tytagof(\dropopt{\sigma}) & \sigma \tsep \sigma_i \vdash \sigma_i' & \tytagof(\sigma)\neq\ident{any}}
+    {\sigma \tsep \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \vdash 
+      \kvd{any}\langl\sigma_1, \ldots, \dropopt{\sigma_i'}, \ldots, \sigma_n\rangl}
+  \;\;
+  \inference
+    {\nexists i . \tytagof(\sigma_i) = \tytagof(\dropopt{\sigma}) & \tytagof(\sigma)\neq\ident{any}}
+    {\sigma \tsep \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \vdash 
+      \kvd{any}\langl\sigma_1, \ldots, \sigma_n, \dropopt{\sigma}\rangl}
+  \end{equation*}
+  \vspace{-2em}
+
+  % two union types
+  \begin{equation*}
+  \begin{array}{l}
+  \inference[(top-2)\;]
+    { (\tytagof(\sigma_i) = \tytagof(\sigma'_j)) \Leftrightarrow (i = j) \wedge (i \leq k)\\
+        \forall i\in\{ 1 .. k \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
+    { \begin{array}{l}
+      \kvd{any}\langl \sigma_1, \ldots, \sigma_k,  \ldots, \sigma_n\rangl \tsep 
+      \kvd{any}\langl \sigma'_1, \ldots, \sigma'_k, \ldots, \sigma'_m\rangl \vdash\\
+      \kvd{any}\langl \sigma''_1, \ldots, \sigma''_k, \sigma_{k+1}, \ldots, \sigma_{n}, \sigma'_{k+1}, \ldots, \sigma'_{m}\rangl
+      \end{array} }
+  \end{array}    
+  \begin{array}{l}
+  \inference[(top-3)\;]
+    {(\forall i\in\{1,2\}) & \tytagof(\sigma_1) \neq \tytagof(\sigma_2) \\ 
+     \tytagof(\sigma_i) \neq \ident{any} & \nexists\sigma_i' . (\sigma_i = \kvd{option}\langl\sigma_i'\rangl)  }
+    {\sigma_1 \tsep \sigma_2 \vdash \kvd{any}\langle\dropopt{\sigma_1}, \dropopt{\sigma_2}\rangle}
+  \end{array}  
+  \end{equation*}
+  \caption{Extending the common preferred shape relation for labelled top shapes}
+  \label{fig:subtyping-cst-var}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Adding labelled top type}
+\label{sec:inference-vars}
+
+When analyzing the structure of shapes, it suffices to consider a single top shape \kvd{any}.
+However, the type providers need more information to provide typed access to the possible 
+alternative shapes of data, such as XML nodes. To support track the possible shapes, we label 
+the \kvd{any} shape:
+%
+\begin{equation*}
+\sigma = \ldots \lsep \kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl
+\end{equation*}
+%
+The shapes $\sigma_1, \ldots, \sigma_n$ represent statically known shapes that appear in the
+sample and that we expose in the provided type. As discussed earlier (\S\ref{sec:providers-xml})
+this is important when reading external \emph{open world} data. The labels do not affect the 
+preferred shape relation and $\kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl$ should still be
+seen as the top type, regardless of the labels.
+
+The common preferred shape relation is extended to find a labelled top shape that best represents 
+the sample. We limit the number of labels and avoid nesting by grouping shapes by the shape tag 
+introduced earlier (Figure~\ref{fig:subtyping-cst}).
+For example, rather than inferring $\kvd{any}\langl\ident{int}, \kvd{any}\langl\ident{bool}, \ident{float}\rangl\rangl$, 
+our algorithm finds the common preferred shape of $\ident{int}$ and $\ident{float}$ and produces 
+$\kvd{any}\langl\ident{float},\ident{bool}\rangl$.
+
+The three rules in Figure~\ref{fig:subtyping-cst-var} replace the earlier (\emph{top}). When 
+combining a top with another shape (\emph{top-1}), the labelled top shape may or may not already 
+contain a case with the tag of the other shape. If it does, the two shapes are combined, otherwise 
+a new case is added. When combining two top types (\emph{top-2}), we group the labels that have 
+shared tags. Finally, (\emph{top-3}) covers the case when we are combining two distinct non-top 
+shapes. As top shapes implicitly permit \kvd{null} values, we use an auxiliary function $\dropopt{-}$ 
+to make nullable shapes non-nullable (when possible) to simplify the label.
+
+\paragraph{Properties.}
+The revised algorithm stil finds a shape which is the least upper bound. This means that 
+labelled top shape is only inferred when there is no other alternative.
+
+Stating properties of the labels requires a finer relation than \emph{preferred shape}. In particular, 
+it can only be done with respect to the \emph{sample}. We leave the details to future work, but we
+note that the algorithm infers the best labels in the sense that there are labels that enable 
+typed access to every possible value in the sample, but not more. The same is the case for nullable
+fields of records.
+
+\subsection{Inferring shapes from values}
+\label{sec:formal-inferval}
+
+The common preferred shape relation is at the core of the shape inference. What remains to be 
+specified is how we obtain the shape from data. We represent JSON, XML and CSV documents using
+the same first-order \emph{data} value (\S\ref{sec:impl-parsing}):
+%
+\begin{equation*}
+\begin{array}{lcl}
+ d &\narrow{=}& i \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
+   &\narrow{|}& [d_1; \ldots; d_n] \lsep \nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}
+\end{array}
+\end{equation*}
+%
+The first few cases represent primitive values ($i$ for integers, $f$ for floating
+point numbers and $s$ for strings) and the \kvd{null} value. A collection is written as a 
+list of values in square brackets. A record starts with a name $\nu$, followed by a 
+sequence of field assignments $\nu_i \mapsto d_i$.
+
+The following defines a mapping $\semalt{d_1,\ldots,d_n}$ which turns a collection of sample 
+data $d_1, \ldots, d_n$ into a shape $\sigma$:
+%
+\begin{equation*}
+\begin{array}{rclcrclcrcl}
+ \semalt{i} &\narrow{=}& \ident{int} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} && \semalt{\kvd{true}} &\narrow{=}& \ident{bool}\\
+ \semalt{f} &\narrow{=}& \ident{float} && \semalt{s} &\narrow{=}& \ident{string} && \semalt{\kvd{false}}  &\narrow{=}& \ident{bool}\\
+\end{array}
+\end{equation*}
+\noindent
+\vspace{-0.5em}
+\begin{equation*}
+\begin{array}{l}
+ \semalt{d_1, \ldots, d_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
+ \qquad\sigma_0 = \bot,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{d_i} \vdash \sigma_i
+ \\[0.5em]
+ \semalt{[d_1; \ldots; d_n]} = [\semalt{d_1, \ldots, d_n}]
+ \\[0.5em]
+ \semalt{\nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}} =\\[0.1em]
+ \qquad\nu~\{ \nu_1:\semalt{d_1}, \ldots, \nu_n :\semalt{d_n} \}
+\end{array}
+\end{equation*}
+%
+We overload the notation and write $\semalt{s}$ when inferring shape from a single value. Primitive 
+values are mapped to their corresponding shapes. For records, we return infer field shapes
+from the individual values.
+
+When infering a shape from multiple samples, we use the common preferred shape relation to find a 
+common shape for all values (starting with $\bot$). This operation is used at the top-level 
+(when calling a type provider with multiple samples) and also when inferring the shape of collection
+elements.
+
+% ==================================================================================================
+% 
+%    #######
+%    #        ####  #####  #    #   ##   #      # ######   ##   ##### #  ####  #    #
+%    #       #    # #    # ##  ##  #  #  #      #     #   #  #    #   # #    # ##   #
+%    #####   #    # #    # # ## # #    # #      #    #   #    #   #   # #    # # #  #
+%    #       #    # #####  #    # ###### #      #   #    ######   #   # #    # #  # #
+%    #       #    # #   #  #    # #    # #      #  #     #    #   #   # #    # #   ##
+%    #        ####  #    # #    # #    # ###### # ###### #    #   #   #  ####  #    #
+%
+% ==================================================================================================
+
+\section{Formalising type providers}
+\label{sec:formal}
+
+In this section, we build the theoretical framework for proving relativised type safety of 
+structural type providers. We discuss the runtime representation of documents and 
+primitive operations (\S\ref{sec:formal-convert}), we embed these into a formal model of an F\# 
+subset (\S\ref{sec:formal-ff}) and we describe how structural type providers turn
+inferred structural types into F\# types (\S\ref{sec:formal-tp}).
 
@@
 \begin{array}{l}
-\ident{hasType}(\nu_? \{ \nu_1 \!:\! \sigma_1, \ldots, \nu_n \!:\! \sigma_n \}, \nu'_? \{ \nu'_1\mapsto s_1, \ldots, \nu'_m\mapsto s_m \}) \reduce (\nu_? = \nu'_?) ~\wedge \\
+\ident{hasType}(\nu \{ \nu_1 \!:\! \sigma_1, \ldots, \nu_n \!:\! \sigma_n \}, \nu'_? \{ \nu'_1\mapsto s_1, \ldots, \nu'_m\mapsto s_m \}) \reduce (\nu = \nu'_?) ~\wedge \\
   \quad (~ ((\nu_1 = \nu'_1) \wedge \ident{hasType}(\sigma_1, s_1)) \vee\ldots\vee ((\nu_1 = \nu'_m) \wedge \ident{hasType}(\sigma_1, s_m)) \vee \ldots \vee\\
@@
 
-\subsection{Common supertype relation}
-\label{sec:inference-commonsuper}
-
-Given two structural types, the \emph{common supertype} relation finds a least upper bound
-(Theorem~\ref{thm:lub}). When possible, it avoids variant types and prefers records 
-(Corollary~\ref{thm:no-unions}), which is important for usability as discussed earlier
-(\S\ref{sec:providers-json}).
-
-\begin{definition}
-A \emph{common supertype} of types $\sigma_1$ and $\sigma_2$ is a type $\sigma$, written 
-$\sigma_1 \triangledown \sigma_2 \vdash \sigma$, obtained according to the inference rules in 
-Figure~\ref{fig:subtyping-cst}.
-\end{definition}
-
-\noindent
-When finding a common supertype of records (\emph{record-1}), we return a record that has the 
-union of their fields. The types of shared fields become common supertypes of their respective 
-types while fields that are present in only one record are marked as optional. The (\emph{order-1})
-rule lets us reorder fields.
-
-Although all variants are mutual subtypes, we find one that best represents the sample. We avoid
-nested variants and limit the number of cases. This is done by grouping types 
-that have a common supertype which is not a variant. For example, rather than inferring 
-$\kvd{any}\langl\ident{int}, \kvd{any}\langl\ident{bool}, \ident{decimal}\rangl\rangl$, 
-our algorithm finds the common supertype of $\ident{int}$ and $\ident{decimal}$ and produces 
-$\kvd{any}\langl\ident{decimal},\ident{bool}\rangl$. To identify types that have a common supertype 
-which is not a variant, we group the types by a tag. The tag of a type is obtained using a function 
-\tytagof{(-)}. The function is not defined on $\bot$ and $\kvd{null}$ as those are hanled by the 
-(\emph{bot}) and (\emph{null-1}) or (\emph{null-2}).
-
-The handling of variants is specified using three rules. When combining a non-variant and a
-variant (\emph{var-1}), the variant may or may not already contain a case with the tag of the other
-type. If it does, the two types are combined, otherwise a new case is added.
-When combining two variants (\emph{var-2}), we group the cases that have shared tags.
-Finally, (\emph{var-3}) covers the case when we are combining two distinct non-variant types.
-As variants implicitly permit \kvd{null} values, we use an auxiliary function $\dropopt{-}$ to
-make nullable types non-nullable (when possible) to simplify the type.
-
-The remaining rules are straightforward. For collections and options, we find the common supertype 
-of the contained value(s); for compatible primitive types, we choose their supertype (\emph{prim})
-and a common supertype with \kvd{null} is either the type itself (\emph{null-1}) or an option 
-(\emph{null-2}).
-
-\paragraph{Properties.}
-The partially ordered set of types does not have a \emph{unique} least upper bound, but it does 
-have a least upper bound with respect to an equivalence between mutual subtypes. The common 
-supertype relation is well-defined (Theorem~\ref{thm:func}) and finds the least upper bound 
-(Theorem~\ref{thm:lub}).
-
-\begin{definition}
-Types $\sigma_1$, $\sigma_2$ are \emph{equivalent}, written $\sigma_1 \equiv \sigma_2$ iff
-they are mutual subtypes, i.e. $\sigma_1 <: \sigma_2 \wedge \sigma_1 :> \sigma_2$.
-An equivalence class of a type $\sigma$ is $\mathcal{E}(\sigma) = \{ \sigma' \;|\; \sigma\equiv\sigma' \}$.
-\end{definition}
-
-\noindent
-The equivalence class $\mathcal{E}$ sheds light on the structure of structural types. It defines
-a lattice with bottom $\{ \bot \}$ and top (set of all variant types). It also joins records with
-reordered fields (of same names and types) and types that contain other equivalent types (e.g.~a
-lists containing variant values).
-
-\begin{theorem}[Function]
-\label{thm:func}
-\raggedright
-For all $\sigma_1$ and $\sigma_2$ there exists exactly one $\sigma$ such that 
-$\sigma_1 \triangledown \sigma_2 \vdash \sigma$.
-Furthermore, if $\sigma_1 \triangledown \sigma_2 \vdash \sigma'$
-then it holds that $\sigma :> \sigma'$ and $\sigma' :> \sigma$.
-\end{theorem}
-\begin{proof}
-The pre-conditions of rules in Figure~\ref{fig:subtyping-cst} are disjoint, with the exception 
-of (\emph{order}), (\emph{sym}) and (\emph{refl}). These rules produce types that are subtypes
-of each other and this property is preserved by rules that use $\triangledown$ recursively.
-\end{proof}
-
-\begin{theorem}[Least upper bound]
-\label{thm:lub}
-\raggedright
-If $\sigma_1 \triangledown \sigma_2 \vdash \sigma$ then $\sigma$ is a least upper bound, i.e. 
-$\sigma :> \sigma_1$ and $\sigma :> \sigma_2$ and for all $\sigma'$ such that $\sigma' :> \sigma_1$
-and $\sigma' :> \sigma_2$, it holds that $\sigma' :> \sigma$.
-\end{theorem}
-\begin{proof}
-By induction over $\vdash$. Note that the algorithm never produces nested variant or nested
-option. When one type is variant, the only common supertype is also a variant (\emph{var-2}); 
-for (\emph{var-3}) no other common supertype exists because the two types have distinct 
-tags and are not options.
-\end{proof}
-
-\begin{corollary}
-\label{thm:no-unions}
-Given $\sigma_1, \sigma_2, \sigma$ such that $\sigma :> \sigma_1$ and $\sigma :> \sigma_2$ and 
-$\sigma$ is not a variant, then $\sigma_1 \tsep \sigma_2 \vdash \sigma'$ 
-and $\sigma' \equiv \sigma$.
-\end{corollary}
-\begin{proof}
-Consequence of Theorem~\ref{thm:lub} and the fact that the set of all variants is the
-top element of $\mathcal{E}$.
-\end{proof}
-
-% ==================================================================================================
-% 
-%    #######
-%    #        ####  #####  #    #   ##   #      # ######   ##   ##### #  ####  #    #
-%    #       #    # #    # ##  ##  #  #  #      #     #   #  #    #   # #    # ##   #
-%    #####   #    # #    # # ## # #    # #      #    #   #    #   #   # #    # # #  #
-%    #       #    # #####  #    # ###### #      #   #    ######   #   # #    # #  # #
-%    #       #    # #   #  #    # #    # #      #  #     #    #   #   # #    # #   ##
-%    #        ####  #    # #    # #    # ###### # ###### #    #   #   #  ####  #    #
-%
-% ==================================================================================================
-
-\section{Formalising type providers}
-\label{sec:formal}
-
-In this section, we build the theoretical framework for proving relativised type safety of 
-structural type providers. We discuss the runtime representation of documents and 
-primitive operations (\S\ref{sec:formal-convert}), we embed these into a formal model of an F\# 
-subset (\S\ref{sec:formal-ff}) and we describe how structural type providers turn
-inferred structural types into F\# types (\S\ref{sec:formal-tp}).
-
-% -------------------------------------------------------------------------------------------------
-
 \subsection{Structural values and conversions}
 \label{sec:formal-convert}
-
-We represent JSON, XML and CSV documents using the same first-order \emph{structural values} defined
-as follows:
-%
-\begin{equation*}
-\begin{array}{lcl}
- s &\narrow{=}& i \lsep d \lsep f \lsep t \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
-   &\narrow{|}& [s_1; \ldots; s_n] \lsep \nu_?\{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \}
-\end{array}
-\end{equation*}
-%
-The first few cases represent primitive values ($i$ for integers, $d$ for decimals, $f$ for floating
-point numbers and $t$ for strings) and the \kvd{null} value. A collection is written as a 
-list of values in square brackets. A record starts with an optional name $\nu_?$, followed by a 
-sequence of field assignments $\nu_i \mapsto s_i$.
 
@@
 \begin{array}{l}
- \sem{\,\nu_? \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}\,}_e = \\[0.1em]
+ \sem{\,\nu \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}\,}_e = \\[0.1em]
  \quad C, \kvd{new}~C(e), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
@@
  \qquad M_i = \kvd{member}~\nu_i:\tau_i=e_i\\[0.1em]
- \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_{e'},\;\; e'=\ident{getField}(\nu_?, \nu_i, e)\\[0.6em]
+ \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_{e'},\;\; e'=\ident{getField}(\nu, \nu_i, e)\\[0.6em]
 \end{array}
@@
 
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Inferring types from values}
-\label{sec:formal-inferval}
-
-The common supertype algorithm (\S\ref{sec:inference}) is at the core of the type inference 
-for structural values, but we have not yet clarified how it is used. Given a JSON, XML or CSV 
-document, F\# Data constructs a structural value $s$ from the sample. The following defines a mapping 
-$\semalt{s_1,\ldots,s_n}$ which turns a collection of sample values $s_1, \ldots, s_n$ into a 
-structural type $\sigma$:
-%
-\begin{equation*}
-\begin{array}{rclcrclcrcl}
- \semalt{i} &\narrow{=}& \ident{int} && \semalt{d} &\narrow{=}& \ident{decimal} && \semalt{b} &\narrow{=}& \ident{bool}\\
- \semalt{f} &\narrow{=}& \ident{float} && \semalt{t} &\narrow{=}& \ident{string} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null}\\
-\end{array}
-\end{equation*}
-\noindent
-\vspace{-0.5em}
-\begin{equation*}
-\begin{array}{l}
- \semalt{s_1, \ldots, s_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
- \qquad\sigma_0 = \bot,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{s_i} \vdash \sigma_i
- \\[0.5em]
- \semalt{[s_1; \ldots; s_n]} = [\semalt{s_1, \ldots, s_n}]
- \\[0.5em]
- \semalt{\nu?~\{ \nu_1 \mapsto s_1, \ldots, \nu_n \mapsto s_n \}} =\\[0.1em]
- \qquad\nu?~\{ \nu_1:\semalt{s_1}, \ldots, \nu_n :\semalt{s_n} \}
-\end{array}
-\end{equation*}
-%
-We overload the notation and write $\semalt{s}$ when inferring type from a single value. Primitive 
-values are mapped to their corresponding types. For records, we return a type with field types 
-inferred based on individual values.
-
-When infering a value from multiple samples, we use the common supertype relation to find a common 
-type for all values (starting with $\bot$). This operation is used at the top-level (when calling 
-a type provider with multiple samples) and also when inferring the type of a collection.
-
 % ==================================================================================================
@@
 \label{thm:tp-correctness}
-Given a sample value $s$ and an input value $s'$ such that $\semalt{s} :> \semalt{s'}$
+Given a sample value $s$ and an input value $s'$ such that $\semalt{s} \sqsupseteq \semalt{s'}$
 and provided type, expression and class declarations $\tau, e, L = \sem{\semalt{s}}_{s'}$, 
@@
 By induction over the structure of $\sem{-}_-$. For primitives, the conversion functions accept all subtypes.
-For other cases, analyse the provided code to see that it can work on all subtypes (\emph{e.g.}~\ident{getChildren}
+For other cases, analyze the provided code to see that it can work on all subtypes (\emph{e.g.}~\ident{getChildren}
 works on \kvd{null} values, \ident{getField} returns \kvd{null} when a field is missing and, for variants,
@@
 
-For all new inputs $s'$ such that $\exists i.(\semalt{s_i} :> \semalt{s'})$, let $e_s=e[x\leftarrow s']$
+For all new inputs $s'$ such that $\exists i.(\semalt{s_i} \sqsupseteq \semalt{s'})$, let $e_s=e[x\leftarrow s']$
 be an expression (of type $\tau$) that wraps the input in a provided type. Then, for any expression $e_c$

### 2015-11-20 pldi-2016/paper.tex
Final edits in section 4 (excluding examples)

@@
 \newcommand{\tytagof}{\ident{tagof}}
-\newcommand{\nameoftag}{\ident{nameof}}
 
@@
 \vspace{-0.5em}
-\caption{Subtype relation between structural types}
+\caption{Important aspects of the preferred shape relation}
 \label{fig:subtyping-diagram}
@@
 
-In this section, we build the theoretical framework for proving relativised type safety of 
-structural type providers. We discuss the runtime representation of documents and 
-primitive operations (\S\ref{sec:formal-convert}), we embed these into a formal model of an F\# 
-subset (\S\ref{sec:formal-ff}) and we describe how structural type providers turn
-inferred structural types into F\# types (\S\ref{sec:formal-tp}).
+This section presents the theoretical model of strucutral type providers. We introduce the Foo calculus 
+which is an object-oriented subset of F\# based on SML \cite{sml} and Featherweight Java \cite{fwjava}. It 
+includes operations for working with data that model the F\# Data runtime. Finally, we describe how 
+type providers turn inferred shapes into Foo classes (\S\ref{sec:formal-tp}).
+
+Type providers map the dirty world of data into a nice world of F\#. To model this, the Foo calculus
+does not have \kvd{null} values and data values $d$ are never directly exposed.
 
@@
 \begin{equation*}
-\begin{array}{l}
-\ident{asFloat}(i) \,\reduce f\quad(f = i)\\
-\ident{asFloat}(d) \reduce f\quad(f = d)\\
-\ident{asFloat}(f) \reduce f
+\hspace{1.1em}
+\begin{array}{l}
+\ident{hasShape}(\nu~ \{ \nu_1 \!:\! \sigma_1, \ldots, \nu_n \!:\! \sigma_n \}, \nu'~\{ \nu'_1\mapsto d_1, \ldots, \nu'_m\mapsto d_m \}) \reduce (\nu = \nu') ~\wedge \\
+  \quad (~ ((\nu_1 = \nu'_1) \wedge \ident{hasShape}(\sigma_1, d_1)) \vee\ldots\vee ((\nu_1 = \nu'_m) \wedge \ident{hasShape}(\sigma_1, d_m)) \vee \ldots \vee\\
+  \quad ~\; ((\nu_n = \nu'_1) \wedge \ident{hasShape}(\sigma_n, d_1)) \vee\ldots\vee ((\nu_n = \nu'_m) \wedge \ident{hasShape}(\sigma_n, d_m))~)
 \\[0.5em]
-\ident{asDec}(i) \,\reduce d\quad(d = i)\\
-\ident{asDec}(d) \reduce d
-\end{array}
-\qquad
-\begin{array}{l}
-\ident{hasType}(\nu \{ \nu_1 \!:\! \sigma_1, \ldots, \nu_n \!:\! \sigma_n \}, \nu'_? \{ \nu'_1\mapsto s_1, \ldots, \nu'_m\mapsto s_m \}) \reduce (\nu = \nu'_?) ~\wedge \\
-  \quad (~ ((\nu_1 = \nu'_1) \wedge \ident{hasType}(\sigma_1, s_1)) \vee\ldots\vee ((\nu_1 = \nu'_m) \wedge \ident{hasType}(\sigma_1, s_m)) \vee \ldots \vee\\
-  \quad ~\; ((\nu_n = \nu'_1) \wedge \ident{hasType}(\sigma_n, s_1)) \vee\ldots\vee ((\nu_n = \nu'_m) \wedge \ident{hasType}(\sigma_n, s_m))~)
-\\[0.5em]
-\ident{hasType}([\sigma], [s_1; \ldots; s_n]) \reduce \ident{hasType}(\sigma, s_1)\wedge\ldots\wedge\ident{hasType}(\sigma, s_n) \\
-\ident{hasType}([\sigma], \kvd{null}) \reduce \kvd{true} \\  
-\end{array}  
-\end{equation*}
-%
-\vspace{-0.5em}
-%
-\begin{equation*}
-\begin{array}{l}
-\ident{isNull}(\kvd{null}) \reduce \kvd{true} \\
-\ident{isNull}(\_) \reduce \kvd{false} 
-\\[0.5em]
-\ident{getField}(\nu,\nu_i, \nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
-\ident{getField}(\bullet, \nu_i, \{\ldots, \nu_i=s_i, \ldots\}) \reduce s_i\\
-\ident{getField}(\nu,\nu', \nu~\{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )\\
-\ident{getField}(\bullet, \nu', \{\ldots, \nu_i=s_i, \ldots\}) \reduce \kvd{null}\quad(\nexists i.\nu_i=\nu' )
-\\[0.5em]
-\ident{getChildren}([s_1; \ldots; s_n]) \reduce [s_1; \ldots; s_n]  \\  
-\ident{getChildren}(\kvd{null}) \reduce [~] 
-\end{array}
-~
-\begin{array}{l}
-\ident{hasType}(\ident{string}, t) \reduce \kvd{true} \\
-\ident{hasType}(\ident{bool}, s) \reduce \kvd{true} \quad(\textnormal{when}~s\in{\kvd{true},\kvd{false}} )\\
-\ident{hasType}(\ident{float}, s) \reduce \kvd{true} \quad(\textnormal{when}~s=i, s=d ~\textnormal{or}~ s=f) \\
-\ident{hasType}(\ident{decimal}, s) \reduce \kvd{true} \quad(\textnormal{when}~s=i ~\textnormal{or}~ s=f) \\
-\ident{hasType}(\ident{int}, i) \reduce \kvd{true}\\
-\ident{hasType}(\kvd{option}\langl\sigma\rangl, s) \reduce \ident{hasType}(\sigma,s) \\
-\ident{hasType}(\kvd{option}\langl\sigma\rangl, s) \reduce \kvd{true} \\
-\ident{hasType}(\kvd{any}\langl\ldots\rangl, s) \reduce \kvd{true} \\
-\ident{hasType}(\_, \_) \reduce \kvd{false} \\
+\ident{hasShape}([\sigma], [d_1; \ldots; d_n]) \reduce \ident{hasShape}(\sigma, d_1)\wedge\ldots\wedge\ident{hasShape}(\sigma, d_n) \\
+\ident{hasShape}([\sigma], \kvd{null}) \reduce \kvd{true} 
+\end{array}  
+\quad
+\begin{array}{l}
+\ident{convFloat}(\ident{float}, i) \reduce f~(f=i) \\
+\ident{convFloat}(\ident{float}, f) \reduce f \\[0.75em]
+\ident{convNull}(\kvd{null}, e) \reduce \ident{None} \\
+\ident{convNull}(d, e) \reduce \ident{Some}(e~d) 
+\end{array}
+\end{equation*}
+%
+\vspace{-0.75em}
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{hasShape}(\ident{string}, s) \reduce \kvd{true} \\
+\ident{hasShape}(\ident{int}, i) \reduce \kvd{true}\\
+\ident{hasShape}(\ident{bool}, d) \reduce \kvd{true} \quad(\textnormal{when}~d\in{\kvd{true},\kvd{false}} )\\
+\ident{hasShape}(\ident{float}, d) \reduce \kvd{true} \quad(\textnormal{when}~d=i ~\textnormal{or}~ d=f) \\
+\ident{hasShape}(\_, \_) \reduce \kvd{false} \\
+\end{array}
+\;\;
+\begin{array}{l}
+\ident{convPrim}(\sigma, d) \reduce d\quad(\sigma,d\in\{ (\ident{int},i),(\ident{string},s),(\ident{bool}, b) \})
+\\[0.25em]
+\ident{convField}(\nu,\nu_i, \nu~\{\ldots, \nu_i=d_i, \ldots\}, e) \reduce e~d_i\\
+\ident{convField}(\nu,\nu', \nu~\{\ldots, \nu_i=d_i, \ldots\}, e) \reduce e~\kvd{null}\quad(\nexists i.\nu_i=\nu' )
+\\[0.25em]
+\ident{convElements}([d_1; \ldots; d_n], e) \reduce e~d_1 :: \ldots :: e~d_n :: \kvd{nil}  \\  
+\ident{convElements}(\kvd{null}) \reduce \kvd{nil}
 \end{array}
@@
 
+% --------------------------------------------------------------------------------------------------
+
+\subsection{The Foo calculus}
+\label{sec:formal-ff}
+
+The syntax of the calculus is shown in Figure~\ref{fig:foo-syntax}. We only need classes with 
+parameter-less members and without inheritance. A class has a single implicit constructor and the 
+declaration closes over constructor parameters.
+
+The type \ident{Data} is the type of all structural data $d$. A class definition $L$ consists of 
+a constructor and zero or more members. Values $v$ include previously defined data $d$; expressions
+$e$ include class construction, member access and the usual functional constructs (functions,
+lists, options) and conditionals. The $op$ constructs are discussed next.
+
 % -------------------------------------------------------------------------------------------------
 
-\subsection{Structural values and conversions}
-\label{sec:formal-convert}
-
-As indicated by the subtyping, our system permits certain runtime conversions (e.g.~an integer
-\num{1} can be treated as a floating-point \num{1.0}). The following primitive operations 
-implement the conversions and other helpers:
-%
-\begin{equation*}
-\begin{array}{lcl}
- op  &\narrow{=}& \ident{asDec}(s) \lsep \ident{asFloat}(s) \lsep \ident{getChildren}(s) \\
-     &\narrow{|}& \ident{getField}(\nu?, \nu, s) \lsep \ident{isNull}(s) \lsep \ident{hasType}(\sigma, s)
-\end{array}
-\end{equation*}
-%
-The operations are used in code generated by type providers. Their behavior is defined
-by the rules in Figure~\ref{fig:op-conversions}. The conversion functions (\ident{asDec}
-and \ident{asFloat}) only perform conversions required by the subtyping (our actual 
-implementation is more lax and performs additional conversions).
-
-The \ident{getField} operation is used to access a field of a record. The operation ensures that
-the actual record name matches the expected name (we write $\bullet$ for the name of an unnamed 
-record). The \ident{getChildren} operation returns elements of a list and turns \kvd{null} into 
-an empty collction.
-
-Finally, we also define two helper functions. The \ident{isNull} operation is used to check whether 
-value is \kvd{null} and \ident{hasType} is a runtime type test. The most complex case is handling 
-of records where we check that for each field $\nu_1, \ldots, \nu_n$ in the type, the actual record 
-value has a field of the same name with a matching type. The last line defines a ``catch all'' 
-pattern, which returns \kvd{false} for all remaining cases. Although not elegant, the rules 
-can be expressed using just Featherweight F\#, discussed in the next section.
+\paragraph{Dynamic data operations.}
+
+The Foo programs can only work with raw data using certain primitive operations. Those are modelled
+by the $op$ primitives. In F\# Data, those are internal and users never access them directly.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure}[!h]
+\noindent
+\begin{equation*}
+\begin{array}{rcl}
+ \tau &\narrow{=}& \ident{int} \lsep \ident{float} \lsep \ident{bool} \lsep \ident{string} \lsep C \lsep \ident{Data} \\[0.0em]
+      &\narrow{|}& \tau_1 \rightarrow \tau_2 \lsep \ident{list}\langl\tau\rangl \lsep \ident{option}\langl\tau\rangl 
+\\[0.6em]
+ L &\narrow{=}& \kvd{type}~C(\overline{x:\tau}) = \overline{M} \\[0.0em]
+ M &\narrow{=}& \kvd{member}~N:\tau=
+\\[0.6em]
+ v &\narrow{=}& d \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \lsep v_1::v_2 \\[0.0em]
+ e &\narrow{=}& d \lsep op \lsep e_1~e_2 \lsep \lambda x.e \lsep e.N \lsep \kvd{new}~C(\overline{e}) \\
+   &\narrow{|}& \ident{None} \lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
+   &\narrow{|}& \ident{Some}(e) \lsep e_1 = e_2 \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3} \lsep \kvd{nil} \\
+   &\narrow{|}& e_1 :: e_2 \lsep \kvd{match}~e~\kvd{with}~x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2 
+\\[0.6em]
+op &\narrow{=}& \ident{convFloat}(\sigma, e) \lsep \ident{convPrim}(\sigma, e) \\
+   &\narrow{|}& \ident{convField}(\nu_1, \nu_2, e, e) \lsep \ident{convNull}(e_1, e_2) \\
+   &\narrow{|}& \ident{convElements}(e_1, e_2) \,\lsep \ident{hasShape}(\sigma, e) 
+\end{array}
+\end{equation*}
+
+\caption{The syntax of the Foo calculus}
+\label{fig:foo-syntax}
+\vspace{-0.5em}
+\end{figure}
+
+% -------------------------------------------------------------------------------------------------
+
+The behaviour of the dynamic data operations is defined by the reduction rules in 
+Figure~\ref{fig:op-conversions}. The operations can convert data values into values of 
+less preferred shape. For example, \ident{convFloat} turns an integer \num{1} into 
+a floating-point \num{1.0} and \ident{convElements} turns \kvd{null} value into an empty list.
+
+All of the dynamic data operations take a data value $d$ and produce a Foo value. We do not 
+make the conversion explicit for primitive values (\ident{convPrim}) and assume that primitive
+values are shared, but we convert all other values. For example \ident{convElements} turns a 
+data collection $[d_1, \ldots, d_n]$ into a Foo list $v_1 :: .. :: v_n :: \kvd{nil}$.
+
+The data operations for collections, record fields and nullable values all take a function
+as the last argument and invoke it to convert the obtained data. For example, \ident{convField} 
+accesses a record field -- it ensures that the actual record name matches the expected name and 
+passes the field value (or \kvd{null} data value) to the provided function.
+
+Finally, we also define a runtime shape test \ident{hasShape}.
+The most complex case is handling of records where we check that for each field $\nu_1, \ldots, \nu_n$ 
+in the record, the actual record value has a field of the same name with a matching shape. The last 
+line defines a ``catch all'' pattern, which returns \kvd{false} for all remaining cases. Although
+not elegant, if we treat $e_1 \vee e_2$ and $e_1 \wedge e_2$ as syntactic sugar for \kvd{if} then
+the rule is expressed using just Foo expressions.
+
+% -------------------------------------------------------------------------------------------------
+
+\paragraph{Reduction.}
+The reduction relation is of the form $L, e \reduce e'$. We often omit class declarations 
+$L$ and we write $L, e \reduce^{*} e'$ to denote 
+the reflexive and transitive closure of $\reduce$.
+
+Figure~\ref{fig:ff-reduction} shows the reduction rules. The (\emph{ctx}) rule models the eager
+evaluation of F\# and performs a reduction inside a sub-expression specified by an evaluation 
+context $E$:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ E &\narrow{=}& v::E \lsep v~E \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e})\\[0.1em]
+   &\narrow{|}&  \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2  \lsep E = e \lsep v = E \\[0.1em]
+   &\narrow{|}& \ident{Some}(E) \lsep op(\overline{v}, E, \overline{e})\\[0.1em]
+   &\narrow{|}& \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
+   &\narrow{|}& \kvd{match}~E~\kvd{with}~x_1 :: x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
+\end{array} 
+\end{equation*}
+
+\noindent
+The evaluation proceeds from left to right as denoted by $\overline{v}, E, \overline{e}$ in 
+constructor and dynamic data operation arguments or $v::E$ in list initialization.
+
+We write $e[\overline{x} \leftarrow \overline{v}]$ for the result of replacing variables $\overline{x}$ by
+values $\overline{v}$ in an expression. The (\emph{member}) rule reduces a member access using a class 
+definition in the assumption to obtain the body of a member. The remaining six rules
+give standard reductions for a functional language.
+
+% -------------------------------------------------------------------------------------------------
+
+\paragraph{Type checking.} 
+Well-typed Foo programs reduce to a value in a finite number of steps or get stuck due to an
+error condition. The stuck states can only be due to the dynamic data operations (e.g.
+$\ident{convFloat}(\ident{float},\kvd{null})$). The relativized type safety (Theorem~\ref{thm:safety})
+characterizes the additional conditions on input data under which Foo programs do not ``go wrong''.
+
+Typing rules in Figure~\ref{fig:ff-typecheck} are written using a judgement
+$L; \Gamma \vdash e : \tau$ where the context also contains a set of class declarations $L$.
+The fragment demonstrates the differences and similarities with SML and Featherweight Java:
+%
+\begin{itemize}
+\item[--] All data values $d$ have the type \ident{Data}, but primitive data values (Booleans, 
+  strings, integers and floats) can be implicitly converted to Foo values and so they also have a 
+  primitive type as illustrated by the rule for $i$ and $f$.
+\item[--] For non-primitive data values (including \kvd{null}, data collections and records),
+  \ident{Data} is the only type.
+\item[--] Operations $op$ (omitted) have a family of types (Doo does not have polymorphic types),
+  accepting \ident{Data} as one of the arguments and producing a non-data Foo type.
+\item[--] Rules for checking class construction and member access are similar to corresponding
+  rules of Featherweight Java.  
+\end{itemize}
+%
+An important part of Featherweight Java that is omitted here is the checking of type declarations
+(ensuring the members are well-typed). We consider only classes generated by our type providers 
+and those are well-typed by construction.
 
@@
  \textnormal{\footnotesize{(member)}}&
- \hspace{-0.4em}
+ \hspace{-0.6em}
  \inference
- { \kvd{type}~C(\overline{x:\tau})=\ldots \kvd{member}~N_i : \tau_i = e_i \ldots }
- { (\kvd{new}~C(\overline{v})).N_i \reduce e_i[\overline{x} \leftarrow \overline{v}] }\\
- \\
- \textnormal{\footnotesize{(eq1)}}&
- v=v\reduce\kvd{true} \qquad \textnormal{\footnotesize{(eq2)}}~~v=v'\reduce\kvd{false}\\
- \\
+ { \kvd{type}~C(\overline{x:\tau})= \kvd{member}~N_i : \tau_i = e_i \ldots \in L }
+ { L, (\kvd{new}~C(\overline{v})).N_i \reduce e_i[\overline{x} \leftarrow \overline{v}] }\\
+ \\[-0.2em]
+ \textnormal{\footnotesize{(match1)}}&
+ \hspace{-1em}
+ \begin{array}{l}
+  \kvd{match}~\ident{None}~\kvd{with} \\
+  \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
+ \end{array} \hspace{-0.5em} ~\reduce~ e_2 \\
+ \\[-0.2em]
+ \textnormal{\footnotesize{(match2)}}&
+ \hspace{-1em}
+ \begin{array}{l}
+    \kvd{match}~\ident{Some}(v)~\kvd{with} \\
+    \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
+ \end{array} \hspace{-0.5em} ~\reduce~ e_1[x\leftarrow v]\\
+ \\[-0.2em]
+ \textnormal{\footnotesize{(match3)}}&
+ \hspace{-1em}
+ \begin{array}{l}
+  \kvd{match}~\kvd{nil}~\kvd{with} \\[0em]
+  x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
+ \end{array} \hspace{-0.5em} ~\reduce~ e_2\\
+ \\[-0.2em]
+ \textnormal{\footnotesize{(match4)}}&
+ \hspace{-1em}
+ \begin{array}{l}
+  \kvd{match}~v_1::v_2~\kvd{with} \\[0em]
+  x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
+ \end{array} \hspace{-0.5em} ~\reduce~ e_1[\overline{x}\leftarrow\overline{v}] \\
+ \\[-0.2em]
  \textnormal{\footnotesize{(cond1)}}&
@@
  \kvd{if}~\kvd{true}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_1 \\
- \\
+ \\[-0.2em]
  \textnormal{\footnotesize{(cond2)}}&
@@
  \kvd{if}~\kvd{false}~\kvd{then}~e_1~\kvd{else}~e_2 ~\reduce~ e_2 \\
- \\
- \textnormal{\footnotesize{(match1)}}&
- \hspace{-1em}
- \begin{array}{l}
-  \kvd{match}~\ident{None}~\kvd{with} \\
-  \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
- \end{array} \hspace{-0.5em} ~\reduce~ e_2 \\
- \\
- \textnormal{\footnotesize{(match2)}}&
- \hspace{-1em}
- \begin{array}{l}
-    \kvd{match}~\ident{Some}(v)~\kvd{with} \\
-    \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
- \end{array} \hspace{-0.5em} ~\reduce~ e_1[x\leftarrow v]\\
- \\
- \textnormal{\footnotesize{(match3)}}&
- \hspace{-1em}
- \begin{array}{l}
-  \kvd{match}~[v_1;\ldots;v_n]~\kvd{with} \\[0em]
-  [x_1;\ldots;x_m ] \rightarrow e_1 \,|\, \_ \rightarrow e_2
- \end{array} \hspace{-0.5em} ~\reduce~ e_2\quad(m\neq n) \\
- \\
- \textnormal{\footnotesize{(match4)}}&
- \hspace{-1em}
- \begin{array}{l}
-  \kvd{match}~[v_1;\ldots;v_n]~\kvd{with} \\[0em]
-  [x_1;\ldots;x_n ] \rightarrow e_1 \,|\, \_ \rightarrow e_2
- \end{array} \hspace{-0.5em} ~\reduce~ e_1[\overline{x}\leftarrow\overline{v}] \\
- \\
- \textnormal{\footnotesize{(map)}}&
+ \\[-0.3em]
+ \textnormal{\footnotesize{(eq1)}}&
+ v=v\reduce\kvd{true} \qquad \textnormal{\footnotesize{(eq2)}}~~v=v'\reduce\kvd{false}\\
+ \\[-0.3em]
+ \textnormal{\footnotesize{(fun)}}&
  \hspace{-0.4em}
- \ident{map}~(\lambda x.e)~[v_1; \ldots] ~\reduce~ [e[x\leftarrow v_1]; \ldots] \\
- \\
+ (\lambda x.e)~v ~\reduce~ e[x\leftarrow v] \\
+ \\[-0.3em]
  \textnormal{\footnotesize{(ctx)}}&
@@
 
-\subsection{Featherweight F\#}
-\label{sec:formal-ff}
-
-We now extend the semantics discussed so far and add a minimal subset of F\# needed to model F\# 
-Data. Type providers use classes and so we focus on the F\# object model and combine aspects of SML 
-\cite{sml} and Featherweight Java \cite{fwjava}.
-
-We only need classes with parameter-less members and without inheritance. A class has a single 
-implicit constructor and the declaration closes over constructor parameters. To avoid including all 
-of ML, we only pick constructs for working with options and lists that we need later.
-%
-\begin{equation*}
-\begin{array}{rcl}
- \tau &\narrow{=}& \ident{int} \lsep \ident{decimal} \lsep \ident{float} \lsep \ident{bool} \lsep \ident{string} \\[0.0em]
-      &\narrow{|}& C \lsep \ident{StructVal} \lsep \ident{list}\langl\tau\rangl \lsep \ident{option}\langl\tau\rangl \\[0.6em]
- L &\narrow{=}& \kvd{type}~C(\overline{x:\tau}) = \overline{M} \\[0.0em]
- M &\narrow{=}& \kvd{member}~N:\tau=e \\[0.6em]
- v &\narrow{=}& s \lsep \ident{None} \lsep \ident{Some}(v) \lsep \kvd{new}~C(\overline{v}) \lsep [v_1; \ldots; v_n] \\[0.0em]
- e &\narrow{=}& s \lsep op \lsep e.N \lsep \kvd{new}~C(\overline{e}) \lsep {\kvd{if}~e_1~\kvd{then}~e_2~\kvd{else}~e_3}\\
-   &\narrow{|}& \ident{None} \lsep\kvd{match}~e~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\
-   &\narrow{|}& \ident{Some}(e) \lsep [e_1;\ldots;e_n]\lsep \ident{map}~(\lambda x\rightarrow e_1)~e_2\\
-   &\narrow{|}& e = e \lsep \kvd{match}~e~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2
-\end{array}
-\end{equation*}
-
-\noindent
-The type \ident{StructVal} is a type of all structural values $s$. A class
-definition $L$ consists of a constructor and zero or more members. Values $v$ include 
-previously defined structural values $s$ and values for the option and list type; finally 
-expressions $e$ include previously defined operations $op$, class construction, member access, 
-conditionals and expressions for working with option values and lists. We include 
-\ident{map} as a special construct to avoid making the language too complex.
-
-\paragraph{Reduction.}
-The reduction relation is of the form $e \reduce e'$. We also write 
-$e \reduce^{*} e'$ to denote the reflexive and transitive closure of $\reduce$. The reduction rules
-for operations $op$ were discussed earlier. Figure~\ref{fig:ff-reduction} shows the remaining
-rules.
-
-The (\emph{ctx}) rule performs a reduction inside a sub-expression specified by an evaluation context.
-This models the eager evaluation of F\#. An evaluation context $E$ is defined as:
-%
-\begin{equation*}
-\begin{array}{rcl}
- E &\narrow{=}& [\overline{v};E;\overline{e}] \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e})\\[0.1em]
-   &\narrow{|}& \ident{map}~(\lambda x\rightarrow e)~E \lsep \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2 \\[0.1em]
-   &\narrow{|}& \ident{Some}(E) \lsep op(E) \lsep E = e \lsep v = E \\[0.1em]
-   &\narrow{|}& \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
-   &\narrow{|}& \kvd{match}~E~\kvd{with}~[x_1; \ldots; x_n] \rightarrow e_1 \,|\, \_ \rightarrow e_2
-\end{array} 
-\end{equation*}
-
-\noindent
-The evaluation first reduces arguments of functions and the evaluation proceeds from left to right 
-as denoted by $\overline{v}, E, \overline{e}$ in constructor arguments or $\overline{v};E;\overline{e}$
-in list initialization.
-
-We write $e[\overline{x} \leftarrow \overline{v}]$ for the result of replacing variables $\overline{x}$ by
-values $\overline{v}$ in an expression. The (\emph{member}) rule reduces a member access using a class 
-definition in the assumption to obtain the body of a member. The remaining six rules
-give standard reductions for conditionals and pattern matching.
-
-We omit expressions $e \vee e$ and $e \wedge e$ that are used in Figure~\ref{fig:op-conversions}, 
-as those can be easily defined as syntactic sugar using \kvd{if}. All expressions reduce to a value in a 
-finite number of steps or get stuck due to an error condition. An error condition can be a wrong 
-argument passed to conditional, pattern matching or one of the primitives from Figure~\ref{fig:op-conversions}.
+\subsection{Type providers}
+\label{sec:formal-tp}
+
+So far, we defined the type inference algorithm which produces a shape $\sigma$ from one 
+or more sample documents (\S\ref{sec:inference}) and we defined a simplified model of evaluation
+of F\# (\S\ref{sec:formal-ff}) and F\# Data runtime (\S\ref{sec:formal-tp}). In this section, we 
+define how the type providers work, linking the two parts.
+
+All F\# Data type providers take (one or more) sample documents, infer a common preferred shape $\sigma$
+and then use it to generate F\# types that are exposed to the programmer.\footnote{The actual 
+implementation provides \emph{erased types} as described in \cite{fsharp-typeprov}. Here, we treat 
+the code as actually generated. This is an acceptable simplification, because F\# Data type providers 
+do not rely on laziness that is available through erased types.}
+
+\paragraph{Type provider mapping.}
+A type provider produces an F\# type $\tau$ together with a Foo expression and a collection of 
+class definitions. We express it using the following mapping:
+%
+\begin{equation*}
+\sem{-} : \sigma \rightarrow (\tau \times e \times L) \qquad (\textnormal{where}~L,\emptyset \vdash e:\ident{Data}\rightarrow\tau)
+\end{equation*}
+%
+The mapping $\sem{\sigma}$ takes an inferred shape $\sigma$. It returns an F\# type $\tau$ and
+a function that turns the input data (value of type \ident{Data}) into a Foo value of type $\tau$. 
+The type provider also generates class definitions that may be used by $e$.
+
+Figure~\ref{fig:tp-generation} defines $\sem{-}$. Primitive types are handled by a single rule that
+inserts an appropriate conversion function; \ident{convPrim} just checks that the shape matches
+and \ident{convFloat} converts numbers to a floating-point.
+
+For records, we generate a class $C$ that takes a data value as constructor parameter. For each 
+record field, we generate a member with the same name as the field.\footnote{The actual
+F\# Data implementation also capitalizes the names.} The body of the member calls \ident{convField}
+with a function obtained from $\sem{\sigma_i}$. This function turns the field value (data of shape
+$\sigma_i$) into a Foo value of type $\tau_i$. The returned expression creates a new instance of 
+$C$ and the mapping returns the class $C$ together with all recursively generated classes. Note that 
+the class name $C$ is not directly accessed by the user and so we can use arbitrary name, although the 
+actual implementation in F\# Data attempts to infer a reasonable name.\footnote{For example, in 
+\ident{\{\str{person}:\{\str{name}:\str{Tomas}\}\}}, the nested record will be named \ident{Person}
+based on the name of the parent record field.}
+
+A collection shape becomes an Foo $\ident{list}\langl\tau\rangl$. The returned expression calls \ident{convElements}
+(which returns empty list for data value \kvd{null}). The last parameter is the recursively obtained
+conversion function for the shape of elements $\sigma$. The handling of the nullable shape is similar,
+but uses \ident{convNull}.
+
+As discussed earlier, labelled top shape are also generated as Foo classes with properties. Given 
+$\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl$, we get corresponding F\# types $\tau_i$ and generate 
+$n$ members of type $\ident{option}\langl \tau_i\rangl$. When the member is accessed, we need to perform
+a runtime shape test using \ident{hasShape} to ensure that the value has the right shape (similarly to runtime 
+type conversions from the top type in languages like Java). If the shape matches, a \ident{Some} value is 
+returned. The shape inference algorithm also guarantees that there is only one case for each shape tag 
+(\S\ref{sec:inference-commonsuper}) and so we can use the tag for the name of the generated member.
 
@@
 \inference
-  {L; \Gamma \vdash e_1 : \ident{StructVal}}
-  {L; \Gamma \vdash [e_1; \ldots; e_n]:\ident{StructVal}}
+  {~}
+  {L; \Gamma \vdash d : \ident{Data}}
+~~
+\inference
+  {~}
+  {L; \Gamma \vdash i : \ident{int}}
+~~
+\inference
+  {~}
+  {L; \Gamma \vdash f : \ident{float}}
+\end{equation*}
+\vspace{0.25em}
+\begin{equation*}
+\inference
+  {L; \Gamma, x : \tau_1 \vdash e : \tau_2}
+  {L; \Gamma \vdash \lambda x.e : \tau_2}
 \quad
 \inference
-  {~}
-  {L; \Gamma \vdash n : \ident{int}}
-\end{equation*}
-\begin{equation*}
-\inference
-  {L; \Gamma \vdash e_1 : \tau}
-  {L; \Gamma \vdash [e_1; \ldots; e_n]:\ident{list}\langl\tau\rangl}
-\quad
-\inference
-  {~}
-  {L; \Gamma \vdash s : \ident{StructVal}}
+  {L; \Gamma \vdash e_2 : \tau_1 \\
+   L; \Gamma \vdash e_1 : \tau_1 \rightarrow \tau_2}
+  {L; \Gamma \vdash e_1~e_2 : \tau_2}
 \end{equation*}
@@
 
-\paragraph{Type checking.} 
-Typing rules in Figure~\ref{fig:ff-typecheck} are written using a judgement
-$L; \Gamma \vdash e : \tau$ where the context also contains a set of class declarations $L$.
-The fragment demonstrates the key differences from Standard ML and Featherweight Java:
-%
-\begin{itemize}
-\item[--] All structural values $s$ have the type \ident{StructVal}, but some have other types
-  (Booleans, strings, integers, etc.) as illustrated by the rule for $n$.
-  For other values, \ident{StructVal} is the only type -- this includes records and \kvd{null}.
-\item[--] A list containing other structural values $[s_1; \ldots; s_n]$ has a type \ident{StructVal},
-  but can also have the $\ident{list}\langl\tau\rangl$ type. Conversely, lists that contain
-  non-structural values like objects or options are not of type \ident{StructVal}.
-\item[--] Operations $op$ (omitted) are treated as functions, accepting 
-  \ident{StructVal} and producing an appropriate result type.
-\item[--] Rules for checking class construction and member access are similar to corresponding
-  rules of Featherweight Java.  
-\end{itemize}
-%
-An important part of Featherweight Java that is omitted here is the checking of type declarations
-(ensuring the members are well-typed). We consider only classes generated by our type providers 
-and those are well-typed by construction.
-
-% -------------------------------------------------------------------------------------------------
-
 \begin{figure*}
-\noindent
-\begin{equation*}
-\quad\qquad\boxed{
-\begin{array}{rcl}
- \nameoftag(\ident{string}) &\narrow{=}& \ident{String} \\
- \nameoftag(\ident{bool}) &\narrow{=}& \ident{Boolean} \\
-\end{array}
-\qquad
-\begin{array}{rcl}
- \nameoftag(\ident{number}) &\narrow{=}& \ident{Number} \\
- \nameoftag(\ident{collection}) &\narrow{=}& \ident{List} \\
-\end{array}
-\qquad
-\begin{array}{rcl}
- \nameoftag(\ident{record}) &\narrow{=}& \ident{Record} \\
- \nameoftag(\ident{named}\;\nu) &\narrow{=}& \nu \\
-\end{array}
-}
-\end{equation*}
-
 \begin{multicols}{2}
@@
 \begin{array}{l} 
- \sem{\sigma_p}_e = \tau_p,op(e),\emptyset\quad\textnormal{where}~\tau_p = \sigma_p \\[0.2em]
-\quad\sigma_p, op\in  \{~(\ident{decimal},\ident{asDec}), (\ident{float},\ident{asFloat})~\}
+ \sem{\sigma_p} = \tau_p,\lambda x\rightarrow op(\sigma_p, x),\emptyset\quad\textnormal{where} \\[0.6em] 
+ \quad\sigma_p, \tau_p, op\in  \{~ (\ident{bool}, \ident{bool}, \ident{convPrim})\\
+ \hspace{2.9em} (\ident{int}, \ident{int}, \ident{convPrim}), (\ident{float},\ident{float},\ident{convFloat}),\\
+ \hspace{2.9em} (\ident{string},\ident{string},\ident{convPrim}) ~\}
 \end{array} 
@@
 \begin{array}{l}
- \sem{\,\nu \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}\,}_e = \\[0.1em]
- \quad C, \kvd{new}~C(e), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
- \qquad \;\;C~\textnormal{is a fresh class name} \\[0.1em]
- \qquad \;\,\,L = \kvd{type}~C(v:\ident{StructVal})~=~M_1\ldots M_n  \\[0.1em]
- \qquad M_i = \kvd{member}~\nu_i:\tau_i=e_i\\[0.1em]
- \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_{e'},\;\; e'=\ident{getField}(\nu, \nu_i, e)\\[0.6em]
+ \sem{\,\nu~ \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}\,} = \\[0.1em]
+ \quad C, \lambda x\rightarrow \kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
+ \qquad \;\;C~\textnormal{is a fresh class name} \\
+ \qquad \;\,\,L = \kvd{type}~C(x:\ident{Data})~=~M_1\ldots M_n  \\
+ \qquad M_i = \kvd{member}~\nu_i:\tau_i=\ident{convField}(\nu, \nu_i, x, e_i),\\
+ \qquad \tau_i, e_i, L_i = \sem{\sigma_i} 
+\end{array}
+\end{equation*}
+%
+% Option values
+\begin{equation*}
+\begin{array}{l}
+ \sem{\kvd{nullable}\langl\hat{\sigma}\rangl} = \\[0.2em]
+ \qquad \ident{option}\langl\tau\rangl, \lambda x \rightarrow \ident{convNull}(x,e), L\\[0.2em] 
+ \qquad \textnormal{where}~\tau, e, L = \sem{\hat{\sigma}}
 \end{array}
@@
 \begin{equation*}
-\begin{array}{l}
- \sem{\,[\sigma]\,}_e = \ident{list}\langl\tau\rangl, e_b, L \;\;\textnormal{where}\\[0.4em]
- \qquad e_b = \ident{map}~(\lambda x\rightarrow e')~(\ident{getChildren}(e)) \\[0.1em]
- \qquad \tau, e', L = \sem{\hat{\sigma}}_x
+\hspace{-1.3em}
+\begin{array}{l}
+ \sem{\,[\sigma]\,} = \ident{list}\langl\tau\rangl, \lambda x \rightarrow \ident{convElements}(x, e'), L \;\;\textnormal{where}\\[0.4em]
+ \qquad \tau, e', L = \sem{\hat{\sigma}}
+\end{array}
+\end{equation*}
+%
+% Sum type
+\begin{equation*}
+\begin{array}{l}
+ \sem{\,\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl\,} = \\
+ \quad C, \lambda x\rightarrow\kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\;\;\textnormal{where}\\[0.6em]
+ \qquad \;\;C~\textnormal{is a fresh class name} \\
+ \qquad \;\,\,L = \kvd{type}~C(x:\ident{Data})~=~M_1\ldots M_n \\
+ \qquad M_i = \kvd{member}~\nu_i:\ident{option}\langl\tau_i\rangl=\\
+ \hspace{5.8em}  \kvd{if}~\ident{hasShape}(\sigma_i, x)~\kvd{then}~ \ident{Some}(e_i~x)~\kvd{else}~\ident{None} \\[0.1em]
+ \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_e,\;\; \nu_i=\tytagof{(\sigma_i)}
 \end{array}
@@
 \begin{equation*}
-\begin{array}{l}
- \sem{\bot}_e = \sem{\kvd{null}}_e = \ident{StructVal}, e, \emptyset
-\end{array}
-\end{equation*}
-
-\noindent
-\begin{equation*}
-\hspace{-2.5em}
-\begin{array}{l}
- \sem{\sigma_p}_e = \tau_p,e,\emptyset\quad\textnormal{where}~\tau_p = \sigma_p \\[0.2em]
-\quad\sigma_p \in  \{~ \ident{string},\ident{bool},\ident{int}~\}\\[0.6em]
-\end{array}
-\end{equation*}
-%
-% Sum type
-\begin{equation*}
-\hspace{-2.5em}
-\begin{array}{l}
- \sem{\,\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl\,}_e = \\[0.1em]
- \quad C, \kvd{new}~C(e), L_1\cup\ldots\cup L_n\cup\{ L \}\;\;\textnormal{where}\\[0.6em]
- \qquad \;\;C~\textnormal{is a fresh class name} \\[0.1em]
- \qquad \;\,\,L = \kvd{type}~C(v:\ident{StructVal})~=~M_1\ldots M_n \\[0.1em]
- \qquad M_i = \kvd{member}~\nu_i:\ident{option}\langl\tau_i\rangl=\\[0.1em]
- \hspace{5.8em}  \kvd{if}~\ident{hasType}(\sigma_i, v)~\kvd{then}~
-     \ident{Some}(e_i)~\kvd{else}~\ident{None} \\[0.1em]
- \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_e,\;\; t_i = \tytagof{(\sigma_i)},\;\; \nu_i=\nameoftag{(t_i)}
-\end{array}
-\end{equation*}
-%
-% Option values
-\begin{equation*}
-\hspace{-2.5em}
-\begin{array}{l}
- \sem{\hat{\sigma}\;\,\kvd{option}}_e = \\[0.2em]
- \hspace{1.25em} \ident{option}\langl\tau\rangl, \kvd{if}~\ident{isNull}~e~\kvd{then}~\ident{None}~\kvd{else}~\ident{Some}(e'), L\\[0.2em] 
- \hspace{1.25em} \textnormal{where}~\tau, e', L = \sem{\hat{\sigma}}_e
+\hspace{-1.3em}
+\begin{array}{l}
+ \sem{\bot} = \sem{\kvd{null}} = C, \lambda x\rightarrow \kvd{new}~C(), \{ L \} \;\;\textnormal{where}\\[0.4em]
+ \qquad C~\textnormal{is a fresh class name} \\[0.1em]
+ \qquad L = \kvd{type}~C(v:\ident{Data})
 \end{array}
@@
 % -------------------------------------------------------------------------------------------------
-
-\subsection{Type providers}
-\label{sec:formal-tp}
-
-So far, we defined the type inference algorithm which produces a structural type $\sigma$ from one 
-or more sample documents (\S\ref{sec:inference}) and we defined a simplified model of evaluation
-of F\# (\S\ref{sec:formal-ff}) and F\# Data runtime (\S\ref{sec:formal-tp}). In this section, we 
-define how the type providers work, linking the two parts.
-
-All F\# Data type providers take (one or more) sample documents, infer a common supertype $\sigma$
-and then use it to generate F\# types that are exposed to the programmer.\footnote{The actual 
-implementation provides \emph{erased types} as described in \cite{fsharp-typeprov}. Here, we treat 
-the code as actually generated. This is an acceptable simplification, because F\# Data type providers 
-do not rely on laziness that is available through erased types.}
-
-\paragraph{Type provider mapping.}
-A type provider produces an F\# type $\tau$ together with an expression that wraps an input
-value (of type \ident{StructVal}) as a value of type $\tau$ and a collection of class definitions. 
-We express it using the following mapping:
-%
-\begin{equation*}
-\sem{-}_{-} : (\sigma\times e) \rightarrow (\tau \times e' \times L)
-\end{equation*}
-%
-The mapping $\sem{\sigma}_e$ takes an inferred structural type $\sigma$ and an expression $e$, 
-which represents code to obtain a structural value that is being wrapped. It returns an F\# type 
-$\tau$, an expression $e'$ which constructs a value of type $\tau$ using $e$ and also a set of 
-generated class definitions $L$.
-
-Figure~\ref{fig:tp-generation} shows the rules that define $\sem{-}_{-}$. Primitive types are all
-handled by two rules -- for \ident{decimal} and \ident{float}, we insert a call to an appropriate 
-conversion function from Figure~\ref{fig:op-conversions}.
-
-For records, we generate a class $C$ that takes a structural value as constructor
-parameter. For each record field, we generate a member with the same name as the field.\footnote{The actual
-F\# Data implementation also capitalizes the names.} The body of the member calls \ident{getField} and then
-wraps this expression using $\sem{\sigma_i}$ which maps the field (structural value of type $\sigma_i$) into 
-an F\# value of type $\tau_i$. The returned expression creates a new instance of $C$ and the mapping returns 
-the class $C$ together with all recursively generated classes. Note that the class
-name $C$ is not directly accessed by the user and so we can use arbitrary name, although the actual 
-implementation in F\# Data attempts to infer a reasonable name.\footnote{For example, in 
-\ident{\{\str{person}:\{\str{name}:\str{Tomas}\}\}}, the nested record will be named \ident{Person}
-based on the name of the parent record field.}
-
-A collection type becomes an F\# $\ident{list}\langl\tau\rangl$. The returned expression calls \ident{getChildren}
-(which turns \kvd{null} into an empty list) and then uses \ident{map} to convert nested values to 
-values of an F\# type $\tau$. The handling of option type is similar; if the value is not \kvd{null},
-we wrap the recursively generated conversion expression $e'$ in the \ident{Some} constructor.
-
-As discussed earlier, variant types are also generated as F\# classes with properties. Given a variant
-$\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl$, we get corresponding F\# types $\tau_i$ and generate 
-$n$ members of type $\ident{option}\langl \tau_i\rangl$. When the member is accessed, we need to perform
-a runtime type test using \ident{hasType} to ensure that the value has the right type (similarly to runtime 
-type conversions from the top type in languages like Java). If the type matches, a \ident{Some} value is 
-returned. The type inference algorithm also guarantees that there is only one case for each type tag 
-(\S\ref{sec:inference-commonsuper}) and so we can use the tag when naming the generated member 
-(using the \ident{nameof} function).
 
@@
  \quad \kvd{member}~\ident{Person}~:~\ident{option}\langl\ident{Person}\rangl~= \\[0.1em]
- \qquad \kvd{if}~\ident{hasType}(\{\ident{Age}\!:\!\kvd{option}\langl\ident{int}\rangl, \ident{Name}\!:\!\ident{string}\}, v)~\\[0.1em]
+ \qquad \kvd{if}~\ident{hasShape}(\{\ident{Age}\!:\!\kvd{option}\langl\ident{int}\rangl, \ident{Name}\!:\!\ident{string}\}, v)~\\[0.1em]
  \qquad\quad \kvd{then}~\ident{Some}(\kvd{new}~\ident{Person}(v))~\kvd{else}~\ident{None} \\[0.1em]
  \quad \kvd{member}~\ident{String}~:~\ident{option}\langl\ident{string}\rangl~= \\[0.1em]
- \qquad \kvd{if}~\ident{hasType}(\ident{string}, v)~\\[0.1em]
+ \qquad \kvd{if}~\ident{hasShape}(\ident{string}, v)~\\[0.1em]
  \qquad\quad \kvd{then}~\ident{Some}(\ident{asStr}(v))~\kvd{else}~\ident{None}
@@
 The \ident{PersonOrString} type contains one member for each of the variant case. In the body, they
-check that the value has the correct type using \ident{hasType}. This also implicitly handles \kvd{null}
+check that the value has the correct type using \ident{hasShape}. This also implicitly handles \kvd{null}
 by returning \kvd{false}. As discussed earlier, variants provide easy access to the known cases
@@
 works on \kvd{null} values, \ident{getField} returns \kvd{null} when a field is missing and, for variants,
-the value has the correct type as guaranteed by \ident{hasType}.
+the value has the correct type as guaranteed by \ident{hasShape}.
 \end{proof}

### 2015-11-20 pldi-2016/paper.tex
Final edits in sections 4 and 5

@@
 For records, we generate a class $C$ that takes a data value as constructor parameter. For each 
-record field, we generate a member with the same name as the field.\footnote{The actual
-F\# Data implementation also capitalizes the names.} The body of the member calls \ident{convField}
+record field, we generate a member with the same name as the field.
+The body of the member calls \ident{convField}
 with a function obtained from $\sem{\sigma_i}$. This function turns the field value (data of shape
-$\sigma_i$) into a Foo value of type $\tau_i$. The returned expression creates a new instance of 
+$\sigma_i$) into a Foo value of type $\tau_i$. For optional fields, we also insert a runtime shape
+check using the \ident{hasShape} operation. The returned expression creates a new instance of 
 $C$ and the mapping returns the class $C$ together with all recursively generated classes. Note that 
@@
 \begin{multicols}{2}
-% Primitive structural types become corresponding F# types and conversion is inserted
-\noindent
-\begin{equation*}
+
+% Records 
+\noindent
+\begin{equation*}
+\begin{array}{l}
+ \sem{\,\nu~ \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}\,} = \\[0.1em]
+ \quad C, \lambda x\rightarrow \kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
+ \qquad \;\;C~\textnormal{is a fresh class name} \\
+ \qquad \;\;k~\textnormal{such that}~(i > k) \Leftrightarrow (\sigma_i = \kvd{nullable}\langl \hat{\sigma_i} \rangl) \\
+ \qquad \;\,\,L = \kvd{type}~C(x_1\!:\!\ident{Data})=M_1 \,.\,.\, M_k, M'_{k+1} \,.\,.\, M'_n  \\
+ \qquad M_i = \kvd{member}~\nu_i:\tau_i=\ident{convField}(\nu, \nu_i, x_1, e_i),\\
+ \qquad M'_i = \kvd{member}~\nu_i:\tau_i=\ident{convField}(\nu, \nu_i, x_1, \lambda x_2 \rightarrow \\
+ \hspace{5.8em}  \kvd{if}~\ident{hasShape}(\hat{\sigma_i}, x_2)~\kvd{then}~ e_i~x_2~\kvd{else}~\ident{None}) \\[0.1em]
+ \qquad \tau_i, e_i, L_i = \sem{\sigma_i} 
+\end{array}
+\end{equation*}
+\vspace{-2em}
+
+% Nullables
+\begin{equation*}
+\hspace{-1.25em}
+\begin{array}{l}
+ \sem{\kvd{nullable}\langl\hat{\sigma}\rangl} = \\[0.2em]
+ \qquad \ident{option}\langl\tau\rangl, \lambda x \rightarrow \ident{convNull}(x,e), L\\[0.2em] 
+ \qquad \textnormal{where}~\tau, e, L = \sem{\hat{\sigma}}
+\end{array}
+\end{equation*}
+\vspace{-2em}
+
+% Collections
+\begin{equation*}
+\hspace{-1.25em}
+\begin{array}{l}
+ \sem{\,[\sigma]\,} = \ident{list}\langl\tau\rangl, \lambda x \rightarrow \ident{convElements}(x, e'), L \;\;\textnormal{where}\\[0.4em]
+ \qquad \tau, e', L = \sem{\hat{\sigma}}
+\end{array}
+\end{equation*}
+
+% Primitives
+\noindent
+\begin{equation*}
+\hspace{-1.3em}
 \begin{array}{l} 
@@
 \end{equation*}
-%
-% Records become classes
-\begin{equation*}
-\begin{array}{l}
- \sem{\,\nu~ \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}\,} = \\[0.1em]
- \quad C, \lambda x\rightarrow \kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
- \qquad \;\;C~\textnormal{is a fresh class name} \\
- \qquad \;\,\,L = \kvd{type}~C(x:\ident{Data})~=~M_1\ldots M_n  \\
- \qquad M_i = \kvd{member}~\nu_i:\tau_i=\ident{convField}(\nu, \nu_i, x, e_i),\\
- \qquad \tau_i, e_i, L_i = \sem{\sigma_i} 
-\end{array}
-\end{equation*}
-%
-% Option values
-\begin{equation*}
-\begin{array}{l}
- \sem{\kvd{nullable}\langl\hat{\sigma}\rangl} = \\[0.2em]
- \qquad \ident{option}\langl\tau\rangl, \lambda x \rightarrow \ident{convNull}(x,e), L\\[0.2em] 
- \qquad \textnormal{where}~\tau, e, L = \sem{\hat{\sigma}}
-\end{array}
-\end{equation*}
-%
-% Collections
-\begin{equation*}
-\hspace{-1.3em}
-\begin{array}{l}
- \sem{\,[\sigma]\,} = \ident{list}\langl\tau\rangl, \lambda x \rightarrow \ident{convElements}(x, e'), L \;\;\textnormal{where}\\[0.4em]
- \qquad \tau, e', L = \sem{\hat{\sigma}}
-\end{array}
-\end{equation*}
-%
-% Sum type
+\vspace{-2em}
+
+% Labelled top
 \begin{equation*}
@@
 \end{equation*}
-%
+\vspace{-2em}
+
 % Anything else
@@
 To illustrate how the mechanism works, we consider two examples. First, assume 
-that the inferred type is a record  
-$\{~\ident{Age}\!:\!\kvd{option}\langl\ident{int}\rangl, \ident{Name}\!:\!\ident{string}~ \}$. 
-The rules from Figure~\ref{fig:tp-generation} produce the following class:
-%
-\begin{equation*}
-\begin{array}{l}
- \kvd{type}~\ident{Person}(\ident{v}:\ident{StructVal})~= \\[0.1em]
+that the inferred shape is a record  
+$\ident{Person}~\{~\ident{Age}\!:\!\kvd{option}\langl\ident{int}\rangl, \ident{Name}\!:\!\ident{string}~ \}$. 
+The rules from Figure~\ref{fig:tp-generation} produce the following class (we reduce some 
+function applications for readability):
+%
+\vspace{-0.1em}
+\begin{equation*}
+\begin{array}{l}
+ \kvd{type}~\ident{Person}(x_1:\ident{Data})~= \\[0.1em]
  \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl\ident{int}\rangl~= \\[0.1em]
- \qquad \kvd{if}~\ident{isNull}~(\ident{getField}(\ident{Person},\ident{Age}, \ident{v}))~\kvd{then}~\ident{None} \\[0.1em]
- \qquad \kvd{else}~\ident{Some}(\ident{asInt}(\ident{getField}(\ident{Person},\ident{Age}, \ident{v}))) \\[0.1em]
+ \qquad \ident{convField}(\ident{Person}, \ident{Age}, x_1, \lambda x_2 \rightarrow \\[0.1em]
+ \qquad \quad \kvd{if}~\ident{hasShape}(\ident{int}, x_2)~\kvd{then} \\[0.1em]
+ \qquad \quad \quad \ident{convNull}(x_2, \lambda x_3\rightarrow\ident{convPrim}(\ident{int}, x_3)) \\[0.1em]
+ \qquad \quad \kvd{else}~\ident{None})\\[0.1em]
  \quad \kvd{member}~\ident{Name}~:~\ident{string}~= \\[0.1em]
- \qquad \ident{asStr}(\ident{getField}(\ident{Person},\ident{Name}, \ident{v}))
-\end{array}
-\end{equation*}
-%
-The body of the \ident{Age} member is produced by the case for optional types applied to an expression
-$\ident{getField}(\ident{Person},\ident{Age},v)$. If the returned field is not \kvd{null}, then the member
-calls \ident{asInt} and wraps the result in \ident{Some}. Note that \ident{getField} is defined even 
-when the field does not exist, but it returns \kvd{null}. This lets us treat missing fields as 
-optional. An F\# type corresponding to the (structural) record is \ident{Person} and a structural 
-value $s$ is wrapped by by calling $\kvd{new}~\ident{Person}(s)$.
-
-\paragraph{Example 2.} The second example illustrates the remaining types including collections and 
-variants. Reusing \ident{Person} from the previous example, consider 
-$[\kvd{any}\langl\ident{Person},\ident{string}\rangl]$, a collection which contains a mix 
-of \ident{Person} and string values:
-%
-\begin{equation*}
-\begin{array}{l}
- \kvd{type}~\ident{PersonOrString}(v:\ident{StructVal})~= \\[0.1em]
+ \qquad \ident{convField}(\ident{Person},\ident{Name}, \lambda x_2 \rightarrow \\[0.1em]
+ \qquad \quad \ident{convPrim}(\ident{string}, x_2)))
+\end{array}
+\end{equation*}
+%
+The body of the \ident{Age} member uses \ident{convField} as specified by the case for optional
+record fields. The field shape is nullable and so \ident{convNull} is used in the continuation to 
+convert the value to \ident{None} if \ident{convField} produces a \kvd{null} data value and 
+\ident{hasShape} is used to ensure that the field has the correct shape. The \ident{Name} value should
+be always available and should have the right shape so \ident{convPrim} appears directly in the 
+continuation. This is where the evaluation can get stuck if the field value was missing. The F\# 
+type corresponding to the record shape is \ident{Person} the function to create it from a data value 
+is $\lambda x \rightarrow \kvd{new}~\ident{Person}(x)$.
+
+\paragraph{Example 2.} The second example illustrates the handling of collections and 
+labelled top types. Reusing \ident{Person} from the previous example, consider 
+$[\kvd{any}\langl\ident{Person}~\{ \ldots \},\ident{string}\rangl]$:
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{type}~\ident{PersonOrString}(x:\ident{Data})~= \\[0.1em]
  \quad \kvd{member}~\ident{Person}~:~\ident{option}\langl\ident{Person}\rangl~= \\[0.1em]
- \qquad \kvd{if}~\ident{hasShape}(\{\ident{Age}\!:\!\kvd{option}\langl\ident{int}\rangl, \ident{Name}\!:\!\ident{string}\}, v)~\\[0.1em]
- \qquad\quad \kvd{then}~\ident{Some}(\kvd{new}~\ident{Person}(v))~\kvd{else}~\ident{None} \\[0.1em]
+ \qquad \kvd{if}~\ident{hasShape}(\ident{Person}~\{ \ldots \}, x)~\kvd{then}\\[0.1em]
+ \qquad\quad \ident{Some}(\kvd{new}~\ident{Person}(x))~\kvd{else}~\ident{None} \\[0.1em]
  \quad \kvd{member}~\ident{String}~:~\ident{option}\langl\ident{string}\rangl~= \\[0.1em]
- \qquad \kvd{if}~\ident{hasShape}(\ident{string}, v)~\\[0.1em]
- \qquad\quad \kvd{then}~\ident{Some}(\ident{asStr}(v))~\kvd{else}~\ident{None}
-\end{array}
-\end{equation*}
-%
-The type provider maps the structural type to an F\# type $\ident{list}\langl\ident{PersonOrString}\rangl$. 
-Given a structural document value $s$, the code to obtain the wrapped F\# value is:
-%
-\begin{equation*}
-\ident{map}~(\lambda x\rightarrow\kvd{new}~\ident{PersonOrString}(x))~(\ident{getChildren}(s))
-\end{equation*}
-
-The \ident{PersonOrString} type contains one member for each of the variant case. In the body, they
-check that the value has the correct type using \ident{hasShape}. This also implicitly handles \kvd{null}
-by returning \kvd{false}. As discussed earlier, variants provide easy access to the known cases
-(\ident{string} or \ident{Person}), but they are robust with respect to unexpected inputs.
+ \qquad \kvd{if}~\ident{hasShape}(\ident{string}, x)~\kvd{then}\\[0.1em]
+ \qquad\quad \ident{Some}(\ident{convPrim}(\ident{string}, x))~\kvd{else}~\ident{None}
+\end{array}
+\end{equation*}
+%
+The type provider maps the collection of labelled top shapes to a type $\ident{list}\langl\ident{PersonOrString}\rangl$
+and returns a function that parses a data value as follows:
+%
+\begin{equation*}
+\lambda x_1\rightarrow \ident{convElements}(x_1\lambda x_2\rightarrow\kvd{new}~\ident{PersonOrString}(x_2))
+\end{equation*}
+%
+The \ident{PersonOrString} class contains one member for each of the labels. In the body, they
+check that the input data value has the correct shape using \ident{hasShape}. This also implicitly 
+handles \kvd{null} by returning \kvd{false}. As discussed earlier, labelled top types provide easy 
+access to the known cases (\ident{string} or \ident{Person}), but they require a runtime shape check.
 
@@
 More formally, given a set of sample documents, code using the provided type is guaranteed to work if 
-the inferred type of the input is a subtype of any of the samples. Going back to 
+the inferred shape of the input is preferred with respect to the shape of any of the samples. Going back to 
 \S\ref{sec:inference-subtyping}, this means that:
 %
-\begin{itemize}[noitemsep]
+\begin{itemize}
 \item[--] Input can contain smaller numerical values (e.g., if a sample contains float, the input can contain an integer)
 \item[--] Records in the input can have additional fields
-\item[--] Records in the input can have fewer fields, provided that the type of the field is missing in some of the samples
-\item[--] When a variant is inferred from the sample, the actual input can also contain any other value  
+\item[--] Records in the input can have fewer fields, provided that the type of the field is nullable in some of the samples
+\item[--] When a labelled top type is inferred from the sample, the actual input can also contain any other value,
+  which implements the open world assumption  
 \end{itemize}
@@
 The following lemma states that the provided code (generated in Figure~\ref{fig:tp-generation})
-works correctly on an input $s'$ that is a subtype of the sample $s$. More formally, the provided
-provided expression (with input $s'$) can be reduced to a value and, if it is a class,
+works correctly on an input $d'$ that is a subshape of $d$. More formally, the provided
+provided expression (with input $d'$) can be reduced to a value and, if it is a class,
 all its members can also be reduced to values.
@@
 \label{thm:tp-correctness}
-Given a sample value $s$ and an input value $s'$ such that $\semalt{s} \sqsupseteq \semalt{s'}$
-and provided type, expression and class declarations $\tau, e, L = \sem{\semalt{s}}_{s'}$, 
-then $e \reduce^{*} v$ and if $\tau$ is a class ($\tau=C$) then for all members $N_i$ of the 
-class $C$, it holds that $e.N_i \reduce^{*} v$.
+Given sample data $d$ and an input data value $d'$ such that $\semalt{d} \sqsupseteq \semalt{d'}$
+and provided type, expression and classes $\tau, e, L = \sem{\semalt{d}}$, 
+then $L, e~d' \reduce^{*} v$ and if $\tau$ is a class ($\tau=C$) then for all members $N_i$ of the 
+class $C$, it holds that $L, (e~d').N_i \reduce^{*} v$.
 \end{lemma}
 \begin{proof}
-By induction over the structure of $\sem{-}_-$. For primitives, the conversion functions accept all subtypes.
-For other cases, analyze the provided code to see that it can work on all subtypes (\emph{e.g.}~\ident{getChildren}
-works on \kvd{null} values, \ident{getField} returns \kvd{null} when a field is missing and, for variants,
-the value has the correct type as guaranteed by \ident{hasShape}.
+By induction over the structure of $\sem{-}$. For primitives, the conversion functions accept all subtypes.
+For other cases, analyze the provided code to see that it can work on all subtypes (for example~\ident{convElements}
+works on \kvd{null} values, \ident{convFloat} accepts an integer); for labelled top types and
+optional record fields, the \ident{hasShape} operation is used to guaranteed the correct shape at runtime.
 \end{proof}
@@
 \noindent
-This shows that the provided types are correct with respect to subtyping. Our key
-theorem states that, for any input (which is a subtype of any of the samples) and 
+This shows that provided types are correct with respect to the preferred shape relation. 
+Our key theorem states that, for any input (which is a subshape of any of the samples) and 
 any expression $e$, a well-typed program that uses the provided types does not ``go wrong''.
 Using standard syntactic type safety  \cite{syntactic}, we prove type preservation 
-(reduction does not change type) and progress (an expression that is not a value can be reduced).
+(reduction does not change type) and progress (an expression can be reduced).
 
@@
 \label{thm:safety}
-Assume $s_1, \ldots, s_n$ are samples, $\sigma=\semalt{s_1, \ldots, s_n}$ is an inferred
-type and $\tau,e,L = \sem{\sigma}_x$ are a type, expression and class definitions generated by a 
+Assume $d_1, \ldots, d_n$ are samples, $\sigma=\semalt{d_1, \ldots, d_n}$ is an inferred
+shape and $\tau,e,L = \sem{\sigma}$ are a type, expression and class definitions generated by a 
 type provider.
 
-For all new inputs $s'$ such that $\exists i.(\semalt{s_i} \sqsupseteq \semalt{s'})$, let $e_s=e[x\leftarrow s']$
-be an expression (of type $\tau$) that wraps the input in a provided type. Then, for any expression $e_c$
-(user code) such that $\emptyset; y:\tau \vdash e_c:\tau'$ and $e_c$ does not contain primitive operations
-$op$ as a sub-expression, it is the case that $e_c[y\leftarrow e_s] \reduce^{*} v$ for some value $v$ and
-also $\emptyset; \vdash v : \tau$.
+For all inputs $d'$ such that $\exists i.(\semalt{d_i} \sqsupseteq \semalt{d'})$ and all expressions $e'$
+(representing the user code) such that $e'$ does not contain primitive operations $op$ as a sub-expression and
+$\emptyset; y\!:\!\tau \vdash e'\!:\!\tau'$, it is the case that $L, e[y\leftarrow e'~d'] \reduce^{*} v$ for some value $v$ and
+also $\emptyset; \vdash v : \tau'$.
 \end{theorem}
@@
 \label{thm:rs-preservation}
-Given the class definitions $L$ generated by a type provider as specified in
-the assumptions of Theorem~\ref{thm:safety}, then if $\Gamma \vdash e : \tau$ and 
-$e \reduce^{*} e'$ then $\Gamma \vdash e' : \tau$.
+Given the $\tau, e, L$ generated by a type provider as specified in
+the assumptions of Theorem~\ref{thm:safety}, then if $L, \Gamma \vdash e : \tau$ and 
+$L, e \reduce^{*} e'$ then $\Gamma \vdash e' : \tau$.
 \end{lemma}
@@
 This is not a problem for the (standard) ML subset and object-oriented subset of the calculus. The 
-problematic part are the primitive functions (Figure~\ref{fig:op-conversions}). Given a structural 
-value (of type \ident{StructVal}), the reduction can get stuck if the value does not have a structure 
-required by a specific primitive.
+problematic part are the dynamic data operations (Figure~\ref{fig:op-conversions}). Given a data
+value (of type \ident{Data}), the reduction can get stuck if the value does not have a structure 
+required by a specific operation.
 
 The Lemma~\ref{thm:tp-correctness} guarantees that this does not happen inside the provided type.
-In Theorem~\ref{thm:safety}, we carefully state that we only consider expressions $e_c$ which 
-``[do] not contain primitive operations $op$ as sub-expressions''. This makes sure that the only 
-code working with structural values is the code generated by the type provider.
+We carefully state that we only consider expressions $e'$ which 
+``[do] not contain primitive operations $op$ as sub-expressions''. This ensure that only
+the code generated by a type provider works directly with data values.
 
@@
 Given the assumptions and definitions from Theorem~\ref{thm:safety}, it is the case that
-$e_c[y\leftarrow e_s] \reduce e_c'$.
+$e[y\leftarrow e'~d'] \reduce e''$.
 \end{lemma}
 \begin{proof}
-Proceed by induction over the typing derivation of $L; \emptyset \vdash e_c[y\leftarrow e_s] : \tau'$. 
+Proceed by induction over the typing derivation of $L; \emptyset \vdash e[y\leftarrow e'~d'] : \tau'$. 
 The cases for the ML subset are standard. For member access, we rely on Lemma~\ref{thm:tp-correctness}.

### 2015-11-20 pldi-2016/paper.tex
Final edits in sections 4 and 5

@@
 \label{thm:rs-progress}
-Given the assumptions and definitions from Theorem~\ref{thm:safety}, it is the case that
+Given the assumptions and definitions from Theorem~\ref{thm:safety}, there exists $e''$ such that
 $e[y\leftarrow e'~d'] \reduce e''$.

### 2015-11-20 pldi-2016/paper.tex
Final changes in section 6

@@
 
-In this section, we discuss our practical experience with the safety guarantees provided by the
+In this section, we discuss our experience with the safety guarantees provided by the
 F\# Data type providers and other notable aspects of the implementation.
@@
 The \emph{relativized safety} property does not guarantee the same amount of safety as traditional
-ML type safety, but it reflects the reality of programming with external data sources that is 
+closed-world type safety, but it reflects the reality of programming with external data that is 
 increasingly important \cite{age-of-web}. Type providers do not reduce the safety -- they simply 
@@
 One such issue is the handling of schema change. With type providers, the sample is captured 
-at compile-time. If the schema changes later (so that the input is no longer a subtype of the 
-sample), the program fails at runtime and developers have to handle the exception. This is the 
-same problem that happens when reading data using any other library.
+at compile-time. If the schema changes later (so that the input is no longer related to the  
+shape of the sample), the program fails at runtime and developers have to handle the exception. 
+The same problem happens when using weakly-typed techniques.
 
@@
 
-In our formalization, we treat XML, JSON and CSV uniformly as \emph{structural values}. With the addition of 
-named records (for XML nodes), the definition of structural values is rich enough to capture all 
+In our formalization, we treat XML, JSON and CSV uniformly as \emph{data values}. With the addition of 
+names for records (for XML nodes), the definition of structural values is rich enough to capture all 
 three formats\footnote{The same mechanism has later been used by the HTML type provider 
@@
 One difference between JSON and CSV is that in CSV, the literals have no data types and so 
-we also need to infer the type of primitive values. For example:
+we also need to infer the shape of primitive values. For example:
 %
@@
 as \kvd{null}. The \ident{Date} column uses mixed formats and is inferred as \ident{string} 
-(we support many date formats and ``May 3'' would be parsed correctly). More interestingly,
+(we support many date formats and ``May 3'' would be parsed as date). More interestingly,
 we also infer \ident{Autofiled} as Boolean, because the sample contains only $0$ and $1$.
-This is handled by adding a \ident{bit} type which is a subtype of both \ident{int} and \ident{bool}.
+This is handled by adding a \ident{bit} shape which is preferred of both \ident{int} and \ident{bool}.
 
@@
 Mapping XML documents to structural values is perhaps the most interesting. For each node, we
-create a named record. Attributes become record fields and the body becomes a field with a special
+create a record. Attributes become record fields and the body becomes a field with a special
 name:
@@
 The nested element contains only the $\bullet$ field with the inner text. As with CSV, we
-infer type of primitive values:
+infer shape of primitive values:
 %
@@
 %
-When generating types for XML documents, we also lifts the members nested under the $\bullet$ 
+When generating F\# types for XML documents, we also lifts the members nested under the $\bullet$ 
 field into the parent type to simplify the structure of the generated type.
@@
 The XML type provider also includes an option to use \emph{global inference}. In that case, 
-the inference from values (\ref{sec:formal-inferval}) unifies the types of all records with the 
+the inference from values (\S\ref{sec:formal-inferval}) unifies the shapes of \emph{all} records with the 
 same name. This is useful because, for example, in XHTML all {\small\ttfamily <table>} elements
@@
 implements a special handling of heterogeneous collections. This allows us to avoid inferring a 
-variant types in many common scenarios. In the earlier example, a sample collection 
+labelled top shapes in many common scenarios. In the earlier example, a sample collection 
 contains a record (with \strf{pages} and \strf{page} fields) and a nested collection with values.
 
-Rather than storing a single type for the collection elements as in $[\sigma]$, heterogeneous
-collections store multiple possible element types together with their \emph{  inferred multiplicity} 
+Rather than storing a single shape for the collection elements as in $[\sigma]$, heterogeneous
+collections store multiple possible element shapes together with their \emph{  inferred multiplicity} 
 (zero or one, exactly one, zero or more):
@@
 %
-We omit the details, but finding a common supertype of two heterogeneous 
-collections is analogous to the handling of variants. We merge cases with the same tag (by finding 
-their common super-type) and calculate their new shared multiplicity (for example, by turning 
+We omit the details, but finding a preferred common shape of two heterogeneous 
+collections is analogous to the handling of labelled top types. We merge cases with the same tag (by finding 
+their common shape) and calculate their new shared multiplicity (for example, by turning 
 $1$ and $1?$ into $1?$).
@@
 \cite{links,linq}, into a programming language, required the user to define the schema
-or it has an ad-hoc extension that reads the schema.
+or had an ad-hoc extension that reads the schema.
 
@@
 structural types (including similar heterogeneous collections), but it does not infer the types
-from a sample and extends the type system of the host language (rather than using general purpose
+from a sample. It also extends the type system of the host language (rather than using a general purpose
 embedding).
@@
 many of the patterns we explore in this paper, but formally specifying the safety property would be
-challenging.
+more challenging.
 
@@
 We explored the F\# Data type providers for XML, CSV and JSON. As most real-world data do not come 
-with an explicit schema, the library uses \emph{type inference} that deduces a type from a set of 
-samples. Our type inference algorithm is based on a common supertype relation. For usability 
-reasons, it prefers records and avoids variant types. The algorithm is predictable, which is 
+with an explicit schema, the library uses \emph{shape inference} that deduces a shape from a set of 
+samples. Our inference algorithm is based on a preferred shape relation. It prefers records to 
+encompass the open world assumption and support developer tooling. The algorithm is predictable, which is 
 important as developers need to understand how changing the samples affects the resulting types.
 
-We explore the programming language theory behind type providers. F\# Data is a prime example of 
-type providers, but our work also demonstrates a more general point. The types generated by type 
+We explore the theory behind type providers. F\# Data is a prime example of 
+type providers, but our work demonstrates a more general point. The types generated by type 
 providers can depend on external input and so we can only guarantee \emph{relativized safety}, 
-which says that a program is safe only if the actual inputs satisfy additional conditions --
-in our case, they have to be subtypes of one of the samples.
-
-Type providers have been described before, but this paper is novel in that it explores concrete 
-type providers from the perspective of programming language theory. This is particularly important
-for F\# Data type providers, which have been widely adopted by the industry.
+which says that a program is safe only if the actual inputs satisfy additional conditions.
+
+Type providers have been described before, but this paper is novel in that it explores the 
+properties of type providers that represent the ``types from data'' approach. Our experience suggests 
+that this significantly broadens the applicability of statically typed languages to real-world 
+problems that are often solved by weakly-typed techniques.
 

### 2015-11-20 pldi-2016/paper.tex

@@
 
-In this paper, we integrate external structured data into the F\# type system. As most real-world data
+In this paper, we integrate external structured data into F\# programming. As most real-world data
 does not come with an explicit schema, we develop a shape inference algorithm that infers a shape from 
-representative sample documents and integrate it into the F\# type system using type providers.
+representative sample documents and transforms the shape into F\# types using type providers.
+We present a formalization and a relative type soundness theorem for a simplified version of this process.
 
@@
 
-Social network clients, applications for finding tomorrow's weather or searching train schedules
-all communicate with external services. Increasing number of such services provide REST-based 
+Applications for social network clients, finding tomorrow's weather or searching train schedules
+all communicate with external services. Increasingly these services provide REST-based 
 end-points that return data as CSV, XML or JSON. Despite many schematization efforts, most 
@@
 %
-$\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider at 
+$\ident{JsonProvider}\langl\str{...}\rangl$ invokes a type provider \cite{fsharp-typeprov} at 
 compile-time with the URL as a sample. The type provider infers the structure of the response
@@
 In short, \emph{the types come from the sample data}. In our experience, this technique is 
-immensely practical and surprisingly effective in achieving sound information interchange 
-in heterogeneous systems. This paper presents our approach:
+both practical and surprisingly effective in achieving more sound information interchange 
+in heterogeneous systems. Our contributions are as follows:
 
@@
   \emph{relativized type safety} for the type providers (\S\ref{sec:safety}).
-  This adapts the ML-style type safety for the context of the web.
 \end{itemize}
@@
 
-\item The type providers prefer F\# objects with properties. This is to allow extensible 
-  (open-world) data formats (\S\ref{sec:providers-xml}). It also interacts well with developer tooling
-  as most F\# editors provide auto-completion on ``.'' and so objects are easier to use than 
-  types that require pattern matching.
-
-\item Finally, the mechanism handles practical concerns important in the real world. This includes 
+\item The type providers prefer F\# object types with properties. This allows extensible 
+  (open-world) data formats (\S\ref{sec:providers-xml}). It also interacts well with developer tooling,
+  as most F\# editors provide auto-completion on ``.'' making values of provided types easy to use.
+  This also makes our techniques applicable any language with nominal object types -- for example
+  variations of Java or C# with a type provider mechanism added.
+
+\item Finally, the mechanism handles practical concerns including
   support for different numerical types, \kvd{null} values and missing data.
@@
  \quad|~ \ident{Number}~\kvd{of}~\ident{float} ~~~~|~ \ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
- \quad|~ \ident{String}~\kvd{of}~\ident{string} ~~\qquad\, |~ \ident{Null} \\[0.1em]
+ \quad|~ \ident{String}~\kvd{of}~\ident{string} ~~~~~~ |~ \ident{Null} \\[0.1em]
  \quad|~ \ident{Record}~\kvd{of}~\ident{Map}\langl\ident{string}, \ident{JsonValue}\rangl \\[0.1em]
@@
 to the type provider. Static parameters are resolved at compile-time and have to be constant. 
-The provider analyzes the sample and generates a type  \ident{People}. Most F\# editors also 
+The provider analyzes the sample and provides a type  \ident{People}. F\# editors also 
 execute the type provider in the background at development-time and use the provided types 
@@
 The example iterates over a collection of elements returned by \ident{root.Doc}. The type of \ident{elem} 
-provides a typed access to elements known from the sample and so we can write \ident{elem.Heading}, 
+provides typed access to elements known from the sample and so we can write \ident{elem.Heading}, 
 which returns an optional string value.
@@
 in the real-world (\emph{missing data}, \emph{inconsistent encoding} of primitive values and 
-\emph{heterogeneous shapes}). In our experience, these are the most common issues. The following 
+\emph{heterogeneous shapes}). The following 
 JSON response with government debt information returned by the World Bank\footnote{Available at 
@@
 
-In addition to type providers for JSON and XM, F\# Data also implements a type provider for CSV 
+In addition to type providers for JSON and XML, F\# Data also implements a type provider for CSV 
 (\S\ref{sec:impl-parsing}). We treat CSV files as lists of records (with field for each column) 
@@
 for finding a common shape.
+
+The shape algebra and inference presented here is influenced by the design principles
+we outlined earlier and the type definitions available in the F\# language.
+If applied to other languages, details may differe, for example
+w.r.t. numerical types and missing data.
 
@@
 The rule that allows preferred shapes to have fewer fields (\ref{eq:sub-record3}) is particularly
-important. This is important, because it allows us to prefer records over the \kvd{any} 
+important. This allows us to prefer records with optional fields over the \kvd{any} 
 shape. Given $\{ \ident{name}\!:\!\ident{string} \}$ and $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int} \}$,
@@
 This is a least upper bound and more usable than the top shape $\kvd{var}$.
-
-Some of the aspects of our system (numeric types and handling of missing data) offer a range of
-choices. For example, OCaml also has different numerical types, but would likely always handle 
-\kvd{null} explicitly.
 
@@
 of (\emph{sym}) and (\emph{refl}), but applying those does not affect the result. Reflexivity is
-preserved recursively and all rules, symmetry does not enable different derrivations, even for
+preserved recursively and all rules, symmetry does not enable different derivations, even for
 non-symmetric (\emph{nullable}) and (\emph{prim}).

### 2015-11-20 pldi-2016/paper.tex
Fix typo and make things fit

@@
 
-Applications for social network clients, finding tomorrow's weather or searching train schedules
-all communicate with external services. Increasingly these services provide REST-based 
-end-points that return data as CSV, XML or JSON. Despite many schematization efforts, most 
-services do not come with an explicit schema. At best, the documentation provides sample responses 
-for typical requests.
-
-For example, \url{http://openweathermap.org/current} provides an end-point to get the current weather.
-The documentation shows one sample to illustrate the typical response. Using 
-standard JSON and web libraries, we might write:
+Applications for social networks, finding tomorrow's weather or searching train schedules
+all communicate with external services. Increasingly, these services provide end-points that return 
+data as CSV, XML or JSON. Most such services do not come with an explicit schema. At best, the 
+documentation provides sample responses for typical requests.
+
+For example, \url{http://openweathermap.org/current} documents an end-point to get the current 
+weather with one example. Using standard libraries, we might call it as:
 
@@
 \item The type providers prefer F\# object types with properties. This allows extensible 
-  (open-world) data formats (\S\ref{sec:providers-xml}). It also interacts well with developer tooling,
-  as most F\# editors provide auto-completion on ``.'' making values of provided types easy to use.
-  This also makes our techniques applicable any language with nominal object types -- for example
-  variations of Java or C# with a type provider mechanism added.
-
-\item Finally, the mechanism handles practical concerns including
-  support for different numerical types, \kvd{null} values and missing data.
+  (open-world) data formats (\S\ref{sec:providers-xml}) and it interacts well with developer tooling
+  (\S\ref{sec:providers-json}).
+
+\item The above also makes our techniques applicable to any language with nominal 
+  object types (e.g.~variations of Java or C\# with a type provider mechanism added).
+
+\item Finally, the we handle practical concerns including
+  support for different numerical types, \kvd{null} and missing data.
 \end{itemize}
@@
 execute the type provider in the background at development-time and use the provided types 
-in auto-completion and background type-checking.
+in auto-completion on ``.'' and background type-checking.
 
@@
 If applied to other languages, details may differe, for example
-w.r.t. numerical types and missing data.
+with respect to numerical types and missing data.
 
@@
 paper, we assume that record fields can be freely reordered:
-%
+
+\noindent
 \begin{equation*}
@@
 Next, we extend the core model (sufficient for the relativized safety) 
-with \emph{labelled top types} discussed earleir.
+with \emph{labelled top types} discussed earlier.
 

### 2015-11-20 pldi-2016/paper.tex
record inference fix

@@
 \newcommand{\sem}[1]{\llbracket #1 \rrbracket}
-\newcommand{\semalt}[1]{\llangl #1 \rrangl}
+\newcommand{\semalt}[1]{S(#1)}
 
@@
 \begin{array}{rcl}
- \hat{\sigma} &\narrow{=}& \nu \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
+ \hat{\sigma} &\narrow{=}& \nu \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n, \addopt{\rho_i}  \} \\[0.1em]
                 &\narrow{|}& \ident{float} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
@@
 JSON records always use a single name $\bullet$.
+
+During inference we use row-variables $\rho_i$ to represent the flexibility
+implied by our interpretation of records in samples. 
+When a record {\small\ttfamily \{ x : 3 \}} occurs in a sample,
+it may be combined in a list with other samples that contain more fields. The
+overall shape inferred must account for the potential that any other fields are not
+present in the first sample. The solutions of $\rho_i$ variables is 
+discussed in \S\ref{sec:inference}. We assume substitution and row-unification is
+defined for these in the usual way , except that as substitutions are made
+the $\addopt{-}$ operator from Figure~\ref{fig:subtyping-cst} is applied to all types
+in the vector. After inference any remaining $\rho_i$ variables are ignored (assumed empty).
 
@@
 \begin{definition}
-We write $\sigma_1 \sqsupseteq \sigma_2$ to denote that $\sigma_2$ is preferred over $\sigma_1$. 
+For ground $\sigma_1$, $\sigma_2$ (i.e. without $\rho_i$ variables), we write $\sigma_1 \sqsupseteq \sigma_2$ to denote that $\sigma_2$ is preferred over $\sigma_1$. 
 The shape preference relation is defined as a transitive reflexive closure of the following rules:
@@
   \label{eq:sub-var-top}\tag{P7}
-  \kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl &\sqsupseteq \sigma 
+  \kvd{any} &\sqsupseteq \sigma 
 \end{align}
@@
 \end{align}
+
 \vspace{-1.5em}
+
 \begin{align}
@@
 \end{align}
-\vspace{-1.5em}
-\begin{align}
-\label{eq:sub-record3}\tag{R3}
-\begin{array}{l}
- \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null} \} \sqsupseteq \\
- \quad\nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
-\end{array}
-\end{align}
+
 \end{definition}
@@
 \inference[(record)\;]
-  { (\nu_i = \nu'_j) \Leftrightarrow (i = j) \wedge (i \leq k)
-      \qquad \forall i\in\{ 1 .. k \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
+  { \forall i\in\{ 1 .. n \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
   { \begin{array}{l}
-    \nu \; \{ \nu_1 \!:\! \sigma_1,  \; \ldots \;, \nu_k \!:\! \sigma_k, \; \ldots \;, \nu_n \!:\! \tau_n \} \tsep
-    \nu \; \{ \nu'_1 \!:\! \sigma'_1, \; \ldots \;, \nu'_k \!:\! \sigma'_k, \; \ldots \;, \nu'_m \!:\! \tau'_m \} \vdash\\
-    \nu \; \{ \nu_1 \!:\! \sigma''_1, \; \ldots \; , \nu_k \!:\! \sigma''_k, 
-                            \nu_{k+1} \!:\! \addopt{\sigma_{k+1}}, \ldots, \nu_n \!:\! \addopt{\sigma_n},
-                            \nu'_{k+1} \!:\! \addopt{\sigma'_{k+1}}, \ldots, \nu'_m \!:\! \addopt{\sigma'_m} \}
+    \nu \; \{ \nu_1 \!:\! \sigma_1,  \ldots \;, \nu_n \!:\! \sigma_n \} \tsep
+    \nu \; \{ \nu'_1 \!:\! \sigma'_1, \; \ldots \;, \nu'_m \!:\! \sigma'_m \} \vdash
+    \nu \; \{ \nu_1 \!:\! \sigma''_1, \; \ldots \; , \nu_k \!:\! \sigma''_n \}
     \end{array} }
@@
 
-\caption{Inference judgements that define the common preferred shape relation}
+\caption{The common preferred shape relation}
 \label{fig:subtyping-cst}
@@
 \item The record shapes are covariant (\ref{eq:sub-record1}) and preferred record can have 
-  additional fields (\ref{eq:sub-record2}). The rule (\ref{eq:sub-record3}) implies that records 
-  can omit fields, provided that \kvd{null} is a valid value for the 
-  field (i.e. the field is optional).
+  additional fields (\ref{eq:sub-record2}). 
 \end{itemize}
@@
 \noindent
-The rule that allows preferred shapes to have fewer fields (\ref{eq:sub-record3}) is particularly
-important. This allows us to prefer records with optional fields over the \kvd{any} 
-shape. Given $\{ \ident{name}\!:\!\ident{string} \}$ and $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int} \}$,
-we find a common shape $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int}~\kvd{option} \}$.
-This is a least upper bound and more usable than the top shape $\kvd{var}$.
 
@@
 
-Given two shapes, the \emph{common preferred shape} relation finds a least upper bound
+Given two ground shapes, the \emph{common preferred shape} relation finds a least upper bound
 (Theorem~\ref{thm:lub}). It prefers records, which is important for usability as discussed
@@
 \begin{definition}
-A \emph{common preferred shape} of shapes $\sigma_1$ and $\sigma_2$ is a shape $\sigma$, written 
-$\sigma_1 \tsep \sigma_2 \vdash \sigma$, obtained according to the inference rules in 
+A \emph{common preferred shape} of two ground shapes $\sigma_1$ and $\sigma_2$ is a shape $\sigma$, written 
+$\sigma_1 \tsep \sigma_2 \vdash \sigma$, by the rules in 
 Figure~\ref{fig:subtyping-cst}.
@@
 We define shape \emph{tags} to identify shapes that have a common preferred shape which is not the
-top shape. When finding a common shape of two records (\emph{record}), we return a record that has 
-the union of their fields. The shape of shared fields becomes the common preferred shape of their 
-respective shapes; fields that are present in only one record are marked as nullable using 
-the function $\addopt{-}$.
+top shape. When finding a common shape of two records (\emph{record}), the shape of fields becomes the common preferred shape of their 
+respective shapes. If the field labels of two record types do not match,
+we can assume a common assignment of fields has been determined
+through shape inference (\S\ref{sec:formal-inferval}).
 
@@
 \paragraph{Properties.}
-The set of shapes does not have a \emph{unique} least upper bound, but it 
-has a least upper bound with respect to an equivalence between mutually preferred shapes. 
-The common preferred shape relation defines a function (Theorem~\ref{thm:func}) that finds a
-least upper bound (Theorem~\ref{thm:lub}).
-
-The equivalence on shapes groups records by their non-nullable fields. Given 
-$\sigma_1 = \ident{A}~\{ \ident{a}\!:\!\ident{int}, \ident{b}\!:\!\kvd{nullable}\langl\ident{int}\rangl \}$,
-$\sigma_2 = \ident{A}~\{ \ident{a}\!:\!\ident{int} \}$ and
-$\sigma_3 = \ident{A}~\{ \ident{a}\!:\!\ident{int}, \ident{b}\!:\!\ident{int} \}$,
-it is the case that $\sigma_1 \sqsubseteq \sigma_2$ and $\sigma_1 \sqsupseteq \sigma_2$ but
-$\sigma_3 \sqsubseteq \sigma_1$ and not vice versa.
-
-Similarly to labels on the labelled top type (\S\ref{sec:inference-vars}), the nullable fields 
-of records serve as annotations that provide access to additional values that may or may not be 
-available in the input. In the provided code, these always have to be protected by a runtime check.
-
-\begin{theorem}[Preferred shape function]
+A set of ground shapes has a \emph{unique} least upper bound by the preferred shape ordering.
+For ground shapes, the common preferred shape relation defines a function (Lemma~\ref{thm:func}) that finds 
+the least upper bound (Lemma~\ref{thm:lub}).
+
+\begin{lemma}[Preferred shape function]
 \label{thm:func}
-For all $\sigma_1$ and $\sigma_2$ there exists exactly one $\sigma$ such that 
+For all ground $\sigma_1$ and $\sigma_2$ there exists exactly one $\sigma$ such that 
 $\sigma_1 \tsep \sigma_2 \vdash \sigma$. 
-\end{theorem}
+\end{lemma}
 \begin{proof}
@@
 
-\begin{theorem}[Least upper bound]
+\begin{lemma}[Least upper bound]
 \label{thm:lub}
-If $\sigma_1 \triangledown \sigma_2 \vdash \sigma$ then $\sigma$ is a least upper bound, i.e. 
-$\sigma \sqsupseteq \sigma_1$ and $\sigma \sqsupseteq \sigma_2$ and for all $\sigma'$ such that $\sigma' \sqsupseteq \sigma_1$
-and $\sigma' \sqsupseteq \sigma_2$, it holds that $\sigma' \sqsupseteq \sigma$.
-\end{theorem}
+For ground $\sigma_1$ and $\sigma_2$, if $\sigma_1 \triangledown \sigma_2 \vdash \sigma$ then $\sigma$ is a least upper bound by $\sqsupseteq$.
+\end{lemma}
 \begin{proof}
@@
 
-\noindent
-Next, we extend the core model (sufficient for the relativized safety) 
-with \emph{labelled top types} discussed earlier.
+\subsection{Inferring shapes from samples}
+\label{sec:formal-inferval}
+
+What remains to be 
+specified is how we obtain the shape from data. We represent JSON, XML and CSV documents using
+the same first-order \emph{data} value (\S\ref{sec:impl-parsing}):
+%
+\begin{equation*}
+\begin{array}{lcl}
+ d &\narrow{=}& i \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
+   &\narrow{|}& [d_1; \ldots; d_n] \lsep \nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}
+\end{array}
+\end{equation*}
+%
+The first few cases of represent primitive values ($i$ for integers, $f$ for floating
+point numbers and $s$ for strings) and the \kvd{null} value. A collection is written as a 
+list of values in square brackets. A record starts with a name $\nu$, followed by a 
+sequence of field assignments $\nu_i \mapsto d_i$.
+
+Figure~\ref{fig:shape-inference} defines a mapping $\semalt{d_1,\ldots,d_n}$ which turns a collection of sample 
+data $d_1, \ldots, d_n$ into a shape $\sigma$. We assume each record in each $d_i$ is marked with a fresh row inference
+variable $\rho_i$.  We then choose a minimal solution $\theta$ for these row variables
+in order to equate the record 
+field labels and satisfy the side conditions in rule (record) when multiple record samples are merged.
+This minimization is implemented via row variable unification, we omit the details here.
+
+\begin{figure}
+\noindent
+\begin{equation*}
+\begin{array}{rclcrclcrcl}
+ \semalt{i} &\narrow{=}& \ident{int} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} && \semalt{\kvd{true}} &\narrow{=}& \ident{bool}\\
+ \semalt{f} &\narrow{=}& \ident{float} && \semalt{s} &\narrow{=}& \ident{string} && \semalt{\kvd{false}}  &\narrow{=}& \ident{bool}\\
+\end{array}
+\end{equation*}
+\noindent
+\vspace{-0.5em}
+\begin{equation*}
+\begin{array}{l}
+ \semalt{[d_1; \ldots; d_n]} = [\semalt{d_1, \ldots, d_n}]
+ \\[0.5em]
+ \semalt{\nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}_{\rho_i}} =\\[0.1em]
+ \qquad\nu~\{ \nu_1:\semalt{d_1}, \ldots, \nu_n :\semalt{d_n}, \addopt{\theta(\rho_i)} \}
+ \\[0.5em]
+ \semalt{d_1, \ldots, d_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
+ \qquad\sigma_0 = \bot,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{d_i} \vdash \sigma_i \\[0.1em]
+ \textnormal{Choose minimal}\ \theta\ \textnormal{by ordering}\ \sqsupseteq\ \textnormal{lifted over substitutions} \\[0.1em]
+\end{array}
+\end{equation*}
+\caption{Shape inference}
+\label{fig:shape inference}
+\end{figure}
+
+%
+Primitive  values are mapped to their corresponding shapes. 
+When infering a shape from multiple samples, we use the common preferred shape relation to find a 
+common shape for all values (starting with $\bot$). This operation is used at the top-level 
+(when calling a type provider with multiple samples) and also when inferring the shape of collection
+elements.
 
@@
 
-\subsection{Adding labelled top type}
+\subsection{Adding labelled top shapes}
 \label{sec:inference-vars}
+
+Next, we indicate how the core model (sufficient for the relativized safety) 
+can be extended with \emph{labelled top shapes} discussed earlier.
 
@@
 preferred shape relation and $\kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl$ should still be
-seen as the top type, regardless of the labels.
+seen as the top shape, regardless of the labels.
 
@@
 contain a case with the tag of the other shape. If it does, the two shapes are combined, otherwise 
-a new case is added. When combining two top types (\emph{top-2}), we group the labels that have 
+a new case is added. When combining two top shapes (\emph{top-2}), we group the labels that have 
 shared tags. Finally, (\emph{top-3}) covers the case when we are combining two distinct non-top 
@@
 
-Stating properties of the labels requires a finer relation than \emph{preferred shape}. In particular, 
-it can only be done with respect to the \emph{sample}. We leave the details to future work, but we
+Stating properties of the labels requires refinements to the \emph{preferred shape} relation.
+We leave the details to future work, but we
 note that the algorithm infers the best labels in the sense that there are labels that enable 
@@
 fields of records.
-
-\subsection{Inferring shapes from values}
-\label{sec:formal-inferval}
-
-The common preferred shape relation is at the core of the shape inference. What remains to be 
-specified is how we obtain the shape from data. We represent JSON, XML and CSV documents using
-the same first-order \emph{data} value (\S\ref{sec:impl-parsing}):
-%
-\begin{equation*}
-\begin{array}{lcl}
- d &\narrow{=}& i \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
-   &\narrow{|}& [d_1; \ldots; d_n] \lsep \nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}
-\end{array}
-\end{equation*}
-%
-The first few cases represent primitive values ($i$ for integers, $f$ for floating
-point numbers and $s$ for strings) and the \kvd{null} value. A collection is written as a 
-list of values in square brackets. A record starts with a name $\nu$, followed by a 
-sequence of field assignments $\nu_i \mapsto d_i$.
-
-The following defines a mapping $\semalt{d_1,\ldots,d_n}$ which turns a collection of sample 
-data $d_1, \ldots, d_n$ into a shape $\sigma$:
-%
-\begin{equation*}
-\begin{array}{rclcrclcrcl}
- \semalt{i} &\narrow{=}& \ident{int} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} && \semalt{\kvd{true}} &\narrow{=}& \ident{bool}\\
- \semalt{f} &\narrow{=}& \ident{float} && \semalt{s} &\narrow{=}& \ident{string} && \semalt{\kvd{false}}  &\narrow{=}& \ident{bool}\\
-\end{array}
-\end{equation*}
-\noindent
-\vspace{-0.5em}
-\begin{equation*}
-\begin{array}{l}
- \semalt{d_1, \ldots, d_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
- \qquad\sigma_0 = \bot,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{d_i} \vdash \sigma_i
- \\[0.5em]
- \semalt{[d_1; \ldots; d_n]} = [\semalt{d_1, \ldots, d_n}]
- \\[0.5em]
- \semalt{\nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}} =\\[0.1em]
- \qquad\nu~\{ \nu_1:\semalt{d_1}, \ldots, \nu_n :\semalt{d_n} \}
-\end{array}
-\end{equation*}
-%
-We overload the notation and write $\semalt{s}$ when inferring shape from a single value. Primitive 
-values are mapped to their corresponding shapes. For records, we return infer field shapes
-from the individual values.
-
-When infering a shape from multiple samples, we use the common preferred shape relation to find a 
-common shape for all values (starting with $\bot$). This operation is used at the top-level 
-(when calling a type provider with multiple samples) and also when inferring the shape of collection
-elements.
 
@@
 This section presents the theoretical model of strucutral type providers. We introduce the Foo calculus 
-which is an object-oriented subset of F\# based on SML \cite{sml} and Featherweight Java \cite{fwjava}. It 
-includes operations for working with data that model the F\# Data runtime. Finally, we describe how 
+which is a subset of F\# using object-property notation with
+added operations for working with weakly typed structured data along the lines of the F\# Data runtime. Finally, we describe how 
 type providers turn inferred shapes into Foo classes (\S\ref{sec:formal-tp}).
 
-Type providers map the dirty world of data into a nice world of F\#. To model this, the Foo calculus
+Type providers map the ``dirty'' world of weakly typed structured data into a nice world of strong types. To model this, the Foo calculus
 does not have \kvd{null} values and data values $d$ are never directly exposed.  
+Furthermore Foo is simply typed: despite using class types and object notation 
+for properties, it has no subtyping.
 
@@
 
-The type \ident{Data} is the type of all structural data $d$. A class definition $L$ consists of 
+The type \ident{Data} is the type of structural data $d$. A class definition $L$ consists of 
 a constructor and zero or more members. Values $v$ include previously defined data $d$; expressions
@@
 
-The Foo programs can only work with raw data using certain primitive operations. Those are modelled
+The Foo programs can only work with \ident{Data} values using certain primitive operations. Those are modelled
 by the $op$ primitives. In F\# Data, those are internal and users never access them directly.

### 2015-11-20 pldi-2016/paper.tex
updates

@@
 $\hat{\sigma}$) and \emph{nullable shapes} that encompass missing and \kvd{null} values 
-(written as $\sigma$). We write $\nu$ for record names and record field names. In the rest of the
-paper, we assume that record fields can be freely reordered:
+(written as $\sigma$). We write $\nu$ for record names and record field names.
 
@@
 \begin{array}{rcl}
- \hat{\sigma} &\narrow{=}& \nu \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n, \addopt{\rho_i}  \} \\[0.1em]
+ \hat{\sigma} &\narrow{=}& \nu \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n, \rho_i  \} \\[0.1em]
                 &\narrow{|}& \ident{float} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
@@
 JSON records always use a single name $\bullet$.
-
-During inference we use row-variables $\rho_i$ to represent the flexibility
-implied by our interpretation of records in samples. 
-When a record {\small\ttfamily \{ x : 3 \}} occurs in a sample,
-it may be combined in a list with other samples that contain more fields. The
-overall shape inferred must account for the potential that any other fields are not
-present in the first sample. The solutions of $\rho_i$ variables is 
-discussed in \S\ref{sec:inference}. We assume substitution and row-unification is
-defined for these in the usual way , except that as substitutions are made
-the $\addopt{-}$ operator from Figure~\ref{fig:subtyping-cst} is applied to all types
-in the vector. After inference any remaining $\rho_i$ variables are ignored (assumed empty).
 
@@
 discussed earlier (\S\ref{sec:providers-xml}).
+
+In the rest of the paper, we assume that record fields can be freely reordered.
+
+During inference we use row-variables $\rho_i$ in record shapes to represent the flexibility
+arising from records in samples. For example, when a 
+record {\small\ttfamily \{ x : 3 \}} occurs in a sample,
+it may be combined through inference with other
+samples such as {\small\ttfamily \{ x : 3, y : 4 \}} that contain more fields. The
+overall shape inferred must account for the fact that any extra fields are optional,
+giving an inferred shape {\small\ttfamily \{ x : int, y : option<int> \}}.
+The process of solving $\rho_i$ variables is  discussed in \S\ref{sec:formal-inferval}.
 
@@
 
-What remains to be 
-specified is how we obtain the shape from data. We represent JSON, XML and CSV documents using
+We now specify how we obtain the shape from data. We represent JSON, XML and CSV documents using
 the same first-order \emph{data} value (\S\ref{sec:impl-parsing}):
@@
 Figure~\ref{fig:shape-inference} defines a mapping $\semalt{d_1,\ldots,d_n}$ which turns a collection of sample 
-data $d_1, \ldots, d_n$ into a shape $\sigma$. We assume each record in each $d_i$ is marked with a fresh row inference
-variable $\rho_i$.  We then choose a minimal solution $\theta$ for these row variables
-in order to equate the record 
-field labels and satisfy the side conditions in rule (record) when multiple record samples are merged.
-This minimization is implemented via row variable unification, we omit the details here.
+data $d_1, \ldots, d_n$ into a shape $\sigma$. Before applying $S$, we assume each record in each $d_i$ is marked with a fresh row inference
+variable $\rho_i$.  
+We then choose a ground, minimal substitution $\theta$ for row variables.
+Because $\rho_i$ variables represent potentially missing fields,
+the $\addopt{-}$ operator from Figure~\ref{fig:subtyping-cst} is applied to all types
+in the vector. 
+This must be sufficient to equate the record field labels and satisfy the side conditions in rule (record) when multiple record samples are merged.
+In practice, $\theta$ is found via row variable unification, we omit the details here.
+No $\rho_i$ variables remain after inference as the substitution chosen is ground.
 
@@
 \caption{Shape inference}
-\label{fig:shape inference}
+\label{fig:shape-inference}
 \end{figure}
@@
 \paragraph{Properties.}
-The revised algorithm stil finds a shape which is the least upper bound. This means that 
+The revised algorithm still finds a shape which is the least upper bound. This means that 
 labelled top shape is only inferred when there is no other alternative.

### 2015-11-20 pldi-2016/paper.tex
Remove shape checks from optional fields

--- 2015-11-20 pldi-2016/paper.tex
@@
 For records, we generate a class $C$ that takes a data value as constructor parameter. For each 
-record field, we generate a member with the same name as the field.
-The body of the member calls \ident{convField}
-with a function obtained from $\sem{\sigma_i}$. This function turns the field value (data of shape
-$\sigma_i$) into a Foo value of type $\tau_i$. For optional fields, we also insert a runtime shape
-check using the \ident{hasShape} operation. The returned expression creates a new instance of 
+record field, we generate a member with the same name as the field. The body of the member calls 
+\ident{convField} with a function obtained from $\sem{\sigma_i}$. This function turns the field 
+value (data of shape $\sigma_i$) into a Foo value of type $\tau_i$. The returned expression creates a new instance of 
 $C$ and the mapping returns the class $C$ together with all recursively generated classes. Note that 
@@
 
-% Records 
-\noindent
-\begin{equation*}
-\begin{array}{l}
- \sem{\,\nu~ \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}\,} = \\[0.1em]
- \quad C, \lambda x\rightarrow \kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
- \qquad \;\;C~\textnormal{is a fresh class name} \\
- \qquad \;\;k~\textnormal{such that}~(i > k) \Leftrightarrow (\sigma_i = \kvd{nullable}\langl \hat{\sigma_i} \rangl) \\
- \qquad \;\,\,L = \kvd{type}~C(x_1\!:\!\ident{Data})=M_1 \,.\,.\, M_k, M'_{k+1} \,.\,.\, M'_n  \\
- \qquad M_i = \kvd{member}~\nu_i:\tau_i=\ident{convField}(\nu, \nu_i, x_1, e_i),\\
- \qquad M'_i = \kvd{member}~\nu_i:\tau_i=\ident{convField}(\nu, \nu_i, x_1, \lambda x_2 \rightarrow \\
- \hspace{5.8em}  \kvd{if}~\ident{hasShape}(\hat{\sigma_i}, x_2)~\kvd{then}~ e_i~x_2~\kvd{else}~\ident{None}) \\[0.1em]
- \qquad \tau_i, e_i, L_i = \sem{\sigma_i} 
-\end{array}
-\end{equation*}
-\vspace{-2em}
-
-% Nullables
-\begin{equation*}
-\hspace{-1.25em}
-\begin{array}{l}
- \sem{\kvd{nullable}\langl\hat{\sigma}\rangl} = \\[0.2em]
- \qquad \ident{option}\langl\tau\rangl, \lambda x \rightarrow \ident{convNull}(x,e), L\\[0.2em] 
- \qquad \textnormal{where}~\tau, e, L = \sem{\hat{\sigma}}
-\end{array}
-\end{equation*}
-\vspace{-2em}
-
-% Collections
-\begin{equation*}
-\hspace{-1.25em}
-\begin{array}{l}
- \sem{\,[\sigma]\,} = \ident{list}\langl\tau\rangl, \lambda x \rightarrow \ident{convElements}(x, e'), L \;\;\textnormal{where}\\[0.4em]
- \qquad \tau, e', L = \sem{\hat{\sigma}}
-\end{array}
-\end{equation*}
-
 % Primitives 
@@
 \begin{equation*}
-\hspace{-1.3em}
 \begin{array}{l} 
@@
 
+% Records
+\begin{equation*}
+\begin{array}{l}
+ \sem{\,\nu~ \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}\,} = \\[0.1em]
+ \quad C, \lambda x\rightarrow \kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
+ \qquad \;\;C~\textnormal{is a fresh class name} \\
+ \qquad \;\,\,L = \kvd{type}~C(x_1\!:\!\ident{Data})=M_1 \ldots M_n \\
+ \qquad M_i = \kvd{member}~\nu_i:\tau_i=\ident{convField}(\nu, \nu_i, x_1, e_i),\\
+ \qquad \tau_i, e_i, L_i = \sem{\sigma_i} 
+\end{array}
+\end{equation*}
+\vspace{-2em}
+
+% Collections
+\begin{equation*}
+\begin{array}{l}
+ \sem{\,[\sigma]\,} = \ident{list}\langl\tau\rangl, \lambda x \rightarrow \ident{convElements}(x, e'), L \;\;\textnormal{where}\\[0.4em]
+ \qquad \tau, e', L = \sem{\hat{\sigma}}
+\end{array}
+\end{equation*}
+
 % Labelled top
+\noindent
 \begin{equation*}
@@
 
+% Nullables
+\begin{equation*}
+\hspace{-1.3em}
+\begin{array}{l}
+ \sem{\kvd{nullable}\langl\hat{\sigma}\rangl} = \\[0.2em]
+ \qquad \ident{option}\langl\tau\rangl, \lambda x \rightarrow \ident{convNull}(x,e), L\\[0.2em] 
+ \qquad \textnormal{where}~\tau, e, L = \sem{\hat{\sigma}}
+\end{array}
+\end{equation*}
+\vspace{-2em}
+
 % Anything else
@@
  \qquad \ident{convField}(\ident{Person}, \ident{Age}, x_1, \lambda x_2 \rightarrow \\[0.1em]
- \qquad \quad \kvd{if}~\ident{hasShape}(\ident{int}, x_2)~\kvd{then} \\[0.1em]
- \qquad \quad \quad \ident{convNull}(x_2, \lambda x_3\rightarrow\ident{convPrim}(\ident{int}, x_3)) \\[0.1em]
- \qquad \quad \kvd{else}~\ident{None})\\[0.1em]
+ \qquad \quad \ident{convNull}(x_2, \lambda x_3\rightarrow\ident{convPrim}(\ident{int}, x_3))~) \\[0.1em]
  \quad \kvd{member}~\ident{Name}~:~\ident{string}~= \\[0.1em]
- \qquad \ident{convField}(\ident{Person},\ident{Name}, \lambda x_2 \rightarrow \\[0.1em]
+ \qquad \ident{convField}(\ident{Person},\ident{Name}, x_1, \lambda x_2 \rightarrow \\[0.1em]
  \qquad \quad \ident{convPrim}(\ident{string}, x_2)))

### 2015-11-20 pldi-2016/paper.tex
updates

--- 2015-11-20 pldi-2016/paper.tex
@@
 worth noting that our shape inference uses labelled top shapes only as the last resort 
-(Theorem~\ref{thm:lub}, \S\ref{sec:impl-hetero}).
+(Lemma~\ref{thm:lub}, \S\ref{sec:impl-hetero}).
 
@@
 Given two ground shapes, the \emph{common preferred shape} relation finds a least upper bound
-(Theorem~\ref{thm:lub}). It prefers records, which is important for usability as discussed
+(Lemma~\ref{thm:lub}). It prefers records, which is important for usability as discussed
 earlier.

### 2015-11-20 pldi-2016/paper.tex
fix 3

@@
 
-This section presents the theoretical model of strucutral type providers. We introduce the Foo calculus 
-which is a subset of F\# using object-property notation with
+This section presents the formal model of type providers for structured data. 
+To represent the programming language that hosts the type provider and the
+provided declarations, we introduce the Foo calculus,
+a subset of F\# using object-property notation with
 added operations for working with weakly typed structured data along the lines of the F\# Data runtime. Finally, we describe how 
@@
 
-Type providers map the ``dirty'' world of weakly typed structured data into a nice world of strong types. To model this, the Foo calculus
+Type providers for structured data map the ``dirty'' world of weakly typed 
+semi-structured data into a ``nice'' world of strong types. To model this, the Foo calculus
 does not have \kvd{null} values and data values $d$ are never directly exposed.  
 Furthermore Foo is simply typed: despite using class types and object notation 
-for properties, it has no subtyping.
+for notational convenience, it has no subtyping.
 
@@
  L &\narrow{=}& \kvd{type}~C(\overline{x:\tau}) = \overline{M} \\[0.0em]
- M &\narrow{=}& \kvd{member}~N:\tau=
+ M &\narrow{=}& \kvd{member}~N:\tau= e
 \\[0.6em]
@@
 in the record, the actual record value has a field of the same name with a matching shape. The last 
-line defines a ``catch all'' pattern, which returns \kvd{false} for all remaining cases. Although
-not elegant, if we treat $e_1 \vee e_2$ and $e_1 \wedge e_2$ as syntactic sugar for \kvd{if} then
+line defines a ``catch all'' pattern, which returns \kvd{false} for all remaining cases. 
+We treat $e_1 \vee e_2$ and $e_1 \wedge e_2$ as syntactic sugar for \kvd{if}-\kvd{then}-\kvd{else} so
 the rule is expressed using just Foo expressions.
@@
 \paragraph{Reduction.}
-The reduction relation is of the form $L, e \reduce e'$. We often omit class declarations 
-$L$ and we write $L, e \reduce^{*} e'$ to denote 
-the reflexive and transitive closure of $\reduce$.
+The reduction relation is of the form $L, e \reduce e'$. We omit class declarations 
+$L$ where implied by context and write $e \reduce^{*} e'$ for
+the reflexive, transitive closure of $\reduce$.
 
@@
 definition in the assumption to obtain the body of a member. The remaining six rules
-give standard reductions for a functional language.
+give standard reductions.
 
@@
 $L; \Gamma \vdash e : \tau$ where the context also contains a set of class declarations $L$.
-The fragment demonstrates the differences and similarities with SML and Featherweight Java:
+The fragment demonstrates the differences and similarities with Featherweight Java \cite{fwjava}:
 %
@@
   \ident{Data} is the only type.
-\item[--] Operations $op$ (omitted) have a family of types (Doo does not have polymorphic types),
-  accepting \ident{Data} as one of the arguments and producing a non-data Foo type.
+\item[--] Operations $op$ (omitted) have a family of types (Foo does not have polymorphic types),
+  accepting \ident{Data} as one of the arguments and producing a non-\ident{Data} Foo type.
 \item[--] Rules for checking class construction and member access are similar to corresponding
@@
 
-\caption{Featherweight F\# -- Remaining reduction rules}
+\caption{Foo -- Remaining reduction rules}
 \label{fig:ff-reduction}
@@
 and then use it to generate F\# types that are exposed to the programmer.\footnote{The actual 
-implementation provides \emph{erased types} as described in \cite{fsharp-typeprov}. Here, we treat 
+implementation provides \emph{lazy erased types} as described in \cite{fsharp-typeprov}. Here, we treat 
 the code as actually generated. This is an acceptable simplification, because F\# Data type providers 
-do not rely on laziness that is available through erased types.}
+do not rely on laziness or erasure of type provision.}
 
@@
 $C$ and the mapping returns the class $C$ together with all recursively generated classes. Note that 
-the class name $C$ is not directly accessed by the user and so we can use arbitrary name, although the 
+the class name $C$ is not directly accessed by the user and so we can use an arbitrary name, although the 
 actual implementation in F\# Data attempts to infer a reasonable name.\footnote{For example, in 
@@
 
-A collection shape becomes an Foo $\ident{list}\langl\tau\rangl$. The returned expression calls \ident{convElements}
-(which returns empty list for data value \kvd{null}). The last parameter is the recursively obtained
+A collection shape becomes a Foo $\ident{list}\langl\tau\rangl$. The returned expression calls \ident{convElements}
+(which returns the empty list for data value \kvd{null}). The last parameter is the recursively obtained
 conversion function for the shape of elements $\sigma$. The handling of the nullable shape is similar,
@@
 
-As discussed earlier, labelled top shape are also generated as Foo classes with properties. Given 
+As discussed earlier, labelled top shapes are also generated as Foo classes with properties. Given 
 $\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl$, we get corresponding F\# types $\tau_i$ and generate 
@@
 
-\caption{Featherweight F\# -- Fragment of type checking}
+\caption{Foo -- Fragment of type checking}
 \label{fig:ff-typecheck}
@@
 
-\caption{Type provider -- generation of featherweight F\# types from inferred structural types}
+\caption{Type provider -- generation of Foo types from inferred structural types}
 \label{fig:tp-generation}
@@
 This shows that provided types are correct with respect to the preferred shape relation. 
+This is an obligation upon the author of a type provider, which is why
+type soundness is ``relative''.
 Our key theorem states that, for any input (which is a subshape of any of the samples) and 
@@
 
-For all inputs $d'$ such that $\exists i.(\semalt{d_i} \sqsupseteq \semalt{d'})$ and all expressions $e'$
+For all inputs $d'$ such that $\sigma \sqsupseteq \semalt{d'}$ and all expressions $e'$
 (representing the user code) such that $e'$ does not contain primitive operations $op$ as a sub-expression and
@@
 \begin{proof}
-By induction over the reduction $\reduce$. The cases for the ML subset of Featherweight F\# 
+By induction over the reduction $\reduce$. The cases for the ML subset of Foo 
 are standard. For (\emph{member}), we check that code generated by type providers

### 2015-11-20 pldi-2016/paper.tex
fix 3

@@
 This shows that provided types are correct with respect to the preferred shape relation. 
-This is an obligation upon the author of a type provider, which is why
-type soundness is ``relative''.
+In general, this is an obligation upon the author of a type provider, which is a second reason why
+the type soundness is ``relative''.
 Our key theorem states that, for any input (which is a subshape of any of the samples) and 
@@
 
-The F\# Data library has been widely adopted by the industry and is one of the most downloaded
+The F\# Data library has been widely adopted by users and is one of the most downloaded
 F\# libraries.\footnote{At the time of writing, the library has over 80,000 downloads on NuGet 
@@
 
-The \emph{relativized safety} property does not guarantee the same amount of safety as traditional
-closed-world type safety, but it reflects the reality of programming with external data that is 
-increasingly important \cite{age-of-web}. Type providers do not reduce the safety -- they simply 
-illuminate the existing issues.
-
-One such issue is the handling of schema change. With type providers, the sample is captured 
-at compile-time. If the schema changes later (so that the input is no longer related to the  
-shape of the sample), the program fails at runtime and developers have to handle the exception. 
-The same problem happens when using weakly-typed techniques.
+The \emph{relativized safety} property does not guarantee safety in the same way as traditional
+closed-world type safety, but it reflects the reality that programming with external data is
+increasingly important \cite{age-of-web}. Type providers increase the safety of this kind of programming,
+seen by example in \S\ref{sec:introduction}.
+
+One such issue is the handling of sample change. With type providers, the sample is captured 
+at compile-time. If the sample changes later (so that the input is no longer related to the  
+shape of the sample), the program can fail at runtime and developers have to handle the exception. 
+The same problem happens when using weakly-typed code with explicit failure cases.
 
@@
 points the JSON type provider at a sample using a live URL. This has the advantage that a re-compilation 
-fails when the schema changes, which is an indication that the program needs to be updated to reflect the
-change.
-
-In general, there is no better solution for plain XML, CSV and JSON data sources. However, some data 
-sources provide versioning support (with meta-data about how the schema changed). For those, a type 
-provider could adapt automatically, but we leave this for future work.
+fails when the sample changes, which is an indication that the program needs to be updated to reflect the
+change. Repeated type-checking of programs by continuous integration systems can be used to help 
+detect such changes.
+
+In general, schema-free XML, CSV and JSON data sources will necessarily require techniques akin
+to those we have shown. However, some data sources provide schema and schema versioning support 
+(with meta-data about how the schema changed). For those, a type 
+provider could be written to adapt automatically, but we leave this for future work.
 
@@
 \paragraph{Reading XML documents.}
-Mapping XML documents to structural values is perhaps the most interesting. For each node, we
+Mapping XML documents to structural values is more interesting. For each node, we
 create a record. Attributes become record fields and the body becomes a field with a special
@@
 
-When introducing type providers (\S\ref{sec:providers-sum}), we mentioned that F\# Data 
-implements a special handling of heterogeneous collections. This allows us to avoid inferring a 
+When introducing type providers (\S\ref{sec:providers-sum}), we mentioned how F\# Data 
+handles heterogeneous collections. This allows us to avoid inferring 
 labelled top shapes in many common scenarios. In the earlier example, a sample collection 
@@
 
-C$\omega$ \cite{comega-xs} is the most similar to F\# Data. It extends C\# with types similar to our
-structural types (including similar heterogeneous collections), but it does not infer the types
-from a sample. It also extends the type system of the host language (rather than using a general purpose
-embedding).
-
-\paragraph{Advanced type systems.}
+C$\omega$ \cite{comega-xs} bears similarity to F\# Data. Like many other
+languages, it extends a host language with advanced types similar to our
+structural shapes (including similar heterogeneous collections), but it 
+does not infer the types from a sample. In contrast, in F\# Data 
+we have avoided extending F\# at all, and the simplicity of the Foo 
+target language shows we have avoided placing strong requirements 
+(such as structural records) on the host language. Instead we provide nominal 
+types based on structure, rather than adding an advanced
+system of structural types into the host language.
+
+\paragraph{Advanced type systems with meta-programming.}
 A number of other advanced type system features could be used to tackle the problem discussed 
@@
 makes formal specification of safety (Theorem~\ref{thm:safety}) difficult. Extending our 
-\emph{relativized safety} to \emph{probabilistic safety} is an interesting future work.
+\emph{relativized safety} to \emph{probabilistic safety} is an interesting future direction.
 
 The PADS project \cite{pads-dsl,pads-ml} tackles a more general problem of handling \emph{any} data format.
-The schema definitions in PADS are similar to our structural type. The structure inference for LearnPADS
+The schema definitions in PADS are similar to our shapes. The structure inference for LearnPADS
 \cite{pads-learn} infers the data format from a flat input stream. A PADS type provider could follow
@@
 
-We explored the F\# Data type providers for XML, CSV and JSON. As most real-world data do not come 
+We explored the F\# Data type providers for XML, CSV and JSON. As most real-world data does not come 
 with an explicit schema, the library uses \emph{shape inference} that deduces a shape from a set of 
@@
 
-We explore the theory behind type providers. F\# Data is a prime example of 
+We explored the theory behind type providers. F\# Data is a prime example of 
 type providers, but our work demonstrates a more general point. The types generated by type 
@@
 that this significantly broadens the applicability of statically typed languages to real-world 
-problems that are often solved by weakly-typed techniques.
+problems that are often solved by error-prone weakly-typed techniques.
 

### 2015-11-20 pldi-2016/paper.tex
Minor changes to make things fit

@@
 does not come with an explicit schema, we develop a shape inference algorithm that infers a shape from 
-representative sample documents and transforms the shape into F\# types using type providers.
+representative sample documents and integrates it into the F\# type system using type providers.
 We present a formalization and a relative type soundness theorem for a simplified version of this process.
@@
 
-For example, \url{http://openweathermap.org/current} documents an end-point to get the current 
-weather with one example. Using standard libraries, we might call it as:
+For example, \url{http://openweathermap.org/current} contains one eample to document an end-point 
+to get the current weather. Using standard libraries, we might call it as:
 
@@
 The code assumes that the response has a particular shape described in the documentation. The
-root node must be a record with a \str{main} field, which has to be another record containing
-a numerical \str{temp} field. When the shape is different, the code simply fails with an exception. 
+root node must be a record with a \strf{main} field, which has to be another record containing
+a numerical \strf{temp} field. When the shape is different, the code simply fails with an exception. 
 While not immediately unsound, the code is manifestly prone to errors if strings are misspelled 
@@
 
-In short, \emph{the types come from the sample data}. In our experience, this technique is 
+In short, the \emph{types} come from the sample \emph{data}. In our experience, this technique is 
 both practical and surprisingly effective in achieving more sound information interchange 
@@
 
-\item Finally, the we handle practical concerns including
+\item Finally, we handle practical concerns including
   support for different numerical types, \kvd{null} and missing data.
@@
  \quad|~ \ident{Number}~\kvd{of}~\ident{float} ~~~~|~ \ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
- \quad|~ \ident{String}~\kvd{of}~\ident{string} ~~~~~~ |~ \ident{Null} \\[0.1em]
+ \quad|~ \ident{String}~\kvd{of}~\ident{string} ~~~~~\, |~ \ident{Null} \\[0.1em]
  \quad|~ \ident{Record}~\kvd{of}~\ident{Map}\langl\ident{string}, \ident{JsonValue}\rangl \\[0.1em]
@@
 \noindent
-The standard approach to print the names and ages would be to pattern match on the parsed 
+The standard way to print the names and ages would be to pattern match on the parsed 
 \ident{JsonValue}, check that the top-level node is a \ident{Array} and iterate over the elements 
@@
 \url{http://openweathermap.org} changes the shape of the response, the code in \S\ref{sec:introduction}
-will not re-compile and the developer knows that the code needs to be changed.
+will not re-compile and the developer knows that the code needs to be corrected.
 
@@
 The sample code is easy to write thanks to the fact that most F\# editors provide auto-completion
-when ``.'' is typed (see the supplementary screencast). The developer does not need to look at the 
+when ``.'' is typed (see the supplementary screencast). The developer does not need to examine the 
 sample JSON file to see what fields are available. To support this scenario, our type providers 
@@
 The example iterates over a collection of elements returned by \ident{root.Doc}. The type of \ident{elem} 
-provides typed access to elements known from the sample and so we can write \ident{elem.Heading}, 
+provides typed access to elements known statically from the sample and so we can write \ident{elem.Heading}, 
 which returns an optional string value.
@@
 when the document structure is defined using XML Schema, documents may contain elements prefixed 
-with other namespaces.} This is the important \emph{open world assumption} about external data. 
+with other namespaces.} This is the fundamental \emph{open world assumption} about external data. 
 Actual input might be an element about which nothing is known.
@@
 Throughout the introduction, we used data sets that demonstrate the typical problems that are frequent
-in the real-world (\emph{missing data}, \emph{inconsistent encoding} of primitive values and 
-\emph{heterogeneous shapes}). The following 
-JSON response with government debt information returned by the World Bank\footnote{Available at 
+in the real-world (missing data, inconsistent encoding of primitive values and heterogeneous shapes). 
+The following JSON response with government debt information returned by the World Bank\footnote{Available at 
 \url{http://data.worldbank.org}} demonstrates all three problems:
@@
 \noindent
-First of all, the top-level element is a collection containing two values of different kind.
+First of all, the top-level element is a collection containing two values of different shape.
 The first is a record with meta-data about the current page and the second is an array with data. 
@@
 in \S\ref{sec:impl-hetero}) and provides a type with properties \ident{Record} for the former and 
-\ident{Array} for the latter. Second, the \ident{value} field is \kvd{null} for some records. Third, 
+\ident{Array} for the latter. Second, the the field \ident{value} is \kvd{null} for some records. Third, 
 numbers can be represented in JSON as numeric literals (without quotes), but here, they are 
@@
 We first define the shape of structured data $\sigma$. We use the term \emph{shape} to distinguish 
-shapes from programming language \emph{types} $\tau$ (type providers generate the latter from the former). 
+shapes of data from programming language \emph{types} $\tau$ (type providers generate the latter from the former). 
 Next, we define the preference relation on shapes $\sigma$ and describe the algorithm 
@@
 we outlined earlier and the type definitions available in the F\# language.
-If applied to other languages, details may differe, for example
+If applied to other languages, details may differ, for example
 with respect to numerical types and missing data.
@@
 \begin{array}{rcl}
- \hat{\sigma} &\narrow{=}& \nu \; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n, \rho_i  \} \\[0.1em]
+ \hat{\sigma} &\narrow{=}& \nu \; \{ \nu_1 \!:\! \sigma_1, \ldots, \nu_n \!:\! \sigma_n,\; \rho_i  \} \\[0.1em]
                 &\narrow{|}& \ident{float} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
@@
 Non-nullable shapes include records (consisting of a name and fields with their shapes) and 
-primitives. Names of records arising from XML are the names of the XML elements while
-JSON records always use a single name $\bullet$.
+primitives. The row variables $\rho_i$ are discussed below. Names of records arising from XML are the names of the XML elements.
+For JSON records we always use a single name $\,\bullet$. We assume that record fields can be freely
+reordered.
 
@@
 
-In the rest of the paper, we assume that record fields can be freely reordered.
-
-During inference we use row-variables $\rho_i$ in record shapes to represent the flexibility
+During inference we use row-variables $\rho_i$ \cite{rows-cardelli} in record shapes to represent the flexibility
 arising from records in samples. For example, when a 
-record {\small\ttfamily \{ x : 3 \}} occurs in a sample,
+record $\ident{Point}\,\{ \ident{x} \mapsto \num{3} \}$ occurs in a sample,
 it may be combined through inference with other
-samples such as {\small\ttfamily \{ x : 3, y : 4 \}} that contain more fields. The
+samples such as $\ident{Point}\,\{\, \ident{x}\mapsto\num{3}, \ident{y}\mapsto\num{4} \}$ that contain more fields. The
 overall shape inferred must account for the fact that any extra fields are optional,
-giving an inferred shape {\small\ttfamily \{ x : int, y : option<int> \}}.
+giving an inferred shape $\ident{Point}\,\{ \ident{x}\!:\!\ident{int},~ \ident{y}\!:\!\kvd{option}\langl\ident{int}\rangl\}$.
 The process of solving $\rho_i$ variables is discussed in \S\ref{sec:formal-inferval}.
@@
   \label{eq:sub-null}\tag{P2}
-  \sigma &\sqsupseteq \kvd{null}  &(\textnormal{iff}~\sigma \neq \hat{\sigma})  \\[-0.2em]
+  \sigma &\sqsupseteq \kvd{null}  &(\textnormal{for}~\sigma \neq \hat{\sigma})  \\[-0.2em]
   \label{eq:sub-opt}\tag{P3}
@@
   all shapes excluding non-nullable shapes $\hat{\sigma}$. Any non-nullable shape is preferred
-  over its nullable version (\ref{eq:sub-opt}); nullable shapes and collections are 
+  over its nullable version (\ref{eq:sub-opt})
+
+\item Nullable shapes and collections are 
   covariant (\ref{eq:sub-opt-cov}, \ref{eq:sub-col}).
@@
 (Lemma~\ref{thm:lub}). It prefers records, which is important for usability as discussed
-earlier.
+earlier (\S\ref{sec:providers-xml}).
 
@@
 top shape. When finding a common shape of two records (\emph{record}), the shape of fields becomes the common preferred shape of their 
-respective shapes. If the field labels of two record types do not match,
-we can assume a common assignment of fields has been determined
-through shape inference (\S\ref{sec:formal-inferval}).
+respective shapes. The field labels of two record types can always be made to match,
+through the shape inference from sample (\S\ref{sec:formal-inferval}),
+\emph{before} finding a common preferred shape using $\triangledown$.
 
@@
 \paragraph{Properties.}
-A set of ground shapes has a \emph{unique} least upper bound by the preferred shape ordering.
-For ground shapes, the common preferred shape relation defines a function (Lemma~\ref{thm:func}) that finds 
+The partially ordered set of ground shapes has a unique least upper bound.
+The $\triangledown$ relation defines a function (Lemma~\ref{thm:func}) that finds 
 the least upper bound (Lemma~\ref{thm:lub}).
@@
 of (\emph{sym}) and (\emph{refl}), but applying those does not affect the result. Reflexivity is
-preserved recursively and all rules, symmetry does not enable different derivations, even for
-non-symmetric (\emph{nullable}) and (\emph{prim}).
+preserved recursively and symmetry does not enable different derivations.
 \end{proof}
@@
 \begin{proof}
-By induction over $\vdash$. The algorithm only infers the top shape \kvd{any} when for non-nullable
+By induction over $\vdash$. Note that the algorithm only infers the top shape \kvd{any} when for non-nullable
 shapes of distinct tags and so no better preferred shape exists.
@@
 
-We now specify how we obtain the shape from data. We represent JSON, XML and CSV documents using
-the same first-order \emph{data} value (\S\ref{sec:impl-parsing}):
+We now specify how we obtain the shape from data. As clrified later (\S\ref{sec:impl-parsing}), 
+we represent JSON, XML and CSV documents using the same first-order \emph{data} value:
 %
@@
 %
-The first few cases of represent primitive values ($i$ for integers, $f$ for floating
-point numbers and $s$ for strings) and the \kvd{null} value. A collection is written as a 
+The definition includes primitive values ($i$ for integers, $f$ for floats
+and $s$ for strings) and \kvd{null}. A collection is written as a 
 list of values in square brackets. A record starts with a name $\nu$, followed by a 
@@
 in the vector. 
-This must be sufficient to equate the record field labels and satisfy the side conditions in rule (record) when multiple record samples are merged.
-In practice, $\theta$ is found via row variable unification, we omit the details here.
+This is sufficient to equate the record field labels and satisfy the side conditions in rule (record) when multiple record samples are merged.
+In practice, $\theta$ is found via row variable unification \cite{rows-remy}. We omit the details here.
 No $\rho_i$ variables remain after inference as the substitution chosen is ground.
 
-\begin{figure}
-\noindent
-\begin{equation*}
-\begin{array}{rclcrclcrcl}
- \semalt{i} &\narrow{=}& \ident{int} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} && \semalt{\kvd{true}} &\narrow{=}& \ident{bool}\\
- \semalt{f} &\narrow{=}& \ident{float} && \semalt{s} &\narrow{=}& \ident{string} && \semalt{\kvd{false}}  &\narrow{=}& \ident{bool}\\
-\end{array}
-\end{equation*}
-\noindent
-\vspace{-0.5em}
-\begin{equation*}
-\begin{array}{l}
- \semalt{[d_1; \ldots; d_n]} = [\semalt{d_1, \ldots, d_n}]
- \\[0.5em]
- \semalt{\nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}_{\rho_i}} =\\[0.1em]
- \qquad\nu~\{ \nu_1:\semalt{d_1}, \ldots, \nu_n :\semalt{d_n}, \addopt{\theta(\rho_i)} \}
- \\[0.5em]
- \semalt{d_1, \ldots, d_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
- \qquad\sigma_0 = \bot,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{d_i} \vdash \sigma_i \\[0.1em]
- \textnormal{Choose minimal}\ \theta\ \textnormal{by ordering}\ \sqsupseteq\ \textnormal{lifted over substitutions} \\[0.1em]
-\end{array}
-\end{equation*}
-\caption{Shape inference}
-\label{fig:shape-inference}
-\end{figure}
-
-%
 Primitive  values are mapped to their corresponding shapes. 
 When infering a shape from multiple samples, we use the common preferred shape relation to find a 
-common shape for all values (starting with $\bot$). This operation is used at the top-level 
-(when calling a type provider with multiple samples) and also when inferring the shape of collection
+common shape for all values (starting with $\bot$). This operation is used  
+when calling a type provider with multiple samples and also when inferring the shape of collection
 elements.
@@
 
-Next, we indicate how the core model (sufficient for the relativized safety) 
-can be extended with \emph{labelled top shapes} discussed earlier.
-
 When analyzing the structure of shapes, it suffices to consider a single top shape \kvd{any}.
 However, the type providers need more information to provide typed access to the possible 
-alternative shapes of data, such as XML nodes. To support track the possible shapes, we label 
-the \kvd{any} shape:
+alternative shapes of data, such as XML nodes.
+
+\begin{figure}[!h]
+\noindent
+\begin{equation*}
+\begin{array}{rclcrclcrcl}
+ \semalt{i} &\narrow{=}& \ident{int} && \semalt{\kvd{null}}  &\narrow{=}& \kvd{null} && \semalt{\kvd{true}} &\narrow{=}& \ident{bool}\\
+ \semalt{f} &\narrow{=}& \ident{float} && \semalt{s} &\narrow{=}& \ident{string} && \semalt{\kvd{false}}  &\narrow{=}& \ident{bool}\\
+\end{array}
+\end{equation*}
+\noindent
+\vspace{-0.5em}
+\begin{equation*}
+\begin{array}{l}
+ \semalt{[d_1; \ldots; d_n]} = [\semalt{d_1, \ldots, d_n}]
+ \\[0.5em]
+ \semalt{\nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}_{\rho_i}} =\\[0.1em]
+ \qquad\nu~\{ \nu_1:\semalt{d_1}, \ldots, \nu_n :\semalt{d_n}, \addopt{\theta(\rho_i)} \}
+ \\[0.5em]
+ \semalt{d_1, \ldots, d_n} = \sigma_n \quad\textnormal{where}\\[0.1em]
+ \qquad\sigma_0 = \bot,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{d_i} \vdash \sigma_i \\[0.6em]
+ \textnormal{Choose minimal}\ \theta\ \textnormal{by ordering}\ \sqsupseteq\ \textnormal{lifted over substitutions} \\[0.1em]
+\end{array}
+\end{equation*}
+\caption{Shape inference from sample data}
+\label{fig:shape-inference}
+\vspace{-0.5em}
+\end{figure}
+
+We extend the core model (sufficient for the relativized safety) 
+with \emph{labelled top shapes} defined as:
 % 
@@
 The common preferred shape relation is extended to find a labelled top shape that best represents 
-the sample. We limit the number of labels and avoid nesting by grouping shapes by the shape tag 
-introduced earlier (Figure~\ref{fig:subtyping-cst}).
-For example, rather than inferring $\kvd{any}\langl\ident{int}, \kvd{any}\langl\ident{bool}, \ident{float}\rangl\rangl$, 
-our algorithm finds the common preferred shape of $\ident{int}$ and $\ident{float}$ and produces 
+the sample. We limit the number of labels and avoid nesting by grouping shapes by the tag 
+introduced earlier.
+Rather than inferring $\kvd{any}\langl\ident{int}, \kvd{any}\langl\ident{bool}, \ident{float}\rangl\rangl$, 
+our algorithm joins $\ident{int}$ and $\ident{float}$ and produces 
 $\kvd{any}\langl\ident{float},\ident{bool}\rangl$.
 
-The three rules in Figure~\ref{fig:subtyping-cst-var} replace the earlier (\emph{top}). When 
-combining a top with another shape (\emph{top-1}), the labelled top shape may or may not already 
+The new rules for \kvd{any} appear in Figure~\ref{fig:subtyping-cst-var}. When 
+combining a top with another shape (\emph{top-1}), the labels may or may not already 
 contain a case with the tag of the other shape. If it does, the two shapes are combined, otherwise 
 a new case is added. When combining two top shapes (\emph{top-2}), we group the labels that have 
-shared tags. Finally, (\emph{top-3}) covers the case when we are combining two distinct non-top 
-shapes. As top shapes implicitly permit \kvd{null} values, we use an auxiliary function $\dropopt{-}$ 
-to make nullable shapes non-nullable (when possible) to simplify the label.
-
-\paragraph{Properties.}
+shared tags. Finally, (\emph{top-3}) combines two distinct non-top 
+shapes. As top shapes implicitly permit \kvd{null} values, we make the labels non-nullable 
+using $\dropopt{-}$.
+
 The revised algorithm still finds a shape which is the least upper bound. This means that 
 labelled top shape is only inferred when there is no other alternative.
-
-Stating properties of the labels requires refinements to the \emph{preferred shape} relation.
-We leave the details to future work, but we
-note that the algorithm infers the best labels in the sense that there are labels that enable 
-typed access to every possible value in the sample, but not more. The same is the case for nullable
-fields of records.
-
-% ==================================================================================================
-% 
-%    #######
-%    #        ####  #####  #    #   ##   #      # ######   ##   ##### #  ####  #    #
-%    #       #    # #    # ##  ##  #  #  #      #     #   #  #    #   # #    # ##   #
-%    #####   #    # #    # # ## # #    # #      #    #   #    #   #   # #    # # #  #
-%    #       #    # #####  #    # ###### #      #   #    ######   #   # #    # #  # #
-%    #       #    # #   #  #    # #    # #      #  #     #    #   #   # #    # #   ##
-%    #        ####  #    # #    # #    # ###### # ###### #    #   #   #  ####  #    #
-%
-% ==================================================================================================
-
-\section{Formalising type providers}
-\label{sec:formal}
-
-This section presents the formal model of type providers for structured data. 
-To represent the programming language that hosts the type provider and the
-provided declarations, we introduce the Foo calculus,
-a subset of F\# using object-property notation with
-added operations for working with weakly typed structured data along the lines of the F\# Data runtime. Finally, we describe how 
-type providers turn inferred shapes into Foo classes (\S\ref{sec:formal-tp}).
-
-Type providers for structured data map the ``dirty'' world of weakly typed 
-semi-structured data into a ``nice'' world of strong types. To model this, the Foo calculus
-does not have \kvd{null} values and data values $d$ are never directly exposed.  
-Furthermore Foo is simply typed: despite using class types and object notation 
-for notational convenience, it has no subtyping.
-
-% -------------------------------------------------------------------------------------------------
-
-\begin{figure*}
-\noindent
-\begin{equation*}
-\hspace{1.1em}
-\begin{array}{l}
-\ident{hasShape}(\nu~ \{ \nu_1 \!:\! \sigma_1, \ldots, \nu_n \!:\! \sigma_n \}, \nu'~\{ \nu'_1\mapsto d_1, \ldots, \nu'_m\mapsto d_m \}) \reduce (\nu = \nu') ~\wedge \\
-  \quad (~ ((\nu_1 = \nu'_1) \wedge \ident{hasShape}(\sigma_1, d_1)) \vee\ldots\vee ((\nu_1 = \nu'_m) \wedge \ident{hasShape}(\sigma_1, d_m)) \vee \ldots \vee\\
-  \quad ~\; ((\nu_n = \nu'_1) \wedge \ident{hasShape}(\sigma_n, d_1)) \vee\ldots\vee ((\nu_n = \nu'_m) \wedge \ident{hasShape}(\sigma_n, d_m))~)
-\\[0.5em]
-\ident{hasShape}([\sigma], [d_1; \ldots; d_n]) \reduce \ident{hasShape}(\sigma, d_1)\wedge\ldots\wedge\ident{hasShape}(\sigma, d_n) \\
-\ident{hasShape}([\sigma], \kvd{null}) \reduce \kvd{true} 
-\end{array}  
-\quad
-\begin{array}{l}
-\ident{convFloat}(\ident{float}, i) \reduce f~(f=i) \\
-\ident{convFloat}(\ident{float}, f) \reduce f \\[0.75em]
-\ident{convNull}(\kvd{null}, e) \reduce \ident{None} \\
-\ident{convNull}(d, e) \reduce \ident{Some}(e~d) 
-\end{array}
-\end{equation*}
-%
-\vspace{-0.75em}
-%
-\begin{equation*}
-\begin{array}{l}
-\ident{hasShape}(\ident{string}, s) \reduce \kvd{true} \\
-\ident{hasShape}(\ident{int}, i) \reduce \kvd{true}\\
-\ident{hasShape}(\ident{bool}, d) \reduce \kvd{true} \quad(\textnormal{when}~d\in{\kvd{true},\kvd{false}} )\\
-\ident{hasShape}(\ident{float}, d) \reduce \kvd{true} \quad(\textnormal{when}~d=i ~\textnormal{or}~ d=f) \\
-\ident{hasShape}(\_, \_) \reduce \kvd{false} \\
-\end{array}
-\;\;
-\begin{array}{l}
-\ident{convPrim}(\sigma, d) \reduce d\quad(\sigma,d\in\{ (\ident{int},i),(\ident{string},s),(\ident{bool}, b) \})
-\\[0.25em]
-\ident{convField}(\nu,\nu_i, \nu~\{\ldots, \nu_i=d_i, \ldots\}, e) \reduce e~d_i\\
-\ident{convField}(\nu,\nu', \nu~\{\ldots, \nu_i=d_i, \ldots\}, e) \reduce e~\kvd{null}\quad(\nexists i.\nu_i=\nu' )
-\\[0.25em]
-\ident{convElements}([d_1; \ldots; d_n], e) \reduce e~d_1 :: \ldots :: e~d_n :: \kvd{nil}  \\  
-\ident{convElements}(\kvd{null}) \reduce \kvd{nil}
-\end{array}
-\end{equation*}
-
-\caption{Reduction rules for conversion functions}
-\label{fig:op-conversions}
-\vspace{-0.5em}
-\end{figure*}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{The Foo calculus}
-\label{sec:formal-ff}
-
-The syntax of the calculus is shown in Figure~\ref{fig:foo-syntax}. We only need classes with 
-parameter-less members and without inheritance. A class has a single implicit constructor and the 
-declaration closes over constructor parameters.
-
-The type \ident{Data} is the type of structural data $d$. A class definition $L$ consists of 
-a constructor and zero or more members. Values $v$ include previously defined data $d$; expressions
-$e$ include class construction, member access and the usual functional constructs (functions,
-lists, options) and conditionals. The $op$ constructs are discussed next.
-
-% -------------------------------------------------------------------------------------------------
-
-\paragraph{Dynamic data operations.}
-
-The Foo programs can only work with \ident{Data} values using certain primitive operations. Those are modelled
-by the $op$ primitives. In F\# Data, those are internal and users never access them directly.
 
@@
 
+Stating properties of the labels requires refinements to the \emph{preferred shape} relation.
+We leave the details to future work, but we
+note that the algorithm infers the best labels in the sense that there are labels that enable 
+typed access to every possible value in the sample, but not more. The same is the case for nullable
+fields of records.
+
+% ==================================================================================================
+% 
+%    #######
+%    #        ####  #####  #    #   ##   #      # ######   ##   ##### #  ####  #    #
+%    #       #    # #    # ##  ##  #  #  #      #     #   #  #    #   # #    # ##   #
+%    #####   #    # #    # # ## # #    # #      #    #   #    #   #   # #    # # #  #
+%    #       #    # #####  #    # ###### #      #   #    ######   #   # #    # #  # #
+%    #       #    # #   #  #    # #    # #      #  #     #    #   #   # #    # #   ##
+%    #        ####  #    # #    # #    # ###### # ###### #    #   #   #  ####  #    #
+%
+% ==================================================================================================
+
+\section{Formalising type providers}
+\label{sec:formal}
+
+This section presents the formal model of type providers for structured data. 
+To represent the programming language that hosts the type provider and the
+provided declarations, we introduce the Foo calculus,
+a subset of F\# with objects and properties, extended with
+operations for working with weakly typed structured data along the lines of the F\# Data runtime. Finally, we describe how 
+type providers turn inferred shapes into Foo classes (\S\ref{sec:formal-tp}).
+
+Type providers for structured data map the ``dirty'' world of weakly typed 
+structured data into a ``nice'' world of strong types. To model this, the Foo calculus
+does not have \kvd{null} values and data values $d$ are never directly exposed.  
+Furthermore Foo is simply typed: despite using class types and object notation 
+for notational convenience, it has no subtyping.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure*}
+\noindent
+\begin{equation*}
+\hspace{1.1em}
+\begin{array}{l}
+\ident{hasShape}(\nu~ \{ \nu_1 \!:\! \sigma_1, \ldots, \nu_n \!:\! \sigma_n \}, \nu'~\{ \nu'_1\mapsto d_1, \ldots, \nu'_m\mapsto d_m \}) \reduce (\nu = \nu') ~\wedge \\
+  \quad (~ ((\nu_1 = \nu'_1) \wedge \ident{hasShape}(\sigma_1, d_1)) \vee\ldots\vee ((\nu_1 = \nu'_m) \wedge \ident{hasShape}(\sigma_1, d_m)) \vee \ldots \vee\\
+  \quad ~\; ((\nu_n = \nu'_1) \wedge \ident{hasShape}(\sigma_n, d_1)) \vee\ldots\vee ((\nu_n = \nu'_m) \wedge \ident{hasShape}(\sigma_n, d_m))~)
+\\[0.5em]
+\ident{hasShape}([\sigma], [d_1; \ldots; d_n]) \reduce \ident{hasShape}(\sigma, d_1)\wedge\ldots\wedge\ident{hasShape}(\sigma, d_n) \\
+\ident{hasShape}([\sigma], \kvd{null}) \reduce \kvd{true} 
+\end{array}  
+\quad
+\begin{array}{l}
+\ident{convFloat}(\ident{float}, i) \reduce f~(f=i) \\
+\ident{convFloat}(\ident{float}, f) \reduce f \\[0.75em]
+\ident{convNull}(\kvd{null}, e) \reduce \ident{None} \\
+\ident{convNull}(d, e) \reduce \ident{Some}(e~d) 
+\end{array}
+\end{equation*}
+%
+\vspace{-0.75em}
+%
+\begin{equation*}
+\begin{array}{l}
+\ident{hasShape}(\ident{string}, s) \reduce \kvd{true} \\
+\ident{hasShape}(\ident{int}, i) \reduce \kvd{true}\\
+\ident{hasShape}(\ident{bool}, d) \reduce \kvd{true} \quad(\textnormal{when}~d\in{\kvd{true},\kvd{false}} )\\
+\ident{hasShape}(\ident{float}, d) \reduce \kvd{true} \quad(\textnormal{when}~d=i ~\textnormal{or}~ d=f) \\
+\ident{hasShape}(\_, \_) \reduce \kvd{false} \\
+\end{array}
+\;\;
+\begin{array}{l}
+\ident{convPrim}(\sigma, d) \reduce d\quad(\sigma,d\in\{ (\ident{int},i),(\ident{string},s),(\ident{bool}, b) \})
+\\[0.25em]
+\ident{convField}(\nu,\nu_i, \nu~\{\ldots, \nu_i=d_i, \ldots\}, e) \reduce e~d_i\\
+\ident{convField}(\nu,\nu', \nu~\{\ldots, \nu_i=d_i, \ldots\}, e) \reduce e~\kvd{null}\quad(\nexists i.\nu_i=\nu' )
+\\[0.25em]
+\ident{convElements}([d_1; \ldots; d_n], e) \reduce e~d_1 :: \ldots :: e~d_n :: \kvd{nil}  \\  
+\ident{convElements}(\kvd{null}) \reduce \kvd{nil}
+\end{array}
+\end{equation*}
+
+\caption{Reduction rules for conversion functions}
+\label{fig:op-conversions}
+\vspace{-0.5em}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{The Foo calculus}
+\label{sec:formal-ff}
+
+The syntax of the calculus is shown in Figure~\ref{fig:foo-syntax}. 
+The type \ident{Data} is the type of structural data $d$. A class definition $L$ consists of 
+a single constructor and zero or more parameter-less members. The declaration implicitly closes
+over the constructor parameters.
+Values $v$ include previously defined data $d$; expressions
+$e$ include class construction, member access, usual functional constructs (functions,
+lists, options) and conditionals. The $op$ constructs are discussed next.
+
+% -------------------------------------------------------------------------------------------------
+
+\paragraph{Dynamic data operations.}
+
+The Foo programs can only work with \ident{Data} values using certain primitive operations. Those are modelled
+by the $op$ primitives. In F\# Data, those are internal and users never access them directly.
+
 The behaviour of the dynamic data operations is defined by the reduction rules in 
@@
 make the conversion explicit for primitive values (\ident{convPrim}) and assume that primitive
-values are shared, but we convert all other values. For example \ident{convElements} turns a 
+values are shared between \emph{data} and Foo values, but we convert all other values. For example \ident{convElements} turns a 
 data collection $[d_1, \ldots, d_n]$ into a Foo list $v_1 :: .. :: v_n :: \kvd{nil}$.
@@
 Finally, we also define a runtime shape test \ident{hasShape}.
-The most complex case is handling of records where we check that for each field $\nu_1, \ldots, \nu_n$ 
+For records, we have to check that for each field $\nu_1, \ldots, \nu_n$ 
 in the record, the actual record value has a field of the same name with a matching shape. The last 
 line defines a ``catch all'' pattern, which returns \kvd{false} for all remaining cases. 
-We treat $e_1 \vee e_2$ and $e_1 \wedge e_2$ as syntactic sugar for \kvd{if}-\kvd{then}-\kvd{else} so
-the rule is expressed using just Foo expressions.
+We treat $e_1 \vee e_2$ and $e_1 \wedge e_2$ as a syntactic sugar for \kvd{if}\,.\,.\,\kvd{then}\,.\,.\,\kvd{else} so
+the result of the reduction is just a Foo expression.
 
@@
 
-Figure~\ref{fig:ff-reduction} shows the reduction rules. The (\emph{ctx}) rule models the eager
+Figure~\ref{fig:ff-reduction} shows the reduction rules.  The (\emph{member}) rule reduces a member access using a class 
+definition in the assumption. The (\emph{ctx}) rule models the eager
 evaluation of F\# and performs a reduction inside a sub-expression specified by an evaluation 
@@
 constructor and dynamic data operation arguments or $v::E$ in list initialization.
-
 We write $e[\overline{x} \leftarrow \overline{v}]$ for the result of replacing variables $\overline{x}$ by
-values $\overline{v}$ in an expression. The (\emph{member}) rule reduces a member access using a class 
-definition in the assumption to obtain the body of a member. The remaining six rules
+values $\overline{v}$ in an expression. The remaining six rules
 give standard reductions.
@@
 error condition. The stuck states can only be due to the dynamic data operations (e.g.
-$\ident{convFloat}(\ident{float},\kvd{null})$). The relativized type safety (Theorem~\ref{thm:safety})
+$\ident{convFloat}(\ident{float},\kvd{null})$). The relativized safety (Theorem~\ref{thm:safety})
 characterizes the additional conditions on input data under which Foo programs do not ``go wrong''.
@@
 %
-\begin{itemize}
+\begin{itemize}[noitemsep]
 \item[--] All data values $d$ have the type \ident{Data}, but primitive data values (Booleans, 
@@
 class definitions. We express it using the following mapping:
-%
+
+\noindent
 \begin{equation*}
@@
 $[\kvd{any}\langl\ident{Person}~\{ \ldots \},\ident{string}\rangl]$:
-%
+
+\noindent
 \begin{equation*}
@@
 This shows that provided types are correct with respect to the preferred shape relation. 
-In general, this is an obligation upon the author of a type provider, which is a second reason why
+This is an obligation upon the author of a type provider, which is a second reason why
 the type soundness is ``relative''.
-Our key theorem states that, for any input (which is a subshape of any of the samples) and 
+Our key theorem states that, for any input which is a subshape the inferred shape and 
 any expression $e$, a well-typed program that uses the provided types does not ``go wrong''.
@@
 \begin{proof}
-By induction over the reduction $\reduce$. The cases for the ML subset of Foo 
+By induction over $\reduce$. The cases for the ML subset of Foo 
 are standard. For (\emph{member}), we check that code generated by type providers
@@
 The progress lemma states that evaluation of a well-typed program does not reach an undefined state. 
-This is not a problem for the (standard) ML subset and object-oriented subset of the calculus. The 
+This is not a problem for the (standard) ML \cite{sml} subset and object-oriented subset \cite{fwjava} of the calculus. The 
 problematic part are the dynamic data operations (Figure~\ref{fig:op-conversions}). Given a data
@@
 increasingly important \cite{age-of-web}. Type providers increase the safety of this kind of programming,
-seen by example in \S\ref{sec:introduction}.
-
-One such issue is the handling of sample change. With type providers, the sample is captured 
-at compile-time. If the sample changes later (so that the input is no longer related to the  
+as seen earlier (\S\ref{sec:introduction}).
+
+One common issue is the handling of schema change. With type providers, the sample is captured 
+at compile-time. If the schema changes (so that the input is no longer related to the  
 shape of the sample), the program can fail at runtime and developers have to handle the exception. 
@@
 
-In general, schema-free XML, CSV and JSON data sources will necessarily require techniques akin
-to those we have shown. However, some data sources provide schema and schema versioning support 
+In general, XML, CSV and JSON data sources without explicit schema will necessarily require techniques akin
+to those we have shown. However, some data sources provide explicit schema with schema versioning support 
 (with meta-data about how the schema changed). For those, a type 

### 2016-02-01 pldi-2016/paper.tex
Deanonymized version
Adding lub function
Working on final edits

@@
-\documentclass[10pt,preprint,blind,clearpagebib]{sigplanconf}
+\documentclass[10pt,preprint,clearpagebib]{sigplanconf}
 
@@
 
-For example, \url{http://openweathermap.org/current} contains one eample to document an end-point 
-to get the current weather. Using standard libraries, we might call it as:
-
-\noindent
-\begin{equation*}
-\begin{array}{l}
- \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC}) \\
+For example, \url{http://openweathermap.org/current} contains one example to document an end-point 
+to get the current weather. Using standard libraries, we might call it as\footnote{We use an abbreviation
+\url{http://api.owm.org/?q=X} for the full URL:
+\url{http://api.openweathermap.org/data/2.5/weather?q=X} 
+}:
+
+\noindent
+\begin{equation*}
+\begin{array}{l}
+ \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC"}) \\
  \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
@@
 in \S\ref{sec:impl-hetero}) and provides a type with properties \ident{Record} for the former and 
-\ident{Array} for the latter. Second, the the field \ident{value} is \kvd{null} for some records. Third, 
+\ident{Array} for the latter. Second, the field \ident{value} is \kvd{null} for some records. Third, 
 numbers can be represented in JSON as numeric literals (without quotes), but here, they are 
@@
 
+{\color{red} TO BE DELETED}
+
 \vspace{-0.5em}
@@
 % option types
+\color{red}
 \inference[(nullable)\;]
@@
 \begin{equation*}
+\color{red}
 \hspace{4em}
@@
 \begin{equation*}
+\color{red}
 \hspace{2em}
@@
  \textnormal{\footnotesize{(prim)}}\;\; \kvd{int} \tsep \kvd{float} \vdash \kvd{float}
+\end{array}
+\end{equation*}
+
+\begin{equation*}
+\begin{array}{rcl}
+\ident{lub}(\sigma, \sigma) &=& \sigma \\
+\ident{lub}([\sigma_1], [\sigma_2]) &=& [\ident{lub}(\sigma_1, \sigma_2)]  \\
+\ident{lub}(\bot, \sigma) = \ident{lub}(\sigma, \bot) &=& \sigma \\
+\ident{lub}(\kvd{null}, \sigma) = \ident{lub}(\sigma, \kvd{null}) &=& \addopt{\sigma} \\
+\ident{lub}(\kvd{any}, \sigma) = \ident{lub}(\sigma, \kvd{any}) &=& \kvd{any} \\
+\ident{lub}(\kvd{float}, \kvd{int}) = \ident{lub}(\kvd{int}, \kvd{float}) &=& \kvd{float} \\
+\ident{lub}(\sigma_2, \kvd{nullable}\langl \hat{\sigma_1} \rangl) = \ident{lub}(\kvd{nullable}\langl \hat{\sigma_1} \rangl, \sigma_2) &=& \addopt{\ident{lub}(\hat{\sigma_1}, \sigma_2)} \\
+\ident{lub}(\nu \; \{ \nu_1 \!:\! \sigma_1,  \ldots \;, \nu_n \!:\! \sigma_n \}, 
+  \nu \; \{ \nu_1 \!:\! \sigma'_1, \; \ldots \;, \nu_n \!:\! \sigma'_n \}) &=& 
+  \nu \; \{ \nu_1 \!:\! \ident{lub}(\sigma_1, \sigma'_1), \; \ldots \; , \nu_n \!:\! \ident{lub}(\sigma_n, \sigma'_n) \}  \\[0.75em]
+\textnormal{Otherwise}~~~~ \ident{lub}(\sigma_1, \sigma_2) &=& \kvd{any} \\
+
 \end{array}
@@
  \textnormal{\footnotesize{(eq1)}}&
- v=v\reduce\kvd{true} \qquad \textnormal{\footnotesize{(eq2)}}~~v=v'\reduce\kvd{false}\\
+ v=v'\reduce\kvd{true} \qquad (\textnormal{when}~v = v')\\
+ \\[-0.3em]
+ \textnormal{\footnotesize{(eq2)}}&
+ v=v'\reduce\kvd{false} \qquad (\textnormal{when}~v \neq v')\\
  \\[-0.3em]
@@
 class definitions. We express it using the following mapping:
-
-\noindent
-\begin{equation*}
-\sem{-} : \sigma \rightarrow (\tau \times e \times L) \qquad (\textnormal{where}~L,\emptyset \vdash e:\ident{Data}\rightarrow\tau)
+%
+\begin{equation*}
+\sem{\sigma} = (\tau, e, L) \qquad (\textnormal{where}~L,\emptyset \vdash e:\ident{Data}\rightarrow\tau)
 \end{equation*}
@@
 The progress lemma states that evaluation of a well-typed program does not reach an undefined state. 
-This is not a problem for the (standard) ML \cite{sml} subset and object-oriented subset \cite{fwjava} of the calculus. The 
+This is not a problem for the Standard ML \cite{sml} subset and object-oriented subset \cite{fwjava} of the calculus. The 
 problematic part are the dynamic data operations (Figure~\ref{fig:op-conversions}). Given a data
@@
 
-We connect two lines of research: integration of external data into statically-typed 
-programming languages and inferring types for real-world data sources. We build on F\# type 
-providers \cite{fsharp-typeprov,fsharp-typeprov-ddfp,idris-tp,liteq}, but our paper is novel 
-in that it discusses the theory and safety of a concrete type provider.
+We connect two lines of research: the work: inferring types for real-world data sources
+and the integration of external data into statically-typed programming langauges.
+
+Early work on F\# Data happened together with the development of the type providers mechanism in F\# 
+\cite{fsharp-typeprov,fsharp-typeprov-ddfp}, but previous publications on type providers focused on 
+the general mechanism \cite{idris-tp,liteq} rather than on concrete type providers. This paper 
+is novel in that it discusses the theory and safety of a concrete type provider from the programming
+language perspective.
 

### 2016-02-04 pldi-2016/paper.tex
Final edits (section 1 and 2)

@@
-\documentclass[10pt,preprint,clearpagebib]{sigplanconf}
+\documentclass[10pt,clearpagebib]{sigplanconf}
 
@@
 
-% TODO: Discuss 'stability' of the inference (minor change in schema does not cause dramatic cahnges)
-% TODO: Say that variant is a last-resort thing
-% TODO: Practical evaluation (lots of downloads, lots of Github - how often people customize samples?)
-% TODO: Explain why we need option type (because T + null is a variant)
-% TODO: In the PLDI version, we delete all the unnecessary runtime conversions
-% but we should say that those happen for practical reasons (e.g. float -> int, or anyhing to string)
-
 \urlstyle{sf}
@@
 \setlength{\pdfpagewidth}{\paperwidth}
-\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country} 
-\copyrightyear{20yy} 
+\conferenceinfo{PLDI '16}{June 13--17, 2016, Santa Barbara, CA, United States} 
+\copyrightyear{2016} 
 \copyrightdata{978-1-nnnn-nnnn-n/yy/mm} 
@@
 
-In this paper, we integrate external structured data into F\# programming. As most real-world data
-does not come with an explicit schema, we develop a shape inference algorithm that infers a shape from 
-representative sample documents and integrates it into the F\# type system using type providers.
-We present a formalization and a relative type soundness theorem for a simplified version of this process.
+We present F\# Data, a library that integrates external structured data into F\#. As most real-world 
+data does not come with an explicit schema, we develop a shape inference algorithm that infers a 
+shape from representative sample documents. We then integrate the inferred shape into the F\# type 
+system using type providers. We formalize the process and prove a relative type soundness theorem.
 
@@
 root node must be a record with a \strf{main} field, which has to be another record containing
-a numerical \strf{temp} field. When the shape is different, the code simply fails with an exception. 
-While not immediately unsound, the code is manifestly prone to errors if strings are misspelled 
-or incorrect shape assumed.
+a numerical \strf{temp} field representing the current temperature. When the shape is different, 
+the code fails. While not immediately unsound, the code is prone to errors if strings are 
+misspelled or incorrect shape assumed.
 
@@
 \end{itemize}
-
-\vspace{-0.1em}
-\noindent
-The supplementary screencast illustrates the practical developer experience using 
-  F\# Data with JSON, XML and CSV.
 
@@
 \begin{itemize}
-\item The mechanism is predictable. The user directly works with the provided types and should 
-  understand why a specific type was produced from a given sample.\footnote{In particular, we do 
-  not use probabilistic methods where adding an additional sample could completely change the 
-  shape of the type.}
+\item The mechanism is predictable (\S\ref{sec:impl-stable}). The user directly works with the 
+  provided types and should understand why a specific type was produced from a given sample.
 
@@
 
+\noindent
+The supplementary screencast provides further illustration of the practical developer
+experience using F\# Data.
+
 % --------------------------------------------------------------------------------------------------
@@
  \kvd{type}~\ident{JsonValue} = \\[0.1em]
- \quad|~ \ident{Number}~\kvd{of}~\ident{float} ~~~~|~ \ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
- \quad|~ \ident{String}~\kvd{of}~\ident{string} ~~~~~\, |~ \ident{Null} \\[0.1em]
+ \quad|~ \ident{Number}~\kvd{of}~\ident{float} \\[0.1em]
+ \quad|~ \ident{Boolean}~\kvd{of}~\ident{bool} \\[0.1em]
+ \quad|~ \ident{String}~\kvd{of}~\ident{string}\\[0.1em]
  \quad|~ \ident{Record}~\kvd{of}~\ident{Map}\langl\ident{string}, \ident{JsonValue}\rangl \\[0.1em]
  \quad|~ \ident{Array}~\kvd{of}~\ident{JsonValue}[] \\[0.1em]
+ \quad|~ \ident{Null} \\[0.1em]
 \end{array}
@@
 \begin{verbatim}
-  [ { "name":"Jan", "age":25 }, { "name":"Tomas" },
+  [ { "name":"Jan", "age":25 }, 
+    { "name":"Tomas" },
     { "name":"Alexander", "age":3.5 } ]
@@
 sample values and is inferred as \ident{string}. The field \ident{Age} is marked as optional,
-because the value is missing in one sample. The two age values are an integer $25$ and a 
-float $3.5$ and so the common inferred type is \ident{float}.
+because the value is missing in one sample. In F\#, we can then use \ident{Option.iter} to call
+the specified function (printing) only when an optional value is available. The two age values 
+are an integer $25$ and a float $3.5$ and so the common inferred type is \ident{float}.
+The names of the properties are normalized to follow standard F\# naming conventions
+as discussed later (\S\ref{sec:impl-naming}).
 
@@
 \paragraph{Error handling.}
-In addition to the structure of the types, the type provider also specifies what code should be 
-executed at run-time in place of \ident{item.Name} and other operations. The runtime behaviour is 
+In addition to the structure of the types, the type provider also specifies the code of 
+operations such as \ident{item.Name}. The runtime behaviour is 
 the same as in the earlier hand-written sample (\S\ref{sec:introduction}) -- a member access 
@@
 %
-This provides access to the elements known statically from the sample. However the type is 
-deliberately `weak', because the user needs to explicitly handle the case when a value is not 
-a statically known element. The above code uses \ident{Option.iter} to skip all unknown elements.
+\ident{Element} is an abstract type with properties. It can represent the statically known
+elements, but it is not limited to them. For a table element, all three properties would return 
+\ident{None}.
+
+Using a type with optional properties provides access to the elements known statically from the 
+sample. However the user needs to explicitly handle the case when a value is not a statically 
+known element. In object-oriented languages, the same could be done by providing a class hierarchy, 
+but this loses the easy discoverability when ``.'' is typed.
 
@@
 
-\subsection{Summary}
+\subsection{Real-world JSON services}
 \label{sec:providers-sum}
-Throughout the introduction, we used data sets that demonstrate the typical problems that are frequent
+
+Throughout the introduction, we used data sets that demonstrate the typical problems frequent
 in the real-world (missing data, inconsistent encoding of primitive values and heterogeneous shapes). 
-The following JSON response with government debt information returned by the World Bank\footnote{Available at 
-\url{http://data.worldbank.org}} demonstrates all three problems:
+The government debt information returned by the World Bank\footnote{Available at 
+\url{http://data.worldbank.org}} demonstrates all three:
 %
@@
 \begin{verbatim}
-  [ { "page": 1, "pages": 5 },
+  [ { "pages": 5 },
     [ { "indicator": "GC.DOD.TOTL.GD.ZS",
@@
 \noindent
-First of all, the top-level element is a collection containing two values of different shape.
-The first is a record with meta-data about the current page and the second is an array with data. 
-The actual F\# Data implementation supports a concept of heterogeneous collections (briefly outlined
-in \S\ref{sec:impl-hetero}) and provides a type with properties \ident{Record} for the former and 
-\ident{Array} for the latter. Second, the field \ident{value} is \kvd{null} for some records. Third, 
-numbers can be represented in JSON as numeric literals (without quotes), but here, they are 
-returned as string literals instead.\footnote{This is often used to avoid non-standard numerical 
-types of JavaScript.}
+First, the field \ident{value} is \kvd{null} for some records. Second, numbers in JSON can be 
+represented as numeric literals (without quotes), but here, they are returned as string literals 
+instead.\footnote{This is often used to avoid non-standard numerical types of JavaScript.}
+Finally, the top-level element is a collection containing two values of different shape.
+The record contains meta-data with the total number of pages and the array contains the data.
+F\# Data supports a concept of heterogeneous collection (outlined in in \S\ref{sec:impl-hetero})
+and provides the following type:
+%
+\begin{equation*}
+\begin{array}{l}
+\kvd{type}~\ident{Record}~=  \\
+ \quad \kvd{member}~\ident{Pages}~:~\ident{int} \\[0.5em]
+\kvd{type}~\ident{Item}~=  \\
+ \quad \kvd{member}~\ident{Date}~:~\ident{int} \\
+ \quad \kvd{member}~\ident{Indicator}~:~\ident{string} \\
+ \quad \kvd{member}~\ident{Value}~:~\ident{option}\langl \ident{float}\rangl\\[0.5em]
+\kvd{type}~\ident{WorldBank}~=  \\
+ \quad \kvd{member}~\ident{Record}~:~\ident{Record}\\
+ \quad \kvd{member}~\ident{Array}~:~\ident{Item}[]\\[0.5em]
+\end{array}
+\end{equation*}
+%
+The inference for heterogeneous collections infers the multiplicities and shapes of nested 
+elements. As there is exactly one record and one array, the provided type \ident{WorldBank} exposes 
+them as properties \ident{Record} and \ident{Array}.
 
@@
 
-\section{Formalising type providers}
+\section{Formalizing type providers}
 \label{sec:formal}
@@
 
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Capitalization} 
+\label{sec:impl-naming}
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Predictability and stability} 
+\label{sec:impl-stable}
+
+In particular, we do 
+not use probabilistic methods where adding an additional sample could completely change the 
+shape of the type.
+
 % ==================================================================================================

### 2016-02-04 pldi-2016/paper.tex
Further edits - sections 3 and 4

@@
 \item We give a formal model (\S\ref{sec:formal}) and use it to prove
-  \emph{relativized type safety} for the type providers (\S\ref{sec:safety}).
+  \emph{relative type safety} for the type providers (\S\ref{sec:safety}).
 \end{itemize}
@@
 
+\subsection{Inferred shapes}
+\label{sec:inference-types}
+
+We distinguish between \emph{non-nullable shapes} that always have a valid value (written as 
+$\hat{\sigma}$) and \emph{nullable shapes} that encompass missing and \kvd{null} values 
+(written as $\sigma$). We write $\nu$ for record names and record field names. 
+%
+\begin{equation*}
+\begin{array}{rcl}
+ \hat{\sigma} &\narrow{=}& \nu \; \{ \nu_1 \!:\! \sigma_1, \ldots, \nu_n \!:\! \sigma_n,\; \rho_i  \} \\[0.1em]
+                &\narrow{|}& \ident{float} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
+ \\[0.6em] 
+       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{nullable}\langl \hat{\sigma} \rangl \lsep [\sigma] \lsep \kvd{any} \lsep \kvd{null}  \lsep ~\bot~
+\end{array}
+\end{equation*}
+
+\noindent
+Non-nullable shapes include records (consisting of a name and fields with their shapes) and 
+primitives. The row variables $\rho_i$ are discussed below. Names of records arising from XML are the names of the XML elements.
+For JSON records we always use a single name $\,\bullet$. We assume that record fields can be freely
+reordered.
+
+We include two numerical primitives, \ident{int} for integers and \ident{float} for floating-point 
+numbers. The two are related by the perference relation and we prefer \ident{int}.
+
+Any non-nullable shape $\hat{\sigma}$ can be wrapped as $\kvd{nullable}\langl\hat{\sigma}\rangl$ to 
+explicitly permit the \kvd{null} value. Type providers map \kvd{nullable} shapes to the F\# option
+type. A collection $[\sigma]$ is also nullable and \kvd{null} values are treated as empty 
+collections. This is motivated by the fact that a \kvd{null} collection is usually
+handled as an empty collection by client code. However there is a range of design alternatives (make collections
+non-nullable or treat \kvd{null} \ident{string} as an empty string).
+
+The shape $\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded 
+notation) and $\bot$ is the bottom shape. The \kvd{any} shape is the top shape, but we later add
+labels for statically known alternative shapes (\S\ref{sec:inference-vars}) as 
+discussed earlier (\S\ref{sec:providers-xml}).
+
+During inference we use row-variables $\rho_i$ \cite{rows-cardelli} in record shapes to represent 
+the flexibility arising from records in samples. For example, when a 
+record $\ident{Point}\,\{ \ident{x} \mapsto \num{3} \}$ occurs in a sample,
+it may be combined with $\ident{Point}\,\{\, \ident{x}\mapsto\num{3}, \ident{y}\mapsto\num{4} \}$ that contain more fields. The
+overall shape inferred must account for the fact that any extra fields are optional,
+giving an inferred shape $\ident{Point}\,\{ \ident{x}\!:\!\ident{int},~ \ident{y}\!:\!\kvd{nullable}\langl\ident{int}\rangl\}$.
+
+% --------------------------------------------------------------------------------------------------
+
 \begin{figure}
@@
 \label{fig:subtyping-diagram}
-\vspace{-0.5em}
 \end{figure}
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Inferred shapes}
-\label{sec:inference-types}
-
-We distinguish between \emph{non-nullable shapes} that always have a valid value (written as 
-$\hat{\sigma}$) and \emph{nullable shapes} that encompass missing and \kvd{null} values 
-(written as $\sigma$). We write $\nu$ for record names and record field names.
-
-\noindent
-\begin{equation*}
-\begin{array}{rcl}
- \hat{\sigma} &\narrow{=}& \nu \; \{ \nu_1 \!:\! \sigma_1, \ldots, \nu_n \!:\! \sigma_n,\; \rho_i  \} \\[0.1em]
-                &\narrow{|}& \ident{float} \lsep \ident{int} \lsep \ident{bool} \lsep \ident{string} 
- \\[0.6em] 
-       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{nullable}\langl \hat{\sigma} \rangl \lsep [\sigma] \lsep \kvd{any} \lsep \kvd{null}  \lsep ~\bot~
-\end{array}
-\end{equation*}
-
-\noindent
-Non-nullable shapes include records (consisting of a name and fields with their shapes) and 
-primitives. The row variables $\rho_i$ are discussed below. Names of records arising from XML are the names of the XML elements.
-For JSON records we always use a single name $\,\bullet$. We assume that record fields can be freely
-reordered.
-
-We include two numerical primitives, \ident{int} for integers and \ident{float} for floating-point 
-numbers. The two are related by the perference relation and we prefer \ident{int}.
-
-Any non-nullable shape $\hat{\sigma}$ can be wrapped as $\kvd{nullable}\langl\hat{\sigma}\rangl$ to 
-explicitly permit the \kvd{null} value. Type providers map \kvd{nullable} shapes to the F\# option
-type. A collection $[\sigma]$ is also nullable and \kvd{null} values are treated as empty 
-collections. The shape $\kvd{null}$ is inhabited by the $\kvd{null}$ value (using an overloaded 
-notation) and $\bot$ is the bottom shape. The \kvd{any} shape is the top shape, but we revisit it 
-later by adding labels for statically known alternative shapes (\S\ref{sec:inference-vars}) as 
-discussed earlier (\S\ref{sec:providers-xml}).
-
-During inference we use row-variables $\rho_i$ \cite{rows-cardelli} in record shapes to represent the flexibility
-arising from records in samples. For example, when a 
-record $\ident{Point}\,\{ \ident{x} \mapsto \num{3} \}$ occurs in a sample,
-it may be combined through inference with other
-samples such as $\ident{Point}\,\{\, \ident{x}\mapsto\num{3}, \ident{y}\mapsto\num{4} \}$ that contain more fields. The
-overall shape inferred must account for the fact that any extra fields are optional,
-giving an inferred shape $\ident{Point}\,\{ \ident{x}\!:\!\ident{int},~ \ident{y}\!:\!\kvd{option}\langl\ident{int}\rangl\}$.
-The process of solving $\rho_i$ variables is discussed in \S\ref{sec:formal-inferval}.
 
@@
 Figure~\ref{fig:subtyping-diagram} provides an intuition about the preference between 
-shapes. The upper part shows non-nullable shapes (with records and primitives) and the lower part 
+shapes. The lower part shows non-nullable shapes (with records and primitives) and the upper part 
 shows nullable shapes with \kvd{null}, collections and nullable shapes. In the diagram, we 
@@
 \begin{definition}
-For ground $\sigma_1$, $\sigma_2$ (i.e. without $\rho_i$ variables), we write $\sigma_1 \sqsupseteq \sigma_2$ to denote that $\sigma_2$ is preferred over $\sigma_1$. 
+For ground $\sigma_1$, $\sigma_2$ (i.e. without $\rho_i$ variables), we write $\sigma_1 \sqsubseteq \sigma_2$ to denote that $\sigma_1$ is preferred over $\sigma_2$. 
 The shape preference relation is defined as a transitive reflexive closure of the following rules:
@@
   \label{eq:sub-prim}\tag{P1}
-  \ident{float}\,&\sqsupseteq\,\ident{int}&\\[-0.2em]
+  \ident{int}\,&\sqsubseteq\,\ident{float}&\\[-0.2em]
   \label{eq:sub-null}\tag{P2}
-  \sigma &\sqsupseteq \kvd{null}  &(\textnormal{for}~\sigma \neq \hat{\sigma})  \\[-0.2em]
+  \kvd{null} &\sqsubseteq \sigma  &(\textnormal{for}~\sigma \neq \hat{\sigma})  \\[-0.2em]
   \label{eq:sub-opt}\tag{P3}
-  \kvd{nullable}\langl\hat{\sigma}\rangl &\sqsupseteq \hat{\sigma}  &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
+  \hat{\sigma} &\sqsubseteq \kvd{nullable}\langl\hat{\sigma}\rangl  &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
   \label{eq:sub-opt-cov}\tag{P4}
-  \kvd{nullable}\langl\hat{\sigma_1}\rangl &\sqsupseteq 
-    \kvd{nullable}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} \sqsupseteq \hat{\sigma_2})\\[-0.2em]
+  \kvd{nullable}\langl\hat{\sigma_1}\rangl &\sqsubseteq 
+    \kvd{nullable}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} \sqsubseteq \hat{\sigma_2})\\[-0.2em]
   \label{eq:sub-col}\tag{P5}
-  [\sigma_1] &\sqsupseteq [\sigma_2]  &(\textnormal{if}~\sigma_1 \sqsupseteq \sigma_2) \\[-0.2em]
+  [\sigma_1] &\sqsubseteq [\sigma_2]  &(\textnormal{if}~\sigma_1 \sqsubseteq \sigma_2) \\[-0.2em]
   \label{eq:sub-bot}\tag{P6}
-  \sigma &\sqsupseteq \bot  &(\textnormal{for all}~\sigma)\\[-0.2em]
+  \bot &\sqsubseteq \sigma  &(\textnormal{for all}~\sigma)\\[-0.2em]
   \label{eq:sub-var-top}\tag{P7}
-  \kvd{any} &\sqsupseteq \sigma 
+  \sigma &\sqsubseteq \kvd{any}
 \end{align}
@@
 \begin{array}{l}
- \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsupseteq \\
+ \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsubseteq \\
  \quad\nu~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
-\end{array} \qquad ~~~~(\textnormal{if}~\sigma_i \sqsupseteq \sigma_i')
+\end{array} \qquad ~~~~(\textnormal{if}~\sigma_i \sqsubseteq \sigma_i')
 \end{align}
@@
 \begin{array}{l}
- \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsupseteq \\
+ \nu~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \sqsubseteq \\
  \quad\nu~\{ \nu_1\!:\!\sigma_1, .., .., \nu_m\!:\!\sigma_m \}
-\end{array} \quad~~\, (\textnormal{when}~m \geq n)
+\end{array} \quad~~\, (\textnormal{when}~m \leq n)
 \end{align}
@@
 \begin{equation*}
-\begin{array}{rclcl}
- \tytag &\narrow{=}& \ident{collection}  &\narrow{\lsep}& \ident{number} \\
-        &\narrow{\lsep}& \ident{nullable} &\narrow{\lsep}& \ident{string}  \\
-        &\narrow{\lsep}& \nu ~\lsep \ident{any} &\narrow{\lsep}& \ident{bool}
-\end{array}
-%
-\quad\;\;\;
-%
-\begin{array}{rcl}
- \tytagof(\ident{string}) &\narrow{=}& \ident{string}\\
- \tytagof(\ident{bool}) &\narrow{=}& \ident{bool}\\
- \tytagof(\ident{int}) &\narrow{=}& \ident{number}\\
- \tytagof(\ident{float}) &\narrow{=}& \ident{number}\\
-\end{array}
-%
-\;\;
-%
-\begin{array}{rcl}
- \tytagof(\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl) &\narrow{=}& \ident{any}\\
- \tytagof(\nu\; \{ \nu_1 : \sigma_1, \; \ldots \; , \nu_n : \sigma_n \}) &\narrow{=}& \nu \\
- \tytagof(\kvd{nullable}\langl\hat{\sigma}\rangl) &\narrow{=}& \ident{nullable}\\
- \tytagof([\sigma]) &\narrow{=}& \ident{collection}\\
-\end{array}
-\end{equation*}
-
-\vspace{-0.5em}
-\begin{equation*}
-\quad\qquad\qquad
+\begin{array}{rcll}
+\ident{csh}(\sigma, \sigma) &=& \sigma & (\emph{eq})\\
+\ident{csh}([\sigma_1], [\sigma_2]) &=& [\ident{csh}(\sigma_1, \sigma_2)] & (\emph{list}) \\
+\ident{csh}(\bot, \sigma) = \ident{csh}(\sigma, \bot) &=& \sigma & (\emph{bot})\\
+\ident{csh}(\kvd{null}, \sigma) = \ident{csh}(\sigma, \kvd{null}) &=& \addopt{\sigma}& (\emph{null}) \\
+\ident{csh}(\kvd{any}, \sigma) = \ident{csh}(\sigma, \kvd{any}) &=& \kvd{any} & (\emph{top})\\
+\ident{csh}(\kvd{float}, \kvd{int}) = \ident{csh}(\kvd{int}, \kvd{float}) &=& \kvd{float} & (\emph{num})\\
+\ident{csh}(\sigma_2, \kvd{nullable}\langl \hat{\sigma_1} \rangl) = \ident{csh}(\kvd{nullable}\langl \hat{\sigma_1} \rangl, \sigma_2) &=& \addopt{\ident{csh}(\hat{\sigma_1}, \sigma_2)} & (\emph{opt})\\
+\ident{csh}(\nu \; \{ \nu_1 \!:\! \sigma_1,  \ldots \;, \nu_n \!:\! \sigma_n \}, 
+  \nu \; \{ \nu_1 \!:\! \sigma'_1, \; \ldots \;, \nu_n \!:\! \sigma'_n \}) &=& 
+  \nu \; \{ \nu_1 \!:\! \ident{csh}(\sigma_1, \sigma'_1), \; \ldots \; , \nu_n \!:\! \ident{csh}(\sigma_n, \sigma'_n) \} & (\emph{recd}) \\
+\ident{csh}(\sigma_1, \sigma_2) &=& \kvd{any} & (\emph{any})\\
+\end{array}
+\end{equation*}
+%
+\begin{equation*}
+\quad\qquad
 \begin{array}{rcll}
@@
 
-{\color{red} TO BE DELETED}
-
-\vspace{-0.5em}
-\begin{equation*}
-% option types
-\color{red}
-\inference[(nullable)\;]
-  {\hat{\sigma_1} \tsep \sigma_2 \vdash \sigma}
-  {\kvd{nullable}\langl\hat{\sigma_1}\rangl \tsep \sigma_2 \vdash \kvd{nullable}\langl\dropopt{\sigma}\rangl}
-\qquad
-\inference[(any)\;]
-  { \tytagof(\sigma_1) \neq \tytagof(\sigma_2) & \tytagof(\sigma_1) \neq \ident{nullable} \neq \tytagof{\sigma_2}}
-  { \sigma_1 \tsep \sigma_2 \vdash \kvd{any} }  
-\end{equation*}
-\vspace{-2em}
-
-\begin{equation*}
-\color{red}
-\hspace{4em}
-\inference[(record)\;]
-  { \forall i\in\{ 1 .. n \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
-  { \begin{array}{l}
-    \nu \; \{ \nu_1 \!:\! \sigma_1,  \ldots \;, \nu_n \!:\! \sigma_n \} \tsep
-    \nu \; \{ \nu'_1 \!:\! \sigma'_1, \; \ldots \;, \nu'_m \!:\! \sigma'_m \} \vdash
-    \nu \; \{ \nu_1 \!:\! \sigma''_1, \; \ldots \; , \nu_k \!:\! \sigma''_n \}
-    \end{array} }
-\end{equation*}
-\vspace{-2em}
-
-% all rules are symmetric and reflexive
-\begin{equation*}
-\color{red}
-\hspace{2em}
-\inference[(list)\;]
-  {\sigma_1 \tsep \sigma_2 \vdash \sigma}
-  {[\sigma_1] \tsep [\sigma_2] \vdash [\sigma]}
-\quad
-\inference[(sym)\;]
-  {\sigma_1 \tsep \sigma_2 \vdash \sigma}
-  {\sigma_2 \tsep \sigma_1 \vdash \sigma}
-%
-\quad
-\begin{array}{l}
- \textnormal{\footnotesize{(refl)}}\;\; \sigma \tsep \sigma \vdash \sigma\\[0.6em]
- % bottom type
- \textnormal{\footnotesize{(bot)}}\;\; \bot \tsep \sigma \vdash \sigma
-\end{array}
-%
-\quad
-% null and primitives
-\begin{array}{l}
- \textnormal{\footnotesize{(null)}}\;\; \sigma \tsep \kvd{null} \vdash \addopt{\sigma} \quad (\sigma \neq \bot)\\[0.6em]
- \textnormal{\footnotesize{(prim)}}\;\; \kvd{int} \tsep \kvd{float} \vdash \kvd{float}
-\end{array}
-\end{equation*}
-
-\begin{equation*}
-\begin{array}{rcl}
-\ident{lub}(\sigma, \sigma) &=& \sigma \\
-\ident{lub}([\sigma_1], [\sigma_2]) &=& [\ident{lub}(\sigma_1, \sigma_2)]  \\
-\ident{lub}(\bot, \sigma) = \ident{lub}(\sigma, \bot) &=& \sigma \\
-\ident{lub}(\kvd{null}, \sigma) = \ident{lub}(\sigma, \kvd{null}) &=& \addopt{\sigma} \\
-\ident{lub}(\kvd{any}, \sigma) = \ident{lub}(\sigma, \kvd{any}) &=& \kvd{any} \\
-\ident{lub}(\kvd{float}, \kvd{int}) = \ident{lub}(\kvd{int}, \kvd{float}) &=& \kvd{float} \\
-\ident{lub}(\sigma_2, \kvd{nullable}\langl \hat{\sigma_1} \rangl) = \ident{lub}(\kvd{nullable}\langl \hat{\sigma_1} \rangl, \sigma_2) &=& \addopt{\ident{lub}(\hat{\sigma_1}, \sigma_2)} \\
-\ident{lub}(\nu \; \{ \nu_1 \!:\! \sigma_1,  \ldots \;, \nu_n \!:\! \sigma_n \}, 
-  \nu \; \{ \nu_1 \!:\! \sigma'_1, \; \ldots \;, \nu_n \!:\! \sigma'_n \}) &=& 
-  \nu \; \{ \nu_1 \!:\! \ident{lub}(\sigma_1, \sigma'_1), \; \ldots \; , \nu_n \!:\! \ident{lub}(\sigma_n, \sigma'_n) \}  \\[0.75em]
-\textnormal{Otherwise}~~~~ \ident{lub}(\sigma_1, \sigma_2) &=& \kvd{any} \\
-
-\end{array}
-\end{equation*}
-
 \caption{The common preferred shape relation}
@@
 
-Given two ground shapes, the \emph{common preferred shape} relation finds a least upper bound
-(Lemma~\ref{thm:lub}). It prefers records, which is important for usability as discussed
-earlier (\S\ref{sec:providers-xml}).
+Given two ground shapes, the \emph{common preferred shape} is the least upper bound of the
+shape with respect to the preferred shape relation. The least upper bound prefers records, 
+which is important for usability as discussed earlier (\S\ref{sec:providers-xml}).
 
 \begin{definition}
-A \emph{common preferred shape} of two ground shapes $\sigma_1$ and $\sigma_2$ is a shape $\sigma$, written 
-$\sigma_1 \tsep \sigma_2 \vdash \sigma$, by the rules in 
-Figure~\ref{fig:subtyping-cst}.
+A \emph{common preferred shape} of two ground shapes $\sigma_1$ and $\sigma_2$ is a shape 
+$\ident{csh}(\sigma_1, \sigma_2)$ obtained according to Figure~\ref{fig:subtyping-cst}. 
+The rules are matched from top to bottom.
 \end{definition}
@@
 \noindent
-We define shape \emph{tags} to identify shapes that have a common preferred shape which is not the
-top shape. When finding a common shape of two records (\emph{record}), the shape of fields becomes the common preferred shape of their 
-respective shapes. The field labels of two record types can always be made to match,
-through the shape inference from sample (\S\ref{sec:formal-inferval}),
-\emph{before} finding a common preferred shape using $\triangledown$.
-
-We can find a common shape of two different numbers (\emph{prim}); for two collections,
-we combine their elements (\emph{list}). When one shape is nullable (\emph{nullable}), we find
-the common non-nullable shape and wrap it in \kvd{nullable}; if one of the shapes is \kvd{null}, 
-we make the other one nullable (\emph{null}). Finally, if the two shapes do not have matching tags 
-(and they are not nullable), we have to infer the \kvd{any} shape as there is no better alternative. 
-The remaining rules for reflexivity, symmetry and the bottom shape are standard.
-
-\paragraph{Properties.}
-The partially ordered set of ground shapes has a unique least upper bound.
-The $\triangledown$ relation defines a function (Lemma~\ref{thm:func}) that finds 
-the least upper bound (Lemma~\ref{thm:lub}).
-
-\begin{lemma}[Preferred shape function]
-\label{thm:func}
-For all ground $\sigma_1$ and $\sigma_2$ there exists exactly one $\sigma$ such that 
-$\sigma_1 \tsep \sigma_2 \vdash \sigma$. 
-\end{lemma}
-\begin{proof}
-The pre-conditions of rules in Figure~\ref{fig:subtyping-cst} are disjoint, with the exception 
-of (\emph{sym}) and (\emph{refl}), but applying those does not affect the result. Reflexivity is
-preserved recursively and symmetry does not enable different derivations.
-\end{proof}
+The fact that the rules of \ident{csh} are matched from top to bottom resolves the ambiguity 
+between certain rules. Most importantly (\emph{any}) is used only as the last resort.
+When finding a common shape of two records (\emph{recd}) we find common preferred shapes of their 
+respective fields. The field labels of two record types can always be made to match, through the 
+shape inference from sample (\S\ref{sec:formal-inferval}), \emph{before} finding a common preferred 
+shape using $\ident{csh}$.
+
+We can find a common shape of two different numbers (\emph{num}); for two collections,
+we combine their elements (\emph{list}). When one shape is nullable (\emph{opt}), we find
+the common non-nullable shape and ensure the result is nullable using $\addopt{-}$, which
+is also applied when one of the shapes is \kvd{null} (\emph{null}).
+
+The \ident{csh} function finds the unique least upper bound of the partially ordered 
+set of ground shapes (Lemma~\ref{thm:lub}).
 
@@
 \end{lemma}
-\begin{proof}
-By induction over $\vdash$. Note that the algorithm only infers the top shape \kvd{any} when for non-nullable
-shapes of distinct tags and so no better preferred shape exists.
-\end{proof}
-
-\subsection{Inferring shapes from samples}
-\label{sec:formal-inferval}
-
-We now specify how we obtain the shape from data. As clrified later (\S\ref{sec:impl-parsing}), 
-we represent JSON, XML and CSV documents using the same first-order \emph{data} value:
-%
-\begin{equation*}
-\begin{array}{lcl}
- d &\narrow{=}& i \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
-   &\narrow{|}& [d_1; \ldots; d_n] \lsep \nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}
-\end{array}
-\end{equation*}
-%
-The definition includes primitive values ($i$ for integers, $f$ for floats
-and $s$ for strings) and \kvd{null}. A collection is written as a 
-list of values in square brackets. A record starts with a name $\nu$, followed by a 
-sequence of field assignments $\nu_i \mapsto d_i$.
-
-Figure~\ref{fig:shape-inference} defines a mapping $\semalt{d_1,\ldots,d_n}$ which turns a collection of sample 
-data $d_1, \ldots, d_n$ into a shape $\sigma$. Before applying $S$, we assume each record in each $d_i$ is marked with a fresh row inference
-variable $\rho_i$.  
-We then choose a ground, minimal substitution $\theta$ for row variables.
-Because $\rho_i$ variables represent potentially missing fields,
-the $\addopt{-}$ operator from Figure~\ref{fig:subtyping-cst} is applied to all types
-in the vector. 
-This is sufficient to equate the record field labels and satisfy the side conditions in rule (record) when multiple record samples are merged.
-In practice, $\theta$ is found via row variable unification \cite{rows-remy}. We omit the details here.
-No $\rho_i$ variables remain after inference as the substitution chosen is ground.
-
-Primitive  values are mapped to their corresponding shapes. 
-When infering a shape from multiple samples, we use the common preferred shape relation to find a 
-common shape for all values (starting with $\bot$). This operation is used  
-when calling a type provider with multiple samples and also when inferring the shape of collection
-elements.
 
@@
 
-\begin{figure*}
-  \noindent
-  \begin{equation*}
-  \textnormal{\footnotesize{(top-1)}}\;\;
-  \inference
-    {\exists i . \tytagof(\sigma_i) = \tytagof(\dropopt{\sigma}) & \sigma \tsep \sigma_i \vdash \sigma_i' & \tytagof(\sigma)\neq\ident{any}}
-    {\sigma \tsep \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \vdash 
-      \kvd{any}\langl\sigma_1, \ldots, \dropopt{\sigma_i'}, \ldots, \sigma_n\rangl}
-  \;\;
-  \inference
-    {\nexists i . \tytagof(\sigma_i) = \tytagof(\dropopt{\sigma}) & \tytagof(\sigma)\neq\ident{any}}
-    {\sigma \tsep \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl \vdash 
-      \kvd{any}\langl\sigma_1, \ldots, \sigma_n, \dropopt{\sigma}\rangl}
-  \end{equation*}
-  \vspace{-2em}
-
-  % two union types
-  \begin{equation*}
-  \begin{array}{l}
-  \inference[(top-2)\;]
-    { (\tytagof(\sigma_i) = \tytagof(\sigma'_j)) \Leftrightarrow (i = j) \wedge (i \leq k)\\
-        \forall i\in\{ 1 .. k \}.(\sigma_i \tsep \sigma'_i \vdash \sigma''_i) }
-    { \begin{array}{l}
-      \kvd{any}\langl \sigma_1, \ldots, \sigma_k,  \ldots, \sigma_n\rangl \tsep 
-      \kvd{any}\langl \sigma'_1, \ldots, \sigma'_k, \ldots, \sigma'_m\rangl \vdash\\
-      \kvd{any}\langl \sigma''_1, \ldots, \sigma''_k, \sigma_{k+1}, \ldots, \sigma_{n}, \sigma'_{k+1}, \ldots, \sigma'_{m}\rangl
-      \end{array} }
-  \end{array}    
-  \begin{array}{l}
-  \inference[(top-3)\;]
-    {(\forall i\in\{1,2\}) & \tytagof(\sigma_1) \neq \tytagof(\sigma_2) \\ 
-     \tytagof(\sigma_i) \neq \ident{any} & \nexists\sigma_i' . (\sigma_i = \kvd{option}\langl\sigma_i'\rangl)  }
-    {\sigma_1 \tsep \sigma_2 \vdash \kvd{any}\langle\dropopt{\sigma_1}, \dropopt{\sigma_2}\rangle}
-  \end{array}  
-  \end{equation*}
-  \caption{Extending the common preferred shape relation for labelled top shapes}
-  \label{fig:subtyping-cst-var}
-\end{figure*}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Adding labelled top shapes}
-\label{sec:inference-vars}
-
-When analyzing the structure of shapes, it suffices to consider a single top shape \kvd{any}.
-However, the type providers need more information to provide typed access to the possible 
-alternative shapes of data, such as XML nodes.
-
 \begin{figure}[!h]
+\vspace{0em}  
 \noindent
@@
  \qquad\sigma_0 = \bot,~ \forall i\in \{ 1.. n \}.~ \sigma_{i-1} \triangledown \semalt{d_i} \vdash \sigma_i \\[0.6em]
- \textnormal{Choose minimal}\ \theta\ \textnormal{by ordering}\ \sqsupseteq\ \textnormal{lifted over substitutions} \\[0.1em]
+ \textnormal{Choose minimal}\ \theta\ \textnormal{by ordering}\ \sqsubseteq\ \textnormal{lifted over substitutions} \\[0.1em]
 \end{array}
@@
 
-We extend the core model (sufficient for the relativized safety) 
-with \emph{labelled top shapes} defined as:
+% --------------------------------------------------------------------------------------------------
+
+\begin{proof}
+By induction over the definition of $\ident{csh}$. Note that the algorithm only infers the top shape 
+\kvd{any} when on of the shapes is the top shape (\emph{top}) or when there is no other option 
+(\emph{any}); a nullable shape is introduced in $\addopt{-}$ only when no non-nullable shape can
+be used (\emph{null}), (\emph{opt}).
+\end{proof}
+
+\subsection{Inferring shapes from samples}
+\label{sec:formal-inferval}
+
+We now specify how we obtain the shape from data. As clrified later (\S\ref{sec:impl-parsing}), 
+we represent JSON, XML and CSV documents using the same first-order \emph{data} value:
+%
+\begin{equation*}
+\begin{array}{lcl}
+ d &\narrow{=}& i \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
+   &\narrow{|}& [d_1; \ldots; d_n] \lsep \nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}
+\end{array}
+\end{equation*}
+%
+The definition includes primitive values ($i$ for integers, $f$ for floats
+and $s$ for strings) and \kvd{null}. A collection is written as a 
+list of values in square brackets. A record starts with a name $\nu$, followed by a 
+sequence of field assignments $\nu_i \mapsto d_i$.
+
+Figure~\ref{fig:shape-inference} defines a mapping $\semalt{d_1,\ldots,d_n}$ which turns a 
+collection of sample data $d_1, \ldots, d_n$ into a shape $\sigma$. Before applying $S$, we assume 
+each record in each $d_i$ is marked with a fresh row inference variable $\rho_i$.
+
+We then choose a ground, minimal, global substitution $\theta$ for row variables. The substitution
+is minimal in the sense that it introduces as few new fields as possible, but no additional fields, 
+and the newly introduced fields have the most specific shape. Because $\rho_i$ variables represent 
+potentially missing fields, the $\addopt{-}$ operator from Figure~\ref{fig:subtyping-cst} is applied 
+to all types in the vector.
+
+This is sufficient to equate the record field labels and satisfy the pre-conditions in rule 
+(\emph{recd}) when multiple record samples are merged. In practice, $\theta$ is found via row 
+variable unification \cite{rows-remy}. We omit the details here. No $\rho_i$ variables remain after 
+inference as the substitution chosen is ground.
+
+Primitive  values are mapped to their corresponding shapes. 
+When infering a shape from multiple samples, we use the common preferred shape relation to find a 
+common shape for all values (starting with $\bot$). This operation is used  
+when calling a type provider with multiple samples and also when inferring the shape of collection
+elements.
+
+% --------------------------------------------------------------------------------------------------
+
+\begin{figure*}
+  \noindent
+  \begin{equation*}
+  \begin{array}{rclcl}
+   \tytag &\narrow{=}& \ident{collection}  &\narrow{\lsep}& \ident{number} \\
+          &\narrow{\lsep}& \ident{nullable} &\narrow{\lsep}& \ident{string}  \\
+          &\narrow{\lsep}& \nu ~\lsep \ident{any} &\narrow{\lsep}& \ident{bool}
+  \end{array}
+  %
+  \quad\;\;\;
+  %
+  \begin{array}{rcl}
+   \tytagof(\ident{string}) &\narrow{=}& \ident{string}\\
+   \tytagof(\ident{bool}) &\narrow{=}& \ident{bool}\\
+   \tytagof(\ident{int}) &\narrow{=}& \ident{number}\\
+   \tytagof(\ident{float}) &\narrow{=}& \ident{number}\\
+  \end{array}
+  %
+  \;\;
+  %
+  \begin{array}{rcl}
+   \tytagof(\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl) &\narrow{=}& \ident{any}\\
+   \tytagof(\nu\; \{ \nu_1 : \sigma_1, \; \ldots \; , \nu_n : \sigma_n \}) &\narrow{=}& \nu \\
+   \tytagof(\kvd{nullable}\langl\hat{\sigma}\rangl) &\narrow{=}& \ident{nullable}\\
+   \tytagof([\sigma]) &\narrow{=}& \ident{collection}\\
+  \end{array}
+  \end{equation*}
+
+  \vspace{-0.5em}
+
+  \begin{equation*}
+  \qquad\qquad
+  \begin{array}{ll}
+  \begin{array}{l}
+    \ident{csh}(\kvd{any}\langl \sigma_1, \ldots, \sigma_k,  \ldots, \sigma_n\rangl, 
+      \kvd{any}\langl \sigma'_1, \ldots, \sigma'_k, \ldots, \sigma'_m\rangl) =\\
+      \qquad \kvd{any}\langl \ident{csh}(\sigma_1, \sigma_1'), \ldots, \ident{csh}(\sigma_k, \sigma_k'), 
+        \sigma_{k+1}, \ldots, \sigma_{n}, \sigma'_{k+1}, \ldots, \sigma'_{m}\rangl \\[0.5em]
+      \qquad \textnormal{For $i,j$ such that }(\tytagof(\sigma_i) = \tytagof(\sigma'_j)) \Leftrightarrow (i = j) \wedge (i \leq k)
+  \end{array}\quad~
+  &  (\emph{top-merge}) 
+  \\[2.5em]    
+  \begin{array}{l}
+    \ident{csh}(\sigma, \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl) = 
+    \ident{csh}(\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl, \sigma) = \\
+    \qquad \kvd{any}\langl\sigma_1, \ldots, \dropopt{\ident{csh}(\sigma, \sigma_i)}, \ldots, \sigma_n\rangl\\[0.5em]
+    \qquad \textnormal{For $i$ such that }\tytagof(\sigma_i) = \tytagof(\dropopt{\sigma})
+  \end{array}
+  &  (\emph{top-incl}) 
+  \\[2.5em] 
+  ~\;\ident{csh}(\sigma, \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl) =
+   \kvd{any}\langl\sigma_1, \ldots, \sigma_n, \dropopt{\sigma}\rangl
+   & (\emph{top-add})\\[0.5em]    
+  ~\;\ident{csh}(\sigma_1,\sigma_2) =
+  \kvd{any}\langle\dropopt{\sigma_1}, \dropopt{\sigma_2}\rangle
+  & (\emph{top-any})
+ \end{array}
+\end{equation*}
+
+\caption{Extending the common preferred shape relation for labelled top shapes}
+\label{fig:subtyping-cst-var}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Adding labelled top shapes}
+\label{sec:inference-vars}
+
+When analyzing the structure of shapes, it suffices to consider a single top shape \kvd{any}.
+The type providers need more information to provide typed access to the possible 
+alternative shapes of data, such as XML nodes.
+
+We extend the core model (sufficient for the discussion of relative safety) with \emph{labelled top 
+shapes} defined as:
 % 
@@
 preferred shape relation and $\kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl$ should still be
-seen as the top shape, regardless of the labels.
-
-The common preferred shape relation is extended to find a labelled top shape that best represents 
-the sample. We limit the number of labels and avoid nesting by grouping shapes by the tag 
-introduced earlier.
-Rather than inferring $\kvd{any}\langl\ident{int}, \kvd{any}\langl\ident{bool}, \ident{float}\rangl\rangl$, 
+seen as the top shape, regardless of the labels\footnote{An alternative would be to add unions of 
+shapes, but doing so in a way that is compatible with the open-world assumption breaks the 
+existence of unique lower bound of the preferred shape relation.}.
+
+The common preferred shape function is extended to find a labelled top shape that best represents 
+the sample. The new rules for \kvd{any} appear in Figure~\ref{fig:subtyping-cst-var}. 
+We define shape \emph{tags} to identify shapes that have a common preferred shape 
+which is not the top shape. We use it to limit the number of labels and avoid nesting by grouping 
+shapes by the shape tag. Rather than inferring $\kvd{any}\langl\ident{int}, \kvd{any}\langl\ident{bool}, \ident{float}\rangl\rangl$, 
 our algorithm joins $\ident{int}$ and $\ident{float}$ and produces 
@@
 
-The new rules for \kvd{any} appear in Figure~\ref{fig:subtyping-cst-var}. When 
-combining a top with another shape (\emph{top-1}), the labels may or may not already 
-contain a case with the tag of the other shape. If it does, the two shapes are combined, otherwise 
-a new case is added. When combining two top shapes (\emph{top-2}), we group the labels that have 
-shared tags. Finally, (\emph{top-3}) combines two distinct non-top 
-shapes. As top shapes implicitly permit \kvd{null} values, we make the labels non-nullable 
-using $\dropopt{-}$.
+When combining two top shapes (\emph{top-merge}), we group the annotations by their tags.
+When combining a top with another shape, the labels may or may not already contain a case with 
+the tag of the other shape. If they doe, the two shapes are combined (\emph{top-incl}), otherwise 
+a new case is added (\emph{top-add}). Finally, (\emph{top-all}) replaces earlier (\emph{any}) 
+and combines two distinct non-top shapes. As top shapes implicitly permit \kvd{null} values,
+we make the labels non-nullable using $\dropopt{-}$.
 
@@
 
+Stating properties of the labels requires refinements to the \emph{preferred shape} relation.
+We leave the details to future work, but we
+note that the algorithm infers the best labels in the sense that there are labels that enable 
+typed access to every possible value in the sample, but not more. The same is the case for nullable
+fields of records.
+
+% ==================================================================================================
+% 
+%    #######
+%    #        ####  #####  #    #   ##   #      # ######   ##   ##### #  ####  #    #
+%    #       #    # #    # ##  ##  #  #  #      #     #   #  #    #   # #    # ##   #
+%    #####   #    # #    # # ## # #    # #      #    #   #    #   #   # #    # # #  #
+%    #       #    # #####  #    # ###### #      #   #    ######   #   # #    # #  # #
+%    #       #    # #   #  #    # #    # #      #  #     #    #   #   # #    # #   ##
+%    #        ####  #    # #    # #    # ###### # ###### #    #   #   #  ####  #    #
+%
+% ==================================================================================================
+
+\section{Formalizing type providers}
+\label{sec:formal}
+
+This section presents the formal model of F\# Data integration. To represent the programming 
+language that hosts the type provider, we introduce the Foo calculus, a subset of F\# with objects 
+and properties, extended with operations for working with weakly typed structured data along the 
+lines of the F\# Data runtime. Finally, we describe how 
+type providers turn inferred shapes into Foo classes (\S\ref{sec:formal-tp}).
+
 % -------------------------------------------------------------------------------------------------
@@
 \begin{figure}[!h]
+\vspace{-0.1em}
 \noindent
@@
 
-Stating properties of the labels requires refinements to the \emph{preferred shape} relation.
-We leave the details to future work, but we
-note that the algorithm infers the best labels in the sense that there are labels that enable 
-typed access to every possible value in the sample, but not more. The same is the case for nullable
-fields of records.
-
-% ==================================================================================================
-% 
-%    #######
-%    #        ####  #####  #    #   ##   #      # ######   ##   ##### #  ####  #    #
-%    #       #    # #    # ##  ##  #  #  #      #     #   #  #    #   # #    # ##   #
-%    #####   #    # #    # # ## # #    # #      #    #   #    #   #   # #    # # #  #
-%    #       #    # #####  #    # ###### #      #   #    ######   #   # #    # #  # #
-%    #       #    # #   #  #    # #    # #      #  #     #    #   #   # #    # #   ##
-%    #        ####  #    # #    # #    # ###### # ###### #    #   #   #  ####  #    #
-%
-% ==================================================================================================
-
-\section{Formalizing type providers}
-\label{sec:formal}
-
-This section presents the formal model of type providers for structured data. 
-To represent the programming language that hosts the type provider and the
-provided declarations, we introduce the Foo calculus,
-a subset of F\# with objects and properties, extended with
-operations for working with weakly typed structured data along the lines of the F\# Data runtime. Finally, we describe how 
-type providers turn inferred shapes into Foo classes (\S\ref{sec:formal-tp}).
-
-Type providers for structured data map the ``dirty'' world of weakly typed 
-structured data into a ``nice'' world of strong types. To model this, the Foo calculus
-does not have \kvd{null} values and data values $d$ are never directly exposed.  
-Furthermore Foo is simply typed: despite using class types and object notation 
-for notational convenience, it has no subtyping.
-
-% -------------------------------------------------------------------------------------------------
-
 \begin{figure*}
@@
 \begin{array}{l}
+\\[-2.5em]
 \ident{convFloat}(\ident{float}, i) \reduce f~(f=i) \\
@@
 
+% -------------------------------------------------------------------------------------------------
+
+Type providers for structured data map the ``dirty'' world of weakly typed 
+structured data into a ``nice'' world of strong types. To model this, the Foo calculus
+does not have \kvd{null} values and data values $d$ are never directly exposed.  
+Furthermore Foo is simply typed: despite using class types and object notation 
+for notational convenience, it has no subtyping.
+
 % --------------------------------------------------------------------------------------------------
@@
 The behaviour of the dynamic data operations is defined by the reduction rules in 
-Figure~\ref{fig:op-conversions}. The operations can convert data values into values of 
-less preferred shape. For example, \ident{convFloat} turns an integer \num{1} into 
-a floating-point \num{1.0} and \ident{convElements} turns \kvd{null} value into an empty list.
-
-All of the dynamic data operations take a data value $d$ and produce a Foo value. We do not 
-make the conversion explicit for primitive values (\ident{convPrim}) and assume that primitive
-values are shared between \emph{data} and Foo values, but we convert all other values. For example \ident{convElements} turns a 
-data collection $[d_1, \ldots, d_n]$ into a Foo list $v_1 :: .. :: v_n :: \kvd{nil}$.
-
-The data operations for collections, record fields and nullable values all take a function
-as the last argument and invoke it to convert the obtained data. For example, \ident{convField} 
-accesses a record field -- it ensures that the actual record name matches the expected name and 
-passes the field value (or \kvd{null} data value) to the provided function.
-
-Finally, we also define a runtime shape test \ident{hasShape}.
-For records, we have to check that for each field $\nu_1, \ldots, \nu_n$ 
-in the record, the actual record value has a field of the same name with a matching shape. The last 
-line defines a ``catch all'' pattern, which returns \kvd{false} for all remaining cases. 
-We treat $e_1 \vee e_2$ and $e_1 \wedge e_2$ as a syntactic sugar for \kvd{if}\,.\,.\,\kvd{then}\,.\,.\,\kvd{else} so
-the result of the reduction is just a Foo expression.
+Figure~\ref{fig:op-conversions}. The typing is shown in Figure~\ref{fig:ff-typecheck} and is 
+discussed later. The \ident{hasShape} function represents a runtime shape test.
+It checks whether a \ident{Data} value $d$ (Section~\ref{sec:formal-inferval}) passed as the second
+argument has a shape specified by the first argument. For records, we have to check that for each 
+field $\nu_1, \ldots, \nu_n$ in the record, the actual record value has a field of the same name 
+with a matching shape. The last line defines a ``catch all'' pattern, which returns \kvd{false} 
+for all remaining cases. We treat $e_1 \vee e_2$ and $e_1 \wedge e_2$ as a syntactic sugar for 
+\kvd{if}\,.\,.\,\kvd{then}\,.\,.\,\kvd{else} so the result of the reduction is just a Foo 
+expression.
+
+The remaining operations convert data values into values of less preferred shape.  
+The \ident{convPrim} and \ident{convFloat} operations take the required shape and a data value. 
+When the data does not match the required type, they do not reduce. For example, 
+$\ident{convPrim}(\kvd{bool}, \num{42})$ represents a stuck state, but $\ident{convFloat}(\kvd{float}, \num{42})$ 
+turns an integer \num{42} into a floating-point numerical value \num{42.0}.
+
+The \ident{convNull}, \ident{convElements} and \ident{convField} operations take an additional
+parameter $e$ which represents a function to be used in order to convert a contained value (non-null
+optional value, list elements or field value); \ident{convNull} turns \kvd{null} data value into
+\ident{None} and \ident{convElements} turns a data collection $[d_1, \ldots, d_n]$ into a Foo list 
+$v_1 :: \ldots :: v_n :: \kvd{nil}$ and a \kvd{null} value into an empty list.
 
@@
 Well-typed Foo programs reduce to a value in a finite number of steps or get stuck due to an
-error condition. The stuck states can only be due to the dynamic data operations (e.g.
-$\ident{convFloat}(\ident{float},\kvd{null})$). The relativized safety (Theorem~\ref{thm:safety})
-characterizes the additional conditions on input data under which Foo programs do not ``go wrong''.
+error condition. The stuck states can only be due to the dynamic data operations (e.g. an attempt
+to convert \kvd{null} value to a number $\ident{convFloat}(\ident{float},\kvd{null})$). The relative safety (Theorem~\ref{thm:safety})
+characterizes the additional conditions on input data under which Foo programs do not get stuck.
 
@@
 $L; \Gamma \vdash e : \tau$ where the context also contains a set of class declarations $L$.
-The fragment demonstrates the differences and similarities with Featherweight Java \cite{fwjava}:
-%
-\begin{itemize}[noitemsep]
+The fragment demonstrates the differences and similarities with Featherweight Java \cite{fwjava} and
+typing rules for the dynamic data operations $op$:
+%
+\begin{itemize}
 \item[--] All data values $d$ have the type \ident{Data}, but primitive data values (Booleans, 
@@
   primitive type as illustrated by the rule for $i$ and $f$.
+
 \item[--] For non-primitive data values (including \kvd{null}, data collections and records),
   \ident{Data} is the only type.
-\item[--] Operations $op$ (omitted) have a family of types (Foo does not have polymorphic types),
-  accepting \ident{Data} as one of the arguments and producing a non-\ident{Data} Foo type.
+
+\item[--] Operations $op$ accept \ident{Data} as one of the arguments and produce a non-\ident{Data} 
+  Foo type. Some of them require a function specifying the conversion for nested values.
+
 \item[--] Rules for checking class construction and member access are similar to corresponding
@@
 \end{equation*}
-%
-The mapping $\sem{\sigma}$ takes an inferred shape $\sigma$. It returns an F\# type $\tau$ and
-a function that turns the input data (value of type \ident{Data}) into a Foo value of type $\tau$. 
-The type provider also generates class definitions that may be used by $e$.
-
-Figure~\ref{fig:tp-generation} defines $\sem{-}$. Primitive types are handled by a single rule that
-inserts an appropriate conversion function; \ident{convPrim} just checks that the shape matches
-and \ident{convFloat} converts numbers to a floating-point.
-
-For records, we generate a class $C$ that takes a data value as constructor parameter. For each 
-record field, we generate a member with the same name as the field. The body of the member calls 
-\ident{convField} with a function obtained from $\sem{\sigma_i}$. This function turns the field 
-value (data of shape $\sigma_i$) into a Foo value of type $\tau_i$. The returned expression creates a new instance of 
-$C$ and the mapping returns the class $C$ together with all recursively generated classes. Note that 
-the class name $C$ is not directly accessed by the user and so we can use an arbitrary name, although the 
-actual implementation in F\# Data attempts to infer a reasonable name.\footnote{For example, in 
-\ident{\{\str{person}:\{\str{name}:\str{Tomas}\}\}}, the nested record will be named \ident{Person}
-based on the name of the parent record field.}
-
-A collection shape becomes a Foo $\ident{list}\langl\tau\rangl$. The returned expression calls \ident{convElements}
-(which returns the empty list for data value \kvd{null}). The last parameter is the recursively obtained
-conversion function for the shape of elements $\sigma$. The handling of the nullable shape is similar,
-but uses \ident{convNull}.
-
-As discussed earlier, labelled top shapes are also generated as Foo classes with properties. Given 
-$\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl$, we get corresponding F\# types $\tau_i$ and generate 
-$n$ members of type $\ident{option}\langl \tau_i\rangl$. When the member is accessed, we need to perform
-a runtime shape test using \ident{hasShape} to ensure that the value has the right shape (similarly to runtime 
-type conversions from the top type in languages like Java). If the shape matches, a \ident{Some} value is 
-returned. The shape inference algorithm also guarantees that there is only one case for each shape tag 
-(\S\ref{sec:inference-commonsuper}) and so we can use the tag for the name of the generated member.
 
@@
 \begin{equation*}
+\begin{array}{c}  
 \inference
@@
   {L; \Gamma \vdash f : \ident{float}}
-\end{equation*}
-\vspace{0.25em}
-\begin{equation*}
+\\[1.75em]
 \inference
@@
   {L; \Gamma \vdash e_1~e_2 : \tau_2}
-\end{equation*}
-\vspace{0.5em}
-\begin{equation*}
+\\[1.75em]
 \inference
@@
   {L; \Gamma \vdash e.N_i:\tau_i}
-\end{equation*}
-\vspace{0.25em}
-\begin{equation*}
+\\[1.75em]
 \inference
@@
   {L; \Gamma \vdash \kvd{new}~C(e_1, \ldots, e_n):C}
+\\[1.75em]
+\inference
+  {L; \Gamma \vdash e : \ident{Data}}
+  {L; \Gamma \vdash \ident{hasShape}(\sigma, e) : \ident{bool}}
+\\[1.75em]
+\inference
+  { L; \Gamma \vdash e : \tau & \tau \in \{ \ident{int}, \ident{float} \} }
+  { L; \Gamma \vdash \ident{convFloat}(\sigma, e) : \ident{float} }
+\\[1.75em]
+\inference
+  { L; \Gamma \vdash e_1 : \ident{Data} & L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
+  { L; \Gamma \vdash \ident{convNull}(e_1, e_2) : \ident{option}\langle\tau\rangle }
+\\[1.75em]
+\inference
+  { L; \Gamma \vdash e : \ident{Data} \\
+    (\sigma,\tau)\in\{ (\ident{int},\ident{int}), (\ident{string},\ident{string}), (\ident{bool},\ident{bool}) \} }
+  { L; \Gamma \vdash \ident{convPrim}(\sigma, e) : \tau }
+\\[1.75em]
+\inference
+  { L; \Gamma \vdash e_1 : \ident{Data} & L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
+  { L; \Gamma \vdash \ident{convElements}(e_1, e_2) : \ident{list}\langle\tau\rangle } 
+\\[1.75em]
+\inference
+  { L; \Gamma \vdash e_1 : \ident{Data} & L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
+  { L; \Gamma \vdash \ident{convField}(\nu, \nu', e_1, e_2) : \tau }
+\end{array}
 \end{equation*}
@@
 
+\noindent
+The mapping $\sem{\sigma}$ takes an inferred shape $\sigma$. It returns an F\# type $\tau$ and
+a function that turns the input data (value of type \ident{Data}) into a Foo value of type $\tau$. 
+The type provider also generates class definitions that may be used by $e$.
+
+Figure~\ref{fig:tp-generation} defines $\sem{-}$. Primitive types are handled by a single rule that
+inserts an appropriate conversion function; \ident{convPrim} just checks that the shape matches
+and \ident{convFloat} converts numbers to a floating-point.
+
+For records, we generate a class $C$ that takes a data value as constructor parameter. For each 
+record field, we generate a member with the same name as the field. The body of the member calls 
+\ident{convField} with a function obtained from $\sem{\sigma_i}$. This function turns the field 
+value (data of shape $\sigma_i$) into a Foo value of type $\tau_i$. The returned expression creates a new instance of 
+$C$ and the mapping returns the class $C$ together with all recursively generated classes. Note that 
+the class name $C$ is not directly accessed by the user and so we can use an arbitrary name, although the 
+actual implementation in F\# Data attempts to infer a reasonable name.\footnote{For example, in 
+\ident{\{\str{person}:\{\str{name}:\str{Tomas}\}\}}, the nested record will be named \ident{Person}
+based on the name of the parent record field.}
+
+A collection shape becomes a Foo $\ident{list}\langl\tau\rangl$. The returned expression calls \ident{convElements}
+(which returns the empty list for data value \kvd{null}). The last parameter is the recursively obtained
+conversion function for the shape of elements $\sigma$. The handling of the nullable shape is similar,
+but uses \ident{convNull}.
+
+As discussed earlier, labelled top shapes are also generated as Foo classes with properties. Given 
+$\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl$, we get corresponding F\# types $\tau_i$ and generate 
+$n$ members of type $\ident{option}\langl \tau_i\rangl$. When the member is accessed, we need to perform
+a runtime shape test using \ident{hasShape} to ensure that the value has the right shape (similarly to runtime 
+type conversions from the top type in languages like Java). If the shape matches, a \ident{Some} value is 
+returned. The shape inference algorithm also guarantees that there is only one case for each shape tag 
+(\S\ref{sec:inference-commonsuper}) and so we can use the tag for the name of the generated member.
+
+% -------------------------------------------------------------------------------------------------
+
 \begin{figure*}
@@
 \begin{array}{l} 
- \sem{\sigma_p} = \tau_p,\lambda x\rightarrow op(\sigma_p, x),\emptyset\quad\textnormal{where} \\[0.6em] 
+ \sem{\sigma_p} = \tau_p,\lambda x. op(\sigma_p, x),\emptyset\quad\textnormal{where} \\[0.6em] 
  \quad\sigma_p, \tau_p, op\in  \{~ (\ident{bool}, \ident{bool}, \ident{convPrim})\\
@@
  \sem{\,\nu~ \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}\,} = \\[0.1em]
- \quad C, \lambda x\rightarrow \kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
+ \quad C, \lambda x. \kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
  \qquad \;\;C~\textnormal{is a fresh class name} \\
@@
 \begin{array}{l}
- \sem{\,[\sigma]\,} = \ident{list}\langl\tau\rangl, \lambda x \rightarrow \ident{convElements}(x, e'), L \;\;\textnormal{where}\\[0.4em]
+ \sem{\,[\sigma]\,} = \ident{list}\langl\tau\rangl, \lambda x . \ident{convElements}(x, e'), L \;\;\textnormal{where}\\[0.4em]
  \qquad \tau, e', L = \sem{\hat{\sigma}}
@@
  \sem{\,\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl\,} = \\
- \quad C, \lambda x\rightarrow\kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\;\;\textnormal{where}\\[0.6em]
+ \quad C, \lambda x.\kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\;\;\textnormal{where}\\[0.6em]
  \qquad \;\;C~\textnormal{is a fresh class name} \\
@@
  \sem{\kvd{nullable}\langl\hat{\sigma}\rangl} = \\[0.2em]
- \qquad \ident{option}\langl\tau\rangl, \lambda x \rightarrow \ident{convNull}(x,e), L\\[0.2em] 
+ \qquad \ident{option}\langl\tau\rangl, \lambda x . \ident{convNull}(x,e), L\\[0.2em] 
  \qquad \textnormal{where}~\tau, e, L = \sem{\hat{\sigma}}
@@
 \begin{array}{l}
- \sem{\bot} = \sem{\kvd{null}} = C, \lambda x\rightarrow \kvd{new}~C(), \{ L \} \;\;\textnormal{where}\\[0.4em]
+ \sem{\bot} = \sem{\kvd{null}} = C, \lambda x. \kvd{new}~C(x), \{ L \} \;\;\textnormal{where}\\[0.4em]
  \qquad C~\textnormal{is a fresh class name} \\[0.1em]
@@
 type corresponding to the record shape is \ident{Person} the function to create it from a data value 
-is $\lambda x \rightarrow \kvd{new}~\ident{Person}(x)$.
+is $\lambda x . \kvd{new}~\ident{Person}(x)$.
 
@@
 
-\section{Relativized type safety}
+\section{Relative type safety}
 \label{sec:safety}
@@
 documents, any code that can be written using the provided types is guaranteed to work. We call this 
-\emph{relativized safety}, because we cannot avoid \emph{all} errors. In particular, one can always
+\emph{relative safety}, because we cannot avoid \emph{all} errors. In particular, one can always
 provide an input that has a different structure than any of the samples. In this case, it is expected 
@@
 
-\begin{theorem}[Relativized safety]
+\begin{theorem}[Relative safety]
 \label{thm:safety}
@@
 
-\subsection{Relativized safety in practice}
+\subsection{Relative safety in practice}
 \label{sec:safety-discuss}
 
-The \emph{relativized safety} property does not guarantee safety in the same way as traditional
+The \emph{relative safety} property does not guarantee safety in the same way as traditional
 closed-world type safety, but it reflects the reality that programming with external data is
@@
 makes formal specification of safety (Theorem~\ref{thm:safety}) difficult. Extending our 
-\emph{relativized safety} to \emph{probabilistic safety} is an interesting future direction.
+\emph{relative safety} to \emph{probabilistic safety} is an interesting future direction.
 
@@
 type providers, but our work demonstrates a more general point. The types generated by type 
-providers can depend on external input and so we can only guarantee \emph{relativized safety}, 
+providers can depend on external input and so we can only guarantee \emph{relative safety}, 
 which says that a program is safe only if the actual inputs satisfy additional conditions.

### 2016-02-08 pldi-2016/paper.tex
More edits in section 4 and 5

@@
 \noindent
-\begin{equation*}
-\hspace{1.1em}
+\begin{center}
+\textbf{Part I.} Reduction rules for conversion functions
+\end{center}
+
+\vspace{0.5em}
+\noindent
+\begin{equation*}
 \begin{array}{l}
@@
 \begin{equation*}
+\hspace{-1.1em}
 \begin{array}{l}
@@
 
-\caption{Reduction rules for conversion functions}
-\label{fig:op-conversions}
-\vspace{-0.5em}
-\end{figure*}
-
-% -------------------------------------------------------------------------------------------------
-
-Type providers for structured data map the ``dirty'' world of weakly typed 
-structured data into a ``nice'' world of strong types. To model this, the Foo calculus
-does not have \kvd{null} values and data values $d$ are never directly exposed.  
-Furthermore Foo is simply typed: despite using class types and object notation 
-for notational convenience, it has no subtyping.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{The Foo calculus}
-\label{sec:formal-ff}
-
-The syntax of the calculus is shown in Figure~\ref{fig:foo-syntax}. 
-The type \ident{Data} is the type of structural data $d$. A class definition $L$ consists of 
-a single constructor and zero or more parameter-less members. The declaration implicitly closes
-over the constructor parameters.
-Values $v$ include previously defined data $d$; expressions
-$e$ include class construction, member access, usual functional constructs (functions,
-lists, options) and conditionals. The $op$ constructs are discussed next.
-
-% -------------------------------------------------------------------------------------------------
-
-\paragraph{Dynamic data operations.}
-
-The Foo programs can only work with \ident{Data} values using certain primitive operations. Those are modelled
-by the $op$ primitives. In F\# Data, those are internal and users never access them directly.
-
-The behaviour of the dynamic data operations is defined by the reduction rules in 
-Figure~\ref{fig:op-conversions}. The typing is shown in Figure~\ref{fig:ff-typecheck} and is 
-discussed later. The \ident{hasShape} function represents a runtime shape test.
-It checks whether a \ident{Data} value $d$ (Section~\ref{sec:formal-inferval}) passed as the second
-argument has a shape specified by the first argument. For records, we have to check that for each 
-field $\nu_1, \ldots, \nu_n$ in the record, the actual record value has a field of the same name 
-with a matching shape. The last line defines a ``catch all'' pattern, which returns \kvd{false} 
-for all remaining cases. We treat $e_1 \vee e_2$ and $e_1 \wedge e_2$ as a syntactic sugar for 
-\kvd{if}\,.\,.\,\kvd{then}\,.\,.\,\kvd{else} so the result of the reduction is just a Foo 
-expression.
-
-The remaining operations convert data values into values of less preferred shape.  
-The \ident{convPrim} and \ident{convFloat} operations take the required shape and a data value. 
-When the data does not match the required type, they do not reduce. For example, 
-$\ident{convPrim}(\kvd{bool}, \num{42})$ represents a stuck state, but $\ident{convFloat}(\kvd{float}, \num{42})$ 
-turns an integer \num{42} into a floating-point numerical value \num{42.0}.
-
-The \ident{convNull}, \ident{convElements} and \ident{convField} operations take an additional
-parameter $e$ which represents a function to be used in order to convert a contained value (non-null
-optional value, list elements or field value); \ident{convNull} turns \kvd{null} data value into
-\ident{None} and \ident{convElements} turns a data collection $[d_1, \ldots, d_n]$ into a Foo list 
-$v_1 :: \ldots :: v_n :: \kvd{nil}$ and a \kvd{null} value into an empty list.
-
-% -------------------------------------------------------------------------------------------------
-
-\paragraph{Reduction.}
-The reduction relation is of the form $L, e \reduce e'$. We omit class declarations 
-$L$ where implied by context and write $e \reduce^{*} e'$ for
-the reflexive, transitive closure of $\reduce$.
-
-Figure~\ref{fig:ff-reduction} shows the reduction rules.  The (\emph{member}) rule reduces a member access using a class 
-definition in the assumption. The (\emph{ctx}) rule models the eager
-evaluation of F\# and performs a reduction inside a sub-expression specified by an evaluation 
-context $E$:
-%
-\begin{equation*}
-\begin{array}{rcl}
- E &\narrow{=}& v::E \lsep v~E \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e})\\[0.1em]
-   &\narrow{|}&  \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2  \lsep E = e \lsep v = E \\[0.1em]
-   &\narrow{|}& \ident{Some}(E) \lsep op(\overline{v}, E, \overline{e})\\[0.1em]
-   &\narrow{|}& \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
-   &\narrow{|}& \kvd{match}~E~\kvd{with}~x_1 :: x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
-\end{array} 
-\end{equation*}
-
-\noindent
-The evaluation proceeds from left to right as denoted by $\overline{v}, E, \overline{e}$ in 
-constructor and dynamic data operation arguments or $v::E$ in list initialization.
-We write $e[\overline{x} \leftarrow \overline{v}]$ for the result of replacing variables $\overline{x}$ by
-values $\overline{v}$ in an expression. The remaining six rules
-give standard reductions.
-
-% -------------------------------------------------------------------------------------------------
-
-\paragraph{Type checking.} 
-Well-typed Foo programs reduce to a value in a finite number of steps or get stuck due to an
-error condition. The stuck states can only be due to the dynamic data operations (e.g. an attempt
-to convert \kvd{null} value to a number $\ident{convFloat}(\ident{float},\kvd{null})$). The relative safety (Theorem~\ref{thm:safety})
-characterizes the additional conditions on input data under which Foo programs do not get stuck.
-
-Typing rules in Figure~\ref{fig:ff-typecheck} are written using a judgement
-$L; \Gamma \vdash e : \tau$ where the context also contains a set of class declarations $L$.
-The fragment demonstrates the differences and similarities with Featherweight Java \cite{fwjava} and
-typing rules for the dynamic data operations $op$:
-%
-\begin{itemize}
-\item[--] All data values $d$ have the type \ident{Data}, but primitive data values (Booleans, 
-  strings, integers and floats) can be implicitly converted to Foo values and so they also have a 
-  primitive type as illustrated by the rule for $i$ and $f$.
-
-\item[--] For non-primitive data values (including \kvd{null}, data collections and records),
-  \ident{Data} is the only type.
-
-\item[--] Operations $op$ accept \ident{Data} as one of the arguments and produce a non-\ident{Data} 
-  Foo type. Some of them require a function specifying the conversion for nested values.
-
-\item[--] Rules for checking class construction and member access are similar to corresponding
-  rules of Featherweight Java.  
-\end{itemize}
-%
-An important part of Featherweight Java that is omitted here is the checking of type declarations
-(ensuring the members are well-typed). We consider only classes generated by our type providers 
-and those are well-typed by construction.
-
-% -------------------------------------------------------------------------------------------------
-
-\begin{figure}
+\noindent
+\begin{center}
+\textbf{Part II.} Reduction rules for the rest of the Foo calculus 
+\end{center}
+
+\vspace{0.5em}
 \noindent
@@
  { L, (\kvd{new}~C(\overline{v})).N_i \reduce e_i[\overline{x} \leftarrow \overline{v}] }\\
- \\[-0.2em]
- \textnormal{\footnotesize{(match1)}}&
- \hspace{-1em}
- \begin{array}{l}
-  \kvd{match}~\ident{None}~\kvd{with} \\
-  \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
- \end{array} \hspace{-0.5em} ~\reduce~ e_2 \\
- \\[-0.2em]
- \textnormal{\footnotesize{(match2)}}&
- \hspace{-1em}
- \begin{array}{l}
-    \kvd{match}~\ident{Some}(v)~\kvd{with} \\
-    \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
- \end{array} \hspace{-0.5em} ~\reduce~ e_1[x\leftarrow v]\\
- \\[-0.2em]
- \textnormal{\footnotesize{(match3)}}&
- \hspace{-1em}
- \begin{array}{l}
-  \kvd{match}~\kvd{nil}~\kvd{with} \\[0em]
-  x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
- \end{array} \hspace{-0.5em} ~\reduce~ e_2\\
- \\[-0.2em]
- \textnormal{\footnotesize{(match4)}}&
- \hspace{-1em}
- \begin{array}{l}
-  \kvd{match}~v_1::v_2~\kvd{with} \\[0em]
-  x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
- \end{array} \hspace{-0.5em} ~\reduce~ e_1[\overline{x}\leftarrow\overline{v}] \\
  \\[-0.2em]
@@
  (\lambda x.e)~v ~\reduce~ e[x\leftarrow v] \\
+\end{array}
+\quad
+\begin{array}{rl}
+ \textnormal{\footnotesize{(match1)}}&
+ \hspace{-1em}
+ \begin{array}{l}
+  \kvd{match}~\ident{None}~\kvd{with} \\
+  \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
+ \end{array} \hspace{-0.5em} ~\reduce~ e_2 \\
+ \\[-0.2em]
+ \textnormal{\footnotesize{(match2)}}&
+ \hspace{-1em}
+ \begin{array}{l}
+    \kvd{match}~\ident{Some}(v)~\kvd{with} \\
+    \ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2
+ \end{array} \hspace{-0.5em} ~\reduce~ e_1[x\leftarrow v]\\
+ \\[-0.2em]
+ \textnormal{\footnotesize{(match3)}}&
+ \hspace{-1em}
+ \begin{array}{l}
+  \kvd{match}~\kvd{nil}~\kvd{with} \\[0em]
+  x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
+ \end{array} \hspace{-0.5em} ~\reduce~ e_2\\
+ \\[-0.2em]
+ \textnormal{\footnotesize{(match4)}}&
+ \hspace{-1em}
+ \begin{array}{l}
+  \kvd{match}~v_1::v_2~\kvd{with} \\[0em]
+  x_1::x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
+ \end{array} \hspace{-0.5em} ~\reduce~ e_1[\overline{x}\leftarrow\overline{v}] \\
 \\[-0.3em]
@@
 
-\caption{Foo -- Remaining reduction rules}
+\caption{Foo -- Reduction rules for the Foo calculus and dynamic data operations}
 \label{fig:ff-reduction}
-\end{figure}
+\end{figure*}
 
@@
 
-\subsection{Type providers}
-\label{sec:formal-tp}
-
-So far, we defined the type inference algorithm which produces a shape $\sigma$ from one 
-or more sample documents (\S\ref{sec:inference}) and we defined a simplified model of evaluation
-of F\# (\S\ref{sec:formal-ff}) and F\# Data runtime (\S\ref{sec:formal-tp}). In this section, we 
-define how the type providers work, linking the two parts.
-
-All F\# Data type providers take (one or more) sample documents, infer a common preferred shape $\sigma$
-and then use it to generate F\# types that are exposed to the programmer.\footnote{The actual 
-implementation provides \emph{lazy erased types} as described in \cite{fsharp-typeprov}. Here, we treat 
-the code as actually generated. This is an acceptable simplification, because F\# Data type providers 
-do not rely on laziness or erasure of type provision.}
-
-\paragraph{Type provider mapping.}
-A type provider produces an F\# type $\tau$ together with a Foo expression and a collection of 
-class definitions. We express it using the following mapping:
-%
-\begin{equation*}
-\sem{\sigma} = (\tau, e, L) \qquad (\textnormal{where}~L,\emptyset \vdash e:\ident{Data}\rightarrow\tau)
-\end{equation*}
+Type providers for structured data map the ``dirty'' world of weakly typed 
+structured data into a ``nice'' world of strong types. To model this, the Foo calculus
+does not have \kvd{null} values and data values $d$ are never directly exposed.  
+Furthermore Foo is simply typed: despite using class types and object notation 
+for notational convenience, it has no subtyping.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{The Foo calculus}
+\label{sec:formal-ff}
+
+The syntax of the calculus is shown in Figure~\ref{fig:foo-syntax}. 
+The type \ident{Data} is the type of structural data $d$. A class definition $L$ consists of 
+a single constructor and zero or more parameter-less members. The declaration implicitly closes
+over the constructor parameters.
+Values $v$ include previously defined data $d$; expressions
+$e$ include class construction, member access, usual functional constructs (functions,
+lists, options) and conditionals. The $op$ constructs are discussed next.
 
@@
 
-\begin{figure}
+\paragraph{Dynamic data operations.}
+
+The Foo programs can only work with \ident{Data} values using certain primitive operations. Those are modelled
+by the $op$ primitives. In F\# Data, those are internal and users never access them directly.
+
+The behaviour of the dynamic data operations is defined by the reduction rules in 
+Figure~\ref{fig:ff-reduction} (Part I). The typing is shown in Figure~\ref{fig:ff-typecheck} and is 
+discussed later. The \ident{hasShape} function represents a runtime shape test.
+It checks whether a \ident{Data} value $d$ (Section~\ref{sec:formal-inferval}) passed as the second
+argument has a shape specified by the first argument. For records, we have to check that for each 
+field $\nu_1, \ldots, \nu_n$ in the record, the actual record value has a field of the same name 
+with a matching shape. The last line defines a ``catch all'' pattern, which returns \kvd{false} 
+for all remaining cases. We treat $e_1 \vee e_2$ and $e_1 \wedge e_2$ as a syntactic sugar for 
+\kvd{if}\,.\,.\,\kvd{then}\,.\,.\,\kvd{else} so the result of the reduction is just a Foo 
+expression.
+
+The remaining operations convert data values into values of less preferred shape.  
+The \ident{convPrim} and \ident{convFloat} operations take the required shape and a data value. 
+When the data does not match the required type, they do not reduce. For example, 
+$\ident{convPrim}(\kvd{bool}, \num{42})$ represents a stuck state, but $\ident{convFloat}(\kvd{float}, \num{42})$ 
+turns an integer \num{42} into a floating-point numerical value \num{42.0}.
+
+The \ident{convNull}, \ident{convElements} and \ident{convField} operations take an additional
+parameter $e$ which represents a function to be used in order to convert a contained value (non-null
+optional value, list elements or field value); \ident{convNull} turns \kvd{null} data value into
+\ident{None} and \ident{convElements} turns a data collection $[d_1, \ldots, d_n]$ into a Foo list 
+$v_1 :: \ldots :: v_n :: \kvd{nil}$ and a \kvd{null} value into an empty list.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure*}
 \noindent  
@@
   {~}
-  {L; \Gamma \vdash d : \ident{Data}}
+  {\hspace{-0.5em}L; \Gamma \vdash d : \ident{Data}\hspace{-0.5em}}
 ~~
@@
   {~}
-  {L; \Gamma \vdash i : \ident{int}}
+  {\hspace{-0.5em}L; \Gamma \vdash i : \ident{int}\hspace{-0.5em}}
 ~~
@@
   {~}
-  {L; \Gamma \vdash f : \ident{float}}
-\\[1.75em]
+  {\hspace{-0.5em}L; \Gamma \vdash f : \ident{float}\hspace{-0.5em}}
+~~
 \inference
@@
   {L; \Gamma \vdash \lambda x.e : \tau_2}
-\quad
+~~
 \inference
-  {L; \Gamma \vdash e_2 : \tau_1 \\
+  {L; \Gamma \vdash e_2 : \tau_1 &
    L; \Gamma \vdash e_1 : \tau_1 \rightarrow \tau_2}
@@
 \inference
-  {L; \Gamma \vdash e : C \\ \kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;.. \in L}
-  {L; \Gamma \vdash e.N_i:\tau_i}
-\\[1.75em]
-\inference
-  {L; \Gamma \vdash e_i : \tau_i & \kvd{type}~C(x_1:\tau_1, \ldots, x_n:\tau_n) = \ldots \in L}
-  {L; \Gamma \vdash \kvd{new}~C(e_1, \ldots, e_n):C}
-\\[1.75em]
-\inference
   {L; \Gamma \vdash e : \ident{Data}}
   {L; \Gamma \vdash \ident{hasShape}(\sigma, e) : \ident{bool}}
-\\[1.75em]
+\quad
 \inference
@@
   { L; \Gamma \vdash \ident{convFloat}(\sigma, e) : \ident{float} }
-\\[1.75em]
+\quad
 \inference
@@
   { L; \Gamma \vdash e : \ident{Data} \\
-    (\sigma,\tau)\in\{ (\ident{int},\ident{int}), (\ident{string},\ident{string}), (\ident{bool},\ident{bool}) \} }
-  { L; \Gamma \vdash \ident{convPrim}(\sigma, e) : \tau }
+    \ident{prim}\in\{ \ident{int}, \ident{string}, \ident{bool} \} }
+  { L; \Gamma \vdash \ident{convPrim}(\ident{prim}, e) : \ident{prim} }
+\quad
+\inference
+  { L; \Gamma \vdash e_1 : \ident{Data} \\ L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
+  { L; \Gamma \vdash \ident{convElements}(e_1, e_2) : \ident{list}\langle\tau\rangle } 
+\quad
+\inference
+  { L; \Gamma \vdash e_1 : \ident{Data} \\ L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
+  { L; \Gamma \vdash \ident{convField}(\nu, \nu', e_1, e_2) : \tau }
 \\[1.75em]
 \inference
-  { L; \Gamma \vdash e_1 : \ident{Data} & L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
-  { L; \Gamma \vdash \ident{convElements}(e_1, e_2) : \ident{list}\langle\tau\rangle } 
-\\[1.75em]
+  {L; \Gamma \vdash e : C \\ \kvd{type}~C(\overline{x:\tau}) = ..\;\kvd{member}~N_i : \tau_i = e_i\;.. \in L}
+  {L; \Gamma \vdash e.N_i:\tau_i}
+\qquad
 \inference
-  { L; \Gamma \vdash e_1 : \ident{Data} & L; \Gamma \vdash e_2 : \ident{Data} \rightarrow \tau }
-  { L; \Gamma \vdash \ident{convField}(\nu, \nu', e_1, e_2) : \tau }
+  {L; \Gamma \vdash e_i : \tau_i & \kvd{type}~C(x_1:\tau_1, \ldots, x_n:\tau_n) = \ldots \in L}
+  {L; \Gamma \vdash \kvd{new}~C(e_1, \ldots, e_n):C}
 \end{array}
@@
 \label{fig:ff-typecheck}
-\end{figure}
+\end{figure*}
+
+% -------------------------------------------------------------------------------------------------
+
+\paragraph{Reduction.}
+The reduction relation is of the form $L, e \reduce e'$. We omit class declarations 
+$L$ where implied by context and write $e \reduce^{*} e'$ for
+the reflexive, transitive closure of $\reduce$.
+
+Figure~\ref{fig:ff-reduction} (Part II) shows the reduction rules.  The (\emph{member}) rule reduces 
+a member access using a class definition in the assumption. The (\emph{ctx}) rule models the eager
+evaluation of F\# and performs a reduction inside a sub-expression specified by an evaluation 
+context $E$:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ E &\narrow{=}& v::E \lsep v~E \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e})\\[0.1em]
+   &\narrow{|}&  \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2  \lsep E = e \lsep v = E \\[0.1em]
+   &\narrow{|}& \ident{Some}(E) \lsep op(\overline{v}, E, \overline{e})\\[0.1em]
+   &\narrow{|}& \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
+   &\narrow{|}& \kvd{match}~E~\kvd{with}~x_1 :: x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
+\end{array} 
+\end{equation*}
+
+\noindent
+The evaluation proceeds from left to right as denoted by $\overline{v}, E, \overline{e}$ in 
+constructor and dynamic data operation arguments or $v::E$ in list initialization.
+We write $e[\overline{x} \leftarrow \overline{v}]$ for the result of replacing variables $\overline{x}$ by
+values $\overline{v}$ in an expression. The remaining six rules
+give standard reductions.
+
+% -------------------------------------------------------------------------------------------------
+
+\paragraph{Type checking.} 
+Well-typed Foo programs reduce to a value in a finite number of steps or get stuck due to an
+error condition. The stuck states can only be due to the dynamic data operations (e.g. an attempt
+to convert \kvd{null} value to a number $\ident{convFloat}(\ident{float},\kvd{null})$). The relative safety (Theorem~\ref{thm:safety})
+characterizes the additional conditions on input data under which Foo programs do not get stuck.
+
+Typing rules in Figure~\ref{fig:ff-typecheck} are written using a judgement
+$L; \Gamma \vdash e : \tau$ where the context also contains a set of class declarations $L$.
+The fragment demonstrates the differences and similarities with Featherweight Java \cite{fwjava} and
+typing rules for the dynamic data operations $op$:
+%
+\begin{itemize}
+\item[--] All data values $d$ have the type \ident{Data}, but primitive data values (Booleans, 
+  strings, integers and floats) can be implicitly converted to Foo values and so they also have a 
+  primitive type as illustrated by the rule for $i$ and $f$.
+
+\item[--] For non-primitive data values (including \kvd{null}, data collections and records),
+  \ident{Data} is the only type.
+
+\item[--] Operations $op$ accept \ident{Data} as one of the arguments and produce a non-\ident{Data} 
+  Foo type. Some of them require a function specifying the conversion for nested values.
+
+\item[--] Rules for checking class construction and member access are similar to corresponding
+  rules of Featherweight Java.  
+\end{itemize}
+%
+An important part of Featherweight Java that is omitted here is the checking of type declarations
+(ensuring the members are well-typed). We consider only classes generated by our type providers 
+and those are well-typed by construction.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Type providers}
+\label{sec:formal-tp}
+
+So far, we defined the type inference algorithm which produces a shape $\sigma$ from one 
+or more sample documents (\S\ref{sec:inference}) and we defined a simplified model of evaluation
+of F\# (\S\ref{sec:formal-ff}) and F\# Data runtime (\S\ref{sec:formal-tp}). In this section, we 
+define how the type providers work, linking the two parts.
+
+All F\# Data type providers take (one or more) sample documents, infer a common preferred shape $\sigma$
+and then use it to generate F\# types that are exposed to the programmer.\footnote{The actual 
+implementation provides \emph{lazy erased types} as described in \cite{fsharp-typeprov}. Here, we treat 
+the code as actually generated. This is an acceptable simplification, because F\# Data type providers 
+do not rely on laziness or erasure of type provision.}
+
+\paragraph{Type provider mapping.}
+A type provider produces an F\# type $\tau$ together with a Foo expression and a collection of 
+class definitions. We express it using the following mapping:
+%
+\begin{equation*}
+\sem{\sigma} = (\tau, e, L) \qquad (\textnormal{where}~L,\emptyset \vdash e:\ident{Data}\rightarrow\tau)
+\end{equation*}
 
@@
 and \ident{convFloat} converts numbers to a floating-point.
+
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure*}
+\begin{multicols}{2}
+
+% Primitives 
+\noindent
+\begin{equation*}
+\begin{array}{l} 
+ \sem{\sigma_p} = \tau_p,\lambda x. op(\sigma_p, x),\emptyset\quad\textnormal{where} \\[0.6em] 
+ \quad\sigma_p, \tau_p, op\in  \{~ (\ident{bool}, \ident{bool}, \ident{convPrim})\\
+ \hspace{2.9em} (\ident{int}, \ident{int}, \ident{convPrim}), (\ident{float},\ident{float},\ident{convFloat}),\\
+ \hspace{2.9em} (\ident{string},\ident{string},\ident{convPrim}) ~\}
+\end{array} 
+\end{equation*}
+\vspace{-2em}
+
+% Records
+\begin{equation*}
+\begin{array}{l}
+ \sem{\,\nu~ \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}\,} = \\[0.1em]
+ \quad C, \lambda x. \kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
+ \qquad \;\;C~\textnormal{is a fresh class name} \\
+ \qquad \;\,\,L = \kvd{type}~C(x_1\!:\!\ident{Data})=M_1 \ldots M_n \\
+ \qquad M_i = \kvd{member}~\nu_i:\tau_i=\ident{convField}(\nu, \nu_i, x_1, e_i),\\
+ \qquad \tau_i, e_i, L_i = \sem{\sigma_i} 
+\end{array}
+\end{equation*}
+\vspace{-2em}
+
+% Collections
+\begin{equation*}
+\begin{array}{l}
+ \sem{\,[\sigma]\,} = \ident{list}\langl\tau\rangl, \lambda x . \ident{convElements}(x, e'), L \;\;\textnormal{where}\\[0.4em]
+ \qquad \tau, e', L = \sem{\hat{\sigma}}
+\end{array}
+\end{equation*}
+
+% Labelled top
+\noindent
+\begin{equation*}
+\begin{array}{l}
+ \sem{\,\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl\,} = \\
+ \quad C, \lambda x.\kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\;\;\textnormal{where}\\[0.6em]
+ \qquad \;\;C~\textnormal{is a fresh class name} \\
+ \qquad \;\,\,L = \kvd{type}~C(x:\ident{Data})~=~M_1\ldots M_n \\
+ \qquad M_i = \kvd{member}~\nu_i:\ident{option}\langl\tau_i\rangl=\\
+ \hspace{5.8em}  \kvd{if}~\ident{hasShape}(\sigma_i, x)~\kvd{then}~ \ident{Some}(e_i~x)~\kvd{else}~\ident{None} \\[0.1em]
+ \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_e,\;\; \nu_i=\tytagof{(\sigma_i)}
+\end{array}
+\end{equation*}
+\vspace{-2em}
+
+% Nullables
+\begin{equation*}
+\hspace{-1.3em}
+\begin{array}{l}
+ \sem{\kvd{nullable}\langl\hat{\sigma}\rangl} = \\[0.2em]
+ \qquad \ident{option}\langl\tau\rangl, \lambda x . \ident{convNull}(x,e), L\\[0.2em] 
+ \qquad \textnormal{where}~\tau, e, L = \sem{\hat{\sigma}}
+\end{array}
+\end{equation*}
+\vspace{-2em}
+
+% Anything else
+\begin{equation*}
+\hspace{-1.3em}
+\begin{array}{l}
+ \sem{\bot} = \sem{\kvd{null}} = C, \lambda x. \kvd{new}~C(x), \{ L \} \;\;\textnormal{where}\\[0.4em]
+ \qquad C~\textnormal{is a fresh class name} \\[0.1em]
+ \qquad L = \kvd{type}~C(v:\ident{Data})
+\end{array}
+\end{equation*}
+\end{multicols}
+
+\caption{Type provider -- generation of Foo types from inferred structural types}
+\label{fig:tp-generation}
+\vspace{-0.5em}
+\end{figure*}
+
+% -------------------------------------------------------------------------------------------------
 
@@
 
-% -------------------------------------------------------------------------------------------------
-
-\begin{figure*}
-\begin{multicols}{2}
-
-% Primitives 
-\noindent
-\begin{equation*}
-\begin{array}{l} 
- \sem{\sigma_p} = \tau_p,\lambda x. op(\sigma_p, x),\emptyset\quad\textnormal{where} \\[0.6em] 
- \quad\sigma_p, \tau_p, op\in  \{~ (\ident{bool}, \ident{bool}, \ident{convPrim})\\
- \hspace{2.9em} (\ident{int}, \ident{int}, \ident{convPrim}), (\ident{float},\ident{float},\ident{convFloat}),\\
- \hspace{2.9em} (\ident{string},\ident{string},\ident{convPrim}) ~\}
-\end{array} 
-\end{equation*}
-\vspace{-2em}
-
-% Records
-\begin{equation*}
-\begin{array}{l}
- \sem{\,\nu~ \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}\,} = \\[0.1em]
- \quad C, \lambda x. \kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
- \qquad \;\;C~\textnormal{is a fresh class name} \\
- \qquad \;\,\,L = \kvd{type}~C(x_1\!:\!\ident{Data})=M_1 \ldots M_n \\
- \qquad M_i = \kvd{member}~\nu_i:\tau_i=\ident{convField}(\nu, \nu_i, x_1, e_i),\\
- \qquad \tau_i, e_i, L_i = \sem{\sigma_i} 
-\end{array}
-\end{equation*}
-\vspace{-2em}
-
-% Collections
-\begin{equation*}
-\begin{array}{l}
- \sem{\,[\sigma]\,} = \ident{list}\langl\tau\rangl, \lambda x . \ident{convElements}(x, e'), L \;\;\textnormal{where}\\[0.4em]
- \qquad \tau, e', L = \sem{\hat{\sigma}}
-\end{array}
-\end{equation*}
-
-% Labelled top
-\noindent
-\begin{equation*}
-\begin{array}{l}
- \sem{\,\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl\,} = \\
- \quad C, \lambda x.\kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\;\;\textnormal{where}\\[0.6em]
- \qquad \;\;C~\textnormal{is a fresh class name} \\
- \qquad \;\,\,L = \kvd{type}~C(x:\ident{Data})~=~M_1\ldots M_n \\
- \qquad M_i = \kvd{member}~\nu_i:\ident{option}\langl\tau_i\rangl=\\
- \hspace{5.8em}  \kvd{if}~\ident{hasShape}(\sigma_i, x)~\kvd{then}~ \ident{Some}(e_i~x)~\kvd{else}~\ident{None} \\[0.1em]
- \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_e,\;\; \nu_i=\tytagof{(\sigma_i)}
-\end{array}
-\end{equation*}
-\vspace{-2em}
-
-% Nullables
-\begin{equation*}
-\hspace{-1.3em}
-\begin{array}{l}
- \sem{\kvd{nullable}\langl\hat{\sigma}\rangl} = \\[0.2em]
- \qquad \ident{option}\langl\tau\rangl, \lambda x . \ident{convNull}(x,e), L\\[0.2em] 
- \qquad \textnormal{where}~\tau, e, L = \sem{\hat{\sigma}}
-\end{array}
-\end{equation*}
-\vspace{-2em}
-
-% Anything else
-\begin{equation*}
-\hspace{-1.3em}
-\begin{array}{l}
- \sem{\bot} = \sem{\kvd{null}} = C, \lambda x. \kvd{new}~C(x), \{ L \} \;\;\textnormal{where}\\[0.4em]
- \qquad C~\textnormal{is a fresh class name} \\[0.1em]
- \qquad L = \kvd{type}~C(v:\ident{Data})
-\end{array}
-\end{equation*}
-\end{multicols}
-
-\caption{Type provider -- generation of Foo types from inferred structural types}
-\label{fig:tp-generation}
-\vspace{-0.5em}
-\end{figure*}
-
-% -------------------------------------------------------------------------------------------------
-
 \paragraph{Example 1.}
@@
 $\ident{Person}~\{~\ident{Age}\!:\!\kvd{option}\langl\ident{int}\rangl, \ident{Name}\!:\!\ident{string}~ \}$. 
-The rules from Figure~\ref{fig:tp-generation} produce the following class (we reduce some 
-function applications for readability):
-%
-\vspace{-0.1em}
+The rules from Figure~\ref{fig:tp-generation} produce the following class \ident{Person} (shown below)
+with two members.
+
+The body of the \ident{Age} member uses \ident{convField} as specified by the case for optional
+record fields. The field shape is nullable and so \ident{convNull} is used in the continuation to 
+convert the value to \ident{None} if \ident{convField} produces a \kvd{null} data value and 
+\ident{hasShape} is used to ensure that the field has the correct shape. The \ident{Name} value should
+be always available and should have the right shape so \ident{convPrim} appears directly in the 
+continuation. This is where the evaluation can get stuck if the field value was missing:
+%
 \begin{equation*}
@@
 %
-The body of the \ident{Age} member uses \ident{convField} as specified by the case for optional
-record fields. The field shape is nullable and so \ident{convNull} is used in the continuation to 
-convert the value to \ident{None} if \ident{convField} produces a \kvd{null} data value and 
-\ident{hasShape} is used to ensure that the field has the correct shape. The \ident{Name} value should
-be always available and should have the right shape so \ident{convPrim} appears directly in the 
-continuation. This is where the evaluation can get stuck if the field value was missing. The F\# 
-type corresponding to the record shape is \ident{Person} the function to create it from a data value 
+The function to create the Foo value \ident{Person} from a data value 
 is $\lambda x . \kvd{new}~\ident{Person}(x)$.
@@
 $[\kvd{any}\langl\ident{Person}~\{ \ldots \},\ident{string}\rangl]$:
-
-\noindent
+%
 \begin{equation*}
@@
 and returns a function that parses a data value as follows:
-%
+
+\noindent
 \begin{equation*}
@@
 \item[--] Records in the input can have additional fields
-\item[--] Records in the input can have fewer fields, provided that the type of the field is nullable in some of the samples
+\item[--] Records in the input can have fewer fields than some of the records in the sample 
+  document, provided that the sample also contains records that do not have the field 
 \item[--] When a labelled top type is inferred from the sample, the actual input can also contain any other value,
@@
 This shows that provided types are correct with respect to the preferred shape relation. 
-This is an obligation upon the author of a type provider, which is a second reason why
-the type soundness is ``relative''.
 Our key theorem states that, for any input which is a subshape the inferred shape and 
@@
 This is not a problem for the Standard ML \cite{sml} subset and object-oriented subset \cite{fwjava} of the calculus. The 
-problematic part are the dynamic data operations (Figure~\ref{fig:op-conversions}). Given a data
+problematic part are the dynamic data operations (Figure~\ref{fig:ff-reduction}). Given a data
 value (of type \ident{Data}), the reduction can get stuck if the value does not have a structure 
@@
 Given the assumptions and definitions from Theorem~\ref{thm:safety}, there exists $e''$ such that
-$e[y\leftarrow e'~d'] \reduce e''$.
+$e'[y\leftarrow e~d'] \reduce e''$.
 \end{lemma}

### 2016-02-09 pldi-2016/paper.tex
Finish minor edits, draft stability

@@
   \nu \; \{ \nu_1 \!:\! \ident{csh}(\sigma_1, \sigma'_1), \; \ldots \; , \nu_n \!:\! \ident{csh}(\sigma_n, \sigma'_n) \} & (\emph{recd}) \\
-\ident{csh}(\sigma_1, \sigma_2) &=& \kvd{any} & (\emph{any})\\
+\ident{csh}(\sigma_1, \sigma_2) &=& \kvd{any} \quad(\textnormal{when}~\sigma_1\neq\nu\;\{ \ldots \}~\textnormal{or}~\sigma_2\neq\nu\;\{ \ldots \} ) & (\emph{any})\\
 \end{array}
@@
 
-\caption{The common preferred shape relation}
+\caption{The rules that define the common preferred shape function}
 \label{fig:subtyping-cst}
@@
 which is important for usability as discussed earlier (\S\ref{sec:providers-xml}).
-
+%
 \begin{definition}
@@
 When finding a common shape of two records (\emph{recd}) we find common preferred shapes of their 
-respective fields. The field labels of two record types can always be made to match, through the 
-shape inference from sample (\S\ref{sec:formal-inferval}), \emph{before} finding a common preferred 
-shape using $\ident{csh}$.
+respective fields. The \ident{csh} function is not defined for two records with mis-matching 
+fields, however, the fields can always be made to match, through the shape inference 
+from sample (\S\ref{sec:formal-inferval}), \emph{before} finding a common preferred shape 
+using $\ident{csh}$.
 
@@
 
-The \ident{csh} function finds the unique least upper bound of the partially ordered 
+When defined, \ident{csh} finds the unique least upper bound of the partially ordered 
 set of ground shapes (Lemma~\ref{thm:lub}).
@@
 \label{thm:lub}
-For ground $\sigma_1$ and $\sigma_2$, if $\sigma_1 \triangledown \sigma_2 \vdash \sigma$ then $\sigma$ is a least upper bound by $\sqsupseteq$.
+For ground $\sigma_1$ and $\sigma_2$, if $\ident{csh}(\sigma_1, \sigma_2) \vdash \sigma$ then $\sigma$ is a least upper bound by $\sqsupseteq$.
 \end{lemma}
+\begin{proof}
+By induction over the structure of the shapes $\sigma_1,\sigma_2$. Note that \ident{csh} only infers the top shape 
+\kvd{any} when on of the shapes is the top shape (\emph{top}) or when there is no other option 
+(\emph{any}); a nullable shape is introduced in $\addopt{-}$ only when no non-nullable shape can
+be used (\emph{null}), (\emph{opt}).
+\end{proof}
+
+\subsection{Inferring shapes from samples}
+\label{sec:formal-inferval}
+
+We now specify how we obtain the shape from data. As clrified later (\S\ref{sec:impl-parsing}), 
+we represent JSON, XML and CSV documents using the same first-order \emph{data} value:
+
+\noindent
+\begin{equation*}
+\begin{array}{lcl}
+  \\[-0.5em]
+ d &\narrow{=}& i \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
+   &\narrow{|}& [d_1; \ldots; d_n] \lsep \nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}
+\end{array}
+\end{equation*}
+%
+The definition includes primitive values ($i$ for integers, $f$ for floats
+and $s$ for strings) and \kvd{null}. A collection is written as a 
+list of values in square brackets. A record starts with a name $\nu$, followed by a 
+sequence of field assignments $\nu_i \mapsto d_i$.
+
+Figure~\ref{fig:shape-inference} defines a mapping $\semalt{d_1,\ldots,d_n}$ which turns a 
+collection of sample data $d_1, \ldots, d_n$ into a shape $\sigma$. Before applying $S$, we assume 
+each record in each $d_i$ is marked with a fresh row inference variable $\rho_i$. 
+We then choose a ground, minimal substitution $\theta$ for row variables. Because $\rho_i$ variables 
+represent potentially missing fields, the $\addopt{-}$ operator from Figure~\ref{fig:subtyping-cst} 
+is applied to all types in the vector.
+
+This is sufficient to equate the record field labels and satisfy the pre-conditions in rule 
+(\emph{recd}) when multiple record samples are merged. In practice, $\theta$ is found via row 
+variable unification \cite{rows-remy}. We omit the details here. No $\rho_i$ variables remain after 
+inference as the substitution chosen is ground.
+
+Primitive  values are mapped to their corresponding shapes. 
+When infering a shape from multiple samples, we use the common preferred shape relation to find a 
+common shape for all values (starting with $\bot$). This operation is used  
+when calling a type provider with multiple samples and also when inferring the shape of collection
+elements.
 
@@
 \begin{figure}[!h]
-\vspace{0em}  
-\noindent
 \begin{equation*}
@@
 \label{fig:shape-inference}
-\vspace{-0.5em}
+\vspace{-1em}
 \end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{proof}
-By induction over the definition of $\ident{csh}$. Note that the algorithm only infers the top shape 
-\kvd{any} when on of the shapes is the top shape (\emph{top}) or when there is no other option 
-(\emph{any}); a nullable shape is introduced in $\addopt{-}$ only when no non-nullable shape can
-be used (\emph{null}), (\emph{opt}).
-\end{proof}
-
-\subsection{Inferring shapes from samples}
-\label{sec:formal-inferval}
-
-We now specify how we obtain the shape from data. As clrified later (\S\ref{sec:impl-parsing}), 
-we represent JSON, XML and CSV documents using the same first-order \emph{data} value:
-%
-\begin{equation*}
-\begin{array}{lcl}
- d &\narrow{=}& i \lsep f \lsep s \lsep \kvd{true} \lsep \kvd{false} \lsep \kvd{null} \\[0.1em]
-   &\narrow{|}& [d_1; \ldots; d_n] \lsep \nu~\{ \nu_1 \mapsto d_1, \ldots, \nu_n \mapsto d_n \}
-\end{array}
-\end{equation*}
-%
-The definition includes primitive values ($i$ for integers, $f$ for floats
-and $s$ for strings) and \kvd{null}. A collection is written as a 
-list of values in square brackets. A record starts with a name $\nu$, followed by a 
-sequence of field assignments $\nu_i \mapsto d_i$.
-
-Figure~\ref{fig:shape-inference} defines a mapping $\semalt{d_1,\ldots,d_n}$ which turns a 
-collection of sample data $d_1, \ldots, d_n$ into a shape $\sigma$. Before applying $S$, we assume 
-each record in each $d_i$ is marked with a fresh row inference variable $\rho_i$.
-
-We then choose a ground, minimal, global substitution $\theta$ for row variables. The substitution
-is minimal in the sense that it introduces as few new fields as possible, but no additional fields, 
-and the newly introduced fields have the most specific shape. Because $\rho_i$ variables represent 
-potentially missing fields, the $\addopt{-}$ operator from Figure~\ref{fig:subtyping-cst} is applied 
-to all types in the vector.
-
-This is sufficient to equate the record field labels and satisfy the pre-conditions in rule 
-(\emph{recd}) when multiple record samples are merged. In practice, $\theta$ is found via row 
-variable unification \cite{rows-remy}. We omit the details here. No $\rho_i$ variables remain after 
-inference as the substitution chosen is ground.
-
-Primitive  values are mapped to their corresponding shapes. 
-When infering a shape from multiple samples, we use the common preferred shape relation to find a 
-common shape for all values (starting with $\bot$). This operation is used  
-when calling a type provider with multiple samples and also when inferring the shape of collection
-elements.
 
@@
 \inference
-  { L; \Gamma \vdash e : \tau & \tau \in \{ \ident{int}, \ident{float} \} }
+  { L; \Gamma \vdash e : \ident{Data} & \tau \in \{ \ident{int}, \ident{float} \} }
   { L; \Gamma \vdash \ident{convFloat}(\sigma, e) : \ident{float} }
@@
 works correctly on an input $d'$ that is a subshape of $d$. More formally, the provided
-provided expression (with input $d'$) can be reduced to a value and, if it is a class,
+expression (with input $d'$) can be reduced to a value and, if it is a class,
 all its members can also be reduced to values.
@@
 \label{thm:tp-correctness}
-Given sample data $d$ and an input data value $d'$ such that $\semalt{d} \sqsupseteq \semalt{d'}$
+Given sample data $d$ and an input data value $d'$ such that $\semalt{d'} \sqsubseteq \semalt{d}$
 and provided type, expression and classes $\tau, e, L = \sem{\semalt{d}}$, 
@@
 For other cases, analyze the provided code to see that it can work on all subtypes (for example~\ident{convElements}
-works on \kvd{null} values, \ident{convFloat} accepts an integer); for labelled top types and
-optional record fields, the \ident{hasShape} operation is used to guaranteed the correct shape at runtime.
+works on \kvd{null} values, \ident{convFloat} accepts an integer). Finally, for labelled top types,
+the \ident{hasShape} operation is used to guaranteed the correct shape at runtime.
 \end{proof}
@@
 
-For all inputs $d'$ such that $\sigma \sqsupseteq \semalt{d'}$ and all expressions $e'$
-(representing the user code) such that $e'$ does not contain primitive operations $op$ as a sub-expression and
-$\emptyset; y\!:\!\tau \vdash e'\!:\!\tau'$, it is the case that $L, e[y\leftarrow e'~d'] \reduce^{*} v$ for some value $v$ and
+For all inputs $d'$ such that $\semalt{d'} \sqsubseteq \sigma$ and all expressions $e'$
+(representing the user code) such that $e'$ does not contain any of the dynamic data operations $op$
+and any \ident{Data} values as sub-expressions and $L; y\!:\!\tau \vdash e'\!:\!\tau'$, it is 
+the case that $L, e[y\leftarrow e'~d'] \reduce^{*} v$ for some value $v$ and
 also $\emptyset; \vdash v : \tau'$.
@@
 This is not a problem for the Standard ML \cite{sml} subset and object-oriented subset \cite{fwjava} of the calculus. The 
-problematic part are the dynamic data operations (Figure~\ref{fig:ff-reduction}). Given a data
+problematic part are the dynamic data operations (Figure~\ref{fig:ff-reduction}, Part I). Given a data
 value (of type \ident{Data}), the reduction can get stuck if the value does not have a structure 
@@
 The F\# Data library has been widely adopted by users and is one of the most downloaded
-F\# libraries.\footnote{At the time of writing, the library has over 80,000 downloads on NuGet 
-(package repository), 1,821 commits and 44 contributors on GitHub.} A practical demonstration of 
+F\# libraries.\footnote{At the time of writing, the library has over 100,000 downloads on NuGet 
+(package repository), 1,844 commits and 44 contributors on GitHub.} A practical demonstration of 
 development using the library can be seen in an attached screencast and additional documentation can be
@@
 closed-world type safety, but it reflects the reality that programming with external data is
-increasingly important \cite{age-of-web}. Type providers increase the safety of this kind of programming,
-as seen earlier (\S\ref{sec:introduction}).
-
-One common issue is the handling of schema change. With type providers, the sample is captured 
-at compile-time. If the schema changes (so that the input is no longer related to the  
-shape of the sample), the program can fail at runtime and developers have to handle the exception. 
-The same problem happens when using weakly-typed code with explicit failure cases.
+increasingly important \cite{age-of-web}. Type providers increase the safety of this kind of 
+programming.
+
+\paragraph{Representative samples.}
+When choosing a representative sample document, the user does not need to provide a sample 
+that represents all possible inputs. They merely need to provide a sample that is representative
+with respect to data they intend to access. This makes the task of choosing a representative
+sample easier.
+
+\paragraph{Schema change.}
+Type providers are invoked at compile-time. If the schema changes (so that inputs are no longer 
+related to the shape of the sample used at compile-time), the program can fail at runtime and 
+developers have to handle the exception. The same problem happens when using weakly-typed code 
+with explicit failure cases.
 
 F\# Data can help discover such errors earlier. Our first example (\S\ref{sec:introduction})
-points the JSON type provider at a sample using a live URL. This has the advantage that a re-compilation 
-fails when the sample changes, which is an indication that the program needs to be updated to reflect the
-change. Repeated type-checking of programs by continuous integration systems can be used to help 
-detect such changes.
-
-In general, XML, CSV and JSON data sources without explicit schema will necessarily require techniques akin
-to those we have shown. However, some data sources provide explicit schema with schema versioning support 
-(with meta-data about how the schema changed). For those, a type 
-provider could be written to adapt automatically, but we leave this for future work.
+points the JSON type provider at a sample using a live URL. This has the advantage that a 
+re-compilation fails when the sample changes, which is an indication that the program needs to be 
+updated to reflect the change.
+
+\paragraph{Richer data sources.}
+In general, XML, CSV and JSON data sources without explicit schema will necessarily require 
+techniques akin to those we have shown. However, some data sources provide explicit schema with 
+schema versioning support. For those, a type provider could be written to adapt automatically, 
+but we leave this for future work.
 
@@
 %
-When generating F\# types for XML documents, we also lifts the members nested under the $\bullet$ 
-field into the parent type to simplify the structure of the generated type.
-
 The XML type provider also includes an option to use \emph{global inference}. In that case, 
@@
 
+\subsection{Providing idiomatic F\# types} 
+\label{sec:impl-naming}
+
+In order to provide types that are easy to use and follow the F\# coding guidelines,
+we perform a number of transformations on the provided types that simplify their structure
+and use more idiomatic naming of fields. For example, the type provided for the XML document in 
+\S\ref{sec:impl-parsing} is:
+%
+\begin{equation*}
+\begin{array}{l}
+ \kvd{type}~\ident{Root}~=  \\
+ \quad \kvd{member}~\ident{Id}~:~\ident{int} \\
+ \quad \kvd{member}~\ident{Item}~:~\ident{string}
+\end{array}
+\end{equation*}
+%
+To obtain the type signature, we used the type provider as defined in Figure~\ref{fig:tp-generation}
+and applied three additional transformations and simplifications:
+
+\begin{itemize}
+\item When a class $C$ contains a member $\bullet$, which is a class with further members, the
+  nested members are lifted into the class $C$. For example, the above type \ident{Root}
+  directly contains \ident{Item} rather than containing a member $\bullet$ returning a 
+  class with a member \ident{Item}.
+
+\item Remaining members named $\bullet$ in the provided classes (typically of primitive
+  types) are renamed to \ident{Value}.
+
+\item Class members are renamed to follow \ident{PascalCase} naming convention, when a 
+  colision occurs, a number is appended to the end as in \ident{PascalCase2}. The provided
+  implementation preforms the lookup using the original name.
+\end{itemize}
+
+Our current implementation also adds an additional member to each class that returns the
+underlying JSON node (called \ident{JsonValue}) or XML element (called \ident{XElement}). 
+Those return standard .NET or F\# representation of the value and can be used to dynamically 
+access data not exposed by the type providers, such as textual values inside mixed-content XML elements.
+
+% -------------------------------------------------------------------------------------------------
+
 \subsection{Heterogeneous collections} 
@@
 labelled top shapes in many common scenarios. In the earlier example, a sample collection 
-contains a record (with \strf{pages} and \strf{page} fields) and a nested collection with values.
+contains a record (with \strf{pages} field) and a nested collection with values.
 
@@
 
-\subsection{Capitalization} 
-\label{sec:impl-naming}
-
-% -------------------------------------------------------------------------------------------------
-
 \subsection{Predictability and stability} 
 \label{sec:impl-stable}
+
+One of the design motivations mentioned in Section~\ref{sec:providers} is that the inference
+mechanism is predictable and stable. Informally this means that ...
 
@@
 shape of the type.
+
+For the purpose of this section, assume conversion function \ident{int} and an error value
+\ident{error} that propagates, i.e. $C[\ident{error}]\reduce\ident{error}$
+
+\begin{theorem}[Stability of inference]
+
+Given set of samples $d_1, \ldots, d_n$
+
+And $\tau_1, e_1, L_1 = \sem{\semalt{d_1, \ldots, d_n}}$
+
+And some code $L_1; x:\tau_1\vdash e : \tau$
+
+~
+
+Now, add $d_{n+1}$ and let
+ $\tau_2, e_2, L_2 = \sem{\semalt{d_1, \ldots, d_n, d_{n+1}}}$
+
+There is $e'$
+
+such that $L_2; x:\tau_2\vdash e' : \tau$
+
+and if for some $d$ it is the case that $e[x\leftarrow e_1~d] \reduce v$ then
+also $e'[x\leftarrow e_2~d] \reduce v$
+
+Obtained by changing
+
+\begin{itemize}
+\item[--]
+$e$ to $\kvd{match}~e~\kvd{with}~\ident{Some}(v) \rightarrow v~|~\ident{None} \rightarrow \ident{wrong}$
+\item[--]
+$e$ to $e.\ident{Member}$ where $\ident{Member} = \tytagof(\sigma)$ for some $\sigma$
+\item[--]
+$e$ to $\ident{int}(e)$
+\end{itemize}
+\end{theorem}
+\begin{proof}
+
+Primitive type - can turn into float or into any with the primitive 
+type exposed as one of the members (because it has unique tag), possibly turned to float
+
+Record - introduce new fields, change types of fields - by induction
+or not record, in which case it becomes member via any
+
+List - element type change or becomes member of any
+
+Any - ok
+
+nullable - stays nullable, or if they're not compatible, part of any
+
+\end{proof}
+
+A full treatment of issue (2) would likely require more space than
+available. However, we would like to add a brief section that makes the
+following point: When a new sample is added to an existing set of samples,
+the inferred shape can change only in certain limited ways (i.e. certain
+parts may become optional). The fact that this change of shape is limited
+shows that the inference algorithm is "stable" (users only need to do small
+changes in their code). This is also relevant for the practical problem of
+"choosing a representative sample". If an initial sample is not
+representative and is refined later, the user will see which inferred types
+became optional - and can adjust the program accordingly.
 
@@
 
-% \acks
-% We would like to thank to the F\# Data contributors on GitHub and other colleagues working on type providers,
-% including Jomo Fisher, Keith Battocchi and Kenji Takeda.
+\acks
+We would like to thank to the F\# Data contributors on GitHub and other colleagues working 
+on type providers, including Jomo Fisher, Keith Battocchi and Kenji Takeda.
 

### 2016-02-09 pldi-2016/paper.tex
Final edits and fix bibtex

@@
-\documentclass[10pt,clearpagebib]{sigplanconf}
+\documentclass[10pt]{sigplanconf}
 
@@
 \usepackage{flushend}
+\usepackage{enumerate}
+\usepackage{etoolbox}
 
@@
 \newtheorem{theorem}{Theorem}
+\newtheorem{remark}{Remark}
 \newtheorem{corollary}[theorem]{Corollary}
 \newtheorem{lemma}[theorem]{Lemma}
+\newenvironment{proofoutline}
+ {\renewcommand\qedsymbol{}\proof[Proof outline]}
+ {\endproof}
 
@@
 
-One of the design motivations mentioned in Section~\ref{sec:providers} is that the inference
-mechanism is predictable and stable. Informally this means that ...
-
-In particular, we do 
-not use probabilistic methods where adding an additional sample could completely change the 
-shape of the type.
-
-For the purpose of this section, assume conversion function \ident{int} and an error value
-\ident{error} that propagates, i.e. $C[\ident{error}]\reduce\ident{error}$
-
-\begin{theorem}[Stability of inference]
-
-Given set of samples $d_1, \ldots, d_n$
-
-And $\tau_1, e_1, L_1 = \sem{\semalt{d_1, \ldots, d_n}}$
-
-And some code $L_1; x:\tau_1\vdash e : \tau$
-
-~
-
-Now, add $d_{n+1}$ and let
- $\tau_2, e_2, L_2 = \sem{\semalt{d_1, \ldots, d_n, d_{n+1}}}$
-
-There is $e'$
-
-such that $L_2; x:\tau_2\vdash e' : \tau$
-
-and if for some $d$ it is the case that $e[x\leftarrow e_1~d] \reduce v$ then
-also $e'[x\leftarrow e_2~d] \reduce v$
-
-Obtained by changing
+As discussed in \S\ref{sec:providers}, our inference algorithm is designed to be predictable
+and stable. When a user writes a program using the provided type and then adds another sample 
+(e.g.~with more missing values), they should not need to restructure their program. 
+For this reason, we keep the algorithm simple. For example, we do not use probabilistic methods to 
+asses similarity of record types, because a small change in the sample could cause a large change 
+in the provided types.
+
+We leave a general theory of stability and predictability of type providers to future work, but
+we formalize a brief observation in this section. Say we write a program using a provided type
+that is based on a collection of samples. When a new sample is added, the program can be modified
+to run as before with only small local changes.
+
+For the purpose of this section, assume that the Foo calculus also contains an \ident{exn} 
+value representing a runtime exception that propagates in the usual way, 
+i.e.~$C[\ident{exn}]\reduce\ident{exn}$, and also a conversion function \ident{int} that
+turns floating-point number into an integer.
+
+\begin{remark}[Stability of inference]
+Assume we have a set of samples $d_1, \ldots, d_n$, provided type based on the samples
+$\tau_1, e_1, L_1 = \sem{\semalt{d_1, \ldots, d_n}}$ and some user code $e$ written using
+the provided type, such that $L_1; x:\tau_1\vdash e : \tau$.
+
+Next, we add a new sample $d_{n+1}$ and consider a new provided type 
+$\tau_2, e_2, L_2 = \sem{\semalt{d_1, \ldots, d_n, d_{n+1}}}$.
+
+Now there exists $e'$ such that $L_2; x:\tau_2\vdash e' : \tau$ and if
+for some $d$ it is the case that $e[x\leftarrow e_1~d] \reduce v$ then
+also $e'[x\leftarrow e_2~d] \reduce v$.
+
+Such $e'$ is obtained by applying one or more transformations to sub-expressions of $e$
+that transform:
+%
+\begin{enumerate}[(i)]
+\item
+$C[e]$ to $C[\kvd{match}~e~\kvd{with}~\ident{Some}(v) \rightarrow v~|~\ident{None} \rightarrow \ident{exn}]$
+\item
+$C[e]$ to $C[e.\ident{M}]$ where $\ident{M} = \tytagof(\sigma)$ for some $\sigma$
+\item
+$C[e]$ to $C[\ident{int}(e)]$
+\end{enumerate}
+\end{remark}
+
+\begin{proof}
+For each case in the type provision (Figure~\ref{fig:tp-generation}) an original shape $\sigma$
+may be replaced by a less preferred shape $\sigma'$. The user code can always be transformed
+to use the newly provided shape:
 
 \begin{itemize}
-\item[--]
-$e$ to $\kvd{match}~e~\kvd{with}~\ident{Some}(v) \rightarrow v~|~\ident{None} \rightarrow \ident{wrong}$
-\item[--]
-$e$ to $e.\ident{Member}$ where $\ident{Member} = \tytagof(\sigma)$ for some $\sigma$
-\item[--]
-$e$ to $\ident{int}(e)$
+\item[--] Primitive shapes can become nullable (i), \ident{int} can become \ident{float} (iii)
+  or can become a part of labelled top type (ii).
+\item[--] Record shape fields can change shape (recursively) and record may become a part
+  of labelled top type (ii).
+\item[--] For list and nullable shape, the shape of the value may change (we apply the 
+  transformations recursively).
+\item[--] For any shape, the original code will continue to work (label is never removed).
 \end{itemize}
-\end{theorem}
-\begin{proof}
-
-Primitive type - can turn into float or into any with the primitive 
-type exposed as one of the members (because it has unique tag), possibly turned to float
-
-Record - introduce new fields, change types of fields - by induction
-or not record, in which case it becomes member via any
-
-List - element type change or becomes member of any
-
-Any - ok
-
-nullable - stays nullable, or if they're not compatible, part of any
-
+\vspace{-1.5em}
 \end{proof}
-
-A full treatment of issue (2) would likely require more space than
-available. However, we would like to add a brief section that makes the
-following point: When a new sample is added to an existing set of samples,
-the inferred shape can change only in certain limited ways (i.e. certain
-parts may become optional). The fact that this change of shape is limited
-shows that the inference algorithm is "stable" (users only need to do small
-changes in their code). This is also relevant for the practical problem of
-"choosing a representative sample". If an initial sample is not
-representative and is refined later, the user will see which inferred types
-became optional - and can adjust the program accordingly.
+%
+\noindent
+Intuitively, the first transformation is needed when the new sample makes a type optional.
+This happens when it contains a \kvd{null} value or a record that does not contain a field 
+that all previous samples have. The second transformation is needed when a shape $\sigma$ 
+becomes $\kvd{any}\langle\sigma, \ldots\rangle$ and the third one is needed when $\ident{int}$
+becomes \ident{float}.
+
+This property also underlines a common way of handling errors when using F\# Data type providers.
+When a program fails on some input, the input can be added as another sample. This makes some
+fields optional and the code can be updated accordingly, using a variation of (i) that uses
+an appropriate default value rather than throwing an exception.
 
@@
 
-We connect two lines of research: the work: inferring types for real-world data sources
-and the integration of external data into statically-typed programming langauges.
-
-Early work on F\# Data happened together with the development of the type providers mechanism in F\# 
-\cite{fsharp-typeprov,fsharp-typeprov-ddfp}, but previous publications on type providers focused on 
-the general mechanism \cite{idris-tp,liteq} rather than on concrete type providers. This paper 
-is novel in that it discusses the theory and safety of a concrete type provider from the programming
-language perspective.
+The F\# Data library connects two lines of research that have been previously disconnected. The first is 
+extending the type systems of programming languages to accommodate external data sources and the second
+is inferring types for real-world data sources.
+
+The type provider mechanism has been introduced in F\# \cite{fsharp-typeprov,fsharp-typeprov-ddfp}, 
+added to Idris  \cite{idris-tp} and used in areas such as semantic web \cite{liteq}. The F\# Data 
+library has been developed as part of the early F\# type provider research, but previous 
+publications focused on the general mechanisms. This paper is novel in that it shows the 
+programming language theory behind a concrete type providers.
 
 \paragraph{Extending the type systems.} 
-Previous work that integrates external data, such as XML \cite{xduce,xduce-ml} and databases 
-\cite{links,linq}, into a programming language, required the user to define the schema
-or had an ad-hoc extension that reads the schema.
-
-C$\omega$ \cite{comega-xs} bears similarity to F\# Data. Like many other
-languages, it extends a host language with advanced types similar to our
-structural shapes (including similar heterogeneous collections), but it 
-does not infer the types from a sample. In contrast, in F\# Data 
+Several systems integrate external data into a programming language. Those include 
+XML \cite{xduce,xduce-ml} and databases \cite{links}. In both of these, the system requires
+the user to explicitly define the schema (using the host language) or it has an ad-hoc extension 
+that reads the schema (\emph{e.g.}~from a database). LINQ \cite{linq} is more general, but relies
+on code generation when importing the schema.
+
+The work that is most similar to F\# Data is the data integration in C$\omega$ \cite{comega-xs}.
+It extends C\# language with types similar to our structural types 
+(including nullable types, choices with subtyping and heterogeneous collections with multiplicities).
+However, C$\omega$ does not infer the types from samples and extends the type system of the host
+language (rather than using a general purpose embedding mechanism).
+
+In contrast, in F\# Data 
 we have avoided extending F\# at all, and the simplicity of the Foo 
@@
 We would like to thank to the F\# Data contributors on GitHub and other colleagues working 
-on type providers, including Jomo Fisher, Keith Battocchi and Kenji Takeda.
+on type providers, including Jomo Fisher, Keith Battocchi and Kenji Takeda. We are also
+grateful to anonymous reviewers of the paper for their valuable feedback.
 

### 2016-02-09 pldi-2016/paper.tex
Typos

@@
 We include two numerical primitives, \ident{int} for integers and \ident{float} for floating-point 
-numbers. The two are related by the perference relation and we prefer \ident{int}.
+numbers. The two are related by the preference relation and we prefer \ident{int}.
 
@@
 
-We now specify how we obtain the shape from data. As clrified later (\S\ref{sec:impl-parsing}), 
+We now specify how we obtain the shape from data. As clarified later (\S\ref{sec:impl-parsing}), 
 we represent JSON, XML and CSV documents using the same first-order \emph{data} value:
@@
 Primitive  values are mapped to their corresponding shapes. 
-When infering a shape from multiple samples, we use the common preferred shape relation to find a 
+When inferring a shape from multiple samples, we use the common preferred shape relation to find a 
 common shape for all values (starting with $\bot$). This operation is used  
@@
 \item Class members are renamed to follow \ident{PascalCase} naming convention, when a 
-  colision occurs, a number is appended to the end as in \ident{PascalCase2}. The provided
+  collision occurs, a number is appended to the end as in \ident{PascalCase2}. The provided
   implementation preforms the lookup using the original name.

### 2016-02-12 pldi-2016/paper.tex
Corrections

@@
 \end{equation*}
-%
+
 \begin{equation*}
@@
 between certain rules. Most importantly (\emph{any}) is used only as the last resort.
+
 When finding a common shape of two records (\emph{recd}) we find common preferred shapes of their 
-respective fields. The \ident{csh} function is not defined for two records with mis-matching 
-fields, however, the fields can always be made to match, through the shape inference 
-from sample (\S\ref{sec:formal-inferval}), \emph{before} finding a common preferred shape 
-using $\ident{csh}$.
-
-We can find a common shape of two different numbers (\emph{num}); for two collections,
+respective fields. We can find a common shape of two different numbers (\emph{num}); for two collections,
 we combine their elements (\emph{list}). When one shape is nullable (\emph{opt}), we find
@@
 \end{equation*}
-%
+
+~
+
+\noindent
 The definition includes primitive values ($i$ for integers, $f$ for floats
@@
 This is sufficient to equate the record field labels and satisfy the pre-conditions in rule 
-(\emph{recd}) when multiple record samples are merged. In practice, $\theta$ is found via row 
+(\emph{recd}) when multiple record shapes are combined. The \ident{csh} function is not defined for 
+two records with mis-matching fields, however, the fields can always be made to match, through 
+a substitution for row variables. In practice, $\theta$ is found via row 
 variable unification \cite{rows-remy}. We omit the details here. No $\rho_i$ variables remain after 
@@
 \begin{remark}[Stability of inference]
-Assume we have a set of samples $d_1, \ldots, d_n$, provided type based on the samples
+Assume we have a set of samples $d_1, \ldots, d_n$, a provided type based on the samples
 $\tau_1, e_1, L_1 = \sem{\semalt{d_1, \ldots, d_n}}$ and some user code $e$ written using
@@
   of labelled top type (ii).
-\item[--] For list and nullable shape, the shape of the value may change (we apply the 
+\item[--] For list and nullable shapes, the shape of the value may change (we apply the 
   transformations recursively).
-\item[--] For any shape, the original code will continue to work (label is never removed).
+\item[--] For the any shape, the original code will continue to work (the label is never removed).
 \end{itemize}

### 2016-04-15 pldi-2016/paper.tex
Published version

@@
 For example, \url{http://openweathermap.org/current} contains one example to document an end-point
-to get the current weather. Using standard libraries, we might call it as\footnote{We use an abbreviation
-\url{http://api.owm.org/?q=X} for the full URL:
-\url{http://api.openweathermap.org/data/2.5/weather?q=X} 
-}:
-
-\noindent
-\begin{equation*}
-\begin{array}{l}
- \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC"}) \\
+to get the current weather. Using standard libraries, we might call it as\footnote{We abbreviate
+the full URL and omit application key (available after registration). The returned JSON is shown
+in Appendix~\ref{sec:appendix-weather} and can be used to run the code against a local file.}:
+
+\noindent
+\begin{equation*}
+\begin{array}{l}
+ \kvd{let}~\ident{doc}=\ident{Http.Request}(\str{http://api.owm.org/?q=NYC}) \\
  \kvd{match}~\ident{JsonValue.Parse}(\ident{doc})~\kvd{with} \\
@@
 JSON and XML. We introduce the necessary aspects of F\# type providers along the way. The examples
-in this section illustrate the key design principles of the shape inference algorithm:
+in this section also illustrate the key design principles of the shape inference algorithm:
 
@@
 
-\item The above also makes our techniques applicable to any language with nominal 
+\item The above makes our techniques applicable to any language with nominal
   object types (e.g.~variations of Java or C\# with a type provider mechanism added).
@@
 The supplementary screencast provides further illustration of the practical developer
-experience using F\# Data.
+experience using F\# Data.\footnote{Available at \url{http://tomasp.net/academic/papers/fsharp-data}.}
 
@@
 %
-In contrast to the earlier example, we now use a local file \strf{people.json} as a sample for 
+We now use a local file as a sample for
 the type inference, but then processes data from another source. The code achieves a similar
@@
 The provider analyzes the sample and provides a type  \ident{People}. F\# editors also
-execute the type provider in the background at development-time and use the provided types 
-in auto-completion on ``.'' and background type-checking.
+execute the type provider at development-time and use the provided types
+for auto-completion on ``.'' and for background type-checking.
 
@@
 \begin{array}{l}
- \kvd{type}~\ident{Entity}~=  \\
- \quad \kvd{member}~\ident{Name}~:~\ident{string} \\
- \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{float}\rangl \\[0.5em]
- \kvd{type}~\ident{People}~=  \\
- \quad \kvd{member}~\ident{GetSample}~:~\ident{unit}~\rightarrow~\ident{Entity}[] \\
- \quad \kvd{member}~\ident{Parse}~:~\ident{string}~\rightarrow~\ident{Entity}[] \\
+ \kvd{type}~\ident{Entity}~=  \\[-0.05em]
+ \quad \kvd{member}~\ident{Name}~:~\ident{string} \\[-0.05em]
+ \quad \kvd{member}~\ident{Age}~:~\ident{option}\langl \ident{float}\rangl \\[0.4em]
+ \kvd{type}~\ident{People}~=  \\[-0.05em]
+ \quad \kvd{member}~\ident{GetSample}~:~\ident{unit}~\rightarrow~\ident{Entity}[] \\[-0.05em]
+ \quad \kvd{member}~\ident{Parse}~:~\ident{string}~\rightarrow~\ident{Entity}[] \\[-0.05em]
 \end{array}
@@
 sample values and is inferred as \ident{string}. The field \ident{Age} is marked as optional,
-because the value is missing in one sample. In F\#, we can then use \ident{Option.iter} to call
+because the value is missing in one sample. In F\#, we use \ident{Option.iter} to call
 the specified function (printing) only when an optional value is available. The two age values
@@
 
-\paragraph{The role of objects with properties.}
+\paragraph{Objects with properties.}
 The sample code is easy to write thanks to the fact that most F\# editors provide auto-completion
@@
 This is demonstrated by the fact that \ident{Age} becomes an optional member.
-An alternative is to provide two different record types (one with \ident{Name} and other with 
+An alternative is to provide two different record types (one with \ident{Name} and one with
 \ident{Name} and \ident{Age}), but this would complicate the processing code.
@@
  \kvd{type}~\ident{Document}~=~\ident{XmlProvider}\langl\str{sample.xml}\rangl\hspace{1em} \\[0.5em]
- \kvd{let}~\ident{root}~=~\ident{Document.Load}(\str{c:/pldi/another.xml})\\
+ \kvd{let}~\ident{root}~=~\ident{Document.Load}(\str{pldi/another.xml})\\
  \kvd{for}~\ident{elem}~\kvd{in}~\ident{root.Doc}~\kvd{do}\\
@@
 The government debt information returned by the World Bank\footnote{Available at
-\url{http://data.worldbank.org}} demonstrates all three:
+\url{http://data.worldbank.org}} includes all three:
 %
@@
 The shape algebra and inference presented here is influenced by the design principles
-we outlined earlier and the type definitions available in the F\# language.
-If applied to other languages, details may differ, for example
+we outlined earlier and by the type definitions available in the F\# language.
+The same principles apply to other languages, but details may differ, for example
 with respect to numerical types and missing data.
@@
 record $\ident{Point}\,\{ \ident{x} \mapsto \num{3} \}$ occurs in a sample,
-it may be combined with $\ident{Point}\,\{\, \ident{x}\mapsto\num{3}, \ident{y}\mapsto\num{4} \}$ that contain more fields. The
+it may be combined with $\ident{Point}\,\{\, \ident{x}\mapsto\num{3}, \ident{y}\mapsto\num{4} \}$ that contains more fields. The
 overall shape inferred must account for the fact that any extra fields are optional,
@@
 \begin{align}
-  \label{eq:sub-prim}\tag{P1}
+  \label{eq:sub-prim}
   \ident{int}\,&\sqsubseteq\,\ident{float}&\\[-0.2em]
-  \label{eq:sub-null}\tag{P2}
+  \label{eq:sub-null}
   \kvd{null} &\sqsubseteq \sigma  &(\textnormal{for}~\sigma \neq \hat{\sigma})  \\[-0.2em]
-  \label{eq:sub-opt}\tag{P3}
+  \label{eq:sub-opt}
   \hat{\sigma} &\sqsubseteq \kvd{nullable}\langl\hat{\sigma}\rangl  &(\textnormal{for all}~\hat{\sigma})\\[-0.2em]
-  \label{eq:sub-opt-cov}\tag{P4}
+  \label{eq:sub-opt-cov}
   \kvd{nullable}\langl\hat{\sigma_1}\rangl &\sqsubseteq
     \kvd{nullable}\langl\hat{\sigma_2}\rangl  &(\textnormal{if}~\hat{\sigma_1} \sqsubseteq \hat{\sigma_2})\\[-0.2em]
-  \label{eq:sub-col}\tag{P5}
+  \label{eq:sub-col}
   [\sigma_1] &\sqsubseteq [\sigma_2]  &(\textnormal{if}~\sigma_1 \sqsubseteq \sigma_2) \\[-0.2em]
-  \label{eq:sub-bot}\tag{P6}
+  \label{eq:sub-bot}
   \bot &\sqsubseteq \sigma  &(\textnormal{for all}~\sigma)\\[-0.2em]
-  \label{eq:sub-var-top}\tag{P7}
+  \label{eq:sub-var-top}
   \sigma &\sqsubseteq \kvd{any}
@@
 \begin{align}
-\label{eq:sub-record1}\tag{R1}
+\label{eq:sub-record1}
 \begin{array}{l}
@@
 \begin{align}
-\label{eq:sub-record2}\tag{R2}
+\label{eq:sub-record2}
 \begin{array}{l}
@@
 when calling a type provider with multiple samples and also when inferring the shape of collection
-elements.
+values.
 
@@
 When combining a top with another shape, the labels may or may not already contain a case with
-the tag of the other shape. If they doe, the two shapes are combined (\emph{top-incl}), otherwise 
+the tag of the other shape. If they do, the two shapes are combined (\emph{top-incl}), otherwise
 a new case is added (\emph{top-add}). Finally, (\emph{top-all}) replaces earlier (\emph{any})
@@
 The reduction relation is of the form $L, e \reduce e'$. We omit class declarations
-$L$ where implied by context and write $e \reduce^{*} e'$ for
+$L$ where implied by the context and write $e \reduce^{*} e'$ for
 the reflexive, transitive closure of $\reduce$.
@@
 and then use it to generate F\# types that are exposed to the programmer.\footnote{The actual
-implementation provides \emph{lazy erased types} as described in \cite{fsharp-typeprov}. Here, we treat 
+implementation provides \emph{erased types} as described in \cite{fsharp-typeprov}. Here, we treat
 the code as actually generated. This is an acceptable simplification, because F\# Data type providers
@@
 
-For records, we generate a class $C$ that takes a data value as constructor parameter. For each 
-record field, we generate a member with the same name as the field. The body of the member calls 
+For records, we generate a class $C$ that takes a data value as a constructor parameter. For each
+field, we generate a member with the same name as the field. The body of the member calls
 \ident{convField} with a function obtained from $\sem{\sigma_i}$. This function turns the field
@@
 that the inferred shape is a record
-$\ident{Person}~\{~\ident{Age}\!:\!\kvd{option}\langl\ident{int}\rangl, \ident{Name}\!:\!\ident{string}~ \}$. 
-The rules from Figure~\ref{fig:tp-generation} produce the following class \ident{Person} (shown below)
+$\ident{Person}~\{~\ident{Age}\!:\!\kvd{option}\langl\ident{int}\rangl,~\ident{Name}\!:\!\ident{string}~ \}$.
+The rules from Figure~\ref{fig:tp-generation} produce the \ident{Person} class shown below
 with two members.
@@
 \begin{itemize}
-\item[--] Input can contain smaller numerical values (e.g., if a sample contains float, the input can contain an integer)
-\item[--] Records in the input can have additional fields
+\item[--] Input can contain smaller numerical values (e.g., if a sample contains float, the input can contain an integer).
+\item[--] Records in the input can have additional fields.
 \item[--] Records in the input can have fewer fields than some of the records in the sample
-  document, provided that the sample also contains records that do not have the field 
+  document, provided that the sample also contains records that do not have the field.
 \item[--] When a labelled top type is inferred from the sample, the actual input can also contain any other value,
-  which implements the open world assumption  
+  which implements the open world assumption.
 \end{itemize}
@@
 \begin{proof}
-By induction over the structure of $\sem{-}$. For primitives, the conversion functions accept all subtypes.
-For other cases, analyze the provided code to see that it can work on all subtypes (for example~\ident{convElements}
+By induction over the structure of $\sem{-}$. For primitives, the conversion functions accept all subshapes.
+For other cases, analyze the provided code to see that it can work on all subshapes (for example~\ident{convElements}
 works on \kvd{null} values, \ident{convFloat} accepts an integer). Finally, for labelled top types,
@@
 The F\# Data library has been widely adopted by users and is one of the most downloaded
-F\# libraries.\footnote{At the time of writing, the library has over 100,000 downloads on NuGet 
+F\# libraries.\footnote{At the time of writing, the library has over 125,000 downloads on NuGet
 (package repository), 1,844 commits and 44 contributors on GitHub.} A practical demonstration of
@@
 The \emph{relative safety} property does not guarantee safety in the same way as traditional
-closed-world type safety, but it reflects the reality that programming with external data is
-increasingly important \cite{age-of-web}. Type providers increase the safety of this kind of 
+closed-world type safety, but it reflects the reality of programming with external data that is
+becoming increasingly important \cite{age-of-web}. Type providers increase the safety of this kind of
 programming.
@@
 \paragraph{Richer data sources.}
-In general, XML, CSV and JSON data sources without explicit schema will necessarily require 
-techniques akin to those we have shown. However, some data sources provide explicit schema with 
-schema versioning support. For those, a type provider could be written to adapt automatically, 
+In general, XML, CSV and JSON data sources without an explicit schema will necessarily require
+techniques akin to those we have shown. However, some data sources provide an explicit schema with
+versioning support. For those, a type provider that adapts automatically could be written,
 but we leave this for future work.
@@
 names for records (for XML nodes), the definition of structural values is rich enough to capture all
-three formats\footnote{The same mechanism has later been used by the HTML type provider 
+three formats.\footnote{The same mechanism has later been used by the HTML type provider
 (\url{http://fsharp.github.io/FSharp.Data/HtmlProvider.html}), which provides similarly easy
-access to data in HTML tables and lists.}. However, parsing real-world data poses a number of practical issues.
+access to data in HTML tables and lists.} However, parsing real-world data poses a number of practical issues.
 
@@
 create a record. Attributes become record fields and the body becomes a field with a special
-name:
+name. For example:
 %
@@
 underlying JSON node (called \ident{JsonValue}) or XML element (called \ident{XElement}).
-Those return standard .NET or F\# representation of the value and can be used to dynamically 
+Those return the standard .NET or F\# representation of the value and can be used to dynamically
 access data not exposed by the type providers, such as textual values inside mixed-content XML elements.
@@
 collections store multiple possible element shapes together with their \emph{  inferred multiplicity}
-(zero or one, exactly one, zero or more):
+(exactly one, zero or one, zero or more):
 %
@@
 For this reason, we keep the algorithm simple. For example, we do not use probabilistic methods to
-asses similarity of record types, because a small change in the sample could cause a large change 
+assess the similarity of record types, because a small change in the sample could cause a large change
 in the provided types.
@@
 
-Such $e'$ is obtained by applying one or more transformations to sub-expressions of $e$
-that transform:
+Such $e'$ is obtained by transforming sub-expressions of $e$ using one of the following
+translation rules:
 %
@@
 \item[--] Primitive shapes can become nullable (i), \ident{int} can become \ident{float} (iii)
-  or can become a part of labelled top type (ii).
+  or become a part of a labelled top type (ii).
 \item[--] Record shape fields can change shape (recursively) and record may become a part
-  of labelled top type (ii).
+  of a labelled top type (ii).
 \item[--] For list and nullable shapes, the shape of the value may change (we apply the
   transformations recursively).
-\item[--] For the any shape, the original code will continue to work (the label is never removed).
+\item[--] For the $\kvd{any}$ shape, the original code will continue to work (none of the labels is ever removed).
 \end{itemize}
@@
 
-The work that is most similar to F\# Data is the data integration in C$\omega$ \cite{comega-xs}.
+The work that is the most similar to F\# Data is the data integration in C$\omega$ \cite{comega-xs}.
 It extends C\# language with types similar to our structural types
@@
 
-In contrast, in F\# Data 
-we have avoided extending F\# at all, and the simplicity of the Foo 
-target language shows we have avoided placing strong requirements 
-(such as structural records) on the host language. Instead we provide nominal 
-types based on structure, rather than adding an advanced
+In contrast, F\# Data type providers do not require any F\# language extensions. The simplicity
+of the Foo calculus shows we have avoided placing strong requirements on the host language. We
+provide nominal types based on the shapes, rather than adding an advanced
 system of structural types into the host language.
 
-\paragraph{Advanced type systems with meta-programming.}
+\paragraph{Advanced type systems and meta-programming.}
 A number of other advanced type system features could be used to tackle the problem discussed
 in this paper. The Ur \cite{ur} language has a rich system for working with records;
-meta-programming \cite{template-hask}, \cite{th-camlp4} and multi-stage programming \cite{multi-stage}
+meta-programming \cite{template-hask,th-camlp4} and multi-stage programming \cite{multi-stage}
 could be used to generate code for the provided types; and gradual typing \cite{gradual,gradual-js}
@@
 samples. Our inference algorithm is based on a preferred shape relation. It prefers records to
-encompass the open world assumption and support developer tooling. The algorithm is predictable, which is 
+encompass the open world assumption and support developer tooling. The inference algorithm is predictable, which is
 important as developers need to understand how changing the samples affects the resulting types.
@@
 \acks
-We would like to thank to the F\# Data contributors on GitHub and other colleagues working 
-on type providers, including Jomo Fisher, Keith Battocchi and Kenji Takeda. We are also
-grateful to anonymous reviewers of the paper for their valuable feedback.
+We thank to the F\# Data contributors on GitHub and other colleagues working
+on type providers, including Jomo Fisher, Keith Battocchi and Kenji Takeda. We are
+grateful to anonymous reviewers of the paper for their valuable feedback and to
+David Walker for shepherding of the paper.
 
@@
 
+\appendix
+
+\section{OpenWeatherMap service response}
+\label{sec:appendix-weather}
+
+The introduction uses the \ident{JsonProvider} to access weather
+information using the OpenWeatherMap service. After registering, you can access the service
+using a URL \url{http://api.openweathermap.org/data/2.5/weather} with query string parameters
+\strf{q} and \strf{APPID} representing the city name and application key. A sample response looks
+as follows:
+
+\vspace{-1em}
+{\small\begin{verbatim}
+{
+  "coord": {
+    "lon": 14.42,
+    "lat": 50.09
+  },
+  "weather": [
+    {
+      "id": 802,
+      "main": "Clouds",
+      "description": "scattered clouds",
+      "icon": "03d"
+    }
+  ],
+  "base": "cmc stations",
+  "main": {
+    "temp": 5,
+    "pressure": 1010,
+    "humidity": 100,
+    "temp_min": 5,
+    "temp_max": 5
+  },
+  "wind": { "speed": 1.5, "deg": 150 },
+  "clouds": { "all": 32 },
+  "dt": 1460700000,
+  "sys": {
+    "type": 1,
+    "id": 5889,
+    "message": 0.0033,
+    "country": "CZ",
+    "sunrise": 1460693287,
+    "sunset": 1460743037
+  },
+  "id": 3067696,
+  "name": "Prague",
+  "cod": 200
+}
+\end{verbatim}}
+
 \end{document}

### 2016-05-03 pldi-2016/paper.tex
Required publication changes for PLDI 2016

@@
-\documentclass[10pt]{sigplanconf}
+\documentclass[pldi-cameraready]{sigplanconf-pldi16}
+
+\RequirePackage[loading]{tracefnt}
 
@@
 \usepackage[hyphens]{url}
-\usepackage[svgnames]{xcolor}
-\definecolor{lcl}{RGB}{140,0,100}
-\usepackage[colorlinks=true,breaklinks,draft=false]{hyperref}
-\hypersetup{urlcolor=lcl,linkcolor=lcl,citecolor=lcl}
 \newcommand{\doi}[1]{doi:~\href{http://dx.doi.org/#1}{\Hurl{#1}}}
@@
 \usepackage{ftnright}
-\usepackage[T1]{fontenc}
 \usepackage{semantic}
@@
 % =================================================================================================
+
+\CopyrightYear{2016}
+%\setcopyright{acmlicensed}
+\conferenceinfo{PLDI '16,}{June 13 - 17, 2016, Santa Barbara, CA, USA}
+\copyrightdata{978-1-4503-4261-2/16/06}
+%\acmPrice{\$15.00}
+\copyrightdoi{2908080.2908115}
 
@@
 
-\special{papersize=8.5in,11in}
-\setlength{\pdfpageheight}{\paperheight}
-\setlength{\pdfpagewidth}{\paperwidth}
-\conferenceinfo{PLDI '16}{June 13--17, 2016, Santa Barbara, CA, United States}
-\copyrightyear{2016}
-\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
-%\doi{nnnnnnn.nnnnnnn}
-
 %\titlebanner{Unpublished draft, March 2015}        % These are ignored unless
@@
 
-\title{Types from data: \textnormal{Making structured data first-class citizens in F\#}}
+\title{Types from Data: Making Structured Data First-Class Citizens in F\#}
 %\subtitle{Subtitle Text, if any}
@@
 \authorinfo{Tomas Petricek}
-           {University of Cambridge}
+           {University of Cambridge, UK}
            {tomas@tomasp.net}
 \authorinfo{Gustavo Guerra}
-           {Microsoft Corporation, London}
+           {Microsoft Corporation, UK}
            {gustavo@codebeside.org}
 \authorinfo{Don Syme}
-           {Microsoft Research, Cambridge}
+           {Microsoft Research Cambridge, UK}
            {dsyme@microsoft.com}
@@
 \category{D.3.3}{Programming Languages}{Language Constructs and Features }
+\terms{Languages, Theory}
 \keywords F\#, Type Providers, Inference, JSON, XML
@@
 in Appendix~\ref{sec:appendix-weather} and can be used to run the code against a local file.}:
-
-\noindent
+%
 \begin{equation*}
@@
 %
-\vspace{-0.1em}
 \begin{equation*}
@@
 
-\section{Type providers for structured data}
+\section{Type Providers for Structured Data}
 \label{sec:providers}
@@
 
-\subsection{Working with JSON documents}
+\subsection{Working with JSON Documents}
 \label{sec:providers-json}
@@
 
-\subsection{Processing XML documents}
+\subsection{Processing XML Documents}
 \label{sec:providers-xml}
@@
 
-\subsection{Real-world JSON services}
+\subsection{Real-world JSON Services}
 \label{sec:providers-sum}
@@
 
-\section{Shape inference for structured data}
+\section{Shape Inference for Structured Data}
 \label{sec:inference}
@@
 
-\subsection{Inferred shapes}
+\subsection{Inferred Shapes}
 \label{sec:inference-types}
@@
 \begin{center}
-\includegraphics[scale=0.80,trim=5mm 5mm 5mm 5mm,clip]{images/hierarchy.pdf} % left bottom right top
+% \includegraphics[scale=0.80,trim=5mm 5mm 5mm 5mm,clip]{images/hierarchy.pdf} % left bottom right top
+\includegraphics[scale=0.80,clip]{images/hierarchy.png} % left bottom right top
 \end{center}
@@
 
-\subsection{Preferred shape relation}
+\subsection{Preferred Shape Relation}
 \label{sec:inference-subtyping}
@@
 
-\subsection{Common preferred shape relation}
+\subsection{Common Preferred Shape Relation}
 \label{sec:inference-commonsuper}
@@
 
-\subsection{Inferring shapes from samples}
+\subsection{Inferring Shapes from Samples}
 \label{sec:formal-inferval}
@@
 
+\begin{figure*}
+  \noindent
+  \begin{equation*}
+  \begin{array}{rclcl}
+   \tytag &\narrow{=}& \ident{collection}  &\narrow{\lsep}& \ident{number} \\
+          &\narrow{\lsep}& \ident{nullable} &\narrow{\lsep}& \ident{string}  \\
+          &\narrow{\lsep}& \nu ~\lsep \ident{any} &\narrow{\lsep}& \ident{bool}
+  \end{array}
+  %
+  \quad\;\;\;
+  %
+  \begin{array}{rcl}
+   \tytagof(\ident{string}) &\narrow{=}& \ident{string}\\
+   \tytagof(\ident{bool}) &\narrow{=}& \ident{bool}\\
+   \tytagof(\ident{int}) &\narrow{=}& \ident{number}\\
+   \tytagof(\ident{float}) &\narrow{=}& \ident{number}\\
+  \end{array}
+  %
+  \;\;
+  %
+  \begin{array}{rcl}
+   \tytagof(\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl) &\narrow{=}& \ident{any}\\
+   \tytagof(\nu\; \{ \nu_1 : \sigma_1, \; \ldots \; , \nu_n : \sigma_n \}) &\narrow{=}& \nu \\
+   \tytagof(\kvd{nullable}\langl\hat{\sigma}\rangl) &\narrow{=}& \ident{nullable}\\
+   \tytagof([\sigma]) &\narrow{=}& \ident{collection}\\
+  \end{array}
+  \end{equation*}
+
+  \vspace{-0.5em}
+
+  \begin{equation*}
+  \qquad\qquad
+  \begin{array}{ll}
+  \begin{array}{l}
+    \ident{csh}(\kvd{any}\langl \sigma_1, \ldots, \sigma_k,  \ldots, \sigma_n\rangl,
+      \kvd{any}\langl \sigma'_1, \ldots, \sigma'_k, \ldots, \sigma'_m\rangl) =\\
+      \qquad \kvd{any}\langl \ident{csh}(\sigma_1, \sigma_1'), \ldots, \ident{csh}(\sigma_k, \sigma_k'),
+        \sigma_{k+1}, \ldots, \sigma_{n}, \sigma'_{k+1}, \ldots, \sigma'_{m}\rangl \\[0.5em]
+      \qquad \textnormal{For $i,j$ such that }(\tytagof(\sigma_i) = \tytagof(\sigma'_j)) \Leftrightarrow (i = j) \wedge (i \leq k)
+  \end{array}\quad~
+  &  (\emph{top-merge})
+  \\[2.5em]
+  \begin{array}{l}
+    \ident{csh}(\sigma, \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl) =
+    \ident{csh}(\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl, \sigma) = \\
+    \qquad \kvd{any}\langl\sigma_1, \ldots, \dropopt{\ident{csh}(\sigma, \sigma_i)}, \ldots, \sigma_n\rangl\\[0.5em]
+    \qquad \textnormal{For $i$ such that }\tytagof(\sigma_i) = \tytagof(\dropopt{\sigma})
+  \end{array}
+  &  (\emph{top-incl})
+  \\[2.5em]
+  ~\;\ident{csh}(\sigma, \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl) =
+   \kvd{any}\langl\sigma_1, \ldots, \sigma_n, \dropopt{\sigma}\rangl
+   & (\emph{top-add})\\[0.5em]
+  ~\;\ident{csh}(\sigma_1,\sigma_2) =
+  \kvd{any}\langle\dropopt{\sigma_1}, \dropopt{\sigma_2}\rangle
+  & (\emph{top-any})
+ \end{array}
+\end{equation*}
+
+\caption{Extending the common preferred shape relation for labelled top shapes}
+\label{fig:subtyping-cst-var}
+\end{figure*}
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{Adding Labelled Top Shapes}
+\label{sec:inference-vars}
+
+When analyzing the structure of shapes, it suffices to consider a single top shape \kvd{any}.
+The type providers need more information to provide typed access to the possible
+alternative shapes of data, such as XML nodes.
+
+We extend the core model (sufficient for the discussion of relative safety) with \emph{labelled top
+shapes} defined as:
+%
+\begin{equation*}
+\sigma = \ldots \lsep \kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl
+\end{equation*}
+%
+The shapes $\sigma_1, \ldots, \sigma_n$ represent statically known shapes that appear in the
+sample and that we expose in the provided type. As discussed earlier (\S\ref{sec:providers-xml})
+this is important when reading external \emph{open world} data. The labels do not affect the
+preferred shape relation and $\kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl$ should still be
+seen as the top shape, regardless of the labels\footnote{An alternative would be to add unions of
+shapes, but doing so in a way that is compatible with the open-world assumption breaks the
+existence of unique lower bound of the preferred shape relation.}.
+
+The common preferred shape function is extended to find a labelled top shape that best represents
+the sample. The new rules for \kvd{any} appear in Figure~\ref{fig:subtyping-cst-var}.
+We define shape \emph{tags} to identify shapes that have a common preferred shape
+which is not the top shape. We use it to limit the number of labels and avoid nesting by grouping
+shapes by the shape tag. Rather than inferring $\kvd{any}\langl\ident{int}, \kvd{any}\langl\ident{bool}, \ident{float}\rangl\rangl$,
+our algorithm joins $\ident{int}$ and $\ident{float}$ and produces
+$\kvd{any}\langl\ident{float},\ident{bool}\rangl$.
+
+When combining two top shapes (\emph{top-merge}), we group the annotations by their tags.
+When combining a top with another shape, the labels may or may not already contain a case with
+the tag of the other shape. If they do, the two shapes are combined (\emph{top-incl}), otherwise
+a new case is added (\emph{top-add}). Finally, (\emph{top-all}) replaces earlier (\emph{any})
+and combines two distinct non-top shapes. As top shapes implicitly permit \kvd{null} values,
+we make the labels non-nullable using $\dropopt{-}$.
+
+The revised algorithm still finds a shape which is the least upper bound. This means that
+labelled top shape is only inferred when there is no other alternative.
+
+Stating properties of the labels requires refinements to the \emph{preferred shape} relation.
+We leave the details to future work, but we
+note that the algorithm infers the best labels in the sense that there are labels that enable
+typed access to every possible value in the sample, but not more. The same is the case for nullable
+fields of records.
+
+% ==================================================================================================
+%
+%    #######
+%    #        ####  #####  #    #   ##   #      # ######   ##   ##### #  ####  #    #
+%    #       #    # #    # ##  ##  #  #  #      #     #   #  #    #   # #    # ##   #
+%    #####   #    # #    # # ## # #    # #      #    #   #    #   #   # #    # # #  #
+%    #       #    # #####  #    # ###### #      #   #    ######   #   # #    # #  # #
+%    #       #    # #   #  #    # #    # #      #  #     #    #   #   # #    # #   ##
+%    #        ####  #    # #    # #    # ###### # ###### #    #   #   #  ####  #    #
+%
+% ==================================================================================================
+
+\section{Formalizing Type Providers}
+\label{sec:formal}
+
+This section presents the formal model of F\# Data integration. To represent the programming
+language that hosts the type provider, we introduce the Foo calculus, a subset of F\# with objects
+and properties, extended with operations for working with weakly typed structured data along the
+lines of the F\# Data runtime. Finally, we describe how
+type providers turn inferred shapes into Foo classes (\S\ref{sec:formal-tp}).
+
+% --------------------------------------------------------------------------------------------------
+
 \begin{figure}[!h]
@@
 \noindent
-\vspace{-0.5em}
 \begin{equation*}
@@
 \label{fig:shape-inference}
-\vspace{-1em}
 \end{figure}
-
-% --------------------------------------------------------------------------------------------------
-
-\begin{figure*}
-  \noindent
-  \begin{equation*}
-  \begin{array}{rclcl}
-   \tytag &\narrow{=}& \ident{collection}  &\narrow{\lsep}& \ident{number} \\
-          &\narrow{\lsep}& \ident{nullable} &\narrow{\lsep}& \ident{string}  \\
-          &\narrow{\lsep}& \nu ~\lsep \ident{any} &\narrow{\lsep}& \ident{bool}
-  \end{array}
-  %
-  \quad\;\;\;
-  %
-  \begin{array}{rcl}
-   \tytagof(\ident{string}) &\narrow{=}& \ident{string}\\
-   \tytagof(\ident{bool}) &\narrow{=}& \ident{bool}\\
-   \tytagof(\ident{int}) &\narrow{=}& \ident{number}\\
-   \tytagof(\ident{float}) &\narrow{=}& \ident{number}\\
-  \end{array}
-  %
-  \;\;
-  %
-  \begin{array}{rcl}
-   \tytagof(\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl) &\narrow{=}& \ident{any}\\
-   \tytagof(\nu\; \{ \nu_1 : \sigma_1, \; \ldots \; , \nu_n : \sigma_n \}) &\narrow{=}& \nu \\
-   \tytagof(\kvd{nullable}\langl\hat{\sigma}\rangl) &\narrow{=}& \ident{nullable}\\
-   \tytagof([\sigma]) &\narrow{=}& \ident{collection}\\
-  \end{array}
-  \end{equation*}
-
-  \vspace{-0.5em}
-
-  \begin{equation*}
-  \qquad\qquad
-  \begin{array}{ll}
-  \begin{array}{l}
-    \ident{csh}(\kvd{any}\langl \sigma_1, \ldots, \sigma_k,  \ldots, \sigma_n\rangl,
-      \kvd{any}\langl \sigma'_1, \ldots, \sigma'_k, \ldots, \sigma'_m\rangl) =\\
-      \qquad \kvd{any}\langl \ident{csh}(\sigma_1, \sigma_1'), \ldots, \ident{csh}(\sigma_k, \sigma_k'),
-        \sigma_{k+1}, \ldots, \sigma_{n}, \sigma'_{k+1}, \ldots, \sigma'_{m}\rangl \\[0.5em]
-      \qquad \textnormal{For $i,j$ such that }(\tytagof(\sigma_i) = \tytagof(\sigma'_j)) \Leftrightarrow (i = j) \wedge (i \leq k)
-  \end{array}\quad~
-  &  (\emph{top-merge})
-  \\[2.5em]
-  \begin{array}{l}
-    \ident{csh}(\sigma, \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl) =
-    \ident{csh}(\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl, \sigma) = \\
-    \qquad \kvd{any}\langl\sigma_1, \ldots, \dropopt{\ident{csh}(\sigma, \sigma_i)}, \ldots, \sigma_n\rangl\\[0.5em]
-    \qquad \textnormal{For $i$ such that }\tytagof(\sigma_i) = \tytagof(\dropopt{\sigma})
-  \end{array}
-  &  (\emph{top-incl})
-  \\[2.5em]
-  ~\;\ident{csh}(\sigma, \kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl) =
-   \kvd{any}\langl\sigma_1, \ldots, \sigma_n, \dropopt{\sigma}\rangl
-   & (\emph{top-add})\\[0.5em]
-  ~\;\ident{csh}(\sigma_1,\sigma_2) =
-  \kvd{any}\langle\dropopt{\sigma_1}, \dropopt{\sigma_2}\rangle
-  & (\emph{top-any})
- \end{array}
-\end{equation*}
-
-\caption{Extending the common preferred shape relation for labelled top shapes}
-\label{fig:subtyping-cst-var}
-\end{figure*}
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{Adding labelled top shapes}
-\label{sec:inference-vars}
-
-When analyzing the structure of shapes, it suffices to consider a single top shape \kvd{any}.
-The type providers need more information to provide typed access to the possible
-alternative shapes of data, such as XML nodes.
-
-We extend the core model (sufficient for the discussion of relative safety) with \emph{labelled top
-shapes} defined as:
-%
-\begin{equation*}
-\sigma = \ldots \lsep \kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl
-\end{equation*}
-%
-The shapes $\sigma_1, \ldots, \sigma_n$ represent statically known shapes that appear in the
-sample and that we expose in the provided type. As discussed earlier (\S\ref{sec:providers-xml})
-this is important when reading external \emph{open world} data. The labels do not affect the
-preferred shape relation and $\kvd{any}\langl \sigma_1, \ldots, \sigma_n\rangl$ should still be
-seen as the top shape, regardless of the labels\footnote{An alternative would be to add unions of
-shapes, but doing so in a way that is compatible with the open-world assumption breaks the
-existence of unique lower bound of the preferred shape relation.}.
-
-The common preferred shape function is extended to find a labelled top shape that best represents
-the sample. The new rules for \kvd{any} appear in Figure~\ref{fig:subtyping-cst-var}.
-We define shape \emph{tags} to identify shapes that have a common preferred shape
-which is not the top shape. We use it to limit the number of labels and avoid nesting by grouping
-shapes by the shape tag. Rather than inferring $\kvd{any}\langl\ident{int}, \kvd{any}\langl\ident{bool}, \ident{float}\rangl\rangl$,
-our algorithm joins $\ident{int}$ and $\ident{float}$ and produces
-$\kvd{any}\langl\ident{float},\ident{bool}\rangl$.
-
-When combining two top shapes (\emph{top-merge}), we group the annotations by their tags.
-When combining a top with another shape, the labels may or may not already contain a case with
-the tag of the other shape. If they do, the two shapes are combined (\emph{top-incl}), otherwise
-a new case is added (\emph{top-add}). Finally, (\emph{top-all}) replaces earlier (\emph{any})
-and combines two distinct non-top shapes. As top shapes implicitly permit \kvd{null} values,
-we make the labels non-nullable using $\dropopt{-}$.
-
-The revised algorithm still finds a shape which is the least upper bound. This means that
-labelled top shape is only inferred when there is no other alternative.
-
-Stating properties of the labels requires refinements to the \emph{preferred shape} relation.
-We leave the details to future work, but we
-note that the algorithm infers the best labels in the sense that there are labels that enable
-typed access to every possible value in the sample, but not more. The same is the case for nullable
-fields of records.
-
-% ==================================================================================================
-%
-%    #######
-%    #        ####  #####  #    #   ##   #      # ######   ##   ##### #  ####  #    #
-%    #       #    # #    # ##  ##  #  #  #      #     #   #  #    #   # #    # ##   #
-%    #####   #    # #    # # ## # #    # #      #    #   #    #   #   # #    # # #  #
-%    #       #    # #####  #    # ###### #      #   #    ######   #   # #    # #  # #
-%    #       #    # #   #  #    # #    # #      #  #     #    #   #   # #    # #   ##
-%    #        ####  #    # #    # #    # ###### # ###### #    #   #   #  ####  #    #
-%
-% ==================================================================================================
-
-\section{Formalizing type providers}
-\label{sec:formal}
-
-This section presents the formal model of F\# Data integration. To represent the programming
-language that hosts the type provider, we introduce the Foo calculus, a subset of F\# with objects
-and properties, extended with operations for working with weakly typed structured data along the
-lines of the F\# Data runtime. Finally, we describe how
-type providers turn inferred shapes into Foo classes (\S\ref{sec:formal-tp}).
 
@@
 \end{figure}
+
+% -------------------------------------------------------------------------------------------------
+
+Type providers for structured data map the ``dirty'' world of weakly typed
+structured data into a ``nice'' world of strong types. To model this, the Foo calculus
+does not have \kvd{null} values and data values $d$ are never directly exposed.
+Furthermore Foo is simply typed: despite using class types and object notation
+for notational convenience, it has no subtyping.
+
+% --------------------------------------------------------------------------------------------------
+
+\subsection{The Foo Calculus}
+\label{sec:formal-ff}
+
+The syntax of the calculus is shown in Figure~\ref{fig:foo-syntax}.
+The type \ident{Data} is the type of structural data $d$. A class definition $L$ consists of
+a single constructor and zero or more parameter-less members. The declaration implicitly closes
+over the constructor parameters.
+Values $v$ include previously defined data $d$; expressions
+$e$ include class construction, member access, usual functional constructs (functions,
+lists, options) and conditionals. The $op$ constructs are discussed next.
+
+% -------------------------------------------------------------------------------------------------
+
+\paragraph{Dynamic data operations.}
+
+The Foo programs can only work with \ident{Data} values using certain primitive operations. Those are modelled
+by the $op$ primitives. In F\# Data, those are internal and users never access them directly.
+
+The behaviour of the dynamic data operations is defined by the reduction rules in
+Figure~\ref{fig:ff-reduction} (Part I). The typing is shown in Figure~\ref{fig:ff-typecheck} and is
+discussed later. The \ident{hasShape} function represents a runtime shape test.
+It checks whether a \ident{Data} value $d$ (Section~\ref{sec:formal-inferval}) passed as the second
+argument has a shape specified by the first argument. For records, we have to check that for each
+field $\nu_1, \ldots, \nu_n$ in the record, the actual record value has a field of the same name
+with a matching shape. The last line defines a ``catch all'' pattern, which returns \kvd{false}
+for all remaining cases. We treat $e_1 \vee e_2$ and $e_1 \wedge e_2$ as a syntactic sugar for
+\kvd{if}\,.\,.\,\kvd{then}\,.\,.\,\kvd{else} so the result of the reduction is just a Foo
+expression.
+
+The remaining operations convert data values into values of less preferred shape.
+The \ident{convPrim} and \ident{convFloat} operations take the required shape and a data value.
+When the data does not match the required type, they do not reduce. For example,
+$\ident{convPrim}(\kvd{bool}, \num{42})$ represents a stuck state, but $\ident{convFloat}(\kvd{float}, \num{42})$
+turns an integer \num{42} into a floating-point numerical value \num{42.0}.
+
+The \ident{convNull}, \ident{convElements} and \ident{convField} operations take an additional
+parameter $e$ which represents a function to be used in order to convert a contained value (non-null
+optional value, list elements or field value); \ident{convNull} turns \kvd{null} data value into
+\ident{None} and \ident{convElements} turns a data collection $[d_1, \ldots, d_n]$ into a Foo list
+$v_1 :: \ldots :: v_n :: \kvd{nil}$ and a \kvd{null} value into an empty list.
 
@@
 
-Type providers for structured data map the ``dirty'' world of weakly typed
-structured data into a ``nice'' world of strong types. To model this, the Foo calculus
-does not have \kvd{null} values and data values $d$ are never directly exposed.
-Furthermore Foo is simply typed: despite using class types and object notation
-for notational convenience, it has no subtyping.
-
-% --------------------------------------------------------------------------------------------------
-
-\subsection{The Foo calculus}
-\label{sec:formal-ff}
-
-The syntax of the calculus is shown in Figure~\ref{fig:foo-syntax}.
-The type \ident{Data} is the type of structural data $d$. A class definition $L$ consists of
-a single constructor and zero or more parameter-less members. The declaration implicitly closes
-over the constructor parameters.
-Values $v$ include previously defined data $d$; expressions
-$e$ include class construction, member access, usual functional constructs (functions,
-lists, options) and conditionals. The $op$ constructs are discussed next.
-
-% -------------------------------------------------------------------------------------------------
-
-\paragraph{Dynamic data operations.}
-
-The Foo programs can only work with \ident{Data} values using certain primitive operations. Those are modelled
-by the $op$ primitives. In F\# Data, those are internal and users never access them directly.
-
-The behaviour of the dynamic data operations is defined by the reduction rules in
-Figure~\ref{fig:ff-reduction} (Part I). The typing is shown in Figure~\ref{fig:ff-typecheck} and is
-discussed later. The \ident{hasShape} function represents a runtime shape test.
-It checks whether a \ident{Data} value $d$ (Section~\ref{sec:formal-inferval}) passed as the second
-argument has a shape specified by the first argument. For records, we have to check that for each
-field $\nu_1, \ldots, \nu_n$ in the record, the actual record value has a field of the same name
-with a matching shape. The last line defines a ``catch all'' pattern, which returns \kvd{false}
-for all remaining cases. We treat $e_1 \vee e_2$ and $e_1 \wedge e_2$ as a syntactic sugar for
-\kvd{if}\,.\,.\,\kvd{then}\,.\,.\,\kvd{else} so the result of the reduction is just a Foo
-expression.
-
-The remaining operations convert data values into values of less preferred shape.
-The \ident{convPrim} and \ident{convFloat} operations take the required shape and a data value.
-When the data does not match the required type, they do not reduce. For example,
-$\ident{convPrim}(\kvd{bool}, \num{42})$ represents a stuck state, but $\ident{convFloat}(\kvd{float}, \num{42})$
-turns an integer \num{42} into a floating-point numerical value \num{42.0}.
-
-The \ident{convNull}, \ident{convElements} and \ident{convField} operations take an additional
-parameter $e$ which represents a function to be used in order to convert a contained value (non-null
-optional value, list elements or field value); \ident{convNull} turns \kvd{null} data value into
-\ident{None} and \ident{convElements} turns a data collection $[d_1, \ldots, d_n]$ into a Foo list
-$v_1 :: \ldots :: v_n :: \kvd{nil}$ and a \kvd{null} value into an empty list.
+\paragraph{Reduction.}
+The reduction relation is of the form $L, e \reduce e'$. We omit class declarations
+$L$ where implied by the context and write $e \reduce^{*} e'$ for
+the reflexive, transitive closure of $\reduce$.
+
+Figure~\ref{fig:ff-reduction} (Part II) shows the reduction rules.  The (\emph{member}) rule reduces
+a member access using a class definition in the assumption. The (\emph{ctx}) rule models the eager
+evaluation of F\# and performs a reduction inside a sub-expression specified by an evaluation
+context $E$:
+%
+\begin{equation*}
+\begin{array}{rcl}
+ E &\narrow{=}& v::E \lsep v~E \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e})\\[0.1em]
+   &\narrow{|}&  \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2  \lsep E = e \lsep v = E \\[0.1em]
+   &\narrow{|}& \ident{Some}(E) \lsep op(\overline{v}, E, \overline{e})\\[0.1em]
+   &\narrow{|}& \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
+   &\narrow{|}& \kvd{match}~E~\kvd{with}~x_1 :: x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
+\end{array}
+\end{equation*}
+
+\noindent
+The evaluation proceeds from left to right as denoted by $\overline{v}, E, \overline{e}$ in
+constructor and dynamic data operation arguments or $v::E$ in list initialization.
+We write $e[\overline{x} \leftarrow \overline{v}]$ for the result of replacing variables $\overline{x}$ by
+values $\overline{v}$ in an expression. The remaining six rules
+give standard reductions.
+
+% -------------------------------------------------------------------------------------------------
+
+\paragraph{Type checking.}
+Well-typed Foo programs reduce to a value in a finite number of steps or get stuck due to an
+error condition. The stuck states can only be due to the dynamic data operations (e.g. an attempt
+to convert \kvd{null} value to a number $\ident{convFloat}(\ident{float},\kvd{null})$). The relative safety (Theorem~\ref{thm:safety})
+characterizes the additional conditions on input data under which Foo programs do not get stuck.
+
+Typing rules in Figure~\ref{fig:ff-typecheck} are written using a judgement
+$L; \Gamma \vdash e : \tau$ where the context also contains a set of class declarations $L$.
+The fragment demonstrates the differences and similarities with Featherweight Java \cite{fwjava} and
+typing rules for the dynamic data operations $op$:
+%
+\begin{itemize}
+\item[--] All data values $d$ have the type \ident{Data}, but primitive data values (Booleans,
+  strings, integers and floats) can be implicitly converted to Foo values and so they also have a
+  primitive type as illustrated by the rule for $i$ and $f$.
+
+\item[--] For non-primitive data values (including \kvd{null}, data collections and records),
+  \ident{Data} is the only type.
+
+\item[--] Operations $op$ accept \ident{Data} as one of the arguments and produce a non-\ident{Data}
+  Foo type. Some of them require a function specifying the conversion for nested values.
+
+\item[--] Rules for checking class construction and member access are similar to corresponding
+  rules of Featherweight Java.
+\end{itemize}
+%
+An important part of Featherweight Java that is omitted here is the checking of type declarations
+(ensuring the members are well-typed). We consider only classes generated by our type providers
+and those are well-typed by construction.
+
+% -------------------------------------------------------------------------------------------------
+
+\subsection{Type Providers}
+\label{sec:formal-tp}
+
+So far, we defined the type inference algorithm which produces a shape $\sigma$ from one
+or more sample documents (\S\ref{sec:inference}) and we defined a simplified model of evaluation
+of F\# (\S\ref{sec:formal-ff}) and F\# Data runtime (\S\ref{sec:formal-tp}). In this section, we
+define how the type providers work, linking the two parts.
+
+All F\# Data type providers take (one or more) sample documents, infer a common preferred shape $\sigma$
+and then use it to generate F\# types that are exposed to the programmer.\footnote{The actual
+implementation provides \emph{erased types} as described in \cite{fsharp-typeprov}. Here, we treat
+the code as actually generated. This is an acceptable simplification, because F\# Data type providers
+do not rely on laziness or erasure of type provision.}
+
+\paragraph{Type provider mapping.}
+A type provider produces an F\# type $\tau$ together with a Foo expression and a collection of
+class definitions. We express it using the following mapping:
+%
+\begin{equation*}
+\sem{\sigma} = (\tau, e, L) \qquad (\textnormal{where}~L,\emptyset \vdash e:\ident{Data}\rightarrow\tau)
+\end{equation*}
+
+% -------------------------------------------------------------------------------------------------
+
+\noindent
+The mapping $\sem{\sigma}$ takes an inferred shape $\sigma$. It returns an F\# type $\tau$ and
+a function that turns the input data (value of type \ident{Data}) into a Foo value of type $\tau$.
+The type provider also generates class definitions that may be used by $e$.
+
+Figure~\ref{fig:tp-generation} defines $\sem{-}$. Primitive types are handled by a single rule that
+inserts an appropriate conversion function; \ident{convPrim} just checks that the shape matches
+and \ident{convFloat} converts numbers to a floating-point.
+
+% -------------------------------------------------------------------------------------------------
 
@@
 
-\paragraph{Reduction.}
-The reduction relation is of the form $L, e \reduce e'$. We omit class declarations
-$L$ where implied by the context and write $e \reduce^{*} e'$ for
-the reflexive, transitive closure of $\reduce$.
-
-Figure~\ref{fig:ff-reduction} (Part II) shows the reduction rules.  The (\emph{member}) rule reduces
-a member access using a class definition in the assumption. The (\emph{ctx}) rule models the eager
-evaluation of F\# and performs a reduction inside a sub-expression specified by an evaluation
-context $E$:
-%
-\begin{equation*}
-\begin{array}{rcl}
- E &\narrow{=}& v::E \lsep v~E \lsep E.N \lsep \kvd{new}~C(\overline{v}, E, \overline{e})\\[0.1em]
-   &\narrow{|}&  \kvd{if}~E~\kvd{then}~e_1~\kvd{else}~e_2  \lsep E = e \lsep v = E \\[0.1em]
-   &\narrow{|}& \ident{Some}(E) \lsep op(\overline{v}, E, \overline{e})\\[0.1em]
-   &\narrow{|}& \kvd{match}~E~\kvd{with}~\ident{Some}(x) \rightarrow e_1 \,|\, \ident{None} \rightarrow e_2 \\[0.1em]
-   &\narrow{|}& \kvd{match}~E~\kvd{with}~x_1 :: x_2 \rightarrow e_1 \,|\, \kvd{nil} \rightarrow e_2
-\end{array}
-\end{equation*}
-
-\noindent
-The evaluation proceeds from left to right as denoted by $\overline{v}, E, \overline{e}$ in
-constructor and dynamic data operation arguments or $v::E$ in list initialization.
-We write $e[\overline{x} \leftarrow \overline{v}]$ for the result of replacing variables $\overline{x}$ by
-values $\overline{v}$ in an expression. The remaining six rules
-give standard reductions.
-
-% -------------------------------------------------------------------------------------------------
-
-\paragraph{Type checking.}
-Well-typed Foo programs reduce to a value in a finite number of steps or get stuck due to an
-error condition. The stuck states can only be due to the dynamic data operations (e.g. an attempt
-to convert \kvd{null} value to a number $\ident{convFloat}(\ident{float},\kvd{null})$). The relative safety (Theorem~\ref{thm:safety})
-characterizes the additional conditions on input data under which Foo programs do not get stuck.
-
-Typing rules in Figure~\ref{fig:ff-typecheck} are written using a judgement
-$L; \Gamma \vdash e : \tau$ where the context also contains a set of class declarations $L$.
-The fragment demonstrates the differences and similarities with Featherweight Java \cite{fwjava} and
-typing rules for the dynamic data operations $op$:
-%
-\begin{itemize}
-\item[--] All data values $d$ have the type \ident{Data}, but primitive data values (Booleans,
-  strings, integers and floats) can be implicitly converted to Foo values and so they also have a
-  primitive type as illustrated by the rule for $i$ and $f$.
-
-\item[--] For non-primitive data values (including \kvd{null}, data collections and records),
-  \ident{Data} is the only type.
-
-\item[--] Operations $op$ accept \ident{Data} as one of the arguments and produce a non-\ident{Data}
-  Foo type. Some of them require a function specifying the conversion for nested values.
-
-\item[--] Rules for checking class construction and member access are similar to corresponding
-  rules of Featherweight Java.
-\end{itemize}
-%
-An important part of Featherweight Java that is omitted here is the checking of type declarations
-(ensuring the members are well-typed). We consider only classes generated by our type providers
-and those are well-typed by construction.
-
-% -------------------------------------------------------------------------------------------------
-
-\subsection{Type providers}
-\label{sec:formal-tp}
-
-So far, we defined the type inference algorithm which produces a shape $\sigma$ from one
-or more sample documents (\S\ref{sec:inference}) and we defined a simplified model of evaluation
-of F\# (\S\ref{sec:formal-ff}) and F\# Data runtime (\S\ref{sec:formal-tp}). In this section, we
-define how the type providers work, linking the two parts.
-
-All F\# Data type providers take (one or more) sample documents, infer a common preferred shape $\sigma$
-and then use it to generate F\# types that are exposed to the programmer.\footnote{The actual
-implementation provides \emph{erased types} as described in \cite{fsharp-typeprov}. Here, we treat
-the code as actually generated. This is an acceptable simplification, because F\# Data type providers
-do not rely on laziness or erasure of type provision.}
-
-\paragraph{Type provider mapping.}
-A type provider produces an F\# type $\tau$ together with a Foo expression and a collection of
-class definitions. We express it using the following mapping:
-%
-\begin{equation*}
-\sem{\sigma} = (\tau, e, L) \qquad (\textnormal{where}~L,\emptyset \vdash e:\ident{Data}\rightarrow\tau)
-\end{equation*}
-
-% -------------------------------------------------------------------------------------------------
-
-\noindent
-The mapping $\sem{\sigma}$ takes an inferred shape $\sigma$. It returns an F\# type $\tau$ and
-a function that turns the input data (value of type \ident{Data}) into a Foo value of type $\tau$.
-The type provider also generates class definitions that may be used by $e$.
-
-Figure~\ref{fig:tp-generation} defines $\sem{-}$. Primitive types are handled by a single rule that
-inserts an appropriate conversion function; \ident{convPrim} just checks that the shape matches
-and \ident{convFloat} converts numbers to a floating-point.
-
-% -------------------------------------------------------------------------------------------------
-
-\begin{figure*}
-\begin{multicols}{2}
-
-% Primitives
-\noindent
-\begin{equation*}
-\begin{array}{l}
- \sem{\sigma_p} = \tau_p,\lambda x. op(\sigma_p, x),\emptyset\quad\textnormal{where} \\[0.6em]
- \quad\sigma_p, \tau_p, op\in  \{~ (\ident{bool}, \ident{bool}, \ident{convPrim})\\
- \hspace{2.9em} (\ident{int}, \ident{int}, \ident{convPrim}), (\ident{float},\ident{float},\ident{convFloat}),\\
- \hspace{2.9em} (\ident{string},\ident{string},\ident{convPrim}) ~\}
-\end{array}
-\end{equation*}
-\vspace{-2em}
-
-% Records
-\begin{equation*}
-\begin{array}{l}
- \sem{\,\nu~ \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}\,} = \\[0.1em]
- \quad C, \lambda x. \kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
- \qquad \;\;C~\textnormal{is a fresh class name} \\
- \qquad \;\,\,L = \kvd{type}~C(x_1\!:\!\ident{Data})=M_1 \ldots M_n \\
- \qquad M_i = \kvd{member}~\nu_i:\tau_i=\ident{convField}(\nu, \nu_i, x_1, e_i),\\
- \qquad \tau_i, e_i, L_i = \sem{\sigma_i}
-\end{array}
-\end{equation*}
-\vspace{-2em}
-
-% Collections
-\begin{equation*}
-\begin{array}{l}
- \sem{\,[\sigma]\,} = \ident{list}\langl\tau\rangl, \lambda x . \ident{convElements}(x, e'), L \;\;\textnormal{where}\\[0.4em]
- \qquad \tau, e', L = \sem{\hat{\sigma}}
-\end{array}
-\end{equation*}
-
-% Labelled top
-\noindent
-\begin{equation*}
-\begin{array}{l}
- \sem{\,\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl\,} = \\
- \quad C, \lambda x.\kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\;\;\textnormal{where}\\[0.6em]
- \qquad \;\;C~\textnormal{is a fresh class name} \\
- \qquad \;\,\,L = \kvd{type}~C(x:\ident{Data})~=~M_1\ldots M_n \\
- \qquad M_i = \kvd{member}~\nu_i:\ident{option}\langl\tau_i\rangl=\\
- \hspace{5.8em}  \kvd{if}~\ident{hasShape}(\sigma_i, x)~\kvd{then}~ \ident{Some}(e_i~x)~\kvd{else}~\ident{None} \\[0.1em]
- \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_e,\;\; \nu_i=\tytagof{(\sigma_i)}
-\end{array}
-\end{equation*}
-\vspace{-2em}
-
-% Nullables
-\begin{equation*}
-\hspace{-1.3em}
-\begin{array}{l}
- \sem{\kvd{nullable}\langl\hat{\sigma}\rangl} = \\[0.2em]
- \qquad \ident{option}\langl\tau\rangl, \lambda x . \ident{convNull}(x,e), L\\[0.2em]
- \qquad \textnormal{where}~\tau, e, L = \sem{\hat{\sigma}}
-\end{array}
-\end{equation*}
-\vspace{-2em}
-
-% Anything else
-\begin{equation*}
-\hspace{-1.3em}
-\begin{array}{l}
- \sem{\bot} = \sem{\kvd{null}} = C, \lambda x. \kvd{new}~C(x), \{ L \} \;\;\textnormal{where}\\[0.4em]
- \qquad C~\textnormal{is a fresh class name} \\[0.1em]
- \qquad L = \kvd{type}~C(v:\ident{Data})
-\end{array}
-\end{equation*}
-\end{multicols}
-
-\caption{Type provider -- generation of Foo types from inferred structural types}
-\label{fig:tp-generation}
-\vspace{-0.5em}
-\end{figure*}
-
-% -------------------------------------------------------------------------------------------------
-
 For records, we generate a class $C$ that takes a data value as a constructor parameter. For each
@@
 
+% -------------------------------------------------------------------------------------------------
+
+\begin{figure*}
+\begin{multicols}{2}
+
+% Primitives
+\noindent
+\begin{equation*}
+\begin{array}{l}
+ \sem{\sigma_p} = \tau_p,\lambda x. op(\sigma_p, x),\emptyset\quad\textnormal{where} \\[0.6em]
+ \quad\sigma_p, \tau_p, op\in  \{~ (\ident{bool}, \ident{bool}, \ident{convPrim})\\
+ \hspace{2.9em} (\ident{int}, \ident{int}, \ident{convPrim}), (\ident{float},\ident{float},\ident{convFloat}),\\
+ \hspace{2.9em} (\ident{string},\ident{string},\ident{convPrim}) ~\}
+\end{array}
+\end{equation*}
+\vspace{-2em}
+
+% Records
+\begin{equation*}
+\begin{array}{l}
+ \sem{\,\nu~ \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \}\,} = \\[0.1em]
+ \quad C, \lambda x. \kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\quad\textnormal{where}\\[0.6em]
+ \qquad \;\;C~\textnormal{is a fresh class name} \\
+ \qquad \;\,\,L = \kvd{type}~C(x_1\!:\!\ident{Data})=M_1 \ldots M_n \\
+ \qquad M_i = \kvd{member}~\nu_i:\tau_i=\ident{convField}(\nu, \nu_i, x_1, e_i),\\
+ \qquad \tau_i, e_i, L_i = \sem{\sigma_i}
+\end{array}
+\end{equation*}
+\vspace{-2em}
+
+% Collections
+\begin{equation*}
+\begin{array}{l}
+ \sem{\,[\sigma]\,} = \ident{list}\langl\tau\rangl, \lambda x . \ident{convElements}(x, e'), L \;\;\textnormal{where}\\[0.4em]
+ \qquad \tau, e', L = \sem{\hat{\sigma}}
+\end{array}
+\end{equation*}
+
+% Labelled top
+\noindent
+\begin{equation*}
+\begin{array}{l}
+ \sem{\,\kvd{any}\langl\sigma_1, \ldots, \sigma_n\rangl\,} = \\
+ \quad C, \lambda x.\kvd{new}~C(x), L_1\cup\ldots\cup L_n\cup\{ L \}\;\;\textnormal{where}\\[0.6em]
+ \qquad \;\;C~\textnormal{is a fresh class name} \\
+ \qquad \;\,\,L = \kvd{type}~C(x:\ident{Data})~=~M_1\ldots M_n \\
+ \qquad M_i = \kvd{member}~\nu_i:\ident{option}\langl\tau_i\rangl=\\
+ \hspace{5.8em}  \kvd{if}~\ident{hasShape}(\sigma_i, x)~\kvd{then}~ \ident{Some}(e_i~x)~\kvd{else}~\ident{None} \\[0.1em]
+ \qquad \tau_i, e_i, L_i = \sem{\sigma_i}_e,\;\; \nu_i=\tytagof{(\sigma_i)}
+\end{array}
+\end{equation*}
+\vspace{-2em}
+
+% Nullables
+\begin{equation*}
+\hspace{-1.3em}
+\begin{array}{l}
+ \sem{\kvd{nullable}\langl\hat{\sigma}\rangl} = \\[0.2em]
+ \qquad \ident{option}\langl\tau\rangl, \lambda x . \ident{convNull}(x,e), L\\[0.2em]
+ \qquad \textnormal{where}~\tau, e, L = \sem{\hat{\sigma}}
+\end{array}
+\end{equation*}
+\vspace{-2em}
+
+% Anything else
+\begin{equation*}
+\hspace{-1.3em}
+\begin{array}{l}
+ \sem{\bot} = \sem{\kvd{null}} = C, \lambda x. \kvd{new}~C(x), \{ L \} \;\;\textnormal{where}\\[0.4em]
+ \qquad C~\textnormal{is a fresh class name} \\[0.1em]
+ \qquad L = \kvd{type}~C(v:\ident{Data})
+\end{array}
+\end{equation*}
+\end{multicols}
+
+\caption{Type provider -- generation of Foo types from inferred structural types}
+\label{fig:tp-generation}
+\vspace{-0.5em}
+\end{figure*}
+
+% -------------------------------------------------------------------------------------------------
+
 % ==================================================================================================
@@
 
-\section{Relative type safety}
+\section{Relative Type Safety}
 \label{sec:safety}
@@
 
-\section{Practical experience}
+\section{Practical Experience}
 \label{sec:impl}
@@
 
-\subsection{Relative safety in practice}
+\subsection{Relative Safety in Practice}
 \label{sec:safety-discuss}
@@
 
-\subsection{Parsing structured data}
+\subsection{Parsing Structured Data}
 \label{sec:impl-parsing}
@@
 
-\subsection{Providing idiomatic F\# types}
+\subsection{Providing Idiomatic F\# Types}
 \label{sec:impl-naming}
@@
 
-\subsection{Heterogeneous collections}
+\subsection{Heterogeneous Collections}
 \label{sec:impl-hetero}
@@
 
-\subsection{Predictability and stability}
+\subsection{Predictability and Stability}
 \label{sec:impl-stable}
@@
 
-\section{Related and future work}
+\section{Related and Future Work}
 \label{sec:related}

### 2016-05-10 pldi-2016/paper.tex
Minor tweaks for arXiv submission

--- 2016-04-15 pldi-2016/paper.tex
@@
-\documentclass[10pt]{sigplanconf}
-
+\documentclass[10pt,nocopyrightspace]{sigplanconf}
+\pdfoutput=1
 \usepackage{listings}
@@
 \usepackage{flushend}
-\usepackage{enumerate}
 \usepackage{etoolbox}
@@
 
-\noindent
 \begin{equation*}
@@
 \begin{center}
-\includegraphics[scale=0.80,trim=5mm 5mm 5mm 5mm,clip]{images/hierarchy.pdf} % left bottom right top
+\includegraphics[scale=0.80,trim=5mm 5mm 5mm 5mm,clip]{hierarchy.pdf} % left bottom right top
 \end{center}
@@
 %
-\begin{enumerate}[(i)]
+\begin{enumerate}
 \item
@@
 \begin{itemize}
-\item[--] Primitive shapes can become nullable (i), \ident{int} can become \ident{float} (iii)
-  or become a part of a labelled top type (ii).
+\item[--] Primitive shapes can become nullable (1), \ident{int} can become \ident{float} (3)
+  or become a part of a labelled top type (2).
 \item[--] Record shape fields can change shape (recursively) and record may become a part
-  of a labelled top type (ii).
+  of a labelled top type (2).
 \item[--] For list and nullable shapes, the shape of the value may change (we apply the
